### 2021-03-05 README.Rmd
Initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas"
```

# berchukHaulout

<!-- [![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio) -->

This repository contains the data and code for our paper:

<!-- > Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx> -->

<!-- Our pre-print is online here: -->

<!-- > Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx> -->

<!-- ### How to cite -->

<!-- Please cite this compendium as: -->

<!-- > Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx> -->

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your broswer or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

The simplest way (not yet enabled) to explore the text, code and data is to click on
[binder](https://mybinder.org/v2/gh///master?urlpath=rstudio)
to open an instance of RStudio in your browser, which will have the
compendium files ready to work with. Binder uses rocker-project.org
Docker images to ensure a consistent and reproducible computational
environment. These Docker images can also be used locally.

<!-- You can download the compendium as a zip from from this URL: -->
<!-- [master.zip](/archive/master.zip). After unzipping: -->
<!-- - open the `.Rproj` file in RStudio -->
<!-- - run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the -->
<!-- [DESCRIPTION](/DESCRIPTION) file). -->
<!-- - finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console -->

### 2021-03-05 analysis/paper/paper.Rmd
Initial commit

---
title: "Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas"
author:
  - Josh M. London:
      email: josh.london@noaa.gov
      institute: [MML]
      correspondence: true
  - Paul B. Conn:
      institute: [MML]
      correspondence: false
  - Stacie K. Hardy:
      institute: [MML]
      correspondence: false
  - Erin L. Richmond:
      institute: [MML]
      correspondence: false
  - Jay M. Ver Hoef:
      institute: [MML]
      correspondence: false
  - Irina S. Trukhanova:
      institute: [MML]
      correspondence: false
  - Michael F. Cameron:
      institute: [MML]
      correspondence: false
  - Peter L. Boveng:
      institute: [MML]
      correspondence: false
institute:
  - MML: Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
output: 
  officedown::rdocx_document:
    reference_docx: ../templates/jml_template.docx
    pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
bibliography: references.json
link-citations: true
csl: ../templates/royal-society-open-science.csl
always_allow_html: true
params:
   run_models: FALSE
---

## Abstract

Ice-associated seals rely on sea ice for a variety of activities, including breeding,
molting, pupping, and resting. In the Arctic, many of these activities occur in spring and
early summer (April-June) as sea ice begins to melt and retreat northward. Rapid
acceleration of climate change in Arctic ecosystems is thus of concern, as the quantity
and quality of suitable habitat is forecast to decrease. In this paper, we use data
collected from satellite-linked telemetry tags deployed between 2005 and 2020 to
investigate the seasonal timing and environmental factors affecting sea ice use by seals
(specifically, bearded, ribbon, and spotted seals) in the Bering, and Chukchi seas. In
addition to providing baseline data on phenology, these data also allow us to refine
availability estimates needed to accurately estimate abundance from aerial survey counts
of seals basking on ice (i.e., to correct for the proportion of animals that are in the
water while surveys are conducted). Using generalized linear mixed pseudo-models to
properly account for temporal autocorrelation, we fit models with a variety of covariates
(e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure,
temperature, precipitation) to examine their ability to explain variation in hourly
haul-out records. We found evidence for strong diurnal and within season patterns in
haul-out behavior, as well as strong weather effects (particularly wind and temperature).
In general, seals were more likely to bask on ice in the middle of the day and when wind
speed was low and temperatures were higher. Haul-out propensity increased through March
and April, peaking in May and early June before declining again. The timing and frequency
of haul-out events also varied substantially based on species and age-sex class. For
ribbon and spotted seals, models with year effects were highly supported, indicating that
the timing and magnitude of haul-out behavior varied among years. Our
analysis also emphasizes the importance of accounting for factors affecting haul-out
behavior when interpreting the number of seals counted in aerial surveys.

keywords: availability; generalized linear mixed pseudo-model; haul-out behavior;
phenology; Phocidae; sea ice

```{r setup, include = FALSE}
library(tidyverse)
library(lubridate)
library(glue)
library(DBI)
library(dbplyr)
library(sf)
library(flextable)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(glmmLDTS)
library(mgcv)
library(ragg)

knitr::opts_chunk$set(dev = "ragg_png")

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  include = FALSE,
  fig.path = "../figures/",
  dpi = 400,
  fig.width = 6.5,
  out.width = "100%"
)

knitr::opts_chunk$set(
  tab.cap.style = "Table Caption",
  tab.cap.pre = "Table ",
  tab.cap.sep = ": "
  )

knitr::opts_chunk$set(
  fig.cap.style = "Image Caption",
  fig.cap.pre = "Figure ",
  fig.cap.sep = ": "
  )

theme_set(theme_minimal(base_family = 'Source Sans Pro'))
theme_replace(plot.title.position = "plot")
```

# Introduction

The global climate crisis is causing considerable reduction in Arctic sea ice extent,
volume, and seasonal presence [@kwok2018erl; @meier2014rg; @wang2017dsrpitsio]. These
changes have ripple effects on Arctic ecosystems, taxa, and the human communities who live
in the region [@huntington2020ncc]. Such disruptions are a particular cause of concern for
ice-associated seals (ringed, bearded, spotted, and ribbon seals) which depend on spring
and early summer sea ice (March-June) in the Bering and Chukchi seas as a platform for
important functions such as pupping, nursing, breeding, and molting [@boveng2009a;
@boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge gaps prevent our
ability to make definitive predictions about the ultimate effects of changes in sea ice on
the behavior, abundance, and distribution of these seals. Knowledge about biological
constraints on the phenology of reproductive and molting behavior is generally lacking, so
it is difficult to predict the readiness with which ice-associated seal species will be
able to adapt to future changes (e.g., by adjusting pupping or molting schedules to
earlier dates or different locales). Additionally, trends in abundance of these species
are largely unknown, so it is difficult to say what effect, if any, declines in sea ice
habitat have had, or will have, on seal densities.

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
of the water almost exclusively on sea ice and are mostly pelagic outside the spring
pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals do
rest on beaches during ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The very remote nature of the Bering and
Chukchi seas mean direct observation of seal behavior is limited. Bio-logging devices are,
thus, especially useful tools for collecting key information on movement and haul-out
behavior for these species. Records of time spent out of the water provide valuable data
for identifying covariates that explain variation in haul-out behavior. For instance, in
the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to
demonstrate greater haul-out propensity in juvenile crabeater seals (*Lobodon
carcinophaga*) than adults, with highest probabilities in February and at times close to
solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson et al.
[-@bengtson2005pb] documented a higher propensity for ringed seal basking near solar noon,
as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded (*Erignathus
barbatus*), ribbon (*Histriophoca fasciata*), and spotted (*Phoca largha*) seals using
much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that the proportion of time
bearded seals spent hauled out progressively increased through spring and summer. And, Ver
Hoef et al. [-@verhoef2014sm] found haul-out probabilities increased gradually starting in
March and peaking in May and June for bearded, ribbon, and spotted seals. However, their
primary focus was estimating seal abundance, and like a subsequent analysis by Conn et al.
[-@conn2014mee], they did not attempt to separately analyze haul-out schedules for
different age and sex classes (although initial models with sex and age as fixed effects
were insignificant at p = 0.05; [@verhoef2014sm]).

Several authors have used logistic regression-style analyses to estimate the probability
of sea ice use from data garnered from bio-loggers. In these models, haul-out
probabilities are expressed as a function of predictive covariates, such as time of day,
day-of-year, sex, age class, and environmental conditions (e.g., [@reder2003pb],
[@bengtson2004pb], [@bengtson2005pb], [@udevitz2009cjz], and [@verhoef2014sm]). However,
sample sizes have often been low and insufficient to permit strong inference about life
history and/or seasonal variation in haul-out probabilities. For instance, Bengtson and
Cameron's [-@bengtson2004pb] study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s [-@bengtson2005pb] study was based on 6 telemetered ringed seals (*Phoca
hispida*) in the Chukchi Sea. These studies are often further limited by logistical
constraints on tagging fieldwork and the attachment duration or operational life of
satellite tags.

Ultimately, knowledge of trends in phenology and abundance (or life history surrogates
such as survival and recruitment) will be necessary to make credible quantitative
predictions about the effects of the climate crisis on the abundance and distribution of
Arctic seal populations. Before we can construct a trend, however, we first need credible
data points. Several studies have contributed estimates of the distribution and abundance
of ice-associated seal species in the Arctic using aerial surveys (e.g.,
[@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]). Such studies are conducted over
very large areas and estimation of absolute abundance requires making inference about
numerous nuisance processes affecting the observation of seals on ice. This includes
availability (only seals basking on ice are available to be counted), detection
probability (observers or automated detection systems may miss some seals on ice), species
misclassification, and possible disturbance of seals by aircraft [@conn2014mee;
@verhoef2014sm].

In this study, we use data from fifteen years of bio-logger records in a comprehensive
investigation of haul-out behavior by three ice-associated seals in the Bering and Chukchi
Seas. We focused our effort on bearded, spotted, and ribbon seals and do not include
ringed seals the analysis. Ringed seals are unique in their use of snow lairs in the
spring months and they warrant a separate study that includes additional environmental
covariates such as the onset of snow melt and snow depth. Our goals are threefold. First,
we wish to establish baseline estimates for the chronology of haul-out behavior in the
critical spring season for each species across different age and sex classes. Second, we
wish to refine estimates of haul-out availability corrections for aerial surveys in order
to improve estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]) accounted for
variables such as the time-of-day and day-of-year, but did not investigate meteorological
variables that have been shown to influence haul-out behavior of walruses [@reder2003pb;
@udevitz2009cjz]. Further, previous correction factor analyses assumed that tagged seals
were representative of the population as a whole, even though the age and sex distribution
of telemetered individuals might differ from the population. Third, we explore the annual
variability in haul-out timing and possible linkage to dramatic changes in the
availability and extent of seasonal sea ice between 2005 and 2020. Our work extends the
scope of previous haul-out analyses, includes the influence of meteorological variability
and possible ramifications of non-representative age-sex samples, and investigates the
potential impact of a changing icescape on the behavior of these species.

# Methods

## Data collection

```{r getdata}
con <- dbConnect(
  odbc::odbc(),
  dsn = "PostgreSQL pep",
  uid = keyringr::get_kc_account("pgpep_londonj"),
  pwd = keyringr::decrypt_kc_pw("pgpep_londonj"),
  sslmode = "disable"
)

# qry removes two Sea of Okhotsk bearded seals
qry <- {
  "SELECT * 
  FROM telem.res_iceseal_haulout_cov 
  WHERE 
  EXTRACT(MONTH FROM haulout_dt) IN (3,4,5,6,7) AND
  rast_vwnd IS NOT NULL AND
  species != 'Ph' AND
  speno != 'EB2009_7010' AND
  speno != 'EB2009_3002' "
}

dat.sf <- sf::st_read(con,query = qry) %>% 
  filter(lubridate::yday(haulout_dt) < 197 ) %>% 
  mutate(age = case_when(
    age == "PUP" ~ "YOUNG OF YEAR",
    age == "YEARLING" ~ "SUBADULT",
    TRUE ~ age
  ))

deploy_meta <- tbl(con, in_schema("telem", "tbl_tag_deployments")) %>% 
  filter(speno %in% !!unique(dat.sf$speno)) %>% 
  group_by(speno) %>% 
  summarise(deploy_dt = min(deploy_dt, na.rm = TRUE)) %>% 
  collect()

dat.sf <- dat.sf %>% left_join(deploy_meta) %>% 
  mutate(bday = as.Date(paste0(lubridate::year(deploy_dt),"-03-01")),
         age_days = difftime(haulout_dt, bday, tz="UTC", units = "days"),
         age = case_when(
           age_days > 360 & age == "YOUNG OF YEAR" ~ "SUBADULT",
           TRUE ~ age
         )) %>% 
  mutate(age = forcats::fct_relevel(age, c("ADULT", "SUBADULT", "YOUNG OF YEAR"))) %>% 
  dplyr::select(-c(bday,age_days)) %>% 
  relocate(deploy_dt, .after = age) %>% 
  filter(!(species == "Bearded seal" & age == "YOUNG OF YEAR"))

# read in world land polygons and buffer in by 100m; want to ensure only locations well
world <- ne_countries(scale = "large", returnclass = "sf") %>% 
  filter(continent != "Antarctica") %>% 
  st_wrap_dateline() %>% st_transform(st_crs(dat.sf)) %>% 
  st_buffer(-100)

not_intersects <- function(x,y) {
  !sf::st_intersects(x,y)
}

dat.sf <- dat.sf %>% 
  sf::st_filter(st_combine(world), .predicate = not_intersects)

save("dat.sf", file = here::here("data/dat_sf.Rdata"))

grid.sf <- sf::st_read(con,
                       query = "SELECT * FROM base.geo_analysis_grid") %>% 
  st_filter(st_bbox(dat.sf) %>% st_as_sfc())

dbDisconnect(con)

deploy_tbl <- dat.sf %>% 
  tibble::as_tibble() %>% 
  dplyr::group_by(speno,species, sex, age) %>% 
  dplyr::summarise(min_dt = min(haulout_dt),
            max_dt = max(haulout_dt),
            n_days = difftime(max_dt, min_dt, units = "days"),
            n_hours = n()
  ) %>% 
  dplyr::select(-c("min_dt","max_dt"))

```

Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was subset to include only records from 1 March to
15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt and limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were usually preferred. Sex and age class were determined at the time of
deployment by experts in the field. Seals determined to be less than one year were
classified as young of the year. subadults are those seals likely greater than one year
of age but not yet sexually mature. Adults are likely sexually mature and older than
approximately four years. For those tags deployed on young of the year and transmitting
into the next year, the age class was advanced to subadult on March 1 of the following
year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

```{r tblDeploy, include=TRUE, tab.id = "tblDeploy", tab.cap = "Summary of bio-logger deployments across seal species and age classification from 1 March to 15 July. Total seal hours represents the sum of available data across all seals."}

deploy_tbl %>% 
  dplyr::group_by(species, sex, age) %>% 
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>% 
  dplyr::mutate(
    deploy_stats = glue::glue("{n} ({seal_hours} seal hours)")
  ) %>% 
  dplyr::select(-c("n","seal_hours")) %>% 
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>% 
  dplyr::ungroup() %>% 
  dplyr::rename(Species = species, Sex = sex, `Young of Year` = `YOUNG OF YEAR`,
                `subadult` = SUBADULT, Adult = ADULT) %>% 
  flextable() %>% 
  add_header(`Young of Year` = "Age Class",
   `subadult` = "Age Class", Adult = "Age Class",
   top = TRUE ) %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "header") %>% 
  font(fontname = "Source Serif Pro Light", part = "all") %>% 
  fontsize(size = 10) %>% 
  autofit()
```

```{r seal-hours}
seal_hours_tbl <- dat.sf %>% 
  tibble::as_tibble() %>% 
  dplyr::mutate(wkofyr = lubridate::epiweek(haulout_dt),
                dayofwk = lubridate::wday(haulout_dt, label = TRUE, abbr = TRUE) %>% 
                  forcats::fct_rev()
  ) %>% 
  dplyr::group_by(species, wkofyr, dayofwk) %>% 
  dplyr::count()
```

Tags that fall off due to molt or mortality and remain on ice or land can still send data
to satellites and we did not want to include data from these tags in our analyses. As
such, start and end times of each deployment were made by experts who examined several
simultaneous data streams (e.g., tag locations and dive behavior) to determine when tags
ceased providing data consistent with seal behavior. Sensor readings outside of these
start and end times were discarded prior to analysis. After approximately 9 months, a few
(7) devices deployed on the rear flipper of bearded seals reported implausible sensor
data. All data after the first instance of unrealistic values were censored from this
analysis. Figure \@ref(fig:dataCal) shows a calendar distribution of these data across the
study season for each species. Observations are concentrated in the months of May and June
due to the timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the water and
more data is transmitted. Molt timing also impacts when many deployments end as any tags
adhered to the hair will fall off.

```{r seal-hours-calendar, fig.height = 4.75, include=TRUE, fig.id = "dataCal", fig.cap="Distribution of of haul-out behavior data from 1 March to 15 July for each species. The higher density of data from May and June coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, a majority of deployments started in April and May. Data were collated across all years between 2005 and 2020."}
ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,4.35)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(13,18,23,27),
                     labels = c("April", "May", "June", "July")) +
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(x = "", y = "day of week", color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("Haul-out behavior data is concentrated in the months of May and June",
          subtitle = "data presented in seal-hours collated across years (2005-2020)")
```

Haul-out behavior data was recorded and transmitted via the Argos satellite network as
hourly percent-dry timelines. For each hour of a day, the wet/dry sensor was polled by the
tag firmware every few seconds and a percent of the hour in a dry state was calculated
(\@ref(fig:examplePlot)). Each data transmission message consisted of a complete 24-hour
record. Tags were deployed and programmed in a manner to maximize data transmission during
the spring pupping and molting period. However, not all days during the deployment were
successfully transmitted. This is due to a variety of factors including satellite
coverage, tag availability (i.e. tags mounted to the rear flipper are not available to the
surface while the seal is at sea), tag performance, duty cycling, and extra-terrestrial
interference.

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, May, and June. Empty areas of the plot represent missing data that was not successfully transmitted from the tag"}
d <- dat.sf %>% 
  filter(speno == "HF2016_1002") %>% 
  dplyr::select(speno, haulout_dt, percent_dry) %>% 
  arrange(haulout_dt) %>% st_drop_geometry() %>% 
  mutate(year = lubridate::year(haulout_dt),
         month = lubridate::month(haulout_dt,label=TRUE),
         day = lubridate::day(haulout_dt),
         hour = lubridate::hour(haulout_dt))

p <- ggplot(d,aes(day,hour,fill=percent_dry))+
  geom_tile(color= "white",size=0) + 
  scale_fill_viridis_c(name = "hourly percent dry", aesthetics = "fill",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15, 
                           barheight = 0.5, title.hjust = 0.5)
    )
p <- p + facet_grid(year~month)
#p <- p + scale_y_continuous(trans = "reverse")
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  ggtitle("Observation records are incomplete because not all data is transmitted",
          subtitle = "percent-dry data for a single ribbon seal across two years") +
  xlab("day of month") + ylab("hour (UTC)")
p
```

Of key interest in this study is the relationship between haul-out behavior and a variety
of environmental covariates that vary with space and time. We explored use of a movement
model to predict locations at specific times. However, the sparse nature of data from some
tags, especially those mounted to the rear flipper, resulted in poor modeling performance
or convergence issues. For this study, we calculated a weighted average daily location
where the inverse of the estimated Argos location error was used for the weight. Any days
where haul-out observations were present without any location data were filled from the
last calculated daily location and any days where the location intersected with land were
removed from the dataset. Figure \@ref(fig:dataMap) shows the spatial distribution of
available haul-out behavior data for each species across the study area.

```{r locs-count-mapdata}
haulout_locs <- dat.sf %>% 
  filter(lubridate::hour(haulout_dt) == 12)

hexgrid <- sf::st_make_grid(st_bbox(haulout_locs) %>% st_as_sfc(), cellsize = 50*1000,
                            what = "polygons", square = FALSE) 
hexgrid <- st_sf(index = 1:length(lengths(hexgrid)), hexgrid)

hexbin <- st_join(haulout_locs, hexgrid, join = st_intersects)

locs_count <- hexgrid %>%
  left_join(
    count(hexbin, index, species) %>%
      as_tibble() %>%
      dplyr::select(species, index, ct=n)
  ) %>% drop_na()
```

```{r dataMap, fig.height = 3.75, include=TRUE, fig.id="dataMap", fig.cap="Spatial distribution of haul-out data during the months of March through July for each of the three species. Data were collated across all years between 2005 and 2020."}

world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(st_crs(locs_count))

ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(
    data = locs_count, size = 0.125,
    aes(fill = ct)
  ) +
  scale_fill_distiller(type = "seq", palette = "YlOrRd", direction = 1,
                    trans = "log10", aesthetics = "fill",
    name = "number of locations",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15, 
                           barheight = 0.5, title.hjust = 0.5)) + 
  facet_grid(~ species) +
  theme(legend.position = "bottom") +
  ggtitle("Spatial distribution of haul-out behavior data")
```

## Explanatory variables

In addition to sex and age class (non-dependent *young of the year*, sexually immature
*subadults*, and mature *adults*), we gathered additional variables that might help
explain variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and solar hour (for diurnal effects). Solar hour was calculated using the {solaR}
package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019] using the
daily locations. We also compiled several meteorological variables shown to affect
haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz]. In
particular, we linked locations from satellite tags to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
data to produce a long-term picture of weather over North America. Numerous weather
variables are made available across the region 8 times daily. For this study, NARR weather
values were subset to the extent of our study area over the Bering and Chukchi Seas at
3-hr intervals based on a grid resolution of 32 km (1024 sq. km) The following
meteorological variables were interpolated and assigned to seal locations using a bilinear
method: 1) air temperature at 2m above the earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the Euclidean norm,
3) barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, and wind chill (represented
by a wind $\times$ temperature interaction; [@udevitz2009cjz]), together with day-of-year and
time of day interactions to permit diurnal patterns to change throughout the year).
Spotted and ribbon seal models included age-sex class and interactions between day-of-year
and age-sex class, but we omitted these from bearded seal models due to poor
representation of age-sex classes (Table \@ref(tab:tblDeploy)). Bearded seal models did
include a latitudinal effect (and an interaction with day-of-year) since bearded seals
occupy a substantial range and we were interested in possible differences in the timing of
haul-out behavior along a latitudinal gradient. We omitted the latitudinal effect from ribbon and
spotted seal models since, during the spring, these species are most prevalent near the
southern ice edge in the Bering Sea [@conn2014mee].

```{r wxCovMap, include=TRUE, fig.height = 5.5, fig.id="wxCovMap", fig.cap="Example spatial distribution of weather covariates across the study region for 2012 May 05. The NARR was used to predict weather covariates for each hour in a percent dry timeline record."}

load(file = here::here('data/cov_means_daily.Rdata'))
grid_plot <- cov_means_daily %>% filter(year == 2012,
                           yday == 125) %>%
  inner_join(grid.sf, by = 'cell') %>%
  sf::st_as_sf()

world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(st_crs(grid_plot))

p_precip <- ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(grid_plot, aes(fill = precip), size = 0) +
  scale_fill_distiller(palette = "RdYlBu") +
  # scale_fill_viridis_c(option = "C") +
  labs(subtitle = "precipitation") +
  theme(legend.position = "none")

p_pressure <- ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(grid_plot, aes(fill = pressure), size = 0) +
  scale_fill_distiller(palette = "RdYlBu") +
  # scale_fill_viridis_c(option = "C") +
  labs(subtitle = "barometric pressure") +
  theme(legend.position = "none")

p_wind <- ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(grid_plot, aes(fill = wind), size = 0) +
  scale_fill_distiller(palette = "RdYlBu") +
  # scale_fill_viridis_c(option = "C") +
  labs(subtitle = "wind") +
  theme(legend.position = "none")

p_temp2 <- ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(grid_plot, aes(fill = temp2), size = 0) +
  scale_fill_distiller(palette = "RdYlBu") +
  # scale_fill_viridis_c(option = "C") +
  labs(subtitle = "temperature") +
  theme(legend.position = "none")

(p_precip + p_pressure) / (p_wind + p_temp2) +
  plot_annotation(
    title = 'Weather varies across the Bering and Chukchi seas',
    caption = "North American Regional Reanalysis (NARR), 2012 May 05")
```

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem counter-intuitive
when modeling the haul-out behavior of seal species with such a close association to sea ice.
Seals haul out in the presence of sea ice and we could assess the local concentration of
sea ice during these events. However, seals also spend time in the water away from
sea ice. Thus, the challenge is how best to associate that seal's at-sea behavior with a
sea ice concentration. Would it be the concentration at the last known haul-out event? The
next known haul-out event? Or, the closest sea ice concentration available? Without a
meaningful means to associate a sea ice covariate with the seal throughout the duration of
the observation record, such parameters are not viable for inclusion in this model.

We did explore whether the annual variation in spring sea ice extent in the Bering Sea
influences the peak timing of seal haul-out behavior. For this, all sea ice concentration
grid cells (25 $km^2$) in the study area with greater than 15% concentration were summed to
get total sea ice extent for each day between 15 February and 15 July across all years 
(`r min(lubridate::year(dat.sf$haulout_dt))`-`r max(lubridate::year(dat.sf$haulout_dt))`). 
The maximum sea ice extent for each year was determined from this calculation. We obtained
sea ice concentration data from the Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave
Dataset, Version 1 [@cavalieri1996].

```{r tblCovariates, include=TRUE, tab.id = "tblCovariates", tab.cap = "Explanatory covariates used in analyses of binary haul-out records for ice-associated seals."}
covariate_tbl <- tibble::tribble(
  ~Covariate, ~Type, ~Source, ~Description,
  "Age-sex class",   "Categorical", "Field Assessment", 
  "Classes included young-of-year, subadult, adult male and adult female. Young-of-year tags surviving past ~Feb 1 of the next year were treated as subadult in subsequent years.",
  "Hour", "Continuous; Fourier basis", "Bio-logger", 
  "We modeled solar hour using 6 variables of a Fourier-series basis to allow circularity, flexibility, and dimension reduction ",
  "Day",   "Continuous", "Bio-logger", 
  "Linear, quadratic, and cubic effects of day-of-year were modeled",
  "Precip", "Continuous", "NARR", 
  "Convective precipitation (kg/m2)",
  "Pressure", "Continuos", "NARR", 
  "Atmospheric pressure at sea level (kPa)",
  "Temp", "Continuous", "NARR",
  "Air temperatures predicted at 2m above the earth’s surface",
  "Wind", "Continuous", "NARR", 
  "Northerly and easterly vector components for wind were downloaded and converted into a single wind speed via the Euclidean norm.",
  "Northing", "Continuous", "Bio-logger","For bearded seals a measurement of northing was caclulated as latitude divided by the mean latitude across all locations"
)

covariate_tbl %>% 
  flextable() %>% 
  font(fontname = "Source Serif Pro Light", part = "all") %>% 
  fontsize(size = 10) %>% 
  set_table_properties(width = .95, layout = "autofit")
```

## Haul-out modeling

```{r runmodels}
if (params$run_models) {
source(here::here('analysis/scripts/model.R'))
} else {
  load(here::here('data/fit_ribbon.Rdata'))
  load(here::here('data/fit_bearded.Rdata'))
  load(here::here('data/fit_spotted.Rdata'))
}
```

Haul-out records for seals are often characterized by multiple hours spent basking on ice
alternating with long periods of foraging (Figure \@ref(fig:examplePlot)). Commonly used
statistical models for binary data (e.g. logistic regression) assume independence among
responses, an assumption that is clearly violated if hourly responses are modeled. Any
analysis that ignores temporal dependence in responses will thus have overstated
precision.

To properly account for temporal dependence within a computationally tractable framework,
we used generalized linear mixed pseudo-models (GLMPMs; [@verhoef2010cs]) to model
variation in haul-out behavior as a function of (1) covariate predictors, (2) temporally
autocorrelated random effects, and (3) individual random effects representing
heterogeneity in individual behavior. We used the {glmmLDTS} package [@verhoef2010cs] to
implement GLMPMs in the R programming environment . Owing to the large number of records,
we fit separate models to bearded, ribbon, and spotted seal data sets. We fit two
different models to our datasets. First, for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted seals (which
had considerably more data than bearded seals), we fitted models that included all the
effects from the first model, but also permitted annual variation in haul-out timing. This
second class of models was used to examine whether haul-out patterns varied by year, and
to determine the annual timing of apparent peaks in haul-out behavior.

We followed Ver Hoef et al. [-@verhoef2014sm] in using linear, quadratic, and cubic
effects of day-of-year to represent seasonal changes in behavior. However, unlike previous
models for harbor seals [@london2012] and ice-associated seals [@verhoef2014sm], which
treated hour-of-day as a 24-level categorical variable to capture diurnal cycles, we
adopted a continuous formulation based on Fourier series that provides a flexible model
while preserving the inherit circularity needed for time-of-day effects (i.e., hour 0
should be equal to hour 24). It also represents hour-of-day with 6 variables, which is a
considerable reduction when compared to a 24 variable model, especially when fitting
models with interactions between hour-of-day and other variables (e.g., age-sex class,
day-of-year). According to this approach, we used the following specification for
hour-of-day effects:

$$
H_t=\sum_{i=1}^{3}{\alpha_{i1}cos{\left(\frac{\pi t}{2\left(i+1\right)}\right)}}+\alpha_{i2}sin{\left(\frac{\pi t}{2\left(i+1\right)}\right)}1
$$

where $H_t$ gives the effect for solar hour $t$ and $α_{ij}$ for $i$ = 1, 2, or 3 and $j$
= 1 or 2 are estimated parameters (regression coefficients).

For the second set of models examining inter-annual variation in sea ice use, we fitted
models with year by day-of-year interactions. However, in this case we only included
year$\times$day and year$\times$day2, omitting the main effects of year as well as year$\times$day3
interactions as models with the latter effects were numerically unstable. However, the
modeled interactions were sufficient to allow shifts in haul-out distribution, as one can
show mathematically that a simple horizontal shift in timing of haul-out distributions
does not affect the main effects or cubic terms in a polynomial regression model.

```{r runyearmodels}
if (params$run_models) {
 source(here::here('analysis/scripts/model_year_hf.R'))
 source(here::here('analysis/scripts/model_year_pl.R'))

} else {
 load(here::here('data/fit_ribbon_year.Rdata'))
 load(here::here('data/fit_spotted_year.Rdata'))
}
```

A typical model fitting exercise would also include a model selection process. However,
AIC (and similar criteria) is not suitable when using pseudo-likelihoods. After fitting
GLMPM models, we used "type III" F-tests to calculate p-values [@verhoef2010cs] and rely
on those insights to evaluate model performance and important terms. We also produced
predictions of haul-out behavior as a function of influential predictors (e.g. solar hour,
day-of-year, age-sex). Weather covariates for these predictions were based on daily or
hourly (GAM) smoothed weather covariate values across the study region. Such predictions
were then used to develop haul-out probability surfaces, explore marginal effects of
weather covariates, and determine annual peaks in haul-out activity. The timing of peak
haul-out behavior was further used to regress against the annual maximum sea ice extent 
in the study region.

## Stable stage distributions

Previous availability correction factors used in aerial surveys for ice-associated Arctic
seals (e.g. [@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]) ignored any age- and
sex-based variation in haul-out probabilities. The implicit assumption is that the
behavior of telemetered animals represents the population as a whole. However, if haul-out
probabilities are related to sex- and age-class, and if there are tangible differences
between the age- and sex-structure of the telemetered sample and that of the population,
correction factors computed in this way may lead to biased abundance estimates.

To determine how much of a potential problem this is for our data, we compared estimates
of haul-out probabilities that (1) ignored sex- and age-structure, and (2) accounted for
sex- and age-structure, adjusting for the relative proportion of seals in each sex- and
age-class. To simplify comparisons, we fitted two additional GLMPMs to ribbon and spotted
seal haul-out datasets. In the first model, we expressed haul-out probabilities as a
function of solar hour and day-of-year. In the second, we also included effects of sex-age
class and sex-age interactions with solar hour and day-of-year. To adjust for the
proportion of seals in each age class, we calculated a predicted population-averaged
haul-out probability at solar noon as a function of day-of-year ($t$) as

$$
\widetilde{p_t}=\sum_{s}\sum_{a}{\omega_{s,a}{\hat{p}}_{s,a,t}}
$$

where $\omega_{s,a}$ is the proportion of the population of seals in age class $a$ and sex
$s$. Typically, we do not know $\omega_{s,a}$ exactly, but can use information about
likely mortality and reproductive schedules of seals to approximate $\omega_{s,a}$ using
stable stage distributions calculated from matrix population models ().

To calculate plausible values for $\omega_{s,a}$ we obtained natural mortality predictions
from a hierarchical meta-analysis of phocid natural mortality [@trukhanova2018ee]. We then
assumed a 50/50 sex ratio at birth, 1 pup per litter, and reproductive schedules as
reported by Fedoseev [-@fedoseev2000] for seals in the western Bering Sea to formulate
Leslie matrices Ai for each species i. Stable age distributions
$v_i={v\_i0,v\_i1,⋯,v\_iA}$ were then determined by normalizing the dominant eigenvector
of $A_i$ so that they summed to one . Given a stable age distribution, we calculated the
proportion of the population by sex and stage (young of the year: 0-1 year; subadult: at
least one year but sexually immature; adult: sexually mature) using data on sexual
maturity from Fedoseev [-@fedoseev2000]. In particular, the expected proportion of the
population that are sex $s$ and young-of-year was set as $\pi_{i,s,yoy}=0.5v_i0$, while
the expected proportion that are subadults and adults were set to

$$
\pi_{i,s,sub}={0.5\sum_{a}{{(1-m}_{i,s,a})}v}_{ia}
$$

and

$$
\pi_{i,s,ad}={0.5\sum_{a}\ m_{i,s,a}v}_{ia}
$$

Here, $m_{i,s,a}$ gives the anticipated proportion of animals of species $i$ and sex $s$
that are sexually mature. Further details on mortality, reproductive and maturity
schedules, and matrix modeling methods are presented in Appendix A.

# Results

```{r createnewdata}
# function to create a prediction data frame for a given age_sex group
create_newdata <- function(data, age_sex) {
  df_list <- vector(mode = "list", length = length({{age_sex}} ))

  for (a_s in {{age_sex}}) {
  range_yday <- data %>% filter(age_sex == a_s ) %>%
    summarize(start_day = min(yday),
              end_day = max(yday))
  start_day <- range_yday$start_day
  end_day <- range_yday$end_day
  n_days = (end_day - start_day) + 1

  # for wx covariates we'll use a gam to get values by day/hour since wx is likely
  # to vary w/in day over the season

  gam.baro <-
    gam(pressure ~ s(yday), data = data)
  gam.temp <-
    gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), data = data)
  gam.wind <-
    gam(wind ~ s(yday) + s(as.numeric(solar_hour)), data = data)
  gam.precip <-
    gam(precip ~ s(yday), data = data)

  wx_new_data <- data.frame(
    solar_hour = rep(0:23, each = n_days),
    yday = rep(start_day:end_day, times = 24)
  )

  temp_pred <- predict(gam.temp, newdata = wx_new_data)
  wind_pred <- predict(gam.wind, newdata = wx_new_data)
  baro_pred <- predict(gam.baro, newdata = wx_new_data)
  precip_pred <- predict(gam.precip, newdata = wx_new_data)

  df <- data.frame(
    age_sex = a_s,
    solar_hour = rep(0:23, each = n_days),
    yday = rep(start_day:end_day, times = 24),
    northing = mean(data$northing),
    temp2 = temp_pred,
    wind = wind_pred,
    pressure = baro_pred,
    precip = precip_pred
  ) %>%
    mutate(
      sin1 = sin(pi * solar_hour / 12),
      cos1 = cos(pi * solar_hour / 12),
      sin2 = sin(pi * solar_hour / 6),
      cos2 = cos(pi * solar_hour / 6),
      sin3 = sin(pi * solar_hour / 4),
      cos3 = sin(pi * solar_hour / 4),
    ) %>%
    mutate(day = (yday - 120) / 10,
           day2 = day ^ 2,
           day3 = day ^ 3)
  df_list[[a_s]] <- df
  }
  if(length({{age_sex}}) > 1) {
  df_out <- bind_rows(df_list) %>%
    mutate(age_sex = forcats::fct_relevel(
      age_sex,c("ADULT.F","ADULT.M","SUBADULT","YOUNG OF YEAR"))
    )
  } else {
    df_out <- bind_rows(df_list)
  }
}
```

Models omitting year effects suggested that day-of-year, solar hour, age-sex class,
temperature, and wind substantially altered haul-out behavior of all three species, with F
tests producing p-values less than 0.05 for variables embodying these effects and/or their
interactions. Haul-out probabilities typically increased throughout March and April,
reaching a peak in May and early June before declining again. Diurnal patterns were
present, with maximum haul-out behavior centered around local solar noon.

## Ribbon Seals

```{r hfnewdata}
# get the ribbon seal fit object
load(here::here("data/fit_ribbon.Rdata"))
ribbon_newdata <- create_newdata(fit_ribbon$dataset,
                                 age_sex = levels(fit_ribbon$dataset$age_sex))

# create the model matrix
ribbon_mm <- model.matrix(fit_ribbon$fixed.formula[-2],
                data = ribbon_newdata)

#clean up extra intercept and get coef
fit_coef <- fit_ribbon$fixed.effects %>%
  filter(!is.na(std.err)) %>%
  pull(estimate)

ribbon_newdata <- ribbon_newdata %>%
  mutate(logit_fits = as.vector(ribbon_mm %*% fit_coef),
         logit_fits_var = diag(ribbon_mm %*% fit_ribbon$covb %*% t(ribbon_mm)),
         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
         ho_prob = plogis(ribbon_mm %*% fit_coef),
         lower95 = plogis(logit_fits_lo95),
         upper95 = plogis(logit_fits_up95)
         )
```

Ribbon seals show a pattern of gradually increasing haul-out probability in April that
peaks in May for subadults and in June for adults. The behavior is clearly centered
around local solar noon and expands to other hours later in the season as seals enter
their molting period. subadults showed an earlier start and more intense haul-out
activity in April and May. The young of the year records begin after weaning and the model
predictions demonstrate the ontogeny of in-water activities (e.g. diving, foraging) in
May. Adult females have a more protracted season compared to males and more time spent
resting in June and July. The model suggests adult male ribbon seals complete their molt
by the end of June. Lastly, for adults of both sexes there is some indication for a shift
to a nocturnal haul-out pattern in late June and July.

```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOCal", fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model."}
plot_df <- ribbon_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label=TRUE),
         day = lubridate::day(date)) %>%
  filter(month != "Aug")

p <- ggplot(plot_df,aes(day,solar_hour,fill=ho_prob))+
  geom_tile(color= "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  xlab("day of month") + ylab("solar hour") +
  ggtitle("Ribbon seals rest on sea ice in late spring and around solar noon.")
p
```

The haul-out probability for ribbon seals was strongly influenced by each of the weather
covariates included within the model. The impact of temperature, wind, and precipitation
was especially noticeable as ribbon seals were less likely to haul out at higher winds and
higher precipitation rates and more likely to be on the ice when temperatures were
relatively warm. Wind chill (temperature $\times$ wind) also had a negative influence on haul-out
probability.

```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Predicted haul-out probability of ribbon seals across a range of weather conditions."}
library(ggsci)

p_temp <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(temp2*27, ho_prob, color = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
  geom_smooth(se = FALSE) +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
  theme(legend.position = "none") +
  xlab("temperature (C)") + ylab("haul-out probability")

p_wind <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(wind, ho_prob, color = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) ),
                 name = "") +
  geom_smooth(se = FALSE) +
  theme(legend.position = "top") +
  xlab("wind") + ylab("")

p_pressure <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(pressure, ho_prob, color = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none") +
  xlab("barometric pressure") + ylab("haul-out probability")

p_precip <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(precip, ho_prob, color = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none") +
  xlab("precipitation") + ylab("")

library(patchwork)

p_temp + p_wind + p_pressure + p_precip +
  plot_annotation(
    title = 'Ribbon seal haul-out behavior is influenced by weather')
```

## Spotted Seals

```{r plnewdata}
# get the spotted seal fit object
load(here::here("data/fit_spotted.Rdata"))

spotted_newdata <- create_newdata(fit_spotted$dataset,
                                 age_sex = levels(fit_spotted$dataset$age_sex))

# create the model matrix
spotted_mm <- model.matrix(fit_spotted$fixed.formula[-2],
                          data = spotted_newdata)

#clean up extra intercept and get coef
fit_coef <- fit_spotted$fixed.effects %>%
  filter(!is.na(std.err)) %>%
  pull(estimate)

spotted_newdata <- spotted_newdata %>%
  mutate(logit_fits = as.vector(spotted_mm %*% fit_coef),
         logit_fits_var = diag(spotted_mm %*% fit_spotted$covb %*% t(spotted_mm)),
         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
         ho_prob = plogis(spotted_mm %*% fit_coef),
         lower95 = plogis(logit_fits_lo95),
         upper95 = plogis(logit_fits_up95)
  )
```

Compared to ribbon seals, spotted seals showed a longer spring haul-out season that is less
intensely centered on solar noon. Adults of both sexes spend considerable time in April
and May resting on the ice. As with ribbon seals, the young of the year records begin
after weaning and the model predictions demonstrate the ontogeny of in-water activities
(e.g. diving, foraging) in May.

```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.id = "spottedHOCal", fig.cap="Predicted haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model."}
plot_df <- spotted_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label=TRUE),
         day = lubridate::day(date)) %>%
  filter(month != "Aug")

p <- ggplot(plot_df,aes(day,solar_hour,fill=ho_prob))+
  geom_tile(color= "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  xlab("day of month") + ylab("solar hour") +
  ggtitle("Spotted seals rest on sea ice in late spring and around solar noon.")
p
```

Spotted seal haul-out behavior appears most strongly influenced by temperature and wind.
Spotted seals were less likely to haul out at higher winds and more likely to be on the
ice when temperatures were relatively warm. Wind chill (temperature $\times$ wind) also had a
negative influence on haul-out probability. Barometric pressure and precipitation were
influential to a slightly lesser extent. Differences in the magnitude of response between
the age-sex classes is present and consistent across each of the weather covariates.

```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Predicted haul-out probability of spotted seals across a range of weather conditions."}
library(ggsci)

p_temp <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(temp2*27, ho_prob, color = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none") +
  xlab("temperature (C)") + ylab("haul-out probability")

p_wind <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(wind, ho_prob, color = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) ),
                 name = "") +
  geom_smooth(se = FALSE) +
  theme(legend.position = "top") +
  xlab("wind") + ylab("")

p_pressure <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(pressure, ho_prob, color = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none") +
  xlab("barometric pressure") + ylab("haul-out probability")

p_precip <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(precip, ho_prob, color = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none") +
  xlab("precipitation") + ylab("")

library(patchwork)

p_temp + p_wind + p_pressure + p_precip +
  plot_annotation(
    title = 'Spotted seal haul-out behavior is influenced by weather')
```

## Bearded Seals

Age and sex class were not included in the model for bearded seals and, as such, results
are shown for all ages. Unlike ribbon and spotted seals, bearded seals were predicted to
have a bi-modal distribution of haul-out probability across the day. In addition to a peak
around local solar noon, the bearded seal model predicts additional haul-out activity
around local midnight. The haul-out behavior is also more protracted throughout the spring
season compared to ribbon and spotted seals.

```{r ebnewdata}
# get the bearded seal fit object
load(here::here("data/fit_bearded.Rdata"))

eb_data <- fit_bearded$dataset %>% 
  mutate(age_sex = forcats::as_factor("ALL AGES"))

bearded_newdata <- create_newdata(eb_data,
                                  age_sex = levels(eb_data$age_sex))

# create the model matrix
bearded_mm <- model.matrix(fit_bearded$fixed.formula[-2],
                           data = bearded_newdata)

#clean up extra intercept and get coef
fit_coef <- fit_bearded$fixed.effects %>%
  filter(!is.na(std.err)) %>%
  pull(estimate)

bearded_newdata <- bearded_newdata %>%
  mutate(logit_fits = as.vector(bearded_mm %*% fit_coef),
         logit_fits_var = diag(bearded_mm %*% fit_bearded$covb %*% t(bearded_mm)),
         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
         ho_prob = plogis(bearded_mm %*% fit_coef),
         lower95 = plogis(logit_fits_lo95),
         upper95 = plogis(logit_fits_up95)
  )
```

```{r beardedHOCal, fig.height = 3.5, include=TRUE, fig.id = "beardedHOCal", fig.cap="Predicted haul-out probability of bearded seals from 1 March to 15 July for each age and sex class used in the model."}
plot_df <- bearded_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label=TRUE),
         day = lubridate::day(date)) %>%
  filter(month != "Aug")

p <- ggplot(plot_df,aes(day,solar_hour,fill = ho_prob))+
  geom_tile(color= "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  xlab("day of month") + ylab("solar hour") +
  ggtitle("Bearded seals rest on sea ice around solar noon and solar midnight")
p
```

Like ribbon and spotted seals, bearded seal haul-out probability appears influenced by
wind and temperature. And, the model also predicted a response to wind chill. Any
influence of barometric pressure or precipitation is less apparent.

```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Predicted haul-out probability of bearded seals across a range of weather conditions."}
library(ggsci)

p_temp <- ggplot(plot_df, aes(temp2*27, ho_prob, color = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none") +
  xlab("temperature (C)") + ylab("haul-out probability")

p_wind <- ggplot(plot_df, aes(wind, ho_prob, color = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none") +
  xlab("wind") + ylab("")

p_pressure <- ggplot(plot_df, aes(pressure, ho_prob, color = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none") +
  xlab("barometric pressure") + ylab("haul-out probability")

p_precip <- ggplot(plot_df, aes(precip, ho_prob, color = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none") +
  xlab("precipitation") + ylab("")

library(patchwork)

((p_temp + p_wind) / (p_pressure + p_precip)) +
  plot_annotation(
    title = 'Bearded seal haul-out behavior is less influeced by weather')
```

## Annual variation in haul-out

```{r createnewdatayr}
create_newdata_year <- function(data, age_sex) {
  df_list <- vector(mode = "list", length = length({{age_sex}} ))
  df_list_years <- vector(mode = "list", length = length(levels(data$year)))

  for (a_s in {{age_sex}}) {
  for (y in levels(data$year)) {
    d <- data %>% filter(year == y, age_sex == a_s)
    if(nrow(d) == 0) { next }
    range_yday <- d %>%
      summarize(start_day = min(yday),
                end_day = max(yday))
    if(range_yday$start_day > 152) {next}
    if(range_yday$end_day < 121) {next}
    print(range_yday)
    start_day <- range_yday$start_day
    end_day <- range_yday$end_day
    n_days = (end_day - start_day) + 1

    # for wx covariates we'll use a gam to get values by day/hour since wx is likely
    # to vary w/in day over the season

    gam.baro <-
      gam(pressure ~ s(yday), data = data)
    gam.temp <-
      gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), data = data)
    gam.wind <-
      gam(wind ~ s(yday) + s(as.numeric(solar_hour)), data = data)
    gam.precip <-
      gam(precip ~ s(yday), data = data)

    wx_new_data <- data.frame(
      solar_hour = 12,
      yday = start_day:end_day
    )

    temp_pred <- predict(gam.temp, newdata = wx_new_data)
    wind_pred <- predict(gam.wind, newdata = wx_new_data)
    baro_pred <- predict(gam.baro, newdata = wx_new_data)
    precip_pred <- predict(gam.precip, newdata = wx_new_data)

    df <- data.frame(
      age_sex = a_s,
      year = y,
      solar_hour = 12,
      yday = start_day:end_day,
      northing = mean(data$northing),
      temp2 = temp_pred,
      wind = wind_pred,
      pressure = baro_pred,
      precip = precip_pred
    ) %>%
      mutate(
        sin1 = sin(pi * solar_hour / 12),
        cos1 = cos(pi * solar_hour / 12),
        sin2 = sin(pi * solar_hour / 6),
        cos2 = cos(pi * solar_hour / 6),
        sin3 = sin(pi * solar_hour / 4),
        cos3 = sin(pi * solar_hour / 4),
      ) %>%
      mutate(day = (yday - 120) / 10,
             day2 = day ^ 2,
             day3 = day ^ 3)
    df_list_years[[y]] <- df
  }
    df_list[[a_s]] <- bind_rows(df_list_years)
  }
  df_out <- bind_rows(df_list) %>%
    mutate(age_sex = forcats::fct_relevel(
      age_sex,c("ADULT.F","ADULT.M","SUBADULT","YOUNG OF YEAR")),
      year = forcats::as_factor(as.numeric(year))
    )

  return(df_out)
}
```

```{r hfpvyear}
load(here::here('data/fit_ribbon_year.Rdata'))

ribbon_newdata_year <- create_newdata_year(fit_ribbon_year$dataset,
                                           age_sex = levels(fit_ribbon_year$dataset$age_sex))

# create the model matrix
ribbon_mm <- model.matrix(fit_ribbon_year$fixed.formula[-2],
                          data = ribbon_newdata_year)

#clean up extra intercept and get coef
fit_coef <- fit_ribbon_year$fixed.effects %>%
  filter(!is.na(std.err)) %>%
  pull(estimate)

# multiply by the model coefficients and convert to probabilities with plogis
# (this came from J Laake code for Hood Canal estimates)
ribbon_newdata_year <- ribbon_newdata_year %>%
  mutate(logit_fits = as.vector(ribbon_mm %*% fit_coef),
         logit_fits_var = diag(ribbon_mm %*% fit_ribbon_year$covb %*% t(ribbon_mm)),
         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
         ho_prob = plogis(ribbon_mm %*% fit_coef),
         lower95 = plogis(logit_fits_lo95),
         upper95 = plogis(logit_fits_up95)
  )

load(here::here('data/fit_spotted_year.Rdata'))

spotted_newdata_year <- create_newdata_year(fit_spotted_year$dataset,
                                            age_sex = levels(fit_spotted_year$dataset$age_sex))

# create the model matrix
spotted_mm <- model.matrix(fit_spotted_year$fixed.formula[-2],
                           data = spotted_newdata_year)

#clean up extra intercept and get coef
fit_coef <- fit_spotted_year$fixed.effects %>%
  filter(!is.na(std.err)) %>%
  pull(estimate)

# multiply by the model coefficients and convert to probabilities with plogis
# (this came from J Laake code for Hood Canal estimates)
spotted_newdata_year <- spotted_newdata_year %>%
  mutate(logit_fits = as.vector(spotted_mm %*% fit_coef),
         logit_fits_var = diag(spotted_mm %*% fit_spotted_year$covb %*% t(spotted_mm)),
         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
         ho_prob = plogis(spotted_mm %*% fit_coef),
         lower95 = plogis(logit_fits_lo95),
         upper95 = plogis(logit_fits_up95)
  )

```

<!-- The second set of models that included annual variation in haul-out patterns produced -->
<!-- p-values that were strongly significant for linear (ribbon seals: F10,76662=9.92, -->
<!-- p\<0.0001; spotted seals: F8,51981=5.23, p\<0.0001) and quadratic (ribbon seals: -->
<!-- F10,76662=8.02, p\<0.0001; spotted seals: F8,51981=3.79, p=0.0002) year-by-day -->
<!-- interactions. Predicted distributions of haul-out activity were largely unimodal, but -->
<!-- varied substantially among and within years with respect to both the timing and magnitude -->
<!-- of haul-out peaks (Figure \@ref(fig:annualHO)). For instance, in 2010 and 2016 spotted -->
<!-- seal haul-out activity peaked in early May, while in other years (e.g. 2011) it appeared -->
<!-- to peak three weeks later. It is important to note that predicted variation in annual -->
<!-- haul-out patterns likely reflects both process error and sampling variability. For -->
<!-- instance, some of the more extreme patterns (e.g. ribbon seals in 2012 , spotted seals in -->
<!-- 2014) were for years with only one tagged seal and are unlikely to represent -->
<!-- population-level processes. -->

```{r annualHO, fig.height = 5.5, include=TRUE, fig.id = "annualHO", fig.cap="Annual variability in the timing of peak haul-out probability for ribbon and spotted seals across 10 years. Predictions are shown for local solar noon and under smoothed weather conditions. Only those groups (age-sex + year) that included observations from more than one seal are shown. Additionally, any groups where the underlying data started after 01-June or ended before 01-May are not included."}
select_groups <- fit_ribbon_year$dataset %>%
  filter(age_sex != "YOUNG OF YEAR") %>%
  group_by(age_sex, year) %>%
  summarise(n = n_distinct(speno), group_id = paste0(age_sex,year)) %>%
  filter(n >1) %>% pull(group_id)

plot_df <- ribbon_newdata_year %>%
  mutate(species = "Ribbon seal",
         date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label=TRUE),
         day = lubridate::day(date)) %>%
  filter(month != "Aug", paste0(age_sex,year) %in% select_groups)

peak_ho_df <- plot_df %>% filter(age_sex != "YOUNG OF YEAR",
                                 yday > 100) %>%
  group_by(species, age_sex, year) %>% slice_max(ho_prob)

select_groups <- fit_spotted_year$dataset %>%
  filter(age_sex != "YOUNG OF YEAR") %>%
  group_by(age_sex, year) %>%
  summarise(n = n_distinct(speno), group_id = paste0(age_sex,year)) %>%
  filter(n >1) %>% pull(group_id)

plot_df2 <- spotted_newdata_year %>%
  mutate(species = "Spotted seal",
         date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label=TRUE),
         day = lubridate::day(date)) %>%
  filter(month != "Aug", paste0(age_sex,year) %in% select_groups)

peak_ho_df2 <- plot_df2 %>% filter(age_sex != "YOUNG OF YEAR",
                                 yday > 100) %>%
  group_by(species, age_sex, year) %>% slice_max(ho_prob)

plot_df <- plot_df %>% bind_rows(plot_df2)
peak_ho_df <- peak_ho_df %>% bind_rows(peak_ho_df2)

p <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(yday, ho_prob)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.25, 
                  stroke = 0, color = "grey80") +
  geom_pointrange(data = peak_ho_df,
                  mapping = aes(yday, ho_prob, ymin = 0, ymax = ho_prob, color = year),
                  shape = 15) +
  scale_color_brewer(palette = "RdYlBu") +
  scale_x_continuous(breaks = c(74,105,135,166),
                     labels = c("15-Mar", "15-Apr", "15-May", "15-Jun")) +
  facet_grid(age_sex~species) +
  theme(legend.position = "bottom") +
  xlab("") + ylab("haul-out probability") +
  ggtitle("Annual variability in the timing of peak haul-out probability")
 p
```

```{r}
load(here::here('data/seaice_extent.Rdata'))

extent_max <- extent_df %>% 
  group_by(year) %>% 
  slice_max(area_km2_epsg3571) %>% 
  ungroup() %>% 
  rename(max_extent_km2 = area_km2_epsg3571,
         max_extent_yday = yday) %>% 
  dplyr::select(-fdate)

extent_lookup <- extent_df %>% 
  dplyr::select(fdate, area_km2_epsg3571) %>% 
  rename(extent_km2 = area_km2_epsg3571)

peak_ho_df <- peak_ho_df %>% left_join(extent_max, 
                         by = "year") %>% 
  left_join(extent_lookup, by = c("date" = "fdate")) %>% 
  mutate(peak_diff_days = yday - max_extent_yday) %>% 
  distinct()

models <- peak_ho_df %>% 
  mutate(max_extent_km2 = max_extent_km2/1e6) %>% 
  group_by(species, age_sex) %>%  
  nest() %>% 
  mutate(model = map(data, ~ lm(max_extent_km2 ~ yday, data = .x))) %>% 
  mutate(glance = map(model, broom::glance)) %>% 
  unnest(glance)

hf_adultF <- models %>% filter(species == "Ribbon seal" & age_sex == "ADULT.F") %>% 
  dplyr::select(r.squared, statistic, p.value)
hf_adultM <- models %>% filter(species == "Ribbon seal" & age_sex == "ADULT.M") %>% 
  dplyr::select(r.squared, statistic, p.value)
hf_subadult <- models %>% filter(species == "Ribbon seal" & age_sex == "SUBADULT") %>% 
  dplyr::select(r.squared, statistic, p.value)
pl_adultF <- models %>% filter(species == "Spotted seal" & age_sex == "ADULT.F") %>% 
  dplyr::select(r.squared, statistic, p.value)
pl_adultM <- models %>% filter(species == "Spotted seal" & age_sex == "ADULT.M") %>% 
  dplyr::select(r.squared, statistic, p.value)
pl_subadult <- models %>% filter(species == "Spotted seal" & age_sex == "SUBADULT") %>% 
  dplyr::select(r.squared, statistic, p.value)

```

The timing of peak haul-out probability for ribbon and spotted seals appeared to have 
only a limited to no relationship with the amount of yearly maximum sea ice extent. Adult female and
subadult spotted seals show a negative trend line but no indication that the observed
trend is meaningful (R-squared and p-values for adult female spotted seals 
($R^{2}$: `r formatC(pl_adultF$r.squared, digits=3)`; $p$: 
`r formatC(pl_adultF$p.value, digits=3)`) and
spotted seal subadults($R^{2}$: `r formatC(pl_subadult$r.squared, digits=3)`; 
$p$: `r formatC(pl_subadult$p.value, digits=3)`)). 
For ribbon seals and adult male spotted seals, p-values were
substantially larger than 0.05 (R-squared and p-values for ribbon seals (adult
females($R^{2}$: `r formatC(hf_adultF$r.squared, digits=3)`; $p$: 
`r formatC(hf_adultF$p.value, digits=3)`), 
adult males($R^{2}$: `r formatC(hf_adultM$r.squared, digits=3)`; $p$: 
`r formatC(hf_adultM$p.value, digits=3)`), 
subadults($R^{2}$: `r formatC(hf_subadult$r.squared, digits=3)`; $p$: `r formatC(hf_subadult$p.value, digits=3)`)) and 
spotted seals (adult males($R^{2}$: `r formatC(pl_adultM$r.squared, digits=3)`; $p$: `r formatC(pl_adultM$p.value, digits=3)`)).

```{r maxextentHO, fig.height = 5.5, include=TRUE, fig.id = "maxextent", fig.cap="Annual timing of peak haul-out behavior analyzed against the yearly maximum area of sea ice extent in the study area. While spotted seal adult females and subadults show a negative trend line, none of the relationships were determined to be significant."}
ggplot(data = peak_ho_df, aes(x = yday, y = max_extent_km2, color = species)) + 
  geom_smooth(method = "lm", alpha = 0.2, aes(fill = species)) +
  geom_point() +
  facet_grid(age_sex~.) +
  scale_x_continuous(breaks = c(125,135,145, 156),
                     labels = c("05-May", "15-May", "25-May", "05-Jun")) +
  scale_color_manual(values = c('#E66100','#5D3A9B'), name = "") +
  scale_fill_manual(values = c('#E66100','#5D3A9B'), name = "") +
  theme(legend.position = "top",
        legend.justification = "right") +
  labs(x = "date of peak haul-out", y = "maximum sea ice extent (km^2 / 1e6)",
       title = "The connection between yearly maximum sea ice extent and the \ntiming of peak haul-out probability is limited")

```

## Stable stage distributions

Stable stage distributions calculated using mortality and reproductive schedules produced
different patterns for the four species of ice-associated seals residing in Alaskan waters
(Fig. X). Ribbon seals mature the fastest, and so can be expected to have the largest
proportion of adults and smallest proportion of subadults; conversely, ringed seals mature
slowly and are predicted to have roughly equal proportions of adults and subadults (note
that we do not attempt to model ringed seal haul-out behavior in this study; see
Discussion). Population-level haul-out distributions were somewhat different depending on
whether we a) adjusted haul-out probabilities based on stable stage distributions or b)
assumed the tagged sample represented the population. In particular, ribbon seal estimates
differed somewhat at the beginning of the time series, but the two approaches produced
similar estimates (Fig. X) within the window that previous aerial surveys in the region
have been conducted (April and May; e.g. Conn et al. 2014). By contrast, during this time
period spotted seal haul-out estimates were often ≈10% higher for the approach that
adjusted for stable age distributions, presumably because subadults had lower predicted
haul-out probabilities than adults but made up a disproportionately high proportion of the
tagged sample.

# Discussion

In this paper, we fit statistical models to data from sensors on satellite tagged
ice-associated seals in order to examine factors affecting the probability that Arctic
seals haul out on ice. Such probabilities are important for understanding the timing and
ecology of seal behavior, since haul out peaks are associated with important life history
functions such as breeding, pupping and molting. Knowledge of factors affecting haul-out
probabilities is a first step in being able to predict possible impacts of losses of sea
ice on ice-associated seals. Understanding the factors affecting haul-out probabilities is
also necessary to properly interpret seal counts made during aerial surveys. For instance,
a count might be low because there are few seals, or because the timing and conditions are
such that many seals are in the water.

Our analysis revealed that seals use sea ice progressively more throughout the spring, and
with peak use near mid-May to early June before declining again. We also found that seals
preferentially haul out on ice shortly after solar noon, which may be associated with
fewer prey in the water column (??? Can someone else who knows more about their feeding
ecology contribute here?). Interestingly, bearded seals appeared to have two peaks in
haul-out activity throughout the day, one shortly after solar noon, and one centered near
solar midnight (related to benthic feeding?). It appeared that adult spotted seals
initiated haul-out behavior earlier in the year than ribbon seals, and maintained this
behavior over a longer duration; however they exhibited stronger diel behavior than ribbon
seals (Fig X). These patterns are similar to those obtained by Conn et al. (2014), who
used a subset of the data we analyzed here.

Unlike previous analyses of seal haul-out data (e.g. Ver Hoef et al. 2009, Conn et al.
2014), we also investigated the influence of sex-age class on haul-out probabilities.
Although it was difficult to discern a pattern for bearded seals owing to low sample size,
it appeared that adult male spotted and ribbon seals typically hauled out for longer
durations, perhaps because they were less constrained by energetic demands associated with
nursing and could sustain longer periods on ice to engage in molting and mating
activities. Subadult ribbon and spotted seals exhibited different behavior when compared
to adults. In particular, subadult ribbon seals exhibited similar haul-out behavior to
adults, whereas subadult spotted seals appeared to haul-out less frequently and sustained
a shorter duration of peak haul-out behavior than adults. The latter result is likely due
to subadults only needing to haul out on ice to molt; pupping and mating are not yet in
their behavioral repertoire.

We also investigated the influence of environmental covariates on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind chill.
These have been investigated for walrus (e.g. Udevitz et al. 2009), but to our knowledge
have never been investigated for ice-associated seals. In particular, we found clear
evidence that seal haul-out probabilities decrease with wind speed and increase with
temperature (Figs X-Y). In addition, we found evidence that ribbon seal haul-out
probabilities are positively related to barometric pressure, and that spotted seal
haul-out probabilities are negatively related to precipitation (Appendix C). The tendency
of seals to haul out in "good" weather may be related to energetic considerations, better
predator detection, or both. Bearded seal haul-out behavior was predicted to change
slightly based on how far north tagged seals were; seals farther north tended to initiate
haul-out behavior slightly later, and for a longer period of time (Fig. X). This response
may be related to different temperature regimes, or to timing of ice melt, which occurs
later in locations farther north.

There appeared to be rather substantial annual deviations in the timing and magnitude of
haul-out peaks (Figs X-Y). For example, peak ribbon seal haul-out behavior appeared to
vary by as much as three weeks between seasons. There thus seems to be a reasonable amount
of plasticity in the timing of haul-out, suggesting some plasticity in life history
functions such as pupping, molting and breeding. However, we were unable to relate the
timing of haul-out conditions to underlying ice conditions, which makes predicting climate
change consequences on life history events difficult. Future investigation into
environmental factors contributing to the timing of haul-out peaks seems warranted. For
instance, Ray et al. (2010) suggested that ribbon and spotted seals were most dependent on
loose pack ice at the sourthern sea ice edge; thus, the relative suitability for haul-out
behavior may have more to do with the characterization, position, and quantity of loose
pack ice rather than the total amount of ice in the Bering Sea. Further work to
characterize remotely sensed sea ice data into relevant ice types (e.g. loose pack,
continuous pack, rounded pack) would be useful.

Previous attempts to estimate the abundance of phocid seals from aerial survey data in the
Bering and Chukchi Seas (e.g. Bengtson et al. 2005, Conn et al. 2014, Ver Hoef et al.
2014) have used estimated haul-out probabilities to correct for the proportion of animals
that are in the water and thus unavailable to be counted during aerial surveys. Although
several of these studies allowed haul-out probabilities to vary by day-of-year and
time-of-day, previous correction factors have not accounted for variability between years,
different weather conditions, and the age-sex class of the sample. In this paper, we have
shown that there can be considerable differences in the number of seals hauled out on ice
based on these factors. We thus recommend that future abundance analyses employ correction
factors that account for these factors. For instance, it is relatively straightforward to
obtain weather reanalysis products for times and locations that are surveyed and to
construct a relevant correction factor based on predictions of GMPMs. Further, one can
apply stable stage distribution estimates from matrix population models to adjust
predictions to the "average" prediction of the population. The latter suggestion is
particularly appropriate when the age structure of the tagged sample does not reflect the
likely age structure of the population. For instance, population weighted estimates can
differ substantially from those based on a naïve analysis of the tagged sample, as was the
case with analysis of spotted seal data (e.g. Fig X). The most challenging element in
developing availability correction factors is with annual variability. It can be difficult
to get a sufficient sample size to estimate year-specific correction factors, particularly
because research teams would likely need to tag seals and conduct aerial surveys
concurrently which requires considerably more personnel and money. One possible suggestion
is to estimate a "shift" parameter within models for aerial survey counts that allow the
peak of haul-out distributions to be adjusted earlier or later in the year based on the
frequency of counts observed over time. Regardless, researchers should anticipate there
being some unmodeled heterogeneity in availability probability present in abundance
estimates obtained from aerial surveys. This may make trend detection difficult, as one
will not know if moderate differences in abundance estimates are attributable to changes
in abundance or changes in haul-out frequencies.

Predictions of absolute haul-out probability in this paper were somewhat different than
those previously reported for these species, especially for bearded seals. For instance,
Ver Hoef et al. (2014) and Conn et al. (2015) used haul-out correction factors with
maximums of 0.66 for bearded seals, 0.62 for ribbon seals, and 0.54 for spotted seals,
where maximums corresponded to times near solar noon in mid-late May. Applying models that
ignore age, sex, and year effects, these probabilities were 0.38, 0.72, and 0.60,
respectively, under the current analysis framework. Our current estimates reflect
increased sample sizes in terms of number of animals, but also improvements to the way
data are prepared prior to analysis. For instance, for bearded seals there were often long
stretches at the end of the tag history where it appeared animals were hauled out 100% of
the time. In reality, these likely represented tags shed during molt on ice floes or
washed up on beaches. In the present study, experts used multiple data streams to
determine effective start and end dates for each deployment, and these long stretches were
often removed. In previous studies, such periods were often retained and likely led to
positively biased estimates of haul-out probability. Given the potential ramifications for
abundance estimation, we strongly suggest conducting such data screening exercises prior
to analyzing haul-out data. Future research should be devoted to automating this
procedure, perhaps using artificial intelligence algorithms to classify tag behavior based
on multiple data streams. Such a framework would require adequate training data, e.g.
using tags purposefully placed on ice floes, beaches, or subject to tag fouling so that
automatic classifiers could appropriately discriminate properly functioning tags from
anomalous behavior. Another approach would be to integrate additional information on pulse
or respiration, as can now be done with life history tags (e.g., Horning and Hill 2005).
At minimum, we suggest reporting haul-out correction factors used when conducting
abundance analyses so that attendant estimates can be revisited as our state of knowledge
about tag function and haul-out behavior evolves.

We focused this paper on haul-out behavior of bearded, ribbon, and spotted seals. Ringed
seals are also present in the Bering and Chukchi Seas but exhibit qualitatively different
haul-out behavior. For instance, many ringed seals build subnivean layers in the snow on
top of the sea ice, which they frequently inhabit before snow melt causes their lairs to
collapse (Kelly 1988). Thus, the wet-dry sensor on a satellite tag could indicate that an
animal is hauled out, but it could still be unavailable to be detected because it is in a
lair. We hope to address availability of ringed seals using data from satellite tags,
replicate survey tracks, and auxiliary information about snow depth and timing of melt in
a future study.

# Author Contributions

# Data Availability

# Acknowledgements

We recognize that the species and ecosystems studied in this paper are within the
traditional range of the Inpuiaq and Yupik people. And, we would like to acknowledge and
thank them as the original stewards of these ecosystems. The deployment of bio-logging
devices used in this study were often done in collaboration with Alaska Native
communities. We would like to especially acknowledge the communities of Kotzebue,
Utqiagvik, and Ulguniq.

We are grateful for the assistance in catching and sampling seals by Michelle Barbieri,
Gavin Brady, Cynthia Christman, Shawn Dahle, Stacy DiRocco, Deb Fauquier, Shannon
Fitzgerald, Tracey Goldstein, Jeff Harris, Markus Horning, John Goodwin, John Jansen,
Charles Littnan, Erin Moreland, Lorrie Rea, Dan Savetilik, and Dave Withrow. We also
appreciate the commitment to science and safety by all officers and crew of the NOAA Ship
Oscar Dyson, the NOAA Ship MacArthur II, and the RV Thomas G. Thompson.

Telemetry data from the Alaska Department of Fish and Game and the North Slope Borough
Department of Wildlife Management were important contributions to the findings presented
here. Deployments in the western Bering Sea and Sea of Okhotsk were done in collaboration
with Russian colleagues and North Pacific Wildlife.

The findings and conclusions in the paper are those of the author(s) and do not
necessarily represent the views of the National Marine Fisheries Service, NOAA. Any use of
trade, product, or firm names does not imply an endorsement by the U.S. Government.
Funding for this study was provided by the U.S. National Oceanic and Atmospheric
Administration. The field work was conducted under the authority of Marine Mammal
Protection Act Research Permits Nos. 782-1676, 782-1765, 15126, and 19309 issued by the
National Marine Fisheries Service, and Letters of Assurance of Compliance with Animal
Welfare Act regulations, Nos. A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries
Science Center/Northwest Fisheries Science Center Institutional Animal Care and Use
Committee.

# References

::: {#refs}
:::

### 2021-03-05 analysis/scripts/get_data/creating_input_data.Rmd
Initial commit

---
title: "Creating Haul-out Timeline Input Data"
description: |
  document the processing of the 'raw' telemetry data into a usable 
  data structure for downstream analysis procedures
author:
  - name: Josh M. London 
    url: https://github.com/jmlondon
    orcid_id: 0000-0003-0645-5666
    affiliation: NOAA Alaska Fisheries Science Center
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    theme: theme.css
params:
   make_data: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(conflicted)
library(rmarkdown)
library(odbc)
library(DBI)
library(dplyr)
conflict_prefer("filter", "dplyr")
conflict_prefer("selec", "dplyr")
library(dbplyr)
library(pingr)
library(keyringr)
library(lubridate)
conflict_prefer("month","lubridate")
library(sf)
library(glue)
library(tibble)
library(tidyr)
library(purrr)
library(forcats)
library(gt)
library(here)
library(ggplot2)
```

## Introduction

The focus of this document is not to analyze the data, but instead to document the
processing of the 'raw' telemetry data into a usable data structure for downstream
analysis procedures in support of the manuscript *\<insert manuscript title here\>*. The
key focus of this step is to merge separate data streams for estimated seal locations,
observed seal haul-out behavior (in the form of percent-dry timelines), and capture
details into a single table ready for ingestion to the PostgreSQL database. This table
should, generally, provide hourly percent-dry records for each seal along with reasonable
location estimates. We will rely on a separate process within the database for, then,
assigning reanalysis weather covariates to each record in this spatial time series.

## Workflow

This process is managed with a `make_data.R` script. We rely on a series of functions to
pull data from the database and, then, tidy and merge into our needed data structure. We
will briefly describe each functional step below. Each function can be examined in detail
by opening the corresponding R file within the R directory.

```{r}
if (params$make_data) {
source(here::here('analysis/scripts/get_data/make_data.R'))
} else {
  load(here::here('data/locs_sf.Rdata'))
  load(here::here('data/timeline_data.Rdata'))
  load(here::here('data/tbl_percent_locs.Rdata'))
}
```

### `get_timeline_data()`

This is a relatively simple pull and merge of telemetry data from the PEP PostgreSQL
database. The focus is on the *percent_dry* timeline data and joining those records with
additional details regarding seal species, age, sex, etc. We also filter the data to only
include records for *Bearded seals*, *Ribbon seals*, and *Spotted seals*. Also, only
records from March, April, May, June, and July are retained. The last step is to create a
*unique_day* column that represents the unique combination of year and day-of-year. Since
we will be calculating daily location averages, this field will be crucial for properly
joining the timeline data with those locations.

The final data structure:

```{r}
glimpse(timeline_data)
```

### `get_locs_sf()`

This is also a simple pull and merge of telemetry data from PostgreSQL database. The focus
here is on location estimates. As such, we rely on the `sf::read_sf()` function and a
custom query:

``` {.sql}
SELECT  deployid,
        locs_dt,
        type,
        quality,
        error_radius,
        geom
FROM telem.geo_wc_locs_qa
```

After this query, the process is similar to the `get_timeline_data()` function. We merge
with the additional details (e.g. seal species, age, sex) and filter to only include our
months of interest (March, April, May, June). Note we also calculate the *unique_day*
field that will become the basis for our location averaging.

The final data structure:

```{r}
glimpse(locs_sf)
```

### `create_db_input_data()`

This is the workhorse function for this part of our workflow. We need to accomplish a few
key objectives with this function:

1.  determine an appropriate error radius value for each location record

    1.  for records with only Argos location quality, use standard error values

    2.  for records with error_radius values, prefer over Argos location quality values

2.  transform the geographic coordinates to the *LAEA Bering Sea* / *EPSG:3571* projection

3.  calculate a weighted mean daily x and y coordinate where the weights are
    error_radius\^-1

4.  full join of the timeline and daily location records

5.  fill any missing coordinates with the most recent known daily location

6.  write the result back to the PEP PostgreSQL database

Our data represents deployments across several years and differing data formats for Argos
locations. Only the newer (since \~2010) deployments provide an record-specific estimate
of error radius (also, error ellipse information but we are simplifying the approach here
and just using radius). Earlier deployments only provide the location quality values

The following table represents our map of Argos location quality values to estimated error
radius (in meters).

```{r}
tribble(
    ~quality, ~error_radius,
    "3", 250,
    "2", 500,
    "1", 1500,
    "0", 2500,
    "A", 2500,
    "B", 2500
  ) 
```

Once we have an error radius value for each record, we can proceed to transform our
spatial coordinates from geographic (latitude, longitude) into an equal-area projection
appropriate for the Bering Sea. We chose the LAEA Bering Sea (espg:3571) projection. This
transforms our spatial coordinates into units of meters and allows for a straightforward
calculation of daily mean coordinates. Since we have our error radius information, we can
take a more thorough approach to calculation of our daily mean and apply the inverse of
the error radius value as a weight to our mean calculation.

At this point, we have two tables: daily, weighted mean locations and hourly percent-dry
values. A full join of these two tables based on seal identification (i.e. *speno*) and
the *unique_day* provides an full table of all timeline records and, where there is a
matching daily mean location, spatial estimates of seal location. In more than a few
cases, there are observed percent-dry data without a corresponding location for that same
day. In these situations, the previously known location is carried forward to fill in the
unknown days.

## Data

First, the table structure:

```{r}
glimpse(db_input_data)
```

Nothing unexpected here. The key is that this dataset is a Simple Features Collection with
each record represented by a geospatial point. This geometry along with the haul-out
date-time data will allow us to properly link these records with spatially explicit
weather covariates.

Next, let's summarize the number deployments for each species by age and sex along with
the number of seal hours represented. This table allows us to compare the spread and
amount of data for each with our expectations.

```{r}
db_input_data %>% 
  st_set_geometry(NULL) %>% 
  mutate(age = dplyr::case_when(
      age == "YEARLING" ~ "SUBADULT",
      TRUE ~ age
    ),
    sex = dplyr::case_when(
      sex == "F" ~ "Female",
      sex == "M" ~ "Male",
      TRUE ~ sex
    )) %>% 
  mutate(age = fct_relevel(age, "PUP","YOUNG OF YEAR","SUBADULT","ADULT")) %>%  
  dplyr::mutate(data_year = lubridate::year(haulout_dt)) %>% 
  dplyr::group_by(speno,data_year,species, sex, age) %>% 
  dplyr::summarise(min_dt = min(haulout_dt),
            max_dt = max(haulout_dt),
            n_days = difftime(max_dt, min_dt, units = "days"),
            n_hours = n()
  ) %>% 
  dplyr::select(-c("min_dt","max_dt")) %>% 
  dplyr::group_by(species, sex, age) %>% 
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>% 
  dplyr::mutate(
    deploy_stats = glue::glue("{n} ({seal_hours} seal hours)")
  ) %>% 
  dplyr::select(-c("n","seal_hours")) %>% 
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>% 
  group_by(species) %>% 
  gt(rowname_col = "sex") %>% 
  tab_header(
    title = "Summary of bio-logger deployment data across seal 
  species, sex, and age classification during the months of March, April, May, 
  and June.",
    subtitle = "total seal hours represents the sum of available data across all
  seals"
  ) %>% 
  tab_options(
    table.border.top.width = px(0),
    heading.title.font.size = px(16),
    column_labels.border.top.width = 3,
    column_labels.border.top.color = "black",
    column_labels.border.bottom.width = 3,
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table.border.bottom.width = px(0),
    table.width = pct(100),
    table.background.color = "white",
    row_group.border.top.color = "black",
    row_group.border.bottom.width = px(0)
  ) %>% 
  tab_style(
    style = list(
      # remove horizontal lines
      cell_borders(
        sides = c("top", "bottom", "right"),
        weight = px(0)
      )),
      #do this for all columns and rows
    locations = list(cells_body(
      columns = everything(),
      rows = everything()),
    cells_stub())
    ) %>% 
  tab_style(
    style = list(
      cell_text(weight = "bold")
      ),
    locations = cells_row_groups()
  )
```

It's also helpful to visually examine the temporal spread of the data to make sure it
meets our expectations and to identify any temporal limits we might need to consider
during the analysis process.

```{r, fig.height=6.5, fig.width=8}
seal_hours_tbl <- db_input_data %>% 
  sf::st_set_geometry(NULL) %>% 
  dplyr::mutate(wkofyr = lubridate::epiweek(haulout_dt),
                dayofwk = lubridate::wday(haulout_dt, label = TRUE, abbr = TRUE) %>% 
                  forcats::fct_rev()
  ) %>% 
  dplyr::group_by(species, wkofyr, dayofwk) %>% 
  dplyr::count()

ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n)) +
  scale_color_distiller(palette = "Blues", direction = 1) +
  scale_x_continuous(breaks = c(13,18,23),
                     labels = c("April", "May", "June")) +
  guides(color = "none",
         size = guide_legend(override.aes = list(color = "#3182bd"),
                             title.position = "bottom",
                             title.hjust = 0.5)) +
  facet_grid(species ~ .
             # labeller=labeller(species = c(Eb = "Bearded",
             #                               Hf = "Ribbon",
             #                               Pl = "Spotted"))
             ) + 
  theme_minimal(12) + 
  theme(axis.text.y = element_blank(),
        panel.grid = element_blank()) +
  theme(plot.title.position = "plot") +
  theme(legend.position = "bottom") +
  labs(x = "", y = "day of week", color = "seal-hours",
       size = "seal hours") +
  ggtitle("Distribution in seal-hours of haul-out behavior data",
          subtitle = "collated across years (2005-2017)")
```

lastly, let's plot a map of our data so we can get a sense of the geographic distribution
for each species

```{r}
haulout_locs <- db_input_data %>% 
  filter(lubridate::hour(haulout_dt) == 12)

hexgrid <- sf::st_make_grid(st_bbox(haulout_locs) %>% st_as_sfc(), cellsize = 50*1000,
                            what = "polygons", square = FALSE) 
hexgrid <- st_sf(index = 1:length(lengths(hexgrid)), hexgrid)

hexbin <- st_join(haulout_locs, hexgrid, join = st_intersects) %>% 
  group_by(species)

locs_count <- hexgrid %>%
  left_join(
    count(hexbin, index) %>%
      as_tibble() %>%
      dplyr::select(index, ct=n)
  ) %>% drop_na()

ggplot() +
  geom_sf(
    data = locs_count, size = 0.125,
    aes(fill = ct)
  ) +
  scale_fill_viridis_c(
    option = "magma", 
    trans = "log10", aesthetics = "fill",
    name = "# locations"
  )

```

We can see that there are two locations well outside the study area. these are likely a
result of very poor quality Argos locations and the averaging used to calculate daily
locations for this analysis. We can remove these by filtering based on the speno
(*PL2018_1002*) and the two dates (*2018-06-04* and *2018-06-06*).

```{r}
haulout_locs <- db_input_data %>% 
  filter(lubridate::hour(haulout_dt) == 12,
         speno != 'PL2018_1002' & !lubridate::date(haulout_dt) %in% c('2018-06-04','2018-06-06'))

hexgrid <- sf::st_make_grid(st_bbox(haulout_locs) %>% st_as_sfc(), cellsize = 50*1000,
                            what = "polygons", square = FALSE) 
hexgrid <- st_sf(index = 1:length(lengths(hexgrid)), hexgrid)

hexbin <- st_join(haulout_locs, hexgrid, join = st_intersects)

locs_count <- hexgrid %>%
  left_join(
    count(hexbin, index, species) %>%
      as_tibble() %>%
      dplyr::select(species, index, ct=n)
  ) %>% drop_na()
```

With the erroneous locations removed, we can plot the distribution of haulout locations
for each of our three species

```{r, fig.height=3, fig.width=8}
library("rnaturalearth")
library("rnaturalearthdata")
library(ggspatial)

world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(st_crs(locs_count))

ggplot() +
  annotation_spatial(world) +
  layer_spatial(
    data = locs_count, size = 0.125,
    aes(fill = ct)
  ) +
  scale_fill_viridis_c(
    option = "magma", 
    trans = "log10", aesthetics = "fill",
    name = "# locations"
  ) + facet_grid(~ species)
```

### 2021-03-05 R/create_db_input_data.R

@@ -1,0 +1,84 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##' @param timeline_data
+##' @param locs_sf
+##' @return
+##' @author Josh.London
+##' @export
+create_db_input_data <- function(timeline_data, locs_sf) {
+
+  loc_qual_tbl <- tribble(
+    ~quality, ~error_radius,
+    "3", 250,
+    "2", 500,
+    "1", 1500,
+    "0", 2500,
+    "A", 2500,
+    "B", 2500
+  )
+
+  locs_daily <- locs_sf %>%
+    filter(quality != "Z") %>%
+    left_join(loc_qual_tbl, by = "quality") %>%
+    st_transform(3571) %>%
+    mutate(x = st_coordinates(.)[,"X"],
+           y = st_coordinates(.)[,"Y"]) %>%
+    st_set_geometry(NULL) %>%
+    mutate(error_radius.x = ifelse(is.na(error_radius.x),
+                                          error_radius.y,
+                                          error_radius.x)) %>%
+    rename(error_radius = error_radius.x) %>%
+    select(-error_radius.y) %>%
+    mutate(error_radius = ifelse(type %in% c("GPS","FastGPS"),
+                                        50,error_radius),
+                  error_radius = ifelse(type == "User",
+                                        50,error_radius)) %>%
+    group_by(speno,unique_day,age,sex,species) %>%
+    summarise(x = weighted.mean(x,1/error_radius),
+                     y = weighted.mean(y,1/error_radius))
+
+  tbl_percent_locs <- timeline_data %>%
+    full_join(locs_daily,
+                     by = c("speno","unique_day","age",
+                            "sex","species")) %>%
+    arrange(speno,unique_day,timeline_start_dt) %>%
+    group_by(speno) %>% nest() %>%
+    mutate(start_idx = map_int(data,~ which.max(!is.na(.x$x))),
+           data = map2(data, start_idx, ~ slice(.x, .y:nrow(.x)))) %>%
+    select(-start_idx) %>%
+    unnest(cols = c(data)) %>%
+    mutate(fill_xy = ifelse(is.na(x), TRUE, FALSE)) %>%
+    group_by(speno) %>%
+    fill(x,y) %>%
+    ungroup() %>%
[…]

### 2021-03-05 R/get_locs_sf.R

@@ -1,0 +1,50 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+
+##' @return
+##' @author Josh.London
+##' @export
+get_locs_sf <- function() {
+
+  stopifnot(
+    "PEP Postgres Database Not Available; did you start VPN? ;)" =
+      is_up("161.55.120.122", "5432")
+  )
+
+  con <- dbConnect(
+    odbc(),
+    dsn = "PostgreSQL pep",
+    uid = get_kc_account("pgpep_londonj"),
+    pwd = decrypt_kc_pw("pgpep_londonj")
+  )
+
+  deployments_db <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>%
+    select(speno, deployid, tag_family, deploy_dt, end_dt) %>% collect()
+
+  spenos_db <- tbl(con, in_schema("capture","for_telem")) %>% collect()
+
+  locs_qry <- "SELECT deployid,
+                locs_dt,
+                type,
+                quality,
+                error_radius,
+                geom
+              FROM telem.geo_wc_locs_qa"
+
+  locs_sf <- read_sf(con, query = locs_qry) %>%
+    left_join(deployments_db, by = 'deployid') %>%
+    left_join(spenos_db, by = 'speno') %>%
+    filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
+    filter(month(locs_dt) %in% c(3,4,5,6,7)) %>%
+    mutate(unique_day =
+             glue("{year(locs_dt)}",
+                  "{yday(locs_dt)}",
+                  .sep = "_"))
+  dbDisconnect(con)
+
+  return(locs_sf)
+
+}

### 2021-03-05 R/get_timeline_data.R

@@ -1,0 +1,52 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+
+##' @return
+##' @author Josh.London
+##' @export
+get_timeline_data <- function() {
+
+  stopifnot(
+    "PEP Postgres Database Not Available; did you start VPN? ;)" =
+      is_up("161.55.120.122", "5432")
+  )
+
+  con <- dbConnect(
+    odbc(),
+    dsn = "PostgreSQL pep",
+    uid = get_kc_account("pgpep_londonj"),
+    pwd = decrypt_kc_pw("pgpep_londonj")
+  )
+
+  timeline_db <- tbl(con, in_schema("telem","tbl_wc_histos_timeline_qa")) %>%
+    dplyr::select(deployid,timeline_start_dt, percent_dry)
+  deployments_db <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>%
+    dplyr::select(speno, deployid, tag_family, deploy_dt, end_dt)
+  spenos_db <- tbl(con, in_schema("capture","for_telem"))
+
+  timeline_data <- timeline_db  %>%
+    left_join(deployments_db, by = 'deployid') %>%
+    left_join(spenos_db, by = 'speno') %>%
+    filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
+    collect()
+
+  timeline_data <- timeline_data %>%
+    group_by(speno, species, sex, age, timeline_start_dt) %>%
+    summarize(n_tags = n(),
+              percent_dry = ifelse(
+                n_tags == 1, percent_dry, percent_dry[tag_family == "SPOT"]
+              )) %>%
+    filter(month(timeline_start_dt) %in% c(3,4,5,6,7)) %>%
+    mutate(unique_day =
+             glue("{year(timeline_start_dt)}",
+                  "{yday(timeline_start_dt)}",
+                  .sep = "_"))
+
+  dbDisconnect(con)
+
+  return(timeline_data)
+
+}

### 2021-03-05 analysis/scripts/create_avg_wx.R

@@ -1,0 +1,53 @@
+library(tidyverse)
+library(DBI)
+library(dbplyr)
+library(sf)
+
+con <- dbConnect(
+  odbc::odbc(),
+  dsn = "PostgreSQL pep",
+  uid = keyringr::get_kc_account("pgpep_londonj"),
+  pwd = keyringr::decrypt_kc_pw("pgpep_londonj"),
+  sslmode = "disable"
+)
+
+load(here::here('data/dat_sf.Rdata'))
+
+grid.sf <- sf::st_read(con,
+                       query = "SELECT * FROM base.geo_analysis_grid") %>%
+  sf::st_crop(st_bbox(dat.sf) %>% st_as_sfc())
+
+cov_means <- tbl(con, in_schema("base","tbl_analysis_grid_cov_wx")) %>%
+  filter(cell %in% !!grid.sf$cell) %>%
+  filter(month(fdatetime_range_start) %in% c(3,4,5,6,7),
+         year(fdatetime_range_start) != '2004') %>%
+  group_by(year = year(fdatetime_range_start),
+           yday = yday(fdatetime_range_start)) %>%
+  summarise(precip = mean(rast_acpcp, na.rm=TRUE),
+            temp2 = (mean(rast_air2m, na.rm=TRUE)-270),
+            temp0 = (mean(rast_airsfc, na.rm=TRUE)-270),
+            pressure = (mean(rast_prmsl, na.rm=TRUE)-100000),
+            wind = sqrt(mean(rast_uwnd, na.rm=TRUE)^2 + mean(rast_vwnd, na.rm=TRUE)^2)
+            )
+
+cov_means_daily <- tbl(con, in_schema("base","tbl_analysis_grid_cov_wx")) %>%
+  filter(cell %in% !!grid.sf$cell) %>%
+  filter(month(fdatetime_range_start) %in% c(3,4,5,6,7),
+         year(fdatetime_range_start) != '2004') %>%
+  group_by(year = year(fdatetime_range_start),
+           yday = yday(fdatetime_range_start),
+           cell) %>%
+  summarise(precip = mean(rast_acpcp, na.rm=TRUE),
+            temp2 = (mean(rast_air2m, na.rm=TRUE)-270),
+            temp_sfc = (mean(rast_airsfc, na.rm=TRUE)-270),
+            pressure = (mean(rast_prmsl, na.rm=TRUE)-100000),
[…]

### 2021-03-05 analysis/scripts/estimate.R

@@ -1,0 +1,93 @@
+library(glmmLDTS)
+
+fit_ribbon <- readRDS(here::here("data/fit_ribbon.rds"))
+#available here: https://drive.google.com/file/d/1rsC7PM11QcSFsxe-XKFiNtJq1ZkeG1QL/view?usp=sharing
+
+# example from Jay's email
+# for(i in 1:ndates) {
+#   results[[i]] <- estimate.glmp.ts(glmp.ts.object = all.fit.1,
+#                                    fixed.effect.list = list(
+#                                      eff1 = list(effect = "MinFHi.Cat",
+#                                                  level = as.character(survcov[i,'MinFHi.Cat']),
+#                                                  weight = 1),
+#                                      eff2 = list(effect = "MFAug15",
+#                                                  weight = survcov[i,'MFAug15']),
+#                                      eff3 = list(effect = "hour",
+#                                                  level = as.character(survcov[i,'hour']),
+#                                                  weight = 1),
+#                                      eff4 = list(effect = "hour:MFAug15",
+#                                                  level = paste(as.character(survcov[i,'hour']),",", sep = ""),
+#                                                  weight = survcov[i,'MFAug15']),
+#                                      eff5 = list(effect = "Year",
+#                                                  level = c("102", "105", "106"),
+#                                                  weight = c(0.333333, 0.333333, 0.333334))
+#                                    )
+#   )
+# }
+
+# let's remind ourselves of the model formula
+fit_ribbon$fixed.formula
+
+# estimate for solar hour 12 on day 106
+#
+# eventually, we want to estimate across all hours for a series of days
+# to create our plots. So, this would become a nested for-loop
+
+fix_solar_hour = 12
+fix_sin1 <- sin(pi*fix_solar_hour/12)
+fix_cos1 <- cos(pi*fix_solar_hour/12)
+fix_sin2 <- sin(pi*fix_solar_hour/6)
+fix_cos2 <- cos(pi*fix_solar_hour/6)
+fix_sin3 <- sin(pi*fix_solar_hour/4)
+fix_cos3 <- sin(pi*fix_solar_hour/4)
+
+# yday 106 is April 15
+fix_day <- (106-120)/10
+fix_day2 <- fix_day^2
+fix_day3 <- fix_day^3
+
+# for now, we'll just use mean wx covariates
[…]

### 2021-03-05 analysis/scripts/fe_plots.R

@@ -1,0 +1,53 @@
+library(dplyr)
+library(tibble)
+library(glue)
+library(ggplot2)
+
+load(here::here("data/fit_ribbon.Rdata"))
+
+age_sex_terms <- tibble(facet_label = "age_sex",
+                        effect_label = c("intercept","ADULT.M", "SUBADULT", "YOUNG OF YEAR")
+)
+
+wx_terms <- tibble(facet_label = "weather",
+                   effect_label = c("temp2","wind","precip","pressure")
+)
+
+time_terms <- tibble(facet_label = "time",
+                     effect_label = c("day","day2","day3","sin1","cos1","sin2","cos2","sin3","cos3")
+)
+
+interact_terms <- tibble(facet_label = "interaction",
+                         effect_label = c("ADULT.M:day", "ADULT.M:day2", "ADULT.M:day3",
+                    "SUBADULT:day","SUBADULT:day2","SUBADULT:day3",
+                    "YOUNG OF YEAR:day","YOUNG OF YEAR:day2","YOUNG OF YEAR:day3",
+                    "sin1:day","sin1:day2","cos1:day","cos1:day2",
+                    "sin2:day","sin2:day2","cos2:day","cos2:day2",
+                    "sin3:day","sin3:day2","cos3:day","cos3:day2",
+                    "temp2:wind")
+)
+
+effects_df <- bind_rows(age_sex_terms,wx_terms,time_terms,interact_terms)
+
+# this needs some help to get the order of the effect_labels right and, also,
+# reorder the facets; probably need to group_by and create an order col, then set
+# the breaks and labels manually
+
+fe <- fit_ribbon$fixed.effects %>%
+  dplyr::select(effect, levels, estimate, std.err) %>%
+  dplyr::filter(!is.na(std.err)) %>%
+  dplyr::mutate(effect_label = glue::glue("{levels}{effect}")) %>%
+  dplyr::mutate(effect_label = stringr::str_remove(effect_label, "age_sex"),
+                effect_label = stringr::str_remove(effect_label, ",")) %>%
+  dplyr::left_join(effects_df) %>%
+  dplyr::mutate(effect_label = forcats::as_factor(effect_label) %>%
+                  forcats::fct_reorder(c(age_sex_terms$effect_label,
+                                                       wx_terms$effect_label,
+                                                       time_terms$effect_label,
+                                                       interact_terms$effect_label))) %>%
[…]

### 2021-03-05 analysis/scripts/get_data/make_data.R

@@ -1,0 +1,29 @@
+
+source(here('R/get_locs_sf.R'))
+source(here('R/get_timeline_data.R'))
+source(here('R/create_db_input_data.R'))
+
+timeline_data = get_timeline_data()
+locs_sf = get_locs_sf()
+db_input_data = create_db_input_data(timeline_data, locs_sf)
+
+save(db_input_data, file = here::here("data/tbl_percent_locs.Rdata"))
+
+stopifnot(
+  "PEP Postgres Database Not Available; did you start VPN? ;)" =
+    is_up("161.55.120.122", "5432")
+)
+
+con <- dbConnect(
+  odbc(),
+  dsn = "PostgreSQL pep",
+  uid = get_kc_account("pgpep_londonj"),
+  pwd = decrypt_kc_pw("pgpep_londonj")
+)
+
+st_write(obj = db_input_data,
+         dsn = con,
+         layer = "res_iceseal_haulout_jml",
+         layer_options = "OVERWRITE=true")
+
+dbDisconnect(con)

### 2021-03-05 analysis/scripts/model.R

@@ -1,0 +1,153 @@
+library(tidyverse)
+library(here)
+library(lubridate)  #some date time functionality
+library(glmmLDTS)
+library(solaR)  #for getting hour of day in solar time
+library(splines)
+library(sf)
+library(DBI)
+library(rnaturalearth)
+library(purler)
+
+create_ar1_id <- function(h_dt) {
+  ar1_id <- tibble(
+    t_seq = seq(from = h_dt[1], to = h_dt[length(h_dt)], by = "hour",)
+  ) %>% left_join(tibble(h_dt = h_dt, haulout_dt = h_dt), by = c("t_seq" = "h_dt")) %>%
+    mutate(x = is.na(haulout_dt),
+           rle = rlenc_id(x)) %>%
+    filter(!is.na(haulout_dt)) %>%
+    pull(rle)
+
+  return(ar1_id)
+}
+
+load(here::here('data/dat_sf.Rdata'))
+
+dat.sf <- dat.sf  %>%
+  mutate(dry = round(percent_dry/100)) %>%
+  mutate(hour_utc = hour(haulout_dt),
+         yday = yday(haulout_dt),
+         year = year(haulout_dt)) %>%
+  mutate(coords_x = st_coordinates(st_transform(.,4326))[,1],
+         coords_y = st_coordinates(st_transform(.,4326))[,2],
+         solar_hour = solaR::local2Solar(haulout_dt,coords_x) %>% lubridate::hour(),
+         sin1 = sin(pi*solar_hour/12),
+         cos1 = cos(pi*solar_hour/12),
+         sin2 = sin(pi*solar_hour/6),
+         cos2 = cos(pi*solar_hour/6),
+         sin3 = sin(pi*solar_hour/4),
+         cos3 = sin(pi*solar_hour/4),
+         day = (yday-120)/10,
+         day2 = day^2,
+         day3 = day^3
+         ) %>%
+  mutate(precip = rast_acpcp,
+         temp2 = (rast_air2m-270)/27,
+         temp0 = (rast_airsfc-270)/27,
+         pressure = (rast_prmsl-100000)/10000,
+         wind = sqrt(rast_uwnd^2 + rast_vwnd^2)/10
+         ) %>%
+  mutate(northing = st_coordinates(.)[,2]/mean(st_coordinates(.)[,2]),
+         north_of_bering = case_when(
+           coords_y > 66.5 ~ TRUE,
+           TRUE ~ FALSE
+         )) %>%
[…]

### 2021-03-05 analysis/scripts/model_year.R

@@ -1,0 +1,106 @@
+library(tidyverse)
+library(here)
+library(lubridate)  #some date time functionality
+library(glmmLDTS)
+library(solaR)  #for getting hour of day in solar time
+library(splines)
+library(sf)
+library(DBI)
+library(rnaturalearth)
+library(purler)
+
+create_ar1_id <- function(h_dt) {
+  ar1_id <- tibble(
+    t_seq = seq(from = h_dt[1], to = h_dt[length(h_dt)], by = "hour",)
+  ) %>% left_join(tibble(h_dt = h_dt, haulout_dt = h_dt), by = c("t_seq" = "h_dt")) %>%
+    mutate(x = is.na(haulout_dt),
+           rle = rlenc_id(x)) %>%
+    filter(!is.na(haulout_dt)) %>%
+    pull(rle)
+
+  return(ar1_id)
+}
+
+load(here::here('data/dat_sf.Rdata'))
+
+dat.sf <- dat.sf  %>%
+  mutate(dry = round(percent_dry/100)) %>%
+  mutate(hour_utc = hour(haulout_dt),
+         yday = yday(haulout_dt),
+         year = year(haulout_dt)) %>%
+  mutate(coords_x = st_coordinates(st_transform(.,4326))[,1],
+         coords_y = st_coordinates(st_transform(.,4326))[,2],
+         solar_hour = solaR::local2Solar(haulout_dt,coords_x) %>% lubridate::hour(),
+         sin1 = sin(pi*solar_hour/12),
+         cos1 = cos(pi*solar_hour/12),
+         sin2 = sin(pi*solar_hour/6),
+         cos2 = cos(pi*solar_hour/6),
+         sin3 = sin(pi*solar_hour/4),
+         cos3 = sin(pi*solar_hour/4),
+         day = (yday-120)/10,
+         day2 = day^2,
+         day3 = day^3
+  ) %>%
+  mutate(precip = rast_acpcp,
+         temp2 = (rast_air2m-270)/27,
+         temp0 = (rast_airsfc-270)/27,
+         pressure = (rast_prmsl-100000)/10000,
+         wind = sqrt(rast_uwnd^2 + rast_vwnd^2)/10
+  ) %>%
+  mutate(northing = st_coordinates(.)[,2]/mean(st_coordinates(.)[,2]),
+         north_of_bering = case_when(
+           coords_y > 66.5 ~ TRUE,
+           TRUE ~ FALSE
+         )) %>%
[…]

### 2021-03-05 analysis/scripts/model_year_pl.R

@@ -1,0 +1,106 @@
+library(tidyverse)
+library(here)
+library(lubridate)  #some date time functionality
+library(glmmLDTS)
+library(solaR)  #for getting hour of day in solar time
+library(splines)
+library(sf)
+library(DBI)
+library(rnaturalearth)
+library(purler)
+
+create_ar1_id <- function(h_dt) {
+  ar1_id <- tibble(
+    t_seq = seq(from = h_dt[1], to = h_dt[length(h_dt)], by = "hour",)
+  ) %>% left_join(tibble(h_dt = h_dt, haulout_dt = h_dt), by = c("t_seq" = "h_dt")) %>%
+    mutate(x = is.na(haulout_dt),
+           rle = rlenc_id(x)) %>%
+    filter(!is.na(haulout_dt)) %>%
+    pull(rle)
+
+  return(ar1_id)
+}
+
+load(here::here('data/dat_sf.Rdata'))
+
+dat.sf <- dat.sf  %>%
+  mutate(dry = round(percent_dry/100)) %>%
+  mutate(hour_utc = hour(haulout_dt),
+         yday = yday(haulout_dt),
+         year = year(haulout_dt)) %>%
+  mutate(coords_x = st_coordinates(st_transform(.,4326))[,1],
+         coords_y = st_coordinates(st_transform(.,4326))[,2],
+         solar_hour = solaR::local2Solar(haulout_dt,coords_x) %>% lubridate::hour(),
+         sin1 = sin(pi*solar_hour/12),
+         cos1 = cos(pi*solar_hour/12),
+         sin2 = sin(pi*solar_hour/6),
+         cos2 = cos(pi*solar_hour/6),
+         sin3 = sin(pi*solar_hour/4),
+         cos3 = sin(pi*solar_hour/4),
+         day = (yday-120)/10,
+         day2 = day^2,
+         day3 = day^3
+  ) %>%
+  mutate(precip = rast_acpcp,
+         temp2 = (rast_air2m-270)/27,
+         temp0 = (rast_airsfc-270)/27,
+         pressure = (rast_prmsl-100000)/10000,
+         wind = sqrt(rast_uwnd^2 + rast_vwnd^2)/10
+  ) %>%
+  mutate(northing = st_coordinates(.)[,2]/mean(st_coordinates(.)[,2]),
+         north_of_bering = case_when(
+           coords_y > 66.5 ~ TRUE,
+           TRUE ~ FALSE
+         )) %>%
[…]

### 2021-03-05 analysis/scripts/predict.R

@@ -1,0 +1,385 @@
+library(glmmLDTS)
+library(mgcv)
+library(dplyr)
+library(forcats)
+library(ggplot2)
+
+# get the ribbon seal fit object
+fit_ribbon <- readRDS(here::here("data/fit_ribbon.Rdata"))
+
+# function to create a prediction data frame for a given age_sex group
+create_newdata <- function(data, age_sex) {
+  df_list <- vector(mode = "list", length = length({{age_sex}} ))
+
+  for (a_s in {{age_sex}}) {
+  range_yday <- data %>% filter(age_sex == a_s ) %>%
+    summarize(start_day = min(yday),
+              end_day = max(yday))
+  start_day <- range_yday$start_day
+  end_day <- range_yday$end_day
+  n_days = (end_day - start_day) + 1
+
+  # for wx covariates we'll use a gam to get values by day/hour since wx is likely
+  # to vary w/in day over the season
+
+  gam.baro <-
+    gam(pressure ~ s(yday), data = data)
+  gam.temp <-
+    gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), data = data)
+  gam.wind <-
+    gam(wind ~ s(yday) + s(as.numeric(solar_hour)), data = data)
+  gam.precip <-
+    gam(precip ~ s(yday), data = data)
+
+  wx_new_data <- data.frame(
+    solar_hour = rep(0:23, each = n_days),
+    yday = rep(start_day:end_day, times = 24)
+  )
+
+  temp_pred <- predict(gam.temp, newdata = wx_new_data)
+  wind_pred <- predict(gam.wind, newdata = wx_new_data)
+  baro_pred <- predict(gam.baro, newdata = wx_new_data)
+  precip_pred <- predict(gam.precip, newdata = wx_new_data)
+
+  df <- data.frame(
+    age_sex = a_s,
+    solar_hour = rep(0:23, each = n_days),
+    yday = rep(start_day:end_day, times = 24),
+    northing = mean(data$northing),
+    temp2 = temp_pred,
+    wind = wind_pred,
+    pressure = baro_pred,
+    precip = precip_pred
+  ) %>%
+    mutate(
+      sin1 = sin(pi * solar_hour / 12),
+      cos1 = cos(pi * solar_hour / 12),
+      sin2 = sin(pi * solar_hour / 6),
+      cos2 = cos(pi * solar_hour / 6),
[…]

### 2021-03-05 analysis/scripts/predict_year.R

@@ -1,0 +1,182 @@
+library(tidyverse)
+library(lubridate)
+library(mgcv)
+library(glmmLDTS)
+
+# function to create a prediction data frame for a given age_sex group
+create_newdata_year <- function(data, age_sex) {
+  df_list <- vector(mode = "list", length = length({{age_sex}} ))
+  df_list_years <- vector(mode = "list", length = length(levels(data$year)))
+
+  for (a_s in {{age_sex}}) {
+  for (y in levels(data$year)) {
+    d <- data %>% filter(year == y, age_sex == a_s)
+    if(nrow(d) == 0) { next }
+    range_yday <- d %>%
+      summarize(start_day = min(yday),
+                end_day = max(yday))
+    if(range_yday$start_day > 152) {next}
+    if(range_yday$end_day < 121) {next}
+    print(range_yday)
+    start_day <- range_yday$start_day
+    end_day <- range_yday$end_day
+    n_days = (end_day - start_day) + 1
+
+    # for wx covariates we'll use a gam to get values by day/hour since wx is likely
+    # to vary w/in day over the season
+
+    gam.baro <-
+      gam(pressure ~ s(yday), data = data)
+    gam.temp <-
+      gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), data = data)
+    gam.wind <-
+      gam(wind ~ s(yday) + s(as.numeric(solar_hour)), data = data)
+    gam.precip <-
+      gam(precip ~ s(yday), data = data)
+
+    wx_new_data <- data.frame(
+      solar_hour = 12,
+      yday = start_day:end_day
+    )
+
+    temp_pred <- predict(gam.temp, newdata = wx_new_data)
+    wind_pred <- predict(gam.wind, newdata = wx_new_data)
+    baro_pred <- predict(gam.baro, newdata = wx_new_data)
+    precip_pred <- predict(gam.precip, newdata = wx_new_data)
+
+    df <- data.frame(
+      age_sex = a_s,
+      year = y,
+      solar_hour = 12,
+      yday = start_day:end_day,
+      northing = mean(data$northing),
+      temp2 = temp_pred,
+      wind = wind_pred,
+      pressure = baro_pred,
+      precip = precip_pred
+    ) %>%
+      mutate(
+        sin1 = sin(pi * solar_hour / 12),
[…]

### 2021-03-05 analysis/scripts/seaice_extent.R

@@ -1,0 +1,41 @@
+library(tidyverse)
+library(lubridate)
+library(glue)
+library(DBI)
+library(dbplyr)
+library(sf)
+
+load(here::here("data/dat_sf.Rdata"))
+
+xmin <- st_bbox(dat.sf)$xmin %>% as.integer()
+ymin <- st_bbox(dat.sf)$ymin %>% as.integer()
+xmax <- st_bbox(dat.sf)$xmax %>% as.integer()
+ymax <- st_bbox(dat.sf)$ymax %>% as.integer()
+epsg <- 3571
+
+con <- dbConnect(
+  odbc::odbc(),
+  dsn = "PostgreSQL pep",
+  uid = keyringr::get_kc_account("pgpep_londonj"),
+  pwd = keyringr::decrypt_kc_pw("pgpep_londonj"),
+  sslmode = "disable"
+)
+
+extent_list <- vector(mode = "list")
+
+for(y in 2005:2020) {
+  print(y)
+  start_date <- glue::glue('{y}-02-15')
+  end_date <- glue::glue('{y}-07-15')
+
+  sql<- glue_sql("
+           SELECT *
+  FROM environ.fxn_sea_ice_extent_by_date(
+    0.15,{xmin},{ymin},{xmax},{ymax},
+    {epsg},
+    {start_date},{end_date} )",
+  .con = con)
+  extent_list[[as.character(y)]] <- DBI::dbGetQuery(con, sql)
+}
+
+dbDisconnect(con)

### 2021-03-05 analysis/paper/paper.Rmd
REMOVE LARGE WX COV FILE B/c >100MB

@@
 
-load(file = here::here('data/cov_means_daily.Rdata'))
+load(file = here::here('data/cov_means_daily_2102.Rdata'))
 grid_plot <- cov_means_daily %>% filter(year == 2012,

### 2021-03-05 README.Rmd
SIMPLIFY README

@@
 
-<!-- [![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio) -->
-
 This repository contains the data and code for our paper:
-
-<!-- > Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx> -->
-
-<!-- Our pre-print is online here: -->
-
-<!-- > Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx> -->
-
-<!-- ### How to cite -->
-
-<!-- Please cite this compendium as: -->
-
-<!-- > Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx> -->
+*Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas*
 
@@
     and tables in this file)
-  - [:file\_folder: data](/analysis/data): Data used in the analysis.
-  - [:file\_folder: figures](/analysis/figures): Plots and other
-    illustrations
-  - [:file\_folder:
-    supplementary-materials](/analysis/supplementary-materials):
-    Supplementary materials including notes and other documents
-    prepared and collected during the analysis.
+  - [:file\_folder: figures](/analysis/figures): figures generated
+  - [:file\_folder: scripts](/analysis/scripts): R scripts in support of analysis.
+  - [:file\_folder: templates](/analysis/templates): template files and scripts.
+
+The **data** directory contains
+
+  - [:file\_folder: data](/data): Data used or generated in analysis
 
@@
 
-The simplest way (not yet enabled) to explore the text, code and data is to click on
+The simplest way (*not yet enabled*) to explore the text, code and data is to click on
 [binder](https://mybinder.org/v2/gh///master?urlpath=rstudio)
@@
 environment. These Docker images can also be used locally.
-
-<!-- You can download the compendium as a zip from from this URL: -->
-<!-- [master.zip](/archive/master.zip). After unzipping: -->
-<!-- - open the `.Rproj` file in RStudio -->
-<!-- - run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the -->
-<!-- [DESCRIPTION](/DESCRIPTION) file). -->
-<!-- - finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console -->

### 2021-03-05 analysis/paper/paper.Rmd
REMOVE REF TO LARGE WX COV MEANS

--- 2021-03-05 analysis/paper/paper.Rmd
@@
 
-load(file = here::here('data/cov_means_daily.Rdata'))
-grid_plot <- cov_means_daily %>% filter(year == 2012,
+load(file = here::here('data/cov_means_daily_2012.Rdata'))
+grid_plot <- cov_means_daily_2012 %>% filter(year == 2012,
                            yday == 125) %>%

### 2021-03-25 analysis/paper/paper.Rmd
FINAL EDITS BEFORE SHARING W/ CO-AUTH

@@
 early summer (April-June) as sea ice begins to melt and retreat northward. Rapid
-acceleration of climate change in Arctic ecosystems is thus of concern, as the quantity
-and quality of suitable habitat is forecast to decrease. In this paper, we use data
-collected from satellite-linked telemetry tags deployed between 2005 and 2020 to
-investigate the seasonal timing and environmental factors affecting sea ice use by seals
-(specifically, bearded, ribbon, and spotted seals) in the Bering, and Chukchi seas. In
-addition to providing baseline data on phenology, these data also allow us to refine
-availability estimates needed to accurately estimate abundance from aerial survey counts
-of seals basking on ice (i.e., to correct for the proportion of animals that are in the
-water while surveys are conducted). Using generalized linear mixed pseudo-models to
-properly account for temporal autocorrelation, we fit models with a variety of covariates
-(e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure,
-temperature, precipitation) to examine their ability to explain variation in hourly
-haul-out records. We found evidence for strong diurnal and within season patterns in
-haul-out behavior, as well as strong weather effects (particularly wind and temperature).
-In general, seals were more likely to bask on ice in the middle of the day and when wind
-speed was low and temperatures were higher. Haul-out propensity increased through March
-and April, peaking in May and early June before declining again. The timing and frequency
-of haul-out events also varied substantially based on species and age-sex class. For
-ribbon and spotted seals, models with year effects were highly supported, indicating that
-the timing and magnitude of haul-out behavior varied among years. Our
-analysis also emphasizes the importance of accounting for factors affecting haul-out
-behavior when interpreting the number of seals counted in aerial surveys.
+acceleration of climate change in Arctic ecosystems is thus of concern as the quantity and
+quality of suitable habitat is forecast to decrease. In this paper, we use data collected
+from satellite-linked telemetry tags deployed between 2005 and 2020 to investigate the
+seasonal timing and environmental factors affecting sea ice use by seals (specifically,
+bearded, ribbon, and spotted seals) in the Bering, and Chukchi seas. In addition to
+providing baseline data on phenology, these data also allow us to refine availability
+estimates needed to accurately estimate abundance from aerial survey counts of seals
+basking on ice (i.e., to correct for the proportion of animals that are in the water while
+surveys are conducted). Using generalized linear mixed pseudo-models to properly account
+for temporal autocorrelation, we fit models with a variety of covariates (e.g.,
+day-of-year, solar hour, age-sex class, wind speed, barometric pressure, temperature,
+precipitation) to examine their ability to explain variation in hourly haul-out records.
+We found evidence for strong diurnal and within season patterns in haul-out behavior, as
+well as strong weather effects (particularly wind and temperature). In general, seals were
+more likely to rest on ice in the middle of the day and when wind speed was low and
+temperatures were higher. Haul-out propensity increased through March and April, peaking
+in May and early June before declining again. The timing and frequency of haul-out events
+also varied based on species and age-sex class. For ribbon and spotted seals, models with
+year effects were highly supported, indicating that the timing and magnitude of haul-out
+behavior varied among years. There was no support linking the variability in haul-out
+timing to annual sea ice extent. Our analysis also emphasizes the importance of accounting
+for weather factors affecting haul-out behavior when interpreting the number of seals
+counted in aerial surveys.
 
@@
 ```{r setup, include = FALSE}
+library(conflicted)
 library(tidyverse)
+conflict_prefer("filter", "dplyr")
+conflict_prefer("select", "dplyr")
 library(lubridate)
+conflict_prefer("month","lubridate")
 library(glue)
@@
   )
+
+options(scipen = 3, digits = 3)
 
@@
 deployment by experts in the field. Seals determined to be less than one year were
-classified as young of the year. subadults are those seals likely greater than one year
-of age but not yet sexually mature. Adults are likely sexually mature and older than
+classified as young of the year. subadults are those seals likely greater than one year of
+age but not yet sexually mature. Adults are likely sexually mature and older than
 approximately four years. For those tags deployed on young of the year and transmitting
@@
 daily locations. We also compiled several meteorological variables shown to affect
-haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz]. In
+haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz; @perry2017]. In
 particular, we linked locations from satellite tags to key weather values from the North
@@
 temperature, wind speed, barometric pressure, precipitation, and wind chill (represented
-by a wind $\times$ temperature interaction; [@udevitz2009cjz]), together with day-of-year and
-time of day interactions to permit diurnal patterns to change throughout the year).
+by a wind $\times$ temperature interaction; [@udevitz2009cjz]), together with day-of-year
+and time of day interactions to permit diurnal patterns to change throughout the year).
 Spotted and ribbon seal models included age-sex class and interactions between day-of-year
@@
 occupy a substantial range and we were interested in possible differences in the timing of
-haul-out behavior along a latitudinal gradient. We omitted the latitudinal effect from ribbon and
-spotted seal models since, during the spring, these species are most prevalent near the
-southern ice edge in the Bering Sea [@conn2014mee].
+haul-out behavior along a latitudinal gradient. We omitted the latitudinal effect from
+ribbon and spotted seal models since, during the spring, these species are most prevalent
+near the southern ice edge in the Bering Sea [@conn2014mee].
 
@@
 representation of sea ice concentration, area, or extent. This may seem counter-intuitive
-when modeling the haul-out behavior of seal species with such a close association to sea ice.
-Seals haul out in the presence of sea ice and we could assess the local concentration of
-sea ice during these events. However, seals also spend time in the water away from
-sea ice. Thus, the challenge is how best to associate that seal's at-sea behavior with a
-sea ice concentration. Would it be the concentration at the last known haul-out event? The
+when modeling the haul-out behavior of seal species with such a close association to sea
+ice. Seals haul out in the presence of sea ice and we could assess the local concentration
+of sea ice during these events. However, seals also spend time in the water away from sea
+ice. Thus, the challenge is how best to associate that seal's at-sea behavior with a sea
+ice concentration. Would it be the concentration at the last known haul-out event? The
 next known haul-out event? Or, the closest sea ice concentration available? Without a
@@
 influences the peak timing of seal haul-out behavior. For this, all sea ice concentration
-grid cells (25 $km^2$) in the study area with greater than 15% concentration were summed to
+grid cells (25 km^2^) in the study area with greater than 15% concentration were summed to
 get total sea ice extent for each day between 15 February and 15 July across all years
@@
 models with year by day-of-year interactions. However, in this case we only included
-year$\times$day and year$\times$day2, omitting the main effects of year as well as year$\times$day3
-interactions as models with the latter effects were numerically unstable. However, the
-modeled interactions were sufficient to allow shifts in haul-out distribution, as one can
-show mathematically that a simple horizontal shift in timing of haul-out distributions
-does not affect the main effects or cubic terms in a polynomial regression model.
+year$\times$day and year$\times$day2, omitting the main effects of year as well as
+year$\times$day3 interactions as models with the latter effects were numerically unstable.
+However, the modeled interactions were sufficient to allow shifts in haul-out
+distribution, as one can show mathematically that a simple horizontal shift in timing of
+haul-out distributions does not affect the main effects or cubic terms in a polynomial
+regression model.
 
@@
 weather covariates, and determine annual peaks in haul-out activity. The timing of peak
-haul-out behavior was further used to regress against the annual maximum sea ice extent 
-in the study region.
+haul-out behavior was further used to regress against the annual maximum sea ice extent in
+the study region.
 
 ## Stable stage distributions
+
+::: {custom-style="comment"}
+I didn't update this at all from Paul's previous text other than to revise some of the
+latex equations and add references. Would be good to confirm all of this is still
+appropriate and makes sense.
+:::
 
@@
 Ribbon seals show a pattern of gradually increasing haul-out probability in April that
-peaks in May for subadults and in June for adults. The behavior is clearly centered
-around local solar noon and expands to other hours later in the season as seals enter
-their molting period. subadults showed an earlier start and more intense haul-out
-activity in April and May. The young of the year records begin after weaning and the model
-predictions demonstrate the ontogeny of in-water activities (e.g. diving, foraging) in
-May. Adult females have a more protracted season compared to males and more time spent
-resting in June and July. The model suggests adult male ribbon seals complete their molt
-by the end of June. Lastly, for adults of both sexes there is some indication for a shift
-to a nocturnal haul-out pattern in late June and July.
+peaks in May for subadults and in June for adults. The behavior is clearly centered around
+local solar noon and expands to other hours later in the season as seals enter their
+molting period. Subadults showed an earlier start and more intense haul-out activity in
+April and May. The young of the year records begin after weaning and the model predictions
+demonstrate the ontogeny of in-water activities (e.g. diving, foraging) in May. Adult
+females have a more protracted season compared to males and more time spent resting in
+June and July. The model suggests adult male ribbon seals complete their molt by the end
+of June. Lastly, for adults of both sexes there is some indication for a shift to a
+nocturnal haul-out pattern in late June and July.
 
@@
 
+```{r}
+ribbon_type3 <- tibble(
+  fit_ribbon$typeIII.hypoth
+) %>% 
+  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
+         Effect = janitor::make_clean_names(Effect)) %>% 
+  column_to_rownames(var = "Effect") 
+ribbon_type3 <- setNames(split(ribbon_type3, 
+                               seq(nrow(ribbon_type3))), 
+                         rownames(ribbon_type3))
+
+spotted_type3 <- tibble(
+  fit_spotted$typeIII.hypoth
+) %>% 
+  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
+         Effect = janitor::make_clean_names(Effect)) %>%  
+  column_to_rownames(var = "Effect")  
+spotted_type3 <- setNames(split(spotted_type3, 
+                               seq(nrow(spotted_type3))), 
+                         rownames(spotted_type3))
+bearded_type3 <- tibble(
+  fit_bearded$typeIII.hypoth
+) %>% 
+  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
+         Effect = janitor::make_clean_names(Effect)) %>%  
+  column_to_rownames(var = "Effect")  
+bearded_type3 <- setNames(split(bearded_type3, 
+                               seq(nrow(bearded_type3))), 
+                         rownames(bearded_type3))
+```
+
 The haul-out probability for ribbon seals was strongly influenced by each of the weather
-covariates included within the model. The impact of temperature, wind, and precipitation
-was especially noticeable as ribbon seals were less likely to haul out at higher winds and
-higher precipitation rates and more likely to be on the ice when temperatures were
-relatively warm. Wind chill (temperature $\times$ wind) also had a negative influence on haul-out
-probability.
+covariates included within the model except precipitation. The impact of temperature
+($F_{`r paste(ribbon_type3$temp2$Num.df, ribbon_type3$temp2$Den.df, sep=",")`}$ = 
+`r ribbon_type3$temp2$F.value`; $p$ = `r ribbon_type3$temp2$Prob.F`), wind
+($F_{`r paste(ribbon_type3$wind$Num.df, ribbon_type3$wind$Den.df, sep=",")`}$ = 
+`r ribbon_type3$wind$F.value`; $p$ = `r ribbon_type3$wind$Prob.F`), and barometric pressure
+($F_{`r paste(ribbon_type3$pressure$Num.df, ribbon_type3$pressure$Den.df, sep=",")`}$ = 
+`r ribbon_type3$pressure$F.value`; $p$ = `r ribbon_type3$pressure$Prob.F`) was especially
+noticeable as ribbon seals were less likely to haul out at higher winds and lower pressure
+values and more likely to be on the ice when temperatures were relatively warm. Wind chill
+($F_{`r paste(ribbon_type3$temp2_wind$Num.df, ribbon_type3$temp2_wind$Den.df, sep=",")`}$ 
+= `r ribbon_type3$temp2_wind$F.value`; $p$ = `r ribbon_type3$temp2_wind$Prob.F`) showed a
+moderately negative influence on haul-out probability.
 
 ```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Predicted haul-out probability of ribbon seals across a range of weather conditions."}
-library(ggsci)
 
@@
 
-Compared to ribbon seals, spotted seals showed a longer spring haul-out season that is less
-intensely centered on solar noon. Adults of both sexes spend considerable time in April
-and May resting on the ice. As with ribbon seals, the young of the year records begin
-after weaning and the model predictions demonstrate the ontogeny of in-water activities
-(e.g. diving, foraging) in May.
+Compared to ribbon seals, spotted seals showed a longer spring haul-out season that is
+less intensely centered on solar noon. Adults of both sexes spend considerable time in
+April and May resting on the ice. As with ribbon seals, the young of the year records
+begin after weaning and the model predictions demonstrate the ontogeny of in-water
+activities (e.g. diving, foraging) in May.
 
@@
 
-Spotted seal haul-out behavior appears most strongly influenced by temperature and wind.
-Spotted seals were less likely to haul out at higher winds and more likely to be on the
-ice when temperatures were relatively warm. Wind chill (temperature $\times$ wind) also had a
-negative influence on haul-out probability. Barometric pressure and precipitation were
-influential to a slightly lesser extent. Differences in the magnitude of response between
-the age-sex classes is present and consistent across each of the weather covariates.
+Spotted seal haul-out behavior appears most strongly influenced by barometric pressure
+($F_{`r paste(spotted_type3$pressure$Num.df, spotted_type3$pressure$Den.df, sep=",")`}$ = 
+`r spotted_type3$pressure$F.value`; $p$ = `r spotted_type3$pressure$Prob.F`) and wind
+($F_{`r paste(spotted_type3$wind$Num.df, spotted_type3$wind$Den.df, sep=",")`}$ =
+`r spotted_type3$wind$F.value`; $p$ = `r spotted_type3$wind$Prob.F`) with temperature
+having a moderate influence
+($F_{`r paste(spotted_type3$temp2$Num.df, spotted_type3$temp2$Den.df, sep=",")`}$ =
+`r spotted_type3$temp2$F.value`; $p$ = `r spotted_type3$temp2$Prob.F`). Spotted seals were
+less likely to haul out at higher winds and more likely to be on the ice when temperatures
+were relatively warm. Wind chill (temperature $\times$ wind) and precipitation were not as
+influential. Differences in the magnitude of response between the age-sex classes is
+present and consistent across each of the weather covariates.
 
@@
 
-Like ribbon and spotted seals, bearded seal haul-out probability appears influenced by
-wind and temperature. And, the model also predicted a response to wind chill. Any
+Like ribbon and spotted seals, bearded seal haul-out probability appears most influenced
+by wind ($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$ =
+`r bearded_type3$wind$F.value`; $p$ = `r bearded_type3$wind$Prob.F`) and temperature
+($F_{`r paste(bearded_type3$temp2$Num.df, bearded_type3$temp2$Den.df, sep=",")`}$ =
+`r bearded_type3$temp2$F.value`; $p$ = `r bearded_type3$temp2$Prob.F`). Additionally, wind
+chill was also a significant influence
+($F_{`r paste(bearded_type3$temp2_wind$Num.df, bearded_type3$temp2_wind$Den.df, sep=",")`}$ 
+= `r bearded_type3$temp2_wind$F.value`; $p$ = `r bearded_type3$temp2_wind$Prob.F`). Any
 influence of barometric pressure or precipitation is less apparent.
@@
 ```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Predicted haul-out probability of bearded seals across a range of weather conditions."}
-library(ggsci)
 
@@
 
-<!-- The second set of models that included annual variation in haul-out patterns produced -->
-<!-- p-values that were strongly significant for linear (ribbon seals: F10,76662=9.92, -->
-<!-- p\<0.0001; spotted seals: F8,51981=5.23, p\<0.0001) and quadratic (ribbon seals: -->
-<!-- F10,76662=8.02, p\<0.0001; spotted seals: F8,51981=3.79, p=0.0002) year-by-day -->
-<!-- interactions. Predicted distributions of haul-out activity were largely unimodal, but -->
-<!-- varied substantially among and within years with respect to both the timing and magnitude -->
-<!-- of haul-out peaks (Figure \@ref(fig:annualHO)). For instance, in 2010 and 2016 spotted -->
-<!-- seal haul-out activity peaked in early May, while in other years (e.g. 2011) it appeared -->
-<!-- to peak three weeks later. It is important to note that predicted variation in annual -->
-<!-- haul-out patterns likely reflects both process error and sampling variability. For -->
-<!-- instance, some of the more extreme patterns (e.g. ribbon seals in 2012 , spotted seals in -->
-<!-- 2014) were for years with only one tagged seal and are unlikely to represent -->
-<!-- population-level processes. -->
+```{r}
+ribbon_year_type3 <- tibble(
+  fit_ribbon_year$typeIII.hypoth
+) %>% 
+  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
+         Effect = janitor::make_clean_names(Effect)) %>% 
+  column_to_rownames(var = "Effect") 
+ribbon_year_type3 <- setNames(split(ribbon_year_type3, 
+                               seq(nrow(ribbon_year_type3))), 
+                         rownames(ribbon_year_type3))
+
+spotted_year_type3 <- tibble(
+  fit_spotted_year$typeIII.hypoth
+) %>% 
+  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
+         Effect = janitor::make_clean_names(Effect)) %>%  
+  column_to_rownames(var = "Effect")  
+spotted_year_type3 <- setNames(split(spotted_year_type3, 
+                               seq(nrow(spotted_year_type3))), 
+                         rownames(spotted_year_type3))
+```
+
+The second set of models that included annual variation in haul-out patterns uncovered
+significant contributions for linear and quadratic interactions between day and year in
+both ribbon seals (day:year,
+$F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$ = 
+`r ribbon_year_type3$day_year$F.value`; $p$ = `r ribbon_year_type3$day_year$Prob.F`;
+day^2^:year,
+$F_{`r paste(ribbon_year_type3$day2_year$Num.df, ribbon_year_type3$day2_year$Den.df, sep=",")`}$ = 
+`r ribbon_year_type3$day2_year$F.value`; $p$ = `r ribbon_year_type3$day2_year$Prob.F`) and
+spotted seals (day:year,
+$F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$ = 
+`r spotted_year_type3$day_year$F.value`; $p$ = `r spotted_year_type3$day_year$Prob.F`;
+day^2^:year,
+$F_{`r paste(spotted_year_type3$day2_year$Num.df, spotted_year_type3$day2_year$Den.df, sep=",")`}$ = 
+`r spotted_year_type3$day2_year$F.value`; $p$ = `r spotted_year_type3$day2_year$Prob.F`).
+Predicted distributions of haul-out activity were largely unimodal, but varied some among
+and within years with respect to the timing and magnitude of haul-out peaks (Figure
+\@ref(fig:annualHO)). It is important to note that predicted variation in annual haul-out
+patterns likely reflects both process error and sampling variability. While we did remove
+any years where only one deployment in a species + age-sex group was present, there are
+still some years where the pattern shown is informed by a small number of tags that may
+not represent population level processes.
 
@@
 
-The timing of peak haul-out probability for ribbon and spotted seals appeared to have 
-only a limited to no relationship with the amount of yearly maximum sea ice extent. Adult female and
-subadult spotted seals show a negative trend line but no indication that the observed
-trend is meaningful (R-squared and p-values for adult female spotted seals 
-($R^{2}$: `r formatC(pl_adultF$r.squared, digits=3)`; $p$: 
-`r formatC(pl_adultF$p.value, digits=3)`) and
-spotted seal subadults($R^{2}$: `r formatC(pl_subadult$r.squared, digits=3)`; 
-$p$: `r formatC(pl_subadult$p.value, digits=3)`)). 
-For ribbon seals and adult male spotted seals, p-values were
-substantially larger than 0.05 (R-squared and p-values for ribbon seals (adult
-females($R^{2}$: `r formatC(hf_adultF$r.squared, digits=3)`; $p$: 
-`r formatC(hf_adultF$p.value, digits=3)`), 
-adult males($R^{2}$: `r formatC(hf_adultM$r.squared, digits=3)`; $p$: 
-`r formatC(hf_adultM$p.value, digits=3)`), 
-subadults($R^{2}$: `r formatC(hf_subadult$r.squared, digits=3)`; $p$: `r formatC(hf_subadult$p.value, digits=3)`)) and 
-spotted seals (adult males($R^{2}$: `r formatC(pl_adultM$r.squared, digits=3)`; $p$: `r formatC(pl_adultM$p.value, digits=3)`)).
+The timing of peak haul-out probability for ribbon and spotted seals appeared to have only
+a limited to no relationship with the amount of yearly maximum sea ice extent. Adult
+female and subadult spotted seals show a negative trend line but no indication that the
+observed trend is meaningful (spotted seal adult female: $R^{2}$ =
+`r pl_adultF$r.squared`, $p$ = `r pl_adultF$p.value`; spotted seal subadults: $R^{2}$ =
+`r pl_subadult$r.squared`, $p$ = `r pl_subadult$p.value`). For ribbon seals and adult male
+spotted seals, $p$-values were substantially larger than 0.05 (ribbon seal adult females;
+$R^{2}$ = `r hf_adultF$r.squared`, $p$ = `r hf_adultF$p.value`; ribbon seal adult males:
+$R^{2}$ = `r hf_adultM$r.squared`, $p$ = `r hf_adultM$p.value`; ribbon seal subadults:
+$R^{2}$ = `r hf_subadult$r.squared`, $p$ = `r hf_subadult$p.value`; spotted seals adult
+males: $R^{2}$ = `r pl_adultM$r.squared`, $p$ = `r pl_adultM$p.value`).
 
@@
 ## Stable stage distributions
+
+::: {custom-style="comment"}
+This section is unchanged from Paul's original write up. I have put off revisiting and
+integrating with my current document/code. Maybe between Paul/Irina/Me we can sort out a
+quick update. Might just need to walk me through the previous code and make sure I get it
+properly incorporated with this version
+:::
 
@@
 
-In this paper, we fit statistical models to data from sensors on satellite tagged
-ice-associated seals in order to examine factors affecting the probability that Arctic
-seals haul out on ice. Such probabilities are important for understanding the timing and
-ecology of seal behavior, since haul out peaks are associated with important life history
-functions such as breeding, pupping and molting. Knowledge of factors affecting haul-out
-probabilities is a first step in being able to predict possible impacts of losses of sea
-ice on ice-associated seals. Understanding the factors affecting haul-out probabilities is
-also necessary to properly interpret seal counts made during aerial surveys. For instance,
-a count might be low because there are few seals, or because the timing and conditions are
-such that many seals are in the water.
-
-Our analysis revealed that seals use sea ice progressively more throughout the spring, and
-with peak use near mid-May to early June before declining again. We also found that seals
-preferentially haul out on ice shortly after solar noon, which may be associated with
-fewer prey in the water column (??? Can someone else who knows more about their feeding
-ecology contribute here?). Interestingly, bearded seals appeared to have two peaks in
-haul-out activity throughout the day, one shortly after solar noon, and one centered near
-solar midnight (related to benthic feeding?). It appeared that adult spotted seals
-initiated haul-out behavior earlier in the year than ribbon seals, and maintained this
-behavior over a longer duration; however they exhibited stronger diel behavior than ribbon
-seals (Fig X). These patterns are similar to those obtained by Conn et al. (2014), who
-used a subset of the data we analyzed here.
+In this paper, we modeled data from bio-logging sensors deployed on bearded, spotted, and
+ribbon seals to examine factors affecting haul-out behavior on sea ice in the Bering and
+Chukchi seas. Our analysis shows seals use sea ice progressively more throughout the
+spring, and with peak use near mid-May to early June before declining again. This pattern
+aligns well with what has been previously documented and confirms our bio-logging
+deployments are likely capturing population level behavioral patterns. Seals
+preferentially haul out on ice shortly after solar noon which coincides with the daily
+period of maximum solar radiation. The association of daily peak haul-out behavior with
+mid to later solar hours is commonly noted in pinnipeds and allows seals to maximize
+absorption of solar radiation. Interestingly, bearded seals appear to have two peaks in
+haul-out activity across the day, one shortly after solar noon, and one centered near
+solar midnight. A similar bi-modal pattern has been seen in ringed seals and suggests that
+bearded and ringed seals may be operating under different constraints than ribbon and
+spotted seals. Bearded and ringed seals are distributed across higher latitudes and the
+extended daylight hours may allow more flexibility in optimizing resting periods with
+foraging. Notably, bearded and ringed seals are subjected to much higher levels of
+predation from polar bears and human harvest.
+
+::: {custom-style="comment"}
+are there references for timing of polar bear activity/kills or hunter activity that might
+lend some additional support for predation as a possible explanation for the bi-modal
+haul-out pattern? Any data on timing of hunting activity by Alaska communities?
+:::
 
 Unlike previous analyses of seal haul-out data (e.g. Ver Hoef et al. 2009, Conn et al.
-2014), we also investigated the influence of sex-age class on haul-out probabilities.
-Although it was difficult to discern a pattern for bearded seals owing to low sample size,
-it appeared that adult male spotted and ribbon seals typically hauled out for longer
-durations, perhaps because they were less constrained by energetic demands associated with
-nursing and could sustain longer periods on ice to engage in molting and mating
-activities. Subadult ribbon and spotted seals exhibited different behavior when compared
-to adults. In particular, subadult ribbon seals exhibited similar haul-out behavior to
-adults, whereas subadult spotted seals appeared to haul-out less frequently and sustained
-a shorter duration of peak haul-out behavior than adults. The latter result is likely due
-to subadults only needing to haul out on ice to molt; pupping and mating are not yet in
-their behavioral repertoire.
+2014), we also investigated the influence of sex-age class on haul-out probabilities for
+all species except bearded seals because of low sample size. While both ribbon and spotted
+seals exhibited a unimodal diel haul-out pattern generally centered around local solar
+noon, there were key differences across species, age, and sex that match expectations
+given what we know of their ecological behavior. Spotted seals are known to form triads
+during the breeding season where a female and dependent pup are accompanied on the ice by
+a suitor male. The male is waiting for the female to wean the pup and enter estrus and
+fending off any other males that may be interested. Triad formation results in both males
+and females spending a large portion of the day hauled out on ice and a protracted spring
+haul-out season for both sexes. Since breeding occurs in the water, females may also be
+less inclined to venture out for foraging trips while still nursing the pup. We can see a
+clear reflection of this in the predicted haul-out behavior with both males and females
+exhibiting a broad distribution of time out of the water throughout the solar day and the
+season. Ribbon seals are not known to form triads and our model predicts a progression of
+increased haul-out behavior with females starting earlier in the season than males.
+Notably, female ribbon seals spend a large portion of the day in the water during the
+pupping period and aligns with the hypothesis that ribbon seal females continue foraging
+while nursing. In the case of both ribbon and spotted seals, subadults are the first to
+begin meaningful haul-out behavior and reflects a typical phocid pattern where subadults
+molt first as they do not have any reproductive constraints.
+
+::: {custom-style="comment"}
+this previous paragraph might benefit from some references e.g. triads in spotted seals,
+ribbon seal females foraging while nursing, etc paper on dive/foraging behavior of
+lactating harp seals: <https://doi.org/10.1006/anbe.1993.1312>
+:::
 
@@
 including wind speed, temperature, barometric pressure, precipitation, and wind chill.
-These have been investigated for walrus (e.g. Udevitz et al. 2009), but to our knowledge
-have never been investigated for ice-associated seals. In particular, we found clear
-evidence that seal haul-out probabilities decrease with wind speed and increase with
-temperature (Figs X-Y). In addition, we found evidence that ribbon seal haul-out
-probabilities are positively related to barometric pressure, and that spotted seal
-haul-out probabilities are negatively related to precipitation (Appendix C). The tendency
-of seals to haul out in "good" weather may be related to energetic considerations, better
-predator detection, or both. Bearded seal haul-out behavior was predicted to change
-slightly based on how far north tagged seals were; seals farther north tended to initiate
-haul-out behavior slightly later, and for a longer period of time (Fig. X). This response
-may be related to different temperature regimes, or to timing of ice melt, which occurs
-later in locations farther north.
-
-There appeared to be rather substantial annual deviations in the timing and magnitude of
-haul-out peaks (Figs X-Y). For example, peak ribbon seal haul-out behavior appeared to
-vary by as much as three weeks between seasons. There thus seems to be a reasonable amount
-of plasticity in the timing of haul-out, suggesting some plasticity in life history
-functions such as pupping, molting and breeding. However, we were unable to relate the
-timing of haul-out conditions to underlying ice conditions, which makes predicting climate
-change consequences on life history events difficult. Future investigation into
-environmental factors contributing to the timing of haul-out peaks seems warranted. For
-instance, Ray et al. (2010) suggested that ribbon and spotted seals were most dependent on
-loose pack ice at the sourthern sea ice edge; thus, the relative suitability for haul-out
-behavior may have more to do with the characterization, position, and quantity of loose
-pack ice rather than the total amount of ice in the Bering Sea. Further work to
-characterize remotely sensed sea ice data into relevant ice types (e.g. loose pack,
-continuous pack, rounded pack) would be useful.
+These have been investigated for walrus (e.g. Udevitz et al. [@udevitz2009cjz]), but to
+our knowledge have never been investigated for ice-associated seals. Ribbon seals seem to
+be the most influenced by weather with wind, temperature, barometric pressure, and
+precipitation all being important components of the model. Spotted seals were most
+affected by wind and barometric pressures. And, for bearded seals, the model indicated
+wind and temperature had the greatest impact. In general, and as might be expected, seals
+were more likely to haul-out when daily temperatures were warmer, winds speeds were less,
+barometric pressure was high, and lower rates of precipitation. These weather conditions
+are likely most beneficial for energetic considerations and may also provide the best
+situation for predator detection. These results highlight the importance for haul-out
+behavior models (and subsequent calculations of availability during surveys) to
+incorporate weather covariates.
+
+Our model uncovered annual deviations in the timing and magnitude of
+haul-out peaks for ribbon and spotted seals. The timing of peak haul-out activity appears
+to fall within a relatively narrow time window of 3-4 weeks in late May and early June. 
+This remarkable level of consistency across 15 years is indicative of the 
+adaptations these species have evolved to synchronize their key life history stages
+with favorable environmental conditions (e.g. availability of sea ice, increased
+oceanographic productivity, favorable weather conditions). Unfortunately, this may also
+be indicative of a limited range of plasticity as climatological conditions change.
+Additionally, while from an ecological perspective the haul-out behavior appears
+consistent, the interannual differences in timing and magnitude are large enough to have
+important ramifications on calculations of abundance and trend.
 
@@
 frequency of counts observed over time. Regardless, researchers should anticipate there
-being some unmodeled heterogeneity in availability probability present in abundance
+being some un-modeled heterogeneity in availability probability present in abundance
 estimates obtained from aerial surveys. This may make trend detection difficult, as one
@@
 increased sample sizes in terms of number of animals, but also improvements to the way
-data are prepared prior to analysis. For instance, for bearded seals there were often long
-stretches at the end of the tag history where it appeared animals were hauled out 100% of
-the time. In reality, these likely represented tags shed during molt on ice floes or
-washed up on beaches. In the present study, experts used multiple data streams to
-determine effective start and end dates for each deployment, and these long stretches were
-often removed. In previous studies, such periods were often retained and likely led to
-positively biased estimates of haul-out probability. Given the potential ramifications for
-abundance estimation, we strongly suggest conducting such data screening exercises prior
-to analyzing haul-out data. Future research should be devoted to automating this
-procedure, perhaps using artificial intelligence algorithms to classify tag behavior based
-on multiple data streams. Such a framework would require adequate training data, e.g.
-using tags purposefully placed on ice floes, beaches, or subject to tag fouling so that
-automatic classifiers could appropriately discriminate properly functioning tags from
-anomalous behavior. Another approach would be to integrate additional information on pulse
-or respiration, as can now be done with life history tags (e.g., Horning and Hill 2005).
-At minimum, we suggest reporting haul-out correction factors used when conducting
-abundance analyses so that attendant estimates can be revisited as our state of knowledge
-about tag function and haul-out behavior evolves.
+data are prepared prior to analysis.
 
@@
 
+::: {custom-style="comment"}
+this could use a better ending with more punch and interest
+:::
+
 # Author Contributions
 
+::: {custom-style="comment"}
+insert broad statement and summary regarding data contribution from AFSC, ADFG, NSB (I 
+still need to lookup details of which org contributed which data). Authors, please
+add your contribution statement below
+:::
+
+*Josh M. London:*
+
+*Paul B. Conn:*
+
+*Stacie K. Hardy:*
+
+*Erin L. Richmond:*
+
+*Jay M. Ver Hoef:*
+
+*Irina S. Trukhanova:*
+
+*Michael F. Cameron:*
+
+*Peter L. Boveng:*
+
 # Data Availability
+
+Data and code are available as a complete research compendium on GitHub and major
+versions are archived at Zenodo. Original data sources for telemetry are archived at
+the United States Animal Telemetry Network.
 
@@
 Fitzgerald, Tracey Goldstein, Jeff Harris, Markus Horning, John Goodwin, John Jansen,
-Charles Littnan, Erin Moreland, Lorrie Rea, Dan Savetilik, and Dave Withrow. We also
-appreciate the commitment to science and safety by all officers and crew of the NOAA Ship
-Oscar Dyson, the NOAA Ship MacArthur II, and the RV Thomas G. Thompson.
+Shawn Johnson, Charles Littnan, Erin Moreland, Lorrie Rea, Dan Savetilik, and Dave
+Withrow. We also appreciate the commitment to science and safety by all officers and crew
+of the NOAA Ship Oscar Dyson, the NOAA Ship MacArthur II, and the RV Thomas G. Thompson.
 

### 2021-03-25 analysis/scripts/create_avg_wx.R

@@ -30,10 +30,10 @@ cov_means <- tbl(con, in_schema("base","tbl_analysis_grid_cov_wx")) %>%
             wind = sqrt(mean(rast_uwnd, na.rm=TRUE)^2 + mean(rast_vwnd, na.rm=TRUE)^2)
             )
 
-cov_means_daily <- tbl(con, in_schema("base","tbl_analysis_grid_cov_wx")) %>%
+cov_means_daily_2012 <- tbl(con, in_schema("base","tbl_analysis_grid_cov_wx")) %>%
   filter(cell %in% !!grid.sf$cell) %>%
   filter(month(fdatetime_range_start) %in% c(3,4,5,6,7),
-         year(fdatetime_range_start) != '2004') %>%
+         year(fdatetime_range_start) == '2012') %>%
   group_by(year = year(fdatetime_range_start),
            yday = yday(fdatetime_range_start),
            cell) %>%
@@ -46,8 +46,8 @@ cov_means_daily <- tbl(con, in_schema("base","tbl_analysis_grid_cov_wx")) %>%
 
 cov_means <- cov_means %>% collect()
 save(cov_means, file = here::here('data/cov_means.Rdata'))
-cov_means_daily <- cov_means_daily %>% collect()
-save(cov_means_daily, file = here::here('data/cov_means_daily.Rdata'))
+cov_means_daily <- cov_means_daily_2012 %>% collect()
+save(cov_means_daily_2012, file = here::here('data/cov_means_daily_2012.Rdata'))
 
 
 

### 2021-03-25 analysis/scripts/estimate.R

@@ -1,93 +1,0 @@
-library(glmmLDTS)
-
-fit_ribbon <- readRDS(here::here("data/fit_ribbon.rds"))
-#available here: https://drive.google.com/file/d/1rsC7PM11QcSFsxe-XKFiNtJq1ZkeG1QL/view?usp=sharing
-
-# example from Jay's email
-# for(i in 1:ndates) {
-#   results[[i]] <- estimate.glmp.ts(glmp.ts.object = all.fit.1,
-#                                    fixed.effect.list = list(
-#                                      eff1 = list(effect = "MinFHi.Cat",
-#                                                  level = as.character(survcov[i,'MinFHi.Cat']),
-#                                                  weight = 1),
-#                                      eff2 = list(effect = "MFAug15",
-#                                                  weight = survcov[i,'MFAug15']),
-#                                      eff3 = list(effect = "hour",
-#                                                  level = as.character(survcov[i,'hour']),
-#                                                  weight = 1),
-#                                      eff4 = list(effect = "hour:MFAug15",
-#                                                  level = paste(as.character(survcov[i,'hour']),",", sep = ""),
-#                                                  weight = survcov[i,'MFAug15']),
-#                                      eff5 = list(effect = "Year",
-#                                                  level = c("102", "105", "106"),
-#                                                  weight = c(0.333333, 0.333333, 0.333334))
-#                                    )
-#   )
-# }
-
-# let's remind ourselves of the model formula
-fit_ribbon$fixed.formula
-
-# estimate for solar hour 12 on day 106
-#
-# eventually, we want to estimate across all hours for a series of days
-# to create our plots. So, this would become a nested for-loop
-
-fix_solar_hour = 12
-fix_sin1 <- sin(pi*fix_solar_hour/12)
-fix_cos1 <- cos(pi*fix_solar_hour/12)
-fix_sin2 <- sin(pi*fix_solar_hour/6)
-fix_cos2 <- cos(pi*fix_solar_hour/6)
-fix_sin3 <- sin(pi*fix_solar_hour/4)
-fix_cos3 <- sin(pi*fix_solar_hour/4)
-
-# yday 106 is April 15
-fix_day <- (106-120)/10
-fix_day2 <- fix_day^2
-fix_day3 <- fix_day^3
-
-# for now, we'll just use mean wx covariates
[…]

### 2021-03-25 README.Rmd
ADD BINDER BADGES TO README

@@
 ```
+
+<!-- badges: start -->
+[![Launch Rstudio Binder](http://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/jmlondon/berchukHaulout/master?urlpath=rstudio)
+<!-- badges: end -->
 
@@
 
-The simplest way (*not yet enabled*) to explore the text, code and data is to click on
-[binder](https://mybinder.org/v2/gh///master?urlpath=rstudio)
+The simplest way to explore the text, code and data is to click on
+[binder](https://mybinder.org/v2/gh/jmlondon/berchukHaulout/master?urlpath=rstudio)
 to open an instance of RStudio in your browser, which will have the

### 2021-03-25 README.Rmd
FIX BRANCH TO 'main' IN BINDER LINK

@@
 <!-- badges: start -->
-[![Launch Rstudio Binder](http://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/jmlondon/berchukHaulout/master?urlpath=rstudio)
+[![Launch Rstudio Binder](http://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/jmlondon/berchukHaulout/main?urlpath=rstudio)
 <!-- badges: end -->

### 2021-03-25 README.Rmd
ADD NOAA DISCLAIMERS

@@
 environment. These Docker images can also be used locally.
+
+### NOAA README
+
+This repository is a scientific product and is not official communication of the National Oceanic and Atmospheric Administration, or the United States Department of Commerce.
+All NOAA GitHub project code is provided on an ‘as is’ basis and the user assumes responsibility for its use.
+Any claims against the Department of Commerce or Department of Commerce bureaus stemming from the use of this GitHub project will be governed by all applicable Federal law.
+Any reference to specific commercial products, processes, or services by service mark, trademark, manufacturer, or otherwise, does not constitute or imply their endorsement, recommendation or favoring by the Department of Commerce.
+The Department of Commerce seal and logo, or the seal and logo of a DOC bureau, shall not be used in any manner to imply endorsement of any commercial product or activity by DOC or the United States Government.
+
+### NOAA License
+
+Software code created by U.S.
+Government employees is not subject to copyright in the United States (17 U.S.C.
+§105).
+The United States/Department of Commerce reserve all rights to seek and obtain copyright protection in countries other than the United States for Software authored in its entirety by the Department of Commerce.
+To this end, the Department of Commerce hereby grants to Recipient a royalty-free, nonexclusive license to use, copy, and create derivative works of the Software outside of the United States.

### 2021-03-25 README.Rmd
ADD ZENODO DOI BADGE

@@
 [![Launch Rstudio Binder](http://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/jmlondon/berchukHaulout/main?urlpath=rstudio)
+[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.4638221.svg)](https://doi.org/10.5281/zenodo.4638221)
 <!-- badges: end -->

### 2021-03-25 README.Rmd
IMPROVING BINDER SUPPORT

@@
 The simplest way to explore the text, code and data is to click on
-[binder](https://mybinder.org/v2/gh/jmlondon/berchukHaulout/master?urlpath=rstudio)
+[binder](https://mybinder.org/v2/gh/jmlondon/berchukHaulout/main?urlpath=rstudio)
 to open an instance of RStudio in your browser, which will have the

### 2021-03-29 README.Rmd
FURTHER UPDATES TO README

@@
 
-# berchukHaulout
+## Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas
+Josh M. London^1^,✉, Paul B. Conn^1^, Stacie K. Hardy^1^, Erin L. Richmond^1^, Jay M. Ver Hoef^1^, Irina S. Trukhanova^1^, Michael F. Cameron^1^, and Peter L. Boveng^1^
+1 Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
+✉ Correspondence: Josh M. London <josh.london@noaa.gov>
 
-This repository contains the data and code for our paper:
-*Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas*
+This repository contains the research compendium in support of the above titled paper.
+The manuscript is currently in preparation and the text, code, analysis, and results
+are under active development.
 
@@
 The Department of Commerce seal and logo, or the seal and logo of a DOC bureau, shall not be used in any manner to imply endorsement of any commercial product or activity by DOC or the United States Government.
-
-### NOAA License
-
-Software code created by U.S.
-Government employees is not subject to copyright in the United States (17 U.S.C.
-§105).
-The United States/Department of Commerce reserve all rights to seek and obtain copyright protection in countries other than the United States for Software authored in its entirety by the Department of Commerce.
-To this end, the Department of Commerce hereby grants to Recipient a royalty-free, nonexclusive license to use, copy, and create derivative works of the Software outside of the United States.

### 2021-03-29 analysis/paper/paper.Rmd
INITIAL EDITS PER P CONN SUGGESTS

@@
 year effects were highly supported, indicating that the timing and magnitude of haul-out
-behavior varied among years. There was no support linking the variability in haul-out
-timing to annual sea ice extent. Our analysis also emphasizes the importance of accounting
-for weather factors affecting haul-out behavior when interpreting the number of seals
+behavior varied among years. However, we did not find evidence that haul-out
+timing was linked annual sea ice extent. Our analysis emphasizes the importance of accounting
+for variation in haul-out behavior when interpreting the number of seals
 counted in aerial surveys.
@@
 Seas. We focused our effort on bearded, spotted, and ribbon seals and do not include
-ringed seals the analysis. Ringed seals are unique in their use of snow lairs in the
+ringed seals in the analysis. Ringed seals are unique in their use of snow lairs in the
 spring months and they warrant a separate study that includes additional environmental
@@
 this study, in cases where both tag types were deployed, percent dry observations from the
-flipper tag were usually preferred. Sex and age class were determined at the time of
+flipper tag were preferred. Sex and age class were determined at the time of
 deployment by experts in the field. Seals determined to be less than one year were
@@
 
-```{r seal-hours-calendar, fig.height = 4.75, include=TRUE, fig.id = "dataCal", fig.cap="Distribution of of haul-out behavior data from 1 March to 15 July for each species.
+```{r seal-hours-calendar, fig.height = 4.75, include=TRUE, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species.
 The higher density of data from May and June coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out.
@@
 Of key interest in this study is the relationship between haul-out behavior and a variety
-of environmental covariates that vary with space and time. We explored use of a movement
+of environmental covariates that vary with space and time. We explored the use of a movement
 model to predict locations at specific times. However, the sparse nature of data from some
@@
 Notably missing from this list of explanatory variables is any spatial-temporal
-representation of sea ice concentration, area, or extent. This may seem counter-intuitive
+representation of sea ice concentration, area, or extent. This may seem counterintuitive
 when modeling the haul-out behavior of seal species with such a close association to sea
@@
 
-We did explore whether the annual variation in spring sea ice extent in the Bering Sea
-influences the peak timing of seal haul-out behavior. For this, all sea ice concentration
+We did explore whether the annual variation in maximum spring sea ice extent in the Bering
+Sea influences the seasonal peak of seal haul-out behavior. In particular, we used sea ice
+concentration data from the Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave Dataset,
+Version 1 [@cavalieri1996] to calculate maximum sea ice extent. All sea ice concentration
 grid cells (25 km^2^) in the study area with greater than 15% concentration were summed to
-get total sea ice extent for each day between 15 February and 15 July across all years
-(`r min(lubridate::year(dat.sf$haulout_dt))`-`r max(lubridate::year(dat.sf$haulout_dt))`).
-The maximum sea ice extent for each year was determined from this calculation. We obtained
-sea ice concentration data from the Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave
-Dataset, Version 1 [@cavalieri1996].
+get total sea ice extent for each day between 15 February and 15 July across all years.
+Maximum spring sea ice extent was simply the largest daily value for each year.
 
@@
   "Convective precipitation (kg/m2)",
-  "Pressure", "Continuos", "NARR", 
+  "Pressure", "Continuous", "NARR", 
   "Atmospheric pressure at sea level (kPa)",
@@
 adopted a continuous formulation based on Fourier series that provides a flexible model
-while preserving the inherit circularity needed for time-of-day effects (i.e., hour 0
+while preserving the inherent circularity needed for time-of-day effects (i.e., hour 0
 should be equal to hour 24). It also represents hour-of-day with 6 variables, which is a
@@
 year$\times$day and year$\times$day2, omitting the main effects of year as well as
-year$\times$day3 interactions as models with the latter effects were numerically unstable.
+year$\times$day3 interactions because models with the latter effects were numerically unstable.
 However, the modeled interactions were sufficient to allow shifts in haul-out
@@
 AIC (and similar criteria) is not suitable when using pseudo-likelihoods. After fitting
-GLMPM models, we used "type III" F-tests to calculate p-values [@verhoef2010cs] and rely
-on those insights to evaluate model performance and important terms. We also produced
-predictions of haul-out behavior as a function of influential predictors (e.g. solar hour,
-day-of-year, age-sex). Weather covariates for these predictions were based on daily or
-hourly (GAM) smoothed weather covariate values across the study region. Such predictions
-were then used to develop haul-out probability surfaces, explore marginal effects of
-weather covariates, and determine annual peaks in haul-out activity. The timing of peak
-haul-out behavior was further used to regress against the annual maximum sea ice extent in
-the study region.
+GLMPM models, we used "type III" F-tests to calculate p-values [@verhoef2010cs] to
+evaluate model performance and important terms. We also produced predictions of haul-out
+behavior as a function of influential predictors (e.g. solar hour, day-of-year, age-sex).
+Weather covariates for these predictions were based on daily or hourly smoothed weather
+covariate values across the study region. Such predictions were then used to develop
+haul-out probability surfaces, explore marginal effects of weather covariates, and
+determine annual peaks in haul-out activity. The timing of peak haul-out behavior was
+further used to regress against the annual maximum sea ice extent in the study region.
 

### 2021-04-07 analysis/paper/paper.Rmd
INITIAL COMMENTS FROM MFC

@@
 early summer (April-June) as sea ice begins to melt and retreat northward. Rapid
-acceleration of climate change in Arctic ecosystems is thus of concern as the quantity and
-quality of suitable habitat is forecast to decrease. In this paper, we use data collected
-from satellite-linked telemetry tags deployed between 2005 and 2020 to investigate the
-seasonal timing and environmental factors affecting sea ice use by seals (specifically,
-bearded, ribbon, and spotted seals) in the Bering, and Chukchi seas. In addition to
-providing baseline data on phenology, these data also allow us to refine availability
-estimates needed to accurately estimate abundance from aerial survey counts of seals
-basking on ice (i.e., to correct for the proportion of animals that are in the water while
-surveys are conducted). Using generalized linear mixed pseudo-models to properly account
-for temporal autocorrelation, we fit models with a variety of covariates (e.g.,
-day-of-year, solar hour, age-sex class, wind speed, barometric pressure, temperature,
-precipitation) to examine their ability to explain variation in hourly haul-out records.
-We found evidence for strong diurnal and within season patterns in haul-out behavior, as
-well as strong weather effects (particularly wind and temperature). In general, seals were
-more likely to rest on ice in the middle of the day and when wind speed was low and
-temperatures were higher. Haul-out propensity increased through March and April, peaking
-in May and early June before declining again. The timing and frequency of haul-out events
-also varied based on species and age-sex class. For ribbon and spotted seals, models with
-year effects were highly supported, indicating that the timing and magnitude of haul-out
-behavior varied among years. However, we did not find evidence that haul-out
-timing was linked annual sea ice extent. Our analysis emphasizes the importance of accounting
-for variation in haul-out behavior when interpreting the number of seals
+acceleration of climate change in Arctic ecosystems is therefore of concern as the
+quantity and quality of suitable habitat is forecast to decrease. In this paper, we use
+data collected from satellite-linked bio-loggers deployed between 2005 and 2020 to
+investigate the seasonal timing and environmental factors affecting sea ice use by seals
+(specifically, bearded, ribbon, and spotted seals) in the Bering and Chukchi seas. In
+addition to providing baseline data on phenology, these data also allow us to refine
+availability estimates needed to accurately estimate abundance from aerial survey counts
+of seals basking on ice (i.e., to correct for the proportion of animals that are in the
+water while surveys are conducted). Using generalized linear mixed pseudo-models to
+properly account for temporal autocorrelation, we fit models with a variety of covariates
+(e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure,
+temperature, precipitation) to examine their ability to explain variation in hourly
+haul-out records. We found evidence for strong diurnal and within season patterns in
+haul-out behavior, as well as strong weather effects (particularly wind and temperature).
+In general, seals were more likely to rest on ice in the middle of the day and when wind
+speed was low and temperatures were higher. Haul-out propensity increased through March
+and April, peaking in May and early June before declining again. The timing and frequency
+of haul-out events also varied based on species and age-sex class. For ribbon and spotted
+seals, models with year effects were highly supported, indicating that the timing and
+magnitude of haul-out behavior varied among years. However, we did not find evidence that
+haul-out timing was linked annual sea ice extent. Our analysis emphasizes the importance
+of accounting for variation in haul-out behavior when interpreting the number of seals
 counted in aerial surveys.
@@
 March and peaking in May and June for bearded, ribbon, and spotted seals. However, their
-primary focus was estimating seal abundance, and like a subsequent analysis by Conn et al.
-[-@conn2014mee], they did not attempt to separately analyze haul-out schedules for
-different age and sex classes (although initial models with sex and age as fixed effects
-were insignificant at p = 0.05; [@verhoef2014sm]).
+primary focus was to develop a correction factor for estimating seal abundance, and like a
+subsequent analysis by Conn et al. [-@conn2014mee], they did not attempt to separately
+analyze haul-out schedules for different age and sex classes (although initial models with
+sex and age as fixed effects were insignificant at p = 0.05; [@verhoef2014sm]).
 
@@
 predictions about the effects of the climate crisis on the abundance and distribution of
-Arctic seal populations. Before we can construct a trend, however, we first need credible
-data points. Several studies have contributed estimates of the distribution and abundance
-of ice-associated seal species in the Arctic using aerial surveys (e.g.,
-[@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]). Such studies are conducted over
-very large areas and estimation of absolute abundance requires making inference about
-numerous nuisance processes affecting the observation of seals on ice. This includes
-availability (only seals basking on ice are available to be counted), detection
-probability (observers or automated detection systems may miss some seals on ice), species
-misclassification, and possible disturbance of seals by aircraft [@conn2014mee;
+Arctic seal populations. Before we can construct a trend, however, we first require a
+baseline. Several studies have contributed estimates of the distribution and abundance of
+ice-associated seal species in the Arctic using aerial surveys (e.g., [@bengtson2005pb],
+[@conn2014mee], and [@verhoef2014sm]). Such studies are conducted over very large areas
+and estimation of absolute abundance requires making inference about numerous nuisance
+processes affecting the observation of seals on ice. This includes availability (only
+seals basking on ice are available to be counted), detection probability (observers or
+automated detection systems may miss some seals on ice), species misclassification, and
+possible disturbance of seals by aircraft [@conn2014mee;
 @verhoef2014sm].
@@
 deployment by experts in the field. Seals determined to be less than one year were
-classified as young of the year. subadults are those seals likely greater than one year of
+classified as young-of-the-year. subadults are those seals likely greater than one year of
 age but not yet sexually mature. Adults are likely sexually mature and older than
-approximately four years. For those tags deployed on young of the year and transmitting
+approximately four years. For those tags deployed on young-of-the-year and transmitting
 into the next year, the age class was advanced to subadult on March 1 of the following
@@
 to satellites and we did not want to include data from these tags in our analyses. As
-such, start and end times of each deployment were made by experts who examined several
+such, start and end times of each deployment were identified by experts who examined several
 simultaneous data streams (e.g., tag locations and dive behavior) to determine when tags
@@
 tag firmware every few seconds and a percent of the hour in a dry state was calculated
-(\@ref(fig:examplePlot)). Each data transmission message consisted of a complete 24-hour
+(Figure \@ref(fig:examplePlot)). Each data transmission message consisted of a complete 24-hour
 record. Tags were deployed and programmed in a manner to maximize data transmission during
@@
 
-In addition to sex and age class (non-dependent *young of the year*, sexually immature
+In addition to sex and age class (non-dependent *young-of-the-year*, sexually immature
 *subadults*, and mature *adults*), we gathered additional variables that might help
@@
 values were subset to the extent of our study area over the Bering and Chukchi Seas at
-3-hr intervals based on a grid resolution of 32 km (1024 sq. km) The following
+3-hr intervals based on the native grid resolution of 32 km (1024 sq. km). The following
 meteorological variables were interpolated and assigned to seal locations using a bilinear
@@
 grid cells (25 km^2^) in the study area with greater than 15% concentration were summed to
-get total sea ice extent for each day between 15 February and 15 July across all years.
+get the total sea ice extent for each day between 15 February and 15 July across all years.
 Maximum spring sea ice extent was simply the largest daily value for each year.
@@
 of $A_i$ so that they summed to one . Given a stable age distribution, we calculated the
-proportion of the population by sex and stage (young of the year: 0-1 year; subadult: at
+proportion of the population by sex and stage (young-of-the-year: 0-1 year; subadult: at
 least one year but sexually immature; adult: sexually mature) using data on sexual
@@
 molting period. Subadults showed an earlier start and more intense haul-out activity in
-April and May. The young of the year records begin after weaning and the model predictions
+April and May. The young-of-the-year records begin after weaning and the model predictions
 demonstrate the ontogeny of in-water activities (e.g. diving, foraging) in May. Adult
@@
 less intensely centered on solar noon. Adults of both sexes spend considerable time in
-April and May resting on the ice. As with ribbon seals, the young of the year records
+April and May resting on the ice. As with ribbon seals, the young-of-the-year records
 begin after weaning and the model predictions demonstrate the ontogeny of in-water
@@
 
-*Josh M. London:*
-
-*Paul B. Conn:*
-
-*Stacie K. Hardy:*
-
-*Erin L. Richmond:*
-
-*Jay M. Ver Hoef:*
-
-*Irina S. Trukhanova:*
-
-*Michael F. Cameron:*
-
-*Peter L. Boveng:*
+*Josh M. London:* led field work, conceptualization, methodology, formal analysis, code development, writing of original draft, visualization, review, revisions, and corresponding author
+
+*Paul B. Conn:* conceptualization, methodology, formal analysis, code development, writing of original draft, review, and revisions
+
+*Stacie K. Hardy:* field work, methodology, data curation, validation, review, and revisions
+
+*Erin L. Richmond:* data curation, conceptualization, methodology, review, and revisions
+
+*Jay M. Ver Hoef:* field work, conceptualization, methodology, code development, review, and revisions
+
+*Irina S. Trukhanova:* conceptualization, formal analysis, code development, review, and revisions
+
+*Michael F. Cameron:* led field work, project administration, methodology, review, and revisions
+
+*Peter L. Boveng:* led field work, supervision, project administration, conceptualization, methodology, review, and revisions
 

### 2021-04-09 analysis/paper/paper.Rmd
ACKNOWLEDGEMENTS UPDATE PER MFC

@@
 frequency of counts observed over time. Regardless, researchers should anticipate there
-being some un-modeled heterogeneity in availability probability present in abundance
+being some unmodeled heterogeneity in availability probability present in abundance
 estimates obtained from aerial surveys. This may make trend detection difficult, as one
@@
 
-*Josh M. London:* led field work, conceptualization, methodology, formal analysis, code development, writing of original draft, visualization, review, revisions, and corresponding author
-
-*Paul B. Conn:* conceptualization, methodology, formal analysis, code development, writing of original draft, review, and revisions
-
-*Stacie K. Hardy:* field work, methodology, data curation, validation, review, and revisions
-
-*Erin L. Richmond:* data curation, conceptualization, methodology, review, and revisions
-
-*Jay M. Ver Hoef:* field work, conceptualization, methodology, code development, review, and revisions
-
-*Irina S. Trukhanova:* conceptualization, formal analysis, code development, review, and revisions
-
-*Michael F. Cameron:* led field work, project administration, methodology, review, and revisions
-
-*Peter L. Boveng:* led field work, supervision, project administration, conceptualization, methodology, review, and revisions
+**Josh M. London:** led field work, conceptualization, methodology, formal analysis, code development, writing of original draft, visualization, review, revisions, and corresponding author
+
+**Paul B. Conn:** conceptualization, methodology, formal analysis, code development, writing of original draft, review, and revisions
+
+**Stacie K. Hardy:** field work, methodology, data curation, validation, review, and revisions
+
+**Erin L. Richmond:** data curation, conceptualization, methodology, review, and revisions
+
+**Jay M. Ver Hoef:** field work, conceptualization, methodology, code development, review, and revisions
+
+**Irina S. Trukhanova:** conceptualization, formal analysis, code development, review, and revisions
+
+**Michael F. Cameron:** led field work, project administration, methodology, review, and revisions
+
+**Peter L. Boveng:** led field work, supervision, project administration, conceptualization, methodology, review, and revisions
 
@@
 communities. We would like to especially acknowledge the communities of Kotzebue,
-Utqiagvik, and Ulguniq.
-
-We are grateful for the assistance in catching and sampling seals by Michelle Barbieri,
-Gavin Brady, Cynthia Christman, Shawn Dahle, Stacy DiRocco, Deb Fauquier, Shannon
-Fitzgerald, Tracey Goldstein, Jeff Harris, Markus Horning, John Goodwin, John Jansen,
-Shawn Johnson, Charles Littnan, Erin Moreland, Lorrie Rea, Dan Savetilik, and Dave
+Utqiagvik, and Ulguniq and the following individuals: James Adams, Jeff Barger, David
+Barr, Wendell Booth, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin, John
+Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr., Virgil
+Naylor Sr., Dan Savetilik, Allen Stone, and Randy Toshavik from Kotzebue, Alaska; Billy
+Adams, Mary-Ellen Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard
+Kittick, Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik, Shawn Oktollik,
+Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross Schaeffer, Bob Shears, and Joe Skin
+from the North Slope Borough, Alaska;
+
+We are grateful for the assistance in catching and sampling seals by James Bailey,
+Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov, Cynthia Christman, Shawn
+Dahle, Rob Delong, Stacy DiRocco, Deb Fauquier, Shannon Fitzgerald, Kathy Frost, Scott
+Gende, Tracey Goldstein, Jeff Harris, Markus Horning, John Jansen, Shawn Johnson, Charles
+Littnan, Lloyd Lowry, Erin Moreland, Lorrie Rea, Gay Sheffield, Brent Stewart, and Dave
 Withrow. We also appreciate the commitment to science and safety by all officers and crew

### 2021-04-10 analysis/paper/paper.Rmd
EDITS PER SKH COMMENTS

@@
 in the region [@huntington2020ncc]. Such disruptions are a particular cause of concern for
-ice-associated seals (ringed, bearded, spotted, and ribbon seals) which depend on spring
+ice-associated seals (ringed, bearded, spotted, and ribbon seals) that depend on spring
 and early summer sea ice (March-June) in the Bering and Chukchi seas as a platform for
-important functions such as pupping, nursing, breeding, and molting [@boveng2009a;
+important functions, such as pupping, nursing, breeding, and molting [@boveng2009a;
 @boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge gaps prevent our
@@
 of the water almost exclusively on sea ice and are mostly pelagic outside the spring
-pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals do
+pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals
 rest on beaches during ice-free months, they strongly prefer sea ice as a resting platform
-during the spring and early summer [@burns2009]. The very remote nature of the Bering and
+during the spring and early summer [@burns2009]. The remote nature of the Bering and
 Chukchi seas mean direct observation of seal behavior is limited. Bio-logging devices are,
@@
 
-In this study, we use data from fifteen years of bio-logger records in a comprehensive
-investigation of haul-out behavior by three ice-associated seals in the Bering and Chukchi
+In this study, we use data from 15 years of bio-logger records in a comprehensive
+investigation of haul-out behavior by 3 ice-associated seals in the Bering and Chukchi
 Seas. We focused our effort on bearded, spotted, and ribbon seals and do not include
 ringed seals in the analysis. Ringed seals are unique in their use of snow lairs in the
-spring months and they warrant a separate study that includes additional environmental
-covariates such as the onset of snow melt and snow depth. Our goals are threefold. First,
+spring months, and they warrant a separate study that includes additional environmental
+covariates, such as the onset of snow melt and snow depth. Our goals are threefold. First,
 we wish to establish baseline estimates for the chronology of haul-out behavior in the
@@
 devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
-during the annual molt and limits the effective length of haul-out timelines.
+during the annual molt, which limits the effective length of haul-out timelines.
 Additionally, sensors attached to the head or dorsal region are often dry while the seal
@@
 deployment by experts in the field. Seals determined to be less than one year were
-classified as young-of-the-year. subadults are those seals likely greater than one year of
+classified as young-of-the-year. Subadults are those seals likely greater than one year of
 age but not yet sexually mature. Adults are likely sexually mature and older than
 approximately four years. For those tags deployed on young-of-the-year and transmitting
-into the next year, the age class was advanced to subadult on March 1 of the following
+into the next year, the age class was advanced to subadult on 1 March of the following
 year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
@@
   dplyr::rename(Species = species, Sex = sex, `Young of Year` = `YOUNG OF YEAR`,
-                `subadult` = SUBADULT, Adult = ADULT) %>% 
+                `Subadult` = SUBADULT, Adult = ADULT) %>% 
   flextable() %>% 
@@
 Tags that fall off due to molt or mortality and remain on ice or land can still send data
-to satellites and we did not want to include data from these tags in our analyses. As
-such, start and end times of each deployment were identified by experts who examined several
-simultaneous data streams (e.g., tag locations and dive behavior) to determine when tags
-ceased providing data consistent with seal behavior. Sensor readings outside of these
-start and end times were discarded prior to analysis. After approximately 9 months, a few
-(7) devices deployed on the rear flipper of bearded seals reported implausible sensor
-data. All data after the first instance of unrealistic values were censored from this
-analysis. Figure \@ref(fig:dataCal) shows a calendar distribution of these data across the
-study season for each species. Observations are concentrated in the months of May and June
-due to the timing of deployment (April and May) and the timing of molt (May and June).
-During molt, seals (and their attached bio-loggers) spend more time out of the water and
-more data is transmitted. Molt timing also impacts when many deployments end as any tags
+to satellites, and we did not want to include data from these tags in our analyses. As
+such, start and end times of each deployment were identified by experts who examined
+several simultaneous data streams (e.g., tag locations and dive behavior) to determine
+when tags ceased providing data consistent with seal behavior. Sensor readings outside of
+these start and end times were discarded prior to analysis. After approximately 9 months,
+7 devices deployed on the rear flipper of bearded seals reported implausible sensor data.
+All data after the first instance of unrealistic values were censored from this analysis.
+Figure \@ref(fig:dataCal) shows a calendar distribution of these data across the study
+season for each species. Observations are concentrated in the months of May and June due
+to the timing of deployment (April and May) and the timing of molt (May and June). During
+molt, seals (and their attached bio-loggers) spend more time out of the water and more
+data is transmitted. Molt timing also impacts when many deployments end as any tags
 adhered to the hair will fall off.
@@
 record. Tags were deployed and programmed in a manner to maximize data transmission during
-the spring pupping and molting period. However, not all days during the deployment were
+the spring pupping and molting period, however, not all days during the deployment were
 successfully transmitted. This is due to a variety of factors including satellite
@@
 where haul-out observations were present without any location data were filled from the
-last calculated daily location and any days where the location intersected with land were
+last calculated daily location, and any days where the location intersected with land were
 removed from the dataset. Figure \@ref(fig:dataMap) shows the spatial distribution of
@@
 Models for all species include the following effects: day-of-year, solar hour,
-temperature, wind speed, barometric pressure, precipitation, and wind chill (represented
-by a wind $\times$ temperature interaction; [@udevitz2009cjz]), together with day-of-year
-and time of day interactions to permit diurnal patterns to change throughout the year).
+temperature, wind speed, barometric pressure, precipitation, wind chill (represented
+by a wind $\times$ temperature interaction [@udevitz2009cjz]), and day-of-year
+and time of day interactions to permit diurnal patterns to change throughout the year.
 Spotted and ribbon seal models included age-sex class and interactions between day-of-year
 and age-sex class, but we omitted these from bearded seal models due to poor
-representation of age-sex classes (Table \@ref(tab:tblDeploy)). Bearded seal models did
-include a latitudinal effect (and an interaction with day-of-year) since bearded seals
+representation of age-sex classes (Table \@ref(tab:tblDeploy)). Bearded seal models
+included a latitudinal effect (and an interaction with day-of-year), since bearded seals
 occupy a substantial range and we were interested in possible differences in the timing of
@@
 
-```{r wxCovMap, include=TRUE, fig.height = 5.5, fig.id="wxCovMap", fig.cap="Example spatial distribution of weather covariates across the study region for 2012 May 05.
+```{r wxCovMap, include=TRUE, fig.height = 5.5, fig.id="wxCovMap", fig.cap="Example spatial distribution of weather covariates across the study region for 05 May 2012.
 The NARR was used to predict weather covariates for each hour in a percent dry timeline record."}
@@
 when modeling the haul-out behavior of seal species with such a close association to sea
-ice. Seals haul out in the presence of sea ice and we could assess the local concentration
+ice; seals haul out in the presence of sea ice and we could assess the local concentration
 of sea ice during these events. However, seals also spend time in the water away from sea
@@
 
-We did explore whether the annual variation in maximum spring sea ice extent in the Bering
+We explored whether the annual variation in maximum spring sea ice extent in the Bering
 Sea influences the seasonal peak of seal haul-out behavior. In particular, we used sea ice
@@
 effects from the first model, but also permitted annual variation in haul-out timing. This
-second class of models was used to examine whether haul-out patterns varied by year, and
+second class of models was used to examine whether haul-out patterns varied by year and
 to determine the annual timing of apparent peaks in haul-out behavior.
@@
 
-To determine how much of a potential problem this is for our data, we compared estimates
-of haul-out probabilities that (1) ignored sex- and age-structure, and (2) accounted for
-sex- and age-structure, adjusting for the relative proportion of seals in each sex- and
-age-class. To simplify comparisons, we fitted two additional GLMPMs to ribbon and spotted
-seal haul-out datasets. In the first model, we expressed haul-out probabilities as a
-function of solar hour and day-of-year. In the second, we also included effects of sex-age
-class and sex-age interactions with solar hour and day-of-year. To adjust for the
-proportion of seals in each age class, we calculated a predicted population-averaged
-haul-out probability at solar noon as a function of day-of-year ($t$) as
+To determine how much of a potential problem this is within our data, we compared
+estimates of haul-out probabilities that (1) ignored sex- and age-structure and (2)
+accounted for sex- and age-structure, adjusting for the relative proportion of seals in
+each sex- and age-class. To simplify comparisons, we fitted two additional GLMPMs to
+ribbon and spotted seal haul-out datasets. In the first model, we expressed haul-out
+probabilities as a function of solar hour and day-of-year. In the second model, we also
+included effects of sex-age class and sex-age interactions with solar hour and
+day-of-year. To adjust for the proportion of seals in each age class, we calculated a
+predicted population-averaged haul-out probability at solar noon as a function of
+day-of-year ($t$) as
 
@@
 
-To calculate plausible values for $\omega_{s,a}$ we obtained natural mortality predictions
+To calculate plausible values for $\omega_{s,a}$, we obtained natural mortality predictions
 from a hierarchical meta-analysis of phocid natural mortality [@trukhanova2018ee]. We then
@@
 of $A_i$ so that they summed to one . Given a stable age distribution, we calculated the
-proportion of the population by sex and stage (young-of-the-year: 0-1 year; subadult: at
+proportion of the population by sex and stage (young-of-the-year: <1 year; subadult: at
 least one year but sexually immature; adult: sexually mature) using data on sexual
@@
 The haul-out probability for ribbon seals was strongly influenced by each of the weather
-covariates included within the model except precipitation. The impact of temperature
+covariates included within the model except precipitation. The impacts of temperature
 ($F_{`r paste(ribbon_type3$temp2$Num.df, ribbon_type3$temp2$Den.df, sep=",")`}$ = 
@@
 ($F_{`r paste(ribbon_type3$pressure$Num.df, ribbon_type3$pressure$Den.df, sep=",")`}$ = 
-`r ribbon_type3$pressure$F.value`; $p$ = `r ribbon_type3$pressure$Prob.F`) was especially
+`r ribbon_type3$pressure$F.value`; $p$ = `r ribbon_type3$pressure$Prob.F`) were especially
 noticeable as ribbon seals were less likely to haul out at higher winds and lower pressure
@@
 significant contributions for linear and quadratic interactions between day and year in
-both ribbon seals (day:year,
+both ribbon seals (day $\times$ year,
 $F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$ = 
@@
 `r ribbon_year_type3$day2_year$F.value`; $p$ = `r ribbon_year_type3$day2_year$Prob.F`) and
-spotted seals (day:year,
+spotted seals (day $\times$ year,
 $F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$ = 
@@
 Only those groups (age-sex + year) that included observations from more than one seal are shown.
-Additionally, any groups where the underlying data started after 01-June or ended before 01-May are not included."}
+Additionally, any groups where the underlying data started after 1 June or ended before 1 May are not included."}
 select_groups <- fit_ribbon_year$dataset %>%
@@
 a limited to no relationship with the amount of yearly maximum sea ice extent. Adult
-female and subadult spotted seals show a negative trend line but no indication that the
-observed trend is meaningful (spotted seal adult female: $R^{2}$ =
+female and subadult spotted seals show a negative trend line, but there is no indication
+that the observed trend is meaningful (spotted seal adult female: $R^{2}$ =
 `r pl_adultF$r.squared`, $p$ = `r pl_adultF$p.value`; spotted seal subadults: $R^{2}$ =
@@
 Chukchi seas. Our analysis shows seals use sea ice progressively more throughout the
-spring, and with peak use near mid-May to early June before declining again. This pattern
+spring, with peak use near mid-May to early June before declining again. This pattern
 aligns well with what has been previously documented and confirms our bio-logging
-deployments are likely capturing population level behavioral patterns. Seals
-preferentially haul out on ice shortly after solar noon which coincides with the daily
+deployments are likely capturing population-level behavioral patterns. Seals
+preferentially haul out on ice shortly after solar noon, which coincides with the daily
 period of maximum solar radiation. The association of daily peak haul-out behavior with
@@
 given what we know of their ecological behavior. Spotted seals are known to form triads
-during the breeding season where a female and dependent pup are accompanied on the ice by
-a suitor male. The male is waiting for the female to wean the pup and enter estrus and
-fending off any other males that may be interested. Triad formation results in both males
+during the breeding season, where a female and dependent pup are accompanied on the ice by
+a suitor male. The male waits for the female to wean the pup and enter estrus and
+fends off any other males that may be interested. Triad formation results in both males
 and females spending a large portion of the day hauled out on ice and a protracted spring
 haul-out season for both sexes. Since breeding occurs in the water, females may also be
-less inclined to venture out for foraging trips while still nursing the pup. We can see a
-clear reflection of this in the predicted haul-out behavior with both males and females
-exhibiting a broad distribution of time out of the water throughout the solar day and the
-season. Ribbon seals are not known to form triads and our model predicts a progression of
+less inclined to venture out for foraging trips while still nursing the pup. We see this
+reflected in the predicted haul-out behavior, with both males and females exhibiting a
+broad distribution of time out of the water throughout the solar day and the season.
+Ribbon seals are not known to form triads and our model predicts a progression of
 increased haul-out behavior with females starting earlier in the season than males.
@@
 while nursing. In the case of both ribbon and spotted seals, subadults are the first to
-begin meaningful haul-out behavior and reflects a typical phocid pattern where subadults
+begin meaningful haul-out behavior and reflect a typical phocid pattern where subadults
 molt first as they do not have any reproductive constraints.
@@
 These have been investigated for walrus (e.g. Udevitz et al. [@udevitz2009cjz]), but to
-our knowledge have never been investigated for ice-associated seals. Ribbon seals seem to
-be the most influenced by weather with wind, temperature, barometric pressure, and
+our knowledge have never been investigated for ice-associated seals. Ribbon seals seemed to
+be the most influenced by weather, with wind, temperature, barometric pressure, and
 precipitation all being important components of the model. Spotted seals were most
@@
 shown that there can be considerable differences in the number of seals hauled out on ice
-based on these factors. We thus recommend that future abundance analyses employ correction
+based on these factors. We recommend that future abundance analyses employ correction
 factors that account for these factors. For instance, it is relatively straightforward to
@@
 increased sample sizes in terms of number of animals, but also improvements to the way
-data are prepared prior to analysis.
+data were prepared prior to analysis.
 

### 2021-04-10 analysis/paper/paper.Rmd
UPDATE NAMES IN ACKNOWLEDGEMENTS

@@
 Utqiagvik, and Ulguniq and the following individuals: James Adams, Jeff Barger, David
-Barr, Wendell Booth, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin, John
-Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr., Virgil
-Naylor Sr., Dan Savetilik, Allen Stone, and Randy Toshavik from Kotzebue, Alaska; Billy
-Adams, Mary-Ellen Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard
-Kittick, Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik, Shawn Oktollik,
-Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross Schaeffer, Bob Shears, and Joe Skin
-from the North Slope Borough, Alaska;
+Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris, Grover Harris, Lee Harris, Tom
+Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin, Pearl Goodwin, Willie
+Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr., Virgil Naylor Sr., Dan Savetilik,
+Allen Stone, and Randy Toshavik from Kotzebue, Alaska; Billy Adams, Mary-Ellen Ahmaogak,
+James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard Kittick, Gilbert Leavitt,
+Isaac Leavitt, J.R. Leavitt, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, Fred Rexford,
+Chuck Schaeffer, Ross Schaeffer, Bob Shears, and Joe Skin from the North Slope Borough,
+Alaska;
 

### 2021-04-13 analysis/paper/paper.Rmd
TEXT EDITS AND ADD FORMAT DEPLOY TABLE INTEGERS

@@
 of the water almost exclusively on sea ice and are mostly pelagic outside the spring
-pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals
-rest on beaches during ice-free months, they strongly prefer sea ice as a resting platform
+pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals rest
+on beaches during ice-free months, they strongly prefer sea ice as a resting platform
 during the spring and early summer [@burns2009]. The remote nature of the Bering and
@@
 automated detection systems may miss some seals on ice), species misclassification, and
-possible disturbance of seals by aircraft [@conn2014mee;
-@verhoef2014sm].
+possible disturbance of seals by aircraft [@conn2014mee; @verhoef2014sm].
 
@@
 this study, in cases where both tag types were deployed, percent dry observations from the
-flipper tag were preferred. Sex and age class were determined at the time of
-deployment by experts in the field. Seals determined to be less than one year were
-classified as young-of-the-year. Subadults are those seals likely greater than one year of
-age but not yet sexually mature. Adults are likely sexually mature and older than
-approximately four years. For those tags deployed on young-of-the-year and transmitting
-into the next year, the age class was advanced to subadult on 1 March of the following
-year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
-from them.
+flipper tag were preferred. Sex and age class were determined at the time of deployment by
+experts in the field. Field identification of age class can be uncertain in some
+situations. The most notable is reliably discerning subadults from adults. In the case of
+ribbon seals, pelage differences between subadults and adults are distinguishable, with
+subadults often having less defined ribbons. Bearded seal subadults will often have a
+spotted pattern in the pelage that is not seen in adults. Spotted seal pelage is
+monomorphic and can not be used to discern age class. Despite these challenges, we feel
+the age classifications used in this analysis are reliable. Seals determined to be less
+than one year were classified as young-of-the-year. Subadults are those seals likely
+greater than one year of age but not yet sexually mature. Adults are likely sexually
+mature and older than approximately four years. For those tags deployed on
+young-of-the-year and transmitting into the next year, the age class was advanced to
+subadult on 1 March of the following year. Table \@ref(tab:tblDeploy) provides a summary
+of these deployments and data received from them.
 
@@
   dplyr::mutate(
-    deploy_stats = glue::glue("{n} ({seal_hours} seal hours)")
+    deploy_stats = glue::glue('{n} ({format(seal_hours, big.mark=",")} seal hours)')
   ) %>% 
@@
   dplyr::ungroup() %>% 
-  dplyr::rename(Species = species, Sex = sex, `Young of Year` = `YOUNG OF YEAR`,
+  dplyr::rename(Species = species, Sex = sex, `Young-of-the-Year` = `YOUNG OF YEAR`,
                 `Subadult` = SUBADULT, Adult = ADULT) %>% 
   flextable() %>% 
-  add_header(`Young of Year` = "Age Class",
-   `subadult` = "Age Class", Adult = "Age Class",
+  add_header(`Young-of-the-Year` = "Age Class",
+   Subadult = "Age Class", Adult = "Age Class",
    top = TRUE ) %>% 
@@
 tag firmware every few seconds and a percent of the hour in a dry state was calculated
-(Figure \@ref(fig:examplePlot)). Each data transmission message consisted of a complete 24-hour
-record. Tags were deployed and programmed in a manner to maximize data transmission during
-the spring pupping and molting period, however, not all days during the deployment were
-successfully transmitted. This is due to a variety of factors including satellite
-coverage, tag availability (i.e. tags mounted to the rear flipper are not available to the
-surface while the seal is at sea), tag performance, duty cycling, and extra-terrestrial
-interference.
+(Figure \@ref(fig:examplePlot)). Each data transmission message consisted of a complete
+24-hour record. Tags were deployed and programmed in a manner to maximize data
+transmission during the spring pupping and molting period, however, not all days during
+the deployment were successfully transmitted. This is due to a variety of factors
+including satellite coverage, tag availability (i.e. tags mounted to the rear flipper are
+not available to the surface while the seal is at sea), tag performance, duty cycling, and
+extra-terrestrial interference.
 
@@
 Of key interest in this study is the relationship between haul-out behavior and a variety
-of environmental covariates that vary with space and time. We explored the use of a movement
-model to predict locations at specific times. However, the sparse nature of data from some
-tags, especially those mounted to the rear flipper, resulted in poor modeling performance
-or convergence issues. For this study, we calculated a weighted average daily location
-where the inverse of the estimated Argos location error was used for the weight. Any days
-where haul-out observations were present without any location data were filled from the
-last calculated daily location, and any days where the location intersected with land were
-removed from the dataset. Figure \@ref(fig:dataMap) shows the spatial distribution of
-available haul-out behavior data for each species across the study area.
+of environmental covariates that vary with space and time. We explored the use of a
+movement model to predict locations at specific times. However, the sparse nature of data
+from some tags, especially those mounted to the rear flipper, resulted in poor modeling
+performance or convergence issues. For this study, we calculated a weighted average daily
+location where the inverse of the estimated Argos location error was used for the weight.
+Any days where haul-out observations were present without any location data were filled
+from the last calculated daily location, and any days where the location intersected with
+land were removed from the dataset. Figure \@ref(fig:dataMap) shows the spatial
+distribution of available haul-out behavior data for each species across the study area.
 
@@
 explain variation in haul-out probabilities. These included day-of-year (for seasonal
-effects) and solar hour (for diurnal effects). Solar hour was calculated using the {solaR}
-package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019] using the
-daily locations. We also compiled several meteorological variables shown to affect
-haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz; @perry2017]. In
-particular, we linked locations from satellite tags to key weather values from the North
-American Regional Reanalysis (NARR) model produced by the National Centers for
-Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
+effects) and local solar hour (for diurnal effects). Solar hour was calculated using the
+{solaR} package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019]
+using the daily locations. We also compiled several meteorological variables shown to
+affect haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz;
+@perry2017]. In particular, we linked locations from satellite tags to key weather values
+from the North American Regional Reanalysis (NARR) model produced by the National Centers
+for Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
 data to produce a long-term picture of weather over North America. Numerous weather
@@
 Models for all species include the following effects: day-of-year, solar hour,
-temperature, wind speed, barometric pressure, precipitation, wind chill (represented
-by a wind $\times$ temperature interaction [@udevitz2009cjz]), and day-of-year
-and time of day interactions to permit diurnal patterns to change throughout the year.
-Spotted and ribbon seal models included age-sex class and interactions between day-of-year
-and age-sex class, but we omitted these from bearded seal models due to poor
-representation of age-sex classes (Table \@ref(tab:tblDeploy)). Bearded seal models
-included a latitudinal effect (and an interaction with day-of-year), since bearded seals
-occupy a substantial range and we were interested in possible differences in the timing of
-haul-out behavior along a latitudinal gradient. We omitted the latitudinal effect from
-ribbon and spotted seal models since, during the spring, these species are most prevalent
-near the southern ice edge in the Bering Sea [@conn2014mee].
+temperature, wind speed, barometric pressure, precipitation, wind chill (represented by a
+wind $\times$ temperature interaction [@udevitz2009cjz]), and day-of-year and time of day
+interactions to permit diurnal patterns to change throughout the year. Spotted and ribbon
+seal models included age-sex class and interactions between day-of-year and age-sex class,
+but we omitted these from bearded seal models due to poor representation of age-sex
+classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
+(and an interaction with day-of-year), since bearded seals occupy a substantial range and
+we were interested in possible differences in the timing of haul-out behavior along a
+latitudinal gradient. We omitted the latitudinal effect from ribbon and spotted seal
+models since, during the spring, these species are most prevalent near the southern ice
+edge in the Bering Sea [@conn2014mee].
 
@@
 grid cells (25 km^2^) in the study area with greater than 15% concentration were summed to
-get the total sea ice extent for each day between 15 February and 15 July across all years.
-Maximum spring sea ice extent was simply the largest daily value for each year.
+get the total sea ice extent for each day between 15 February and 15 July across all
+years. Maximum spring sea ice extent was simply the largest daily value for each year.
 
@@
   "Age-sex class",   "Categorical", "Field Assessment", 
-  "Classes included young-of-year, subadult, adult male and adult female. Young-of-year tags surviving past ~Feb 1 of the next year were treated as subadult in subsequent years.",
+  "Classes included young-of-the-year, subadult, adult male and adult female. Young-of-the-year tags surviving past 1 March of the next year were treated as subadult in subsequent years.",
   "Hour", "Continuous; Fourier basis", "Bio-logger", 
-  "We modeled solar hour using 6 variables of a Fourier-series basis to allow circularity, flexibility, and dimension reduction ",
+  "Local solar hour using 6 variables of a Fourier-series basis to allow circularity, flexibility, and dimension reduction ",
   "Day",   "Continuous", "Bio-logger", 
@@
 heterogeneity in individual behavior. We used the {glmmLDTS} package [@verhoef2010cs] to
-implement GLMPMs in the R programming environment . Owing to the large number of records,
-we fit separate models to bearded, ribbon, and spotted seal data sets. We fit two
-different models to our datasets. First, for each species, we fitted a year-independent
-model that predicted average haul-out behavior as a function of demographic,
-environmental, seasonal, and diurnal effects. Second, for ribbon and spotted seals (which
-had considerably more data than bearded seals), we fitted models that included all the
-effects from the first model, but also permitted annual variation in haul-out timing. This
-second class of models was used to examine whether haul-out patterns varied by year and
-to determine the annual timing of apparent peaks in haul-out behavior.
+implement GLMPMs in the R programming environment [@rcoreteam2021]. Owing to the large
+number of records, we fit separate models to bearded, ribbon, and spotted seal data sets.
+We fit two different models to our datasets. First, for each species, we fitted a
+year-independent model that predicted average haul-out behavior as a function of
+demographic, environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
+seals (which had considerably more data than bearded seals), we fitted models that
+included all the effects from the first model, but also permitted annual variation in
+haul-out timing. This second class of models was used to examine whether haul-out patterns
+varied by year and to determine the annual timing of apparent peaks in haul-out behavior.
 
@@
 year$\times$day and year$\times$day2, omitting the main effects of year as well as
-year$\times$day3 interactions because models with the latter effects were numerically unstable.
-However, the modeled interactions were sufficient to allow shifts in haul-out
+year$\times$day3 interactions because models with the latter effects were numerically
+unstable. However, the modeled interactions were sufficient to allow shifts in haul-out
 distribution, as one can show mathematically that a simple horizontal shift in timing of
@@
 A typical model fitting exercise would also include a model selection process. However,
-AIC (and similar criteria) is not suitable when using pseudo-likelihoods. After fitting
-GLMPM models, we used "type III" F-tests to calculate p-values [@verhoef2010cs] to
-evaluate model performance and important terms. We also produced predictions of haul-out
-behavior as a function of influential predictors (e.g. solar hour, day-of-year, age-sex).
-Weather covariates for these predictions were based on daily or hourly smoothed weather
-covariate values across the study region. Such predictions were then used to develop
-haul-out probability surfaces, explore marginal effects of weather covariates, and
-determine annual peaks in haul-out activity. The timing of peak haul-out behavior was
-further used to regress against the annual maximum sea ice extent in the study region.
+AIC (and similar criteria) is not suitable when using pseudo-likelihoods, because
+pseudo-data generated in the model fitting process [@verhoef2010cs] differ between models
+[@teneyck2017]. After fitting GLMPM models, we instead used "type III" F-tests to
+calculate p-values [@verhoef2010cs] to evaluate model performance and important terms. We
+also produced predictions of haul-out behavior as a function of influential predictors
+(e.g. solar hour, day-of-year, age-sex). Weather covariates for these predictions were
+based on daily or hourly smoothed weather covariate values across the study region. Such
+predictions were then used to develop haul-out probability surfaces, explore marginal
+effects of weather covariates, and determine annual peaks in haul-out activity. The timing
+of peak haul-out behavior was further used to regress against the annual maximum sea ice
+extent in the study region.
 
@@
 
-To calculate plausible values for $\omega_{s,a}$, we obtained natural mortality predictions
-from a hierarchical meta-analysis of phocid natural mortality [@trukhanova2018ee]. We then
-assumed a 50/50 sex ratio at birth, 1 pup per litter, and reproductive schedules as
-reported by Fedoseev [-@fedoseev2000] for seals in the western Bering Sea to formulate
-Leslie matrices Ai for each species i. Stable age distributions
+To calculate plausible values for $\omega_{s,a}$, we obtained natural mortality
+predictions from a hierarchical meta-analysis of phocid natural mortality
+[@trukhanova2018ee]. We then assumed a 50/50 sex ratio at birth, 1 pup per litter, and
+reproductive schedules as reported by Fedoseev [-@fedoseev2000] for seals in the western
+Bering Sea to formulate Leslie matrices Ai for each species i. Stable age distributions
 $v_i={v\_i0,v\_i1,⋯,v\_iA}$ were then determined by normalizing the dominant eigenvector
 of $A_i$ so that they summed to one . Given a stable age distribution, we calculated the
-proportion of the population by sex and stage (young-of-the-year: <1 year; subadult: at
+proportion of the population by sex and stage (young-of-the-year: \<1 year; subadult: at
 least one year but sexually immature; adult: sexually mature) using data on sexual
@@
     mutate(age_sex = forcats::fct_relevel(
-      age_sex,c("ADULT.F","ADULT.M","SUBADULT","YOUNG OF YEAR"))
+      age_sex,c("ADULT.F","ADULT.M","SUBADULT","YOUNG-OF-THE YEAR"))
     )
@@
 Models omitting year effects suggested that day-of-year, solar hour, age-sex class,
-temperature, and wind substantially altered haul-out behavior of all three species, with F
-tests producing p-values less than 0.05 for variables embodying these effects and/or their
-interactions. Haul-out probabilities typically increased throughout March and April,
-reaching a peak in May and early June before declining again. Diurnal patterns were
+temperature, and wind substantially altered haul-out behavior of all three species, with
+$F$ tests producing $p$-values less than 0.05 for variables embodying these effects
+and/or their interactions. Haul-out probabilities typically increased throughout March and
+April, reaching a peak in May and early June before declining again. Diurnal patterns were
 present, with maximum haul-out behavior centered around local solar noon.
@@
 ($F_{`r paste(ribbon_type3$wind$Num.df, ribbon_type3$wind$Den.df, sep=",")`}$ =
-`r ribbon_type3$wind$F.value`; $p$ = `r ribbon_type3$wind$Prob.F`), and barometric pressure
+`r ribbon_type3$wind$F.value`; $p$ = `r ribbon_type3$wind$Prob.F`), and barometric
+pressure
 ($F_{`r paste(ribbon_type3$pressure$Num.df, ribbon_type3$pressure$Den.df, sep=",")`}$ =
@@
 both ribbon seals (day $\times$ year,
-$F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$ = 
-`r ribbon_year_type3$day_year$F.value`; $p$ = `r ribbon_year_type3$day_year$Prob.F`;
+$F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$
+= `r ribbon_year_type3$day_year$F.value`; $p$ = `r ribbon_year_type3$day_year$Prob.F`;
 day^2^:year,
-$F_{`r paste(ribbon_year_type3$day2_year$Num.df, ribbon_year_type3$day2_year$Den.df, sep=",")`}$ = 
-`r ribbon_year_type3$day2_year$F.value`; $p$ = `r ribbon_year_type3$day2_year$Prob.F`) and
-spotted seals (day $\times$ year,
-$F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$ = 
-`r spotted_year_type3$day_year$F.value`; $p$ = `r spotted_year_type3$day_year$Prob.F`;
+$F_{`r paste(ribbon_year_type3$day2_year$Num.df, ribbon_year_type3$day2_year$Den.df, sep=",")`}$
+= `r ribbon_year_type3$day2_year$F.value`; $p$ = `r ribbon_year_type3$day2_year$Prob.F`)
+and spotted seals (day $\times$ year,
+$F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$
+= `r spotted_year_type3$day_year$F.value`; $p$ = `r spotted_year_type3$day_year$Prob.F`;
 day^2^:year,
-$F_{`r paste(spotted_year_type3$day2_year$Num.df, spotted_year_type3$day2_year$Den.df, sep=",")`}$ = 
-`r spotted_year_type3$day2_year$F.value`; $p$ = `r spotted_year_type3$day2_year$Prob.F`).
-Predicted distributions of haul-out activity were largely unimodal, but varied some among
-and within years with respect to the timing and magnitude of haul-out peaks (Figure
-\@ref(fig:annualHO)). It is important to note that predicted variation in annual haul-out
-patterns likely reflects both process error and sampling variability. While we did remove
-any years where only one deployment in a species + age-sex group was present, there are
-still some years where the pattern shown is informed by a small number of tags that may
-not represent population level processes.
+$F_{`r paste(spotted_year_type3$day2_year$Num.df, spotted_year_type3$day2_year$Den.df, sep=",")`}$
+= `r spotted_year_type3$day2_year$F.value`; $p$ =
+`r spotted_year_type3$day2_year$Prob.F`). Predicted distributions of haul-out activity
+were largely unimodal, but varied some among and within years with respect to the timing
+and magnitude of haul-out peaks (Figure \@ref(fig:annualHO)). It is important to note that
+predicted variation in annual haul-out patterns likely reflects both process error and
+sampling variability. While we did remove any years where only one deployment in a species
++ age-sex group was present, there are still some years where the pattern shown is
+informed by a small number of tags that may not represent population level processes.
 
@@
 during the breeding season, where a female and dependent pup are accompanied on the ice by
-a suitor male. The male waits for the female to wean the pup and enter estrus and
-fends off any other males that may be interested. Triad formation results in both males
-and females spending a large portion of the day hauled out on ice and a protracted spring
+a suitor male. The male waits for the female to wean the pup and enter estrus and fends
+off any other males that may be interested. Triad formation results in both males and
+females spending a large portion of the day hauled out on ice and a protracted spring
 haul-out season for both sexes. Since breeding occurs in the water, females may also be
@@
 These have been investigated for walrus (e.g. Udevitz et al. [@udevitz2009cjz]), but to
-our knowledge have never been investigated for ice-associated seals. Ribbon seals seemed to
-be the most influenced by weather, with wind, temperature, barometric pressure, and
+our knowledge have never been investigated for ice-associated seals. Ribbon seals seemed
+to be the most influenced by weather, with wind, temperature, barometric pressure, and
 precipitation all being important components of the model. Spotted seals were most
@@
 
-Our model uncovered annual deviations in the timing and magnitude of
-haul-out peaks for ribbon and spotted seals. The timing of peak haul-out activity appears
-to fall within a relatively narrow time window of 3-4 weeks in late May and early June. 
-This remarkable level of consistency across 15 years is indicative of the 
-adaptations these species have evolved to synchronize their key life history stages
-with favorable environmental conditions (e.g. availability of sea ice, increased
-oceanographic productivity, favorable weather conditions). Unfortunately, this may also
-be indicative of a limited range of plasticity as climatological conditions change.
-Additionally, while from an ecological perspective the haul-out behavior appears
-consistent, the interannual differences in timing and magnitude are large enough to have
-important ramifications on calculations of abundance and trend.
+Our model uncovered annual deviations in the timing and magnitude of haul-out peaks for
+ribbon and spotted seals. The timing of peak haul-out activity appears to fall within a
+relatively narrow time window of 3-4 weeks in late May and early June. This remarkable
+level of consistency across 15 years is indicative of the adaptations these species have
+evolved to synchronize their key life history stages with favorable environmental
+conditions (e.g. availability of sea ice, increased oceanographic productivity, favorable
+weather conditions). Unfortunately, this may also be indicative of a limited range of
+plasticity as climatological conditions change. Additionally, while from an ecological
+perspective the haul-out behavior appears consistent, the interannual differences in
+timing and magnitude are large enough to have important ramifications on calculations of
+abundance and trend.
 
@@
 insert broad statement and summary regarding data contribution from AFSC, ADFG, NSB (I
-still need to lookup details of which org contributed which data). Authors, please
-add your contribution statement below
+still need to lookup details of which org contributed which data). Authors, please add
+your contribution statement below
 :::
 
-**Josh M. London:** led field work, conceptualization, methodology, formal analysis, code development, writing of original draft, visualization, review, revisions, and corresponding author
-
-**Paul B. Conn:** conceptualization, methodology, formal analysis, code development, writing of original draft, review, and revisions
-
-**Stacie K. Hardy:** field work, methodology, data curation, validation, review, and revisions
+**Josh M. London:** led field work, conceptualization, methodology, formal analysis, code
+development, writing of original draft, visualization, review, revisions, and
+corresponding author
+
+**Paul B. Conn:** conceptualization, methodology, formal analysis, code development,
+writing of original draft, review, and revisions
+
+**Stacie K. Hardy:** field work, methodology, data curation, validation, review, and
+revisions
 
@@
 
-**Jay M. Ver Hoef:** field work, conceptualization, methodology, code development, review, and revisions
-
-**Irina S. Trukhanova:** conceptualization, formal analysis, code development, review, and revisions
-
-**Michael F. Cameron:** led field work, project administration, methodology, review, and revisions
-
-**Peter L. Boveng:** led field work, supervision, project administration, conceptualization, methodology, review, and revisions
+**Jay M. Ver Hoef:** field work, conceptualization, methodology, code development, review,
+and revisions
+
+**Irina S. Trukhanova:** conceptualization, formal analysis, code development, review, and
+revisions
+
+**Michael F. Cameron:** led field work, project administration, methodology, review, and
+revisions
+
+**Peter L. Boveng:** led field work, supervision, project administration,
+conceptualization, methodology, review, and revisions
 
@@
 
-Data and code are available as a complete research compendium on GitHub and major
-versions are archived at Zenodo. Original data sources for telemetry are archived at
-the United States Animal Telemetry Network.
+Data and code are available as a complete research compendium on GitHub and major versions
+are archived at Zenodo. Original data sources for telemetry are archived at the United
+States Animal Telemetry Network.
 

### 2021-04-14 analysis/paper/paper.Rmd
EXPLORED CHANGES TO CALENDAR PLOT
explored adding day of week labels back. left as is for now.

@@
                      labels = c("April", "May", "June", "July")) +
+  #scale_y_discrete(labels = c("Sat","","","Wed","","","Sun"))+
   guides(color = "none",
@@
   facet_grid(species ~ .) + 
-  theme(axis.text.y = element_blank(),
+  theme(
+        axis.text.y = element_blank(),
         panel.grid = element_blank(),

### 2021-04-14 analysis/paper/paper.Rmd
EXPLORE EXAMPLE DATA COLOR OPTIONS
there is a commented out section that relies on the latest viridisLite. But, no success or significatn improvement re: discerning variation in the lower values.

@@
   geom_tile(color= "white",size=0) + 
+  # scale_fill_gradientn(colors = viridisLite::turbo(n=256),
+  #                      name = "hourly percent dry", aesthetics = "fill",
+  #   guide = guide_colorbar(title.position = "bottom", barwidth = 15, 
+  #                          barheight = 0.5, title.hjust = 0.5)
+  #   )
   scale_fill_viridis_c(name = "hourly percent dry", aesthetics = "fill",

### 2021-04-14 analysis/paper/paper.Rmd
UPDATE EXPLANTORY COV TABLE

@@
   "Age-sex class",   "Categorical", "Field Assessment", 
-  "Classes included young-of-the-year, subadult, adult male and adult female. Young-of-the-year tags surviving past 1 March of the next year were treated as subadult in subsequent years.",
+  "young-of-the-year, subadult, adult male and adult female",
   "Hour", "Continuous; Fourier basis", "Bio-logger", 
-  "Local solar hour using 6 variables of a Fourier-series basis to allow circularity, flexibility, and dimension reduction ",
+  "local solar hour using 6 variables of a Fourier-series basis",
   "Day",   "Continuous", "Bio-logger", 
-  "Linear, quadratic, and cubic effects of day-of-year were modeled",
+  "linear, quadratic, and cubic effects of day-of-year",
   "Precip", "Continuous", "NARR", 
-  "Convective precipitation (kg/m2)",
+  "convective precipitation (kg/m2)",
   "Pressure", "Continuous", "NARR", 
-  "Atmospheric pressure at sea level (kPa)",
+  "atmospheric pressure at sea level (kPa)",
   "Temp", "Continuous", "NARR",
-  "Air temperatures predicted at 2m above the earth’s surface",
+  "air temperatures at 2m above the earth’s surface",
   "Wind", "Continuous", "NARR", 
-  "Northerly and easterly vector components for wind were downloaded and converted into a single wind speed via the Euclidean norm.",
-  "Northing", "Continuous", "Bio-logger","For bearded seals a measurement of northing was caclulated as latitude divided by the mean latitude across all locations"
+  "northerly and easterly vector components for wind converted into a single wind speed via the Euclidean norm",
+  "Northing", "Continuous", "Bio-logger","latitude divided by the mean latitude across all locations (for bearded seals only)"
 )

### 2021-04-14 analysis/paper/paper.Rmd
CHANGE COLOR PALETTE TO YlGnBu
SKH suggested finding a better color palette for the example data record in figure. YlGnBu from colorbrewer provides better visualization for the lower end of value range.

@@
   geom_tile(color= "white",size=0) + 
-  # scale_fill_gradientn(colors = viridisLite::turbo(n=256),
-  #                      name = "hourly percent dry", aesthetics = "fill",
+  scale_fill_distiller(palette = "YlGnBu",
+                       name = "hourly percent dry", aesthetics = "fill",
+    guide = guide_colorbar(title.position = "bottom", barwidth = 15,
+                           barheight = 0.5, title.hjust = 0.5)
+    )
+  # scale_fill_viridis_c(name = "hourly percent dry", aesthetics = "fill",
   #   guide = guide_colorbar(title.position = "bottom", barwidth = 15,
@@
   # )
-  scale_fill_viridis_c(name = "hourly percent dry", aesthetics = "fill",
-    guide = guide_colorbar(title.position = "bottom", barwidth = 15,
-                           barheight = 0.5, title.hjust = 0.5)
-  )
 p <- p + facet_grid(year~month)
@@
 
-In addition to sex and age class (non-dependent *young-of-the-year*, sexually immature
-*subadults*, and mature *adults*), we gathered additional variables that might help
+In addition to sex and age class, we gathered additional variables that might help
 explain variation in haul-out probabilities. These included day-of-year (for seasonal
@@
 heterogeneity in individual behavior. We used the {glmmLDTS} package [@verhoef2010cs] to
-implement GLMPMs in the R programming environment [@rcoreteam2021]. Owing to the large
-number of records, we fit separate models to bearded, ribbon, and spotted seal data sets.
-We fit two different models to our datasets. First, for each species, we fitted a
-year-independent model that predicted average haul-out behavior as a function of
-demographic, environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
-seals (which had considerably more data than bearded seals), we fitted models that
-included all the effects from the first model, but also permitted annual variation in
-haul-out timing. This second class of models was used to examine whether haul-out patterns
-varied by year and to determine the annual timing of apparent peaks in haul-out behavior.
+implement GLMPMs in the R programming environment [@rcoreteam2021]. We explored two
+different model formulations for our data, and owing to the large number of records, we
+fit separate models to bearded, ribbon, and spotted seal data sets.  In our first model
+formulation, for each species, we fitted a year-independent model that predicted average
+haul-out behavior as a function of demographic, environmental, seasonal, and diurnal
+effects. Second, for ribbon and spotted seals (which had considerably more data than
+bearded seals), we fitted models that included all the effects from the first model, but
+also permitted annual variation in haul-out timing. This second class of models was used
+to examine whether haul-out patterns varied by year and to determine the annual timing of
+apparent peaks in haul-out behavior.
 
@@
 + age-sex group was present, there are still some years where the pattern shown is
-informed by a small number of tags that may not represent population level processes.
+informed by a small number of individuals that may not represent population level processes.
 

### 2021-04-14 analysis/paper/paper.Rmd
REMOVE REFERENCES TO STABLE STAGE
stable stage distributions and application to haul-out corrections are being moved to a separate paper for the Western Bering.

@@
 variables that have been shown to influence haul-out behavior of walruses [@reder2003pb;
-@udevitz2009cjz]. Further, previous correction factor analyses assumed that tagged seals
-were representative of the population as a whole, even though the age and sex distribution
-of telemetered individuals might differ from the population. Third, we explore the annual
-variability in haul-out timing and possible linkage to dramatic changes in the
-availability and extent of seasonal sea ice between 2005 and 2020. Our work extends the
-scope of previous haul-out analyses, includes the influence of meteorological variability
-and possible ramifications of non-representative age-sex samples, and investigates the
-potential impact of a changing icescape on the behavior of these species.
+@udevitz2009cjz]. Third, we explore the annual variability in haul-out timing and possible
+linkage to dramatic changes in the availability and extent of seasonal sea ice between
+2005 and 2020. Our work extends the scope of previous haul-out analyses, includes the
+influence of meteorological variability and possible ramifications of non-representative
+age-sex samples, and investigates the potential impact of a changing icescape on the
+behavior of these species.
 
@@
 extent in the study region.
-
-## Stable stage distributions
-
-::: {custom-style="comment"}
-I didn't update this at all from Paul's previous text other than to revise some of the
-latex equations and add references. Would be good to confirm all of this is still
-appropriate and makes sense.
-:::
-
-Previous availability correction factors used in aerial surveys for ice-associated Arctic
-seals (e.g. [@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]) ignored any age- and
-sex-based variation in haul-out probabilities. The implicit assumption is that the
-behavior of telemetered animals represents the population as a whole. However, if haul-out
-probabilities are related to sex- and age-class, and if there are tangible differences
-between the age- and sex-structure of the telemetered sample and that of the population,
-correction factors computed in this way may lead to biased abundance estimates.
-
-To determine how much of a potential problem this is within our data, we compared
-estimates of haul-out probabilities that (1) ignored sex- and age-structure and (2)
-accounted for sex- and age-structure, adjusting for the relative proportion of seals in
-each sex- and age-class. To simplify comparisons, we fitted two additional GLMPMs to
-ribbon and spotted seal haul-out datasets. In the first model, we expressed haul-out
-probabilities as a function of solar hour and day-of-year. In the second model, we also
-included effects of sex-age class and sex-age interactions with solar hour and
-day-of-year. To adjust for the proportion of seals in each age class, we calculated a
-predicted population-averaged haul-out probability at solar noon as a function of
-day-of-year ($t$) as
-
-$$
-\widetilde{p_t}=\sum_{s}\sum_{a}{\omega_{s,a}{\hat{p}}_{s,a,t}}
-$$
-
-where $\omega_{s,a}$ is the proportion of the population of seals in age class $a$ and sex
-$s$. Typically, we do not know $\omega_{s,a}$ exactly, but can use information about
-likely mortality and reproductive schedules of seals to approximate $\omega_{s,a}$ using
-stable stage distributions calculated from matrix population models ().
-
-To calculate plausible values for $\omega_{s,a}$, we obtained natural mortality
-predictions from a hierarchical meta-analysis of phocid natural mortality
-[@trukhanova2018ee]. We then assumed a 50/50 sex ratio at birth, 1 pup per litter, and
-reproductive schedules as reported by Fedoseev [-@fedoseev2000] for seals in the western
-Bering Sea to formulate Leslie matrices Ai for each species i. Stable age distributions
-$v_i={v\_i0,v\_i1,⋯,v\_iA}$ were then determined by normalizing the dominant eigenvector
-of $A_i$ so that they summed to one . Given a stable age distribution, we calculated the
-proportion of the population by sex and stage (young-of-the-year: \<1 year; subadult: at
-least one year but sexually immature; adult: sexually mature) using data on sexual
-maturity from Fedoseev [-@fedoseev2000]. In particular, the expected proportion of the
-population that are sex $s$ and young-of-year was set as $\pi_{i,s,yoy}=0.5v_i0$, while
-the expected proportion that are subadults and adults were set to
-
-$$
-\pi_{i,s,sub}={0.5\sum_{a}{{(1-m}_{i,s,a})}v}_{ia}
-$$
-
-and
-
-$$
-\pi_{i,s,ad}={0.5\sum_{a}\ m_{i,s,a}v}_{ia}
-$$
-
-Here, $m_{i,s,a}$ gives the anticipated proportion of animals of species $i$ and sex $s$
-that are sexually mature. Further details on mortality, reproductive and maturity
-schedules, and matrix modeling methods are presented in Appendix A.
 
@@
 
-## Stable stage distributions
-
-::: {custom-style="comment"}
-This section is unchanged from Paul's original write up. I have put off revisiting and
-integrating with my current document/code. Maybe between Paul/Irina/Me we can sort out a
-quick update. Might just need to walk me through the previous code and make sure I get it
-properly incorporated with this version
-:::
-
-Stable stage distributions calculated using mortality and reproductive schedules produced
-different patterns for the four species of ice-associated seals residing in Alaskan waters
-(Fig. X). Ribbon seals mature the fastest, and so can be expected to have the largest
-proportion of adults and smallest proportion of subadults; conversely, ringed seals mature
-slowly and are predicted to have roughly equal proportions of adults and subadults (note
-that we do not attempt to model ringed seal haul-out behavior in this study; see
-Discussion). Population-level haul-out distributions were somewhat different depending on
-whether we a) adjusted haul-out probabilities based on stable stage distributions or b)
-assumed the tagged sample represented the population. In particular, ribbon seal estimates
-differed somewhat at the beginning of the time series, but the two approaches produced
-similar estimates (Fig. X) within the window that previous aerial surveys in the region
-have been conducted (April and May; e.g. Conn et al. 2014). By contrast, during this time
-period spotted seal haul-out estimates were often ≈10% higher for the approach that
-adjusted for stable age distributions, presumably because subadults had lower predicted
-haul-out probabilities than adults but made up a disproportionately high proportion of the
-tagged sample.
-
 # Discussion
@@
 obtain weather reanalysis products for times and locations that are surveyed and to
-construct a relevant correction factor based on predictions of GMPMs. Further, one can
-apply stable stage distribution estimates from matrix population models to adjust
-predictions to the "average" prediction of the population. The latter suggestion is
-particularly appropriate when the age structure of the tagged sample does not reflect the
-likely age structure of the population. For instance, population weighted estimates can
-differ substantially from those based on a naïve analysis of the tagged sample, as was the
-case with analysis of spotted seal data (e.g. Fig X). The most challenging element in
-developing availability correction factors is with annual variability. It can be difficult
-to get a sufficient sample size to estimate year-specific correction factors, particularly
-because research teams would likely need to tag seals and conduct aerial surveys
-concurrently which requires considerably more personnel and money. One possible suggestion
-is to estimate a "shift" parameter within models for aerial survey counts that allow the
-peak of haul-out distributions to be adjusted earlier or later in the year based on the
-frequency of counts observed over time. Regardless, researchers should anticipate there
-being some unmodeled heterogeneity in availability probability present in abundance
+construct a relevant correction factor based on predictions of GMPMs. The most challenging
+element in developing availability correction factors is with annual variability. It can
+be difficult to get a sufficient sample size to estimate year-specific correction factors,
+particularly because research teams would likely need to tag seals and conduct aerial
+surveys concurrently which requires considerably more personnel and money. One possible
+suggestion is to estimate a "shift" parameter within models for aerial survey counts that
+allow the peak of haul-out distributions to be adjusted earlier or later in the year based
+on the frequency of counts observed over time. Regardless, researchers should anticipate
+there being some unmodeled heterogeneity in availability probability present in abundance
 estimates obtained from aerial surveys. This may make trend detection difficult, as one

### 2021-04-14 analysis/paper/paper.Rmd
FLIP AXES FOR PEAK HAUL OUT vs EXTENT
per suggestion from Paul. Also adjusted the sea ice extent to smaller numbers and updated axis labels

@@
   mutate(peak_diff_days = yday - max_extent_yday) %>% 
-  distinct()
+  distinct() %>% 
+  mutate(max_extent_km2 = max_extent_km2/1000)
 
 models <- peak_ho_df %>% 
-  mutate(max_extent_km2 = max_extent_km2/1e6) %>% 
   group_by(species, age_sex) %>%  
@@
 While spotted seal adult females and subadults show a negative trend line, none of the relationships were determined to be significant."}
-ggplot(data = peak_ho_df, aes(x = yday, y = max_extent_km2, color = species)) + 
+ggplot(data = peak_ho_df, aes(y = yday, x = max_extent_km2, color = species)) + 
   geom_smooth(method = "lm", alpha = 0.2, aes(fill = species)) +
@@
   facet_grid(age_sex~.) +
-  scale_x_continuous(breaks = c(125,135,145, 156),
+  scale_y_continuous(breaks = c(125,135,145, 156),
                      labels = c("05-May", "15-May", "25-May", "05-Jun")) +
@@
         legend.justification = "right") +
-  labs(x = "date of peak haul-out", y = "maximum sea ice extent (km^2 / 1e6)",
+  labs(y = "date of peak haul-out", x = "maximum sea ice extent (km^2 / 1000)",
        title = "The connection between yearly maximum sea ice extent and the \ntiming of peak haul-out probability is limited")

### 2021-04-15 analysis/paper/paper.Rmd
IMPROVING FIG CAPTIONS + ODDS & ENDS

@@
 library(gghighlight)
+library(ggtext)
 library(glmmLDTS)
@@
 Any days where haul-out observations were present without any location data were filled
-from the last calculated daily location, and any days where the location intersected with
-land were removed from the dataset. Figure \@ref(fig:dataMap) shows the spatial
-distribution of available haul-out behavior data for each species across the study area.
+from the last calculated weighted average daily location, and any days where the location
+intersected with land were removed from the dataset. Figure \@ref(fig:dataMap) shows the
+spatial distribution of available haul-out behavior data for each species across the study
+area.
 
@@
 pseudo-data generated in the model fitting process [@verhoef2010cs] differ between models
-[@teneyck2017]. After fitting GLMPM models, we instead used "type III" F-tests to
-calculate p-values [@verhoef2010cs] to evaluate model performance and important terms. We
+[@teneyck2017]. After fitting GLMPM models, we instead used "type III" $F$-tests to
+calculate $p$-values [@verhoef2010cs] to evaluate model performance and important terms. We
 also produced predictions of haul-out behavior as a function of influential predictors
@@
   theme(axis.ticks=element_blank()) +
-  xlab("day of month") + ylab("solar hour") +
+  xlab("day of month") + ylab("local solar hour") +
   ggtitle("Ribbon seals rest on sea ice in late spring and around solar noon.")
@@
 
-```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Predicted haul-out probability of ribbon seals across a range of weather conditions."}
+```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Variability in predicted haul-out probability of ribbon seals across a range of weather conditions thea range of weather conditions encountered in the observed data.
+Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
+Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}
 
@@
   theme(axis.ticks=element_blank()) +
-  xlab("day of month") + ylab("solar hour") +
+  xlab("day of month") + ylab("local solar hour") +
   ggtitle("Spotted seals rest on sea ice in late spring and around solar noon.")
@@
 
-```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Predicted haul-out probability of spotted seals across a range of weather conditions."}
+```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Variability in the predicted haul-out probability of spotted seals across a range of weather conditions thea range of weather conditions encountered in the observed data.
+Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
+Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}
 library(ggsci)
@@
   theme(axis.ticks=element_blank()) +
-  xlab("day of month") + ylab("solar hour") +
+  xlab("day of month") + ylab("local solar hour") +
   ggtitle("Bearded seals rest on sea ice around solar noon and solar midnight")
@@
 
-```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Predicted haul-out probability of bearded seals across a range of weather conditions."}
+```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across a range of weather conditions thea range of weather conditions encountered in the observed data.
+Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
+Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}
 
@@
         legend.justification = "right") +
-  labs(y = "date of peak haul-out", x = "maximum sea ice extent (km^2 / 1000)",
-       title = "The connection between yearly maximum sea ice extent and the \ntiming of peak haul-out probability is limited")
+  labs(y = "date of peak haul-out", x = "maximum sea ice extent (km^2^ / 1000)",
+       title = "The connection between yearly maximum sea ice extent and the \ntiming of peak haul-out probability is limited") +
+  theme(axis.text.x = element_markdown())
 

### 2021-04-15 analysis/paper/paper.Rmd
EXPAND FIG CAPS/ADD {ggtext} SUPPORT

@@
     mutate(age_sex = forcats::fct_relevel(
-      age_sex,c("ADULT.F","ADULT.M","SUBADULT","YOUNG-OF-THE YEAR"))
+      age_sex,c("ADULT.F","ADULT.M","SUBADULT","YOUNG OF YEAR"))
     )
@@
 
-```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Variability in predicted haul-out probability of ribbon seals across a range of weather conditions thea range of weather conditions encountered in the observed data.
+```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Variability in predicted haul-out probability of ribbon seals across a range of weather conditions the range of weather conditions encountered in the observed data.
 Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
@@
 
-```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Variability in the predicted haul-out probability of spotted seals across a range of weather conditions thea range of weather conditions encountered in the observed data.
+```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Variability in the predicted haul-out probability of spotted seals across a range of weather conditions the range of weather conditions encountered in the observed data.
 Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
@@
 
-```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across a range of weather conditions thea range of weather conditions encountered in the observed data.
+```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across a range of weather conditions the range of weather conditions encountered in the observed data.
 Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
@@
         legend.justification = "right") +
-  labs(y = "date of peak haul-out", x = "maximum sea ice extent (km^2^ / 1000)",
+  labs(y = "date of peak haul-out", x = "maximum sea ice extent (km<sup>2</sup> / 1000)",
        title = "The connection between yearly maximum sea ice extent and the \ntiming of peak haul-out probability is limited") +
-  theme(axis.text.x = element_markdown())
+  theme(axis.title.x = element_markdown())
 

### 2021-04-27 analysis/paper/paper.Rmd
VARIOUS, MINOR EDITS

@@
 library(glue)
+library(scales)
 library(DBI)
@@
 The global climate crisis is causing considerable reduction in Arctic sea ice extent,
-volume, and seasonal presence [@kwok2018erl; @meier2014rg; @wang2017dsrpitsio]. These
+volume, and seasonal presence [@kwok2018erl, @meier2014rg, @wang2017dsrpitsio]. These
 changes have ripple effects on Arctic ecosystems, taxa, and the human communities who live
 in the region [@huntington2020ncc]. Such disruptions are a particular cause of concern for
-ice-associated seals (ringed, bearded, spotted, and ribbon seals) that depend on spring
-and early summer sea ice (March-June) in the Bering and Chukchi seas as a platform for
+ice-associated seals (ringed (*Phoca hispida*), bearded (*Erignathus barbatus*), spotted
+(*Phoca largha*), and ribbon seals (*Histriophoca fasciata*)) that depend on spring and
+early summer sea ice (March-June) in the Bering and Chukchi seas as a platform for
 important functions, such as pupping, nursing, breeding, and molting [@boveng2009a;
@@
 [-@bengtson2005pb] documented a higher propensity for ringed seal basking near solar noon,
-as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded (*Erignathus
-barbatus*), ribbon (*Histriophoca fasciata*), and spotted (*Phoca largha*) seals using
-much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that the proportion of time
-bearded seals spent hauled out progressively increased through spring and summer. And, Ver
-Hoef et al. [-@verhoef2014sm] found haul-out probabilities increased gradually starting in
-March and peaking in May and June for bearded, ribbon, and spotted seals. However, their
-primary focus was to develop a correction factor for estimating seal abundance, and like a
-subsequent analysis by Conn et al. [-@conn2014mee], they did not attempt to separately
-analyze haul-out schedules for different age and sex classes (although initial models with
-sex and age as fixed effects were insignificant at p = 0.05; [@verhoef2014sm]).
+as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded , ribbon , and spotted
+seals using much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that the
+proportion of time bearded seals spent hauled out progressively increased through spring
+and summer. And, Ver Hoef et al. [-@verhoef2014sm] found haul-out probabilities increased
+gradually starting in March and peaking in May and June for bearded, ribbon, and spotted
+seals. However, their primary focus was to develop a correction factor for estimating seal
+abundance, and like a subsequent analysis by Conn et al. [-@conn2014mee], they did not
+attempt to separately analyze haul-out schedules for different age and sex classes
+(although initial models with sex and age as fixed effects were insignificant at p = 0.05
+[@verhoef2014sm]).
 
@@
 of sea ice use from data garnered from bio-loggers. In these models, haul-out
-probabilities are expressed as a function of predictive covariates, such as time of day,
+probabilities were expressed as a function of predictive covariates, such as time of day,
 day-of-year, sex, age class, and environmental conditions (e.g., [@reder2003pb],
@@
 Cameron's [-@bengtson2004pb] study included 5 adult and 2 juvenile crabeater seals, while
-Bengtson et al.'s [-@bengtson2005pb] study was based on 6 telemetered ringed seals (*Phoca
-hispida*) in the Chukchi Sea. These studies are often further limited by logistical
-constraints on tagging fieldwork and the attachment duration or operational life of
-satellite tags.
+Bengtson et al.'s [-@bengtson2005pb] study was based on 6 telemetered ringed seals in the
+Chukchi Sea. These studies were often further limited by logistical constraints on tagging
+fieldwork and the attachment duration or operational life of satellite tags.
 
@@
 ice-associated seal species in the Arctic using aerial surveys (e.g., [@bengtson2005pb],
-[@conn2014mee], and [@verhoef2014sm]). Such studies are conducted over very large areas
-and estimation of absolute abundance requires making inference about numerous nuisance
-processes affecting the observation of seals on ice. This includes availability (only
-seals basking on ice are available to be counted), detection probability (observers or
-automated detection systems may miss some seals on ice), species misclassification, and
-possible disturbance of seals by aircraft [@conn2014mee; @verhoef2014sm].
-
-In this study, we use data from 15 years of bio-logger records in a comprehensive
-investigation of haul-out behavior by 3 ice-associated seals in the Bering and Chukchi
-Seas. We focused our effort on bearded, spotted, and ribbon seals and do not include
-ringed seals in the analysis. Ringed seals are unique in their use of snow lairs in the
-spring months, and they warrant a separate study that includes additional environmental
-covariates, such as the onset of snow melt and snow depth. Our goals are threefold. First,
-we wish to establish baseline estimates for the chronology of haul-out behavior in the
-critical spring season for each species across different age and sex classes. Second, we
-wish to refine estimates of haul-out availability corrections for aerial surveys in order
-to improve estimates of seal abundance. Previously estimated availability correction
-factors (e.g., [@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]) accounted for
-variables such as the time-of-day and day-of-year, but did not investigate meteorological
-variables that have been shown to influence haul-out behavior of walruses [@reder2003pb;
-@udevitz2009cjz]. Third, we explore the annual variability in haul-out timing and possible
-linkage to dramatic changes in the availability and extent of seasonal sea ice between
-2005 and 2020. Our work extends the scope of previous haul-out analyses, includes the
-influence of meteorological variability and possible ramifications of non-representative
-age-sex samples, and investigates the potential impact of a changing icescape on the
-behavior of these species.
+[@conn2014mee], and [@verhoef2014sm]). Such studies were conducted over very large areas
+and estimation of absolute abundance required making inference about numerous nuisance
+processes affecting the observation of seals on ice. This included availability (only
+seals basking on ice were available to be counted), detection probability (observers or
+automated detection systems may have missed some seals on ice), species misclassification,
+and possible disturbance of seals by aircraft [@conn2014mee; @verhoef2014sm].
+
+In this study, we used data from 15 years of bio-logger records in a comprehensive
+investigation of haul-out behavior by 3 species of ice-associated seals in the Bering and
+Chukchi seas. We focused our effort on bearded, spotted, and ribbon seals and did not
+include ringed seals in the analysis. Ringed seals are unique in their use of snow lairs
+in the spring months, and they warrant a separate study that includes additional
+environmental covariates, such as the onset of snow melt and snow depth. Our goals are
+threefold. First, we wish to establish baseline estimates for the chronology of haul-out
+behavior in the critical spring season for each species across different age and sex
+classes. Second, we wish to refine estimates of haul-out availability corrections for
+aerial surveys in order to improve estimates of seal abundance. Previously estimated
+availability correction factors (e.g., [@bengtson2005pb], [@conn2014mee], and
+[@verhoef2014sm]) accounted for variables such as the time-of-day and day-of-year, but did
+not investigate meteorological variables that have been shown to influence haul-out
+behavior of walruses [@reder2003pb; @udevitz2009cjz]. Third, we explored the annual
+variability in haul-out timing and possible linkage to dramatic changes in the
+availability and extent of seasonal sea ice between 2005 and 2020. Our work extends the
+scope of previous haul-out analyses, includes the influence of meteorological variability,
+and investigates the potential impact of a changing icescape on the behavior of these
+species.
 
@@
 
-```{r seal-hours-calendar, fig.height = 4.75, include=TRUE, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species.
+```{r seal-hours-calendar, include=TRUE, fig.height = 4.75, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species.
 The higher density of data from May and June coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out.
@@
 ```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, May, and June.
-Empty areas of the plot represent missing data that was not successfully transmitted from the tag"}
+Empty areas of the plot represent missing data that were not successfully transmitted from the tag"}
 d <- dat.sf %>% 
@@
   theme(axis.ticks=element_blank()) +
-  ggtitle("Observation records are incomplete because not all data is transmitted",
+  ggtitle("Observation records are incomplete because not all data are transmitted",
           subtitle = "percent-dry data for a single ribbon seal across two years") +
@@
 intersected with land were removed from the dataset. Figure \@ref(fig:dataMap) shows the
-spatial distribution of available haul-out behavior data for each species across the study
-area.
+spatial distribution of determined locations and available haul-out behavior data for each
+species across the study area.
 
@@
 
-In addition to sex and age class, we gathered additional variables that might help
-explain variation in haul-out probabilities. These included day-of-year (for seasonal
-effects) and local solar hour (for diurnal effects). Solar hour was calculated using the
-{solaR} package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019]
-using the daily locations. We also compiled several meteorological variables shown to
-affect haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz;
-@perry2017]. In particular, we linked locations from satellite tags to key weather values
-from the North American Regional Reanalysis (NARR) model produced by the National Centers
-for Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
+In addition to sex and age class, we gathered additional variables that might help explain
+variation in haul-out probabilities. These included day-of-year (for seasonal effects) and
+local solar hour (for diurnal effects). Solar hour was calculated using the {solaR}
+package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019] using the
+daily locations. We also compiled several meteorological variables shown to affect
+haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz; @perry2017].
+In particular, we linked locations from satellite tags to key weather values from the
+North American Regional Reanalysis (NARR) model produced by the National Centers for
+Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
 data to produce a long-term picture of weather over North America. Numerous weather
@@
 temperature, wind speed, barometric pressure, precipitation, wind chill (represented by a
-wind $\times$ temperature interaction [@udevitz2009cjz]), and day-of-year and time of day
+_wind:temperature_ interaction [@udevitz2009cjz]), and day-of-year and time of day
 interactions to permit diurnal patterns to change throughout the year. Spotted and ribbon
@@
 
-```{r tblCovariates, include=TRUE, tab.id = "tblCovariates", tab.cap = "Explanatory covariates used in analyses of binary haul-out records for ice-associated seals."}
+```{r tblCovariates, include=TRUE, tab.id = "tblCovariates", tab.cap = "Explanatory covariates used in analyses of binary haul-out records for ice-associated seals.
+Note that we also considered select interactions (see article text) between these primary covariates.
+For instance, wind chill was represented by the interaction _temperature:wind_."}
 covariate_tbl <- tibble::tribble(
@@
 responses, an assumption that is clearly violated if hourly responses are modeled. Any
-analysis that ignores temporal dependence in responses will thus have overstated
-precision.
+analysis that ignores temporal dependence in responses will thus have overstated precision
+[@betts2006].
 
@@
 while preserving the inherent circularity needed for time-of-day effects (i.e., hour 0
-should be equal to hour 24). It also represents hour-of-day with 6 variables, which is a
-considerable reduction when compared to a 24 variable model, especially when fitting
+should be equal to hour 24). It also represents hour-of-day with 6 parameters, which is a
+considerable reduction when compared to a 24 parameter variable, especially when fitting
 models with interactions between hour-of-day and other variables (e.g., age-sex class,
@@
 $$
-H_t=\sum_{i=1}^{3}{\alpha_{i1}cos{\left(\frac{\pi t}{2\left(i+1\right)}\right)}}+\alpha_{i2}sin{\left(\frac{\pi t}{2\left(i+1\right)}\right)}1
+H_t=\alpha_1 \cos(\frac{\pi t}{4}) + \alpha_2 \sin(\frac{\pi t}{4}) + \alpha_3 \cos(\frac{\pi t}{6}) + \alpha_4 \sin(\frac{\pi t}{6}) + \alpha_5 \cos(\frac{\pi t}{12}) + \alpha_6 \sin(\frac{\pi t}{12})
 $$ 
-
-where $H_t$ gives the effect for solar hour $t$ and $α_{ij}$ for $i$ = 1, 2, or 3 and $j$
-= 1 or 2 are estimated parameters (regression coefficients).
+where $H_t$ gives the effect for solar hour $t$ and $α_{i}$ are estimated parameters
+(regression coefficients).
 
@@
 models with year by day-of-year interactions. However, in this case we only included
-year$\times$day and year$\times$day2, omitting the main effects of year as well as
-year$\times$day3 interactions because models with the latter effects were numerically
+_year:day_ and _year:day2_, omitting the main effects of year as well as
+_year:day3_ interactions because models with the latter effects were numerically
 unstable. However, the modeled interactions were sufficient to allow shifts in haul-out
@@
 [@teneyck2017]. After fitting GLMPM models, we instead used "type III" $F$-tests to
-calculate $p$-values [@verhoef2010cs] to evaluate model performance and important terms. We
-also produced predictions of haul-out behavior as a function of influential predictors
+calculate $p$-values [@verhoef2010cs] to evaluate model performance and important terms.
+We also produced predictions of haul-out behavior as a function of influential predictors
 (e.g. solar hour, day-of-year, age-sex). Weather covariates for these predictions were
 based on daily or hourly smoothed weather covariate values across the study region. Such
-predictions were then used to develop haul-out probability surfaces, explore marginal
+predictions were then used to develop haul-out probability surfaces, explore conditional
 effects of weather covariates, and determine annual peaks in haul-out activity. The timing
@@
 temperature, and wind substantially altered haul-out behavior of all three species, with
-$F$ tests producing $p$-values less than 0.05 for variables embodying these effects
-and/or their interactions. Haul-out probabilities typically increased throughout March and
-April, reaching a peak in May and early June before declining again. Diurnal patterns were
+$F$ tests producing $p$-values less than 0.05 for variables embodying these effects and/or
+their interactions. Haul-out probabilities typically increased throughout March and April,
+reaching a peak in May and early June before declining again. Diurnal patterns were
 present, with maximum haul-out behavior centered around local solar noon.
@@
   mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
-         month = lubridate::month(date,label=TRUE),
+         month = lubridate::month(date,label = TRUE),
          day = lubridate::day(date)) %>%
@@
 
-p <- ggplot(plot_df,aes(day,solar_hour,fill=ho_prob))+
-  geom_tile(color= "white",size = 0) +
+p <- ggplot(plot_df, aes(day,solar_hour,fill = ho_prob)) +
+  geom_tile(color = "white",size = 0) +
   scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
@@
 p <- p + theme(legend.position = "bottom") +
-  theme(strip.background = element_rect(colour="white")) +
-  theme(axis.ticks=element_blank()) +
+  theme(strip.background = element_rect(colour = "white")) +
+  theme(axis.ticks = element_blank()) +
   xlab("day of month") + ylab("local solar hour") +
@@
 ) %>% 
-  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
+  mutate(Prob.F = scales::pvalue(Prob.F),
          Effect = janitor::make_clean_names(Effect)) %>% 
@@
 ) %>% 
-  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
+  mutate(Prob.F = scales::pvalue(Prob.F),
          Effect = janitor::make_clean_names(Effect)) %>%  
@@
 less likely to haul out at higher winds and more likely to be on the ice when temperatures
-were relatively warm. Wind chill (temperature $\times$ wind) and precipitation were not as
+were relatively warm. Wind chill (_temperature:wind_) and precipitation were not as
 influential. Differences in the magnitude of response between the age-sex classes is
@@
 ) %>% 
-  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
+  mutate(# Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
+         Prob.F = scales::pvalue(Prob.F),
          Effect = janitor::make_clean_names(Effect)) %>% 
@@
 ) %>% 
-  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
+  mutate(# Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
+         Prob.F = scales::pvalue(Prob.F),
          Effect = janitor::make_clean_names(Effect)) %>%  
@@
 + age-sex group was present, there are still some years where the pattern shown is
-informed by a small number of individuals that may not represent population level processes.
+informed by a small number of individuals that may not represent population level
+processes.
 
@@
 deployments are likely capturing population-level behavioral patterns. Seals
-preferentially haul out on ice shortly after solar noon, which coincides with the daily
-period of maximum solar radiation. The association of daily peak haul-out behavior with
-mid to later solar hours is commonly noted in pinnipeds and allows seals to maximize
-absorption of solar radiation. Interestingly, bearded seals appear to have two peaks in
-haul-out activity across the day, one shortly after solar noon, and one centered near
-solar midnight. A similar bi-modal pattern has been seen in ringed seals and suggests that
-bearded and ringed seals may be operating under different constraints than ribbon and
-spotted seals. Bearded and ringed seals are distributed across higher latitudes and the
-extended daylight hours may allow more flexibility in optimizing resting periods with
-foraging. Notably, bearded and ringed seals are subjected to much higher levels of
-predation from polar bears and human harvest.
-
-::: {custom-style="comment"}
-are there references for timing of polar bear activity/kills or hunter activity that might
-lend some additional support for predation as a possible explanation for the bi-modal
-haul-out pattern? Any data on timing of hunting activity by Alaska communities?
-:::
-
-Unlike previous analyses of seal haul-out data (e.g. Ver Hoef et al. 2009, Conn et al.
-2014), we also investigated the influence of sex-age class on haul-out probabilities for
-all species except bearded seals because of low sample size. While both ribbon and spotted
-seals exhibited a unimodal diel haul-out pattern generally centered around local solar
-noon, there were key differences across species, age, and sex that match expectations
-given what we know of their ecological behavior. Spotted seals are known to form triads
-during the breeding season, where a female and dependent pup are accompanied on the ice by
-a suitor male. The male waits for the female to wean the pup and enter estrus and fends
-off any other males that may be interested. Triad formation results in both males and
-females spending a large portion of the day hauled out on ice and a protracted spring
-haul-out season for both sexes. Since breeding occurs in the water, females may also be
-less inclined to venture out for foraging trips while still nursing the pup. We see this
-reflected in the predicted haul-out behavior, with both males and females exhibiting a
-broad distribution of time out of the water throughout the solar day and the season.
-Ribbon seals are not known to form triads and our model predicts a progression of
-increased haul-out behavior with females starting earlier in the season than males.
-Notably, female ribbon seals spend a large portion of the day in the water during the
-pupping period and aligns with the hypothesis that ribbon seal females continue foraging
-while nursing. In the case of both ribbon and spotted seals, subadults are the first to
-begin meaningful haul-out behavior and reflect a typical phocid pattern where subadults
-molt first as they do not have any reproductive constraints.
-
-::: {custom-style="comment"}
-this previous paragraph might benefit from some references e.g. triads in spotted seals,
-ribbon seal females foraging while nursing, etc paper on dive/foraging behavior of
-lactating harp seals: <https://doi.org/10.1006/anbe.1993.1312>
-:::
+preferentially haul out on ice shortly after solar noon, which allows seals to maximize
+absorption of solar radiation [@feltz1966]. Interestingly, bearded seals appear to have
+two peaks in haul-out activity across the day, one shortly after solar noon, and one
+centered near solar midnight. A similar bi-modal pattern has been seen in ringed seals
+[@vonduyke2020] and suggests that bearded and ringed seals may be operating under
+different constraints than ribbon and spotted seals. Bearded and ringed seals are
+distributed across higher latitudes and the extended daylight hours may allow more
+flexibility in optimizing resting periods with foraging. Notably, bearded and ringed seals
+are subjected to much higher levels of predation from polar bears and human harvest.
+
+Unlike previous analyses of seal haul-out data (e.g. [@verhoef2010cs], [@conn2014mee]), we
+also investigated the influence of sex-age class on haul-out probabilities for all species
+except bearded seals because of low sample size. While both ribbon and spotted seals
+exhibited a unimodal diel haul-out pattern generally centered around local solar noon,
+there were key differences across species, age, and sex that match expectations given what
+we know of their ecological behavior. Spotted seals are known to form triads during the
+breeding season [@burns2009, @mcclintock2015] where a female and dependent pup are
+accompanied on the ice by a suitor male. The male waits for the female to wean the pup and
+enter estrus and fends off any other males that may be interested. Triad formation results
+in both males and females spending a large portion of the day hauled out on ice and a
+protracted spring haul-out season for both sexes. Since breeding occurs in the water,
+females may also be less inclined to venture out for foraging trips while still nursing
+the pup. We see this reflected in the predicted haul-out behavior, with both males and
+females exhibiting a broad distribution of time out of the water throughout the solar day
+and the season. Ribbon seals are not known to form triads and our model predicts a
+progression of increased haul-out behavior with females starting earlier in the season
+than males. Notably, female ribbon seals spend a large portion of the day in the water
+during the pupping period and aligns with the hypothesis that ribbon seal females continue
+foraging while nursing. In the case of both ribbon and spotted seals, subadults are the
+first to begin meaningful haul-out behavior and reflect a typical phocid pattern where
+subadults molt first as they do not have any reproductive constraints.
 
@@
 including wind speed, temperature, barometric pressure, precipitation, and wind chill.
-These have been investigated for walrus (e.g. Udevitz et al. [@udevitz2009cjz]), but to
-our knowledge have never been investigated for ice-associated seals. Ribbon seals seemed
-to be the most influenced by weather, with wind, temperature, barometric pressure, and
+These have been investigated for walrus (e.g. Udevitz et al. [@udevitz2009cjz]) and a few
+select studies of ice-associated seals [@hamilton2018, @perry2017]. Ribbon seals seemed to
+be the most influenced by weather, with wind, temperature, barometric pressure, and
 precipitation all being important components of the model. Spotted seals were most
@@
 Previous attempts to estimate the abundance of phocid seals from aerial survey data in the
-Bering and Chukchi Seas (e.g. Bengtson et al. 2005, Conn et al. 2014, Ver Hoef et al.
-2014) have used estimated haul-out probabilities to correct for the proportion of animals
-that are in the water and thus unavailable to be counted during aerial surveys. Although
-several of these studies allowed haul-out probabilities to vary by day-of-year and
-time-of-day, previous correction factors have not accounted for variability between years,
-different weather conditions, and the age-sex class of the sample. In this paper, we have
-shown that there can be considerable differences in the number of seals hauled out on ice
-based on these factors. We recommend that future abundance analyses employ correction
-factors that account for these factors. For instance, it is relatively straightforward to
-obtain weather reanalysis products for times and locations that are surveyed and to
-construct a relevant correction factor based on predictions of GMPMs. The most challenging
-element in developing availability correction factors is with annual variability. It can
-be difficult to get a sufficient sample size to estimate year-specific correction factors,
-particularly because research teams would likely need to tag seals and conduct aerial
-surveys concurrently which requires considerably more personnel and money. One possible
-suggestion is to estimate a "shift" parameter within models for aerial survey counts that
-allow the peak of haul-out distributions to be adjusted earlier or later in the year based
-on the frequency of counts observed over time. Regardless, researchers should anticipate
-there being some unmodeled heterogeneity in availability probability present in abundance
+Bering and Chukchi Seas (e.g. [@bengtson2005pb], [@conn2014mee], [@verhoef2014sm]) have
+used estimated haul-out probabilities to correct for the proportion of animals that are in
+the water and thus unavailable to be counted during aerial surveys. Although several of
+these studies allowed haul-out probabilities to vary by day-of-year and time-of-day,
+previous correction factors have not accounted for variability between years, different
+weather conditions, and the age-sex class of the sample. In this paper, we have shown that
+there can be considerable differences in the number of seals hauled out on ice based on
+these factors. We recommend that future abundance analyses employ correction factors that
+account for these factors. For instance, it is relatively straightforward to obtain
+weather reanalysis products for times and locations that are surveyed and to construct a
+relevant correction factor based on predictions of GLMPMs. The most challenging element in
+developing availability correction factors is with annual variability. It can be difficult
+to get a sufficient sample size to estimate year-specific correction factors, particularly
+because research teams would likely need to tag seals and conduct aerial surveys
+concurrently which requires considerably more personnel and money. One possible suggestion
+is to estimate a "shift" parameter within models for aerial survey counts that allow the
+peak of haul-out distributions to be adjusted earlier or later in the year based on the
+frequency of counts observed over time. Regardless, researchers should anticipate there
+being some unmodeled heterogeneity in availability probability present in abundance
 estimates obtained from aerial surveys. This may make trend detection difficult, as one
@@
 those previously reported for these species, especially for bearded seals. For instance,
-Ver Hoef et al. (2014) and Conn et al. (2015) used haul-out correction factors with
-maximums of 0.66 for bearded seals, 0.62 for ribbon seals, and 0.54 for spotted seals,
-where maximums corresponded to times near solar noon in mid-late May. Applying models that
-ignore age, sex, and year effects, these probabilities were 0.38, 0.72, and 0.60,
-respectively, under the current analysis framework. Our current estimates reflect
-increased sample sizes in terms of number of animals, but also improvements to the way
-data were prepared prior to analysis.
+Ver Hoef et al. [-@verhoef2014sm)] and Conn et al. [-@conn2014mee] used haul-out
+correction factors with maximums of 0.66 for bearded seals, 0.62 for ribbon seals, and
+0.54 for spotted seals, where maximums corresponded to times near solar noon in mid-late
+May. Applying models that ignore age, sex, and year effects, these probabilities were
+0.38, 0.72, and 0.60, respectively, under the current analysis framework. Our current
+estimates reflect increased sample sizes in terms of number of animals, but also
+improvements to the way data were prepared prior to analysis.
 
@@
 top of the sea ice, which they frequently inhabit before snow melt causes their lairs to
-collapse (Kelly 1988). Thus, the wet-dry sensor on a satellite tag could indicate that an
-animal is hauled out, but it could still be unavailable to be detected because it is in a
-lair. We hope to address availability of ringed seals using data from satellite tags,
-replicate survey tracks, and auxiliary information about snow depth and timing of melt in
-a future study.
+collapse. Thus, the wet-dry sensor on a satellite tag could indicate that an animal is
+hauled out, but it could still be unavailable to be detected because it is in a lair. We
+hope to address availability of ringed seals using data from satellite tags, replicate
+survey tracks, and auxiliary information about snow depth and timing of melt in a future
+study.
 
@@
 
-::: {custom-style="comment"}
-insert broad statement and summary regarding data contribution from AFSC, ADFG, NSB (I
-still need to lookup details of which org contributed which data). Authors, please add
-your contribution statement below
-:::
-
-**Josh M. London:** led field work, conceptualization, methodology, formal analysis, code
-development, writing of original draft, visualization, review, revisions, and
-corresponding author
-
-**Paul B. Conn:** conceptualization, methodology, formal analysis, code development,
-writing of original draft, review, and revisions
-
-**Stacie K. Hardy:** field work, methodology, data curation, validation, review, and
-revisions
-
-**Erin L. Richmond:** data curation, conceptualization, methodology, review, and revisions
-
-**Jay M. Ver Hoef:** field work, conceptualization, methodology, code development, review,
-and revisions
-
-**Irina S. Trukhanova:** conceptualization, formal analysis, code development, review, and
-revisions
-
-**Michael F. Cameron:** led field work, project administration, methodology, review, and
-revisions
-
-**Peter L. Boveng:** led field work, supervision, project administration,
-conceptualization, methodology, review, and revisions
+**Josh M. London**:  investigation, conceptualization, methodology, formal analysis, validation, software, writing: original draft, writing: review and editing, visualization, and data curation
+**Paul B. Conn**:  conceptualization, methodology, formal analysis, software, validation, writing: original draft, writing: review and editing
+**Stacie K. Hardy**: investigation, data curation, methodology, validation, writing: review and editing
+**Erin L. Richmond**: data curation, investigation, methodology, validation, writing: review and editing
+**Jay M. Ver Hoef**: conceptualization, methodology, software, writing: review and editing
+**Irina S. Trukhanova**: conceptualization, writing: review and editing
+**Michael F. Cameron**: investigation, project administration, writing: review and editing
+**Peter L. Boveng**: investigation, conceptualization, supervision, project administration, writing: review and editing
 

### 2021-04-28 analysis/paper/paper.Rmd
KNIT BEFORE UPLOADING RPTS DRAFT

@@
 ```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, May, and June.
-Empty areas of the plot represent missing data that were not successfully transmitted from the tag"}
+Empty areas of the plot represent missing data that were not successfully transmitted from the tag."}
 d <- dat.sf %>% 
@@
 temperature, wind speed, barometric pressure, precipitation, wind chill (represented by a
-_wind:temperature_ interaction [@udevitz2009cjz]), and day-of-year and time of day
+*wind:temperature* interaction [@udevitz2009cjz]), and day-of-year and time of day
 interactions to permit diurnal patterns to change throughout the year. Spotted and ribbon
@@
 when modeling the haul-out behavior of seal species with such a close association to sea
-ice; seals haul out in the presence of sea ice and we could assess the local concentration
-of sea ice during these events. However, seals also spend time in the water away from sea
-ice. Thus, the challenge is how best to associate that seal's at-sea behavior with a sea
-ice concentration. Would it be the concentration at the last known haul-out event? The
-next known haul-out event? Or, the closest sea ice concentration available? Without a
-meaningful means to associate a sea ice covariate with the seal throughout the duration of
-the observation record, such parameters are not viable for inclusion in this model.
+ice; seals haul out in the presence of sea ice, and we could assess the local
+concentration of sea ice during these events (see [@olnes2020]). However, a major focus of
+this study is to develop models applicable for aerial survey correction factors and using
+sea ice as a covariate would almost certainly bias haul-out predictions towards those
+seals that are on or near ice and therefore more likely to haul out.  Since aerial surveys
+can only detect seals on ice, abundance estimates would be missing a correction for those
+seals that are away from ice (e.g., on foraging trips). Additionally, many of the
+bio-logger deployments used in this study were single attachments to the rear flipper of
+the seal and do not provide at-sea locations, limiting our ability to fully evaluate
+habitat preferences related to sea-ice. It should be noted that the aerial survey counts
+are modeled with sea-ice and other habitat parameters. So, any final abundance estimates
+that combine models of survey counts and our model for proportion of seals hauled out on
+ice do account for the strong association between these species and sea-ice in the Bering
+and Chukchi seas. Lastly, our study was limited to the spring season when seal haul-out
+tendencies are strongly influenced by pupping, nursing, breeding, and molt and these
+drivers are likely more influential than specific sea-ice concentration. Crawford et al
+[-@crawford2019] compared haul-out probability models for bearded seals and found those
+that only included season (and not sea-ice concentration) where the most parsimonious. 
+For these reasons, we have elected not to use sea ice concentration as a predictor for
+haul-out probability; studies to evaluate habitat selection in relation to sea-ice
+parameters and other environmental covariates are in progress.
 
@@
 H_t=\alpha_1 \cos(\frac{\pi t}{4}) + \alpha_2 \sin(\frac{\pi t}{4}) + \alpha_3 \cos(\frac{\pi t}{6}) + \alpha_4 \sin(\frac{\pi t}{6}) + \alpha_5 \cos(\frac{\pi t}{12}) + \alpha_6 \sin(\frac{\pi t}{12})
-$$ 
-where $H_t$ gives the effect for solar hour $t$ and $α_{i}$ are estimated parameters
+$$ where $H_t$ gives the effect for solar hour $t$ and $α_{i}$ are estimated parameters
 (regression coefficients).
@@
 models with year by day-of-year interactions. However, in this case we only included
-_year:day_ and _year:day2_, omitting the main effects of year as well as
-_year:day3_ interactions because models with the latter effects were numerically
-unstable. However, the modeled interactions were sufficient to allow shifts in haul-out
-distribution, as one can show mathematically that a simple horizontal shift in timing of
-haul-out distributions does not affect the main effects or cubic terms in a polynomial
-regression model.
+*year:day* and *year:day2*, omitting the main effects of year as well as *year:day3*
+interactions because models with the latter effects were numerically unstable. However,
+the modeled interactions were sufficient to allow shifts in haul-out distribution, as one
+can show mathematically that a simple horizontal shift in timing of haul-out distributions
+does not affect the main effects or cubic terms in a polynomial regression model.
 
@@
                          rownames(spotted_type3))
+
 bearded_type3 <- tibble(
@@
 less likely to haul out at higher winds and more likely to be on the ice when temperatures
-were relatively warm. Wind chill (_temperature:wind_) and precipitation were not as
+were relatively warm. Wind chill (*temperature:wind*) and precipitation were not as
 influential. Differences in the magnitude of response between the age-sex classes is
@@
 were more likely to haul-out when daily temperatures were warmer, winds speeds were less,
-barometric pressure was high, and lower rates of precipitation. These weather conditions
+barometric pressure was high, and rates of precipitation were. These weather conditions
 are likely most beneficial for energetic considerations and may also provide the best
@@
 top of the sea ice, which they frequently inhabit before snow melt causes their lairs to
-collapse. Thus, the wet-dry sensor on a satellite tag could indicate that an animal is
-hauled out, but it could still be unavailable to be detected because it is in a lair. We
-hope to address availability of ringed seals using data from satellite tags, replicate
-survey tracks, and auxiliary information about snow depth and timing of melt in a future
-study.
+collapse [@kelly1990]. Thus, the wet-dry sensor on a satellite tag could indicate that an
+animal is hauled out, but it could still be unavailable to be detected because it is in a
+lair. We hope to address availability of ringed seals using data from satellite tags,
+replicate survey tracks, and auxiliary information about snow depth and timing of melt in
+a future study.
 
@@
 
-**Josh M. London**:  investigation, conceptualization, methodology, formal analysis, validation, software, writing: original draft, writing: review and editing, visualization, and data curation
-**Paul B. Conn**:  conceptualization, methodology, formal analysis, software, validation, writing: original draft, writing: review and editing
-**Stacie K. Hardy**: investigation, data curation, methodology, validation, writing: review and editing
-**Erin L. Richmond**: data curation, investigation, methodology, validation, writing: review and editing
-**Jay M. Ver Hoef**: conceptualization, methodology, software, writing: review and editing
-**Irina S. Trukhanova**: conceptualization, writing: review and editing
-**Michael F. Cameron**: investigation, project administration, writing: review and editing
-**Peter L. Boveng**: investigation, conceptualization, supervision, project administration, writing: review and editing
+**Josh M. London**: investigation, conceptualization, methodology, formal analysis,
+validation, software, writing: original draft, writing: review and editing, visualization,
+and data curation **Paul B. Conn**: conceptualization, methodology, formal analysis,
+software, validation, writing: original draft, writing: review and editing **Stacie K.
+Hardy**: investigation, data curation, methodology, validation, writing: review and
+editing **Erin L. Richmond**: data curation, investigation, methodology, validation,
+writing: review and editing **Jay M. Ver Hoef**: conceptualization, methodology, software,
+writing: review and editing **Irina S. Trukhanova**: conceptualization, writing: review
+and editing **Michael F. Cameron**: investigation, project administration, writing: review
+and editing **Peter L. Boveng**: investigation, conceptualization, supervision, project
+administration, writing: review and editing
 

### 2021-05-07 data_raw/adfg/adfg_tidydata.Rmd
ADDED ADDITIONAL ADFG DATA SOURCES
up to this point, some ADFG data sources were not included in the analysis. This is the initial attempt to include those data.

---
title: "Import and Tidy ADFG Supplied Location and Haul-out Data"
description: | 
  This article describes the import and tidying of location and haul-out data provided to
  MML by ADFG for use in manuscript analyzing haul-out behavior of bearded, ribbon, and
  spotted seals in the Bering and Chukchi seas
date: May 6, 2021
author:
  - name: "Josh M. London"
    url: https://github.com/jmlondon
    affiliation: Alaska Fisheries Science Center
    affiliation_url: https://www.fisheries.noaa.gov/about/alaska-fisheries-science-center
    orcid_id: 0000-0002-3647-5046
  - name: "Justin Crawford"
    url: http://www.adfg.alaska.gov/index.cfm%3Fadfg%3Dmarinemammalprogram.icesealresearch
    affiliation: Alaska Department of Fish & Game
    affiliation_url: http://www.adfg.alaska.gov
output: distill::distill_article
---

```{r packages, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
```

## Introduction

The focus of this document is not to analyze the data, but instead to document
the processing of the 'raw' data (as provided) into a usable data structure for
downstream merging with other datasets and analysis procedures. We will focus
on reading in the csv files, tidying any data structures or values that need it,
and then merging into a single usable tibble.

## Read the CSV Files

There are three files provided by ADF&G

1. ADFG_SealDeploymentTable3.csv
2. ADFG_SealLocationData3.csv
3. ADFG_SealTimelineData3.csv

### ADFG Seal Deployments

Let's read this file in and take a look at the structure

```{r}
adfg_deployments <- readr::read_csv('ADFG_SealDeploymentTable3.csv')
```

Let's look at some of the data and see if we might want to adjust some of the
automatically determined column specifications

```{r}
glimpse(adfg_deployments)
```

Looks like the only thing we would want to change is the `DateTagged` column.
This is clearly a date-time field, so we'll create a custom `readr::cols`
function to specify our column types. Note, we will then set the proper time
zone to *America/Anchorage* and then convert to a `Date` object.

```{r}
adfg_deploy_cols <- readr::cols(
  DeployIDs = readr::col_character(),
  MapID = readr::col_character(),
  DateTagged = readr::col_datetime(format = "%m/%d/%Y %H:%M"),
  Species = readr::col_character(),
  Age = readr::col_character(),
  Sex = readr::col_character(),
  PrimaryTagType = readr::col_character(),
  FlipperTagType = readr::col_character()
)

adfg_deployments <- readr::read_csv('ADFG_SealDeploymentTable3.csv',
                                    col_types = adfg_deploy_cols) %>% 
  dplyr::mutate(DateTagged = lubridate::force_tz(DateTagged, 
                                                 'America/Anchorage')) %>% 
  dplyr::mutate(DateTagged = lubridate::as_date(DateTagged, 
                                                tz = 'America/Anchorage'))

glimpse(adfg_deployments)
```

The first author (JML) likes to stick with snake case for column names and variables.
So, we'll do some additional editing of the data to rename columns. While at it,
we'll also convert some of the columns to lower case to limit typo issues down the
road.

Another issue we need to tidy up is the way satellite tag type and attachment
position is recorded in the dataset. These data are stored within two columns:
`primary_tag_type` and `flipper_tag_type`. For a proper tidy structure, we need
to get these data into two columns: `tag_position` and `tag_type`. To do this,
we'll use the `tidyr::gather` function.

```{r}
adfg_deployments <- adfg_deployments %>% 
  dplyr::rename(deployid = DeployIDs,
                mapid = MapID,
                date_tagged = DateTagged,
                primary_tag_type = PrimaryTagType,
                flipper_tag_type = FlipperTagType
                ) %>% 
  dplyr::rename_all(tolower) %>% 
  dplyr::mutate(species = tolower(species),
                age = tolower(age),
                sex = tolower(sex)) %>% 
  tidyr::gather(tag_position, tag_type, c(primary_tag_type,flipper_tag_type)) %>% 
  dplyr::mutate(tag_position = ifelse(tag_position == "primary_tag_type",
                   "primary","flipper")) %>% 
  dplyr::filter(!is.na(tag_type)) %>% 
  dplyr::arrange(date_tagged,deployid)

glimpse(adfg_deployments)
```

### ADFG Seal Locations

Next, we'll go through a similar process with the locations data.

```{r}
adfg_locations <- readr::read_csv('ADFG_SealLocationData3.csv')
```

```{r}
glimpse(adfg_locations)
```

As with the deployment data, the `Date` column needs to be converted to a proper
date-time type. Additionally, we can see that `Ptt` was brought in as a `double`
when, in fact, this is an `integer`. There is no way to do the `Date` or `Ptt` type conversion via the `readr::col_*` functions, so we will do a mutate after 
reading the data in.

```{r}
adfg_locs_cols <- readr::cols(
  DeployID = readr::col_character(),
  LocationQuality = readr::col_character(),
  Date = readr::col_double(),
  Latitude = readr::col_double(),
  Longitude = readr::col_double(),
  Ptt = readr::col_double(),
  TagType = readr::col_character(),
  MonthNumb = readr::col_character()
)

adfg_locations <- readr::read_csv('ADFG_SealLocationData3.csv',
                                  col_types = adfg_locs_cols) %>% 
  dplyr::mutate(Date = as.Date(Date, origin = "1899-12-30", tz ="UTC") %>% 
                  lubridate::as_datetime(),
                Ptt = as.integer(Ptt))

glimpse(adfg_locations)
```

As before, we'll rename a few columns to snake_case and then all to lower case.

```{r}
adfg_locations <- adfg_locations %>% 
  dplyr::rename(deployid = DeployID,
                location_quality = LocationQuality,
                date_time = Date,
                tag_type = TagType) %>% 
  dplyr::filter(location_quality != "Z") %>% 
  dplyr::select(-MonthNumb) %>% 
  dplyr::rename_all(tolower)

glimpse(adfg_locations)
```

At this point, we have our locations data assembled. Let's take a moment and plot
all the data on a map

```{r}
library(sf)
library(rnaturalearth)
library(ggspatial)

world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(3571)

locs_sf <- adfg_locations %>% 
  dplyr::left_join(adfg_deployments, by = c("deployid","tag_type")) %>% 
  sf::st_as_sf(coords = c("longitude","latitude")) %>% 
  sf::st_set_crs(4326) %>% sf::st_transform(3571)

ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(
    data = locs_sf, size = 0.125
  ) +
  facet_grid(~ species) +
  theme(legend.position = "bottom") +
  ggtitle("ADFG Locations (Argos)",
            subtitle = "all locations from bearded, ringed, and spotted seals")

```

### ADFG Seal Timelines

Last, we'll go through a similar process with the timeline data.

```{r}
adfg_timelines <- readr::read_csv('ADFG_SealTimelineData3.csv')
```

For this dataset, we'll need to convert the `GMTDate` to a proper date-time
format. And, as before the same with the `Ptt` as an integer.

```{r}
adfg_timeline_cols <- cols(
  .default = readr::col_double(),
  DeployIDs = readr::col_character(),
  TagType = readr::col_character(),
  HistType = readr::col_character(),
  MonthNumb = readr::col_character(),
  LocationQuality = readr::col_character(),
  Sum = readr::col_character()
)

adfg_timelines <- readr::read_csv('ADFG_SealTimelineData3.csv',
                                  col_types = adfg_timeline_cols) %>% 
  dplyr::mutate(GMTDate = as.Date(GMTDate, origin = "1899-12-30", 
                                     tz = "UTC") %>% 
                  lubridate::as_datetime(),
                Ptt = as.integer(Ptt),
                TagType = toupper(TagType)) %>% 
  dplyr::rename(deployid = DeployIDs,
                tag_type = TagType,
                hist_type = HistType,
                date_time = GMTDate) %>% 
  dplyr::rename_all(tolower) %>% 
  dplyr::select(-c(monthnumb,latitude,longitude,locationquality,sum,hist_type))

```

We need to do some additional processing of the timeline data to work with the
odd non-tidy data structure. In this case, each `bin*` column represents an hour
of each day.

```{r}
bins <- list(bin = paste("bin", 1:24, sep=""), hour=0:23) %>% as_tibble()

adfg_timelines <- adfg_timelines %>% 
        tidyr::gather(bin, percent_dry, starts_with('bin')) %>%
        dplyr::left_join(bins, by = "bin") %>%
        dplyr::mutate(date_time = date_time + lubridate::hours(hour),
                      data_date = lubridate::floor_date(date_time,"day")) %>%
        dplyr::select(c(deployid,ptt,tag_type,date_time,data_date,percent_dry)) %>% 
        dplyr::arrange(deployid,date_time)
```

```{r}
glimpse(adfg_timelines)
```

```{r}
usethis::use_data(adfg_deployments, overwrite = TRUE)
usethis::use_data(adfg_timelines, overwrite = TRUE)
usethis::use_data(adfg_locations, overwrite = TRUE)
```

### 2021-05-07 R/get_locs_sf.R

@@ -17,8 +17,8 @@ get_locs_sf <- function() {
   con <- dbConnect(
     odbc(),
     dsn = "PostgreSQL pep",
-    uid = get_kc_account("pgpep_londonj"),
-    pwd = decrypt_kc_pw("pgpep_londonj")
+    uid = keyringr::get_kc_account("pgpep_londonj"),
+    pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
   )
 
   deployments_db <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>%
@@ -26,13 +26,16 @@ get_locs_sf <- function() {
 
   spenos_db <- tbl(con, in_schema("capture","for_telem")) %>% collect()
 
-  locs_qry <- "SELECT deployid,
-                locs_dt,
-                type,
-                quality,
-                error_radius,
-                geom
-              FROM telem.geo_wc_locs_qa"
+  locs_qry <- "SELECT deployid, type, error_radius, locs_dt, quality, geom
+              FROM telem.geo_wc_locs_qa
+              UNION ALL
+              SELECT deployid, NULL AS type, NULL AS error_radius,
+                     loc_dt as locs_dt, argos_lc as quality, loc_geom as geom
+              FROM telem.adfg_locs
+              UNION ALL
+              SELECT deployid::VARCHAR, NULL AS type, NULL AS error_radius,
+                     loc_dt as locs_dt, argos_lc as quality, loc_geom as geom
+              FROM telem.nsb_locs;"
 
   locs_sf <- read_sf(con, query = locs_qry) %>%
     left_join(deployments_db, by = 'deployid') %>%

### 2021-05-17 data_raw/adfg/adfg_tidydata.Rmd
PROGRESS WITH ADFG DATA

@@
 adfg_deployments <- adfg_deployments %>% 
-  dplyr::rename(deployid = DeployIDs,
+  dplyr::rename(speno = DeployIDs,
                 mapid = MapID,
@@
   dplyr::mutate(tag_position = ifelse(tag_position == "primary_tag_type",
-                   "primary","flipper")) %>% 
+                   "primary","flipper"),
+                deployid = paste(speno,tag_type, sep = "_")) %>% 
+  dplyr::relocate(deployid, .after = mapid) %>% 
   dplyr::filter(!is.na(tag_type)) %>% 
@@
 adfg_locations <- adfg_locations %>% 
-  dplyr::rename(deployid = DeployID,
+  dplyr::rename(speno = DeployID,
                 location_quality = LocationQuality,
@@
   dplyr::select(-MonthNumb) %>% 
-  dplyr::rename_all(tolower)
+  dplyr::rename_all(tolower) %>% 
+  dplyr::mutate(deployid = paste(speno,tag_type, sep = "_")) %>% 
+  dplyr::relocate(deployid)
 
@@
 locs_sf <- adfg_locations %>% 
-  dplyr::left_join(adfg_deployments, by = c("deployid","tag_type")) %>% 
+  dplyr::left_join(adfg_deployments, by = c("deployid","speno","tag_type")) %>% 
   sf::st_as_sf(coords = c("longitude","latitude")) %>% 

### 2021-05-17 R/get_locs_sf.R

@@ -26,27 +26,29 @@ get_locs_sf <- function() {
 
   spenos_db <- tbl(con, in_schema("capture","for_telem")) %>% collect()
 
-  locs_qry <- "SELECT deployid, type, error_radius, locs_dt, quality, geom
-              FROM telem.geo_wc_locs_qa
-              UNION ALL
-              SELECT deployid, NULL AS type, NULL AS error_radius,
-                     loc_dt as locs_dt, argos_lc as quality, loc_geom as geom
-              FROM telem.adfg_locs
-              UNION ALL
-              SELECT deployid::VARCHAR, NULL AS type, NULL AS error_radius,
-                     loc_dt as locs_dt, argos_lc as quality, loc_geom as geom
-              FROM telem.nsb_locs;"
+  locs_qry <- "SELECT deployid, ptt, type, error_radius, locs_dt, quality, geom
+              FROM telem.geo_wc_locs_qa;"
 
   locs_sf <- read_sf(con, query = locs_qry) %>%
     left_join(deployments_db, by = 'deployid') %>%
     left_join(spenos_db, by = 'speno') %>%
     filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
-    filter(month(locs_dt) %in% c(3,4,5,6,7)) %>%
+    filter(lubridate::month(locs_dt) %in% c(3,4,5,6,7)) %>%
     mutate(unique_day =
-             glue("{year(locs_dt)}",
-                  "{yday(locs_dt)}",
+             glue::glue("{lubridate::year(locs_dt)}",
+                  "{lubridate::yday(locs_dt)}",
                   .sep = "_"))
   dbDisconnect(con)
+
+  load(file = here::here('data/adfg_locations.rda'))
+  load(file = here::here('data/adfg_deployments.rda'))
+
+  adfg_locations <- adfg_locations %>%
+    rename(quality = location_quality, locs_dt = date_time, type = tag_type) %>%
+    left_join(adfg_deployments, by = 'deployid')
+    sf::st_as_sf(coords = c('longitude','latitude')) %>%
+    sf::st_set_crs(4326) %>%
+    sf::st_transform(3571)
 
   return(locs_sf)
 

### 2021-05-17 analysis/scripts/model.R

@@ -12,7 +12,8 @@ library(purler)
 create_ar1_id <- function(h_dt) {
   ar1_id <- tibble(
     t_seq = seq(from = h_dt[1], to = h_dt[length(h_dt)], by = "hour",)
-  ) %>% left_join(tibble(h_dt = h_dt, haulout_dt = h_dt), by = c("t_seq" = "h_dt")) %>%
+  ) %>%
+    left_join(tibble(h_dt = h_dt, haulout_dt = h_dt), by = c("t_seq" = "h_dt")) %>%
     mutate(x = is.na(haulout_dt),
            rle = rlenc_id(x)) %>%
     filter(!is.na(haulout_dt)) %>%
@@ -105,6 +106,18 @@ fit_ribbon <- glmmLDTS(fixed.formula = dry ~ age_sex + sin1 + cos1 + sin2 + cos2
 
 save(fit_ribbon, file = here::here('data/fit_ribbon.Rdata'))
 
+fit_ribbon_noagesex <- glmmLDTS(fixed.formula = dry ~ sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + day + day2+ day3 + temp2 + wind + pressure + precip + wind*temp2 +
+                                  sin1*day + cos1*day + sin2*day + cos2*day + sin3*day + cos3*day +
+                                  sin1*day2 + cos1*day2 + sin2*day2 + cos2*day2 + sin3*day2 + cos3*day2,
+                                random.formula = dry ~ speno,
+                                data = HO_ribbon,
+                                EstMeth = "REML",
+                                timecol = "time_vec",
+                                #ridge.reg = "global",
+                                #lambda = 0.5,
+                                group.vec = "ar1_id")
+
+save(fit_ribbon_noagesex, file = here::here('data/fit_ribbon_noagesex.Rdata'))
 
 
 HO_spotted = dat.sf %>%
@@ -128,6 +141,18 @@ fit_spotted <- glmmLDTS(fixed.formula = dry ~ age_sex + sin1 + cos1 + sin2 + cos
                        #lambda = 0.5,
                        group.vec = "ar1_id")
 save(fit_spotted, file = here::here('data/fit_spotted.Rdata'))
+
+fit_spotted_noagesex <- glmmLDTS(fixed.formula = dry ~ sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + day + day2+ day3 + temp2 + wind + pressure + precip + wind*temp2 +
[…]

### 2021-06-08 data_raw/adfg/adfg_tidydata.Rmd
FINALIZING ADFG DATA PROCESSING

@@
                 TagType = toupper(TagType)) %>% 
-  dplyr::rename(deployid = DeployIDs,
+  dplyr::rename(speno = DeployIDs,
                 tag_type = TagType,
@@
   dplyr::rename_all(tolower) %>% 
+  dplyr::mutate(deployid = paste(speno,tag_type, sep = "_")) %>% 
+  dplyr::relocate(deployid) %>% 
   dplyr::select(-c(monthnumb,latitude,longitude,locationquality,sum,hist_type))
@@
                       data_date = lubridate::floor_date(date_time,"day")) %>%
-        dplyr::select(c(deployid,ptt,tag_type,date_time,data_date,percent_dry)) %>% 
+        dplyr::select(c(deployid,speno,ptt,tag_type,date_time,data_date,percent_dry)) %>% 
         dplyr::arrange(deployid,date_time)

### 2021-06-08 analysis/paper/paper.Rmd
INITIAL EDITS PER PLB COMMENTS

@@
   - Jay M. Ver Hoef:
-      institute: [MML]
-      correspondence: false
-  - Irina S. Trukhanova:
       institute: [MML]
@@
 Ice-associated seals rely on sea ice for a variety of activities, including breeding,
-molting, pupping, and resting. In the Arctic, many of these activities occur in spring and
-early summer (April-June) as sea ice begins to melt and retreat northward. Rapid
-acceleration of climate change in Arctic ecosystems is therefore of concern as the
-quantity and quality of suitable habitat is forecast to decrease. In this paper, we use
-data collected from satellite-linked bio-loggers deployed between 2005 and 2020 to
-investigate the seasonal timing and environmental factors affecting sea ice use by seals
-(specifically, bearded, ribbon, and spotted seals) in the Bering and Chukchi seas. In
-addition to providing baseline data on phenology, these data also allow us to refine
-availability estimates needed to accurately estimate abundance from aerial survey counts
-of seals basking on ice (i.e., to correct for the proportion of animals that are in the
-water while surveys are conducted). Using generalized linear mixed pseudo-models to
-properly account for temporal autocorrelation, we fit models with a variety of covariates
-(e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure,
-temperature, precipitation) to examine their ability to explain variation in hourly
-haul-out records. We found evidence for strong diurnal and within season patterns in
-haul-out behavior, as well as strong weather effects (particularly wind and temperature).
-In general, seals were more likely to rest on ice in the middle of the day and when wind
-speed was low and temperatures were higher. Haul-out propensity increased through March
-and April, peaking in May and early June before declining again. The timing and frequency
-of haul-out events also varied based on species and age-sex class. For ribbon and spotted
-seals, models with year effects were highly supported, indicating that the timing and
-magnitude of haul-out behavior varied among years. However, we did not find evidence that
-haul-out timing was linked annual sea ice extent. Our analysis emphasizes the importance
-of accounting for variation in haul-out behavior when interpreting the number of seals
-counted in aerial surveys.
+molting, pupping, and resting. In the Arctic, many of these activities occur in spring
+(April-June) as sea ice begins to melt and retreat northward. Rapid acceleration of
+climate change in Arctic ecosystems is therefore of concern as the quantity and quality of
+suitable habitat is forecast to decrease. In this paper, we use data collected from
+satellite-linked bio-loggers deployed between 2005 and 2020 to investigate the seasonal
+timing and environmental factors affecting sea ice use by seals (specifically, bearded,
+ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to providing
+baseline data on phenology, these data also allow us to quantify ‘availability’, which is
+needed to accurately estimate abundance from aerial survey counts of seals basking on ice
+(i.e., to correct for the proportion of animals that are in the water while surveys are
+conducted). Using generalized linear mixed pseudo-models to properly account for temporal
+autocorrelation, we fit models with a variety of covariates (e.g., day-of-year, solar
+hour, age-sex class, wind speed, barometric pressure, temperature, precipitation) to
+examine their ability to explain variation in hourly haul-out records. We found evidence
+for strong diurnal and within-season patterns in haul-out behavior, as well as strong
+weather effects (particularly wind and temperature). In general, seals were more likely to
+rest on ice in the middle of the day and when wind speed was low and temperatures were
+higher. Haul-out propensity increased through March and April, peaking in May and early
+June before declining again. The timing and frequency of haul-out events also varied based
+on species and age-sex class. For ribbon and spotted seals, models with year effects were
+highly supported, indicating that the timing and magnitude of haul-out behavior varied
+among years. However, we did not find evidence that haul-out timing was linked to annual
+sea ice extent. Our analysis emphasizes the importance of accounting for variation in
+haul-out behavior when interpreting the number of seals counted in aerial surveys.
 
@@
 
-The global climate crisis is causing considerable reduction in Arctic sea ice extent,
+The global climate disruption is causing considerable reduction in Arctic sea ice extent,
 volume, and seasonal presence [@kwok2018erl, @meier2014rg, @wang2017dsrpitsio]. These
-changes have ripple effects on Arctic ecosystems, taxa, and the human communities who live
-in the region [@huntington2020ncc]. Such disruptions are a particular cause of concern for
-ice-associated seals (ringed (*Phoca hispida*), bearded (*Erignathus barbatus*), spotted
-(*Phoca largha*), and ribbon seals (*Histriophoca fasciata*)) that depend on spring and
-early summer sea ice (March-June) in the Bering and Chukchi seas as a platform for
-important functions, such as pupping, nursing, breeding, and molting [@boveng2009a;
-@boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge gaps prevent our
-ability to make definitive predictions about the ultimate effects of changes in sea ice on
-the behavior, abundance, and distribution of these seals. Knowledge about biological
-constraints on the phenology of reproductive and molting behavior is generally lacking, so
-it is difficult to predict the readiness with which ice-associated seal species will be
-able to adapt to future changes (e.g., by adjusting pupping or molting schedules to
-earlier dates or different locales). Additionally, trends in abundance of these species
-are largely unknown, so it is difficult to say what effect, if any, declines in sea ice
-habitat have had, or will have, on seal densities.
+changes have ripple effects on Arctic organisms, ecosystems, and the human communities who
+live in the region [@huntington2020ncc]. Such disruptions are a particular cause of
+concern for ice-associated seals (ringed (*Phoca hispida*), bearded (*Erignathus
+barbatus*), spotted (*Phoca largha*), and ribbon seals (*Histriophoca fasciata*)) that
+depend on spring and early summer sea ice (March-June) in the Bering and Chukchi seas as a
+platform for important functions, such as pupping, nursing, breeding, and molting
+[@boveng2009a;
+@boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge gaps complicate
+predictions about the ultimate effects of changes in sea ice on the behavior, abundance,
+and distribution of these seals. Knowledge about evolutionary constraints on the timing of
+reproductive and molting behavior is generally lacking, so it is difficult to predict the
+readiness with which ice-associated seal species will be able to adapt to future changes
+(e.g., by adjusting pupping or molting schedules to earlier dates or different locales).
+Additionally, trends in abundance of these species are largely unknown, so it is difficult
+to say what effect, if any, declines in sea ice habitat have had, or will have, on seal
+densities.
 
@@
 on beaches during ice-free months, they strongly prefer sea ice as a resting platform
-during the spring and early summer [@burns2009]. The remote nature of the Bering and
-Chukchi seas mean direct observation of seal behavior is limited. Bio-logging devices are,
-thus, especially useful tools for collecting key information on movement and haul-out
-behavior for these species. Records of time spent out of the water provide valuable data
-for identifying covariates that explain variation in haul-out behavior. For instance, in
-the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to
-demonstrate greater haul-out propensity in juvenile crabeater seals (*Lobodon
-carcinophaga*) than adults, with highest probabilities in February and at times close to
-solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated seasonal
-changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson et al.
-[-@bengtson2005pb] documented a higher propensity for ringed seal basking near solar noon,
-as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded , ribbon , and spotted
-seals using much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that the
-proportion of time bearded seals spent hauled out progressively increased through spring
-and summer. And, Ver Hoef et al. [-@verhoef2014sm] found haul-out probabilities increased
-gradually starting in March and peaking in May and June for bearded, ribbon, and spotted
-seals. However, their primary focus was to develop a correction factor for estimating seal
-abundance, and like a subsequent analysis by Conn et al. [-@conn2014mee], they did not
-attempt to separately analyze haul-out schedules for different age and sex classes
-(although initial models with sex and age as fixed effects were insignificant at p = 0.05
-[@verhoef2014sm]).
+during the spring and early summer [@burns2009]. The remoteness of the Bering and Chukchi
+seas mean direct scientific observation of seal behavior is impractical. Bio-logging
+devices are, thus, especially useful tools for collecting key information on movement and
+haul-out behavior for these species. Bio-logging records of time spent out of the water
+provide valuable data for identifying covariates that explain variation in haul-out
+behavior. For instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied
+on bio-logging data to demonstrate greater haul-out propensity in juvenile crabeater seals
+(*Lobodon carcinophaga*) than adults, with highest probabilities in February and at times
+close to solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated
+seasonal changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson
+et al. [-@bengtson2005pb] documented a higher propensity for ringed seal basking near
+solar noon, as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded , ribbon ,
+and spotted seals using much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that
+the proportion of time bearded seals spent hauled out progressively increased through
+spring and summer. And, Ver Hoef et al. [-@verhoef2014sm] found haul-out probabilities
+increased gradually starting in March and peaking in May and June for bearded, ribbon, and
+spotted seals. However, their primary focus was to develop a correction factor for
+estimating seal abundance, and like a subsequent analysis by Conn et al. [-@conn2014mee],
+they did not attempt to separately analyze haul-out schedules for different age and sex
+classes (although initial models with sex and age as fixed effects were insignificant at p
+= 0.05 [@verhoef2014sm]).
 

### 2021-06-25 data_raw/adfg/adfg_tidydata.Rmd
CONT TO INCORPORATE RAW DATA FROM ADFG/NSB
a large collection of code and data updates as we work to fully incoporate data share from ADFG and NSB

--- 2021-05-07 data_raw/adfg/adfg_tidydata.Rmd
@@
 adfg_deployments <- adfg_deployments %>% 
-  dplyr::rename(deployid = DeployIDs,
+  dplyr::rename(speno = DeployIDs,
                 mapid = MapID,
-                date_tagged = DateTagged,
+                deploy_dt = DateTagged,
                 primary_tag_type = PrimaryTagType,
@@
                 sex = tolower(sex)) %>% 
-  tidyr::gather(tag_position, tag_type, c(primary_tag_type,flipper_tag_type)) %>% 
+  tidyr::gather(tag_position, tag_family, c(primary_tag_type,flipper_tag_type)) %>% 
   dplyr::mutate(tag_position = ifelse(tag_position == "primary_tag_type",
                    "primary","flipper")) %>% 
-  dplyr::filter(!is.na(tag_type)) %>% 
-  dplyr::arrange(date_tagged,deployid)
+  dplyr::filter(!is.na(tag_family)) %>% 
+  dplyr::arrange(deploy_dt)
 
@@
 adfg_locations <- adfg_locations %>% 
-  dplyr::rename(deployid = DeployID,
-                location_quality = LocationQuality,
-                date_time = Date,
-                tag_type = TagType) %>% 
-  dplyr::filter(location_quality != "Z") %>% 
+  dplyr::rename(speno = DeployID,
+                quality = LocationQuality,
+                locs_dt = Date,
+                tag_family = TagType) %>% 
+  dplyr::filter(quality != "Z") %>% 
   dplyr::select(-MonthNumb) %>% 
-  dplyr::rename_all(tolower)
+  dplyr::rename_all(tolower) %>% 
+  dplyr::left_join(adfg_deployments, by = c("speno","tag_family")) %>% 
+  mutate(species_code = case_when(
+        species == "bearded" ~ "EB",
+        species == "ribbon" ~ "HF",
+        species == "spotted" ~ "PL"
+    )) %>% 
+    mutate(deployid = glue::glue("{species_code}",
+                            "{lubridate::year(deploy_dt)}_{ptt}")) %>% 
+    mutate(unique_day =
+             glue::glue("{lubridate::year(locs_dt)}",
+                  "{lubridate::yday(locs_dt)}",
+                  .sep = "_")) %>% 
+  select(-species_code) %>% 
+  relocate(deployid,ptt) %>% 
+  filter(species %in% c('bearded', 'ribbon', 'spotted')) %>%
+    filter(lubridate::month(locs_dt) %in% c(3,4,5,6,7)) %>%
+  sf::st_as_sf(coords = c("longitude","latitude")) %>% 
+  sf::st_set_crs(value = "EPSG:4326")
 
@@
 
-locs_sf <- adfg_locations %>% 
-  dplyr::left_join(adfg_deployments, by = c("deployid","tag_type")) %>% 
-  sf::st_as_sf(coords = c("longitude","latitude")) %>% 
-  sf::st_set_crs(4326) %>% sf::st_transform(3571)
-
 ggplot() +
@@
   layer_spatial(
-    data = locs_sf, size = 0.125
+    data = sf::st_transform(adfg_locations,3571), size = 0.125
   ) +
@@
 ```{r}
-adfg_timelines <- readr::read_csv('ADFG_SealTimelineData3.csv')
+adfg_timelines <- readr::read_csv(here::here('data_raw/adfg/ADFG_SealTimelineData3.csv'))
 ```
@@
 
-adfg_timelines <- readr::read_csv('ADFG_SealTimelineData3.csv',
+adfg_timelines <- readr::read_csv(here::here('data_raw/adfg/ADFG_SealTimelineData3.csv'),
                                   col_types = adfg_timeline_cols) %>% 
@@
                 TagType = toupper(TagType)) %>% 
-  dplyr::rename(deployid = DeployIDs,
-                tag_type = TagType,
+  dplyr::rename(speno = DeployIDs,
+                tag_family = TagType,
                 hist_type = HistType,
-                date_time = GMTDate) %>% 
+                timeline_start_dt = GMTDate) %>% 
   dplyr::rename_all(tolower) %>% 
+  dplyr::left_join(adfg_deployments, by = c("speno","tag_family")) %>% 
+  mutate(species_code = case_when(
+        species == "bearded" ~ "EB",
+        species == "ribbon" ~ "HF",
+        species == "spotted" ~ "PL"
+    )) %>% 
+  mutate(deployid = glue::glue("{species_code}",
+                            "{lubridate::year(deploy_dt)}_{ptt}")) %>%
+  mutate(unique_day =
+             glue::glue("{lubridate::year(timeline_start_dt)}",
+                  "{lubridate::yday(timeline_start_dt)}",
+                  .sep = "_")) %>%
+  dplyr::relocate(deployid) %>% 
+  filter(species %in% c('bearded', 'ribbon', 'spotted')) %>%
+    filter(lubridate::month(timeline_start_dt) %in% c(3,4,5,6,7)) %>%
   dplyr::select(-c(monthnumb,latitude,longitude,locationquality,sum,hist_type))
@@
         dplyr::left_join(bins, by = "bin") %>%
-        dplyr::mutate(date_time = date_time + lubridate::hours(hour),
-                      data_date = lubridate::floor_date(date_time,"day")) %>%
-        dplyr::select(c(deployid,ptt,tag_type,date_time,data_date,percent_dry)) %>% 
-        dplyr::arrange(deployid,date_time)
+        dplyr::mutate(timeline_start_dt = timeline_start_dt + lubridate::hours(hour)) %>%
+        dplyr::select(c(deployid,speno,species,age,sex,deploy_dt,ptt,tag_family,timeline_start_dt,unique_day,percent_dry)) %>% 
+        dplyr::arrange(deployid,timeline_start_dt)
 ```

### 2021-06-25 data_raw/nsb/nsb_tidydata.Rmd
CONT TO INCORPORATE RAW DATA FROM ADFG/NSB
a large collection of code and data updates as we work to fully incoporate data share from ADFG and NSB

---
title: "Import and Tidy NSB Supplied Location and Haul-out Data"
description: | 
  This article describes the import and tidying of location and haul-out data provided to
  MML by North Slope Burough for use in manuscript analyzing haul-out behavior of bearded, ribbon, and
  spotted seals in the Bering and Chukchi seas
date: June 14, 2021
author:
  - name: "Josh M. London"
    url: https://github.com/jmlondon
    affiliation: Alaska Fisheries Science Center
    affiliation_url: https://www.fisheries.noaa.gov/about/alaska-fisheries-science-center
    orcid_id: 0000-0002-3647-5046
output: distill::distill_article
---

```{r packages, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
```

## Introduction

The focus of this document is not to analyze the data, but instead to document
the processing of the 'raw' data (as provided) into a usable data structure for
downstream merging with other datasets and analysis procedures. We will focus
on reading from the database, tidying any data structures or values that need it,
and then merging into a single usable tibble.

## Read from Database

```{r}
library(DBI)
library(odbc)
library(pingr)
library(dbplyr)
library(sf)
stopifnot(
"PEP Postgres Database Not Available; did you start VPN? ;)" = 
pingr::is_up("161.55.120.122", "5432")
)

con <- dbConnect(
  odbc::odbc(),
  dsn = "PostgreSQL pep",
  uid = keyringr::get_kc_account("pgpep_londonj"),
  pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
)

deploy_qry <- "select speno, deployid, common_name as species, age_class as age, sex, tag_family, deploy_dt, end_dt
from telem.nsb_capture c
left join telem.nsb_deploy d
on c.id = d.capture_id
left join telem.lku_tag_family
using (tag_family_lku)
left join capture.lku_species
using (species_lku)
left join capture.lku_age_class
using (age_class_lku)
left join capture.lku_sex
using (sex_lku)"

nsb_deployments <- tbl(con, sql(deploy_qry)) %>% 
    dplyr::select(speno, deployid, species, age, sex, tag_family, deploy_dt, end_dt) %>% 
  collect()

nsb_timelines <- tbl(con, in_schema("telem","nsb_histos_timeline")) %>%
    dplyr::select(deployid,timeline_start_dt, percent_dry) %>% 
  collect()

locs_qry <- "SELECT deployid, loc_dt as locs_dt, argos_lc as quality, loc_geom as geometry
              FROM telem.nsb_locs;"

nsb_locations <- read_sf(con, query = locs_qry) %>%
    left_join(nsb_deployments, by = 'deployid') %>%
    filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
    filter(lubridate::month(locs_dt) %in% c(3,4,5,6,7)) %>%
    rename(ptt = deployid) %>% 
    mutate(species_code = case_when(
        species == "Bearded seal" ~ "EB",
        species == "Ribbon seal" ~ "HF",
        species == "Spotted seal" ~ "PL"
    )) %>% 
    mutate(deployid = glue::glue("{species_code}",
                            "{lubridate::year(locs_dt)}_{ptt}")) %>% 
    mutate(unique_day =
             glue::glue("{lubridate::year(locs_dt)}",
                  "{lubridate::yday(locs_dt)}",
                  .sep = "_")) %>% 
  select(-species_code) %>% 
  relocate(deployid,ptt)

nsb_timelines <- nsb_timelines %>% 
  left_join(nsb_deployments, by = 'deployid') %>% 
  filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
  filter(lubridate::month(timeline_start_dt) %in% c(3,4,5,6,7)) %>% 
  rename(ptt = deployid) %>% 
    mutate(species_code = case_when(
        species == "Bearded seal" ~ "EB",
        species == "Ribbon seal" ~ "HF",
        species == "Spotted seal" ~ "PL"
    )) %>% 
    mutate(deployid = glue::glue("{species_code}",
                            "{lubridate::year(timeline_start_dt)}_{ptt}")) %>% 
    mutate(unique_day =
             glue::glue("{lubridate::year(timeline_start_dt)}",
                  "{lubridate::yday(timeline_start_dt)}",
                  .sep = "_")) %>% 
  select(-species_code) %>% 
  relocate(deployid,ptt) %>% 
  dplyr::arrange(deployid,timeline_start_dt)

usethis::use_data(nsb_deployments, overwrite = TRUE)
usethis::use_data(nsb_timelines, overwrite = TRUE)
usethis::use_data(nsb_locations, overwrite = TRUE)
```

### 2021-06-25 R/create_db_input_data.R

@@ -34,7 +34,7 @@ create_db_input_data <- function(timeline_data, locs_sf) {
     select(-error_radius.y) %>%
     mutate(error_radius = ifelse(type %in% c("GPS","FastGPS"),
                                         50,error_radius),
-                  error_radius = ifelse(type == "User",
+                  error_radius = ifelse(type %in% c("User"),
                                         50,error_radius)) %>%
     group_by(speno,unique_day,age,sex,species) %>%
     summarise(x = weighted.mean(x,1/error_radius),
@@ -70,8 +70,8 @@ create_db_input_data <- function(timeline_data, locs_sf) {
   con <- dbConnect(
     odbc(),
     dsn = "PostgreSQL pep",
-    uid = get_kc_account("pgpep_londonj"),
-    pwd = decrypt_kc_pw("pgpep_londonj")
+    uid = keyringr::get_kc_account("pgpep_londonj"),
+    pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
   )
 
   st_write(obj = tbl_percent_locs,

### 2021-06-25 R/get_locs_sf.R

@@ -11,7 +11,7 @@ get_locs_sf <- function() {
 
   stopifnot(
     "PEP Postgres Database Not Available; did you start VPN? ;)" =
-      is_up("161.55.120.122", "5432")
+      pingr::is_up("161.55.120.122", "5432")
   )
 
   con <- dbConnect(
@@ -22,11 +22,12 @@ get_locs_sf <- function() {
   )
 
   deployments_db <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>%
-    select(speno, deployid, tag_family, deploy_dt, end_dt) %>% collect()
+    select(speno, deployid, tag_family, deploy_dt, end_dt) %>%
+    rename(tag_type = tag_family) %>% collect()
 
   spenos_db <- tbl(con, in_schema("capture","for_telem")) %>% collect()
 
-  locs_qry <- "SELECT deployid, ptt, type, error_radius, locs_dt, quality, geom
+  locs_qry <- "SELECT deployid, ptt, type, error_radius, locs_dt, quality, geom as geometry
               FROM telem.geo_wc_locs_qa;"
 
   locs_sf <- read_sf(con, query = locs_qry) %>%
@@ -41,14 +42,10 @@ get_locs_sf <- function() {
   dbDisconnect(con)
 
   load(file = here::here('data/adfg_locations.rda'))
-  load(file = here::here('data/adfg_deployments.rda'))
+  load(file = here::here('data/nsb_locations.rda'))
 
-  adfg_locations <- adfg_locations %>%
-    rename(quality = location_quality, locs_dt = date_time, type = tag_type) %>%
-    left_join(adfg_deployments, by = 'deployid')
-    sf::st_as_sf(coords = c('longitude','latitude')) %>%
-    sf::st_set_crs(4326) %>%
-    sf::st_transform(3571)
+  locs_sf <- locs_sf %>% bind_rows(adfg_locations) %>%
+    bind_rows(nsb_locations)
 
   return(locs_sf)
 

### 2021-06-25 analysis/scripts/get_data/Haulout_LoadAndProcessCov_20180926_SKH.R

@@ -1,0 +1,30 @@
+#library(icesealhaulout)
+library(RPostgreSQL)
+library(tidyverse)
+
+con <- RPostgreSQL::dbConnect(PostgreSQL(), 
+                              dbname = Sys.getenv("pep_db"), 
+                              host = Sys.getenv("pep_ip"), 
+                              user = Sys.getenv("pep_admin"), 
+                              rstudioapi::askForPassword(paste("Enter your DB password for user account: ", Sys.getenv("pep_admin"), sep = "")))
+
+#data("percent_locs") 
+# load("//AKC0SS-N086/NMML_Users/Stacie.Hardy/Desktop/icesealhaulout/data/percent_locs.rda")
+# percent_locs %>% dplyr::ungroup() 
+# percent_locs$fill_xy <- as.character(percent_locs$fill_xy)
+# percent_locs$id <- 1:nrow(percent_locs)
+# percent_locs <- percent_locs[, c(10, 1, 5, 6, 7, 3, 4, 8, 9)]
+# colnames(percent_locs)[[6]] <- "haulout_dt"
+# RPostgreSQL::dbSendQuery(con, "DROP TABLE IF EXISTS telem.res_iceseal_haulout")
+# sf::st_write_db(con, percent_locs, c("telem", "res_iceseal_haulout"), row.names = FALSE)
+# RPostgreSQL::dbSendQuery(con, "ALTER TABLE telem.res_iceseal_haulout RENAME COLUMN geometry TO geom")
+
+# Create index on haulout data
+RPostgreSQL::dbSendQuery(con, "SELECT telem.fxn_iceseal_pred_idx();")
+RPostgreSQL::dbClearResult(RPostgreSQL::dbListResults(con)[[1]])
+
+# Create covariate tables
+RPostgreSQL::dbSendQuery(con, "SELECT telem.fxn_iceseal_haulout_cov();")
+RPostgreSQL::dbClearResult(RPostgreSQL::dbListResults(con)[[1]])
+
+RPostgreSQL::dbDisconnect(con)

### 2021-06-29 analysis/paper/paper.Rmd
UPDATED DATA AND RE-RUN MODELS
at this point the ADFG and NSB collaborative data have been incorporated and models re-run.

--- 2021-04-28 analysis/paper/paper.Rmd
@@
   mutate(age = case_when(
-    age == "PUP" ~ "YOUNG OF YEAR",
-    age == "YEARLING" ~ "SUBADULT",
-    TRUE ~ age
-  ))
-
-deploy_meta <- tbl(con, in_schema("telem", "tbl_tag_deployments")) %>% 
+    age %in% c("pup", "PUP") ~ "YOUNG OF YEAR",
+    age %in% c("yearling","YEARLING","subadult","Subadult") ~ "SUBADULT",
+    age %in% c("adult","ADULT","Adult") ~ "ADULT",
+    TRUE ~ age),
+    species = case_when(
+    species == "bearded" ~ "Bearded seal",
+    species == "spotted" ~ "Spotted seal",
+    TRUE ~ species),
+    sex = case_when(
+    sex %in% c("Female","female") ~ "F",
+    sex %in% c("Male","male") ~ "M",
+    TRUE ~ sex)
+    ) %>% 
+  filter(!is.na(age))
+
+load(here::here('data/locs_sf.Rdata'))
+
+deploy_meta <- locs_sf %>% 
+  as_data_frame() %>% 
+  select(speno, deploy_dt) %>% 
   filter(speno %in% !!unique(dat.sf$speno)) %>% 
   group_by(speno) %>% 
-  summarise(deploy_dt = min(deploy_dt, na.rm = TRUE)) %>% 
-  collect()
+  summarise(deploy_dt = min(deploy_dt, na.rm = TRUE))
 
@@
 H_t=\alpha_1 \cos(\frac{\pi t}{4}) + \alpha_2 \sin(\frac{\pi t}{4}) + \alpha_3 \cos(\frac{\pi t}{6}) + \alpha_4 \sin(\frac{\pi t}{6}) + \alpha_5 \cos(\frac{\pi t}{12}) + \alpha_6 \sin(\frac{\pi t}{12})
-$$ where $H_t$ gives the effect for solar hour $t$ and $α_{i}$ are estimated parameters
+$$ 
+where $H_t$ gives the effect for solar hour $t$ and $α_{i}$ are estimated parameters
 (regression coefficients).

### 2021-06-29 data_raw/adfg/adfg_tidydata.Rmd
UPDATED DATA AND RE-RUN MODELS
at this point the ADFG and NSB collaborative data have been incorporated and models re-run.

@@
                                                  'America/Anchorage')) %>% 
-  dplyr::mutate(DateTagged = lubridate::as_date(DateTagged, 
+  dplyr::mutate(DateTagged = lubridate::as_datetime(DateTagged, 
                                                 tz = 'America/Anchorage'))

### 2021-06-29 R/create_db_input_data.R

@@ -70,14 +70,15 @@ create_db_input_data <- function(timeline_data, locs_sf) {
   con <- dbConnect(
     odbc(),
     dsn = "PostgreSQL pep",
-    uid = keyringr::get_kc_account("pgpep_londonj"),
-    pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
+    uid = keyringr::get_kc_account("pgpep_sa"),
+    pwd = keyringr::decrypt_kc_pw("pgpep_sa")
   )
 
   st_write(obj = tbl_percent_locs,
            dsn = con,
-           layer = "res_iceseal_haulout_jml",
-           layer_options = "OVERWRITE=true")
+           layer = SQL("telem.res_iceseal_haulout"),
+           layer_options = c("OVERWRITE=true","GEOMETRY_NAME=geom")
+  )
 
   return(tbl_percent_locs)
 

### 2021-06-29 analysis/scripts/get_data/Haulout_LoadAndProcessCov_20180926_SKH.R

@@ -2,15 +2,15 @@
 library(RPostgreSQL)
 library(tidyverse)
 
-con <- RPostgreSQL::dbConnect(PostgreSQL(), 
-                              dbname = Sys.getenv("pep_db"), 
-                              host = Sys.getenv("pep_ip"), 
-                              user = Sys.getenv("pep_admin"), 
+con <- RPostgreSQL::dbConnect(PostgreSQL(),
+                              dbname = Sys.getenv("pep_db"),
+                              host = Sys.getenv("pep_ip"),
+                              user = Sys.getenv("pep_admin"),
                               rstudioapi::askForPassword(paste("Enter your DB password for user account: ", Sys.getenv("pep_admin"), sep = "")))
 
-#data("percent_locs") 
+#data("percent_locs")
 # load("//AKC0SS-N086/NMML_Users/Stacie.Hardy/Desktop/icesealhaulout/data/percent_locs.rda")
-# percent_locs %>% dplyr::ungroup() 
+# percent_locs %>% dplyr::ungroup()
 # percent_locs$fill_xy <- as.character(percent_locs$fill_xy)
 # percent_locs$id <- 1:nrow(percent_locs)
 # percent_locs <- percent_locs[, c(10, 1, 5, 6, 7, 3, 4, 8, 9)]
@@ -20,7 +20,7 @@ con <- RPostgreSQL::dbConnect(PostgreSQL(),
 # RPostgreSQL::dbSendQuery(con, "ALTER TABLE telem.res_iceseal_haulout RENAME COLUMN geometry TO geom")
 
 # Create index on haulout data
-RPostgreSQL::dbSendQuery(con, "SELECT telem.fxn_iceseal_pred_idx();")
+dbSendQuery(con, "SELECT telem.fxn_iceseal_pred_idx();")
 RPostgreSQL::dbClearResult(RPostgreSQL::dbListResults(con)[[1]])
 
 # Create covariate tables

### 2021-06-29 analysis/scripts/model.R

@@ -26,9 +26,9 @@ load(here::here('data/dat_sf.Rdata'))
 
 dat.sf <- dat.sf  %>%
   mutate(dry = round(percent_dry/100)) %>%
-  mutate(hour_utc = hour(haulout_dt),
-         yday = yday(haulout_dt),
-         year = year(haulout_dt)) %>%
+  mutate(hour_utc = lubridate::hour(haulout_dt),
+         yday = lubridate::yday(haulout_dt),
+         year = lubridate::year(haulout_dt)) %>%
   mutate(coords_x = st_coordinates(st_transform(.,4326))[,1],
          coords_y = st_coordinates(st_transform(.,4326))[,2],
          solar_hour = solaR::local2Solar(haulout_dt,coords_x) %>% lubridate::hour(),

### 2021-06-30 analysis/scripts/get_data/creating_input_data.Rmd

@@
   source(here::here('analysis/scripts/get_data/make_data.R'))
+  save(locs_sf, file = here::here('data/locs_sf.Rdata'))
+  save(timeline_data, file = here::here('timeline_data.Rdata'))
 } else {

### 2021-08-12 analysis/paper/paper.Rmd
CONTINUED UPDATES TO PAPER/FIGURES

@@
 ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to providing
-baseline data on phenology, these data also allow us to quantify ‘availability’, which is
+baseline data on phenology, these data also allow us to quantify 'availability', which is
 needed to accurately estimate abundance from aerial survey counts of seals basking on ice
@@
 live in the region [@huntington2020ncc]. Such disruptions are a particular cause of
-concern for ice-associated seals (ringed (*Phoca hispida*), bearded (*Erignathus
+concern for ice-associated seals (ringed (*Pusa hispida*), bearded (*Erignathus
 barbatus*), spotted (*Phoca largha*), and ribbon seals (*Histriophoca fasciata*)) that
@@
 platform for important functions, such as pupping, nursing, breeding, and molting
-[@boveng2009a;
-@boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge gaps complicate
-predictions about the ultimate effects of changes in sea ice on the behavior, abundance,
-and distribution of these seals. Knowledge about evolutionary constraints on the timing of
-reproductive and molting behavior is generally lacking, so it is difficult to predict the
-readiness with which ice-associated seal species will be able to adapt to future changes
-(e.g., by adjusting pupping or molting schedules to earlier dates or different locales).
-Additionally, trends in abundance of these species are largely unknown, so it is difficult
-to say what effect, if any, declines in sea ice habitat have had, or will have, on seal
-densities.
+[@boveng2009a; @boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge
+gaps complicate predictions about the ultimate effects of changes in sea ice on the
+behavior, abundance, and distribution of these seals. Knowledge about evolutionary
+constraints on the timing of reproductive and molting behavior is generally lacking, so it
+is difficult to predict the readiness with which ice-associated seal species will be able
+to adapt to future changes (e.g., by adjusting pupping or molting schedules to earlier
+dates or different locales). Additionally, trends in abundance of these species are
+largely unknown, so it is difficult to say what effect, if any, declines in sea ice
+habitat have had, or will have, on seal densities.
 
@@
 devices are, thus, especially useful tools for collecting key information on movement and
-haul-out behavior for these species. Bio-logging records of time spent out of the water
-provide valuable data for identifying covariates that explain variation in haul-out
-behavior. For instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied
-on bio-logging data to demonstrate greater haul-out propensity in juvenile crabeater seals
-(*Lobodon carcinophaga*) than adults, with highest probabilities in February and at times
-close to solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated
-seasonal changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson
-et al. [-@bengtson2005pb] documented a higher propensity for ringed seal basking near
-solar noon, as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded , ribbon ,
-and spotted seals using much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that
-the proportion of time bearded seals spent hauled out progressively increased through
-spring and summer. And, Ver Hoef et al. [-@verhoef2014sm] found haul-out probabilities
-increased gradually starting in March and peaking in May and June for bearded, ribbon, and
-spotted seals. However, their primary focus was to develop a correction factor for
-estimating seal abundance, and like a subsequent analysis by Conn et al. [-@conn2014mee],
-they did not attempt to separately analyze haul-out schedules for different age and sex
-classes (although initial models with sex and age as fixed effects were insignificant at p
-= 0.05 [@verhoef2014sm]).
-
-Several authors have used logistic regression-style analyses to estimate the probability
-of sea ice use from data garnered from bio-loggers. In these models, haul-out
-probabilities were expressed as a function of predictive covariates, such as time of day,
-day-of-year, sex, age class, and environmental conditions (e.g., [@reder2003pb],
-[@bengtson2004pb], [@bengtson2005pb], [@udevitz2009cjz], and [@verhoef2014sm]). However,
-sample sizes have often been low and insufficient to permit strong inference about life
-history and/or seasonal variation in haul-out probabilities. For instance, Bengtson and
-Cameron's [-@bengtson2004pb] study included 5 adult and 2 juvenile crabeater seals, while
-Bengtson et al.'s [-@bengtson2005pb] study was based on 6 telemetered ringed seals in the
-Chukchi Sea. These studies were often further limited by logistical constraints on tagging
-fieldwork and the attachment duration or operational life of satellite tags.
+haul-out behavior for these species.
+
+Bio-logging records of time spent out of the water provide valuable data for identifying
+covariates that explain variation in haul-out behavior. For instance, in the Antarctic,
+Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to demonstrate greater
+haul-out propensity in juvenile crabeater seals (*Lobodon carcinophaga*) than adults, with
+highest probabilities in February and at times close to solar noon. In the Arctic, Von
+Duyke et al. [-@vonduyke2020ee] used seal-borne, satellite-linked loggers to corroborate
+seasonal changes between diurnal and nocturnal haul-out behavior of ringed seals
+previously described by Kelly et al. [-@kelly2010] using labor-intensive VHF radio tags
+and direct observation. Bengtson et al. [-@bengtson2005pb] documented a higher propensity
+for ringed seal basking near solar noon, as did Ver Hoef et al. [-@verhoef2014sm] in an
+analysis of bearded, ribbon, and spotted seals using much larger sample sizes. Olnes et
+al. [-@olnes2020pb] showed that the proportion of time bearded seals spent hauled out
+progressively increased through spring and summer. And, Ver Hoef et al. [-@verhoef2014sm]
+found haul-out probabilities increased gradually starting in March and peaking in May and
+June for bearded, ribbon, and spotted seals.
+
+While bio-logging of haul-out patterns clearly has utility for description of
+ice-associated seals' natural history and ecology, some studies have focused, as we do
+here, specifically on quantifying the probability of seals being hauled out as a means of
+accounting for seals in the water, and therefore missed, during aerial surveys. These
+studies typically use logistic regression-style analyses to estimate the time-specific
+probability of being hauled out base on 'wet/dry' data relayed by bio-loggers. In these
+models, haul-out probabilities were expressed as a function of predictive covariates, such
+as time of day, day-of-year, sex, age class, and environmental conditions (e.g.,
+[@reder2003pb], [@bengtson2004pb], [@bengtson2005pb], [@udevitz2009cjz], [@verhoef2014sm],
+and [@southwell2008]). However, sample sizes have often been low and insufficient to
+permit strong inference about life history and/or seasonal variation in haul-out
+probabilities. For instance, Bengtson and Cameron's [-@bengtson2004pb] study included 5
+adult and 2 juvenile crabeater seals, while Bengtson et al.'s [-@bengtson2005pb] study was
+based on 6 telemetered ringed seals in the Chukchi Sea. These studies were often further
+limited by logistical constraints on tagging fieldwork and the attachment duration or
+operational life of satellite tags. In this study, we address some of these limitations by
+deploying small bio-loggers designed for attachment on a portion of the study individuals.
+These devices are designed to collect data through the molt period (when those adhered to
+the hair would fall off) and, in some situations, provide multiple years of data.
 
@@
 Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
-on bearded, spotted, and ribbon seals was subset to include only records from 1 March to
+on bearded, spotted, and ribbon seals was filtered to include only records from 1 March to
 15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
@@
 this study, in cases where both tag types were deployed, percent dry observations from the
-flipper tag were preferred. Sex and age class were determined at the time of deployment by
-experts in the field. Field identification of age class can be uncertain in some
-situations. The most notable is reliably discerning subadults from adults. In the case of
-ribbon seals, pelage differences between subadults and adults are distinguishable, with
-subadults often having less defined ribbons. Bearded seal subadults will often have a
-spotted pattern in the pelage that is not seen in adults. Spotted seal pelage is
-monomorphic and can not be used to discern age class. Despite these challenges, we feel
-the age classifications used in this analysis are reliable. Seals determined to be less
-than one year were classified as young-of-the-year. Subadults are those seals likely
-greater than one year of age but not yet sexually mature. Adults are likely sexually
-mature and older than approximately four years. For those tags deployed on
-young-of-the-year and transmitting into the next year, the age class was advanced to
-subadult on 1 March of the following year. Table \@ref(tab:tblDeploy) provides a summary
-of these deployments and data received from them.
+flipper tag were preferred. Sex and age class (non-dependent *young-of-the-year*, sexually
+immature *subadults*, and mature *adults*) were determined at the time of deployment by
+subjective judgement based on length, and pelage characteristics for some species. Field
+identification of age class can be inexact, particularly when discerning subadults from
+adults. In the case of ribbon seals, pelage differences between subadults and adults are
+distinguishable, with subadults often having less defined ribbons. Bearded seal subadults
+will often have a spotted pattern in the pelage that is not seen in adults. Spotted seal
+pelage is monomorphic and can not be used to discern age class. Despite these challenges,
+we feel the age classifications used in this analysis are useful categories in testing for
+age-related effects on haul-out behavior. Seals determined to be less than one year were
+classified as young-of-the-year. Subadults are those seals likely greater than one year of
+age but not yet sexually mature. Adults are likely sexually mature and older than
+approximately four years. For those tags deployed on young-of-the-year and transmitting
+into the next year, the age class was advanced to subadult on 1 March of the following
+year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
+from them.
 
@@
                 dayofwk = lubridate::wday(haulout_dt, label = TRUE, abbr = TRUE) %>% 
-                  forcats::fct_rev()
+                  forcats::fct_relevel("Sun","Mon","Tue","Wed","Thu","Fri","Sat")
   ) %>% 
@@
 to satellites, and we did not want to include data from these tags in our analyses. As
-such, start and end times of each deployment were identified by experts who examined
-several simultaneous data streams (e.g., tag locations and dive behavior) to determine
-when tags ceased providing data consistent with seal behavior. Sensor readings outside of
-these start and end times were discarded prior to analysis. After approximately 9 months,
-7 devices deployed on the rear flipper of bearded seals reported implausible sensor data.
+such, end times of each deployment were identified by experts who examined several
+simultaneous data streams (e.g., tag locations and dive behavior) to determine when tags
+ceased providing data consistent with seal behavior. Sensor readings outside of these
+start and end times were discarded prior to analysis. After approximately 9 months, 7
+devices deployed on the rear flipper of bearded seals reported implausible sensor data.
 All data after the first instance of unrealistic values were censored from this analysis.
@@
   geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
-  scale_size(range = c(1,4.35)) +
+  scale_size(range = c(1,4.3), breaks = c(100,500,1000,1500)) +
   scale_color_gradient(low = "#6baed6", high = "#08306b") +
@@
                      labels = c("April", "May", "June", "July")) +
-  #scale_y_discrete(labels = c("Sat","","","Wed","","","Sun"))+
+  scale_y_discrete(limits = rev) +
   guides(color = "none",
@@
   theme(
-        axis.text.y = element_blank(),
+        #axis.text.y = element_blank(),
         panel.grid = element_blank(),
         legend.position = "bottom") +
-  labs(x = "", y = "day of week", color = "seal-hours",
+  labs(x = "", y = "", color = "seal-hours",
        size = "seal hours of bio-logger data") +
@@
 (Figure \@ref(fig:examplePlot)). Each data transmission message consisted of a complete
-24-hour record. Tags were deployed and programmed in a manner to maximize data
+24-hour record. Tags were, generally, deployed and programmed in a manner to maximize data
 transmission during the spring pupping and molting period, however, not all days during
@@
 
-```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, May, and June.
-Empty areas of the plot represent missing data that were not successfully transmitted from the tag."}
+```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, and May.
+Empty areas of the plot within a month represent missing data that were not successfully transmitted from the tag."}
 d <- dat.sf %>% 
@@
          day = lubridate::day(haulout_dt),
-         hour = lubridate::hour(haulout_dt))
+         hour = lubridate::hour(haulout_dt)) %>% 
+  filter(month != 'Jun')
 
@@
     )
-  # scale_fill_viridis_c(name = "hourly percent dry", aesthetics = "fill",
-  #   guide = guide_colorbar(title.position = "bottom", barwidth = 15,
-  #                          barheight = 0.5, title.hjust = 0.5)
-  # )
-p <- p + facet_grid(year~month)
-#p <- p + scale_y_continuous(trans = "reverse")
-p <- p + scale_x_continuous(breaks = c(1,10,20,30))
+
+p <- p + facet_grid(year~month, scales = "free_x")
+#p <- p + scale_y_continuous(trans = "reverse", expand = c(0,0))
+p <- p + scale_x_continuous(breaks = c(5,15,25), expand = c(0,0)) +
+  scale_y_continuous(breaks = c(4,12,20))
 p <- p + theme(legend.position = "bottom") +
@@
   theme(axis.ticks=element_blank()) +
+  theme(panel.spacing = unit(0,"line")) +
   ggtitle("Observation records are incomplete because not all data are transmitted",
@@
 
-```{r wxCovMap, include=TRUE, fig.height = 5.5, fig.id="wxCovMap", fig.cap="Example spatial distribution of weather covariates across the study region for 05 May 2012.
+```{r wxCovMap, fig.height = 5.5, fig.id="wxCovMap", fig.cap="Example spatial distribution of weather covariates across the study region for 05 May 2012.
 The NARR was used to predict weather covariates for each hour in a percent dry timeline record."}
@@
 [-@crawford2019] compared haul-out probability models for bearded seals and found those
-that only included season (and not sea-ice concentration) where the most parsimonious. 
-For these reasons, we have elected not to use sea ice concentration as a predictor for
+that only included season (and not sea-ice concentration) were the most parsimonious. For
+these reasons, we have elected not to use sea ice concentration as a predictor for
 haul-out probability; studies to evaluate habitat selection in relation to sea-ice
@@
 Models omitting year effects suggested that day-of-year, solar hour, age-sex class,
-temperature, and wind substantially altered haul-out behavior of all three species, with
-$F$ tests producing $p$-values less than 0.05 for variables embodying these effects and/or
-their interactions. Haul-out probabilities typically increased throughout March and April,
-reaching a peak in May and early June before declining again. Diurnal patterns were
+temperature, and wind substantially influenced haul-out behavior of all three species,
+with $F$ tests producing $p$-values less than 0.05 for variables embodying these effects
+and/or their interactions. Haul-out probabilities typically increased throughout March and
+April, reaching a peak in May and early June before declining again. Diurnal patterns were
 present, with maximum haul-out behavior centered around local solar noon.
@@
 Ribbon seals show a pattern of gradually increasing haul-out probability in April that
-peaks in May for subadults and in June for adults. The behavior is clearly centered around
-local solar noon and expands to other hours later in the season as seals enter their
-molting period. Subadults showed an earlier start and more intense haul-out activity in
-April and May. The young-of-the-year records begin after weaning and the model predictions
-demonstrate the ontogeny of in-water activities (e.g. diving, foraging) in May. Adult
-females have a more protracted season compared to males and more time spent resting in
-June and July. The model suggests adult male ribbon seals complete their molt by the end
-of June. Lastly, for adults of both sexes there is some indication for a shift to a
-nocturnal haul-out pattern in late June and July.
-
-```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOCal", fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model."}
+peaks in late May for subadults and in early June for adults (Figure
+\@ref(fig:ribbonHOCal)). The behavior is clearly centered around local solar noon and
+expands to other hours later in the season as seals enter their molting period. Subadults
+showed an earlier start and more intense haul-out activity in April and May. The
+young-of-the-year records begin after weaning and the model predictions demonstrate the
+ontogeny of in-water activities (e.g. diving, foraging) in May. Adult females have a more
+protracted haul-out season compared to males, and more time spent resting in June and
+July. The model suggests adult male ribbon seals complete their molt by the end of June.
+Lastly, for adults of both sexes there is some indication for a shift to a crepuscular
+haul-out pattern in late June and July.
+
+```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOCal", fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model.
+Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males.
+'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}
 plot_df <- ribbon_newdata %>%
@@
 p <- p + facet_grid(age_sex~month)
-p <- p + scale_x_continuous(breaks = c(1,10,20,30))
+p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
+  scale_y_continuous(breaks = c(4,12,20))
 p <- p + theme(legend.position = "bottom") +
@@
 
-```{r}
+```{r type3}
 ribbon_type3 <- tibble(
@@
 
-```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Variability in predicted haul-out probability of ribbon seals across a range of weather conditions the range of weather conditions encountered in the observed data.
+Visualizing the marginal effect of each weather covariate on haul-out probability is difficult
+in this analysis because of the collinearity between covariates as well as the temporal variation 
+in covariate values during the study period. The relationship of each weather covariate with haul-out
+probability, averaged over the other weather conditions, is more variable than simply
+showing model coefficients would imply. Figure \@ref(fig:ribbonHOwx)) presents the
+predicted haul-out probability of ribbon seals across the range of weather conditions
+encountered in the observed data. The collinearity and temporal variation are apparent
+in the visual patterns that emerge. By
+
+```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Variability in predicted haul-out probability of ribbon seals across the range of weather conditions encountered in the observed data.
 Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
@@
 Compared to ribbon seals, spotted seals showed a longer spring haul-out season that is
-less intensely centered on solar noon. Adults of both sexes spend considerable time in
+less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)). Adults of both sexes spend considerable time in
 April and May resting on the ice. As with ribbon seals, the young-of-the-year records
@@
 
-```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.id = "spottedHOCal", fig.cap="Predicted haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model."}
+```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.id = "spottedHOCal", fig.cap="Predicted haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model.
+Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males.
+'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}
 plot_df <- spotted_newdata %>%
@@
 
-Spotted seal haul-out behavior appears most strongly influenced by barometric pressure
+Spotted seal haul-out behavior appears most strongly influenced by temperature
+($F_{`r paste(spotted_type3$temp2$Num.df, spotted_type3$temp2$Den.df, sep=",")`}$ =
+`r spotted_type3$temp2$F.value`; $p$ = `r spotted_type3$temp2$Prob.F`) and wind
+($F_{`r paste(spotted_type3$wind$Num.df, spotted_type3$wind$Den.df, sep=",")`}$ =
+`r spotted_type3$wind$F.value`; $p$ = `r spotted_type3$wind$Prob.F`) with barometric
+pressure having a moderate influence
 ($F_{`r paste(spotted_type3$pressure$Num.df, spotted_type3$pressure$Den.df, sep=",")`}$ =
-`r spotted_type3$pressure$F.value`; $p$ = `r spotted_type3$pressure$Prob.F`) and wind
-($F_{`r paste(spotted_type3$wind$Num.df, spotted_type3$wind$Den.df, sep=",")`}$ =
-`r spotted_type3$wind$F.value`; $p$ = `r spotted_type3$wind$Prob.F`) with temperature
-having a moderate influence
-($F_{`r paste(spotted_type3$temp2$Num.df, spotted_type3$temp2$Den.df, sep=",")`}$ =
-`r spotted_type3$temp2$F.value`; $p$ = `r spotted_type3$temp2$Prob.F`). Spotted seals were
-less likely to haul out at higher winds and more likely to be on the ice when temperatures
-were relatively warm. Wind chill (*temperature:wind*) and precipitation were not as
-influential. Differences in the magnitude of response between the age-sex classes is
-present and consistent across each of the weather covariates.
-
-```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Variability in the predicted haul-out probability of spotted seals across a range of weather conditions the range of weather conditions encountered in the observed data.
+`r spotted_type3$pressure$F.value`; $p$ = `r spotted_type3$pressure$Prob.F`). Spotted
+seals were less likely to haul out at higher winds and more likely to be on the ice when
+temperatures were relatively warm. Wind chill (*temperature:wind*) and precipitation were
+not as influential as the other covariates. Differences in the magnitude of response
+between the age-sex classes are present and consistent across each of the weather
+covariates (Figure \@ref(fig:spottedHOwx)).
+
+```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Variability in the predicted haul-out probability of spotted seals across the range of weather conditions encountered in the observed data.
 Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
@@
 are shown for all ages. Unlike ribbon and spotted seals, bearded seals were predicted to
-have a bi-modal distribution of haul-out probability across the day. In addition to a peak
+have a bi-modal distribution of haul-out probability across the day (Figure \@ref(fig:beardedHOCal)). In addition to a peak
 around local solar noon, the bearded seal model predicts additional haul-out activity
@@
 
-```{r beardedHOCal, fig.height = 3.5, include=TRUE, fig.id = "beardedHOCal", fig.cap="Predicted haul-out probability of bearded seals from 1 March to 15 July for each age and sex class used in the model."}
+```{r beardedHOCal, fig.height = 3.5, include=TRUE, fig.id = "beardedHOCal", fig.cap="Predicted haul-out probability of bearded seals from 1 March to 15 July for all and sex classes."}
 plot_df <- bearded_newdata %>%
@@
 = `r bearded_type3$temp2_wind$F.value`; $p$ = `r bearded_type3$temp2_wind$Prob.F`). Any
-influence of barometric pressure or precipitation is less apparent.
-
-```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across a range of weather conditions the range of weather conditions encountered in the observed data.
+influence of barometric pressure or precipitation is less apparent (Figure \@ref(fig:beardedHOwx)).
+
+```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across the range of weather conditions encountered in the observed data.
 Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
@@
   )
-
-```
-
-```{r}
+```
+
+```{r type3_year}
 ribbon_year_type3 <- tibble(
@@
                   shape = 15) +
-  scale_color_brewer(palette = "RdYlBu") +
+  scale_color_brewer(palette = "Paired") +
   scale_x_continuous(breaks = c(74,105,135,166),
@@
 
-```{r maxextentHO, fig.height = 5.5, include=TRUE, fig.id = "maxextent", fig.cap="Annual timing of peak haul-out behavior analyzed against the yearly maximum area of sea ice extent in the study area.
+```{r maxextentHO, fig.height = 5.5, include=FALSE, fig.id = "maxextent", fig.cap="Annual timing of peak haul-out behavior analyzed against the yearly maximum area of sea ice extent in the study area.
 While spotted seal adult females and subadults show a negative trend line, none of the relationships were determined to be significant."}
@@
 ribbon seals to examine factors affecting haul-out behavior on sea ice in the Bering and
-Chukchi seas. Our analysis shows seals use sea ice progressively more throughout the
+Chukchi seas. Our analysis shows seals use sea ice progressively more through the
 spring, with peak use near mid-May to early June before declining again. This pattern
@@
 distributed across higher latitudes and the extended daylight hours may allow more
-flexibility in optimizing resting periods with foraging. Notably, bearded and ringed seals
-are subjected to much higher levels of predation from polar bears and human harvest.
+flexibility in optimizing resting periods with foraging.
 
@@
 accompanied on the ice by a suitor male. The male waits for the female to wean the pup and
-enter estrus and fends off any other males that may be interested. Triad formation results
+enter estrus, and fends off any other males that may be interested. Triad formation results
 in both males and females spending a large portion of the day hauled out on ice and a
-protracted spring haul-out season for both sexes. Since breeding occurs in the water,
+protracted spring haul-out season for both sexes. Because mating occurs in the water,
 females may also be less inclined to venture out for foraging trips while still nursing
@@
 than males. Notably, female ribbon seals spend a large portion of the day in the water
-during the pupping period and aligns with the hypothesis that ribbon seal females continue
+during the pupping period, aligning with the hypothesis that ribbon seal females continue
 foraging while nursing. In the case of both ribbon and spotted seals, subadults are the
-first to begin meaningful haul-out behavior and reflect a typical phocid pattern where
+first to begin consistent haul-out behavior and follow a typical phocid pattern where
 subadults molt first as they do not have any reproductive constraints.
@@
 wind and temperature had the greatest impact. In general, and as might be expected, seals
-were more likely to haul-out when daily temperatures were warmer, winds speeds were less,
-barometric pressure was high, and rates of precipitation were. These weather conditions
+were more likely to haul-out when daily temperatures were warmer, winds speeds were lower,
+barometric pressure was higher, and precipitation was lower. These weather conditions
 are likely most beneficial for energetic considerations and may also provide the best
@@
 
-::: {custom-style="comment"}
-this could use a better ending with more punch and interest
-:::
-
 # Author Contributions
@@
 validation, software, writing: original draft, writing: review and editing, visualization,
-and data curation **Paul B. Conn**: conceptualization, methodology, formal analysis,
-software, validation, writing: original draft, writing: review and editing **Stacie K.
-Hardy**: investigation, data curation, methodology, validation, writing: review and
-editing **Erin L. Richmond**: data curation, investigation, methodology, validation,
-writing: review and editing **Jay M. Ver Hoef**: conceptualization, methodology, software,
-writing: review and editing **Irina S. Trukhanova**: conceptualization, writing: review
-and editing **Michael F. Cameron**: investigation, project administration, writing: review
-and editing **Peter L. Boveng**: investigation, conceptualization, supervision, project
+and data curation
+
+**Paul B. Conn**: conceptualization, methodology, formal analysis,
+software, validation, writing: original draft, writing: review and editing
+
+**Stacie K. Hardy**: investigation, data curation, methodology, validation, writing: review and
+editing
+
+**Erin L. Richmond**: data curation, investigation, methodology, validation,
+writing: review and editing
+
+**Jay M. Ver Hoef**: conceptualization, methodology, software,
+writing: review and editing
+
+**Irina S. Trukhanova**: conceptualization, writing: review
+and editing
+
+**Michael F. Cameron**: investigation, project administration, writing: review
+and editing
+
+**Peter L. Boveng**: investigation, conceptualization, supervision, project
 administration, writing: review and editing
@@
 
-Data and code are available as a complete research compendium on GitHub and major versions
-are archived at Zenodo. Original data sources for telemetry are archived at the United
+Data and code will be available as a complete research compendium on GitHub and major versions
+archived at Zenodo. Original data sources for telemetry are archived at the United
 States Animal Telemetry Network.
@@
 Department of Wildlife Management were important contributions to the findings presented
-here. Deployments in the western Bering Sea and Sea of Okhotsk were done in collaboration
-with Russian colleagues and North Pacific Wildlife.
+here. Both agencies collaborate closely with Alaska Native communities and the
+contributions of local hunters to the successful capture, deployment, and understanding of seal ecology are
+invaluable. Deployments in the western Bering Sea and Sea of Okhotsk were done in
+collaboration with Russian colleagues and North Pacific Wildlife.
 

### 2021-08-14 analysis/paper/paper.Rmd
IMPLEMENT {targets} PIPELINE; UPDATE DATA ETL w/ NEW ADFG RECORDS
major revamp of compendium and code inspired by updated records from ADFG. implemented the {targets} package workflow.

@@
 ```{r setup, include = FALSE}
+library(targets)
 library(conflicted)
@@
 library(dbplyr)
+library(RPostgres)
+library(pingr)
 library(sf)
@@
 ```{r getdata}
-con <- dbConnect(
-  odbc::odbc(),
-  dsn = "PostgreSQL pep",
-  uid = keyringr::get_kc_account("pgpep_londonj"),
-  pwd = keyringr::decrypt_kc_pw("pgpep_londonj"),
-  sslmode = "disable"
-)
-
-# qry removes two Sea of Okhotsk bearded seals
-qry <- {
-  "SELECT * 
-  FROM telem.res_iceseal_haulout_cov 
-  WHERE 
-  EXTRACT(MONTH FROM haulout_dt) IN (3,4,5,6,7) AND
-  rast_vwnd IS NOT NULL AND
-  species != 'Ph' AND
-  speno != 'EB2009_7010' AND
-  speno != 'EB2009_3002' "
-}
-
-dat.sf <- sf::st_read(con,query = qry) %>% 
-  filter(lubridate::yday(haulout_dt) < 197 ) %>% 
-  mutate(age = case_when(
-    age %in% c("pup", "PUP") ~ "YOUNG OF YEAR",
-    age %in% c("yearling","YEARLING","subadult","Subadult") ~ "SUBADULT",
-    age %in% c("adult","ADULT","Adult") ~ "ADULT",
-    TRUE ~ age),
-    species = case_when(
-    species == "bearded" ~ "Bearded seal",
-    species == "spotted" ~ "Spotted seal",
-    TRUE ~ species),
-    sex = case_when(
-    sex %in% c("Female","female") ~ "F",
-    sex %in% c("Male","male") ~ "M",
-    TRUE ~ sex)
-    ) %>% 
-  filter(!is.na(age))
-
-load(here::here('data/locs_sf.Rdata'))
-
-deploy_meta <- locs_sf %>% 
-  as_data_frame() %>% 
-  select(speno, deploy_dt) %>% 
-  filter(speno %in% !!unique(dat.sf$speno)) %>% 
-  group_by(speno) %>% 
-  summarise(deploy_dt = min(deploy_dt, na.rm = TRUE))
-
-dat.sf <- dat.sf %>% left_join(deploy_meta) %>% 
-  mutate(bday = as.Date(paste0(lubridate::year(deploy_dt),"-03-01")),
-         age_days = difftime(haulout_dt, bday, tz="UTC", units = "days"),
-         age = case_when(
-           age_days > 360 & age == "YOUNG OF YEAR" ~ "SUBADULT",
-           TRUE ~ age
-         )) %>% 
-  mutate(age = forcats::fct_relevel(age, c("ADULT", "SUBADULT", "YOUNG OF YEAR"))) %>% 
-  dplyr::select(-c(bday,age_days)) %>% 
-  relocate(deploy_dt, .after = age) %>% 
-  filter(!(species == "Bearded seal" & age == "YOUNG OF YEAR"))
-
-# read in world land polygons and buffer in by 100m; want to ensure only locations well
-world <- ne_countries(scale = "large", returnclass = "sf") %>% 
-  filter(continent != "Antarctica") %>% 
-  st_wrap_dateline() %>% st_transform(st_crs(dat.sf)) %>% 
-  st_buffer(-100)
-
-not_intersects <- function(x,y) {
-  !sf::st_intersects(x,y)
-}
-
-dat.sf <- dat.sf %>% 
-  sf::st_filter(st_combine(world), .predicate = not_intersects)
-
-save("dat.sf", file = here::here("data/dat_sf.Rdata"))
-
-grid.sf <- sf::st_read(con,
-                       query = "SELECT * FROM base.geo_analysis_grid") %>% 
-  st_filter(st_bbox(dat.sf) %>% st_as_sfc())
-
-dbDisconnect(con)
-
-deploy_tbl <- dat.sf %>% 
-  tibble::as_tibble() %>% 
-  dplyr::group_by(speno,species, sex, age) %>% 
-  dplyr::summarise(min_dt = min(haulout_dt),
-            max_dt = max(haulout_dt),
-            n_days = difftime(max_dt, min_dt, units = "days"),
-            n_hours = n()
-  ) %>% 
-  dplyr::select(-c("min_dt","max_dt"))
-
-```
-
-Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
+targets::tar_load(analysis_data, store = here::here("_targets"))
+targets::tar_load(grid, store = here::here("_targets"))
+targets::tar_load(deploy_table, store = here::here("_targets"))
+```
+
+Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)` bio-loggers deployed
 on bearded, spotted, and ribbon seals was filtered to include only records from 1 March to
-15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
-`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
+15 July between `r min(lubridate::year(analysis_data$haulout_dt))` and
+`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
 family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
@@
 
-deploy_tbl %>% 
+deploy_table %>% 
   dplyr::group_by(species, sex, age) %>% 
@@
 ```{r seal-hours}
-seal_hours_tbl <- dat.sf %>% 
+seal_hours_tbl <- analysis_data %>% 
   tibble::as_tibble() %>% 
@@
 Empty areas of the plot within a month represent missing data that were not successfully transmitted from the tag."}
-d <- dat.sf %>% 
+d <- analysis_data %>% 
   filter(speno == "HF2016_1002") %>% 
@@
 ```{r locs-count-mapdata}
-haulout_locs <- dat.sf %>% 
+haulout_locs <- analysis_data %>% 
   filter(lubridate::hour(haulout_dt) == 12)

### 2021-08-14 R/adfg_clean.R

@@ -1,0 +1,185 @@
+adfg_clean_deploy <- function(file1,file2) {
+  file1 <- fs::path(file1)
+  file2 <- fs::path(file2)
+
+ cols <- readr::cols(
+    DeployIDs = readr::col_character(),
+    MapID = readr::col_character(),
+    Species = readr::col_character(),
+    Age = readr::col_character(),
+    Sex = readr::col_character(),
+    PrimaryTagType = readr::col_character(),
+    FlipperTagType = readr::col_character()
+  )
+
+  adfg_file1 <- readr::read_csv(file1,col_types = cols) %>%
+    dplyr::mutate(DateTagged = parse_datetime(DateTagged,format = "%m/%d/%Y %H:%M")) %>%
+    dplyr::mutate(DateTagged = lubridate::force_tz(DateTagged,
+                                                   'America/Anchorage')) %>%
+    dplyr::mutate(DateTagged = lubridate::as_datetime(DateTagged,
+                                                      tz = 'America/Anchorage')) %>%
+    dplyr::filter( ! (DeployIDs %in% c("EB16SMK-158429")) )
+
+  cols <- readr::cols(
+    DeployIDs = readr::col_character(),
+    Species = readr::col_character(),
+    Age = readr::col_character(),
+    Sex = readr::col_character(),
+    FlipperTagType = readr::col_character()
+  )
+
+  adfg_file2 <- readr::read_csv(file2,col_types = cols) %>%
+    dplyr::mutate(DateTagged = parse_datetime(DateTagged,format = "%m/%d/%y")) %>%
+    dplyr::mutate(DateTagged = lubridate::force_tz(DateTagged,
+                                                   'America/Anchorage')) %>%
+    dplyr::mutate(DateTagged = lubridate::as_datetime(DateTagged,
+                                                      tz = 'America/Anchorage'))
+
+  adfg_deployments <- adfg_file1 %>%
+    dplyr::bind_rows(adfg_file2) %>%
+    dplyr::rename(speno = DeployIDs,
+                  mapid = MapID,
+                  deploy_dt = DateTagged,
+                  primary_tag_type = PrimaryTagType,
+                  flipper_tag_type = FlipperTagType,
+                  flipper_last_signal = FlipperSatTagLastSignal
+    ) %>%
+    dplyr::rename_all(tolower) %>%
[…]

### 2021-08-14 R/create_data_sf.R

@@ -1,0 +1,52 @@
+create_data_sf <- function(locs_sf, source_data) {
+  dat.sf <- source_data %>%
+    filter(lubridate::yday(haulout_dt) < 197 ) %>%
+    mutate(age = case_when(
+      age %in% c("pup", "PUP", "p") ~ "YOUNG OF YEAR",
+      age %in% c("yearling","YEARLING","subadult","Subadult") ~ "SUBADULT",
+      age %in% c("adult","ADULT","Adult") ~ "ADULT",
+      TRUE ~ age),
+      species = case_when(
+        species == "bearded" ~ "Bearded seal",
+        species == "spotted" ~ "Spotted seal",
+        TRUE ~ species),
+      sex = case_when(
+        sex %in% c("Female","female") ~ "F",
+        sex %in% c("Male","male","m") ~ "M",
+        TRUE ~ sex)
+    ) %>%
+    filter(!is.na(age))
+
+  deploy_meta <- locs_sf %>%
+    as_tibble() %>%
+    select(speno, deploy_dt) %>%
+    filter(speno %in% !!unique(dat.sf$speno)) %>%
+    group_by(speno) %>%
+    summarise(deploy_dt = min(deploy_dt, na.rm = TRUE))
+
+  dat.sf <- dat.sf %>% left_join(deploy_meta) %>%
+    mutate(bday = as.Date(paste0(lubridate::year(deploy_dt),"-03-01")),
+           age_days = difftime(haulout_dt, bday, tz="UTC", units = "days"),
+           age = case_when(
+             age_days > 360 & age == "YOUNG OF YEAR" ~ "SUBADULT",
+             TRUE ~ age
+           )) %>%
+    mutate(age = forcats::fct_relevel(age, c("ADULT", "SUBADULT", "YOUNG OF YEAR"))) %>%
+    dplyr::select(-c(bday,age_days)) %>%
+    relocate(deploy_dt, .after = age) %>%
+    filter(!(species == "Bearded seal" & age == "YOUNG OF YEAR"))
+
+  # read in world land polygons and buffer in by 100m; want to ensure only locations well
+  world <- ne_countries(scale = "large", returnclass = "sf") %>%
+    filter(continent != "Antarctica") %>%
+    st_wrap_dateline() %>% st_transform(st_crs(dat.sf)) %>%
+    st_buffer(-100)
+
+
+  not_intersects <- function(x,y) {
+    !sf::st_intersects(x,y)
[…]

### 2021-08-14 R/create_db_input_data.R

@@ -1,14 +1,4 @@
-##' .. content for \description{} (no empty lines) ..
-##'
-##' .. content for \details{} ..
-##'
-##' @title
-##' @param timeline_data
-##' @param locs_sf
-##' @return
-##' @author Josh.London
-##' @export
-create_db_input_data <- function(timeline_data, locs_sf) {
+create_source_data <- function(locs_sf, timeline_data) {
 
   loc_qual_tbl <- tribble(
     ~quality, ~error_radius,
@@ -21,7 +11,7 @@ create_db_input_data <- function(timeline_data, locs_sf) {
   )
 
   locs_daily <- locs_sf %>%
-    filter(quality != "Z") %>%
+    dplyr::filter(quality %in% c("3","2","1","0","A","B")) %>%
     left_join(loc_qual_tbl, by = "quality") %>%
     st_transform(3571) %>%
     mutate(x = st_coordinates(.)[,"X"],
@@ -54,9 +44,9 @@ create_db_input_data <- function(timeline_data, locs_sf) {
     group_by(speno) %>%
     fill(x,y) %>%
     ungroup() %>%
-    filter(!is.na(percent_dry)) %>%
-    filter(!speno %in% c("EB2005_5995","PL2006_5984","PL2006_5987")) %>%
-    filter(!is.na(x)) %>%
+    dplyr::filter(!is.na(percent_dry)) %>%
+    dplyr::filter(!speno %in% c("EB2005_5995","PL2006_5984","PL2006_5987")) %>%
+    dplyr::filter(!is.na(x)) %>%
     st_as_sf(coords = c("x","y")) %>%
     st_set_crs(3571) %>%
     rename(haulout_dt = timeline_start_dt) %>%
@@ -67,19 +57,33 @@ create_db_input_data <- function(timeline_data, locs_sf) {
       is_up("161.55.120.122", "5432")
   )
 
-  con <- dbConnect(
-    odbc(),
-    dsn = "PostgreSQL pep",
-    uid = keyringr::get_kc_account("pgpep_sa"),
-    pwd = keyringr::decrypt_kc_pw("pgpep_sa")
-  )
+  con <- dbConnect(RPostgres::Postgres(),dbname = 'pep',
+                   host = '161.55.120.122',
+                   port = 5432,
+                   user = keyringr::get_kc_account("pgpep_sa"),
[…]

### 2021-08-14 R/create_deploy_tbl.R

@@ -1,0 +1,11 @@
+create_deploy_tbl <- function(data_sf) {
+  deploy_tbl <- data_sf %>%
+    tibble::as_tibble() %>%
+    dplyr::group_by(speno,species, sex, age) %>%
+    dplyr::summarise(min_dt = min(haulout_dt),
+                     max_dt = max(haulout_dt),
+                     n_days = difftime(max_dt, min_dt, units = "days"),
+                     n_hours = n()
+    ) %>%
+    dplyr::select(-c("min_dt","max_dt"))
+}

### 2021-08-14 R/get_locs_sf.R

@@ -1,13 +1,4 @@
-##' .. content for \description{} (no empty lines) ..
-##'
-##' .. content for \details{} ..
-##'
-##' @title
-
-##' @return
-##' @author Josh.London
-##' @export
-get_locs_sf <- function() {
+get_locs_sf <- function(adfg_locations, nsb_locations) {
 
   stopifnot(
     "PEP Postgres Database Not Available; did you start VPN? ;)" =
@@ -31,6 +22,7 @@ get_locs_sf <- function() {
               FROM telem.geo_wc_locs_qa;"
 
   locs_sf <- read_sf(con, query = locs_qry) %>%
+    filter(!deployid %in% c("PL2017_9001_16U2112")) %>%
     left_join(deployments_db, by = 'deployid') %>%
     left_join(spenos_db, by = 'speno') %>%
     filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
@@ -40,9 +32,6 @@ get_locs_sf <- function() {
                   "{lubridate::yday(locs_dt)}",
                   .sep = "_"))
   dbDisconnect(con)
-
-  load(file = here::here('data/adfg_locations.rda'))
-  load(file = here::here('data/nsb_locations.rda'))
 
   locs_sf <- locs_sf %>% bind_rows(adfg_locations) %>%
     bind_rows(nsb_locations)

### 2021-08-14 R/get_nsb_data.R

@@ -1,0 +1,109 @@
+get_nsb_deployments <- function(con) {
+  stopifnot(
+    "PEP Postgres Database Not Available; did you start VPN? ;)" =
+      pingr::is_up("161.55.120.122", "5432")
+  )
+
+  con <- dbConnect(
+    odbc::odbc(),
+    dsn = "PostgreSQL pep",
+    uid = keyringr::get_kc_account("pgpep_londonj"),
+    pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
+  )
+  on.exit(odbc::dbDisconnect(con))
+
+  deploy_qry <- "select speno, deployid, common_name as species, age_class as age, sex, tag_family, deploy_dt, end_dt
+from telem.nsb_capture c
+left join telem.nsb_deploy d
+on c.id = d.capture_id
+left join telem.lku_tag_family
+using (tag_family_lku)
+left join capture.lku_species
+using (species_lku)
+left join capture.lku_age_class
+using (age_class_lku)
+left join capture.lku_sex
+using (sex_lku)"
+
+  nsb_deployments <- tbl(con, sql(deploy_qry)) %>%
+    dplyr::select(speno, deployid, species, age, sex, tag_family, deploy_dt, end_dt) %>%
+    collect()
+
+  return(nsb_deployments)
+}
+
+get_nsb_locs <- function(nsb_deployments) {
+  stopifnot(
+    "PEP Postgres Database Not Available; did you start VPN? ;)" =
+      pingr::is_up("161.55.120.122", "5432")
+  )
+
+  con <- dbConnect(
+    odbc::odbc(),
+    dsn = "PostgreSQL pep",
+    uid = keyringr::get_kc_account("pgpep_londonj"),
+    pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
+  )
+  on.exit(odbc::dbDisconnect(con))
+
+  locs_qry <- "SELECT deployid, loc_dt as locs_dt, argos_lc as quality, loc_geom as geometry
+              FROM telem.nsb_locs;"
+
+  nsb_locations <- read_sf(con, query = locs_qry) %>%
+    left_join(nsb_deployments, by = 'deployid') %>%
+    filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
+    filter(lubridate::month(locs_dt) %in% c(3,4,5,6,7)) %>%
+    rename(ptt = deployid) %>%
+    mutate(species_code = case_when(
+      species == "Bearded seal" ~ "EB",
[…]

### 2021-08-14 R/get_timeline_data.R

@@ -1,13 +1,4 @@
-##' .. content for \description{} (no empty lines) ..
-##'
-##' .. content for \details{} ..
-##'
-##' @title
-
-##' @return
-##' @author Josh.London
-##' @export
-get_timeline_data <- function() {
+get_timeline_data <- function(adfg_timelines, nsb_timelines) {
 
   stopifnot(
     "PEP Postgres Database Not Available; did you start VPN? ;)" =
@@ -30,6 +21,7 @@ get_timeline_data <- function() {
   timeline_data <- timeline_db  %>%
     left_join(deployments_db, by = 'deployid') %>%
     left_join(spenos_db, by = 'speno') %>%
+    filter(!deployid %in% c("PL2017_9001_16U2112")) %>%
     filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
     collect() %>%
     filter(lubridate::month(timeline_start_dt) %in% c(3,4,5,6,7)) %>%
@@ -37,9 +29,6 @@ get_timeline_data <- function() {
              glue::glue("{lubridate::year(timeline_start_dt)}",
                         "{lubridate::yday(timeline_start_dt)}",
                         .sep = "_"))
-
-  load(file = here::here('data/adfg_timelines.rda'))
-  load(file = here::here('data/nsb_timelines.rda'))
 
   timeline_data <- timeline_data %>%
     bind_rows(adfg_timelines) %>%

### 2021-08-14 R/run_wxcov_query.R

@@ -1,0 +1,24 @@
+get_data_wxcov <- function() {
+  con <- dbConnect(RPostgres::Postgres(),dbname = 'pep',
+                   host = '161.55.120.122',
+                   port = 5432,
+                   user = keyringr::get_kc_account("pgpep_sa"),
+                   password = keyringr::decrypt_kc_pw("pgpep_sa"))
+  on.exit(dbDisconnect(con))
+
+  dbExecute(con, "SELECT telem.fxn_iceseal_pred_idx();")
+  dbExecute(con, "SELECT telem.fxn_iceseal_haulout_cov();")
+
+  qry <- {
+    "SELECT *
+  FROM telem.res_iceseal_haulout_cov
+  WHERE
+  EXTRACT(MONTH FROM haulout_dt) IN (3,4,5,6,7) AND
+  rast_vwnd IS NOT NULL AND
+  species != 'Ph' AND
+  speno != 'EB2009_7010' AND
+  speno != 'EB2009_3002' "
+  }
+
+  sf::st_read(con, query = qry)
+}

### 2021-08-14 _targets.R

@@ -19,8 +19,14 @@ tar_option_set(packages = c(
   "sf",
   "lubridate",
   "odbc",
+  "RPostgres",
+  "DBI",
+  "dbplyr",
   "here",
-  "keyringr"
+  "keyringr",
+  "pingr",
+  "rnaturalearth",
+  "rnaturalearthdata"
   )
 )
 
@@ -28,5 +34,24 @@ tar_option_set(packages = c(
 list(
   tar_target(adfg_deploy_file1, "data_raw/adfg/ADFG_SealDeploymentTable3.csv",format = "file"),
   tar_target(adfg_deploy_file2, "data_raw/adfg/ADFG_SealDeploymentTable_NewRecords.csv", format = "file"),
-  tar_target(adfg_clean_deployments, adfg_clean_deploy(adfg_deploy_file1, adfg_deploy_file2))
+  tar_target(adfg_deployments, adfg_clean_deploy(adfg_deploy_file1, adfg_deploy_file2)),
+  tar_target(adfg_locs_file1, "data_raw/adfg/ADFG_SealLocationData3.csv",format = "file"),
+  tar_target(adfg_locs_file2, "data_raw/adfg/ADFG_SealLocationData_NewRecords.csv", format = "file"),
+  tar_target(adfg_locations, adfg_clean_locs(adfg_locs_file1, adfg_locs_file2, adfg_deployments)),
+  tar_target(adfg_timelines_file1, "data_raw/adfg/ADFG_SealTimelineData3.csv",format = "file"),
+  tar_target(adfg_timelines_file2, "data_raw/adfg/ADFG_SealTimelineData_NewRecords.csv", format = "file"),
+  tar_target(adfg_timelines, adfg_clean_tl(adfg_timelines_file1, adfg_timelines_file2, adfg_deployments)),
+
+  tar_target(nsb_deployments, get_nsb_deployments()),
+  tar_target(nsb_locations, get_nsb_locs(nsb_deployments)),
+  tar_target(nsb_timelines, get_nsb_timelines(nsb_deployments)),
+
+  tar_target(locs_sf, get_locs_sf(adfg_locations, nsb_locations)),
+  tar_target(timeline_data, get_timeline_data(adfg_timelines, nsb_timelines)),
+
+  tar_target(source_data, create_source_data(locs_sf, timeline_data)),
+
+  tar_target(analysis_data, create_data_sf(locs_sf, source_data)),
[…]

### 2021-10-28 analysis/paper/paper.Rmd
REVAMP PROJECT FOR {targets}
major revamp of everything to support a {targets} pipeline approach. Previous functionality and pipeline steps that were within Rmd / Scripts have been ported to specific functions within R/

@@
 
-```{r wxCovMap, fig.height = 5.5, fig.id="wxCovMap", fig.cap="Example spatial distribution of weather covariates across the study region for 05 May 2012.
-The NARR was used to predict weather covariates for each hour in a percent dry timeline record."}
-
-load(file = here::here('data/cov_means_daily_2012.Rdata'))
-grid_plot <- cov_means_daily_2012 %>% filter(year == 2012,
-                           yday == 125) %>%
-  inner_join(grid.sf, by = 'cell') %>%
-  sf::st_as_sf()
-
-world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
-  sf::st_transform(st_crs(grid_plot))
-
-p_precip <- ggplot() +
-  annotation_spatial(world, 
-                     size = 0, 
-                     fill = "grey70") +
-  layer_spatial(grid_plot, aes(fill = precip), size = 0) +
-  scale_fill_distiller(palette = "RdYlBu") +
-  # scale_fill_viridis_c(option = "C") +
-  labs(subtitle = "precipitation") +
-  theme(legend.position = "none")
-
-p_pressure <- ggplot() +
-  annotation_spatial(world, 
-                     size = 0, 
-                     fill = "grey70") +
-  layer_spatial(grid_plot, aes(fill = pressure), size = 0) +
-  scale_fill_distiller(palette = "RdYlBu") +
-  # scale_fill_viridis_c(option = "C") +
-  labs(subtitle = "barometric pressure") +
-  theme(legend.position = "none")
-
-p_wind <- ggplot() +
-  annotation_spatial(world, 
-                     size = 0, 
-                     fill = "grey70") +
-  layer_spatial(grid_plot, aes(fill = wind), size = 0) +
-  scale_fill_distiller(palette = "RdYlBu") +
-  # scale_fill_viridis_c(option = "C") +
-  labs(subtitle = "wind") +
-  theme(legend.position = "none")
-
-p_temp2 <- ggplot() +
-  annotation_spatial(world, 
-                     size = 0, 
-                     fill = "grey70") +
-  layer_spatial(grid_plot, aes(fill = temp2), size = 0) +
-  scale_fill_distiller(palette = "RdYlBu") +
-  # scale_fill_viridis_c(option = "C") +
-  labs(subtitle = "temperature") +
-  theme(legend.position = "none")
-
-(p_precip + p_pressure) / (p_wind + p_temp2) +
-  plot_annotation(
-    title = 'Weather varies across the Bering and Chukchi seas',
-    caption = "North American Regional Reanalysis (NARR), 2012 May 05")
-```
-
 Notably missing from this list of explanatory variables is any spatial-temporal
@@
 
-```{r runmodels}
-if (params$run_models) {
-source(here::here('analysis/scripts/model.R'))
-} else {
-  load(here::here('data/fit_ribbon.Rdata'))
-  load(here::here('data/fit_bearded.Rdata'))
-  load(here::here('data/fit_spotted.Rdata'))
-}
+```{r modelfits}
+tar_load(bearded_fit, store = here::here("_targets"))
+tar_load(ribbon_fit, store = here::here("_targets"))
+tar_load(spotted_fit, store = here::here("_targets"))
+tar_load(spotted_year_fit, store = here::here("_targets"))
+tar_load(ribbon_year_fit, store = here::here("_targets"))
 ```
@@
 
-```{r runyearmodels}
-if (params$run_models) {
- source(here::here('analysis/scripts/model_year_hf.R'))
- source(here::here('analysis/scripts/model_year_pl.R'))
-
-} else {
- load(here::here('data/fit_ribbon_year.Rdata'))
- load(here::here('data/fit_spotted_year.Rdata'))
-}
-```
-
 A typical model fitting exercise would also include a model selection process. However,
@@
 
-```{r createnewdata}
-# function to create a prediction data frame for a given age_sex group
-create_newdata <- function(data, age_sex) {
-  df_list <- vector(mode = "list", length = length({{age_sex}} ))
-
-  for (a_s in {{age_sex}}) {
-  range_yday <- data %>% filter(age_sex == a_s ) %>%
-    summarize(start_day = min(yday),
-              end_day = max(yday))
-  start_day <- range_yday$start_day
-  end_day <- range_yday$end_day
-  n_days = (end_day - start_day) + 1
-
-  # for wx covariates we'll use a gam to get values by day/hour since wx is likely
-  # to vary w/in day over the season
-
-  gam.baro <-
-    gam(pressure ~ s(yday), data = data)
-  gam.temp <-
-    gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), data = data)
-  gam.wind <-
-    gam(wind ~ s(yday) + s(as.numeric(solar_hour)), data = data)
-  gam.precip <-
-    gam(precip ~ s(yday), data = data)
-
-  wx_new_data <- data.frame(
-    solar_hour = rep(0:23, each = n_days),
-    yday = rep(start_day:end_day, times = 24)
-  )
-
-  temp_pred <- predict(gam.temp, newdata = wx_new_data)
-  wind_pred <- predict(gam.wind, newdata = wx_new_data)
-  baro_pred <- predict(gam.baro, newdata = wx_new_data)
-  precip_pred <- predict(gam.precip, newdata = wx_new_data)
-
-  df <- data.frame(
-    age_sex = a_s,
-    solar_hour = rep(0:23, each = n_days),
-    yday = rep(start_day:end_day, times = 24),
-    northing = mean(data$northing),
-    temp2 = temp_pred,
-    wind = wind_pred,
-    pressure = baro_pred,
-    precip = precip_pred
-  ) %>%
-    mutate(
-      sin1 = sin(pi * solar_hour / 12),
-      cos1 = cos(pi * solar_hour / 12),
-      sin2 = sin(pi * solar_hour / 6),
-      cos2 = cos(pi * solar_hour / 6),
-      sin3 = sin(pi * solar_hour / 4),
-      cos3 = sin(pi * solar_hour / 4),
-    ) %>%
-    mutate(day = (yday - 120) / 10,
-           day2 = day ^ 2,
-           day3 = day ^ 3)
-  df_list[[a_s]] <- df
-  }
-  if(length({{age_sex}}) > 1) {
-  df_out <- bind_rows(df_list) %>%
-    mutate(age_sex = forcats::fct_relevel(
-      age_sex,c("ADULT.F","ADULT.M","SUBADULT","YOUNG OF YEAR"))
-    )
-  } else {
-    df_out <- bind_rows(df_list)
-  }
-}
-```
-
 Models omitting year effects suggested that day-of-year, solar hour, age-sex class,
@@
 ```{r hfnewdata}
-# get the ribbon seal fit object
-load(here::here("data/fit_ribbon.Rdata"))
-ribbon_newdata <- create_newdata(fit_ribbon$dataset,
-                                 age_sex = levels(fit_ribbon$dataset$age_sex))
-
-# create the model matrix
-ribbon_mm <- model.matrix(fit_ribbon$fixed.formula[-2],
-                data = ribbon_newdata)
-
-#clean up extra intercept and get coef
-fit_coef <- fit_ribbon$fixed.effects %>%
-  filter(!is.na(std.err)) %>%
-  pull(estimate)
-
-ribbon_newdata <- ribbon_newdata %>%
-  mutate(logit_fits = as.vector(ribbon_mm %*% fit_coef),
-         logit_fits_var = diag(ribbon_mm %*% fit_ribbon$covb %*% t(ribbon_mm)),
-         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
-         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
-         ho_prob = plogis(ribbon_mm %*% fit_coef),
-         lower95 = plogis(logit_fits_lo95),
-         upper95 = plogis(logit_fits_up95)
-         )
+targets::tar_load(ribbon_newdata, store = here::here("_targets"))
 ```
@@
 ribbon_type3 <- tibble(
-  fit_ribbon$typeIII.hypoth
+  ribbon_fit$typeIII.hypoth
 ) %>% 
@@
 spotted_type3 <- tibble(
-  fit_spotted$typeIII.hypoth
+  spotted_fit$typeIII.hypoth
 ) %>% 
@@
 bearded_type3 <- tibble(
-  fit_bearded$typeIII.hypoth
+  bearded_fit$typeIII.hypoth
 ) %>% 
@@
 ```{r plnewdata}
-# get the spotted seal fit object
-load(here::here("data/fit_spotted.Rdata"))
-
-spotted_newdata <- create_newdata(fit_spotted$dataset,
-                                 age_sex = levels(fit_spotted$dataset$age_sex))
-
-# create the model matrix
-spotted_mm <- model.matrix(fit_spotted$fixed.formula[-2],
-                          data = spotted_newdata)
-
-#clean up extra intercept and get coef
-fit_coef <- fit_spotted$fixed.effects %>%
-  filter(!is.na(std.err)) %>%
-  pull(estimate)
-
-spotted_newdata <- spotted_newdata %>%
-  mutate(logit_fits = as.vector(spotted_mm %*% fit_coef),
-         logit_fits_var = diag(spotted_mm %*% fit_spotted$covb %*% t(spotted_mm)),
-         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
-         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
-         ho_prob = plogis(spotted_mm %*% fit_coef),
-         lower95 = plogis(logit_fits_lo95),
-         upper95 = plogis(logit_fits_up95)
-  )
+targets::tar_load(spotted_newdata, store = here::here("_targets"))
 ```
@@
 ```{r ebnewdata}
-# get the bearded seal fit object
-load(here::here("data/fit_bearded.Rdata"))
-
-eb_data <- fit_bearded$dataset %>% 
-  mutate(age_sex = forcats::as_factor("ALL AGES"))
-
-bearded_newdata <- create_newdata(eb_data,
-                                  age_sex = levels(eb_data$age_sex))
-
-# create the model matrix
-bearded_mm <- model.matrix(fit_bearded$fixed.formula[-2],
-                           data = bearded_newdata)
-
-#clean up extra intercept and get coef
-fit_coef <- fit_bearded$fixed.effects %>%
-  filter(!is.na(std.err)) %>%
-  pull(estimate)
-
-bearded_newdata <- bearded_newdata %>%
-  mutate(logit_fits = as.vector(bearded_mm %*% fit_coef),
-         logit_fits_var = diag(bearded_mm %*% fit_bearded$covb %*% t(bearded_mm)),
-         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
-         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
-         ho_prob = plogis(bearded_mm %*% fit_coef),
-         lower95 = plogis(logit_fits_lo95),
-         upper95 = plogis(logit_fits_up95)
-  )
+targets::tar_load(bearded_newdata, store = here::here("_targets"))
 ```
@@
 
-```{r createnewdatayr}
-create_newdata_year <- function(data, age_sex) {
-  df_list <- vector(mode = "list", length = length({{age_sex}} ))
-  df_list_years <- vector(mode = "list", length = length(levels(data$year)))
-
-  for (a_s in {{age_sex}}) {
-  for (y in levels(data$year)) {
-    d <- data %>% filter(year == y, age_sex == a_s)
-    if(nrow(d) == 0) { next }
-    range_yday <- d %>%
-      summarize(start_day = min(yday),
-                end_day = max(yday))
-    if(range_yday$start_day > 152) {next}
-    if(range_yday$end_day < 121) {next}
-    print(range_yday)
-    start_day <- range_yday$start_day
-    end_day <- range_yday$end_day
-    n_days = (end_day - start_day) + 1
-
-    # for wx covariates we'll use a gam to get values by day/hour since wx is likely
-    # to vary w/in day over the season
-
-    gam.baro <-
-      gam(pressure ~ s(yday), data = data)
-    gam.temp <-
-      gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), data = data)
-    gam.wind <-
-      gam(wind ~ s(yday) + s(as.numeric(solar_hour)), data = data)
-    gam.precip <-
-      gam(precip ~ s(yday), data = data)
-
-    wx_new_data <- data.frame(
-      solar_hour = 12,
-      yday = start_day:end_day
-    )
-
-    temp_pred <- predict(gam.temp, newdata = wx_new_data)
-    wind_pred <- predict(gam.wind, newdata = wx_new_data)
-    baro_pred <- predict(gam.baro, newdata = wx_new_data)
-    precip_pred <- predict(gam.precip, newdata = wx_new_data)
-
-    df <- data.frame(
-      age_sex = a_s,
-      year = y,
-      solar_hour = 12,
-      yday = start_day:end_day,
-      northing = mean(data$northing),
-      temp2 = temp_pred,
-      wind = wind_pred,
-      pressure = baro_pred,
-      precip = precip_pred
-    ) %>%
-      mutate(
-        sin1 = sin(pi * solar_hour / 12),
-        cos1 = cos(pi * solar_hour / 12),
-        sin2 = sin(pi * solar_hour / 6),
-        cos2 = cos(pi * solar_hour / 6),
-        sin3 = sin(pi * solar_hour / 4),
-        cos3 = sin(pi * solar_hour / 4),
-      ) %>%
-      mutate(day = (yday - 120) / 10,
-             day2 = day ^ 2,
-             day3 = day ^ 3)
-    df_list_years[[y]] <- df
-  }
-    df_list[[a_s]] <- bind_rows(df_list_years)
-  }
-  df_out <- bind_rows(df_list) %>%
-    mutate(age_sex = forcats::fct_relevel(
-      age_sex,c("ADULT.F","ADULT.M","SUBADULT","YOUNG OF YEAR")),
-      year = forcats::as_factor(as.numeric(year))
-    )
-
-  return(df_out)
-}
-```
-
 ```{r hfpvyear}
-load(here::here('data/fit_ribbon_year.Rdata'))
-
-ribbon_newdata_year <- create_newdata_year(fit_ribbon_year$dataset,
-                                           age_sex = levels(fit_ribbon_year$dataset$age_sex))
-
-# create the model matrix
-ribbon_mm <- model.matrix(fit_ribbon_year$fixed.formula[-2],
-                          data = ribbon_newdata_year)
-
-#clean up extra intercept and get coef
-fit_coef <- fit_ribbon_year$fixed.effects %>%
-  filter(!is.na(std.err)) %>%
-  pull(estimate)
-
-# multiply by the model coefficients and convert to probabilities with plogis
-# (this came from J Laake code for Hood Canal estimates)
-ribbon_newdata_year <- ribbon_newdata_year %>%
-  mutate(logit_fits = as.vector(ribbon_mm %*% fit_coef),
-         logit_fits_var = diag(ribbon_mm %*% fit_ribbon_year$covb %*% t(ribbon_mm)),
-         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
-         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
-         ho_prob = plogis(ribbon_mm %*% fit_coef),
-         lower95 = plogis(logit_fits_lo95),
-         upper95 = plogis(logit_fits_up95)
-  )
-
-load(here::here('data/fit_spotted_year.Rdata'))
-
-spotted_newdata_year <- create_newdata_year(fit_spotted_year$dataset,
-                                            age_sex = levels(fit_spotted_year$dataset$age_sex))
-
-# create the model matrix
-spotted_mm <- model.matrix(fit_spotted_year$fixed.formula[-2],
-                           data = spotted_newdata_year)
-
-#clean up extra intercept and get coef
-fit_coef <- fit_spotted_year$fixed.effects %>%
-  filter(!is.na(std.err)) %>%
-  pull(estimate)
-
-# multiply by the model coefficients and convert to probabilities with plogis
-# (this came from J Laake code for Hood Canal estimates)
-spotted_newdata_year <- spotted_newdata_year %>%
-  mutate(logit_fits = as.vector(spotted_mm %*% fit_coef),
-         logit_fits_var = diag(spotted_mm %*% fit_spotted_year$covb %*% t(spotted_mm)),
-         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
-         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
-         ho_prob = plogis(spotted_mm %*% fit_coef),
-         lower95 = plogis(logit_fits_lo95),
-         upper95 = plogis(logit_fits_up95)
-  )
+targets::tar_load(ribbon_newdata_year, store = here::here("_targets"))
+targets::tar_load(spotted_newdata_year, store = here::here("_targets"))
 ```
@@
 ribbon_year_type3 <- tibble(
-  fit_ribbon_year$typeIII.hypoth
+  ribbon_year_fit$typeIII.hypoth
 ) %>% 
@@
 spotted_year_type3 <- tibble(
-  fit_spotted_year$typeIII.hypoth
+  spotted_year_fit$typeIII.hypoth
 ) %>% 
@@
 Additionally, any groups where the underlying data started after 1 June or ended before 1 May are not included."}
-select_groups <- fit_ribbon_year$dataset %>%
+select_groups <- ribbon_year_fit$dataset %>%
   filter(age_sex != "YOUNG OF YEAR") %>%
@@
 
-select_groups <- fit_spotted_year$dataset %>%
+select_groups <- spotted_year_fit$dataset %>%
   filter(age_sex != "YOUNG OF YEAR") %>%
@@
 
-plot_df <- plot_df %>% bind_rows(plot_df2)
-peak_ho_df <- peak_ho_df %>% bind_rows(peak_ho_df2)
+plot_df <- plot_df %>% 
+  bind_rows(plot_df2) %>% 
+  mutate(year = forcats::fct_relevel(.$year, sort))
+peak_ho_df <- peak_ho_df %>% 
+  bind_rows(peak_ho_df2) %>% ungroup() %>% 
+  mutate(year = forcats::fct_relevel(.$year, sort))
 

### 2021-10-28 R/adfg_clean.R

@@ -135,6 +135,7 @@ adfg_clean_tl <- function(file1,file2, adfg_deployments) {
   cols <- cols(
     .default = readr::col_double(),
     DeployIDs = readr::col_character(),
+    GMTDate = readr::col_datetime(format = "%m/%d/%y"),
     TagType = readr::col_character(),
     HistType = readr::col_character(),
     LocationQuality = readr::col_character(),
@@ -145,9 +146,7 @@ adfg_clean_tl <- function(file1,file2, adfg_deployments) {
 
   adfg_timelines <- csv_files %>%
     purrr::map_dfr(read_csv, col_types = cols) %>%
-    dplyr::mutate(GMTDate = as.Date(GMTDate, origin = "1899-12-30",
-                                    tz = "UTC") %>%
-                    lubridate::as_datetime(),
+    dplyr::mutate(GMTDate = lubridate::as_datetime(GMTDate) %>% lubridate::force_tz("UTC"),
                   Ptt = as.integer(Ptt),
                   TagType = toupper(TagType)) %>%
     dplyr::rename(speno = DeployIDs,

### 2021-10-28 R/create_data_sf.R

@@ -2,7 +2,7 @@ create_data_sf <- function(locs_sf, source_data) {
   dat.sf <- source_data %>%
     filter(lubridate::yday(haulout_dt) < 197 ) %>%
     mutate(age = case_when(
-      age %in% c("pup", "PUP", "p") ~ "YOUNG OF YEAR",
+      age %in% c("pup", "PUP", "p", "Pup") ~ "YOUNG OF YEAR",
       age %in% c("yearling","YEARLING","subadult","Subadult") ~ "SUBADULT",
       age %in% c("adult","ADULT","Adult") ~ "ADULT",
       TRUE ~ age),
@@ -19,7 +19,7 @@ create_data_sf <- function(locs_sf, source_data) {
 
   deploy_meta <- locs_sf %>%
     as_tibble() %>%
-    select(speno, deploy_dt) %>%
+    dplyr::select(speno, deploy_dt) %>%
     filter(speno %in% !!unique(dat.sf$speno)) %>%
     group_by(speno) %>%
     summarise(deploy_dt = min(deploy_dt, na.rm = TRUE))

### 2021-10-28 R/create_db_input_data.R

@@ -21,7 +21,7 @@ create_source_data <- function(locs_sf, timeline_data) {
                                           error_radius.y,
                                           error_radius.x)) %>%
     rename(error_radius = error_radius.x) %>%
-    select(-error_radius.y) %>%
+    dplyr::select(-error_radius.y) %>%
     mutate(error_radius = ifelse(type %in% c("GPS","FastGPS"),
                                         50,error_radius),
                   error_radius = ifelse(type %in% c("User"),
@@ -38,7 +38,7 @@ create_source_data <- function(locs_sf, timeline_data) {
     group_by(speno) %>% nest() %>%
     mutate(start_idx = map_int(data,~ which.max(!is.na(.x$x))),
            data = map2(data, start_idx, ~ slice(.x, .y:nrow(.x)))) %>%
-    select(-start_idx) %>%
+    dplyr::select(-start_idx) %>%
     unnest(cols = c(data)) %>%
     mutate(fill_xy = ifelse(is.na(x), TRUE, FALSE)) %>%
     group_by(speno) %>%
@@ -50,7 +50,7 @@ create_source_data <- function(locs_sf, timeline_data) {
     st_as_sf(coords = c("x","y")) %>%
     st_set_crs(3571) %>%
     rename(haulout_dt = timeline_start_dt) %>%
-    select(speno,species,age,sex,haulout_dt,percent_dry,n_tags,fill_xy)
+    dplyr::select(speno,species,age,sex,haulout_dt,percent_dry,n_tags,fill_xy)
 
   stopifnot(
     "PEP Postgres Database Not Available; did you start VPN? ;)" =

### 2021-10-28 R/create_model_input.R

@@ -1,0 +1,59 @@
+create_model_input <- function(dat.sf) {
+  dat.sf  %>%
+  mutate(dry = round(percent_dry/100)) %>%
+  mutate(hour_utc = lubridate::hour(haulout_dt),
+         yday = lubridate::yday(haulout_dt),
+         year = lubridate::year(haulout_dt)) %>%
+  mutate(coords_x = st_coordinates(st_transform(.,4326))[,1],
+         coords_y = st_coordinates(st_transform(.,4326))[,2],
+         solar_hour = solaR::local2Solar(haulout_dt,coords_x) %>% lubridate::hour(),
+         sin1 = sin(pi*solar_hour/12),
+         cos1 = cos(pi*solar_hour/12),
+         sin2 = sin(pi*solar_hour/6),
+         cos2 = cos(pi*solar_hour/6),
+         sin3 = sin(pi*solar_hour/4),
+         cos3 = sin(pi*solar_hour/4),
+         day = (yday-120)/10,
+         day2 = day^2,
+         day3 = day^3
+  ) %>%
+  mutate(precip = rast_acpcp,
+         temp2 = (rast_air2m-270)/27,
+         temp0 = (rast_airsfc-270)/27,
+         pressure = (rast_prmsl-100000)/10000,
+         wind = sqrt(rast_uwnd^2 + rast_vwnd^2)/10
+  ) %>%
+  mutate(northing = st_coordinates(.)[,2]/mean(st_coordinates(.)[,2]),
+         north_of_bering = case_when(
+           coords_y > 66.5 ~ TRUE,
+           TRUE ~ FALSE
+         )) %>%
+  mutate(species = forcats::as_factor(species),
+         year = forcats::as_factor(year),
+         solar_hour = forcats::as_factor(solar_hour)) %>%
+  arrange(speno,haulout_dt) %>%
+  group_by(speno) %>%
+  mutate(ar1_id = paste(speno,create_ar1_id(haulout_dt),sep="_")) %>%
+  ungroup() %>%
+  mutate(ar1_id = forcats::as_factor(ar1_id),
+         speno = forcats::as_factor(speno)) %>%
+  relocate(ar1_id, .after = speno) %>%
+  dplyr::select(-c(coords_x, coords_y)) %>%
[…]

### 2021-10-28 R/create_newdata.R

@@ -1,0 +1,273 @@
+# function to create a prediction data frame for a given age_sex group
+create_newdata <- function(data, age_sex) {
+  df_list <- vector(mode = "list", length = length({{age_sex}} ))
+
+  for (a_s in {{age_sex}}) {
+    range_yday <- data %>% filter(age_sex == a_s ) %>%
+      summarize(start_day = min(yday),
+                end_day = max(yday))
+    start_day <- range_yday$start_day
+    end_day <- range_yday$end_day
+    n_days = (end_day - start_day) + 1
+
+    # for wx covariates we'll use a gam to get values by day/hour since wx is likely
+    # to vary w/in day over the season
+
+    gam.baro <-
+      gam(pressure ~ s(yday), data = data)
+    gam.temp <-
+      gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), data = data)
+    gam.wind <-
+      gam(wind ~ s(yday) + s(as.numeric(solar_hour)), data = data)
+    gam.precip <-
+      gam(precip ~ s(yday), data = data)
+
+    wx_new_data <- data.frame(
+      solar_hour = rep(0:23, each = n_days),
+      yday = rep(start_day:end_day, times = 24)
+    )
+
+    temp_pred <- predict(gam.temp, newdata = wx_new_data)
+    wind_pred <- predict(gam.wind, newdata = wx_new_data)
+    baro_pred <- predict(gam.baro, newdata = wx_new_data)
+    precip_pred <- predict(gam.precip, newdata = wx_new_data)
+
+    df <- data.frame(
+      age_sex = a_s,
+      solar_hour = rep(0:23, each = n_days),
+      yday = rep(start_day:end_day, times = 24),
+      northing = mean(data$northing),
+      temp2 = temp_pred,
+      wind = wind_pred,
+      pressure = baro_pred,
+      precip = precip_pred
+    ) %>%
+      mutate(
+        sin1 = sin(pi * solar_hour / 12),
+        cos1 = cos(pi * solar_hour / 12),
+        sin2 = sin(pi * solar_hour / 6),
+        cos2 = cos(pi * solar_hour / 6),
+        sin3 = sin(pi * solar_hour / 4),
+        cos3 = sin(pi * solar_hour / 4),
+      ) %>%
+      mutate(day = (yday - 120) / 10,
+             day2 = day ^ 2,
+             day3 = day ^ 3)
[…]

### 2021-10-28 R/fit_ribbon_models.R

@@ -1,0 +1,59 @@
+create_ribbon_data <- function(dat.sf) {
+  dat.sf %>%
+    filter(species == "Ribbon seal") %>%
+    st_set_geometry(NULL) %>%
+    as.data.frame() %>%
+    droplevels()
+}
+
+fit_ribbon <- function(HO_ribbon) {
+  glmmLDTS(
+    fixed.formula = dry ~ age_sex + sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + day +
+      day2 + day3 + temp2 + wind + pressure + precip + wind * temp2 +
+      sin1 * day + cos1 * day + sin2 * day + cos2 * day + sin3 *
+      day + cos3 * day +
+      sin1 * day2 + cos1 * day2 + sin2 * day2 + cos2 *
+      day2 + sin3 * day2 + cos3 * day2 +
+      age_sex:day + age_sex:day2 + age_sex:day3,
+    random.formula = dry ~ speno,
+    data = HO_ribbon,
+    EstMeth = "REML",
+    timecol = "time_vec",
+    #ridge.reg = "global",
+    #lambda = 0.5,
+    group.vec = "ar1_id"
+  )
+}
+
+
+fit_ribbon_noagesex <- function(HO_ribbon) {
+  glmmLDTS(
+    fixed.formula = dry ~ sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + day + day2 +
+      day3 + temp2 + wind + pressure + precip + wind * temp2 +
+      sin1 * day + cos1 * day + sin2 * day + cos2 *
+      day + sin3 * day + cos3 * day +
+      sin1 * day2 + cos1 * day2 + sin2 * day2 + cos2 *
+      day2 + sin3 * day2 + cos3 * day2,
+    random.formula = dry ~ speno,
+    data = HO_ribbon,
+    EstMeth = "REML",
+    timecol = "time_vec",
+    #ridge.reg = "global",
+    #lambda = 0.5,
+    group.vec = "ar1_id"
+  )
+}
+
+fit_ribbon_year <- function(HO_ribbon) {
+  glmmLDTS(fixed.formula = dry ~ age_sex + sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + day + day2+ day3 + temp2 + wind + pressure + precip + wind*temp2 +
+             sin1*day + cos1*day + sin2*day + cos2*day + sin3*day + cos3*day +
[…]

### 2021-10-28 R/fit_spotted_models.R

@@ -1,0 +1,48 @@
+create_spotted_data <- function(dat.sf) {
+  dat.sf %>%
+    filter(species == "Spotted seal") %>%
+    st_set_geometry(NULL) %>%
+    as.data.frame() %>%
+    droplevels()
+}
+
+fit_spotted <- function(HO_spotted) {
+  glmmLDTS(fixed.formula = dry ~ age_sex + sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + day + day2+ day3 + temp2 + wind + pressure + precip + wind*temp2 +
+             sin1*day + cos1*day + sin2*day + cos2*day + sin3*day + cos3*day +
+             sin1*day2 + cos1*day2 + sin2*day2 + cos2*day2 + sin3*day2 + cos3*day2 +
+             age_sex:day + age_sex:day2 + age_sex:day3,
+           random.formula = dry ~ speno,
+           data = HO_spotted,
+           EstMeth = "REML",
+           timecol = "time_vec",
+           #ridge.reg = "global",
+           #lambda = 0.5,
+           group.vec = "ar1_id")
+}
+
+fit_spotted_noagesex <- function(HO_spotted) {
+  glmmLDTS(fixed.formula = dry ~ sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + day + day2+ day3 + temp2 + wind + pressure + precip + wind*temp2 +
+             sin1*day + cos1*day + sin2*day + cos2*day + sin3*day + cos3*day +
+             sin1*day2 + cos1*day2 + sin2*day2 + cos2*day2 + sin3*day2 + cos3*day2,
+           random.formula = dry ~ speno,
+           data = HO_spotted,
+           EstMeth = "REML",
+           timecol = "time_vec",
+           #ridge.reg = "global",
+           #lambda = 0.5,
+           group.vec = "ar1_id")
+}
+
+fit_spotted_year <- function(HO_spotted) {
+  glmmLDTS(fixed.formula = dry ~ age_sex + sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + day + day2+ day3 + temp2 + wind + pressure + precip + wind*temp2 +
+             sin1*day + cos1*day + sin2*day + cos2*day + sin3*day + cos3*day +
+             sin1*day2 + cos1*day2 + sin2*day2 + cos2*day2 + sin3*day2 + cos3*day2 +
+             age_sex:day + age_sex:day2 + age_sex:day3 + year*day + year*day2,
[…]

### 2021-10-28 R/get_locs_sf.R

@@ -13,7 +13,7 @@ get_locs_sf <- function(adfg_locations, nsb_locations) {
   )
 
   deployments_db <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>%
-    select(speno, deployid, tag_family, deploy_dt, end_dt) %>%
+    dplyr::select(speno, deployid, tag_family, deploy_dt, end_dt) %>%
     rename(tag_type = tag_family) %>% collect()
 
   spenos_db <- tbl(con, in_schema("capture","for_telem")) %>% collect()

### 2021-10-28 R/get_nsb_data.R

@@ -28,6 +28,15 @@ using (sex_lku)"
   nsb_deployments <- tbl(con, sql(deploy_qry)) %>%
     dplyr::select(speno, deployid, species, age, sex, tag_family, deploy_dt, end_dt) %>%
     collect()
+  nsb_deployments2 <- tibble::tribble(
+    ~speno, ~deployid, ~species, ~age, ~sex, ~tag_family, ~deploy_dt, ~end_dt,
+    "EB2019BW02", 149452, "Bearded seal", "Pup", "Female", "SPLASH", "2019-09-19 18:30:00", "2020-05-24 17:30:00",
+    "EB2019BW02", 171333, "Bearded seal", "Pup", "Female", "SPOT", "2019-09-19 18:30:00", "2020-08-08 06:30:00"
+  ) %>%
+    dplyr::mutate(deploy_dt = lubridate::ymd_hms(deploy_dt),
+                  end_dt = lubridate::ymd_hms(end_dt))
+
+  nsb_deployments <- dplyr::bind_rows(nsb_deployments, nsb_deployments2)
 
   return(nsb_deployments)
 }
@@ -46,10 +55,21 @@ get_nsb_locs <- function(nsb_deployments) {
   )
   on.exit(odbc::dbDisconnect(con))
 
+  locs2 <- readr::read_csv(here::here("data_raw/nsb/iceSeals_NSB_Alaska_vonDuyke.csv")) %>%
+    janitor::clean_names() %>%
+    dplyr::filter(tag_local_identifier %in% c(149452, 171333)) %>%
+    dplyr::select(deployid = tag_local_identifier,
+                  locs_dt = timestamp,
+                  quality = argos_lc,
+                  location_long, location_lat) %>%
+    dplyr::mutate(deployid = as.integer(deployid)) %>%
+    sf::st_as_sf(coords = c("location_long","location_lat"), crs = 4326)
+
   locs_qry <- "SELECT deployid, loc_dt as locs_dt, argos_lc as quality, loc_geom as geometry
               FROM telem.nsb_locs;"
 
   nsb_locations <- read_sf(con, query = locs_qry) %>%
+    dplyr::bind_rows(locs2) %>%
     left_join(nsb_deployments, by = 'deployid') %>%
     filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
     filter(lubridate::month(locs_dt) %in% c(3,4,5,6,7)) %>%
[…]

### 2021-10-28 R/run_wxcov_query.R

@@ -1,24 +1,0 @@
-get_data_wxcov <- function() {
-  con <- dbConnect(RPostgres::Postgres(),dbname = 'pep',
-                   host = '161.55.120.122',
-                   port = 5432,
-                   user = keyringr::get_kc_account("pgpep_sa"),
-                   password = keyringr::decrypt_kc_pw("pgpep_sa"))
-  on.exit(dbDisconnect(con))
-
-  dbExecute(con, "SELECT telem.fxn_iceseal_pred_idx();")
-  dbExecute(con, "SELECT telem.fxn_iceseal_haulout_cov();")
-
-  qry <- {
-    "SELECT *
-  FROM telem.res_iceseal_haulout_cov
-  WHERE
-  EXTRACT(MONTH FROM haulout_dt) IN (3,4,5,6,7) AND
-  rast_vwnd IS NOT NULL AND
-  species != 'Ph' AND
-  speno != 'EB2009_7010' AND
-  speno != 'EB2009_3002' "
-  }
-
-  sf::st_read(con, query = qry)
-}

### 2021-10-28 _targets.R

@@ -1,17 +1,10 @@
 library(targets)
-# This is an example _targets.R file. Every
-# {targets} pipeline needs one.
-# Use tar_script() to create _targets.R and tar_edit()
-# to open it again for editing.
-# Then, run tar_make() to run the pipeline
-# and tar_read(summary) to view the results.
+library(future)
+library(future.callr)
+plan(callr)
 
-# Define custom functions and other global objects.
-# This is where you write source(\"R/functions.R\")
-# if you keep your functions in external scripts.
 files <- fs::dir_ls("R",recurse = TRUE, glob = "*.R")
 sapply(files, source)
-
 
 # Set target-specific options such as packages.
 tar_option_set(packages = c(
@@ -26,7 +19,13 @@ tar_option_set(packages = c(
   "keyringr",
   "pingr",
   "rnaturalearth",
-  "rnaturalearthdata"
+  "rnaturalearthdata",
+  "glmmLDTS",
+  "mgcv",
+  "solaR",
+  "splines",
+  "purler",
+  "wcUtils"
   )
 )
 
@@ -53,5 +52,25 @@ list(
 
   tar_target(analysis_data, create_data_sf(locs_sf, source_data)),
   tar_target(grid, create_grid_sf(analysis_data)),
-  tar_target(deploy_table, create_deploy_tbl(analysis_data))
+  tar_target(deploy_table, create_deploy_tbl(analysis_data)),
+
+  tar_target(model_data, create_model_input(analysis_data)),
+
+  tar_target(ribbon_model_data, create_ribbon_data(model_data)),
+  tar_target(ribbon_fit, fit_ribbon(ribbon_model_data)),
+  tar_target(ribbon_noagesex_fit, fit_ribbon_noagesex(ribbon_model_data)),
+  tar_target(ribbon_year_fit, fit_ribbon_year(ribbon_model_data)),
+  tar_target(spotted_model_data, create_spotted_data(model_data)),
+  tar_target(spotted_fit, fit_spotted(spotted_model_data)),
+  tar_target(spotted_noagesex_fit, fit_spotted_noagesex(spotted_model_data)),
+  tar_target(spotted_year_fit, fit_spotted_year(spotted_model_data)),
+  tar_target(bearded_model_data, create_bearded_data(model_data)),
+  tar_target(bearded_fit, fit_bearded(bearded_model_data)),
+
[…]

### 2021-10-28 data_raw/adfg/adfg_tidydata.Rmd
NEW DATA FROM NSB & ADFG
new deployment data from ADFG via Justin and NSB provided by Andy and downloaded from movebank

--- 2021-06-25 data_raw/adfg/adfg_tidydata.Rmd
@@
 
-There are three files provided by ADF&G
+There are six files provided by ADF&G
 
@@
 3. ADFG_SealTimelineData3.csv
+4. ADFG_SealDeploymentTable_NewRecords.csv
+5. ADFG_SealLocationData_NewRecords.csv
+6. ADFG_SealTimelineData_NewRecords.csv
 
@@
 
-Let's read this file in and take a look at the structure
+Let's read these files in and take a look at the structure
 
@@
 adfg_deployments <- readr::read_csv('ADFG_SealDeploymentTable3.csv')
+adfg_deployments_new <- readr::read_csv('ADFG_SealDeploymentTable_NewRecords.csv')
 ```
@@
   MapID = readr::col_character(),
-  DateTagged = readr::col_datetime(format = "%m/%d/%Y %H:%M"),
+#  DateTagged = readr::col_datetime(format = "%m/%d/%Y %H:%M"),
   Species = readr::col_character(),
@@
                                     col_types = adfg_deploy_cols) %>% 
+  mutate(DateTagged = parse_datetime(DateTagged,format = "%m/%d/%Y %H:%M")) %>% 
   dplyr::mutate(DateTagged = lubridate::force_tz(DateTagged, 
                                                  'America/Anchorage')) %>% 
-  dplyr::mutate(DateTagged = lubridate::as_date(DateTagged, 
+  dplyr::mutate(DateTagged = lubridate::as_datetime(DateTagged, 
                                                 tz = 'America/Anchorage'))
@@
 glimpse(adfg_deployments)
+```
+
+```{r}
+glimpse(adfg_deployments_new)
+```
+```{r}
+adfg_deploy_cols <- readr::cols(
+  DeployIDs = readr::col_character(),
+  Species = readr::col_character(),
+  Age = readr::col_character(),
+  Sex = readr::col_character(),
+  FlipperTagType = readr::col_character()
+)
+
+adfg_deployments_new <- readr::read_csv('ADFG_SealDeploymentTable_NewRecords.csv',
+                                    col_types = adfg_deploy_cols) %>% 
+  mutate(DateTagged = parse_datetime(DateTagged,format = "%m/%d/%y")) %>% 
+  dplyr::mutate(DateTagged = lubridate::force_tz(DateTagged, 
+                                                 'America/Anchorage')) %>% 
+  dplyr::mutate(DateTagged = lubridate::as_datetime(DateTagged, 
+                                                tz = 'America/Anchorage'))
+
+glimpse(adfg_deployments_new)
+```
+
+```{r}
+adfg_deployments <- adfg_deployments %>% 
+  bind_rows(adfg_deployments_new)
 ```
@@
                 primary_tag_type = PrimaryTagType,
-                flipper_tag_type = FlipperTagType
+                flipper_tag_type = FlipperTagType,
+                flipper_last_signal = FlipperSatTagLastSignal
                 ) %>% 
@@
 adfg_locations <- readr::read_csv('ADFG_SealLocationData3.csv')
+adfg_locations_new <- readr::read_csv('ADFG_SealLocationData_NewRecords.csv')
 ```
@@
 
+```{r}
+glimpse(adfg_locations_new)
+```
+
+```{r}
+adfg_locs_cols <- readr::cols(
+  DeployID = readr::col_character(),
+  LocationQuality = readr::col_character(),
+  Date = readr::col_double(),
+  Latitude = readr::col_double(),
+  Longitude = readr::col_double(),
+  Ptt = readr::col_double(),
+  TagType = readr::col_character(),
+  SerialNum = readr::col_character()
+)
+
+adfg_locations_new <- readr::read_csv('ADFG_SealLocationData_NewRecords.csv',
+                                  col_types = adfg_locs_cols) %>% 
+  dplyr::mutate(Date = as.Date(Date, origin = "1899-12-30", tz ="UTC") %>% 
+                  lubridate::as_datetime(),
+                Ptt = as.integer(Ptt))
+
+glimpse(adfg_locations_new)
+```
+
+```{r}
+adfg_locations <- adfg_locations %>% 
+  bind_rows(adfg_locations_new)
+```
+
 As before, we'll rename a few columns to snake_case and then all to lower case.
@@
   dplyr::filter(quality != "Z") %>% 
-  dplyr::select(-MonthNumb) %>% 
   dplyr::rename_all(tolower) %>% 

### 2021-10-28 analysis/scripts/full_deployment_table.R
ADD HELPER SCRIPT TO CREATE DEPLOY TBL

@@ -1,0 +1,31 @@
+View(dat.sf)
+
+library(DBI)
+library(odbc)
+library(pingr)
+stopifnot(
+"PEP Postgres Database Not Available; did you start VPN? ;)" =
+pingr::is_up("161.55.120.122", "5432")
+)
+
+con <- dbConnect(
+  odbc::odbc(),
+  dsn = "PostgreSQL pep",
+  uid = keyringr::get_kc_account("pgpep_londonj"),
+  pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
+)
+
+qry <- "SELECT d.speno, d.serial_num, d.ptt, d.field_effort_lku, effort_lead, meta_project
+FROM telem.tbl_tag_deployments d
+LEFT JOIN capture.tbl_event USING (speno)
+ORDER BY speno"
+
+res <- con %>% tbl(dplyr::sql(qry)) %>%
+  collect()
+
+dat.sf %>% as_tibble() %>%
+  group_by(speno,species,age,sex,deploy_dt) %>% count() %>%
+  left_join(res, by="speno") %>%
+  arrange(deploy_dt) %>%
+  write_csv("haulout_deployments.csv")
+

### 2021-11-02 data_raw/nsb/nsb_tidydata.Rmd

@@
 
+nsb_deployments2 <- tibble::tribble(
+  ~speno, ~deployid, ~species, ~age, ~sex, ~tag_family, ~deploy_dt, ~end_dt,
+  "EB2019BW02", 149452, "Bearded seal", "Pup", "Female", "SPLASH", "2019-09-19 18:30:00", "2020-05-24 17:30:00",
+  "EB2019BW02", 171333, "Bearded seal", "Pup", "Female", "SPLASH", "2019-09-19 18:30:00", "2020-08-08 06:30:00"
+) %>% 
+  dplyr::mutate(deploy_dt = lubridate::ymd_hms(deploy_dt),
+                end_dt = lubridate::ymd_hms(end_dt))
+
+nsb_deployments <- dplyr::bind_rows(nsb_deployments, nsb_deployments2)
+
 nsb_timelines <- tbl(con, in_schema("telem","nsb_histos_timeline")) %>%
@@
   collect()
+
+nsb_timelines2 <- wcUtils::read_histos(
+  here::here('data_raw/nsb/iceSeals_NSB_Alaska_vonDuyke_updatesOnly_20211008-Histos.csv'),
+  dt_fmt = "%Y-%m-%d %H:%M:%S") %>% 
+  wcUtils::tidyTimelines() %>% 
+  dplyr::mutate(deployid = as.integer(deployid)) %>% 
+  dplyr::filter(deployid %in% c(149452, 171333)) %>% 
+  dplyr::select(-hist_type) %>% as_tibble()
+
+nsb_timelines <- dplyr::bind_rows(nsb_timelines, nsb_timelines2)
 

### 2021-11-02 analysis/paper/paper.Rmd
ADD CO-AUTHORS

@@
       correspondence: false
+  - Justin Crawford:
+      institute: [ADFG]
+      correspondence: false
+  - Andrew L. Von Duyke:
+      institute: [NSB]
+      correspondence: false
+  - Lori T. Quakenbush:
+      institute: [ADFG]
+      correspondence: false
   - Peter L. Boveng:
@@
   - MML: Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
+  - ADFG: Arctic Marine Mammals Program, Alaska Department of Fish and Game, Fairbanks, Alaska, USA
+  - NSB: Department of Wildlife Management, North Slope Borough, Utqiaġvik, Alaska, USA
 output: 

### 2021-11-04 analysis/paper/paper.Rmd
UPDATE REFERENCES JSON
things had gotten out of sync and the built-in RStudio editor citation workflow was causing problems. Now, relying on the {rbbt} package.

@@
 link-citations: true
-csl: ../templates/royal-society-open-science.csl
+csl: ../templates/proceedings-of-the-royal-society-b.csl
 always_allow_html: true
-params:
-   run_models: FALSE
 ---
@@
 
-Ice-associated seals rely on sea ice for a variety of activities, including breeding,
-molting, pupping, and resting. In the Arctic, many of these activities occur in spring
-(April-June) as sea ice begins to melt and retreat northward. Rapid acceleration of
-climate change in Arctic ecosystems is therefore of concern as the quantity and quality of
-suitable habitat is forecast to decrease. In this paper, we use data collected from
-satellite-linked bio-loggers deployed between 2005 and 2020 to investigate the seasonal
-timing and environmental factors affecting sea ice use by seals (specifically, bearded,
-ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to providing
-baseline data on phenology, these data also allow us to quantify 'availability', which is
-needed to accurately estimate abundance from aerial survey counts of seals basking on ice
-(i.e., to correct for the proportion of animals that are in the water while surveys are
-conducted). Using generalized linear mixed pseudo-models to properly account for temporal
-autocorrelation, we fit models with a variety of covariates (e.g., day-of-year, solar
-hour, age-sex class, wind speed, barometric pressure, temperature, precipitation) to
-examine their ability to explain variation in hourly haul-out records. We found evidence
-for strong diurnal and within-season patterns in haul-out behavior, as well as strong
-weather effects (particularly wind and temperature). In general, seals were more likely to
-rest on ice in the middle of the day and when wind speed was low and temperatures were
-higher. Haul-out propensity increased through March and April, peaking in May and early
-June before declining again. The timing and frequency of haul-out events also varied based
-on species and age-sex class. For ribbon and spotted seals, models with year effects were
-highly supported, indicating that the timing and magnitude of haul-out behavior varied
-among years. However, we did not find evidence that haul-out timing was linked to annual
-sea ice extent. Our analysis emphasizes the importance of accounting for variation in
-haul-out behavior when interpreting the number of seals counted in aerial surveys.
-
-keywords: availability; generalized linear mixed pseudo-model; haul-out behavior;
-phenology; Phocidae; sea ice
+Ice-associated seals rely on sea ice for a variety of activities,
+including breeding, molting, pupping, and resting. In the Arctic, many
+of these activities occur in spring (April-June) as sea ice begins to
+melt and retreat northward. Rapid acceleration of climate change in
+Arctic ecosystems is therefore of concern as the quantity and quality of
+suitable habitat is forecast to decrease. In this paper, we use data
+collected from satellite-linked bio-loggers deployed between 2005 and
+2020 to investigate the seasonal timing and environmental factors
+affecting sea ice use by seals (specifically, bearded, ribbon, and
+spotted seals) in the Bering and Chukchi seas. In addition to providing
+baseline data on phenology, these data also allow us to quantify
+'availability', which is needed to accurately estimate abundance from
+aerial survey counts of seals basking on ice (i.e., to correct for the
+proportion of animals that are in the water while surveys are
+conducted). Using generalized linear mixed pseudo-models to properly
+account for temporal autocorrelation, we fit models with a variety of
+covariates (e.g., day-of-year, solar hour, age-sex class, wind speed,
+barometric pressure, temperature, precipitation) to examine their
+ability to explain variation in hourly haul-out records. We found
+evidence for strong diurnal and within-season patterns in haul-out
+behavior, as well as strong weather effects (particularly wind and
+temperature). In general, seals were more likely to rest on ice in the
+middle of the day and when wind speed was low and temperatures were
+higher. Haul-out propensity increased through March and April, peaking
+in May and early June before declining again. The timing and frequency
+of haul-out events also varied based on species and age-sex class. For
+ribbon and spotted seals, models with year effects were highly
+supported, indicating that the timing and magnitude of haul-out behavior
+varied among years. However, we did not find evidence that haul-out
+timing was linked to annual sea ice extent. Our analysis emphasizes the
+importance of accounting for variation in haul-out behavior when
+interpreting the number of seals counted in aerial surveys.
+
+keywords: availability; generalized linear mixed pseudo-model; haul-out
+behavior; phenology; Phocidae; sea ice
 
@@
 
-The global climate disruption is causing considerable reduction in Arctic sea ice extent,
-volume, and seasonal presence [@kwok2018erl, @meier2014rg, @wang2017dsrpitsio]. These
-changes have ripple effects on Arctic organisms, ecosystems, and the human communities who
-live in the region [@huntington2020ncc]. Such disruptions are a particular cause of
-concern for ice-associated seals (ringed (*Pusa hispida*), bearded (*Erignathus
-barbatus*), spotted (*Phoca largha*), and ribbon seals (*Histriophoca fasciata*)) that
-depend on spring and early summer sea ice (March-June) in the Bering and Chukchi seas as a
-platform for important functions, such as pupping, nursing, breeding, and molting
-[@boveng2009a; @boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge
-gaps complicate predictions about the ultimate effects of changes in sea ice on the
-behavior, abundance, and distribution of these seals. Knowledge about evolutionary
-constraints on the timing of reproductive and molting behavior is generally lacking, so it
-is difficult to predict the readiness with which ice-associated seal species will be able
-to adapt to future changes (e.g., by adjusting pupping or molting schedules to earlier
-dates or different locales). Additionally, trends in abundance of these species are
-largely unknown, so it is difficult to say what effect, if any, declines in sea ice
+The global climate disruption is causing considerable reduction in
+Arctic sea ice extent, volume, and seasonal presence [@kwok2018,
+@meier2014, @wang2017]. These changes have ripple effects on
+Arctic organisms, ecosystems, and the human communities who live in the
+region [@]. Such disruptions are a particular cause
+of concern for ice-associated seals (ringed (*Pusa hispida*), bearded
+(*Erignathus barbatus*), spotted (*Phoca largha*), and ribbon seals
+(*Histriophoca fasciata*)) that depend on spring and early summer sea
+ice (March-June) in the Bering and Chukchi seas as a platform for
+important functions, such as pupping, nursing, breeding, and molting
+[@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data
+and large knowledge gaps complicate predictions about the ultimate
+effects of changes in sea ice on the behavior, abundance, and
+distribution of these seals. Knowledge about evolutionary constraints on
+the timing of reproductive and molting behavior is generally lacking, so
+it is difficult to predict the readiness with which ice-associated seal
+species will be able to adapt to future changes (e.g., by adjusting
+pupping or molting schedules to earlier dates or different locales).
+Additionally, trends in abundance of these species are largely unknown,
+so it is difficult to say what effect, if any, declines in sea ice
 habitat have had, or will have, on seal densities.
 
-For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
-of the water almost exclusively on sea ice and are mostly pelagic outside the spring
-pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals rest
-on beaches during ice-free months, they strongly prefer sea ice as a resting platform
-during the spring and early summer [@burns2009]. The remoteness of the Bering and Chukchi
-seas mean direct scientific observation of seal behavior is impractical. Bio-logging
-devices are, thus, especially useful tools for collecting key information on movement and
-haul-out behavior for these species.
-
-Bio-logging records of time spent out of the water provide valuable data for identifying
-covariates that explain variation in haul-out behavior. For instance, in the Antarctic,
-Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to demonstrate greater
-haul-out propensity in juvenile crabeater seals (*Lobodon carcinophaga*) than adults, with
-highest probabilities in February and at times close to solar noon. In the Arctic, Von
-Duyke et al. [-@vonduyke2020ee] used seal-borne, satellite-linked loggers to corroborate
-seasonal changes between diurnal and nocturnal haul-out behavior of ringed seals
-previously described by Kelly et al. [-@kelly2010] using labor-intensive VHF radio tags
-and direct observation. Bengtson et al. [-@bengtson2005pb] documented a higher propensity
-for ringed seal basking near solar noon, as did Ver Hoef et al. [-@verhoef2014sm] in an
-analysis of bearded, ribbon, and spotted seals using much larger sample sizes. Olnes et
-al. [-@olnes2020pb] showed that the proportion of time bearded seals spent hauled out
-progressively increased through spring and summer. And, Ver Hoef et al. [-@verhoef2014sm]
-found haul-out probabilities increased gradually starting in March and peaking in May and
-June for bearded, ribbon, and spotted seals.
-
-While bio-logging of haul-out patterns clearly has utility for description of
-ice-associated seals' natural history and ecology, some studies have focused, as we do
-here, specifically on quantifying the probability of seals being hauled out as a means of
-accounting for seals in the water, and therefore missed, during aerial surveys. These
-studies typically use logistic regression-style analyses to estimate the time-specific
-probability of being hauled out base on 'wet/dry' data relayed by bio-loggers. In these
-models, haul-out probabilities were expressed as a function of predictive covariates, such
-as time of day, day-of-year, sex, age class, and environmental conditions (e.g.,
-[@reder2003pb], [@bengtson2004pb], [@bengtson2005pb], [@udevitz2009cjz], [@verhoef2014sm],
-and [@southwell2008]). However, sample sizes have often been low and insufficient to
-permit strong inference about life history and/or seasonal variation in haul-out
-probabilities. For instance, Bengtson and Cameron's [-@bengtson2004pb] study included 5
-adult and 2 juvenile crabeater seals, while Bengtson et al.'s [-@bengtson2005pb] study was
-based on 6 telemetered ringed seals in the Chukchi Sea. These studies were often further
-limited by logistical constraints on tagging fieldwork and the attachment duration or
-operational life of satellite tags. In this study, we address some of these limitations by
-deploying small bio-loggers designed for attachment on a portion of the study individuals.
-These devices are designed to collect data through the molt period (when those adhered to
-the hair would fall off) and, in some situations, provide multiple years of data.
-
-Ultimately, knowledge of trends in phenology and abundance (or life history surrogates
-such as survival and recruitment) will be necessary to make credible quantitative
-predictions about the effects of the climate crisis on the abundance and distribution of
-Arctic seal populations. Before we can construct a trend, however, we first require a
-baseline. Several studies have contributed estimates of the distribution and abundance of
-ice-associated seal species in the Arctic using aerial surveys (e.g., [@bengtson2005pb],
-[@conn2014mee], and [@verhoef2014sm]). Such studies were conducted over very large areas
-and estimation of absolute abundance required making inference about numerous nuisance
-processes affecting the observation of seals on ice. This included availability (only
-seals basking on ice were available to be counted), detection probability (observers or
-automated detection systems may have missed some seals on ice), species misclassification,
-and possible disturbance of seals by aircraft [@conn2014mee; @verhoef2014sm].
-
-In this study, we used data from 15 years of bio-logger records in a comprehensive
-investigation of haul-out behavior by 3 species of ice-associated seals in the Bering and
-Chukchi seas. We focused our effort on bearded, spotted, and ribbon seals and did not
-include ringed seals in the analysis. Ringed seals are unique in their use of snow lairs
-in the spring months, and they warrant a separate study that includes additional
-environmental covariates, such as the onset of snow melt and snow depth. Our goals are
-threefold. First, we wish to establish baseline estimates for the chronology of haul-out
-behavior in the critical spring season for each species across different age and sex
-classes. Second, we wish to refine estimates of haul-out availability corrections for
-aerial surveys in order to improve estimates of seal abundance. Previously estimated
-availability correction factors (e.g., [@bengtson2005pb], [@conn2014mee], and
-[@verhoef2014sm]) accounted for variables such as the time-of-day and day-of-year, but did
-not investigate meteorological variables that have been shown to influence haul-out
-behavior of walruses [@reder2003pb; @udevitz2009cjz]. Third, we explored the annual
-variability in haul-out timing and possible linkage to dramatic changes in the
-availability and extent of seasonal sea ice between 2005 and 2020. Our work extends the
-scope of previous haul-out analyses, includes the influence of meteorological variability,
-and investigates the potential impact of a changing icescape on the behavior of these
-species.
+For ice-associated seals, haul out is tightly linked with sea ice.
+Ribbon seals rest out of the water almost exclusively on sea ice and are
+mostly pelagic outside the spring pupping, breeding, and molting season
+[@burns1981g]. While spotted and bearded seals rest on beaches during
+ice-free months, they strongly prefer sea ice as a resting platform
+during the spring and early summer [@burns2009]. The remoteness of the
+Bering and Chukchi seas mean direct scientific observation of seal
+behavior is impractical. Bio-logging devices are, thus, especially
+useful tools for collecting key information on movement and haul-out
+behavior for these species.
+
+Bio-logging records of time spent out of the water provide valuable data
+for identifying covariates that explain variation in haul-out behavior.
+For instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004]
+relied on bio-logging data to demonstrate greater haul-out propensity in
+juvenile crabeater seals (*Lobodon carcinophaga*) than adults, with
+highest probabilities in February and at times close to solar noon. In
+the Arctic, Von Duyke et al. [-@vonduyke2020] used seal-borne,
+satellite-linked loggers to corroborate seasonal changes between diurnal
+and nocturnal haul-out behavior of ringed seals previously described by
+Kelly et al. [-@kelly2010] using labor-intensive VHF radio tags and
+direct observation. Bengtson et al. [-@bengtson2005a] documented a
+higher propensity for ringed seal basking near solar noon, as did Ver
+Hoef et al. [-@verhoef2014a] in an analysis of bearded, ribbon, and
+spotted seals using much larger sample sizes. Olnes et al.
+[-@olnes2020] showed that the proportion of time bearded seals spent
+hauled out progressively increased through spring and summer. And, Ver
+Hoef et al. [-@verhoef2014a] found haul-out probabilities increased
+gradually starting in March and peaking in May and June for bearded,
+ribbon, and spotted seals.
+
+While bio-logging of haul-out patterns clearly has utility for
+description of ice-associated seals' natural history and ecology, some
+studies have focused, as we do here, specifically on quantifying the
+probability of seals being hauled out as a means of accounting for seals
+in the water, and therefore missed, during aerial surveys. These studies
+typically use logistic regression-style analyses to estimate the
+time-specific probability of being hauled out base on 'wet/dry' data
+relayed by bio-loggers. In these models, haul-out probabilities were
+expressed as a function of predictive covariates, such as time of day,
+day-of-year, sex, age class, and environmental conditions (e.g.,
+[@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
+[@verhoef2014a], and [@southwell2008]). However, sample sizes have
+often been low and insufficient to permit strong inference about life
+history and/or seasonal variation in haul-out probabilities. For
+instance, Bengtson and Cameron's [-@bengtson2004] study included 5
+adult and 2 juvenile crabeater seals, while Bengtson et al.'s
+[-@bengtson2005a] study was based on 6 telemetered ringed seals in the
+Chukchi Sea. These studies were often further limited by logistical
+constraints on tagging fieldwork and the attachment duration or
+operational life of satellite tags. In this study, we address some of
+these limitations by deploying small bio-loggers designed for attachment
+on a portion of the study individuals. These devices are designed to
+collect data through the molt period (when those adhered to the hair
+would fall off) and, in some situations, provide multiple years of data.
+
+Ultimately, knowledge of trends in phenology and abundance (or life
+history surrogates such as survival and recruitment) will be necessary
+to make credible quantitative predictions about the effects of the
+climate crisis on the abundance and distribution of Arctic seal
+populations. Before we can construct a trend, however, we first require
+a baseline. Several studies have contributed estimates of the
+distribution and abundance of ice-associated seal species in the Arctic
+using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
+[@verhoef2014a]). Such studies were conducted over very large areas
+and estimation of absolute abundance required making inference about
+numerous nuisance processes affecting the observation of seals on ice.
+This included availability (only seals basking on ice were available to
+be counted), detection probability (observers or automated detection
+systems may have missed some seals on ice), species misclassification,
+and possible disturbance of seals by aircraft [@conn2014a;
+@verhoef2014a].
+
+In this study, we used data from 15 years of bio-logger records in a
+comprehensive investigation of haul-out behavior by 3 species of
+ice-associated seals in the Bering and Chukchi seas. We focused our
+effort on bearded, spotted, and ribbon seals and did not include ringed
+seals in the analysis. Ringed seals are unique in their use of snow
+lairs in the spring months, and they warrant a separate study that
+includes additional environmental covariates, such as the onset of snow
+melt and snow depth. Our goals are threefold. First, we wish to
+establish baseline estimates for the chronology of haul-out behavior in
+the critical spring season for each species across different age and sex
+classes. Second, we wish to refine estimates of haul-out availability
+corrections for aerial surveys in order to improve estimates of seal
+abundance. Previously estimated availability correction factors (e.g.,
+[@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
+for variables such as the time-of-day and day-of-year, but did not
+investigate meteorological variables that have been shown to influence
+haul-out behavior of walruses [@reder2003; @udevitz2009]. Third,
+we explored the annual variability in haul-out timing and possible
+linkage to dramatic changes in the availability and extent of seasonal
+sea ice between 2005 and 2020. Our work extends the scope of previous
+haul-out analyses, includes the influence of meteorological variability,
+and investigates the potential impact of a changing icescape on the
+behavior of these species.
 
@@
 
-Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)` bio-loggers deployed
-on bearded, spotted, and ribbon seals was filtered to include only records from 1 March to
-15 July between `r min(lubridate::year(analysis_data$haulout_dt))` and
-`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
-family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
-adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
-use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
-devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
-during the annual molt, which limits the effective length of haul-out timelines.
-Additionally, sensors attached to the head or dorsal region are often dry while the seal
-is at the surface and can slightly bias the percent-dry values recorded by the device. For
-this study, in cases where both tag types were deployed, percent dry observations from the
-flipper tag were preferred. Sex and age class (non-dependent *young-of-the-year*, sexually
-immature *subadults*, and mature *adults*) were determined at the time of deployment by
-subjective judgement based on length, and pelage characteristics for some species. Field
-identification of age class can be inexact, particularly when discerning subadults from
-adults. In the case of ribbon seals, pelage differences between subadults and adults are
-distinguishable, with subadults often having less defined ribbons. Bearded seal subadults
-will often have a spotted pattern in the pelage that is not seen in adults. Spotted seal
-pelage is monomorphic and can not be used to discern age class. Despite these challenges,
-we feel the age classifications used in this analysis are useful categories in testing for
-age-related effects on haul-out behavior. Seals determined to be less than one year were
-classified as young-of-the-year. Subadults are those seals likely greater than one year of
-age but not yet sexually mature. Adults are likely sexually mature and older than
-approximately four years. For those tags deployed on young-of-the-year and transmitting
-into the next year, the age class was advanced to subadult on 1 March of the following
-year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
-from them.
+Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
+bio-loggers deployed on bearded, spotted, and ribbon seals was filtered
+to include only records from 1 March to 15 July between
+`r min(lubridate::year(analysis_data$haulout_dt))` and
+`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of
+the 'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers
+(Redmond, Washington, USA) and either adhered to the hair on the seal or
+attached to the rear flipper inter-digital webbing. The use of
+bio-loggers adhered to the back or head provide some benefits over
+flipper mounted devices (e.g. increased satellite transmittal rates,
+locations at sea) but these are lost during the annual molt, which
+limits the effective length of haul-out timelines. Additionally, sensors
+attached to the head or dorsal region are often dry while the seal is at
+the surface and can slightly bias the percent-dry values recorded by the
+device. For this study, in cases where both tag types were deployed,
+percent dry observations from the flipper tag were preferred. Sex and
+age class (non-dependent *young-of-the-year*, sexually immature
+*subadults*, and mature *adults*) were determined at the time of
+deployment by subjective judgement based on length, and pelage
+characteristics for some species. Field identification of age class can
+be inexact, particularly when discerning subadults from adults. In the
+case of ribbon seals, pelage differences between subadults and adults
+are distinguishable, with subadults often having less defined ribbons.
+Bearded seal subadults will often have a spotted pattern in the pelage
+that is not seen in adults. Spotted seal pelage is monomorphic and can
+not be used to discern age class. Despite these challenges, we feel the
+age classifications used in this analysis are useful categories in
+testing for age-related effects on haul-out behavior. Seals determined
+to be less than one year were classified as young-of-the-year. Subadults
+are those seals likely greater than one year of age but not yet sexually
+mature. Adults are likely sexually mature and older than approximately
+four years. For those tags deployed on young-of-the-year and
+transmitting into the next year, the age class was advanced to subadult
+on 1 March of the following year. Table \@ref(tab:tblDeploy) provides a
+summary of these deployments and data received from them.
 
@@
 
-Tags that fall off due to molt or mortality and remain on ice or land can still send data
-to satellites, and we did not want to include data from these tags in our analyses. As
-such, end times of each deployment were identified by experts who examined several
-simultaneous data streams (e.g., tag locations and dive behavior) to determine when tags
-ceased providing data consistent with seal behavior. Sensor readings outside of these
-start and end times were discarded prior to analysis. After approximately 9 months, 7
-devices deployed on the rear flipper of bearded seals reported implausible sensor data.
-All data after the first instance of unrealistic values were censored from this analysis.
-Figure \@ref(fig:dataCal) shows a calendar distribution of these data across the study
-season for each species. Observations are concentrated in the months of May and June due
-to the timing of deployment (April and May) and the timing of molt (May and June). During
-molt, seals (and their attached bio-loggers) spend more time out of the water and more
-data is transmitted. Molt timing also impacts when many deployments end as any tags
-adhered to the hair will fall off.
+Tags that fall off due to molt or mortality and remain on ice or land
+can still send data to satellites, and we did not want to include data
+from these tags in our analyses. As such, end times of each deployment
+were identified by experts who examined several simultaneous data
+streams (e.g., tag locations and dive behavior) to determine when tags
+ceased providing data consistent with seal behavior. Sensor readings
+outside of these start and end times were discarded prior to analysis.
+After approximately 9 months, 7 devices deployed on the rear flipper of
+bearded seals reported implausible sensor data. All data after the first
+instance of unrealistic values were censored from this analysis. Figure
+\@ref(fig:dataCal) shows a calendar distribution of these data across
+the study season for each species. Observations are concentrated in the
+months of May and June due to the timing of deployment (April and May)
+and the timing of molt (May and June). During molt, seals (and their
+attached bio-loggers) spend more time out of the water and more data is
+transmitted. Molt timing also impacts when many deployments end as any
+tags adhered to the hair will fall off.
 
@@
 
-Haul-out behavior data was recorded and transmitted via the Argos satellite network as
-hourly percent-dry timelines. For each hour of a day, the wet/dry sensor was polled by the
-tag firmware every few seconds and a percent of the hour in a dry state was calculated
-(Figure \@ref(fig:examplePlot)). Each data transmission message consisted of a complete
-24-hour record. Tags were, generally, deployed and programmed in a manner to maximize data
-transmission during the spring pupping and molting period, however, not all days during
-the deployment were successfully transmitted. This is due to a variety of factors
-including satellite coverage, tag availability (i.e. tags mounted to the rear flipper are
-not available to the surface while the seal is at sea), tag performance, duty cycling, and
-extra-terrestrial interference.
+Haul-out behavior data was recorded and transmitted via the Argos
+satellite network as hourly percent-dry timelines. For each hour of a
+day, the wet/dry sensor was polled by the tag firmware every few seconds
+and a percent of the hour in a dry state was calculated (Figure
+\@ref(fig:examplePlot)). Each data transmission message consisted of a
+complete 24-hour record. Tags were, generally, deployed and programmed
+in a manner to maximize data transmission during the spring pupping and
+molting period, however, not all days during the deployment were
+successfully transmitted. This is due to a variety of factors including
+satellite coverage, tag availability (i.e. tags mounted to the rear
+flipper are not available to the surface while the seal is at sea), tag
+performance, duty cycling, and extra-terrestrial interference.
 
@@
 
-Of key interest in this study is the relationship between haul-out behavior and a variety
-of environmental covariates that vary with space and time. We explored the use of a
-movement model to predict locations at specific times. However, the sparse nature of data
-from some tags, especially those mounted to the rear flipper, resulted in poor modeling
-performance or convergence issues. For this study, we calculated a weighted average daily
-location where the inverse of the estimated Argos location error was used for the weight.
-Any days where haul-out observations were present without any location data were filled
-from the last calculated weighted average daily location, and any days where the location
-intersected with land were removed from the dataset. Figure \@ref(fig:dataMap) shows the
-spatial distribution of determined locations and available haul-out behavior data for each
-species across the study area.
+Of key interest in this study is the relationship between haul-out
+behavior and a variety of environmental covariates that vary with space
+and time. We explored the use of a movement model to predict locations
+at specific times. However, the sparse nature of data from some tags,
+especially those mounted to the rear flipper, resulted in poor modeling
+performance or convergence issues. For this study, we calculated a
+weighted average daily location where the inverse of the estimated Argos
+location error was used for the weight. Any days where haul-out
+observations were present without any location data were filled from the
+last calculated weighted average daily location, and any days where the
+location intersected with land were removed from the dataset. Figure
+\@ref(fig:dataMap) shows the spatial distribution of determined
+locations and available haul-out behavior data for each species across
+the study area.
 
@@
 
-In addition to sex and age class, we gathered additional variables that might help explain
-variation in haul-out probabilities. These included day-of-year (for seasonal effects) and
-local solar hour (for diurnal effects). Solar hour was calculated using the {solaR}
-package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019] using the
-daily locations. We also compiled several meteorological variables shown to affect
-haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz; @perry2017].
-In particular, we linked locations from satellite tags to key weather values from the
-North American Regional Reanalysis (NARR) model produced by the National Centers for
-Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
-data to produce a long-term picture of weather over North America. Numerous weather
-variables are made available across the region 8 times daily. For this study, NARR weather
-values were subset to the extent of our study area over the Bering and Chukchi Seas at
-3-hr intervals based on the native grid resolution of 32 km (1024 sq. km). The following
-meteorological variables were interpolated and assigned to seal locations using a bilinear
-method: 1) air temperature at 2m above the earth's surface, 2) wind consisting of
-northerly and easterly vector components converted to wind speed using the Euclidean norm,
-3) barometric pressure at sea level, and 4) precipitation (Table
+In addition to sex and age class, we gathered additional variables that
+might help explain variation in haul-out probabilities. These included
+day-of-year (for seasonal effects) and local solar hour (for diurnal
+effects). Solar hour was calculated using the {solaR} package
+[@perpinan2012] within the R statistical environment
+[@rcoreteam2021] using the daily locations. We also compiled several
+meteorological variables shown to affect haul-out behavior in other
+Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017]. In
+particular, we linked locations from satellite tags to key weather
+values from the North American Regional Reanalysis (NARR) model produced
+by the National Centers for Environmental Prediction
+[@mesinger2006]. The NARR model assimilates observational data to
+produce a long-term picture of weather over North America. Numerous
+weather variables are made available across the region 8 times daily.
+For this study, NARR weather values were subset to the extent of our
+study area over the Bering and Chukchi Seas at 3-hr intervals based on
+the native grid resolution of 32 km (1024 sq. km). The following
+meteorological variables were interpolated and assigned to seal
+locations using a bilinear method: 1) air temperature at 2m above the
+earth's surface, 2) wind consisting of northerly and easterly vector
+components converted to wind speed using the Euclidean norm, 3)
+barometric pressure at sea level, and 4) precipitation (Table
 \@ref(tab:tblCovariates)).
 
-Models for all species include the following effects: day-of-year, solar hour,
-temperature, wind speed, barometric pressure, precipitation, wind chill (represented by a
-*wind:temperature* interaction [@udevitz2009cjz]), and day-of-year and time of day
-interactions to permit diurnal patterns to change throughout the year. Spotted and ribbon
-seal models included age-sex class and interactions between day-of-year and age-sex class,
-but we omitted these from bearded seal models due to poor representation of age-sex
-classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
-(and an interaction with day-of-year), since bearded seals occupy a substantial range and
-we were interested in possible differences in the timing of haul-out behavior along a
-latitudinal gradient. We omitted the latitudinal effect from ribbon and spotted seal
-models since, during the spring, these species are most prevalent near the southern ice
-edge in the Bering Sea [@conn2014mee].
-
-Notably missing from this list of explanatory variables is any spatial-temporal
-representation of sea ice concentration, area, or extent. This may seem counterintuitive
-when modeling the haul-out behavior of seal species with such a close association to sea
-ice; seals haul out in the presence of sea ice, and we could assess the local
-concentration of sea ice during these events (see [@olnes2020]). However, a major focus of
-this study is to develop models applicable for aerial survey correction factors and using
-sea ice as a covariate would almost certainly bias haul-out predictions towards those
-seals that are on or near ice and therefore more likely to haul out. Since aerial surveys
-can only detect seals on ice, abundance estimates would be missing a correction for those
-seals that are away from ice (e.g., on foraging trips). Additionally, many of the
-bio-logger deployments used in this study were single attachments to the rear flipper of
-the seal and do not provide at-sea locations, limiting our ability to fully evaluate
-habitat preferences related to sea-ice. It should be noted that the aerial survey counts
-are modeled with sea-ice and other habitat parameters. So, any final abundance estimates
-that combine models of survey counts and our model for proportion of seals hauled out on
-ice do account for the strong association between these species and sea-ice in the Bering
-and Chukchi seas. Lastly, our study was limited to the spring season when seal haul-out
-tendencies are strongly influenced by pupping, nursing, breeding, and molt and these
-drivers are likely more influential than specific sea-ice concentration. Crawford et al
-[-@crawford2019] compared haul-out probability models for bearded seals and found those
-that only included season (and not sea-ice concentration) were the most parsimonious. For
-these reasons, we have elected not to use sea ice concentration as a predictor for
-haul-out probability; studies to evaluate habitat selection in relation to sea-ice
-parameters and other environmental covariates are in progress.
-
-We explored whether the annual variation in maximum spring sea ice extent in the Bering
-Sea influences the seasonal peak of seal haul-out behavior. In particular, we used sea ice
-concentration data from the Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave Dataset,
-Version 1 [@cavalieri1996] to calculate maximum sea ice extent. All sea ice concentration
-grid cells (25 km^2^) in the study area with greater than 15% concentration were summed to
-get the total sea ice extent for each day between 15 February and 15 July across all
-years. Maximum spring sea ice extent was simply the largest daily value for each year.
+Models for all species include the following effects: day-of-year, solar
+hour, temperature, wind speed, barometric pressure, precipitation, wind
+chill (represented by a *wind:temperature* interaction
+[@udevitz2009]), and day-of-year and time of day interactions to
+permit diurnal patterns to change throughout the year. Spotted and
+ribbon seal models included age-sex class and interactions between
+day-of-year and age-sex class, but we omitted these from bearded seal
+models due to poor representation of age-sex classes (Table
+\@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
+(and an interaction with day-of-year), since bearded seals occupy a
+substantial range and we were interested in possible differences in the
+timing of haul-out behavior along a latitudinal gradient. We omitted the
+latitudinal effect from ribbon and spotted seal models since, during the
+spring, these species are most prevalent near the southern ice edge in
+the Bering Sea [@conn2014a].
+
+Notably missing from this list of explanatory variables is any
+spatial-temporal representation of sea ice concentration, area, or
+extent. This may seem counterintuitive when modeling the haul-out
+behavior of seal species with such a close association to sea ice; seals
+haul out in the presence of sea ice, and we could assess the local
+concentration of sea ice during these events (see [@olnes2020]).
+However, a major focus of this study is to develop models applicable for
+aerial survey correction factors and using sea ice as a covariate would
+almost certainly bias haul-out predictions towards those seals that are
+on or near ice and therefore more likely to haul out. Since aerial
+surveys can only detect seals on ice, abundance estimates would be
+missing a correction for those seals that are away from ice (e.g., on
+foraging trips). Additionally, many of the bio-logger deployments used
+in this study were single attachments to the rear flipper of the seal
+and do not provide at-sea locations, limiting our ability to fully
+evaluate habitat preferences related to sea-ice. It should be noted that
+the aerial survey counts are modeled with sea-ice and other habitat
+parameters. So, any final abundance estimates that combine models of
+survey counts and our model for proportion of seals hauled out on ice do
+account for the strong association between these species and sea-ice in
+the Bering and Chukchi seas. Lastly, our study was limited to the spring
+season when seal haul-out tendencies are strongly influenced by pupping,
+nursing, breeding, and molt and these drivers are likely more
+influential than specific sea-ice concentration. Crawford et al
+[-@crawford2019] compared haul-out probability models for bearded seals
+and found those that only included season (and not sea-ice
+concentration) were the most parsimonious. For these reasons, we have
+elected not to use sea ice concentration as a predictor for haul-out
+probability; studies to evaluate habitat selection in relation to
+sea-ice parameters and other environmental covariates are in progress.
+
+We explored whether the annual variation in maximum spring sea ice
+extent in the Bering Sea influences the seasonal peak of seal haul-out
+behavior. In particular, we used sea ice concentration data from the
+Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave Dataset, Version 1
+[@cavalieri1996] to calculate maximum sea ice extent. All sea ice
+concentration grid cells (25 km^2^) in the study area with greater than
+15% concentration were summed to get the total sea ice extent for each
+day between 15 February and 15 July across all years. Maximum spring sea
+ice extent was simply the largest daily value for each year.
 
@@
 
-Haul-out records for seals are often characterized by multiple hours spent basking on ice
-alternating with long periods of foraging (Figure \@ref(fig:examplePlot)). Commonly used
-statistical models for binary data (e.g. logistic regression) assume independence among
-responses, an assumption that is clearly violated if hourly responses are modeled. Any
-analysis that ignores temporal dependence in responses will thus have overstated precision
-[@betts2006].
-
-To properly account for temporal dependence within a computationally tractable framework,
-we used generalized linear mixed pseudo-models (GLMPMs; [@verhoef2010cs]) to model
-variation in haul-out behavior as a function of (1) covariate predictors, (2) temporally
-autocorrelated random effects, and (3) individual random effects representing
-heterogeneity in individual behavior. We used the {glmmLDTS} package [@verhoef2010cs] to
-implement GLMPMs in the R programming environment [@rcoreteam2021]. We explored two
-different model formulations for our data, and owing to the large number of records, we
-fit separate models to bearded, ribbon, and spotted seal data sets. In our first model
-formulation, for each species, we fitted a year-independent model that predicted average
-haul-out behavior as a function of demographic, environmental, seasonal, and diurnal
-effects. Second, for ribbon and spotted seals (which had considerably more data than
-bearded seals), we fitted models that included all the effects from the first model, but
-also permitted annual variation in haul-out timing. This second class of models was used
-to examine whether haul-out patterns varied by year and to determine the annual timing of
-apparent peaks in haul-out behavior.
-
-We followed Ver Hoef et al. [-@verhoef2014sm] in using linear, quadratic, and cubic
-effects of day-of-year to represent seasonal changes in behavior. However, unlike previous
-models for harbor seals [@london2012] and ice-associated seals [@verhoef2014sm], which
-treated hour-of-day as a 24-level categorical variable to capture diurnal cycles, we
-adopted a continuous formulation based on Fourier series that provides a flexible model
-while preserving the inherent circularity needed for time-of-day effects (i.e., hour 0
-should be equal to hour 24). It also represents hour-of-day with 6 parameters, which is a
-considerable reduction when compared to a 24 parameter variable, especially when fitting
-models with interactions between hour-of-day and other variables (e.g., age-sex class,
-day-of-year). According to this approach, we used the following specification for
-hour-of-day effects:
+Haul-out records for seals are often characterized by multiple hours
+spent basking on ice alternating with long periods of foraging (Figure
+\@ref(fig:examplePlot)). Commonly used statistical models for binary
+data (e.g. logistic regression) assume independence among responses, an
+assumption that is clearly violated if hourly responses are modeled. Any
+analysis that ignores temporal dependence in responses will thus have
+overstated precision [@betts2006].
+
+To properly account for temporal dependence within a computationally
+tractable framework, we used generalized linear mixed pseudo-models
+(GLMPMs; [@verhoef2010a]) to model variation in haul-out behavior as a
+function of (1) covariate predictors, (2) temporally autocorrelated
+random effects, and (3) individual random effects representing
+heterogeneity in individual behavior. We used the {glmmLDTS} package
+[@verhoef2010a] to implement GLMPMs. We explored two different model formulations for our
+data, and owing to the large number of records, we fit separate models
+to bearded, ribbon, and spotted seal data sets. In our first model
+formulation, for each species, we fitted a year-independent model that
+predicted average haul-out behavior as a function of demographic,
+environmental, seasonal, and diurnal effects. Second, for ribbon and
+spotted seals (which had considerably more data than bearded seals), we
+fitted models that included all the effects from the first model, but
+also permitted annual variation in haul-out timing. This second class of
+models was used to examine whether haul-out patterns varied by year and
+to determine the annual timing of apparent peaks in haul-out behavior.
+
+We followed Ver Hoef et al. [-@verhoef2014a] in using linear,
+quadratic, and cubic effects of day-of-year to represent seasonal
+changes in behavior. However, unlike previous models for harbor seals
+[@london2012] and ice-associated seals [@verhoef2014a], which treated
+hour-of-day as a 24-level categorical variable to capture diurnal
+cycles, we adopted a continuous formulation based on Fourier series that
+provides a flexible model while preserving the inherent circularity
+needed for time-of-day effects (i.e., hour 0 should be equal to hour
+24). It also represents hour-of-day with 6 parameters, which is a
+considerable reduction when compared to a 24 parameter variable,
+especially when fitting models with interactions between hour-of-day and
+other variables (e.g., age-sex class, day-of-year). According to this
+approach, we used the following specification for hour-of-day effects:
 
@@
 H_t=\alpha_1 \cos(\frac{\pi t}{4}) + \alpha_2 \sin(\frac{\pi t}{4}) + \alpha_3 \cos(\frac{\pi t}{6}) + \alpha_4 \sin(\frac{\pi t}{6}) + \alpha_5 \cos(\frac{\pi t}{12}) + \alpha_6 \sin(\frac{\pi t}{12})
-$$ 
-where $H_t$ gives the effect for solar hour $t$ and $α_{i}$ are estimated parameters
-(regression coefficients).
-
-For the second set of models examining inter-annual variation in sea ice use, we fitted
-models with year by day-of-year interactions. However, in this case we only included
-*year:day* and *year:day2*, omitting the main effects of year as well as *year:day3*
-interactions because models with the latter effects were numerically unstable. However,
-the modeled interactions were sufficient to allow shifts in haul-out distribution, as one
-can show mathematically that a simple horizontal shift in timing of haul-out distributions
-does not affect the main effects or cubic terms in a polynomial regression model.
-
-A typical model fitting exercise would also include a model selection process. However,
-AIC (and similar criteria) is not suitable when using pseudo-likelihoods, because
-pseudo-data generated in the model fitting process [@verhoef2010cs] differ between models
-[@teneyck2017]. After fitting GLMPM models, we instead used "type III" $F$-tests to
-calculate $p$-values [@verhoef2010cs] to evaluate model performance and important terms.
-We also produced predictions of haul-out behavior as a function of influential predictors
-(e.g. solar hour, day-of-year, age-sex). Weather covariates for these predictions were
-based on daily or hourly smoothed weather covariate values across the study region. Such
-predictions were then used to develop haul-out probability surfaces, explore conditional
-effects of weather covariates, and determine annual peaks in haul-out activity. The timing
-of peak haul-out behavior was further used to regress against the annual maximum sea ice
-extent in the study region.
+$$ where $H_t$ gives the effect for solar hour $t$ and $α_{i}$ are
+estimated parameters (regression coefficients).
+
+For the second set of models examining inter-annual variation in sea ice
+use, we fitted models with year by day-of-year interactions. However, in
+this case we only included *year:day* and *year:day2*, omitting the main
+effects of year as well as *year:day3* interactions because models with
+the latter effects were numerically unstable. However, the modeled
+interactions were sufficient to allow shifts in haul-out distribution,
+as one can show mathematically that a simple horizontal shift in timing
+of haul-out distributions does not affect the main effects or cubic
+terms in a polynomial regression model.
+
+A typical model fitting exercise would also include a model selection
+process. However, AIC (and similar criteria) is not suitable when using
+pseudo-likelihoods, because pseudo-data generated in the model fitting
+process [@verhoef2010a] differ between models [@teneyck2018]. After
+fitting GLMPM models, we instead used "type III" $F$-tests to calculate
+$p$-values [@verhoef2010a] to evaluate model performance and important
+terms. We also produced predictions of haul-out behavior as a function
+of influential predictors (e.g. solar hour, day-of-year, age-sex).
+Weather covariates for these predictions were based on daily or hourly
+smoothed weather covariate values across the study region. Such
+predictions were then used to develop haul-out probability surfaces,
+explore conditional effects of weather covariates, and determine annual
+peaks in haul-out activity. The timing of peak haul-out behavior was
+further used to regress against the annual maximum sea ice extent in the
+study region.
 
@@
 
-Models omitting year effects suggested that day-of-year, solar hour, age-sex class,
-temperature, and wind substantially influenced haul-out behavior of all three species,
-with $F$ tests producing $p$-values less than 0.05 for variables embodying these effects
-and/or their interactions. Haul-out probabilities typically increased throughout March and
-April, reaching a peak in May and early June before declining again. Diurnal patterns were
-present, with maximum haul-out behavior centered around local solar noon.
+Models omitting year effects suggested that day-of-year, solar hour,
+age-sex class, temperature, and wind substantially influenced haul-out
+behavior of all three species, with $F$ tests producing $p$-values less
+than 0.05 for variables embodying these effects and/or their
+interactions. Haul-out probabilities typically increased throughout
+March and April, reaching a peak in May and early June before declining
+again. Diurnal patterns were present, with maximum haul-out behavior
+centered around local solar noon.
 
@@
 
-Ribbon seals show a pattern of gradually increasing haul-out probability in April that
-peaks in late May for subadults and in early June for adults (Figure
-\@ref(fig:ribbonHOCal)). The behavior is clearly centered around local solar noon and
-expands to other hours later in the season as seals enter their molting period. Subadults
-showed an earlier start and more intense haul-out activity in April and May. The
-young-of-the-year records begin after weaning and the model predictions demonstrate the
-ontogeny of in-water activities (e.g. diving, foraging) in May. Adult females have a more
-protracted haul-out season compared to males, and more time spent resting in June and
-July. The model suggests adult male ribbon seals complete their molt by the end of June.
-Lastly, for adults of both sexes there is some indication for a shift to a crepuscular
-haul-out pattern in late June and July.
+Ribbon seals show a pattern of gradually increasing haul-out probability
+in April that peaks in late May for subadults and in early June for
+adults (Figure \@ref(fig:ribbonHOCal)). The behavior is clearly centered
+around local solar noon and expands to other hours later in the season
+as seals enter their molting period. Subadults showed an earlier start
+and more intense haul-out activity in April and May. The
+young-of-the-year records begin after weaning and the model predictions
+demonstrate the ontogeny of in-water activities (e.g. diving, foraging)
+in May. Adult females have a more protracted haul-out season compared to
+males, and more time spent resting in June and July. The model suggests
+adult male ribbon seals complete their molt by the end of June. Lastly,
+for adults of both sexes there is some indication for a shift to a
+crepuscular haul-out pattern in late June and July.
 
@@
 
-The haul-out probability for ribbon seals was strongly influenced by each of the weather
-covariates included within the model except precipitation. The impacts of temperature
-($F_{`r paste(ribbon_type3$temp2$Num.df, ribbon_type3$temp2$Den.df, sep=",")`}$ =
-`r ribbon_type3$temp2$F.value`; $p$ = `r ribbon_type3$temp2$Prob.F`), wind
-($F_{`r paste(ribbon_type3$wind$Num.df, ribbon_type3$wind$Den.df, sep=",")`}$ =
-`r ribbon_type3$wind$F.value`; $p$ = `r ribbon_type3$wind$Prob.F`), and barometric
-pressure
-($F_{`r paste(ribbon_type3$pressure$Num.df, ribbon_type3$pressure$Den.df, sep=",")`}$ =
-`r ribbon_type3$pressure$F.value`; $p$ = `r ribbon_type3$pressure$Prob.F`) were especially
-noticeable as ribbon seals were less likely to haul out at higher winds and lower pressure
-values and more likely to be on the ice when temperatures were relatively warm. Wind chill
+The haul-out probability for ribbon seals was strongly influenced by
+each of the weather covariates included within the model except
+precipitation. The impacts of temperature
+($F_{`r paste(ribbon_type3$temp2$Num.df, ribbon_type3$temp2$Den.df, sep=",")`}$
+= `r ribbon_type3$temp2$F.value`; $p$ = `r ribbon_type3$temp2$Prob.F`),
+wind
+($F_{`r paste(ribbon_type3$wind$Num.df, ribbon_type3$wind$Den.df, sep=",")`}$
+= `r ribbon_type3$wind$F.value`; $p$ = `r ribbon_type3$wind$Prob.F`),
+and barometric pressure
+($F_{`r paste(ribbon_type3$pressure$Num.df, ribbon_type3$pressure$Den.df, sep=",")`}$
+= `r ribbon_type3$pressure$F.value`; $p$ =
+`r ribbon_type3$pressure$Prob.F`) were especially noticeable as ribbon
+seals were less likely to haul out at higher winds and lower pressure
+values and more likely to be on the ice when temperatures were
+relatively warm. Wind chill
 ($F_{`r paste(ribbon_type3$temp2_wind$Num.df, ribbon_type3$temp2_wind$Den.df, sep=",")`}$
-= `r ribbon_type3$temp2_wind$F.value`; $p$ = `r ribbon_type3$temp2_wind$Prob.F`) showed a
-moderately negative influence on haul-out probability.
-
-Visualizing the marginal effect of each weather covariate on haul-out probability is difficult
-in this analysis because of the collinearity between covariates as well as the temporal variation 
-in covariate values during the study period. The relationship of each weather covariate with haul-out
-probability, averaged over the other weather conditions, is more variable than simply
-showing model coefficients would imply. Figure \@ref(fig:ribbonHOwx)) presents the
-predicted haul-out probability of ribbon seals across the range of weather conditions
-encountered in the observed data. The collinearity and temporal variation are apparent
-in the visual patterns that emerge. By
+= `r ribbon_type3$temp2_wind$F.value`; $p$ =
+`r ribbon_type3$temp2_wind$Prob.F`) showed a moderately negative
+influence on haul-out probability.
+
+Visualizing the marginal effect of each weather covariate on haul-out
+probability is difficult in this analysis because of the collinearity
+between covariates as well as the temporal variation in covariate values
+during the study period. The relationship of each weather covariate with
+haul-out probability, averaged over the other weather conditions, is
+more variable than simply showing model coefficients would imply. Figure
+\@ref(fig:ribbonHOwx)) presents the predicted haul-out probability of
+ribbon seals across the range of weather conditions encountered in the
+observed data. The collinearity and temporal variation are apparent in
+the visual patterns that emerge. By
 
@@
 
-Compared to ribbon seals, spotted seals showed a longer spring haul-out season that is
-less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)). Adults of both sexes spend considerable time in
-April and May resting on the ice. As with ribbon seals, the young-of-the-year records
-begin after weaning and the model predictions demonstrate the ontogeny of in-water
-activities (e.g. diving, foraging) in May.
+Compared to ribbon seals, spotted seals showed a longer spring haul-out
+season that is less intensely centered on solar noon (Figure
+\@ref(fig:spottedHOCal)). Adults of both sexes spend considerable time
+in April and May resting on the ice. As with ribbon seals, the
+young-of-the-year records begin after weaning and the model predictions
+demonstrate the ontogeny of in-water activities (e.g. diving, foraging)
+in May.
 
@@
 
-Spotted seal haul-out behavior appears most strongly influenced by temperature
-($F_{`r paste(spotted_type3$temp2$Num.df, spotted_type3$temp2$Den.df, sep=",")`}$ =
-`r spotted_type3$temp2$F.value`; $p$ = `r spotted_type3$temp2$Prob.F`) and wind
-($F_{`r paste(spotted_type3$wind$Num.df, spotted_type3$wind$Den.df, sep=",")`}$ =
-`r spotted_type3$wind$F.value`; $p$ = `r spotted_type3$wind$Prob.F`) with barometric
-pressure having a moderate influence
-($F_{`r paste(spotted_type3$pressure$Num.df, spotted_type3$pressure$Den.df, sep=",")`}$ =
-`r spotted_type3$pressure$F.value`; $p$ = `r spotted_type3$pressure$Prob.F`). Spotted
-seals were less likely to haul out at higher winds and more likely to be on the ice when
-temperatures were relatively warm. Wind chill (*temperature:wind*) and precipitation were
-not as influential as the other covariates. Differences in the magnitude of response
-between the age-sex classes are present and consistent across each of the weather
-covariates (Figure \@ref(fig:spottedHOwx)).
+Spotted seal haul-out behavior appears most strongly influenced by
+temperature
+($F_{`r paste(spotted_type3$temp2$Num.df, spotted_type3$temp2$Den.df, sep=",")`}$
+= `r spotted_type3$temp2$F.value`; $p$ = `r spotted_type3$temp2$Prob.F`)
+and wind
+($F_{`r paste(spotted_type3$wind$Num.df, spotted_type3$wind$Den.df, sep=",")`}$
+= `r spotted_type3$wind$F.value`; $p$ = `r spotted_type3$wind$Prob.F`)
+with barometric pressure having a moderate influence
+($F_{`r paste(spotted_type3$pressure$Num.df, spotted_type3$pressure$Den.df, sep=",")`}$
+= `r spotted_type3$pressure$F.value`; $p$ =
+`r spotted_type3$pressure$Prob.F`). Spotted seals were less likely to
+haul out at higher winds and more likely to be on the ice when
+temperatures were relatively warm. Wind chill (*temperature:wind*) and
+precipitation were not as influential as the other covariates.
+Differences in the magnitude of response between the age-sex classes are
+present and consistent across each of the weather covariates (Figure
+\@ref(fig:spottedHOwx)).
 
@@
 
-Age and sex class were not included in the model for bearded seals and, as such, results
-are shown for all ages. Unlike ribbon and spotted seals, bearded seals were predicted to
-have a bi-modal distribution of haul-out probability across the day (Figure \@ref(fig:beardedHOCal)). In addition to a peak
-around local solar noon, the bearded seal model predicts additional haul-out activity
-around local midnight. The haul-out behavior is also more protracted throughout the spring
-season compared to ribbon and spotted seals.
+Age and sex class were not included in the model for bearded seals and,
+as such, results are shown for all ages. Unlike ribbon and spotted
+seals, bearded seals were predicted to have a bi-modal distribution of
+haul-out probability across the day (Figure \@ref(fig:beardedHOCal)). In
+addition to a peak around local solar noon, the bearded seal model
+predicts additional haul-out activity around local midnight. The
+haul-out behavior is also more protracted throughout the spring season
+compared to ribbon and spotted seals.
 
@@
 
-Like ribbon and spotted seals, bearded seal haul-out probability appears most influenced
-by wind ($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$ =
-`r bearded_type3$wind$F.value`; $p$ = `r bearded_type3$wind$Prob.F`) and temperature
-($F_{`r paste(bearded_type3$temp2$Num.df, bearded_type3$temp2$Den.df, sep=",")`}$ =
-`r bearded_type3$temp2$F.value`; $p$ = `r bearded_type3$temp2$Prob.F`). Additionally, wind
-chill was also a significant influence
+Like ribbon and spotted seals, bearded seal haul-out probability appears
+most influenced by wind
+($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$
+= `r bearded_type3$wind$F.value`; $p$ = `r bearded_type3$wind$Prob.F`)
+and temperature
+($F_{`r paste(bearded_type3$temp2$Num.df, bearded_type3$temp2$Den.df, sep=",")`}$
+= `r bearded_type3$temp2$F.value`; $p$ =
+`r bearded_type3$temp2$Prob.F`). Additionally, wind chill was also a
+significant influence
 ($F_{`r paste(bearded_type3$temp2_wind$Num.df, bearded_type3$temp2_wind$Den.df, sep=",")`}$
-= `r bearded_type3$temp2_wind$F.value`; $p$ = `r bearded_type3$temp2_wind$Prob.F`). Any
-influence of barometric pressure or precipitation is less apparent (Figure \@ref(fig:beardedHOwx)).
+= `r bearded_type3$temp2_wind$F.value`; $p$ =
+`r bearded_type3$temp2_wind$Prob.F`). Any influence of barometric
+pressure or precipitation is less apparent (Figure
+\@ref(fig:beardedHOwx)).
 
@@
 
-The second set of models that included annual variation in haul-out patterns uncovered
-significant contributions for linear and quadratic interactions between day and year in
-both ribbon seals (day $\times$ year,
+The second set of models that included annual variation in haul-out
+patterns uncovered significant contributions for linear and quadratic
+interactions between day and year in both ribbon seals (day $\times$
+year,
 $F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$
-= `r ribbon_year_type3$day_year$F.value`; $p$ = `r ribbon_year_type3$day_year$Prob.F`;
-day^2^:year,
+= `r ribbon_year_type3$day_year$F.value`; $p$ =
+`r ribbon_year_type3$day_year$Prob.F`; day^2^:year,
 $F_{`r paste(ribbon_year_type3$day2_year$Num.df, ribbon_year_type3$day2_year$Den.df, sep=",")`}$
-= `r ribbon_year_type3$day2_year$F.value`; $p$ = `r ribbon_year_type3$day2_year$Prob.F`)
-and spotted seals (day $\times$ year,
+= `r ribbon_year_type3$day2_year$F.value`; $p$ =
+`r ribbon_year_type3$day2_year$Prob.F`) and spotted seals (day $\times$
+year,
 $F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$
-= `r spotted_year_type3$day_year$F.value`; $p$ = `r spotted_year_type3$day_year$Prob.F`;
-day^2^:year,
+= `r spotted_year_type3$day_year$F.value`; $p$ =
+`r spotted_year_type3$day_year$Prob.F`; day^2^:year,
 $F_{`r paste(spotted_year_type3$day2_year$Num.df, spotted_year_type3$day2_year$Den.df, sep=",")`}$
 = `r spotted_year_type3$day2_year$F.value`; $p$ =
-`r spotted_year_type3$day2_year$Prob.F`). Predicted distributions of haul-out activity
-were largely unimodal, but varied some among and within years with respect to the timing
-and magnitude of haul-out peaks (Figure \@ref(fig:annualHO)). It is important to note that
-predicted variation in annual haul-out patterns likely reflects both process error and
-sampling variability. While we did remove any years where only one deployment in a species
-+ age-sex group was present, there are still some years where the pattern shown is
-informed by a small number of individuals that may not represent population level
-processes.
+`r spotted_year_type3$day2_year$Prob.F`). Predicted distributions of
+haul-out activity were largely unimodal, but varied some among and
+within years with respect to the timing and magnitude of haul-out peaks
+(Figure \@ref(fig:annualHO)). It is important to note that predicted
+variation in annual haul-out patterns likely reflects both process error
+and sampling variability. While we did remove any years where only one
+deployment in a species + age-sex group was present, there are still
+some years where the pattern shown is informed by a small number of
+individuals that may not represent population level processes.
 
@@
 
-The timing of peak haul-out probability for ribbon and spotted seals appeared to have only
-a limited to no relationship with the amount of yearly maximum sea ice extent. Adult
-female and subadult spotted seals show a negative trend line, but there is no indication
-that the observed trend is meaningful (spotted seal adult female: $R^{2}$ =
-`r pl_adultF$r.squared`, $p$ = `r pl_adultF$p.value`; spotted seal subadults: $R^{2}$ =
-`r pl_subadult$r.squared`, $p$ = `r pl_subadult$p.value`). For ribbon seals and adult male
-spotted seals, $p$-values were substantially larger than 0.05 (ribbon seal adult females;
-$R^{2}$ = `r hf_adultF$r.squared`, $p$ = `r hf_adultF$p.value`; ribbon seal adult males:
-$R^{2}$ = `r hf_adultM$r.squared`, $p$ = `r hf_adultM$p.value`; ribbon seal subadults:
-$R^{2}$ = `r hf_subadult$r.squared`, $p$ = `r hf_subadult$p.value`; spotted seals adult
-males: $R^{2}$ = `r pl_adultM$r.squared`, $p$ = `r pl_adultM$p.value`).
+The timing of peak haul-out probability for ribbon and spotted seals
+appeared to have only a limited to no relationship with the amount of
+yearly maximum sea ice extent. Adult female and subadult spotted seals
+show a negative trend line, but there is no indication that the observed
+trend is meaningful (spotted seal adult female: $R^{2}$ =
+`r pl_adultF$r.squared`, $p$ = `r pl_adultF$p.value`; spotted seal
+subadults: $R^{2}$ = `r pl_subadult$r.squared`, $p$ =
+`r pl_subadult$p.value`). For ribbon seals and adult male spotted seals,
+$p$-values were substantially larger than 0.05 (ribbon seal adult
+females; $R^{2}$ = `r hf_adultF$r.squared`, $p$ = `r hf_adultF$p.value`;
+ribbon seal adult males: $R^{2}$ = `r hf_adultM$r.squared`, $p$ =
+`r hf_adultM$p.value`; ribbon seal subadults: $R^{2}$ =
+`r hf_subadult$r.squared`, $p$ = `r hf_subadult$p.value`; spotted seals
+adult males: $R^{2}$ = `r pl_adultM$r.squared`, $p$ =
+`r pl_adultM$p.value`).
 
@@
 
-In this paper, we modeled data from bio-logging sensors deployed on bearded, spotted, and
-ribbon seals to examine factors affecting haul-out behavior on sea ice in the Bering and
-Chukchi seas. Our analysis shows seals use sea ice progressively more through the
-spring, with peak use near mid-May to early June before declining again. This pattern
-aligns well with what has been previously documented and confirms our bio-logging
-deployments are likely capturing population-level behavioral patterns. Seals
-preferentially haul out on ice shortly after solar noon, which allows seals to maximize
-absorption of solar radiation [@feltz1966]. Interestingly, bearded seals appear to have
-two peaks in haul-out activity across the day, one shortly after solar noon, and one
-centered near solar midnight. A similar bi-modal pattern has been seen in ringed seals
-[@vonduyke2020] and suggests that bearded and ringed seals may be operating under
-different constraints than ribbon and spotted seals. Bearded and ringed seals are
-distributed across higher latitudes and the extended daylight hours may allow more
-flexibility in optimizing resting periods with foraging.
-
-Unlike previous analyses of seal haul-out data (e.g. [@verhoef2010cs], [@conn2014mee]), we
-also investigated the influence of sex-age class on haul-out probabilities for all species
-except bearded seals because of low sample size. While both ribbon and spotted seals
-exhibited a unimodal diel haul-out pattern generally centered around local solar noon,
-there were key differences across species, age, and sex that match expectations given what
-we know of their ecological behavior. Spotted seals are known to form triads during the
-breeding season [@burns2009, @mcclintock2015] where a female and dependent pup are
-accompanied on the ice by a suitor male. The male waits for the female to wean the pup and
-enter estrus, and fends off any other males that may be interested. Triad formation results
-in both males and females spending a large portion of the day hauled out on ice and a
-protracted spring haul-out season for both sexes. Because mating occurs in the water,
-females may also be less inclined to venture out for foraging trips while still nursing
-the pup. We see this reflected in the predicted haul-out behavior, with both males and
-females exhibiting a broad distribution of time out of the water throughout the solar day
-and the season. Ribbon seals are not known to form triads and our model predicts a
-progression of increased haul-out behavior with females starting earlier in the season
-than males. Notably, female ribbon seals spend a large portion of the day in the water
-during the pupping period, aligning with the hypothesis that ribbon seal females continue
-foraging while nursing. In the case of both ribbon and spotted seals, subadults are the
-first to begin consistent haul-out behavior and follow a typical phocid pattern where
-subadults molt first as they do not have any reproductive constraints.
-
-We also investigated the influence of environmental covariates on haul-out probabilities,
-including wind speed, temperature, barometric pressure, precipitation, and wind chill.
-These have been investigated for walrus (e.g. Udevitz et al. [@udevitz2009cjz]) and a few
-select studies of ice-associated seals [@hamilton2018, @perry2017]. Ribbon seals seemed to
-be the most influenced by weather, with wind, temperature, barometric pressure, and
-precipitation all being important components of the model. Spotted seals were most
-affected by wind and barometric pressures. And, for bearded seals, the model indicated
-wind and temperature had the greatest impact. In general, and as might be expected, seals
-were more likely to haul-out when daily temperatures were warmer, winds speeds were lower,
-barometric pressure was higher, and precipitation was lower. These weather conditions
-are likely most beneficial for energetic considerations and may also provide the best
-situation for predator detection. These results highlight the importance for haul-out
-behavior models (and subsequent calculations of availability during surveys) to
-incorporate weather covariates.
-
-Our model uncovered annual deviations in the timing and magnitude of haul-out peaks for
-ribbon and spotted seals. The timing of peak haul-out activity appears to fall within a
-relatively narrow time window of 3-4 weeks in late May and early June. This remarkable
-level of consistency across 15 years is indicative of the adaptations these species have
-evolved to synchronize their key life history stages with favorable environmental
-conditions (e.g. availability of sea ice, increased oceanographic productivity, favorable
-weather conditions). Unfortunately, this may also be indicative of a limited range of
-plasticity as climatological conditions change. Additionally, while from an ecological
-perspective the haul-out behavior appears consistent, the interannual differences in
-timing and magnitude are large enough to have important ramifications on calculations of
-abundance and trend.
-
-Previous attempts to estimate the abundance of phocid seals from aerial survey data in the
-Bering and Chukchi Seas (e.g. [@bengtson2005pb], [@conn2014mee], [@verhoef2014sm]) have
-used estimated haul-out probabilities to correct for the proportion of animals that are in
-the water and thus unavailable to be counted during aerial surveys. Although several of
-these studies allowed haul-out probabilities to vary by day-of-year and time-of-day,
-previous correction factors have not accounted for variability between years, different
-weather conditions, and the age-sex class of the sample. In this paper, we have shown that
-there can be considerable differences in the number of seals hauled out on ice based on
-these factors. We recommend that future abundance analyses employ correction factors that
-account for these factors. For instance, it is relatively straightforward to obtain
-weather reanalysis products for times and locations that are surveyed and to construct a
-relevant correction factor based on predictions of GLMPMs. The most challenging element in
-developing availability correction factors is with annual variability. It can be difficult
-to get a sufficient sample size to estimate year-specific correction factors, particularly
-because research teams would likely need to tag seals and conduct aerial surveys
-concurrently which requires considerably more personnel and money. One possible suggestion
-is to estimate a "shift" parameter within models for aerial survey counts that allow the
-peak of haul-out distributions to be adjusted earlier or later in the year based on the
-frequency of counts observed over time. Regardless, researchers should anticipate there
-being some unmodeled heterogeneity in availability probability present in abundance
-estimates obtained from aerial surveys. This may make trend detection difficult, as one
-will not know if moderate differences in abundance estimates are attributable to changes
-in abundance or changes in haul-out frequencies.
-
-Predictions of absolute haul-out probability in this paper were somewhat different than
-those previously reported for these species, especially for bearded seals. For instance,
-Ver Hoef et al. [-@verhoef2014sm)] and Conn et al. [-@conn2014mee] used haul-out
-correction factors with maximums of 0.66 for bearded seals, 0.62 for ribbon seals, and
-0.54 for spotted seals, where maximums corresponded to times near solar noon in mid-late
-May. Applying models that ignore age, sex, and year effects, these probabilities were
-0.38, 0.72, and 0.60, respectively, under the current analysis framework. Our current
-estimates reflect increased sample sizes in terms of number of animals, but also
-improvements to the way data were prepared prior to analysis.
-
-We focused this paper on haul-out behavior of bearded, ribbon, and spotted seals. Ringed
-seals are also present in the Bering and Chukchi Seas but exhibit qualitatively different
-haul-out behavior. For instance, many ringed seals build subnivean layers in the snow on
-top of the sea ice, which they frequently inhabit before snow melt causes their lairs to
-collapse [@kelly1990]. Thus, the wet-dry sensor on a satellite tag could indicate that an
-animal is hauled out, but it could still be unavailable to be detected because it is in a
-lair. We hope to address availability of ringed seals using data from satellite tags,
-replicate survey tracks, and auxiliary information about snow depth and timing of melt in
-a future study.
+In this paper, we modeled data from bio-logging sensors deployed on
+bearded, spotted, and ribbon seals to examine factors affecting haul-out
+behavior on sea ice in the Bering and Chukchi seas. Our analysis shows
+seals use sea ice progressively more through the spring, with peak use
+near mid-May to early June before declining again. This pattern aligns
+well with what has been previously documented and confirms our
+bio-logging deployments are likely capturing population-level behavioral
+patterns. Seals preferentially haul out on ice shortly after solar noon,
+which allows seals to maximize absorption of solar radiation
+[@feltz1966]. Interestingly, bearded seals appear to have two peaks in
+haul-out activity across the day, one shortly after solar noon, and one
+centered near solar midnight. A similar bi-modal pattern has been seen
+in ringed seals [@vonduyke2020] and suggests that bearded and ringed
+seals may be operating under different constraints than ribbon and
+spotted seals. Bearded and ringed seals are distributed across higher
+latitudes and the extended daylight hours may allow more flexibility in
+optimizing resting periods with foraging.
+
+Unlike previous analyses of seal haul-out data (e.g. [@verhoef2010a],
+[@conn2014a]), we also investigated the influence of sex-age class
+on haul-out probabilities for all species except bearded seals because
+of low sample size. While both ribbon and spotted seals exhibited a
+unimodal diel haul-out pattern generally centered around local solar
+noon, there were key differences across species, age, and sex that match
+expectations given what we know of their ecological behavior. Spotted
+seals are known to form triads during the breeding season [@burns2009,
+@mcclintock2015] where a female and dependent pup are accompanied on the
+ice by a suitor male. The male waits for the female to wean the pup and
+enter estrus, and fends off any other males that may be interested.
+Triad formation results in both males and females spending a large
+portion of the day hauled out on ice and a protracted spring haul-out
+season for both sexes. Because mating occurs in the water, females may
+also be less inclined to venture out for foraging trips while still
+nursing the pup. We see this reflected in the predicted haul-out
+behavior, with both males and females exhibiting a broad distribution of
+time out of the water throughout the solar day and the season. Ribbon
+seals are not known to form triads and our model predicts a progression
+of increased haul-out behavior with females starting earlier in the
+season than males. Notably, female ribbon seals spend a large portion of
+the day in the water during the pupping period, aligning with the
+hypothesis that ribbon seal females continue foraging while nursing. In
+the case of both ribbon and spotted seals, subadults are the first to
+begin consistent haul-out behavior and follow a typical phocid pattern
+where subadults molt first as they do not have any reproductive
+constraints.
+
+We also investigated the influence of environmental covariates on
+haul-out probabilities, including wind speed, temperature, barometric
+pressure, precipitation, and wind chill. These have been investigated
+for walrus (e.g. Udevitz et al. [-@udevitz2009]) and a few select
+studies of ice-associated seals [@hamilton2018, @perry2017]. Ribbon
+seals seemed to be the most influenced by weather, with wind,
+temperature, barometric pressure, and precipitation all being important
+components of the model. Spotted seals were most affected by wind and
+barometric pressures. And, for bearded seals, the model indicated wind
+and temperature had the greatest impact. In general, and as might be
+expected, seals were more likely to haul-out when daily temperatures
+were warmer, winds speeds were lower, barometric pressure was higher,
+and precipitation was lower. These weather conditions are likely most
+beneficial for energetic considerations and may also provide the best
+situation for predator detection. These results highlight the importance
+for haul-out behavior models (and subsequent calculations of
+availability during surveys) to incorporate weather covariates.
+
+Our model uncovered annual deviations in the timing and magnitude of
+haul-out peaks for ribbon and spotted seals. The timing of peak haul-out
+activity appears to fall within a relatively narrow time window of 3-4
+weeks in late May and early June. This remarkable level of consistency
+across 15 years is indicative of the adaptations these species have
+evolved to synchronize their key life history stages with favorable
+environmental conditions (e.g. availability of sea ice, increased
+oceanographic productivity, favorable weather conditions).
+Unfortunately, this may also be indicative of a limited range of
+plasticity as climatological conditions change. Additionally, while from
+an ecological perspective the haul-out behavior appears consistent, the
+interannual differences in timing and magnitude are large enough to have
+important ramifications on calculations of abundance and trend.
+
+Previous attempts to estimate the abundance of phocid seals from aerial
+survey data in the Bering and Chukchi Seas (e.g. [@bengtson2005a],
+[@conn2014a], [@verhoef2014a]) have used estimated haul-out
+probabilities to correct for the proportion of animals that are in the
+water and thus unavailable to be counted during aerial surveys. Although
+several of these studies allowed haul-out probabilities to vary by
+day-of-year and time-of-day, previous correction factors have not
+accounted for variability between years, different weather conditions,
+and the age-sex class of the sample. In this paper, we have shown that
+there can be considerable differences in the number of seals hauled out
+on ice based on these factors. We recommend that future abundance
+analyses employ correction factors that account for these factors. For
+instance, it is relatively straightforward to obtain weather reanalysis
+products for times and locations that are surveyed and to construct a
+relevant correction factor based on predictions of GLMPMs. The most
+challenging element in developing availability correction factors is
+with annual variability. It can be difficult to get a sufficient sample
+size to estimate year-specific correction factors, particularly because
+research teams would likely need to tag seals and conduct aerial surveys
+concurrently which requires considerably more personnel and money. One
+possible suggestion is to estimate a "shift" parameter within models for
+aerial survey counts that allow the peak of haul-out distributions to be
+adjusted earlier or later in the year based on the frequency of counts
+observed over time. Regardless, researchers should anticipate there
+being some unmodeled heterogeneity in availability probability present
+in abundance estimates obtained from aerial surveys. This may make trend
+detection difficult, as one will not know if moderate differences in
+abundance estimates are attributable to changes in abundance or changes
+in haul-out frequencies.
+
+Predictions of absolute haul-out probability in this paper were somewhat
+different than those previously reported for these species, especially
+for bearded seals. For instance, Ver Hoef et al. [-@verhoef2014a)] and
+Conn et al. [-@conn2014a] used haul-out correction factors with
+maximums of 0.66 for bearded seals, 0.62 for ribbon seals, and 0.54 for
+spotted seals, where maximums corresponded to times near solar noon in
+mid-late May. Applying models that ignore age, sex, and year effects,
+these probabilities were 0.38, 0.72, and 0.60, respectively, under the
+current analysis framework. Our current estimates reflect increased
+sample sizes in terms of number of animals, but also improvements to the
+way data were prepared prior to analysis.
+
+We focused this paper on haul-out behavior of bearded, ribbon, and
+spotted seals. Ringed seals are also present in the Bering and Chukchi
+Seas but exhibit qualitatively different haul-out behavior. For
+instance, many ringed seals build subnivean layers in the snow on top of
+the sea ice, which they frequently inhabit before snow melt causes their
+lairs to collapse [@kelly1990]. Thus, the wet-dry sensor on a satellite
+tag could indicate that an animal is hauled out, but it could still be
+unavailable to be detected because it is in a lair. We hope to address
+availability of ringed seals using data from satellite tags, replicate
+survey tracks, and auxiliary information about snow depth and timing of
+melt in a future study.
 
@@
 
-**Josh M. London**: investigation, conceptualization, methodology, formal analysis,
-validation, software, writing: original draft, writing: review and editing, visualization,
-and data curation
+**Josh M. London**: investigation, conceptualization, methodology,
+formal analysis, validation, software, writing: original draft, writing:
+review and editing, visualization, and data curation
 
 **Paul B. Conn**: conceptualization, methodology, formal analysis,
-software, validation, writing: original draft, writing: review and editing
-
-**Stacie K. Hardy**: investigation, data curation, methodology, validation, writing: review and
+software, validation, writing: original draft, writing: review and
 editing
 
-**Erin L. Richmond**: data curation, investigation, methodology, validation,
-writing: review and editing
-
-**Jay M. Ver Hoef**: conceptualization, methodology, software,
-writing: review and editing
-
-**Irina S. Trukhanova**: conceptualization, writing: review
-and editing
-
-**Michael F. Cameron**: investigation, project administration, writing: review
-and editing
-
-**Peter L. Boveng**: investigation, conceptualization, supervision, project
-administration, writing: review and editing
+**Stacie K. Hardy**: investigation, data curation, methodology,
+validation, writing: review and editing
+
+**Erin L. Richmond**: data curation, investigation, methodology,
+validation, writing: review and editing
+
+**Jay M. Ver Hoef**: conceptualization, methodology, software, writing:
+review and editing
+
+**Irina S. Trukhanova**: conceptualization, writing: review and editing
+
+**Michael F. Cameron**: investigation, project administration, writing:
+review and editing
+
+**Peter L. Boveng**: investigation, conceptualization, supervision,
+project administration, writing: review and editing
 
@@
 
-Data and code will be available as a complete research compendium on GitHub and major versions
-archived at Zenodo. Original data sources for telemetry are archived at the United
-States Animal Telemetry Network.
+Data and code will be available as a complete research compendium on
+GitHub and major versions archived at Zenodo. Original data sources for
+telemetry are archived at the United States Animal Telemetry Network.
 
@@
 
-We recognize that the species and ecosystems studied in this paper are within the
-traditional range of the Inpuiaq and Yupik people. And, we would like to acknowledge and
-thank them as the original stewards of these ecosystems. The deployment of bio-logging
-devices used in this study were often done in collaboration with Alaska Native
-communities. We would like to especially acknowledge the communities of Kotzebue,
-Utqiagvik, and Ulguniq and the following individuals: James Adams, Jeff Barger, David
-Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris, Grover Harris, Lee Harris, Tom
-Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin, Pearl Goodwin, Willie
-Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr., Virgil Naylor Sr., Dan Savetilik,
-Allen Stone, and Randy Toshavik from Kotzebue, Alaska; Billy Adams, Mary-Ellen Ahmaogak,
-James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard Kittick, Gilbert Leavitt,
-Isaac Leavitt, J.R. Leavitt, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, Fred Rexford,
-Chuck Schaeffer, Ross Schaeffer, Bob Shears, and Joe Skin from the North Slope Borough,
+We recognize that the species and ecosystems studied in this paper are
+within the traditional range of the Inpuiaq and Yupik people. And, we
+would like to acknowledge and thank them as the original stewards of
+these ecosystems. The deployment of bio-logging devices used in this
+study were often done in collaboration with Alaska Native communities.
+We would like to especially acknowledge the communities of Kotzebue,
+Utqiagvik, and Ulguniq and the following individuals: James Adams, Jeff
+Barger, David Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris,
+Grover Harris, Lee Harris, Tom Jones, Frank Garfield, Brenda Goodwin,
+Henry Goodwin, John Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk,
+Noah Naylor, Virgil Naylor Jr., Virgil Naylor Sr., Dan Savetilik, Allen
+Stone, and Randy Toshavik from Kotzebue, Alaska; Billy Adams, Mary-Ellen
+Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard
+Kittick, Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik,
+Shawn Oktollik, Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross
+Schaeffer, Bob Shears, and Joe Skin from the North Slope Borough,
 Alaska;
 
-We are grateful for the assistance in catching and sampling seals by James Bailey,
-Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov, Cynthia Christman, Shawn
-Dahle, Rob Delong, Stacy DiRocco, Deb Fauquier, Shannon Fitzgerald, Kathy Frost, Scott
-Gende, Tracey Goldstein, Jeff Harris, Markus Horning, John Jansen, Shawn Johnson, Charles
-Littnan, Lloyd Lowry, Erin Moreland, Lorrie Rea, Gay Sheffield, Brent Stewart, and Dave
-Withrow. We also appreciate the commitment to science and safety by all officers and crew
-of the NOAA Ship Oscar Dyson, the NOAA Ship MacArthur II, and the RV Thomas G. Thompson.
-
-Telemetry data from the Alaska Department of Fish and Game and the North Slope Borough
-Department of Wildlife Management were important contributions to the findings presented
-here. Both agencies collaborate closely with Alaska Native communities and the
-contributions of local hunters to the successful capture, deployment, and understanding of seal ecology are
-invaluable. Deployments in the western Bering Sea and Sea of Okhotsk were done in
-collaboration with Russian colleagues and North Pacific Wildlife.
-
-The findings and conclusions in the paper are those of the author(s) and do not
-necessarily represent the views of the National Marine Fisheries Service, NOAA. Any use of
-trade, product, or firm names does not imply an endorsement by the U.S. Government.
-Funding for this study was provided by the U.S. National Oceanic and Atmospheric
-Administration. The field work was conducted under the authority of Marine Mammal
-Protection Act Research Permits Nos. 782-1676, 782-1765, 15126, and 19309 issued by the
-National Marine Fisheries Service, and Letters of Assurance of Compliance with Animal
-Welfare Act regulations, Nos. A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries
-Science Center/Northwest Fisheries Science Center Institutional Animal Care and Use
-Committee.
+We are grateful for the assistance in catching and sampling seals by
+James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir
+Burkanov, Cynthia Christman, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
+Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey
+Goldstein, Jeff Harris, Markus Horning, John Jansen, Shawn Johnson,
+Charles Littnan, Lloyd Lowry, Erin Moreland, Lorrie Rea, Gay Sheffield,
+Brent Stewart, and Dave Withrow. We also appreciate the commitment to
+science and safety by all officers and crew of the NOAA Ship Oscar
+Dyson, the NOAA Ship MacArthur II, and the RV Thomas G. Thompson.
+
+Telemetry data from the Alaska Department of Fish and Game and the North
+Slope Borough Department of Wildlife Management were important
+contributions to the findings presented here. Both agencies collaborate
+closely with Alaska Native communities and the contributions of local
+hunters to the successful capture, deployment, and understanding of seal
+ecology are invaluable. Deployments in the western Bering Sea and Sea of
+Okhotsk were done in collaboration with Russian colleagues and North
+Pacific Wildlife.
+
+The findings and conclusions in the paper are those of the author(s) and
+do not necessarily represent the views of the National Marine Fisheries
+Service, NOAA. Any use of trade, product, or firm names does not imply
+an endorsement by the U.S. Government. Funding for this study was
+provided by the U.S. National Oceanic and Atmospheric Administration.
+The field work was conducted under the authority of Marine Mammal
+Protection Act Research Permits Nos. 782-1676, 782-1765, 15126, and
+19309 issued by the National Marine Fisheries Service, and Letters of
+Assurance of Compliance with Animal Welfare Act regulations, Nos. A/NW
+2010-3 and A/NW 2016-1 from the Alaska Fisheries Science
+Center/Northwest Fisheries Science Center Institutional Animal Care and
+Use Committee.
 

### 2021-11-04 analysis/paper/paper.Rmd
SPECIES ORDER ADJUSTED;MINOR EDITS

@@
 
-Ice-associated seals rely on sea ice for a variety of activities,
-including breeding, molting, pupping, and resting. In the Arctic, many
-of these activities occur in spring (April-June) as sea ice begins to
-melt and retreat northward. Rapid acceleration of climate change in
-Arctic ecosystems is therefore of concern as the quantity and quality of
-suitable habitat is forecast to decrease. In this paper, we use data
-collected from satellite-linked bio-loggers deployed between 2005 and
-2020 to investigate the seasonal timing and environmental factors
-affecting sea ice use by seals (specifically, bearded, ribbon, and
-spotted seals) in the Bering and Chukchi seas. In addition to providing
-baseline data on phenology, these data also allow us to quantify
-'availability', which is needed to accurately estimate abundance from
-aerial survey counts of seals basking on ice (i.e., to correct for the
-proportion of animals that are in the water while surveys are
-conducted). Using generalized linear mixed pseudo-models to properly
-account for temporal autocorrelation, we fit models with a variety of
-covariates (e.g., day-of-year, solar hour, age-sex class, wind speed,
-barometric pressure, temperature, precipitation) to examine their
-ability to explain variation in hourly haul-out records. We found
-evidence for strong diurnal and within-season patterns in haul-out
-behavior, as well as strong weather effects (particularly wind and
-temperature). In general, seals were more likely to rest on ice in the
-middle of the day and when wind speed was low and temperatures were
-higher. Haul-out propensity increased through March and April, peaking
-in May and early June before declining again. The timing and frequency
-of haul-out events also varied based on species and age-sex class. For
-ribbon and spotted seals, models with year effects were highly
-supported, indicating that the timing and magnitude of haul-out behavior
-varied among years. However, we did not find evidence that haul-out
-timing was linked to annual sea ice extent. Our analysis emphasizes the
-importance of accounting for variation in haul-out behavior when
-interpreting the number of seals counted in aerial surveys.
+Ice-associated seals rely on sea ice for a variety of activities, including
+breeding, molting, pupping, and resting. In the Arctic, many of these activities
+occur in spring (April-June) as sea ice begins to melt and retreat northward.
+Rapid acceleration of climate change in Arctic ecosystems is therefore of
+concern as the quantity and quality of suitable habitat is forecast to decrease.
+In this paper, we use data collected from satellite-linked bio-loggers deployed
+between 2005 and 2020 to investigate the seasonal timing and environmental
+factors affecting sea ice use by seals (specifically, bearded, ribbon, and
+spotted seals) in the Bering and Chukchi seas. In addition to providing baseline
+data on phenology, these data also allow us to quantify 'availability', which is
+needed to accurately estimate abundance from aerial survey counts of seals
+basking on ice (i.e., to correct for the proportion of animals that are in the
+water while surveys are conducted). Using generalized linear mixed pseudo-models
+to properly account for temporal autocorrelation, we fit models with a variety
+of covariates (e.g., day-of-year, solar hour, age-sex class, wind speed,
+barometric pressure, temperature, precipitation) to examine their ability to
+explain variation in hourly haul-out records. We found evidence for strong
+diurnal and within-season patterns in haul-out behavior, as well as strong
+weather effects (particularly wind and temperature). In general, seals were more
+likely to rest on ice in the middle of the day and when wind speed was low and
+temperatures were higher. Haul-out propensity increased through March and April,
+peaking in May and early June before declining again. The timing and frequency
+of haul-out events also varied based on species and age-sex class. For ribbon
+and spotted seals, models with year effects were highly supported, indicating
+that the timing and magnitude of haul-out behavior varied among years. However,
+we did not find evidence that haul-out timing was linked to annual sea ice
+extent. Our analysis emphasizes the importance of accounting for variation in
+haul-out behavior when interpreting the number of seals counted in aerial
+surveys.
 
@@
 
-The global climate disruption is causing considerable reduction in
-Arctic sea ice extent, volume, and seasonal presence [@kwok2018,
-@meier2014, @wang2017]. These changes have ripple effects on
-Arctic organisms, ecosystems, and the human communities who live in the
-region [@]. Such disruptions are a particular cause
-of concern for ice-associated seals (ringed (*Pusa hispida*), bearded
-(*Erignathus barbatus*), spotted (*Phoca largha*), and ribbon seals
-(*Histriophoca fasciata*)) that depend on spring and early summer sea
-ice (March-June) in the Bering and Chukchi seas as a platform for
-important functions, such as pupping, nursing, breeding, and molting
-[@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data
-and large knowledge gaps complicate predictions about the ultimate
-effects of changes in sea ice on the behavior, abundance, and
-distribution of these seals. Knowledge about evolutionary constraints on
-the timing of reproductive and molting behavior is generally lacking, so
-it is difficult to predict the readiness with which ice-associated seal
-species will be able to adapt to future changes (e.g., by adjusting
-pupping or molting schedules to earlier dates or different locales).
-Additionally, trends in abundance of these species are largely unknown,
-so it is difficult to say what effect, if any, declines in sea ice
-habitat have had, or will have, on seal densities.
-
-For ice-associated seals, haul out is tightly linked with sea ice.
-Ribbon seals rest out of the water almost exclusively on sea ice and are
-mostly pelagic outside the spring pupping, breeding, and molting season
-[@burns1981g]. While spotted and bearded seals rest on beaches during
-ice-free months, they strongly prefer sea ice as a resting platform
-during the spring and early summer [@burns2009]. The remoteness of the
-Bering and Chukchi seas mean direct scientific observation of seal
-behavior is impractical. Bio-logging devices are, thus, especially
-useful tools for collecting key information on movement and haul-out
-behavior for these species.
-
-Bio-logging records of time spent out of the water provide valuable data
-for identifying covariates that explain variation in haul-out behavior.
-For instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004]
-relied on bio-logging data to demonstrate greater haul-out propensity in
-juvenile crabeater seals (*Lobodon carcinophaga*) than adults, with
-highest probabilities in February and at times close to solar noon. In
-the Arctic, Von Duyke et al. [-@vonduyke2020] used seal-borne,
-satellite-linked loggers to corroborate seasonal changes between diurnal
-and nocturnal haul-out behavior of ringed seals previously described by
-Kelly et al. [-@kelly2010] using labor-intensive VHF radio tags and
-direct observation. Bengtson et al. [-@bengtson2005a] documented a
-higher propensity for ringed seal basking near solar noon, as did Ver
-Hoef et al. [-@verhoef2014a] in an analysis of bearded, ribbon, and
-spotted seals using much larger sample sizes. Olnes et al.
-[-@olnes2020] showed that the proportion of time bearded seals spent
-hauled out progressively increased through spring and summer. And, Ver
-Hoef et al. [-@verhoef2014a] found haul-out probabilities increased
-gradually starting in March and peaking in May and June for bearded,
+The global climate disruption is causing considerable reduction in Arctic sea
+ice extent, volume, and seasonal presence [@kwok2018;
+@meier2014; @overland2021; @wang2017]. These changes have ripple effects on
+Arctic organisms, ecosystems, and the human communities who live in the region
+[@huntington2020]. Such disruptions are a particular cause of concern for
+ice-associated seals (ringed (*Pusa hispida*), bearded (*Erignathus barbatus*),
+ribbon seals (*Histriophoca fasciata*), and spotted (*Phoca largha*)) that
+depend on spring and early summer sea ice (March-June) in the Bering and Chukchi
+seas as a platform for important functions, such as pupping, nursing, breeding,
+and molting [@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data
+and large knowledge gaps complicate predictions about the ultimate effects of
+changes in sea ice on the behavior, abundance, and distribution of these seals.
+Knowledge about evolutionary constraints on the timing of reproductive and
+molting behavior is generally lacking, so it is difficult to predict the
+readiness with which ice-associated seal species will be able to adapt to future
+changes (e.g., by adjusting pupping or molting schedules to earlier dates or
+different locales). Additionally, trends in abundance of these species are
+largely unknown, so it is difficult to say what effect, if any, declines in sea
+ice habitat have had, or will have, on seal densities.
+
+For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals
+rest out of the water almost exclusively on sea ice and are mostly pelagic
+outside the spring pupping, breeding, and molting season [@burns1981g]. While
+spotted and bearded seals rest on beaches during ice-free months, they strongly
+prefer sea ice as a resting platform during the spring and early summer
+[@burns2009]. The remoteness of the Bering and Chukchi seas mean direct
+scientific observation of seal behavior is impractical. Bio-logging devices are,
+thus, especially useful tools for collecting key information on movement and
+haul-out behavior for these species.
+
+Bio-logging records of time spent out of the water provide valuable data for
+identifying covariates that explain variation in haul-out behavior. For
+instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
+bio-logging data to demonstrate greater haul-out propensity in juvenile
+crabeater seals (*Lobodon carcinophaga*) than adults, with highest probabilities
+in February and at times close to solar noon. In the Arctic, Von Duyke et al.
+[-@vonduyke2020] used seal-borne, satellite-linked loggers to corroborate
+seasonal changes between diurnal and nocturnal haul-out behavior of ringed seals
+previously described by Kelly et al. [-@kelly2010] using labor-intensive VHF
+radio tags and direct observation. Bengtson et al. [-@bengtson2005a] documented
+a higher propensity for ringed seal basking near solar noon, as did Ver Hoef et
+al. [-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using
+much larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion
+of time bearded seals spent hauled out progressively increased through spring
+and summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
+increased gradually starting in March and peaking in May and June for bearded,
 ribbon, and spotted seals.
 
-While bio-logging of haul-out patterns clearly has utility for
-description of ice-associated seals' natural history and ecology, some
-studies have focused, as we do here, specifically on quantifying the
-probability of seals being hauled out as a means of accounting for seals
-in the water, and therefore missed, during aerial surveys. These studies
-typically use logistic regression-style analyses to estimate the
-time-specific probability of being hauled out base on 'wet/dry' data
-relayed by bio-loggers. In these models, haul-out probabilities were
-expressed as a function of predictive covariates, such as time of day,
-day-of-year, sex, age class, and environmental conditions (e.g.,
-[@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
-[@verhoef2014a], and [@southwell2008]). However, sample sizes have
-often been low and insufficient to permit strong inference about life
-history and/or seasonal variation in haul-out probabilities. For
-instance, Bengtson and Cameron's [-@bengtson2004] study included 5
-adult and 2 juvenile crabeater seals, while Bengtson et al.'s
-[-@bengtson2005a] study was based on 6 telemetered ringed seals in the
-Chukchi Sea. These studies were often further limited by logistical
-constraints on tagging fieldwork and the attachment duration or
-operational life of satellite tags. In this study, we address some of
-these limitations by deploying small bio-loggers designed for attachment
-on a portion of the study individuals. These devices are designed to
-collect data through the molt period (when those adhered to the hair
-would fall off) and, in some situations, provide multiple years of data.
-
-Ultimately, knowledge of trends in phenology and abundance (or life
-history surrogates such as survival and recruitment) will be necessary
-to make credible quantitative predictions about the effects of the
-climate crisis on the abundance and distribution of Arctic seal
-populations. Before we can construct a trend, however, we first require
-a baseline. Several studies have contributed estimates of the
-distribution and abundance of ice-associated seal species in the Arctic
-using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
-[@verhoef2014a]). Such studies were conducted over very large areas
-and estimation of absolute abundance required making inference about
-numerous nuisance processes affecting the observation of seals on ice.
-This included availability (only seals basking on ice were available to
-be counted), detection probability (observers or automated detection
-systems may have missed some seals on ice), species misclassification,
-and possible disturbance of seals by aircraft [@conn2014a;
+While bio-logging of haul-out patterns clearly has utility for description of
+ice-associated seals' natural history and ecology, some studies have focused, as
+we do here, specifically on quantifying the probability of seals being hauled
+out as a means of accounting for seals in the water, and therefore missed,
+during aerial surveys. These studies typically use logistic regression-style
+analyses to estimate the time-specific probability of being hauled out base on
+'wet/dry' data relayed by bio-loggers. In these models, haul-out probabilities
+were expressed as a function of predictive covariates, such as time of day,
+day-of-year, sex, age class, and environmental conditions (e.g., [@reder2003],
+[@bengtson2004], [@bengtson2005a], [@udevitz2009], [@verhoef2014a], and
+[@southwell2008]). However, sample sizes have often been low and insufficient to
+permit strong inference about life history and/or seasonal variation in haul-out
+probabilities. For instance, Bengtson and Cameron's [-@bengtson2004] study
+included 5 adult and 2 juvenile crabeater seals, while Bengtson et al.'s
+[-@bengtson2005a] study was based on 6 telemetered ringed seals in the Chukchi
+Sea. These studies were often further limited by logistical constraints on
+tagging fieldwork and the attachment duration or operational life of satellite
+tags. In this study, we address some of these limitations by deploying small
+bio-loggers designed for attachment on a portion of the study individuals. These
+devices are designed to collect data through the molt period (when those adhered
+to the hair would fall off) and, in some situations, provide multiple years of
+data.
+
+Ultimately, knowledge of trends in phenology and abundance (or life history
+surrogates such as survival and recruitment) will be necessary to make credible
+quantitative predictions about the effects of the climate crisis on the
+abundance and distribution of Arctic seal populations. Before we can construct a
+trend, however, we first require a baseline. Several studies have contributed
+estimates of the distribution and abundance of ice-associated seal species in
+the Arctic using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
+[@verhoef2014a]). Such studies were conducted over very large areas and
+estimation of absolute abundance required making inference about numerous
+nuisance processes affecting the observation of seals on ice. This included
+availability (only seals basking on ice were available to be counted), detection
+probability (observers or automated detection systems may have missed some seals
+on ice), species misclassification, and possible disturbance of seals by
+aircraft [@conn2014a;
 @verhoef2014a].
 
-In this study, we used data from 15 years of bio-logger records in a
-comprehensive investigation of haul-out behavior by 3 species of
-ice-associated seals in the Bering and Chukchi seas. We focused our
-effort on bearded, spotted, and ribbon seals and did not include ringed
-seals in the analysis. Ringed seals are unique in their use of snow
-lairs in the spring months, and they warrant a separate study that
-includes additional environmental covariates, such as the onset of snow
-melt and snow depth. Our goals are threefold. First, we wish to
-establish baseline estimates for the chronology of haul-out behavior in
-the critical spring season for each species across different age and sex
-classes. Second, we wish to refine estimates of haul-out availability
-corrections for aerial surveys in order to improve estimates of seal
-abundance. Previously estimated availability correction factors (e.g.,
-[@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
-for variables such as the time-of-day and day-of-year, but did not
-investigate meteorological variables that have been shown to influence
-haul-out behavior of walruses [@reder2003; @udevitz2009]. Third,
-we explored the annual variability in haul-out timing and possible
-linkage to dramatic changes in the availability and extent of seasonal
-sea ice between 2005 and 2020. Our work extends the scope of previous
-haul-out analyses, includes the influence of meteorological variability,
-and investigates the potential impact of a changing icescape on the
-behavior of these species.
+In this study, we used data from 16 years of bio-logger records in a
+comprehensive investigation of haul-out behavior by 3 species of ice-associated
+seals in the Bering and Chukchi seas. We focused our effort on bearded, ribbon,
+and spotted seals and did not include ringed seals in the analysis. Ringed seals
+are unique in their use of snow lairs in the spring months, and they warrant a
+separate study that includes additional environmental covariates, such as the
+onset of snow melt and snow depth. Our goals are threefold. First, we wish to
+establish baseline estimates for the chronology of haul-out behavior in the
+critical spring season for each species across different age and sex classes.
+Second, we wish to refine estimates of haul-out availability corrections for
+aerial surveys in order to improve estimates of seal abundance. Previously
+estimated availability correction factors (e.g., [@bengtson2005a], [@conn2014a],
+and [@verhoef2014a]) accounted for variables such as the time-of-day and
+day-of-year, but did not investigate meteorological variables that have been
+shown to influence haul-out behavior of walruses [@reder2003; @udevitz2009].
+Third, we explored the annual variability in haul-out timing and possible
+linkage to dramatic changes in the availability and extent of seasonal sea ice
+between 2005 and 2020. Our work extends the scope of previous haul-out analyses,
+includes the influence of meteorological variability, and investigates the
+potential impact of a changing icescape on the behavior of these species.
 
@@
 Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
-bio-loggers deployed on bearded, spotted, and ribbon seals was filtered
-to include only records from 1 March to 15 July between
-`r min(lubridate::year(analysis_data$haulout_dt))` and
-`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of
-the 'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers
-(Redmond, Washington, USA) and either adhered to the hair on the seal or
-attached to the rear flipper inter-digital webbing. The use of
-bio-loggers adhered to the back or head provide some benefits over
-flipper mounted devices (e.g. increased satellite transmittal rates,
-locations at sea) but these are lost during the annual molt, which
-limits the effective length of haul-out timelines. Additionally, sensors
-attached to the head or dorsal region are often dry while the seal is at
-the surface and can slightly bias the percent-dry values recorded by the
-device. For this study, in cases where both tag types were deployed,
-percent dry observations from the flipper tag were preferred. Sex and
-age class (non-dependent *young-of-the-year*, sexually immature
-*subadults*, and mature *adults*) were determined at the time of
-deployment by subjective judgement based on length, and pelage
-characteristics for some species. Field identification of age class can
-be inexact, particularly when discerning subadults from adults. In the
-case of ribbon seals, pelage differences between subadults and adults
-are distinguishable, with subadults often having less defined ribbons.
-Bearded seal subadults will often have a spotted pattern in the pelage
-that is not seen in adults. Spotted seal pelage is monomorphic and can
-not be used to discern age class. Despite these challenges, we feel the
-age classifications used in this analysis are useful categories in
-testing for age-related effects on haul-out behavior. Seals determined
-to be less than one year were classified as young-of-the-year. Subadults
-are those seals likely greater than one year of age but not yet sexually
-mature. Adults are likely sexually mature and older than approximately
-four years. For those tags deployed on young-of-the-year and
-transmitting into the next year, the age class was advanced to subadult
-on 1 March of the following year. Table \@ref(tab:tblDeploy) provides a
-summary of these deployments and data received from them.
+bio-loggers deployed on bearded, ribbon, and spotted seals was filtered to
+include only records from 1 March to 15 July between `r min(lubridate::year(analysis_data$haulout_dt))` 
+and `r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
+'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
+Washington, USA) and either adhered to the hair on the seal or attached to the
+rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
+or head provide some benefits over flipper mounted devices (e.g. increased
+satellite transmittal rates, locations at sea) but these are lost during the
+annual molt, which limits the effective length of haul-out timelines.
+Additionally, sensors attached to the head or dorsal region are often dry while
+the seal is at the surface and can slightly bias the percent-dry values recorded
+by the device. For this study, in cases where both tag types were deployed,
+percent dry observations from the flipper tag were preferred. Sex and age class
+(non-dependent *young-of-the-year*, sexually immature *subadults*, and mature
+*adults*) were determined at the time of deployment by subjective judgement
+based on length, and pelage characteristics for some species. Field
+identification of age class can be inexact, particularly when discerning
+subadults from adults. In the case of ribbon seals, pelage differences between
+subadults and adults are distinguishable, with subadults often having less
+defined ribbons. Bearded seal subadults will often have a spotted pattern in the
+pelage that is not seen in adults. Spotted seal pelage is monomorphic and can
+not be used to discern age class. Despite these challenges, we feel the age
+classifications used in this analysis are useful categories in testing for
+age-related effects on haul-out behavior. Seals determined to be less than one
+year were classified as young-of-the-year. Subadults are those seals likely
+greater than one year of age but not yet sexually mature. Adults are likely
+sexually mature and older than approximately four years. For those tags deployed
+on young-of-the-year and transmitting into the next year, the age class was
+advanced to subadult on 1 March of the following year. Table
+\@ref(tab:tblDeploy) provides a summary of these deployments and data received
+from them.
 
@@
 
-Tags that fall off due to molt or mortality and remain on ice or land
-can still send data to satellites, and we did not want to include data
-from these tags in our analyses. As such, end times of each deployment
-were identified by experts who examined several simultaneous data
-streams (e.g., tag locations and dive behavior) to determine when tags
-ceased providing data consistent with seal behavior. Sensor readings
-outside of these start and end times were discarded prior to analysis.
-After approximately 9 months, 7 devices deployed on the rear flipper of
-bearded seals reported implausible sensor data. All data after the first
-instance of unrealistic values were censored from this analysis. Figure
-\@ref(fig:dataCal) shows a calendar distribution of these data across
-the study season for each species. Observations are concentrated in the
-months of May and June due to the timing of deployment (April and May)
-and the timing of molt (May and June). During molt, seals (and their
-attached bio-loggers) spend more time out of the water and more data is
-transmitted. Molt timing also impacts when many deployments end as any
-tags adhered to the hair will fall off.
+Tags that fall off due to molt or mortality and remain on ice or land can still
+send data to satellites, and we did not want to include data from these tags in
+our analyses. As such, end times of each deployment were identified by experts
+who examined several simultaneous data streams (e.g., tag locations and dive
+behavior) to determine when tags ceased providing data consistent with seal
+behavior. Sensor readings outside of these start and end times were discarded
+prior to analysis. After approximately 9 months, 7 devices deployed on the rear
+flipper of bearded seals reported implausible sensor data. All data after the
+first instance of unrealistic values were censored from this analysis. Figure
+\@ref(fig:dataCal) shows a calendar-like distribution of these data across the
+study season for each species. Observations for ribbon and spotted seals are
+concentrated in the months of May and June due to the timing of deployment
+(April and May) and the timing of molt (May and June). During molt, seals (and
+their attached bio-loggers) spend more time out of the water and more data is
+transmitted. Molt timing also impacts when many deployments end as any tags
+adhered to the hair will fall off. Bearded seal deployments are more evenly
+spread with fewer observations in May and June. The majority of bearded seal
+deployments start later in the summer and by May tags have either fallen off or
+the batteries are depleted.
 
 ```{r seal-hours-calendar, include=TRUE, fig.height = 4.75, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species.
-The higher density of data from May and June coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out.
+The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out.
 Additionally, a majority of deployments started in April and May.
@@
        size = "seal hours of bio-logger data") +
-  ggtitle("Haul-out behavior data is concentrated in the months of May and June",
+  ggtitle("The number of haul-out behavior observations varies from March through July",
           subtitle = "data presented in seal-hours collated across years (2005-2020)")
@@
 
-Haul-out behavior data was recorded and transmitted via the Argos
-satellite network as hourly percent-dry timelines. For each hour of a
-day, the wet/dry sensor was polled by the tag firmware every few seconds
-and a percent of the hour in a dry state was calculated (Figure
-\@ref(fig:examplePlot)). Each data transmission message consisted of a
-complete 24-hour record. Tags were, generally, deployed and programmed
-in a manner to maximize data transmission during the spring pupping and
-molting period, however, not all days during the deployment were
-successfully transmitted. This is due to a variety of factors including
-satellite coverage, tag availability (i.e. tags mounted to the rear
+Haul-out behavior data was recorded and transmitted via the Argos satellite
+network as hourly percent-dry timelines. For each hour of a day, the wet/dry
+sensor was polled by the tag firmware every few seconds and a percent of the
+hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). Each data
+transmission message consisted of a complete 24-hour record. Tags were,
+generally, deployed and programmed in a manner to maximize data transmission
+during the spring pupping and molting period, however, not all days during the
+deployment were successfully transmitted. This is due to a variety of factors
+including satellite coverage, tag availability (i.e. tags mounted to the rear
 flipper are not available to the surface while the seal is at sea), tag
@@
 
-Of key interest in this study is the relationship between haul-out
-behavior and a variety of environmental covariates that vary with space
-and time. We explored the use of a movement model to predict locations
-at specific times. However, the sparse nature of data from some tags,
-especially those mounted to the rear flipper, resulted in poor modeling
-performance or convergence issues. For this study, we calculated a
-weighted average daily location where the inverse of the estimated Argos
-location error was used for the weight. Any days where haul-out
-observations were present without any location data were filled from the
-last calculated weighted average daily location, and any days where the
-location intersected with land were removed from the dataset. Figure
-\@ref(fig:dataMap) shows the spatial distribution of determined
-locations and available haul-out behavior data for each species across
-the study area.
+Of key interest in this study is the relationship between haul-out behavior and
+a variety of environmental covariates that vary with space and time. We explored
+the use of a movement model to predict locations at specific times. However, the
+sparse nature of data from some tags, especially those mounted to the rear
+flipper, resulted in poor modeling performance or convergence issues. For this
+study, we calculated a weighted average daily location where the inverse of the
+estimated Argos location error was used for the weight. Any days where haul-out
+observations were present without any location data were filled from the last
+calculated weighted average daily location, and any days where the location
+intersected with land were removed from the dataset. Figure \@ref(fig:dataMap)
+shows the spatial distribution of determined locations and available haul-out
+behavior data for each species across the study area.
 
@@
 
-In addition to sex and age class, we gathered additional variables that
-might help explain variation in haul-out probabilities. These included
-day-of-year (for seasonal effects) and local solar hour (for diurnal
-effects). Solar hour was calculated using the {solaR} package
-[@perpinan2012] within the R statistical environment
-[@rcoreteam2021] using the daily locations. We also compiled several
-meteorological variables shown to affect haul-out behavior in other
-Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017]. In
-particular, we linked locations from satellite tags to key weather
-values from the North American Regional Reanalysis (NARR) model produced
-by the National Centers for Environmental Prediction
-[@mesinger2006]. The NARR model assimilates observational data to
-produce a long-term picture of weather over North America. Numerous
-weather variables are made available across the region 8 times daily.
-For this study, NARR weather values were subset to the extent of our
-study area over the Bering and Chukchi Seas at 3-hr intervals based on
-the native grid resolution of 32 km (1024 sq. km). The following
-meteorological variables were interpolated and assigned to seal
-locations using a bilinear method: 1) air temperature at 2m above the
-earth's surface, 2) wind consisting of northerly and easterly vector
-components converted to wind speed using the Euclidean norm, 3)
+In addition to sex and age class, we gathered additional variables that might
+help explain variation in haul-out probabilities. These included day-of-year
+(for seasonal effects) and local solar hour (for diurnal effects). Solar hour
+was calculated using the {solaR} package [@perpinan2012] within the R
+statistical environment [@rcoreteam2021] using the daily locations. We also
+compiled several meteorological variables shown to affect haul-out behavior in
+other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017]. In particular, we
+linked locations from satellite tags to key weather values from the North
+American Regional Reanalysis (NARR) model produced by the National Centers for
+Environmental Prediction [@mesinger2006]. The NARR model assimilates
+observational data to produce a long-term picture of weather over North America.
+Numerous weather variables are made available across the region 8 times daily.
+For this study, NARR weather values were subset to the extent of our study area
+over the Bering and Chukchi Seas at 3-hr intervals based on the native grid
+resolution of 32 km (1024 sq. km). The following meteorological variables were
+interpolated and assigned to seal locations using a bilinear method: 1) air
+temperature at 2m above the earth's surface, 2) wind consisting of northerly and
+easterly vector components converted to wind speed using the Euclidean norm, 3)
 barometric pressure at sea level, and 4) precipitation (Table
@@
 
-Models for all species include the following effects: day-of-year, solar
-hour, temperature, wind speed, barometric pressure, precipitation, wind
-chill (represented by a *wind:temperature* interaction
-[@udevitz2009]), and day-of-year and time of day interactions to
-permit diurnal patterns to change throughout the year. Spotted and
-ribbon seal models included age-sex class and interactions between
-day-of-year and age-sex class, but we omitted these from bearded seal
-models due to poor representation of age-sex classes (Table
-\@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
-(and an interaction with day-of-year), since bearded seals occupy a
-substantial range and we were interested in possible differences in the
-timing of haul-out behavior along a latitudinal gradient. We omitted the
-latitudinal effect from ribbon and spotted seal models since, during the
-spring, these species are most prevalent near the southern ice edge in
-the Bering Sea [@conn2014a].
-
-Notably missing from this list of explanatory variables is any
-spatial-temporal representation of sea ice concentration, area, or
-extent. This may seem counterintuitive when modeling the haul-out
-behavior of seal species with such a close association to sea ice; seals
-haul out in the presence of sea ice, and we could assess the local
-concentration of sea ice during these events (see [@olnes2020]).
-However, a major focus of this study is to develop models applicable for
-aerial survey correction factors and using sea ice as a covariate would
-almost certainly bias haul-out predictions towards those seals that are
-on or near ice and therefore more likely to haul out. Since aerial
-surveys can only detect seals on ice, abundance estimates would be
-missing a correction for those seals that are away from ice (e.g., on
-foraging trips). Additionally, many of the bio-logger deployments used
-in this study were single attachments to the rear flipper of the seal
-and do not provide at-sea locations, limiting our ability to fully
-evaluate habitat preferences related to sea-ice. It should be noted that
-the aerial survey counts are modeled with sea-ice and other habitat
-parameters. So, any final abundance estimates that combine models of
-survey counts and our model for proportion of seals hauled out on ice do
-account for the strong association between these species and sea-ice in
-the Bering and Chukchi seas. Lastly, our study was limited to the spring
-season when seal haul-out tendencies are strongly influenced by pupping,
-nursing, breeding, and molt and these drivers are likely more
-influential than specific sea-ice concentration. Crawford et al
-[-@crawford2019] compared haul-out probability models for bearded seals
-and found those that only included season (and not sea-ice
-concentration) were the most parsimonious. For these reasons, we have
+Models for all species include the following effects: day-of-year, solar hour,
+temperature, wind speed, barometric pressure, precipitation, wind chill
+(represented by a *wind:temperature* interaction [@udevitz2009]), and
+day-of-year and time of day interactions to permit diurnal patterns to change
+throughout the year. Ribbon and spotted seal models included age-sex class and
+interactions between day-of-year and age-sex class, but we omitted these from
+bearded seal models due to poor representation of age-sex classes (Table
+\@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect (and an
+interaction with day-of-year), since bearded seals occupy a substantial range
+and we were interested in possible differences in the timing of haul-out
+behavior along a latitudinal gradient. We omitted the latitudinal effect from
+ribbon and spotted seal models since, during the spring, these species are most
+prevalent near the southern ice edge in the Bering Sea [@conn2014a].
+
+Notably missing from this list of explanatory variables is any spatial-temporal
+representation of sea ice concentration, area, or extent. This may seem
+counterintuitive when modeling the haul-out behavior of seal species with such a
+close association to sea ice; seals haul out in the presence of sea ice, and we
+could assess the local concentration of sea ice during these events (see
+[@olnes2020]). However, a major focus of this study is to develop models
+applicable for aerial survey correction factors and using sea ice as a covariate
+would almost certainly bias haul-out predictions towards those seals that are on
+or near ice and therefore more likely to haul out. Since aerial surveys can only
+detect seals on ice, abundance estimates would be missing a correction for those
+seals that are away from ice (e.g., on foraging trips). Additionally, many of
+the bio-logger deployments used in this study were single attachments to the
+rear flipper of the seal and do not provide at-sea locations, limiting our
+ability to fully evaluate habitat preferences related to sea-ice. It should be
+noted that the aerial survey counts are modeled with sea-ice and other habitat
+parameters. So, any final abundance estimates that combine models of survey
+counts and our model for proportion of seals hauled out on ice do account for
+the strong association between these species and sea-ice in the Bering and
+Chukchi seas. Lastly, our study was limited to the spring season when seal
+haul-out tendencies are strongly influenced by pupping, nursing, breeding, and
+molt and these drivers are likely more influential than specific sea-ice
+concentration. Crawford et al [-@crawford2019] compared haul-out probability
+models for bearded seals and found those that only included season (and not
+sea-ice concentration) were the most parsimonious. For these reasons, we have
 elected not to use sea ice concentration as a predictor for haul-out
-probability; studies to evaluate habitat selection in relation to
-sea-ice parameters and other environmental covariates are in progress.
-
-We explored whether the annual variation in maximum spring sea ice
-extent in the Bering Sea influences the seasonal peak of seal haul-out
-behavior. In particular, we used sea ice concentration data from the
-Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave Dataset, Version 1
-[@cavalieri1996] to calculate maximum sea ice extent. All sea ice
-concentration grid cells (25 km^2^) in the study area with greater than
-15% concentration were summed to get the total sea ice extent for each
-day between 15 February and 15 July across all years. Maximum spring sea
-ice extent was simply the largest daily value for each year.
+probability; studies to evaluate habitat selection in relation to sea-ice
+parameters and other environmental covariates are in progress.
+
+We explored whether the annual variation in maximum spring sea ice extent in the
+Bering Sea influences the seasonal peak of seal haul-out behavior. In
+particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
+SSM/I-SSMIS Passive Microwave Dataset, Version 1 [@cavalieri1996] to calculate
+maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
+study area with greater than 15% concentration were summed to get the total sea
+ice extent for each day between 15 February and 15 July across all years.
+Maximum spring sea ice extent was simply the largest daily value for each year.
 
@@
 
-Haul-out records for seals are often characterized by multiple hours
-spent basking on ice alternating with long periods of foraging (Figure
-\@ref(fig:examplePlot)). Commonly used statistical models for binary
-data (e.g. logistic regression) assume independence among responses, an
-assumption that is clearly violated if hourly responses are modeled. Any
-analysis that ignores temporal dependence in responses will thus have
-overstated precision [@betts2006].
-
-To properly account for temporal dependence within a computationally
-tractable framework, we used generalized linear mixed pseudo-models
-(GLMPMs; [@verhoef2010a]) to model variation in haul-out behavior as a
-function of (1) covariate predictors, (2) temporally autocorrelated
-random effects, and (3) individual random effects representing
-heterogeneity in individual behavior. We used the {glmmLDTS} package
-[@verhoef2010a] to implement GLMPMs. We explored two different model formulations for our
-data, and owing to the large number of records, we fit separate models
-to bearded, ribbon, and spotted seal data sets. In our first model
-formulation, for each species, we fitted a year-independent model that
-predicted average haul-out behavior as a function of demographic,
-environmental, seasonal, and diurnal effects. Second, for ribbon and
-spotted seals (which had considerably more data than bearded seals), we
-fitted models that included all the effects from the first model, but
-also permitted annual variation in haul-out timing. This second class of
-models was used to examine whether haul-out patterns varied by year and
-to determine the annual timing of apparent peaks in haul-out behavior.
-
-We followed Ver Hoef et al. [-@verhoef2014a] in using linear,
-quadratic, and cubic effects of day-of-year to represent seasonal
-changes in behavior. However, unlike previous models for harbor seals
-[@london2012] and ice-associated seals [@verhoef2014a], which treated
-hour-of-day as a 24-level categorical variable to capture diurnal
-cycles, we adopted a continuous formulation based on Fourier series that
-provides a flexible model while preserving the inherent circularity
-needed for time-of-day effects (i.e., hour 0 should be equal to hour
-24). It also represents hour-of-day with 6 parameters, which is a
-considerable reduction when compared to a 24 parameter variable,
-especially when fitting models with interactions between hour-of-day and
-other variables (e.g., age-sex class, day-of-year). According to this
-approach, we used the following specification for hour-of-day effects:
+Haul-out records for seals are often characterized by multiple hours spent
+basking on ice alternating with long periods of foraging (Figure
+\@ref(fig:examplePlot)). Commonly used statistical models for binary data (e.g.
+logistic regression) assume independence among responses, an assumption that is
+clearly violated if hourly responses are modeled. Any analysis that ignores
+temporal dependence in responses will thus have overstated precision
+[@betts2006].
+
+To properly account for temporal dependence within a computationally tractable
+framework, we used generalized linear mixed pseudo-models (GLMPMs;
+[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
+covariate predictors, (2) temporally autocorrelated random effects, and (3)
+individual random effects representing heterogeneity in individual behavior. We
+used the {glmmLDTS} package [@verhoef2010a] to implement GLMPMs. We explored two
+different model formulations for our data, and owing to the large number of
+records, we fit separate models to bearded, ribbon, and spotted seal data sets.
+In our first model formulation, for each species, we fitted a year-independent
+model that predicted average haul-out behavior as a function of demographic,
+environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
+seals (which had considerably more data than bearded seals), we fitted models
+that included all the effects from the first model, but also permitted annual
+variation in haul-out timing. This second class of models was used to examine
+whether haul-out patterns varied by year and to determine the annual timing of
+apparent peaks in haul-out behavior.
+
+We followed Ver Hoef et al. [-@verhoef2014a] in using linear, quadratic, and
+cubic effects of day-of-year to represent seasonal changes in behavior. However,
+unlike previous models for harbor seals [@london2012] and ice-associated seals
+[@verhoef2014a], which treated hour-of-day as a 24-level categorical variable to
+capture diurnal cycles, we adopted a continuous formulation based on Fourier
+series that provides a flexible model while preserving the inherent circularity
+needed for time-of-day effects (i.e., hour 0 should be equal to hour 24). It
+also represents hour-of-day with 6 parameters, which is a considerable reduction
+when compared to a 24 parameter variable, especially when fitting models with
+interactions between hour-of-day and other variables (e.g., age-sex class,
+day-of-year). According to this approach, we used the following specification
+for hour-of-day effects:
 
@@
 H_t=\alpha_1 \cos(\frac{\pi t}{4}) + \alpha_2 \sin(\frac{\pi t}{4}) + \alpha_3 \cos(\frac{\pi t}{6}) + \alpha_4 \sin(\frac{\pi t}{6}) + \alpha_5 \cos(\frac{\pi t}{12}) + \alpha_6 \sin(\frac{\pi t}{12})
-$$ where $H_t$ gives the effect for solar hour $t$ and $α_{i}$ are
+$$ 
+where $H_t$ gives the effect for solar hour $t$ and $α_{i}$ are
 estimated parameters (regression coefficients).
 
-For the second set of models examining inter-annual variation in sea ice
-use, we fitted models with year by day-of-year interactions. However, in
-this case we only included *year:day* and *year:day2*, omitting the main
-effects of year as well as *year:day3* interactions because models with
-the latter effects were numerically unstable. However, the modeled
-interactions were sufficient to allow shifts in haul-out distribution,
-as one can show mathematically that a simple horizontal shift in timing
-of haul-out distributions does not affect the main effects or cubic
-terms in a polynomial regression model.
-
-A typical model fitting exercise would also include a model selection
-process. However, AIC (and similar criteria) is not suitable when using
-pseudo-likelihoods, because pseudo-data generated in the model fitting
-process [@verhoef2010a] differ between models [@teneyck2018]. After
-fitting GLMPM models, we instead used "type III" $F$-tests to calculate
-$p$-values [@verhoef2010a] to evaluate model performance and important
-terms. We also produced predictions of haul-out behavior as a function
-of influential predictors (e.g. solar hour, day-of-year, age-sex).
-Weather covariates for these predictions were based on daily or hourly
-smoothed weather covariate values across the study region. Such
-predictions were then used to develop haul-out probability surfaces,
-explore conditional effects of weather covariates, and determine annual
-peaks in haul-out activity. The timing of peak haul-out behavior was
-further used to regress against the annual maximum sea ice extent in the
-study region.
+For the second set of models examining inter-annual variation in sea ice use, we
+fitted models with year by day-of-year interactions. However, in this case we
+only included *year:day* and *year:day2*, omitting the main effects of year as
+well as *year:day3* interactions because models with the latter effects were
+numerically unstable. However, the modeled interactions were sufficient to allow
+shifts in haul-out distribution, as one can show mathematically that a simple
+horizontal shift in timing of haul-out distributions does not affect the main
+effects or cubic terms in a polynomial regression model.
+
+A typical model fitting exercise would also include a model selection process.
+However, AIC (and similar criteria) is not suitable when using
+pseudo-likelihoods, because pseudo-data generated in the model fitting process
+[@verhoef2010a] differ between models [@teneyck2018]. After fitting GLMPM
+models, we instead used "type III" $F$-tests to calculate $p$-values
+[@verhoef2010a] to evaluate model performance and important terms. We also
+produced predictions of haul-out behavior as a function of influential
+predictors (e.g. solar hour, day-of-year, age-sex). Weather covariates for these
+predictions were based on daily or hourly smoothed weather covariate values
+across the study region. Such predictions were then used to develop haul-out
+probability surfaces, explore conditional effects of weather covariates, and
+determine annual peaks in haul-out activity. The timing of peak haul-out
+behavior was further used to regress against the annual maximum sea ice extent
+in the study region.
 
@@
 
-Models omitting year effects suggested that day-of-year, solar hour,
-age-sex class, temperature, and wind substantially influenced haul-out
-behavior of all three species, with $F$ tests producing $p$-values less
-than 0.05 for variables embodying these effects and/or their
-interactions. Haul-out probabilities typically increased throughout
-March and April, reaching a peak in May and early June before declining
-again. Diurnal patterns were present, with maximum haul-out behavior
-centered around local solar noon.
+Models omitting year effects suggested that day-of-year, solar hour, age-sex
+class, temperature, and wind substantially influenced haul-out behavior of all
+three species, with $F$ tests producing $p$-values less than 0.05 for variables
+embodying these effects and/or their interactions. Haul-out probabilities
+typically increased throughout March and April, reaching a peak in May and early
+June before declining again. Diurnal patterns were present, with maximum
+haul-out behavior centered around local solar noon.
+
+```{r type3}
+ribbon_type3 <- tibble(
+  ribbon_fit$typeIII.hypoth
+) %>% 
+  mutate(Prob.F = scales::pvalue(Prob.F),
+         Effect = janitor::make_clean_names(Effect)) %>% 
+  column_to_rownames(var = "Effect") 
+ribbon_type3 <- setNames(split(ribbon_type3, 
+                               seq(nrow(ribbon_type3))), 
+                         rownames(ribbon_type3))
+
+spotted_type3 <- tibble(
+  spotted_fit$typeIII.hypoth
+) %>% 
+  mutate(Prob.F = scales::pvalue(Prob.F),
+         Effect = janitor::make_clean_names(Effect)) %>%  
+  column_to_rownames(var = "Effect")  
+spotted_type3 <- setNames(split(spotted_type3, 
+                               seq(nrow(spotted_type3))), 
+                         rownames(spotted_type3))
+
+bearded_type3 <- tibble(
+  bearded_fit$typeIII.hypoth
+) %>% 
+  mutate(Prob.F = scales::pvalue(Prob.F),
+         Effect = janitor::make_clean_names(Effect)) %>%  
+  column_to_rownames(var = "Effect")  
+bearded_type3 <- setNames(split(bearded_type3, 
+                               seq(nrow(bearded_type3))), 
+                         rownames(bearded_type3))
+```
+
+## Bearded Seals
+
+Age and sex class were not included in the model for bearded seals and, as such,
+results are shown for all ages. Bearded seals were predicted to have a bi-modal
+distribution of haul-out probability across the day (Figure
+\@ref(fig:beardedHOCal)). In addition to a peak around local solar noon, the
+bearded seal model predicts additional haul-out activity around local midnight.
+The haul-out behavior is also more protracted throughout the spring season
+compared to ribbon and spotted seals.
+
+```{r ebnewdata}
+targets::tar_load(bearded_newdata, store = here::here("_targets"))
+```
+
+```{r beardedHOCal, fig.height = 3.5, include=TRUE, fig.id = "beardedHOCal", fig.cap="Predicted haul-out probability of bearded seals from 1 March to 15 July for all and sex classes."}
+plot_df <- bearded_newdata %>%
+  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+         month = lubridate::month(date,label=TRUE),
+         day = lubridate::day(date)) %>%
+  filter(month != "Aug")
+
+p <- ggplot(plot_df,aes(day,solar_hour,fill = ho_prob))+
+  geom_tile(color= "white",size = 0) +
+  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
+                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
+                                              barheight = 0.5, title.hjust = 0.5)
+  )
+p <- p + facet_grid(age_sex~month)
+p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
+  scale_y_continuous(breaks = c(4,12,20))
+p <- p + theme(legend.position = "bottom") +
+  theme(strip.background = element_rect(colour="white")) +
+  theme(axis.ticks=element_blank()) +
+  xlab("day of month") + ylab("local solar hour") +
+  ggtitle("Bearded seals rest on sea ice around solar noon and solar midnight")
+p
+```
+
+Like ribbon and spotted seals, bearded seal haul-out probability appears
+most influenced by wind
+($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$
+= `r bearded_type3$wind$F.value`; $p$ = `r bearded_type3$wind$Prob.F`)
+and temperature
+($F_{`r paste(bearded_type3$temp2$Num.df, bearded_type3$temp2$Den.df, sep=",")`}$
+= `r bearded_type3$temp2$F.value`; $p$ =
+`r bearded_type3$temp2$Prob.F`). Additionally, wind chill was also a
+significant influence
+($F_{`r paste(bearded_type3$temp2_wind$Num.df, bearded_type3$temp2_wind$Den.df, sep=",")`}$
+= `r bearded_type3$temp2_wind$F.value`; $p$ =
+`r bearded_type3$temp2_wind$Prob.F`). Any influence of barometric
+pressure or precipitation is less apparent.
+
+```{r beardedHOwx,fig.height = 5.5, include=FALSE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across the range of weather conditions encountered in the observed data.
+Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
+Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}
+
+p_temp <- ggplot(plot_df, aes(temp2*27, ho_prob, color = age_sex)) +
+  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
+  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
+  geom_smooth(se = FALSE) +
+  theme(legend.position = "none") +
+  xlab("temperature (C)") + ylab("haul-out probability")
+
+p_wind <- ggplot(plot_df, aes(wind, ho_prob, color = age_sex)) +
+  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
+  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
+  geom_smooth(se = FALSE) +
+  theme(legend.position = "none") +
+  xlab("wind") + ylab("")
+
+p_pressure <- ggplot(plot_df, aes(pressure, ho_prob, color = age_sex)) +
+  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
+  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
+  geom_smooth(se = FALSE) +
+  theme(legend.position = "none") +
+  xlab("barometric pressure") + ylab("haul-out probability")
+
+p_precip <- ggplot(plot_df, aes(precip, ho_prob, color = age_sex)) +
+  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
+  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
+  geom_smooth(se = FALSE) +
+  theme(legend.position = "none") +
+  xlab("precipitation") + ylab("")
+
+library(patchwork)
+
+((p_temp + p_wind) / (p_pressure + p_precip)) +
+  plot_annotation(
+    title = 'Bearded seal haul-out behavior is less influeced by weather')
+```
 
@@
 p
-```
-
-```{r type3}
-ribbon_type3 <- tibble(
-  ribbon_fit$typeIII.hypoth
-) %>% 
-  mutate(Prob.F = scales::pvalue(Prob.F),
-         Effect = janitor::make_clean_names(Effect)) %>% 
-  column_to_rownames(var = "Effect") 
-ribbon_type3 <- setNames(split(ribbon_type3, 
-                               seq(nrow(ribbon_type3))), 
-                         rownames(ribbon_type3))
-
-spotted_type3 <- tibble(
-  spotted_fit$typeIII.hypoth
-) %>% 
-  mutate(Prob.F = scales::pvalue(Prob.F),
-         Effect = janitor::make_clean_names(Effect)) %>%  
-  column_to_rownames(var = "Effect")  
-spotted_type3 <- setNames(split(spotted_type3, 
-                               seq(nrow(spotted_type3))), 
-                         rownames(spotted_type3))
-
-bearded_type3 <- tibble(
-  bearded_fit$typeIII.hypoth
-) %>% 
-  mutate(Prob.F = scales::pvalue(Prob.F),
-         Effect = janitor::make_clean_names(Effect)) %>%  
-  column_to_rownames(var = "Effect")  
-bearded_type3 <- setNames(split(bearded_type3, 
-                               seq(nrow(bearded_type3))), 
-                         rownames(bearded_type3))
 ```
@@
 more variable than simply showing model coefficients would imply. Figure
-\@ref(fig:ribbonHOwx)) presents the predicted haul-out probability of
+\@ref(fig:ribbonHOwx) presents the predicted haul-out probability of
 ribbon seals across the range of weather conditions encountered in the
 observed data. The collinearity and temporal variation are apparent in
-the visual patterns that emerge. By
+the visual patterns that emerge.
 
@@
 
-Compared to ribbon seals, spotted seals showed a longer spring haul-out
-season that is less intensely centered on solar noon (Figure
-\@ref(fig:spottedHOCal)). Adults of both sexes spend considerable time
-in April and May resting on the ice. As with ribbon seals, the
-young-of-the-year records begin after weaning and the model predictions
-demonstrate the ontogeny of in-water activities (e.g. diving, foraging)
-in May.
+Compared to ribbon seals, spotted seals showed a longer spring haul-out season
+that is less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)).
+Adults of both sexes spend considerable time in April and May resting on the
+ice. As with ribbon seals, the young-of-the-year records begin after weaning and
+the model predictions demonstrate the ontogeny of in-water activities (e.g.
+diving, foraging) in May.
 
@@
     title = 'Spotted seal haul-out behavior is influenced by weather')
-```
-
-## Bearded Seals
-
-Age and sex class were not included in the model for bearded seals and,
-as such, results are shown for all ages. Unlike ribbon and spotted
-seals, bearded seals were predicted to have a bi-modal distribution of
-haul-out probability across the day (Figure \@ref(fig:beardedHOCal)). In
-addition to a peak around local solar noon, the bearded seal model
-predicts additional haul-out activity around local midnight. The
-haul-out behavior is also more protracted throughout the spring season
-compared to ribbon and spotted seals.
-
-```{r ebnewdata}
-targets::tar_load(bearded_newdata, store = here::here("_targets"))
-```
-
-```{r beardedHOCal, fig.height = 3.5, include=TRUE, fig.id = "beardedHOCal", fig.cap="Predicted haul-out probability of bearded seals from 1 March to 15 July for all and sex classes."}
-plot_df <- bearded_newdata %>%
-  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
-         month = lubridate::month(date,label=TRUE),
-         day = lubridate::day(date)) %>%
-  filter(month != "Aug")
-
-p <- ggplot(plot_df,aes(day,solar_hour,fill = ho_prob))+
-  geom_tile(color= "white",size = 0) +
-  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
-                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
-                                              barheight = 0.5, title.hjust = 0.5)
-  )
-p <- p + facet_grid(age_sex~month)
-p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
-  scale_y_continuous(breaks = c(4,12,20))
-p <- p + theme(legend.position = "bottom") +
-  theme(strip.background = element_rect(colour="white")) +
-  theme(axis.ticks=element_blank()) +
-  xlab("day of month") + ylab("local solar hour") +
-  ggtitle("Bearded seals rest on sea ice around solar noon and solar midnight")
-p
-```
-
-Like ribbon and spotted seals, bearded seal haul-out probability appears
-most influenced by wind
-($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$
-= `r bearded_type3$wind$F.value`; $p$ = `r bearded_type3$wind$Prob.F`)
-and temperature
-($F_{`r paste(bearded_type3$temp2$Num.df, bearded_type3$temp2$Den.df, sep=",")`}$
-= `r bearded_type3$temp2$F.value`; $p$ =
-`r bearded_type3$temp2$Prob.F`). Additionally, wind chill was also a
-significant influence
-($F_{`r paste(bearded_type3$temp2_wind$Num.df, bearded_type3$temp2_wind$Den.df, sep=",")`}$
-= `r bearded_type3$temp2_wind$F.value`; $p$ =
-`r bearded_type3$temp2_wind$Prob.F`). Any influence of barometric
-pressure or precipitation is less apparent (Figure
-\@ref(fig:beardedHOwx)).
-
-```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across the range of weather conditions encountered in the observed data.
-Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
-Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}
-
-p_temp <- ggplot(plot_df, aes(temp2*27, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
-  xlab("temperature (C)") + ylab("haul-out probability")
-
-p_wind <- ggplot(plot_df, aes(wind, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
-  xlab("wind") + ylab("")
-
-p_pressure <- ggplot(plot_df, aes(pressure, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
-  xlab("barometric pressure") + ylab("haul-out probability")
-
-p_precip <- ggplot(plot_df, aes(precip, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
-  xlab("precipitation") + ylab("")
-
-library(patchwork)
-
-((p_temp + p_wind) / (p_pressure + p_precip)) +
-  plot_annotation(
-    title = 'Bearded seal haul-out behavior is less influeced by weather')
 ```
@@
 
+pal <- c(RColorBrewer::brewer.pal(12, "Paired"), "black")
+
 p <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(yday, ho_prob)) +
@@
                   shape = 15) +
-  scale_color_brewer(palette = "Paired") +
+  scale_color_discrete(type = pal) +
   scale_x_continuous(breaks = c(74,105,135,166),
@@
 
-In this paper, we modeled data from bio-logging sensors deployed on
-bearded, spotted, and ribbon seals to examine factors affecting haul-out
-behavior on sea ice in the Bering and Chukchi seas. Our analysis shows
-seals use sea ice progressively more through the spring, with peak use
-near mid-May to early June before declining again. This pattern aligns
-well with what has been previously documented and confirms our
-bio-logging deployments are likely capturing population-level behavioral
-patterns. Seals preferentially haul out on ice shortly after solar noon,
-which allows seals to maximize absorption of solar radiation
-[@feltz1966]. Interestingly, bearded seals appear to have two peaks in
-haul-out activity across the day, one shortly after solar noon, and one
-centered near solar midnight. A similar bi-modal pattern has been seen
-in ringed seals [@vonduyke2020] and suggests that bearded and ringed
-seals may be operating under different constraints than ribbon and
-spotted seals. Bearded and ringed seals are distributed across higher
-latitudes and the extended daylight hours may allow more flexibility in
-optimizing resting periods with foraging.
+In this paper, we modeled data from bio-logging sensors deployed on bearded,
+ribbon, and spotted seals to examine factors affecting haul-out behavior on sea
+ice in the Bering and Chukchi seas. Our analysis shows seals use sea ice
+progressively more through the spring, with peak use near mid-May to early June
+before declining again. This pattern aligns well with what has been previously
+documented and confirms our bio-logging deployments are likely capturing
+population-level behavioral patterns. Seals preferentially haul out on ice
+shortly after solar noon, which allows seals to maximize absorption of solar
+radiation [@feltz1966]. Interestingly, bearded seals appear to have two peaks in
+haul-out activity across the day, one shortly after solar noon, and one centered
+near solar midnight. A similar bi-modal pattern has been seen in ringed seals
+[@vonduyke2020] and suggests that bearded and ringed seals may be operating
+under different constraints than ribbon and spotted seals. Bearded and ringed
+seals are distributed across higher latitudes and the extended daylight hours
+may allow more flexibility in optimizing resting periods with foraging. Other
+factors such as polar bear predation (which is rare for ribbon and spotted
+seals in the Bering Sea) may also explain differing haul-out patterns.
 
 Unlike previous analyses of seal haul-out data (e.g. [@verhoef2010a],
-[@conn2014a]), we also investigated the influence of sex-age class
-on haul-out probabilities for all species except bearded seals because
-of low sample size. While both ribbon and spotted seals exhibited a
-unimodal diel haul-out pattern generally centered around local solar
-noon, there were key differences across species, age, and sex that match
-expectations given what we know of their ecological behavior. Spotted
+[@conn2014a]), we also investigated the influence of sex-age class on haul-out
+probabilities for all species except bearded seals because of low sample size.
+While both ribbon and spotted seals exhibited a unimodal diel haul-out pattern
+generally centered around local solar noon, there were key differences across
+species, age, and sex that match expectations given what we know of their
+ecological behavior. Ribbon seals are not known to form triads and our model
+predicts a progression of increased haul-out behavior with females starting
+earlier in the season than males. Notably, female ribbon seals spend a large
+portion of the day in the water during the pupping period, aligning with the
+hypothesis that ribbon seal females continue foraging while nursing. Spotted
 seals are known to form triads during the breeding season [@burns2009,
-@mcclintock2015] where a female and dependent pup are accompanied on the
-ice by a suitor male. The male waits for the female to wean the pup and
-enter estrus, and fends off any other males that may be interested.
-Triad formation results in both males and females spending a large
-portion of the day hauled out on ice and a protracted spring haul-out
-season for both sexes. Because mating occurs in the water, females may
-also be less inclined to venture out for foraging trips while still
-nursing the pup. We see this reflected in the predicted haul-out
-behavior, with both males and females exhibiting a broad distribution of
-time out of the water throughout the solar day and the season. Ribbon
-seals are not known to form triads and our model predicts a progression
-of increased haul-out behavior with females starting earlier in the
-season than males. Notably, female ribbon seals spend a large portion of
-the day in the water during the pupping period, aligning with the
-hypothesis that ribbon seal females continue foraging while nursing. In
-the case of both ribbon and spotted seals, subadults are the first to
-begin consistent haul-out behavior and follow a typical phocid pattern
-where subadults molt first as they do not have any reproductive
-constraints.
-
-We also investigated the influence of environmental covariates on
-haul-out probabilities, including wind speed, temperature, barometric
-pressure, precipitation, and wind chill. These have been investigated
-for walrus (e.g. Udevitz et al. [-@udevitz2009]) and a few select
-studies of ice-associated seals [@hamilton2018, @perry2017]. Ribbon
-seals seemed to be the most influenced by weather, with wind,
-temperature, barometric pressure, and precipitation all being important
-components of the model. Spotted seals were most affected by wind and
-barometric pressures. And, for bearded seals, the model indicated wind
-and temperature had the greatest impact. In general, and as might be
-expected, seals were more likely to haul-out when daily temperatures
-were warmer, winds speeds were lower, barometric pressure was higher,
-and precipitation was lower. These weather conditions are likely most
-beneficial for energetic considerations and may also provide the best
-situation for predator detection. These results highlight the importance
-for haul-out behavior models (and subsequent calculations of
-availability during surveys) to incorporate weather covariates.
-
-Our model uncovered annual deviations in the timing and magnitude of
-haul-out peaks for ribbon and spotted seals. The timing of peak haul-out
-activity appears to fall within a relatively narrow time window of 3-4
-weeks in late May and early June. This remarkable level of consistency
-across 15 years is indicative of the adaptations these species have
-evolved to synchronize their key life history stages with favorable
-environmental conditions (e.g. availability of sea ice, increased
-oceanographic productivity, favorable weather conditions).
-Unfortunately, this may also be indicative of a limited range of
-plasticity as climatological conditions change. Additionally, while from
-an ecological perspective the haul-out behavior appears consistent, the
-interannual differences in timing and magnitude are large enough to have
-important ramifications on calculations of abundance and trend.
-
-Previous attempts to estimate the abundance of phocid seals from aerial
-survey data in the Bering and Chukchi Seas (e.g. [@bengtson2005a],
-[@conn2014a], [@verhoef2014a]) have used estimated haul-out
-probabilities to correct for the proportion of animals that are in the
-water and thus unavailable to be counted during aerial surveys. Although
-several of these studies allowed haul-out probabilities to vary by
-day-of-year and time-of-day, previous correction factors have not
-accounted for variability between years, different weather conditions,
-and the age-sex class of the sample. In this paper, we have shown that
-there can be considerable differences in the number of seals hauled out
-on ice based on these factors. We recommend that future abundance
-analyses employ correction factors that account for these factors. For
-instance, it is relatively straightforward to obtain weather reanalysis
-products for times and locations that are surveyed and to construct a
-relevant correction factor based on predictions of GLMPMs. The most
-challenging element in developing availability correction factors is
-with annual variability. It can be difficult to get a sufficient sample
-size to estimate year-specific correction factors, particularly because
-research teams would likely need to tag seals and conduct aerial surveys
-concurrently which requires considerably more personnel and money. One
-possible suggestion is to estimate a "shift" parameter within models for
-aerial survey counts that allow the peak of haul-out distributions to be
-adjusted earlier or later in the year based on the frequency of counts
-observed over time. Regardless, researchers should anticipate there
-being some unmodeled heterogeneity in availability probability present
-in abundance estimates obtained from aerial surveys. This may make trend
-detection difficult, as one will not know if moderate differences in
-abundance estimates are attributable to changes in abundance or changes
-in haul-out frequencies.
+@mcclintock2015] where a female and dependent pup are accompanied on the ice by
+a suitor male. The male waits for the female to wean the pup and enter estrus,
+and fends off any other males that may be interested. Triad formation results in
+both males and females spending a large portion of the day hauled out on ice and
+a protracted spring haul-out season for both sexes. Because mating occurs in the
+water, females may also be less inclined to venture out for foraging trips while
+still nursing the pup. We see this reflected in the predicted haul-out behavior,
+with both males and females exhibiting a broad distribution of time out of the
+water throughout the solar day and the season.  In the case of both ribbon and
+spotted seals, subadults are the first to begin consistent haul-out behavior and
+follow a typical phocid pattern where subadults molt first as they do not have
+any reproductive constraints.
+
+We also investigated the influence of environmental covariates on haul-out
+probabilities, including wind speed, temperature, barometric pressure,
+precipitation, and wind chill. These have been investigated for walrus (e.g.
+Udevitz et al. [-@udevitz2009]) and a few select studies of ice-associated seals
+[@hamilton2018, @perry2017]. Ribbon seals seemed to be the most influenced by
+weather, with wind, temperature, barometric pressure, and precipitation all
+being important components of the model. Spotted seals were most affected by
+wind and barometric pressures. And, for bearded seals, the model indicated wind
+and temperature had the greatest impact. In general, and as might be expected,
+seals were more likely to haul-out when daily temperatures were warmer, winds
+speeds were lower, barometric pressure was higher, and precipitation was lower.
+These weather conditions are likely most beneficial for energetic considerations
+and may also provide the best situation for predator detection. These results
+highlight the importance for haul-out behavior models (and subsequent
+calculations of availability during surveys) to incorporate weather covariates.
+
+Our model uncovered annual deviations in the timing and magnitude of haul-out
+peaks for ribbon and spotted seals. The timing of peak haul-out activity appears
+to fall within a relatively narrow time window of 3-4 weeks in late May and
+early June. This consistency across 15 years is indicative of the adaptations
+these species have evolved to synchronize their key life history stages with
+favorable environmental conditions (e.g. availability of sea ice, increased
+oceanographic productivity, favorable weather conditions). Unfortunately, this
+may also be indicative of a limited range of plasticity as climatological
+conditions change. Additionally, while from an ecological perspective the
+haul-out behavior appears consistent, the interannual differences in timing and
+magnitude are large enough to have important ramifications on calculations of
+abundance and trend.
+
+Previous attempts to estimate the abundance of phocid seals from aerial survey
+data in the Bering and Chukchi Seas (e.g. [@bengtson2005a], [@conn2014a],
+[@verhoef2014a]) have used estimated haul-out probabilities to correct for the
+proportion of animals that are in the water and thus unavailable to be counted
+during aerial surveys. Although several of these studies allowed haul-out
+probabilities to vary by day-of-year and time-of-day, previous correction
+factors have not accounted for variability between years, different weather
+conditions, and the age-sex class of the sample. In this paper, we have shown
+that there can be considerable differences in the number of seals hauled out on
+ice based on these factors. We recommend that future abundance analyses employ
+correction factors that account for these factors. For instance, it is
+relatively straightforward to obtain weather reanalysis products for times and
+locations that are surveyed and to construct a relevant correction factor based
+on predictions of GLMPMs. The most challenging element in developing
+availability correction factors is with annual variability. It can be difficult
+to get a sufficient sample size to estimate year-specific correction factors,
+particularly because research teams would likely need to tag seals and conduct
+aerial surveys concurrently which requires considerably more personnel and
+money. One possible suggestion is to estimate a "shift" parameter within models
+for aerial survey counts that allow the peak of haul-out distributions to be
+adjusted earlier or later in the year based on the frequency of counts observed
+over time. Regardless, researchers should anticipate there being some unmodeled
+heterogeneity in availability probability present in abundance estimates
+obtained from aerial surveys. This may make trend detection difficult, as one
+will not know if moderate differences in abundance estimates are attributable to
+changes in abundance or changes in haul-out frequencies.
 

### 2021-12-13 analysis/paper/paper.Rmd

@@
 between 2005 and 2020 to investigate the seasonal timing and environmental
-factors affecting sea ice use by seals (specifically, bearded, ribbon, and
-spotted seals) in the Bering and Chukchi seas. In addition to providing baseline
-data on phenology, these data also allow us to quantify 'availability', which is
-needed to accurately estimate abundance from aerial survey counts of seals
-basking on ice (i.e., to correct for the proportion of animals that are in the
-water while surveys are conducted). Using generalized linear mixed pseudo-models
-to properly account for temporal autocorrelation, we fit models with a variety
-of covariates (e.g., day-of-year, solar hour, age-sex class, wind speed,
-barometric pressure, temperature, precipitation) to examine their ability to
-explain variation in hourly haul-out records. We found evidence for strong
-diurnal and within-season patterns in haul-out behavior, as well as strong
-weather effects (particularly wind and temperature). In general, seals were more
-likely to rest on ice in the middle of the day and when wind speed was low and
-temperatures were higher. Haul-out propensity increased through March and April,
-peaking in May and early June before declining again. The timing and frequency
-of haul-out events also varied based on species and age-sex class. For ribbon
-and spotted seals, models with year effects were highly supported, indicating
-that the timing and magnitude of haul-out behavior varied among years. However,
-we did not find evidence that haul-out timing was linked to annual sea ice
-extent. Our analysis emphasizes the importance of accounting for variation in
-haul-out behavior when interpreting the number of seals counted in aerial
-surveys.
+factors affecting haul-out behavior sea ice by seals (specifically, bearded,
+ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to
+providing baseline data on phenology, these data also allow us to quantify
+'availability', which is needed to accurately estimate abundance from aerial
+survey counts of seals basking on ice (i.e., to correct for the proportion of
+animals that are in the water while surveys are conducted). Using generalized
+linear mixed pseudo-models to properly account for temporal autocorrelation, we
+fit models with a variety of covariates (e.g., day-of-year, solar hour, age-sex
+class, wind speed, barometric pressure, temperature, precipitation) to examine
+their ability to explain variation in hourly haul-out records. We found evidence
+for strong diurnal and within-season patterns in haul-out behavior, as well as
+strong weather effects (particularly wind and temperature). In general, seals
+were more likely to haul out on ice in the middle of the day and when wind speed
+was low and temperatures were higher. Haul-out propensity increased through
+March and April, peaking in May and early June before declining again. The
+timing and frequency of haul-out events also varied based on species and age-sex
+class. For ribbon and spotted seals, models with year effects were highly
+supported, indicating that the timing and magnitude of haul-out behavior varied
+among years. However, we did not find evidence that haul-out timing was linked
+to annual sea ice extent. Our analysis emphasizes the importance of accounting
+for variation in haul-out behavior when interpreting the number of seals counted
+in aerial surveys.
 
@@
 
-For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals
-rest out of the water almost exclusively on sea ice and are mostly pelagic
-outside the spring pupping, breeding, and molting season [@burns1981g]. While
-spotted and bearded seals rest on beaches during ice-free months, they strongly
-prefer sea ice as a resting platform during the spring and early summer
-[@burns2009]. The remoteness of the Bering and Chukchi seas mean direct
-scientific observation of seal behavior is impractical. Bio-logging devices are,
-thus, especially useful tools for collecting key information on movement and
-haul-out behavior for these species.
+For ice-associated seals, haul-out behavior is tightly linked with sea ice.
+Ribbon seals haul out of the water almost exclusively on sea ice and are mostly
+pelagic outside the spring pupping, breeding, and molting season [@boveng2018a].
+While spotted and bearded seals rest on beaches during ice-free months, they
+strongly prefer sea ice as a resting platform during the spring and early summer
+[@frost2018]. The remoteness of the Bering and Chukchi seas mean direct
+scientific observation of seal behavior is impractical. Thus, bio-logging
+devices are, thus, especially useful tools for collecting key information on
+movement and haul-out behavior for these species.
 
@@
 in February and at times close to solar noon. In the Arctic, Von Duyke et al.
-[-@vonduyke2020] used seal-borne, satellite-linked loggers to corroborate
-seasonal changes between diurnal and nocturnal haul-out behavior of ringed seals
-previously described by Kelly et al. [-@kelly2010] using labor-intensive VHF
-radio tags and direct observation. Bengtson et al. [-@bengtson2005a] documented
-a higher propensity for ringed seal basking near solar noon, as did Ver Hoef et
-al. [-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using
-much larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion
-of time bearded seals spent hauled out progressively increased through spring
-and summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
+[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
+changes between diurnal and nocturnal haul-out behavior of ringed seals
+previously described by Kelly et al. [-@kelly2010] using VHF radio tags and
+direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
+propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
+[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
+larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
+time bearded seals spent hauled out progressively increased through spring and
+summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
 increased gradually starting in March and peaking in May and June for bearded,
@@
 
-While bio-logging of haul-out patterns clearly has utility for description of
-ice-associated seals' natural history and ecology, some studies have focused, as
-we do here, specifically on quantifying the probability of seals being hauled
-out as a means of accounting for seals in the water, and therefore missed,
-during aerial surveys. These studies typically use logistic regression-style
-analyses to estimate the time-specific probability of being hauled out base on
-'wet/dry' data relayed by bio-loggers. In these models, haul-out probabilities
-were expressed as a function of predictive covariates, such as time of day,
-day-of-year, sex, age class, and environmental conditions (e.g., [@reder2003],
-[@bengtson2004], [@bengtson2005a], [@udevitz2009], [@verhoef2014a], and
-[@southwell2008]). However, sample sizes have often been low and insufficient to
-permit strong inference about life history and/or seasonal variation in haul-out
-probabilities. For instance, Bengtson and Cameron's [-@bengtson2004] study
-included 5 adult and 2 juvenile crabeater seals, while Bengtson et al.'s
-[-@bengtson2005a] study was based on 6 telemetered ringed seals in the Chukchi
-Sea. These studies were often further limited by logistical constraints on
-tagging fieldwork and the attachment duration or operational life of satellite
-tags. In this study, we address some of these limitations by deploying small
-bio-loggers designed for attachment on a portion of the study individuals. These
-devices are designed to collect data through the molt period (when those adhered
-to the hair would fall off) and, in some situations, provide multiple years of
-data.
+While analyzing haul-out patterns clearly has utility for understanding natural
+history and ecology, some studies have focused, as we do here, specifically on
+quantifying the probability of seals being hauled out as a means of quantifying
+seals in the water that are not counted during aerial surveys. These studies
+typically use logistic regression-style analyses to estimate the time-specific
+probability of being hauled out based on 'wet/dry' data relayed by bio-loggers.
+In these models, haul-out probabilities were expressed as a function of
+predictive covariates, such as time of day, day-of-year, sex, age class, and
+environmental conditions (e.g., [@reder2003], [@bengtson2004], [@bengtson2005a],
+[@udevitz2009], [@verhoef2014a], and [@southwell2008]). However, sample sizes
+have often been insufficient to permit strong inference about life history
+and/or seasonal variation in haul-out probabilities. For instance, Bengtson and
+Cameron's [-@bengtson2004] study included 5 adult and 2 juvenile crabeater
+seals, while Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed
+seals in the Chukchi Sea. These studies were often further limited by logistical
+constraints on tagging fieldwork and the attachment duration or operational life
+of bio-loggers. In this study, we address some of these limitations by deploying
+small bio-loggers designed for longer-term attachment on a subset of the study
+individuals. These devices are designed to collect data through the molt period
+(when those adhered to the hair would fall off) and, in some situations, provide
+multiple years of data.
 
@@
 surrogates such as survival and recruitment) will be necessary to make credible
-quantitative predictions about the effects of the climate crisis on the
+quantitative predictions about the effects of climate disruptions on the
 abundance and distribution of Arctic seal populations. Before we can construct a
@@
 [@verhoef2014a]). Such studies were conducted over very large areas and
-estimation of absolute abundance required making inference about numerous
-nuisance processes affecting the observation of seals on ice. This included
-availability (only seals basking on ice were available to be counted), detection
-probability (observers or automated detection systems may have missed some seals
-on ice), species misclassification, and possible disturbance of seals by
-aircraft [@conn2014a;
-@verhoef2014a].
-
-In this study, we used data from 16 years of bio-logger records in a
-comprehensive investigation of haul-out behavior by 3 species of ice-associated
-seals in the Bering and Chukchi seas. We focused our effort on bearded, ribbon,
-and spotted seals and did not include ringed seals in the analysis. Ringed seals
-are unique in their use of snow lairs in the spring months, and they warrant a
-separate study that includes additional environmental covariates, such as the
-onset of snow melt and snow depth. Our goals are threefold. First, we wish to
-establish baseline estimates for the chronology of haul-out behavior in the
-critical spring season for each species across different age and sex classes.
-Second, we wish to refine estimates of haul-out availability corrections for
-aerial surveys in order to improve estimates of seal abundance. Previously
-estimated availability correction factors (e.g., [@bengtson2005a], [@conn2014a],
-and [@verhoef2014a]) accounted for variables such as the time-of-day and
-day-of-year, but did not investigate meteorological variables that have been
-shown to influence haul-out behavior of walruses [@reder2003; @udevitz2009].
-Third, we explored the annual variability in haul-out timing and possible
-linkage to dramatic changes in the availability and extent of seasonal sea ice
-between 2005 and 2020. Our work extends the scope of previous haul-out analyses,
-includes the influence of meteorological variability, and investigates the
-potential impact of a changing icescape on the behavior of these species.
+estimation of absolute abundance required making inference about numerous issues
+affecting the observation of seals on ice. These included availability (only
+seals basking on ice were available to be counted), detection probability
+(observers or automated detection systems may have missed some seals on ice),
+species misclassification, and possible disturbance of seals by aircraft
+[@conn2014a; @verhoef2014a].
+
+In this study, we used 16 years of bio-logging data to investigate the haul-out
+behavior of 3 species of ice-associated seals in the Bering and Chukchi seas. We
+focused our effort on bearded, ribbon, and spotted seals and did not include
+ringed seals in the analysis. Ringed seals are unique in their use of snow lairs
+in the spring months, and they warrant a separate study that includes additional
+environmental covariates, such as the onset of snow melt and snow depth. Our
+goals are threefold. First, we wish to establish baseline estimates for the
+chronology of haul-out behavior in the critical spring season for each species
+across different age and sex classes. Second, we wish to refine estimates of
+haul-out availability corrections for aerial surveys in order to improve
+estimates of seal abundance. Previously estimated availability correction
+factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
+for variables such as the time-of-day and day-of-year, but did not investigate
+meteorological variables that have been shown to influence haul-out behavior of
+walruses [@reder2003; @udevitz2009]. Third, we explored the annual variability
+in haul-out timing and possible linkage to changes in the extent of seasonal sea
+ice between 2005 and 2020. Our work extends the scope of previous haul-out
+analyses, includes the influence of meteorological variability, and investigates
+the potential impact of a changing icescape on the behavior of these species.
 
@@
 bio-loggers deployed on bearded, ribbon, and spotted seals was filtered to
-include only records from 1 March to 15 July between `r min(lubridate::year(analysis_data$haulout_dt))` 
-and `r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
+include only records from 1 March to 15 July between `r
+min(lubridate::year(analysis_data$haulout_dt))` and `r
+max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
 'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
@@
 annual molt, which limits the effective length of haul-out timelines.
-Additionally, sensors attached to the head or dorsal region are often dry while
-the seal is at the surface and can slightly bias the percent-dry values recorded
-by the device. For this study, in cases where both tag types were deployed,
-percent dry observations from the flipper tag were preferred. Sex and age class
-(non-dependent *young-of-the-year*, sexually immature *subadults*, and mature
-*adults*) were determined at the time of deployment by subjective judgement
-based on length, and pelage characteristics for some species. Field
-identification of age class can be inexact, particularly when discerning
-subadults from adults. In the case of ribbon seals, pelage differences between
-subadults and adults are distinguishable, with subadults often having less
-defined ribbons. Bearded seal subadults will often have a spotted pattern in the
-pelage that is not seen in adults. Spotted seal pelage is monomorphic and can
-not be used to discern age class. Despite these challenges, we feel the age
-classifications used in this analysis are useful categories in testing for
-age-related effects on haul-out behavior. Seals determined to be less than one
-year were classified as young-of-the-year. Subadults are those seals likely
-greater than one year of age but not yet sexually mature. Adults are likely
-sexually mature and older than approximately four years. For those tags deployed
-on young-of-the-year and transmitting into the next year, the age class was
-advanced to subadult on 1 March of the following year. Table
+Additionally, bio-loggers attached to the head or dorsal region are often dry
+while the seal is at the surface and can slightly bias the percent-dry values
+recorded by the device. For this study, in cases where both tag types were
+deployed, percent dry observations from the flipper tag were preferred. Sex and
+age class (non-dependent *young-of-the-year*, sexually immature *subadults*, and
+mature *adults*) were determined at the time of deployment by varying
+combinations of length, claw annuli, and pelage characteristics for some
+species. Field identification of age class can be inexact, particularly when
+discerning subadults from adults. In the case of ribbon seals, pelage
+differences between subadults and adults are distinguishable, with subadults
+often having less defined ribbons. Bearded seal subadults will often have a
+spotted pattern in the pelage that is not seen in adults. Spotted seal pelage is
+monomorphic and can not be used to discern age class. Despite these challenges,
+we feel the age classifications used in this analysis are useful categories in
+testing for age-related effects on haul-out behavior. Seals determined to be
+less than one year were classified as young-of-the-year. Subadults are those
+seals likely greater than one year of age but not yet sexually mature. Adults
+are likely sexually mature and older than four years. For those
+bio-loggers deployed on young-of-the-year and transmitting into the next year,
+the age class was advanced to subadult on 1 March of the following year. Table
 \@ref(tab:tblDeploy) provides a summary of these deployments and data received
@@
 
-```{r tblDeploy, include=TRUE, tab.id = "tblDeploy", tab.cap = "Summary of bio-logger deployments across seal species and age classification from 1 March to 15 July.
+```{r tblDeploy, include=TRUE, tab.id="tblDeploy",tab.cap = "Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2020.
 Total seal hours represents the sum of available data across all seals."}
@@
 send data to satellites, and we did not want to include data from these tags in
-our analyses. As such, end times of each deployment were identified by experts
-who examined several simultaneous data streams (e.g., tag locations and dive
-behavior) to determine when tags ceased providing data consistent with seal
-behavior. Sensor readings outside of these start and end times were discarded
-prior to analysis. After approximately 9 months, 7 devices deployed on the rear
-flipper of bearded seals reported implausible sensor data. All data after the
-first instance of unrealistic values were censored from this analysis. Figure
-\@ref(fig:dataCal) shows a calendar-like distribution of these data across the
-study season for each species. Observations for ribbon and spotted seals are
-concentrated in the months of May and June due to the timing of deployment
-(April and May) and the timing of molt (May and June). During molt, seals (and
-their attached bio-loggers) spend more time out of the water and more data is
-transmitted. Molt timing also impacts when many deployments end as any tags
-adhered to the hair will fall off. Bearded seal deployments are more evenly
-spread with fewer observations in May and June. The majority of bearded seal
-deployments start later in the summer and by May tags have either fallen off or
-the batteries are depleted.
-
-```{r seal-hours-calendar, include=TRUE, fig.height = 4.75, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species.
+our analyses. As such, end times of each deployment were identified by examining
+tag locations and dive behavior to determine when tags ceased providing data
+consistent with seal behavior. Sensor readings outside of these start and end
+times were discarded prior to analysis. After approximately 9 months, 7 devices
+deployed on the rear flipper of bearded seals reported implausible sensor data.
+All data after the first instance of unrealistic values were censored from this
+analysis. Figure \@ref(fig:dataCal) shows a calendar-like distribution of all
+haul-out data across the study season for each species. Observations for ribbon
+and spotted seals are concentrated in the months of May and June due to the
+timing of deployment (April and May) and the timing of molt (May and June).
+During molt, seals (and their attached bio-loggers) spend more time out of the
+water and more data is transmitted. Molt timing also impacts when many
+deployments end as any tags adhered to the hair will fall off. Bearded seal
+deployments are more evenly spread with fewer observations in May and June. The
+majority of bearded seal deployments start later in the summer and by May tags
+have either fallen off or the batteries are depleted.
+
+```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of haul-out behavior data from 1 March to 15 July for each species.
 The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out.
-Additionally, a majority of deployments started in April and May.
+Additionally, many bio-logger deployments started in April and May.
 Data were collated across all years between 2005 and 2020."}
+
 ggplot(data = seal_hours_tbl) + 
@@
        size = "seal hours of bio-logger data") +
-  ggtitle("The number of haul-out behavior observations varies from March through July",
+  ggtitle("Density of haul-out behavior observations varies from March through July",
           subtitle = "data presented in seal-hours collated across years (2005-2020)")
@@
 hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). Each data
-transmission message consisted of a complete 24-hour record. Tags were,
+transmission message consisted of a complete 24-hour record. Bio-loggers were,
 generally, deployed and programmed in a manner to maximize data transmission
-during the spring pupping and molting period, however, not all days during the
+during the spring pupping and molting period, however, not all days during a
 deployment were successfully transmitted. This is due to a variety of factors
@@
 ```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, and May.
-Empty areas of the plot within a month represent missing data that were not successfully transmitted from the tag."}
+Areas of the plot indicated with an 'X' represent missing data that were not successfully transmitted from the tag."}
+
 d <- analysis_data %>% 
@@
 
+d_deploy <- analysis_data %>% 
+  sf::st_drop_geometry() %>% 
+  filter(speno == "HF2016_1002") %>% 
+  dplyr::select(speno,deploy_dt,haulout_dt,percent_dry) %>% 
+  mutate(year = lubridate::year(deploy_dt),
+         month = lubridate::month(deploy_dt, label = TRUE),
+         day = lubridate::day(deploy_dt),
+         hour = lubridate::hour(deploy_dt))
+
+missing_hours <- tibble(haulout_dt = seq(min(d_deploy$haulout_dt),
+              max(d_deploy$haulout_dt),by = "hour")) %>% 
+  filter(!haulout_dt %in% d_deploy$haulout_dt) %>% 
+  filter(lubridate::month(haulout_dt) %in% c(3,4,5)) %>% 
+  mutate(percent_dry = 0) %>% 
+  mutate(year = lubridate::year(haulout_dt),
+         month = lubridate::month(haulout_dt,label=TRUE),
+         day = lubridate::day(haulout_dt),
+         hour = lubridate::hour(haulout_dt))
+
 p <- ggplot(d,aes(day,hour,fill=percent_dry))+
@@
 p <- p + facet_grid(year~month, scales = "free_x")
-#p <- p + scale_y_continuous(trans = "reverse", expand = c(0,0))
 p <- p + scale_x_continuous(breaks = c(5,15,25), expand = c(0,0)) +
@@
   xlab("day of month") + ylab("hour (UTC)") 
+p <- p + 
+  geom_point(aes(x = day, y = hour), data = missing_hours, 
+             color = "black", shape = 4, size=1)
 p
@@
 Of key interest in this study is the relationship between haul-out behavior and
-a variety of environmental covariates that vary with space and time. We explored
+environmental covariates that vary with space and time. We explored
 the use of a movement model to predict locations at specific times. However, the
@@
 
-```{r dataMap, fig.height = 3.75, include=TRUE, fig.id="dataMap", fig.cap="Spatial distribution of haul-out data during the months of March through July for each of the three species.
+```{r dataMap, fig.height = 3.75, fig.id="dataMap", include=TRUE, fig.cap="Spatial distribution of haul-out data during the months of March through July 15 for each of the three species.
 Data were collated across all years between 2005 and 2020."}
@@
 
-In addition to sex and age class, we gathered additional variables that might
+In addition to sex and age class, we analyzed variables that might
 help explain variation in haul-out probabilities. These included day-of-year
@@
 was calculated using the {solaR} package [@perpinan2012] within the R
-statistical environment [@rcoreteam2021] using the daily locations. We also
-compiled several meteorological variables shown to affect haul-out behavior in
+statistical environment [@rcoreteam2021] using the daily locations. We included
+several meteorological variables shown to affect haul-out behavior in
 other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017]. In particular, we
-linked locations from satellite tags to key weather values from the North
+linked locations from our biologgers to key weather values from the North
 American Regional Reanalysis (NARR) model produced by the National Centers for
@@
 For this study, NARR weather values were subset to the extent of our study area
-over the Bering and Chukchi Seas at 3-hr intervals based on the native grid
+over the Bering and Chukchi seas at 3-hr intervals based on the native grid
 resolution of 32 km (1024 sq. km). The following meteorological variables were
@@
 seals that are away from ice (e.g., on foraging trips). Additionally, many of
-the bio-logger deployments used in this study were single attachments to the
-rear flipper of the seal and do not provide at-sea locations, limiting our
+the bio-logger deployments consisted of a single device attached to the
+rear flipper of the seal which do not provide at-sea locations, limiting our
 ability to fully evaluate habitat preferences related to sea-ice. It should be
@@
 Chukchi seas. Lastly, our study was limited to the spring season when seal
-haul-out tendencies are strongly influenced by pupping, nursing, breeding, and
+haul-out tendencies are strongly influenced by pupping, nursing, breeding behavior, and
 molt and these drivers are likely more influential than specific sea-ice
 concentration. Crawford et al [-@crawford2019] compared haul-out probability
-models for bearded seals and found those that only included season (and not
+models for ringed seals and found those that only included season (and not
 sea-ice concentration) were the most parsimonious. For these reasons, we have
@@
 
-```{r tblCovariates, include=TRUE, tab.id = "tblCovariates", tab.cap = "Explanatory covariates used in analyses of binary haul-out records for ice-associated seals.
+```{r tblCovariates, tab.id = "tblCovariates", include=TRUE, tab.cap = "Explanatory covariates used in analyses of binary haul-out records for bearded, spotted, and ribbon seals.
 Note that we also considered select interactions (see article text) between these primary covariates.
 For instance, wind chill was represented by the interaction _temperature:wind_."}
+
 covariate_tbl <- tibble::tribble(
@@
 
-Haul-out records for seals are often characterized by multiple hours spent
-basking on ice alternating with long periods of foraging (Figure
+Haul-out records for seals are often characterized by sequential hours spent
+basking on ice alternating with long periods in the water (Figure
 \@ref(fig:examplePlot)). Commonly used statistical models for binary data (e.g.
@@
 fitted models with year by day-of-year interactions. However, in this case we
-only included *year:day* and *year:day2*, omitting the main effects of year as
-well as *year:day3* interactions because models with the latter effects were
+only included *year:day* and *year:day^2^*, omitting the main effects of year as
+well as *year:day^3^* interactions because models with the latter effects were
 numerically unstable. However, the modeled interactions were sufficient to allow
@@
 horizontal shift in timing of haul-out distributions does not affect the main
-effects or cubic terms in a polynomial regression model.
+effects or cubic terms in a polynomial regression model. Bearded seals were
+not included in this examination of inter-annual variation because of limited
+data across many years in the study.
 
@@
 
-Age and sex class were not included in the model for bearded seals and, as such,
+Age and sex class were not included in the model for bearded seals due to our
+lower sample size for adult and young-of-year age classes. As such,
 results are shown for all ages. Bearded seals were predicted to have a bi-modal
@@
 
-```{r beardedHOCal, fig.height = 3.5, include=TRUE, fig.id = "beardedHOCal", fig.cap="Predicted haul-out probability of bearded seals from 1 March to 15 July for all and sex classes."}
+```{r beardedHOCal, fig.height = 3.5, include=TRUE, fig.id = "beardedHOCal", fig.cap="Predicted haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes."}
+
 plot_df <- bearded_newdata %>%
@@
 
-Like ribbon and spotted seals, bearded seal haul-out probability appears
-most influenced by wind
+Bearded seal haul-out probability appears most influenced by wind
 ($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$
@@
 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}
+
 plot_df <- ribbon_newdata %>%
@@
 seals were less likely to haul out at higher winds and lower pressure
-values and more likely to be on the ice when temperatures were
+values and more likely to haul out when temperatures were
 relatively warm. Wind chill
@@
 that is less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)).
-Adults of both sexes spend considerable time in April and May resting on the
-ice. As with ribbon seals, the young-of-the-year records begin after weaning and
+Adults of both sexes spend considerable time in April, May and June resting on the
+ice. In contrast to ribibon seals, adult spotted seal males have a more protracted
+haul-out season compared to females, and more time spent resting in June and
+July (Figures \@ref(fig:ribbonHOCal) and \@ref(fig:spottedHOCal)). As with 
+ribbon seals, the young-of-the-year records begin after weaning and
 the model predictions demonstrate the ontogeny of in-water activities (e.g.
@@
 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}
+
 plot_df <- spotted_newdata %>%
@@
 Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}
+
 library(ggsci)
@@
 Additionally, any groups where the underlying data started after 1 June or ended before 1 May are not included."}
+
 select_groups <- ribbon_year_fit$dataset %>%
@@
 The timing of peak haul-out probability for ribbon and spotted seals
-appeared to have only a limited to no relationship with the amount of
+appeared to have a limited to no relationship with the amount of
 yearly maximum sea ice extent. Adult female and subadult spotted seals
@@
 While spotted seal adult females and subadults show a negative trend line, none of the relationships were determined to be significant."}
+
 ggplot(data = peak_ho_df, aes(y = yday, x = max_extent_km2, color = species)) + 
@@
 
-In this paper, we modeled data from bio-logging sensors deployed on bearded,
-ribbon, and spotted seals to examine factors affecting haul-out behavior on sea
-ice in the Bering and Chukchi seas. Our analysis shows seals use sea ice
-progressively more through the spring, with peak use near mid-May to early June
-before declining again. This pattern aligns well with what has been previously
-documented and confirms our bio-logging deployments are likely capturing
-population-level behavioral patterns. Seals preferentially haul out on ice
-shortly after solar noon, which allows seals to maximize absorption of solar
-radiation [@feltz1966]. Interestingly, bearded seals appear to have two peaks in
-haul-out activity across the day, one shortly after solar noon, and one centered
-near solar midnight. A similar bi-modal pattern has been seen in ringed seals
+We modeled data from bio-loggers deployed on bearded, ribbon, and spotted seals
+to examine factors affecting haul-out behavior on sea ice in the Bering and
+Chukchi seas. Our analysis shows seals haul out progressively more through the
+spring and peak near mid-May to early June before declining again. This pattern
+aligns well with what has been previously documented and confirms our haul-out
+data are likely capturing population-level behavioral patterns. Seals
+preferentially haul out on ice shortly after solar noon, which allows seals to
+maximize absorption of solar radiation thought to facilitate the molting process
+[@feltz1966]. Interestingly, bearded seals appear to have two peaks in haul-out
+activity within a day, one shortly after solar noon, and one centered near solar
+midnight. A similar bi-modal pattern has been seen in ringed seals
 [@vonduyke2020] and suggests that bearded and ringed seals may be operating
@@
 may allow more flexibility in optimizing resting periods with foraging. Other
-factors such as polar bear predation (which is rare for ribbon and spotted
-seals in the Bering Sea) may also explain differing haul-out patterns.
+factors such as polar bear predation (which is rare for ribbon and spotted seals
+in the Bering Sea) may also explain differing haul-out patterns. Bearded seals also
+showed a more protracted season of haul-out behavior compared to ribbon and
+spotted seals. This aligns with findings from Thometz et al. [@thometz2021] who
+observed a mean molting period of 119$\pm$2 days and a relatively stable resting
+metabolic rate during that time. In the same study, spotted and ringed seals
+underwent molt periods of 33$\pm$4 and 28$\pm$6 days and exhibited increased
+resting metabolic rates.
 
@@
 probabilities, including wind speed, temperature, barometric pressure,
-precipitation, and wind chill. These have been investigated for walrus (e.g.
+precipitation, and wind chill. These have been investigated for walruses (e.g.
 Udevitz et al. [-@udevitz2009]) and a few select studies of ice-associated seals
@@
 
-Our model uncovered annual deviations in the timing and magnitude of haul-out
-peaks for ribbon and spotted seals. The timing of peak haul-out activity appears
-to fall within a relatively narrow time window of 3-4 weeks in late May and
-early June. This consistency across 15 years is indicative of the adaptations
-these species have evolved to synchronize their key life history stages with
-favorable environmental conditions (e.g. availability of sea ice, increased
-oceanographic productivity, favorable weather conditions). Unfortunately, this
-may also be indicative of a limited range of plasticity as climatological
-conditions change. Additionally, while from an ecological perspective the
-haul-out behavior appears consistent, the interannual differences in timing and
-magnitude are large enough to have important ramifications on calculations of
-abundance and trend.
+Our models detected small deviations in the timing and magnitude of annual peaks
+in haul-out behavior for ribbon and spotted seals. The timing of peak haul-out
+activity appears to fall within a relatively narrow time window of 3-4 weeks in
+late May and early June. This consistency across 15 years is likely a reflection
+of the relationship between a critical photoperiod and the timing of important
+life history stages [@bronson2009; @temte1994]. However, along with a critical
+photoperiod, ribbon and spotted seals are dependent upon the presence of sea ice
+for pupping and molt. We didn't find any support in our models for a
+relationship between the timing of peaks in haul-out behavior and the amount of
+yearly maximum sea ice. This could indicate that, while the extent of spring sea
+ice in the Bering sea varied widely during our study period, seals were still
+able to locate sea ice and haul out. We should expect, however, that some
+minimal threshold in the spatial extent or density of sea ice will have a
+meaningful impact on the timing of peak haul-out behavior -- if there is no sea
+ice, seals will not haul out. Additionally, while from an ecological perspective
+the haul-out behavior appears consistent, the interannual differences in timing
+and magnitude are large enough to have important ramifications on calculations
+of abundance and trend. Those ramifications will only be exacerbated if climate
+variability amplifies those interannual differences.
 
 Previous attempts to estimate the abundance of phocid seals from aerial survey
-data in the Bering and Chukchi Seas (e.g. [@bengtson2005a], [@conn2014a],
+data in the Bering and Chukchi seas (e.g. [@bengtson2005a], [@conn2014a],
 [@verhoef2014a]) have used estimated haul-out probabilities to correct for the
-proportion of animals that are in the water and thus unavailable to be counted
-during aerial surveys. Although several of these studies allowed haul-out
-probabilities to vary by day-of-year and time-of-day, previous correction
-factors have not accounted for variability between years, different weather
-conditions, and the age-sex class of the sample. In this paper, we have shown
-that there can be considerable differences in the number of seals hauled out on
-ice based on these factors. We recommend that future abundance analyses employ
-correction factors that account for these factors. For instance, it is
-relatively straightforward to obtain weather reanalysis products for times and
-locations that are surveyed and to construct a relevant correction factor based
-on predictions of GLMPMs. The most challenging element in developing
-availability correction factors is with annual variability. It can be difficult
-to get a sufficient sample size to estimate year-specific correction factors,
-particularly because research teams would likely need to tag seals and conduct
-aerial surveys concurrently which requires considerably more personnel and
-money. One possible suggestion is to estimate a "shift" parameter within models
-for aerial survey counts that allow the peak of haul-out distributions to be
-adjusted earlier or later in the year based on the frequency of counts observed
-over time. Regardless, researchers should anticipate there being some unmodeled
-heterogeneity in availability probability present in abundance estimates
-obtained from aerial surveys. This may make trend detection difficult, as one
-will not know if moderate differences in abundance estimates are attributable to
-changes in abundance or changes in haul-out frequencies.
+proportion of animals that are in the water and thus unavailable to be counted.
+Although several of these studies allowed haul-out probabilities to vary by
+day-of-year and time-of-day, they have not accounted for variability among
+years, in weather conditions, and in the age-sex class of the sample. In this
+paper, we have shown that there can be considerable differences in the number of
+seals hauled out on ice based on these factors. We recommend that future
+abundance analyses employ availability models that account for these factors. For
+instance, it is relatively straightforward to obtain weather reanalysis products
+for times and locations that are surveyed and to construct a relevant correction
+factor based on predictions of GLMPMs. The most challenging element in
+developing availability correction factors is with annual variability. It can be
+difficult to get a sufficient sample size to estimate year-specific correction
+factors, particularly because research teams would likely need to tag seals and
+conduct aerial surveys concurrently which requires considerably more personnel
+and money. One possible suggestion is to estimate a "shift" parameter within
+models for aerial survey counts that allow the peak of haul-out distributions to
+be adjusted earlier or later in the year based on the frequency of counts
+observed over time. Regardless, researchers should anticipate there being some
+unmodeled heterogeneity in availability probability present in abundance
+estimates obtained from aerial surveys. This may make trend detection difficult,
+as one will not know if moderate differences in abundance estimates are
+attributable to changes in abundance or changes in haul-out behavior.
 
 Predictions of absolute haul-out probability in this paper were somewhat
-different than those previously reported for these species, especially
-for bearded seals. For instance, Ver Hoef et al. [-@verhoef2014a)] and
-Conn et al. [-@conn2014a] used haul-out correction factors with
-maximums of 0.66 for bearded seals, 0.62 for ribbon seals, and 0.54 for
-spotted seals, where maximums corresponded to times near solar noon in
-mid-late May. Applying models that ignore age, sex, and year effects,
-these probabilities were 0.38, 0.72, and 0.60, respectively, under the
-current analysis framework. Our current estimates reflect increased
-sample sizes in terms of number of animals, but also improvements to the
-way data were prepared prior to analysis.
-
-We focused this paper on haul-out behavior of bearded, ribbon, and
-spotted seals. Ringed seals are also present in the Bering and Chukchi
-Seas but exhibit qualitatively different haul-out behavior. For
-instance, many ringed seals build subnivean layers in the snow on top of
-the sea ice, which they frequently inhabit before snow melt causes their
-lairs to collapse [@kelly1990]. Thus, the wet-dry sensor on a satellite
-tag could indicate that an animal is hauled out, but it could still be
-unavailable to be detected because it is in a lair. We hope to address
-availability of ringed seals using data from satellite tags, replicate
-survey tracks, and auxiliary information about snow depth and timing of
-melt in a future study.
+different than those previously reported for these species, especially for
+bearded seals. For instance, Ver Hoef et al. [-@verhoef2014a)] and Conn et al.
+[-@conn2014a] used haul-out correction factors with maximums of 0.66 for bearded
+seals, 0.62 for ribbon seals, and 0.54 for spotted seals, where maximums
+corresponded to times near solar noon in mid-late May. Applying models that
+ignore age, sex, and year effects, these probabilities were 0.38, 0.72, and
+0.60, respectively, under the current analysis framework. Our current estimates
+estimates of haul-out probability reflect increased sample sizes in terms of
+number of animals, but also improvements to the way data were prepared prior to
+analysis.
+
+We focused this paper on haul-out behavior of bearded, ribbon, and spotted
+seals. Ringed seals are also present in the Bering and Chukchi seas but exhibit
+a unique complicating factor. Adult ringed seals build subnivean lairs in the
+snow on top of the sea ice, where they haul out and pup until snow melt causes
+their lairs to collapse [@kelly1990]. Thus, the wet-dry sensor on a bio-logger
+could indicate that an animal is hauled out, but if it is within a lair, it is
+not available to be detected during an aerial survey. We hope to address
+availability of ringed seals using data from satellite tags, replicate survey
+tracks, and auxiliary information about snow depth and timing of melt in a
+future study.
 
@@
 these ecosystems. The deployment of bio-logging devices used in this
-study were often done in collaboration with Alaska Native communities.
+study were often done in collaboration with Alaska Native seal hungers and their communities.
 We would like to especially acknowledge the communities of Kotzebue,
-Utqiagvik, and Ulguniq and the following individuals: James Adams, Jeff
+Utqiaġvik, and Ulguniq (Wainwright) and the following individuals: James Adams, Jeff
 Barger, David Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris,
@@
 
-Telemetry data from the Alaska Department of Fish and Game and the North
-Slope Borough Department of Wildlife Management were important
+Telemetry data from the Alaska Department of Fish and Game (ADFG) and the North
+Slope Borough Department of Wildlife Management (NSB) were important
 contributions to the findings presented here. Both agencies collaborate
@@
 
-The findings and conclusions in the paper are those of the author(s) and
-do not necessarily represent the views of the National Marine Fisheries
-Service, NOAA. Any use of trade, product, or firm names does not imply
-an endorsement by the U.S. Government. Funding for this study was
-provided by the U.S. National Oceanic and Atmospheric Administration.
-The field work was conducted under the authority of Marine Mammal
-Protection Act Research Permits Nos. 782-1676, 782-1765, 15126, and
-19309 issued by the National Marine Fisheries Service, and Letters of
-Assurance of Compliance with Animal Welfare Act regulations, Nos. A/NW
-2010-3 and A/NW 2016-1 from the Alaska Fisheries Science
-Center/Northwest Fisheries Science Center Institutional Animal Care and
-Use Committee.
+The findings and conclusions in the paper are those of the author(s) and do not
+necessarily represent the views of the National Marine Fisheries Service, NOAA.
+Any use of trade, product, or firm names does not imply an endorsement by the
+U.S. Government. Funding for this study was provided by the U.S. National
+Oceanic and Atmospheric Administration. The field work was conducted under the
+authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
+782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
+Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
+A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
+Fisheries Science Center Institutional Animal Care and Use Committee (IACUC). ).
+Funding to ADFG for tagging seals was provided by the Bureau of Ocean Energy
+Management (No. M13PC0015) and the Office of Naval Research (No.
+N00014-16-1-3019). ADFG and NSB field work was covered by Research Permits Nos.
+358-1585, 358-1787, 15324, and 20466 and by ADFG IACUC permits Nos. 06-16,
+09-21, 2010-13, 2012-020, 0027-2019-41, 2013-020, 2014-03,2015-25, 2016-23,
+0027-2017-27, 0027-2018-29, 0027-2019-041.
 

### 2021-12-15 analysis/paper/paper.Rmd
EDITS THROUGH METHODS
initial edits in response to Justin C comments

@@
 between 2005 and 2020 to investigate the seasonal timing and environmental
-factors affecting haul-out behavior sea ice by seals (specifically, bearded,
+factors affecting haul-out behavior by ice-associated seals (specifically, bearded,
 ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to
@@
 linear mixed pseudo-models to properly account for temporal autocorrelation, we
-fit models with a variety of covariates (e.g., day-of-year, solar hour, age-sex
+fit models with covariates of interest (e.g., day-of-year, solar hour, age-sex
 class, wind speed, barometric pressure, temperature, precipitation) to examine
-their ability to explain variation in hourly haul-out records. We found evidence
+their ability to explain variation in hourly haul-out probability. We found evidence
 for strong diurnal and within-season patterns in haul-out behavior, as well as
@@
 were more likely to haul out on ice in the middle of the day and when wind speed
-was low and temperatures were higher. Haul-out propensity increased through
+was low and temperatures were higher. Haul-out probability increased through
 March and April, peaking in May and early June before declining again. The
@@
 to annual sea ice extent. Our analysis emphasizes the importance of accounting
-for variation in haul-out behavior when interpreting the number of seals counted
+for seasonal and temporal variation in haul-out behavior, as well as associated
+environmental covariates, when interpreting the number of seals counted
 in aerial surveys.
@@
 and large knowledge gaps complicate predictions about the ultimate effects of
-changes in sea ice on the behavior, abundance, and distribution of these seals.
-Knowledge about evolutionary constraints on the timing of reproductive and
-molting behavior is generally lacking, so it is difficult to predict the
-readiness with which ice-associated seal species will be able to adapt to future
-changes (e.g., by adjusting pupping or molting schedules to earlier dates or
-different locales). Additionally, trends in abundance of these species are
-largely unknown, so it is difficult to say what effect, if any, declines in sea
-ice habitat have had, or will have, on seal densities.
+changes in sea ice on the behavior, health, abundance, and distribution of these
+seals. To date, indices of seal health sampled during periods of declining sea
+ice differ regionally [@crawford2015; @harwood2020]. Knowledge about
+evolutionary constraints on the timing of reproductive and molting behavior is
+generally lacking, so it is difficult to predict the readiness with which
+ice-associated seal species will be able to adapt to future changes (e.g., by
+adjusting pupping or molting schedules to earlier dates or different locales).
+Additionally, trends in abundance of these species are largely unknown, so it is
+difficult to assess the effect, if any, declines in sea ice habitat have had, or
+will have, on seal densities.
 
@@
 pelagic outside the spring pupping, breeding, and molting season [@boveng2018a].
-While spotted and bearded seals rest on beaches during ice-free months, they
+While spotted and bearded seals rest on coastal features, they
 strongly prefer sea ice as a resting platform during the spring and early summer
@@
 In these models, haul-out probabilities were expressed as a function of
-predictive covariates, such as time of day, day-of-year, sex, age class, and
+predictive covariates, such as time-of-day, day-of-year, sex, age class, and
 environmental conditions (e.g., [@reder2003], [@bengtson2004], [@bengtson2005a],
@@
 seals in the Chukchi Sea. These studies were often further limited by logistical
-constraints on tagging fieldwork and the attachment duration or operational life
+constraints on fieldwork and the attachment duration or operational life
 of bio-loggers. In this study, we address some of these limitations by deploying
@@
 the Arctic using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
-[@verhoef2014a]). Such studies were conducted over very large areas and
+[@verhoef2014a]). Such abundance studies were conducted over very large areas and
 estimation of absolute abundance required making inference about numerous issues
@@
 species misclassification, and possible disturbance of seals by aircraft
-[@conn2014a; @verhoef2014a].
+[@conn2014a; @verhoef2014a]. Refining these inferences will improve the
+accuracy of abundance estimates in the Arctic.
 
@@
 environmental covariates, such as the onset of snow melt and snow depth. Our
-goals are threefold. First, we wish to establish baseline estimates for the
+goals were threefold. First, we wish to establish baseline estimates for the
 chronology of haul-out behavior in the critical spring season for each species
-across different age and sex classes. Second, we wish to refine estimates of
+across different age and sex classes. Second, we refined estimates of
 haul-out availability corrections for aerial surveys in order to improve
@@
 meteorological variables that have been shown to influence haul-out behavior of
-walruses [@reder2003; @udevitz2009]. Third, we explored the annual variability
+walruses [@reder2003; @udevitz2009]. Third, we assessed the annual variability
 in haul-out timing and possible linkage to changes in the extent of seasonal sea
@@
 ```
+Data used in our study came from bio-loggers deployed on bearded, ribbon, and
+spotted seals in the Bering, Chukchi, and Beaufort seas by multiple
+organizations as part of collaborative investigations during 2005 and 2020.
+Seals were captured using nets and instrumented during ship-based capture trips
+and from coastal Alaska communities. Ship-based capture events occurred during
+spring near the southern ice edge in the Bering Sea from 20xx to 20xx (maybe
+provide sample sizes for each species). Land-based capture events occurred from
+May to October, generally between the coastal communities of Scammon Bay,
+Alaska, in the Bering Sea, and Utqiaġvik, Alaska, near the junction of the
+Chukchi and Beaufort seas, from 2005 to 2020. We refer readers to the primary
+literature for detailed capture and bio-logger attachment methods (Supplemental
+Material, Table).
 
 Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
-bio-loggers deployed on bearded, ribbon, and spotted seals was filtered to
+bio-loggers deployed on bearded, ribbon, and spotted seals were subset to
 include only records from 1 March to 15 July between `r
@@
 satellite transmittal rates, locations at sea) but these are lost during the
-annual molt, which limits the effective length of haul-out timelines.
+annual molt, which limits the duration of haul-out data they provide.
 Additionally, bio-loggers attached to the head or dorsal region are often dry
-while the seal is at the surface and can slightly bias the percent-dry values
-recorded by the device. For this study, in cases where both tag types were
-deployed, percent dry observations from the flipper tag were preferred. Sex and
+while the seal is at the surface and can slightly bias the hourly percent-dry values
+reported by the bio-logger. For this study, in cases where both bio-logger types were
+deployed, hourly percent dry observations from the flipper tag were preferred. Sex and
 age class (non-dependent *young-of-the-year*, sexually immature *subadults*, and
 mature *adults*) were determined at the time of deployment by varying
-combinations of length, claw annuli, and pelage characteristics for some
+combinations of length, claw annuli [@burns1969; @burns1983; @mclaren1958a], and pelage characteristics for some
 species. Field identification of age class can be inexact, particularly when
@@
 Tags that fall off due to molt or mortality and remain on ice or land can still
-send data to satellites, and we did not want to include data from these tags in
-our analyses. As such, end times of each deployment were identified by examining
-tag locations and dive behavior to determine when tags ceased providing data
-consistent with seal behavior. Sensor readings outside of these start and end
+send data to satellites; i.e., the bio-logger will be on ice or land and dry, therefore, it will record and transmit data suggesting the seal is hauled out.
+As such, end times of each deployment were identified by examining
+bio-logger locations and dive behavior to determine when bio-loggers ceased providing data
+consistent with seal behavior. Data outside of the deployment start and end
 times were discarded prior to analysis. After approximately 9 months, 7 devices
-deployed on the rear flipper of bearded seals reported implausible sensor data.
+deployed on the rear flipper of bearded seals reported implausible hourly percent-dry data.
 All data after the first instance of unrealistic values were censored from this
@@
 During molt, seals (and their attached bio-loggers) spend more time out of the
-water and more data is transmitted. Molt timing also impacts when many
-deployments end as any tags adhered to the hair will fall off. Bearded seal
+water and more data are transmitted. Molt timing also impacts when many
+deployments end as any bio-loggers adhered to the hair will fall off. Bearded seal
 deployments are more evenly spread with fewer observations in May and June. The
-majority of bearded seal deployments start later in the summer and by May tags
+majority of bearded seal deployments start later in the summer and by May bio-loggers
 have either fallen off or the batteries are depleted.
@@
 sensor was polled by the tag firmware every few seconds and a percent of the
-hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). Each data
-transmission message consisted of a complete 24-hour record. Bio-loggers were,
-generally, deployed and programmed in a manner to maximize data transmission
-during the spring pupping and molting period, however, not all days during a
-deployment were successfully transmitted. This is due to a variety of factors
-including satellite coverage, tag availability (i.e. tags mounted to the rear
-flipper are not available to the surface while the seal is at sea), tag
-performance, duty cycling, and extra-terrestrial interference.
+hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). Hourly
+percent-dry data were usually rounded to the nearest 10 percent inclusive of 0
+and 100 percent and additional values at 3 and 98 percent. This compression
+resulted in additional data transmission as each message consisted of two
+complete 24-hour records. Bio-loggers were, generally, deployed and programmed
+in a manner to maximize data transmission during the spring pupping and molting
+period, however, hourly percent-dry data for days during a deployment were not
+successfully transmitted. This is due to a variety of factors including
+satellite coverage, tag availability (i.e. tags mounted to the rear flipper are
+not available to the surface while the seal is at sea), tag performance, duty
+cycling, and extra-terrestrial atmospheric interference.
 
@@
 Of key interest in this study is the relationship between haul-out behavior and
-environmental covariates that vary with space and time. We explored
-the use of a movement model to predict locations at specific times. However, the
-sparse nature of data from some tags, especially those mounted to the rear
-flipper, resulted in poor modeling performance or convergence issues. For this
-study, we calculated a weighted average daily location where the inverse of the
-estimated Argos location error was used for the weight. Any days where haul-out
-observations were present without any location data were filled from the last
-calculated weighted average daily location, and any days where the location
-intersected with land were removed from the dataset. Figure \@ref(fig:dataMap)
-shows the spatial distribution of determined locations and available haul-out
-behavior data for each species across the study area.
+environmental covariates that vary with space and time. We explored the use of a
+continuous-time correlated random walk [@devins.johnson2008] movement model to
+predict locations at specific times. However, the sparse nature of data from
+some bio-loggers, especially those mounted to the rear flipper, resulted in poor
+modeling performance or convergence issues. For this study, we calculated a
+weighted average daily location where the inverse of the estimated Argos or
+FastLoc GPS location error was used for the weight. Each Argos location estimate
+was assigned an error radius based on either the categorical location quality (
+_3_=250m, _2_=500m, _1_=1500m, _0_=2500m, _A_=2500m, _B_=2500m) or, when
+available, the estimated error radius from the Argos Kalman filter algorithm.
+Location estimates from FastLoc GPS were all assigned an error radius of 50m.
+Any days where haul-out observations were present but location data were missing
+we used the last calculated weighted average daily location, and any days where
+the location intersected with land were removed from the data set. We recognized
+that bearded and spotted seals haul out on land. However, assessing the
+relationship between haul-out behavior and environmental covariates and their
+availability for aerial surveys on land was outside the scope of this study.
+Additionally, any daily locations on land are likely more reflective of
+coordinate averaging and measurement error instead of actual use of coastal
+features. Figure \@ref(fig:dataMap) shows the spatial distribution weighted
+locations with available haul-out behavior data for each species across the
+study area.
 
@@
 
-In addition to sex and age class, we analyzed variables that might
-help explain variation in haul-out probabilities. These included day-of-year
-(for seasonal effects) and local solar hour (for diurnal effects). Solar hour
-was calculated using the {solaR} package [@perpinan2012] within the R
-statistical environment [@rcoreteam2021] using the daily locations. We included
-several meteorological variables shown to affect haul-out behavior in
-other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017]. In particular, we
-linked locations from our biologgers to key weather values from the North
-American Regional Reanalysis (NARR) model produced by the National Centers for
-Environmental Prediction [@mesinger2006]. The NARR model assimilates
+In addition to sex and age class, we analyzed variables that might help explain
+variation in haul-out probabilities. These included day-of-year (for seasonal
+effects) and local solar hour (for diurnal effects). Solar hour was calculated
+using the {solaR} package [@perpinan2012] within the R statistical environment
+[@rcoreteam2021] using the daily locations. We included several meteorological
+variables shown to affect haul-out behavior in other Arctic pinnipeds
+[@reder2003; @udevitz2009; @perry2017]. In particular, we linked the weighted
+average daily locations from our bio-loggers to key weather values from the
+North American Regional Reanalysis (NARR) model produced by the National Centers
+for Environmental Prediction [@mesinger2006]. The NARR model assimilates
 observational data to produce a long-term picture of weather over North America.
@@
 (represented by a *wind:temperature* interaction [@udevitz2009]), and
-day-of-year and time of day interactions to permit diurnal patterns to change
-throughout the year. Ribbon and spotted seal models included age-sex class and
-interactions between day-of-year and age-sex class, but we omitted these from
-bearded seal models due to poor representation of age-sex classes (Table
-\@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect (and an
-interaction with day-of-year), since bearded seals occupy a substantial range
-and we were interested in possible differences in the timing of haul-out
+day-of-year and time-of-day (solar hour) interactions to permit diurnal patterns
+to change from March to July. Ribbon and spotted seal models included age-sex
+class and interactions between day-of-year and age-sex class, but we omitted
+these from bearded seal models due to poor representation of age-sex classes
+(Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
+(and an interaction with day-of-year), since bearded seals occupy a substantial
+range and we were interested in possible differences in the timing of haul-out
 behavior along a latitudinal gradient. We omitted the latitudinal effect from
@@
 seals that are away from ice (e.g., on foraging trips). Additionally, many of
-the bio-logger deployments consisted of a single device attached to the
-rear flipper of the seal which do not provide at-sea locations, limiting our
-ability to fully evaluate habitat preferences related to sea-ice. It should be
-noted that the aerial survey counts are modeled with sea-ice and other habitat
-parameters. So, any final abundance estimates that combine models of survey
-counts and our model for proportion of seals hauled out on ice do account for
-the strong association between these species and sea-ice in the Bering and
-Chukchi seas. Lastly, our study was limited to the spring season when seal
-haul-out tendencies are strongly influenced by pupping, nursing, breeding behavior, and
-molt and these drivers are likely more influential than specific sea-ice
-concentration. Crawford et al [-@crawford2019] compared haul-out probability
-models for ringed seals and found those that only included season (and not
-sea-ice concentration) were the most parsimonious. For these reasons, we have
-elected not to use sea ice concentration as a predictor for haul-out
-probability; studies to evaluate habitat selection in relation to sea-ice
-parameters and other environmental covariates are in progress.
-
-We explored whether the annual variation in maximum spring sea ice extent in the
+the deployments consisted of a single device attached to the rear flipper of the
+seal which do not provide at-sea locations, limiting our ability to fully
+evaluate habitat preferences related to sea-ice. It should be noted that the
+aerial survey counts are modeled with sea-ice and other habitat parameters. So,
+any final abundance estimates that combine models of survey counts and our model
+for proportion of seals hauled out on ice do account for the strong association
+between these species and sea-ice in the Bering and Chukchi seas. Lastly, our
+study was limited to the spring season when seal haul-out tendencies are
+strongly influenced by pupping, nursing, breeding behavior, and molt and these
+drivers are likely more influential than specific sea-ice concentration.
+Crawford et al [-@crawford2019] compared haul-out probability models for ringed
+seals and found those that only included season (and not sea-ice concentration)
+were the most parsimonious. Further, studies to evaluate habitat selection in
+relation sea-ice parameters and other environmental covariates are in progress.
+For these reasons, we have elected not to use sea ice concentration as a
+predictor for haul-out probability.
+
+We assessed whether the annual variation in maximum spring sea ice extent in the
 Bering Sea influences the seasonal peak of seal haul-out behavior. In
@@
 maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
-study area with greater than 15% concentration were summed to get the total sea
+study area with greater than 15% concentration were counted daily to get the total sea
 ice extent for each day between 15 February and 15 July across all years.
-Maximum spring sea ice extent was simply the largest daily value for each year.
+Maximum spring sea ice extent was simply the largest daily count of grid cells
+with greater than 15 percent concentration for each year.
 
@@
 We followed Ver Hoef et al. [-@verhoef2014a] in using linear, quadratic, and
-cubic effects of day-of-year to represent seasonal changes in behavior. However,
+cubic effects of day-of-year to represent temporal changes in behavior. However,
 unlike previous models for harbor seals [@london2012] and ice-associated seals

### 2021-12-18 analysis/paper/paper.Rmd
most comments addressed

@@
 whether haul-out patterns varied by year and to determine the annual timing of
-apparent peaks in haul-out behavior.
+apparent peaks in haul-out behavior. For both models, the response variable was
+the transmitted hourly percent-dry value.
 
@@
 lower sample size for adult and young-of-year age classes. As such,
-results are shown for all ages. Bearded seals were predicted to have a bi-modal
+results are shown for all ages. Bearded seals had a bi-modal
 distribution of haul-out probability across the day (Figure
@@
 that is less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)).
-Adults of both sexes spend considerable time in April, May and June resting on the
-ice. In contrast to ribibon seals, adult spotted seal males have a more protracted
-haul-out season compared to females, and more time spent resting in June and
+Adults of both sexes spend considerable time in April, May and June hauled out.
+In contrast to ribibon seals, adult spotted seal males have a more protracted
+haul-out season compared to females, and more time out of the water in June and
 July (Figures \@ref(fig:ribbonHOCal) and \@ref(fig:spottedHOCal)). As with
-ribbon seals, the young-of-the-year records begin after weaning and
-the model predictions demonstrate the ontogeny of in-water activities (e.g.
-diving, foraging) in May.
+ribbon seals, the young-of-the-year records begin after weaning and the model
+predictions demonstrate the ontogeny of in-water activities (e.g. diving,
+foraging) in May.
 
@@
 to examine factors affecting haul-out behavior on sea ice in the Bering and
-Chukchi seas. Our analysis shows seals haul out progressively more through the
-spring and peak near mid-May to early June before declining again. This pattern
-aligns well with what has been previously documented and confirms our haul-out
-data are likely capturing population-level behavioral patterns. Seals
-preferentially haul out on ice shortly after solar noon, which allows seals to
-maximize absorption of solar radiation thought to facilitate the molting process
-[@feltz1966]. Interestingly, bearded seals appear to have two peaks in haul-out
-activity within a day, one shortly after solar noon, and one centered near solar
-midnight. A similar bi-modal pattern has been seen in ringed seals
-[@vonduyke2020] and suggests that bearded and ringed seals may be operating
-under different constraints than ribbon and spotted seals. Bearded and ringed 
-seals are distributed across higher latitudes and the extended daylight hours
-may allow more flexibility in optimizing resting periods with foraging. Other
-factors such as polar bear predation (which is rare for ribbon and spotted seals
-in the Bering Sea) may also explain differing haul-out patterns. Bearded seals also
-showed a more protracted season of haul-out behavior compared to ribbon and
-spotted seals. This aligns with findings from Thometz et al. [@thometz2021] who
-observed a mean molting period of 119$\pm$2 days and a relatively stable resting
-metabolic rate during that time. In the same study, spotted and ringed seals
+Chukchi seas. Our analysis shows all three species of seal haul out
+progressively more through the spring and peak near mid-May to early June before
+declining again. This pattern aligns well with what has been previously
+documented and confirms our haul-out data are likely capturing population-level
+behavioral patterns. Seals preferentially haul out on ice shortly after solar
+noon, which allows seals to maximize absorption of solar radiation thought to
+facilitate the molting process [@feltz1966]. Interestingly, bearded seals appear
+to have two peaks in haul-out activity within a day, one shortly after solar
+noon, and one centered near solar midnight. A similar bi-modal pattern has been
+seen in ringed seals [@vonduyke2020] and suggests that bearded and ringed seals
+may be operating under different constraints than ribbon and spotted seals.
+Bearded and ringed seals are distributed across higher latitudes and the
+extended daylight hours may allow more flexibility in optimizing resting periods
+with foraging. Other factors such as polar bear predation (which is rare for
+ribbon and spotted seals in the Bering Sea) may also explain differing haul-out
+patterns. Bearded seals also showed a more protracted season of haul-out
+behavior compared to ribbon and spotted seals. This aligns with findings from
+Thometz et al. [@thometz2021] who observed a mean molting period of 119$\pm$2
+days and a relatively stable resting metabolic rate during that time. While
+ribbon seals were not considered in the same study, spotted and ringed seals
 underwent molt periods of 33$\pm$4 and 28$\pm$6 days and exhibited increased
@@
 
-Unlike previous analyses of seal haul-out data (e.g. [@verhoef2010a],
+Unlike previous analyses of seal haul-out behavior (e.g. [@verhoef2010a],
 [@conn2014a]), we also investigated the influence of sex-age class on haul-out
@@
 a suitor male. The male waits for the female to wean the pup and enter estrus,
-and fends off any other males that may be interested. Triad formation results in
-both males and females spending a large portion of the day hauled out on ice and
-a protracted spring haul-out season for both sexes. Because mating occurs in the
-water, females may also be less inclined to venture out for foraging trips while
-still nursing the pup. We see this reflected in the predicted haul-out behavior,
-with both males and females exhibiting a broad distribution of time out of the
-water throughout the solar day and the season.  In the case of both ribbon and
-spotted seals, subadults are the first to begin consistent haul-out behavior and
-follow a typical phocid pattern where subadults molt first as they do not have
-any reproductive constraints.
+and fends off any other potential suitor males. Triad formation results in both
+males and females spending a large portion of the day hauled out on ice and a
+protracted spring haul-out season for both sexes. Because mating occurs in the
+water, females may also be less inclined to interrupt their haul-out with
+foraging trips while still nursing the pup. We see this reflected in the
+predicted haul-out behavior, with both males and females exhibiting a broad
+distribution of time out of the water throughout the solar day and the season.
+In the case of both ribbon and spotted seals, subadults are the first to begin
+consistent haul-out behavior and follow a typical phocid pattern where subadults
+molt first as they do not have any reproductive constraints.
 
@@
 
-We recognize that the species and ecosystems studied in this paper are
-within the traditional range of the Inpuiaq and Yupik people. And, we
-would like to acknowledge and thank them as the original stewards of
-these ecosystems. The deployment of bio-logging devices used in this
-study were often done in collaboration with Alaska Native seal hungers and their communities.
-We would like to especially acknowledge the communities of Kotzebue,
-Utqiaġvik, and Ulguniq (Wainwright) and the following individuals: James Adams, Jeff
-Barger, David Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris,
-Grover Harris, Lee Harris, Tom Jones, Frank Garfield, Brenda Goodwin,
-Henry Goodwin, John Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk,
-Noah Naylor, Virgil Naylor Jr., Virgil Naylor Sr., Dan Savetilik, Allen
-Stone, and Randy Toshavik from Kotzebue, Alaska; Billy Adams, Mary-Ellen
-Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard
-Kittick, Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik,
-Shawn Oktollik, Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross
-Schaeffer, Bob Shears, and Joe Skin from the North Slope Borough,
-Alaska;
+We recognize that the species and ecosystems studied in this paper are within
+the traditional range of the Inpuiaq and Yupik people. And, we would like to
+acknowledge and thank them as the original stewards of these ecosystems. The
+deployment of bio-logging devices used in this study were often done in
+collaboration with Alaska Native seal hungers and their communities. We would
+like to especially acknowledge the communities of Kotzebue, Koyuk, Utqiaġvik,
+and Ulguniq (Wainwright) and the following individuals: James Adams, Jeff
+Barger, David Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris, Grover
+Harris, Lee Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin,
+John Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil
+Naylor Jr., Virgil Naylor Sr., Dan Savetilik, Allen Stone, and Randy Toshavik
+from Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Billy Adams, Mary-Ellen
+Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard Kittick,
+Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik, Shawn Oktollik,
+Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross Schaeffer, Bob Shears, and
+Joe Skin from the North Slope Borough, Alaska;
 
@@
 Goldstein, Jeff Harris, Markus Horning, John Jansen, Shawn Johnson,
-Charles Littnan, Lloyd Lowry, Erin Moreland, Lorrie Rea, Gay Sheffield,
+Charles Littnan, Lloyd Lowry, Erin Moreland, Mark Nelson, Lorrie Rea, Gay Sheffield,
 Brent Stewart, and Dave Withrow. We also appreciate the commitment to
@@
 N00014-16-1-3019). ADFG and NSB field work was covered by Research Permits Nos.
-358-1585, 358-1787, 15324, and 20466 and by ADFG IACUC permits Nos. 06-16,
-09-21, 2010-13, 2012-020, 0027-2019-41, 2013-020, 2014-03,2015-25, 2016-23,
-0027-2017-27, 0027-2018-29, 0027-2019-041.
+358-1585, 358-1787, 15324, and 20466 and by ADF&G IACUC permits Nos. 06-16,
+09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27, 0027-2018-29, 0027-2019-041.
 

### 2021-12-18 analysis/paper/paper.Rmd
IMPROVE FIG 1
feedback re: fig 1 was that the week/day-of-week organization didn't make sense. So, re-org day-of-year into pseudo 10-day weeks and re-adjusted the plot

@@
 bio-loggers deployed on bearded, ribbon, and spotted seals were subset to
-include only records from 1 March to 15 July between `r
-min(lubridate::year(analysis_data$haulout_dt))` and `r
-max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
+include only records from 1 March to 15 July between 
+`r min(lubridate::year(analysis_data$haulout_dt))` and 
+`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
 'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
@@
 ```{r seal-hours}
+create_yday_groups <- function(analysis_data) {
+  d <- analysis_data %>%
+    tibble::as_tibble() %>%
+    dplyr::mutate(yday = lubridate::yday(haulout_dt)) %>%
+    dplyr::select(yday) %>%
+    dplyr::distinct() %>%
+    dplyr::arrange(yday)
+
+  d <- split(d$yday, ceiling(seq_along(d$yday) / 8)) %>%
+    dplyr::as_tibble() %>%
+    tidyr::pivot_longer(
+      cols = 1:18,
+      names_to = "col_group",
+      values_to = "yday",
+      names_transform = list(col_group = as.integer),
+      values_transform = list(yday = as.integer)
+    ) %>%
+    dplyr::arrange(yday) %>% 
+    transform(row_group = letters[1:8]) %>% 
+    dplyr::filter(col_group < 18)
+  return(d)
+}
+
+yday_groups <- create_yday_groups(analysis_data)
+
 seal_hours_tbl <- analysis_data %>% 
   tibble::as_tibble() %>% 
-  dplyr::mutate(wkofyr = lubridate::epiweek(haulout_dt),
-                dayofwk = lubridate::wday(haulout_dt, label = TRUE, abbr = TRUE) %>% 
-                  forcats::fct_relevel("Sun","Mon","Tue","Wed","Thu","Fri","Sat")
-  ) %>% 
-  dplyr::group_by(species, wkofyr, dayofwk) %>% 
+  dplyr::mutate(yday = lubridate::yday(haulout_dt)) %>% 
+  dplyr::left_join(yday_groups) %>% 
+  dplyr::group_by(species, col_group, row_group) %>% 
   dplyr::count()
@@
 
-```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of haul-out behavior data from 1 March to 15 July for each species.
+```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of hourly percent-dry biologger data from 1 March to 15 July for each species.
+Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020.
 The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out.
-Additionally, many bio-logger deployments started in April and May.
-Data were collated across all years between 2005 and 2020."}
+Additionally, many bio-logger deployments started in April and May."}
 
 ggplot(data = seal_hours_tbl) + 
-  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
-  scale_size(range = c(1,4.3), breaks = c(100,500,1000,1500)) +
+  geom_point(aes(x = col_group, y = row_group, color = n, size = n), stroke = 0, shape = 15) +
+  scale_size(range = c(1,3.5), breaks = c(100,500,1000,1500)) +
   scale_color_gradient(low = "#6baed6", high = "#08306b") +
-  scale_x_continuous(breaks = c(13,18,23,27),
-                     labels = c("April", "May", "June", "July")) +
+  scale_x_continuous(breaks = c(1,4,8,12,16),
+                     labels = c("March", "April", "May", "June", "July")) +
   scale_y_discrete(limits = rev) +
@@
   theme(
-        #axis.text.y = element_blank(),
+        axis.text.y = element_blank(),
         panel.grid = element_blank(),
@@
   ggtitle("Density of haul-out behavior observations varies from March through July",
-          subtitle = "data presented in seal-hours collated across years (2005-2020)")
+          subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")
 ```

### 2021-12-19 analysis/paper/paper.Rmd

@@
 spotted seals in the Bering, Chukchi, and Beaufort seas by multiple
-organizations as part of collaborative investigations during 2005 and 2020.
-Seals were captured using nets and instrumented during ship-based capture trips
-and from coastal Alaska communities. Ship-based capture events occurred during
-spring near the southern ice edge in the Bering Sea from 20xx to 20xx (maybe
-provide sample sizes for each species). Land-based capture events occurred from
-May to October, generally between the coastal communities of Scammon Bay,
-Alaska, in the Bering Sea, and Utqiaġvik, Alaska, near the junction of the
-Chukchi and Beaufort seas, from 2005 to 2020. We refer readers to the primary
-literature for detailed capture and bio-logger attachment methods (Supplemental
-Material, Table).
+organizations as part of collaborative investigations from 2005 through 2020.
+Seals were captured using nets and bio-loggers attached during ship-based
+capture trips and from coastal Alaska communities. Ship-based capture events
+occurred during spring near the southern ice edge in the Bering Sea. Land-based
+capture events occurred from May to October, generally between the coastal
+communities of Scammon Bay, Alaska, in the Bering Sea, and Utqiaġvik, Alaska,
+near the junction of the Chukchi and Beaufort seas, from 2005 to 2020. We refer
+readers to the primary literature for detailed capture and bio-logger attachment
+methods (Supplemental Material, Table).
 
@@
 reported by the bio-logger. For this study, in cases where both bio-logger types were
-deployed, hourly percent dry observations from the flipper tag were preferred. Sex and
-age class (non-dependent *young-of-the-year*, sexually immature *subadults*, and
-mature *adults*) were determined at the time of deployment by varying
-combinations of length, claw annuli [@burns1969; @burns1983; @mclaren1958a], and pelage characteristics for some
-species. Field identification of age class can be inexact, particularly when
-discerning subadults from adults. In the case of ribbon seals, pelage
-differences between subadults and adults are distinguishable, with subadults
-often having less defined ribbons. Bearded seal subadults will often have a
-spotted pattern in the pelage that is not seen in adults. Spotted seal pelage is
-monomorphic and can not be used to discern age class. Despite these challenges,
-we feel the age classifications used in this analysis are useful categories in
-testing for age-related effects on haul-out behavior. Seals determined to be
-less than one year were classified as young-of-the-year. Subadults are those
-seals likely greater than one year of age but not yet sexually mature. Adults
-are likely sexually mature and older than four years. For those
-bio-loggers deployed on young-of-the-year and transmitting into the next year,
-the age class was advanced to subadult on 1 March of the following year. Table
+deployed, hourly percent dry observations from the flipper tag were preferred.
+Sex and age class (non-dependent *young-of-the-year*, sexually immature
+*subadults*, and mature *adults*) were determined at the time of deployment by
+varying combinations of length, claw annuli [@burns1969; @burns1983;
+@mclaren1958a], and pelage characteristics for some species. Field
+identification of age class can be inexact, particularly when discerning
+subadults from adults. In the case of ribbon seals, pelage differences between
+subadults and adults are distinguishable, with subadults often having less
+defined ribbons. Bearded seal subadults will often have a spotted pattern in the
+pelage that is not seen in adults. Spotted seal pelage is monomorphic and can
+not be used to discern age class. Despite these challenges, we feel the age
+classifications used in this analysis are useful categories in testing for
+age-related effects on haul-out behavior. Seals determined to be less than one
+year were classified as young-of-the-year. Subadults are those seals likely
+greater than one year of age but not yet sexually mature. Adults are likely
+sexually mature and older than four years. For those bio-loggers deployed on
+young-of-the-year and transmitting into the next year, the age class was
+advanced to subadult on 1 March of the following year. Table
 \@ref(tab:tblDeploy) provides a summary of these deployments and data received
@@
 
-Tags that fall off due to molt or mortality and remain on ice or land can still
-send data to satellites; i.e., the bio-logger will be on ice or land and dry, therefore, it will record and transmit data suggesting the seal is hauled out.
-As such, end times of each deployment were identified by examining
-bio-logger locations and dive behavior to determine when bio-loggers ceased providing data
-consistent with seal behavior. Data outside of the deployment start and end
-times were discarded prior to analysis. After approximately 9 months, 7 devices
-deployed on the rear flipper of bearded seals reported implausible hourly percent-dry data.
-All data after the first instance of unrealistic values were censored from this
-analysis. Figure \@ref(fig:dataCal) shows a calendar-like distribution of all
-haul-out data across the study season for each species. Observations for ribbon
-and spotted seals are concentrated in the months of May and June due to the
-timing of deployment (April and May) and the timing of molt (May and June).
-During molt, seals (and their attached bio-loggers) spend more time out of the
-water and more data are transmitted. Molt timing also impacts when many
-deployments end as any bio-loggers adhered to the hair will fall off. Bearded seal
-deployments are more evenly spread with fewer observations in May and June. The
-majority of bearded seal deployments start later in the summer and by May bio-loggers
-have either fallen off or the batteries are depleted.
+Tags that fall off due to molt, attachment failure, or seal mortality and remain
+on ice or land can still send data to satellites; i.e., the bio-logger will be
+on ice or land and dry, therefore, it will record and transmit data suggesting
+the seal is hauled out. As such, end times of each deployment were identified by
+examining bio-logger locations and dive behavior to determine when bio-loggers
+ceased providing data consistent with seal behavior. Data outside of the
+deployment start and end times were discarded prior to analysis. After
+approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
+reported implausible hourly percent-dry data. All data after the first instance
+of unrealistic values were censored from this analysis. Figure
+\@ref(fig:dataCal) shows a calendar-like distribution of all haul-out data
+across the study season for each species. Observations for ribbon and spotted
+seals are concentrated in the months of May and June due to the timing of
+deployment (April and May) and the timing of molt (May and June). During molt,
+seals (and their attached bio-loggers) spend more time out of the water and more
+data are transmitted. Molt timing also impacts when many deployments end as any
+bio-loggers adhered to the hair will fall off. Bearded seal deployments are more
+evenly spread with fewer observations in May and June. The majority of bearded
+seal deployments start later in the summer and by May bio-loggers have either
+fallen off or the batteries are depleted.
 
@@
   geom_point(aes(x = col_group, y = row_group, color = n, size = n), stroke = 0, shape = 15) +
-  scale_size(range = c(1,3.5), breaks = c(100,500,1000,1500)) +
+  scale_size(range = c(1,3.4), breaks = c(100,500,1000,1500)) +
   scale_color_gradient(low = "#6baed6", high = "#08306b") +
@@
         axis.text.y = element_blank(),
+        axis.text.x = element_text(hjust = 0),
         panel.grid = element_blank(),
@@
 sensor was polled by the tag firmware every few seconds and a percent of the
-hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). Hourly
-percent-dry data were usually rounded to the nearest 10 percent inclusive of 0
-and 100 percent and additional values at 3 and 98 percent. This compression
-resulted in additional data transmission as each message consisted of two
-complete 24-hour records. Bio-loggers were, generally, deployed and programmed
-in a manner to maximize data transmission during the spring pupping and molting
-period, however, hourly percent-dry data for days during a deployment were not
-successfully transmitted. This is due to a variety of factors including
-satellite coverage, tag availability (i.e. tags mounted to the rear flipper are
-not available to the surface while the seal is at sea), tag performance, duty
-cycling, and extra-terrestrial atmospheric interference.
+hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). On board the
+bio-logger, hourly percent-dry data were rounded to the nearest 10 percent
+inclusive of 0 and 100 percent and additional values at 3 and 98 percent. This
+compression resulted in additional data transmission as each message consisted
+of two complete 24-hour records. Bio-loggers were, generally, deployed and
+programmed in a manner to maximize data transmission during the spring pupping
+and molting period, however, hourly percent-dry data for days during a
+deployment were not successfully transmitted. This is due to a variety of
+factors including satellite coverage, tag availability (i.e. tags mounted to the
+rear flipper are not available to the surface while the seal is at sea), tag
+performance, duty cycling, and extra-terrestrial atmospheric interference.
 
@@
   geom_point(aes(x = day, y = hour), data = missing_hours, 
-             color = "black", shape = 4, size=1)
+             color = "black", shape = 4, size=0.8)
 p
@@
 Of key interest in this study is the relationship between haul-out behavior and
-environmental covariates that vary with space and time. We explored the use of a
+weather covariates that vary with space and time. We explored the use of a
 continuous-time correlated random walk [@devins.johnson2008] movement model to
@@
 that bearded and spotted seals haul out on land. However, assessing the
-relationship between haul-out behavior and environmental covariates and their
+relationship between haul-out behavior and weather covariates and their
 availability for aerial surveys on land was outside the scope of this study.
@@
 using the {solaR} package [@perpinan2012] within the R statistical environment
-[@rcoreteam2021] using the daily locations. We included several meteorological
+[@rcoreteam2021] using the daily locations. We included several weather
 variables shown to affect haul-out behavior in other Arctic pinnipeds
@@
   "northerly and easterly vector components for wind converted into a single wind speed via the Euclidean norm",
-  "Northing", "Continuous", "Bio-logger","latitude divided by the mean latitude across all locations (for bearded seals only)"
+  "Northing", "Continuous", "Bio-logger","latitude divided by the mean latitude across all locations (for bearded seals only)",
+  "Year", "Continuous", "Bio-logger", "For the set of models examining inter-annual variation in sea ice use, we fitted models with the addition of year by day-of-year interactions." 
 )
@@
   fontsize(size = 10) %>% 
+  hline(i=8) %>% 
   set_table_properties(width = .95, layout = "autofit")
@@
 
-Bearded seal haul-out probability appears most influenced by wind
+When exploring the influence of weather, Bearded seal haul-out probability appears most influenced by wind
 ($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$
@@
 `r bearded_type3$temp2_wind$Prob.F`). Any influence of barometric
-pressure or precipitation is less apparent.
+pressure or precipitation was less apparent.
 
@@
 Adults of both sexes spend considerable time in April, May and June hauled out.
-In contrast to ribibon seals, adult spotted seal males have a more protracted
+In contrast to ribbon seals, adult spotted seal males have a more protracted
 haul-out season compared to females, and more time out of the water in June and
@@
 behavior compared to ribbon and spotted seals. This aligns with findings from
-Thometz et al. [@thometz2021] who observed a mean molting period of 119$\pm$2
+Thometz et al. [@thometz2021] who observed a mean molting period of 119&pm;2
 days and a relatively stable resting metabolic rate during that time. While
 ribbon seals were not considered in the same study, spotted and ringed seals
-underwent molt periods of 33$\pm$4 and 28$\pm$6 days and exhibited increased
+underwent molt periods of 33&pm;4 and 28&pm;6 days and exhibited increased
 resting metabolic rates.
@@
 
-We also investigated the influence of environmental covariates on haul-out
+We also investigated the influence of weather on haul-out
 probabilities, including wind speed, temperature, barometric pressure,
@@
 These weather conditions are likely most beneficial for energetic considerations
-and may also provide the best situation for predator detection. These results
+and may also provide an optimal situation for predator detection. These results
 highlight the importance for haul-out behavior models (and subsequent

### 2021-12-19 README.Rmd
UPDATE README

@@
 <!-- badges: start -->
-[![Launch Rstudio Binder](http://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/jmlondon/berchukHaulout/main?urlpath=rstudio)
 [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.4638221.svg)](https://doi.org/10.5281/zenodo.4638221)
@@
 ## Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas
-Josh M. London^1^,✉, Paul B. Conn^1^, Stacie K. Hardy^1^, Erin L. Richmond^1^, Jay M. Ver Hoef^1^, Irina S. Trukhanova^1^, Michael F. Cameron^1^, and Peter L. Boveng^1^
-1 Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
+Josh M.
+London^1^,✉, Paul B.
+Conn^1^, Stacie K.
+Hardy^1^, Erin L.
+Richmond^1^, Jay M.
+Ver Hoef^1^, Michael F.
+Cameron^1^, Justin Crawford^2^, Andrew L.
+Von Duyke^3^, Lori T.
+Quakenbush^2^, and Peter L.
+Boveng^1^
+
+1. Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
+2. Arctic Marine Mammals Program, Alaska Department of Fish and Game, Fairbanks, Alaska, USA
+3. Department of Wildlife Management, North Slope Borough, Utqiaġvik, Alaska, USA
+
 ✉ Correspondence: Josh M. London <josh.london@noaa.gov>
@@
     generated by the analysis. It also has a rendered version,
-    `paper.docx`, suitable for reading (the code is replaced by figures
+    `London_HauloutBehavior.docx`, suitable for reading (the code is replaced by figures
     and tables in this file)
@@
 
-  - [:file\_folder: data](/data): Data used or generated in analysis
+  - [:file\_folder: data](/data): data used for model fits and final model output
 
-## How to run in your broswer or download and run locally
+### {targets} and {renv} packages
 
-This research compendium has been developed using the statistical programming
-language R. To work with the compendium, you will need
-installed on your computer the [R software](https://cloud.r-project.org/)
-itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).
+This repository relies heavily on the {targets} and {renv} packages for management
+of analysis pipelines and reproducibility.
 
-The simplest way to explore the text, code and data is to click on
-[binder](https://mybinder.org/v2/gh/jmlondon/berchukHaulout/main?urlpath=rstudio)
-to open an instance of RStudio in your browser, which will have the
-compendium files ready to work with. Binder uses rocker-project.org
-Docker images to ensure a consistent and reproducible computational
-environment. These Docker images can also be used locally.
+  - [:file\_folder: R](/R): pipeline functions developed in support or target pipeline
+  - [:page\_facing\_up: _targets.R](/_targets.R): target pipeline
+  - [:page\_facing\_up: renv.lock](/renv.lock): renv lockfile
 

### 2021-12-21 analysis/paper/London_HauloutBeahvior.Rmd
update docx template

--- 2021-12-19 analysis/paper/paper.Rmd
@@
       - --lua-filter=../templates/author-info-blocks.lua
+    page_size:
+      width: 8.5
+      height: 11
+      orient: "portrait"
+    page_margins:
+      bottom: 1.25
+      top: 1.25
+      right: 1.0
+      left: 0.5
+      header: 0.5
+      footer: 0.5
+      gutter: 0.5
 bibliography: references.json

### 2021-12-21 analysis/paper/London_HauloutBeahvior.Rmd
minor updates before Andy comments

@@
 patterns uncovered significant contributions for linear and quadratic
-interactions between day and year in both ribbon seals (day $\times$
-year,
+interactions between day and year in both ribbon seals (day:year,
 $F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$
@@
 = `r ribbon_year_type3$day2_year$F.value`; $p$ =
-`r ribbon_year_type3$day2_year$Prob.F`) and spotted seals (day $\times$
-year,
+`r ribbon_year_type3$day2_year$Prob.F`) and spotted seals (day:year,
 $F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$
@@
 
-**Irina S. Trukhanova**: conceptualization, writing: review and editing
-
 **Michael F. Cameron**: investigation, project administration, writing:
@@
 
+**Justin Crawford**: investigation, methodology, validation, data curation, 
+writing: review and editing
+
+**Lori T. Quakenbush**: investigation, methodology, supervision, project
+administration, writing: review and editing
+
+**Andrew L. Von Duyke**: investigation, methodology, validation, data curation,
+writing: review and editing
+
 **Peter L. Boveng**: investigation, conceptualization, supervision,
@@
 
-Data and code will be available as a complete research compendium on
-GitHub and major versions archived at Zenodo. Original data sources for
-telemetry are archived at the United States Animal Telemetry Network.
+This manuscript was developed as a reproducible research compendium. Data and
+code are available on GitHub (https://github.com/jmlondon/berchukHaulout) and
+major versions archived at Zenodo (https://doi.org/10.5281/zenodo.4638221).
+Original data sources for telemetry are archived at the United States Animal
+Telemetry Network.
 

### 2021-12-22 analysis/paper/London_HauloutBeahvior.Rmd
Andy VD comments incorporated

@@
 between 2005 and 2020 to investigate the seasonal timing and environmental
-factors affecting haul-out behavior by ice-associated seals (specifically, bearded,
-ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to
-providing baseline data on phenology, these data also allow us to quantify
-'availability', which is needed to accurately estimate abundance from aerial
-survey counts of seals basking on ice (i.e., to correct for the proportion of
-animals that are in the water while surveys are conducted). Using generalized
-linear mixed pseudo-models to properly account for temporal autocorrelation, we
-fit models with covariates of interest (e.g., day-of-year, solar hour, age-sex
-class, wind speed, barometric pressure, temperature, precipitation) to examine
-their ability to explain variation in hourly haul-out probability. We found evidence
-for strong diurnal and within-season patterns in haul-out behavior, as well as
+factors affecting haul-out behavior by ice-associated seals. We specifically
+focused on bearded (_Erignathus barbatus_), ribbon (_Histriophoca fasciata_),
+and spotted seals (_Phoca largha_)) in the Bering and Chukchi seas. Because
+ringed seals (_Phoca hispida_) are unique in their use of snow lairs, they were
+not included and a separate analysis is warranted. In addition to providing
+baseline data on phenology, these data also allow us to quantify 'availability',
+which is needed to accurately estimate abundance from aerial survey counts of
+these seals basking on ice (i.e., to correct for the proportion of animals that
+are in the water while surveys are conducted). Using generalized linear mixed
+pseudo-models to properly account for temporal autocorrelation, we fit models
+with covariates of interest (e.g., day-of-year, solar hour, age-sex class, wind
+speed, barometric pressure, temperature, precipitation) to examine their ability
+to explain variation in hourly haul-out probability. We found evidence for
+strong diurnal and within-season patterns in haul-out behavior, as well as
 strong weather effects (particularly wind and temperature). In general, seals
@@
 for seasonal and temporal variation in haul-out behavior, as well as associated
-environmental covariates, when interpreting the number of seals counted
-in aerial surveys.
+environmental covariates, when interpreting the number of seals counted in
+aerial surveys.
 
@@
 Arctic organisms, ecosystems, and the human communities who live in the region
-[@huntington2020]. Such disruptions are a particular cause of concern for
-ice-associated seals (ringed (*Pusa hispida*), bearded (*Erignathus barbatus*),
-ribbon seals (*Histriophoca fasciata*), and spotted (*Phoca largha*)) that
-depend on spring and early summer sea ice (March-June) in the Bering and Chukchi
-seas as a platform for important functions, such as pupping, nursing, breeding,
-and molting [@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data
-and large knowledge gaps complicate predictions about the ultimate effects of
-changes in sea ice on the behavior, health, abundance, and distribution of these
-seals. To date, indices of seal health sampled during periods of declining sea
-ice differ regionally [@crawford2015; @harwood2020]. Knowledge about
-evolutionary constraints on the timing of reproductive and molting behavior is
-generally lacking, so it is difficult to predict the readiness with which
-ice-associated seal species will be able to adapt to future changes (e.g., by
-adjusting pupping or molting schedules to earlier dates or different locales).
-Additionally, trends in abundance of these species are largely unknown, so it is
-difficult to assess the effect, if any, declines in sea ice habitat have had, or
-will have, on seal densities.
+[@huntington2020]. Such disruptions are a particular cause of concern for the
+ice-associated seals that depend on spring and early summer sea ice (March-June)
+in the Bering and Chukchi seas as a platform for important life history
+functions, such as pupping, nursing, breeding, and molting [@boveng2009;
+@boveng2013a; @cameron2010; @kelly2010]. Limited data and large knowledge gaps
+complicate predictions about the ultimate effects of changes in sea ice on the
+behavior, health, abundance, and distribution of these seals. To date, indices
+of seal health sampled during periods of declining sea ice differ regionally
+[@crawford2015; @harwood2020]. Knowledge about evolutionary constraints on the
+timing of reproductive and molting behavior is generally lacking, so it is
+difficult to predict the readiness with which ice-associated seal species will
+be able to adapt to future changes (e.g., by adjusting pupping or molting
+schedules to earlier dates or different locales). This is further complicated by
+the spatio-temporal variation in the phenology of these life history events 
+within regions and throughout their full ranges. Additionally, trends in
+abundance from large scale surveys of these species are unknown, so it is difficult to assess the
+effect, if any, declines in sea ice habitat have had, or will have, on seal
+densities.
 
 For ice-associated seals, haul-out behavior is tightly linked with sea ice.
-Ribbon seals haul out of the water almost exclusively on sea ice and are mostly
-pelagic outside the spring pupping, breeding, and molting season [@boveng2018a].
-While spotted and bearded seals rest on coastal features, they
-strongly prefer sea ice as a resting platform during the spring and early summer
-[@frost2018]. The remoteness of the Bering and Chukchi seas mean direct
-scientific observation of seal behavior is impractical. Thus, bio-logging
-devices are, thus, especially useful tools for collecting key information on
-movement and haul-out behavior for these species.
+Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
+on sea ice and are mostly pelagic outside the spring pupping, breeding, and
+molting season [@boveng2018a]. While spotted (_Phoca largha_) and bearded
+(_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
+ice as a resting platform during the spring and early summer [@frost2018].
+Ringed seals (_Phoca hispida_) --- not included in this study --- bask on
+sea-ice but also within snow lairs built during the spring. The remoteness of
+the Bering and Chukchi seas mean direct scientific observation of seal behavior
+is impractical. Thus, bio-logging devices are, thus, especially useful tools for
+collecting key information on movement and haul-out behavior for these species.
 
@@
 environmental covariates, such as the onset of snow melt and snow depth. Our
-goals were threefold. First, we wish to establish baseline estimates for the
+goals were threefold. First, we wished to establish baseline estimates for the
 chronology of haul-out behavior in the critical spring season for each species
@@
 satellite transmittal rates, locations at sea) but these are lost during the
-annual molt, which limits the duration of haul-out data they provide.
-Additionally, bio-loggers attached to the head or dorsal region are often dry
-while the seal is at the surface and can slightly bias the hourly percent-dry values
-reported by the bio-logger. For this study, in cases where both bio-logger types were
-deployed, hourly percent dry observations from the flipper tag were preferred.
-Sex and age class (non-dependent *young-of-the-year*, sexually immature
-*subadults*, and mature *adults*) were determined at the time of deployment by
-varying combinations of length, claw annuli [@burns1969; @burns1983;
-@mclaren1958a], and pelage characteristics for some species. Field
-identification of age class can be inexact, particularly when discerning
-subadults from adults. In the case of ribbon seals, pelage differences between
-subadults and adults are distinguishable, with subadults often having less
-defined ribbons. Bearded seal subadults will often have a spotted pattern in the
-pelage that is not seen in adults. Spotted seal pelage is monomorphic and can
-not be used to discern age class. Despite these challenges, we feel the age
-classifications used in this analysis are useful categories in testing for
-age-related effects on haul-out behavior. Seals determined to be less than one
-year were classified as young-of-the-year. Subadults are those seals likely
-greater than one year of age but not yet sexually mature. Adults are likely
-sexually mature and older than four years. For those bio-loggers deployed on
-young-of-the-year and transmitting into the next year, the age class was
-advanced to subadult on 1 March of the following year. Table
-\@ref(tab:tblDeploy) provides a summary of these deployments and data received
-from them.
+following annual molt, which, depending on deployment date, limits the duration
+of haul-out data they provide. Additionally, bio-loggers attached to the head or
+dorsal region are often dry while the seal is at the surface and can slightly
+bias the hourly percent-dry values reported by the bio-logger. For this study,
+in cases where both bio-logger types were deployed, hourly percent dry
+observations from the flipper tag were preferred.
+
+Field identification of age class can be inexact, particularly when discerning
+subadults from adults. Sex and age class (non-dependent _young-of-the-year_,
+sexually immature _subadults_, and mature _adults_) were estimated at the time
+of deployment by various combinations of length, claw growth ridges [@burns1969;
+@burns1983; @mclaren1958a], and pelage characteristics for some species. In the
+case of ribbon seals, subadults often have less distinct ribbons than adults.
+Bearded seal subadults will often have a spotted pattern in the pelage that is
+not seen in adults. Spotted seal pelage cannot be used to reliably discern
+adults from subadults. Despite these challenges, we feel the age classifications
+used in this analysis are useful in testing for age-related effects on haul-out
+behavior. Seals determined to be less than one year were classified as
+young-of-the-year. Subadults are those seals likely greater than one year of age
+but not yet sexually mature. Adults are likely sexually mature and older than
+four years. For those bio-loggers deployed on young-of-the-year and transmitting
+into the next year, the age class was advanced to subadult on 1 March of the
+following year. Table \@ref(tab:tblDeploy) provides a summary of these
+deployments and data received from them.
 
@@
 of unrealistic values were censored from this analysis. Figure
-\@ref(fig:dataCal) shows a calendar-like distribution of all haul-out data
+\@ref(fig:dataCal) shows the distribution of all haul-out data
 across the study season for each species. Observations for ribbon and spotted
@@
 
-```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of hourly percent-dry biologger data from 1 March to 15 July for each species.
+```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of hourly percent-dry biologger data from 1 March (upper left) to 15 July (lower right) for each species.
 Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020.
@@
   scale_x_continuous(breaks = c(1,4,8,12,16),
-                     labels = c("March", "April", "May", "June", "July")) +
-  scale_y_discrete(limits = rev) +
+                     labels = c("March", "April", "May", "June", "July"),
+                     expand = expansion(mult = c(0.01, 0.025))) +
   guides(color = "none",
@@
         axis.text.x = element_text(hjust = 0),
+        axis.title = element_blank(),
         panel.grid = element_blank(),
         legend.position = "bottom") +
-  labs(x = "", y = "", color = "seal-hours",
+  labs(color = "seal-hours",
        size = "seal hours of bio-logger data") +
@@
 
-Of key interest in this study is the relationship between haul-out behavior and
+Of key interest in this study was the relationship between haul-out behavior and
 weather covariates that vary with space and time. We explored the use of a
@@
 to have two peaks in haul-out activity within a day, one shortly after solar
-noon, and one centered near solar midnight. A similar bi-modal pattern has been
-seen in ringed seals [@vonduyke2020] and suggests that bearded and ringed seals
-may be operating under different constraints than ribbon and spotted seals.
-Bearded and ringed seals are distributed across higher latitudes and the
-extended daylight hours may allow more flexibility in optimizing resting periods
-with foraging. Other factors such as polar bear predation (which is rare for
-ribbon and spotted seals in the Bering Sea) may also explain differing haul-out
-patterns. Bearded seals also showed a more protracted season of haul-out
-behavior compared to ribbon and spotted seals. This aligns with findings from
-Thometz et al. [@thometz2021] who observed a mean molting period of 119&pm;2
-days and a relatively stable resting metabolic rate during that time. While
-ribbon seals were not considered in the same study, spotted and ringed seals
-underwent molt periods of 33&pm;4 and 28&pm;6 days and exhibited increased
-resting metabolic rates.
+noon, and one centered near solar midnight. This, of course, could be an
+artifact of our limited sample size for bearded seal deployments across all age
+classes. However, a similar bi-modal pattern has been seen in ringed seals
+[@vonduyke2020] and suggests that bearded and ringed seals may be operating
+under different constraints than ribbon and spotted seals. Bearded and ringed
+seals are distributed across higher latitudes and the extended daylight hours
+may allow more flexibility in optimizing resting periods with foraging. Other
+factors such as predation by polar bears (which is rare for ribbon and spotted
+seals in the Bering Sea) may also explain differing haul-out patterns. Bearded
+seals also showed a more protracted season of haul-out behavior compared to
+ribbon and spotted seals. This aligns with findings from Thometz et al.
+[@thometz2021] who observed a mean molting period of 119&pm;2 days and a
+relatively stable resting metabolic rate during that time. While ribbon seals
+were not considered in the same study, spotted and ringed seals underwent molt
+periods of 33&pm;4 and 28&pm;6 days and exhibited increased resting metabolic
+rates.
 
@@
 hypothesis that ribbon seal females continue foraging while nursing. Spotted
-seals are known to form triads during the breeding season [@burns2009,
-@mcclintock2015] where a female and dependent pup are accompanied on the ice by
-a suitor male. The male waits for the female to wean the pup and enter estrus,
-and fends off any other potential suitor males. Triad formation results in both
-males and females spending a large portion of the day hauled out on ice and a
-protracted spring haul-out season for both sexes. Because mating occurs in the
-water, females may also be less inclined to interrupt their haul-out with
-foraging trips while still nursing the pup. We see this reflected in the
-predicted haul-out behavior, with both males and females exhibiting a broad
-distribution of time out of the water throughout the solar day and the season.
-In the case of both ribbon and spotted seals, subadults are the first to begin
-consistent haul-out behavior and follow a typical phocid pattern where subadults
-molt first as they do not have any reproductive constraints.
-
-We also investigated the influence of weather on haul-out
-probabilities, including wind speed, temperature, barometric pressure,
-precipitation, and wind chill. These have been investigated for walruses (e.g.
-Udevitz et al. [-@udevitz2009]) and a few select studies of ice-associated seals
+seals are known to form triads during the breeding season [@burns1973] where a
+female and dependent pup are accompanied on the ice by a suitor male. The male
+waits for the female to wean the pup and enter estrus, and fends off any other
+potential suitor males. Triad formation results in both males and females
+spending a large portion of the day hauled out on ice and a protracted spring
+haul-out season for both sexes. Because mating occurs in the water, females may
+also be less inclined to interrupt their haul-out with foraging trips while
+still nursing the pup. We see this reflected in the predicted haul-out behavior,
+with both males and females exhibiting a broad distribution of time out of the
+water throughout the solar day and the season. In the case of both ribbon and
+spotted seals, subadults are the first to begin consistent haul-out behavior and
+follow a typical phocid pattern where subadults molt first as they do not have
+any reproductive constraints.
+
+We also investigated the influence of weather on haul-out probabilities,
+including wind speed, temperature, barometric pressure, precipitation, and wind
+chill. These have been investigated for walruses (e.g. Udevitz et al.
+[-@udevitz2009]) and a few select studies of ice-associated seals
 [@hamilton2018, @perry2017]. Ribbon seals seemed to be the most influenced by
@@
 speeds were lower, barometric pressure was higher, and precipitation was lower.
-These weather conditions are likely most beneficial for energetic considerations
-and may also provide an optimal situation for predator detection. These results
-highlight the importance for haul-out behavior models (and subsequent
-calculations of availability during surveys) to incorporate weather covariates.
+These weather conditions are general indicators of increased solar radiation
+which provides energetic benefits. Low winds and precipitation may also
+provide an optimal situation for predator detection. Our results highlight the
+importance for haul-out behavior models (and subsequent calculations of
+availability during surveys) to incorporate weather covariates.
 

### 2021-12-22 README.Rmd
UPDATE README
trying to include the targets dependency graph

@@
 output: github_document
+always_allow_html: true
 ---
@@
 )
-# Please put your title here to include it in the file below.
-Title <- "Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas"
 ```
@@
 
-  - [:file\_folder: R](/R): pipeline functions developed in support or target pipeline
+  - [:file\_folder: R](/R): pipeline functions developed in support of target pipeline
   - [:page\_facing\_up: _targets.R](/_targets.R): target pipeline
   - [:page\_facing\_up: renv.lock](/renv.lock): renv lockfile
+
+The dependency graph for our pipeline is shown below
+
+```{r, echo=FALSE}
+library(targets)
+targets::tar_glimpse()
+```
 

### 2021-12-22 analysis/scripts/create_avg_wx.R
DELETE analysis/scripts FOLDER
all of this functionality is now w/in the R directory and the targets pipeline

@@ -1,53 +1,0 @@
-library(tidyverse)
-library(DBI)
-library(dbplyr)
-library(sf)
-
-con <- dbConnect(
-  odbc::odbc(),
-  dsn = "PostgreSQL pep",
-  uid = keyringr::get_kc_account("pgpep_londonj"),
-  pwd = keyringr::decrypt_kc_pw("pgpep_londonj"),
-  sslmode = "disable"
-)
-
-load(here::here('data/dat_sf.Rdata'))
-
-grid.sf <- sf::st_read(con,
-                       query = "SELECT * FROM base.geo_analysis_grid") %>%
-  sf::st_crop(st_bbox(dat.sf) %>% st_as_sfc())
-
-cov_means <- tbl(con, in_schema("base","tbl_analysis_grid_cov_wx")) %>%
-  filter(cell %in% !!grid.sf$cell) %>%
-  filter(month(fdatetime_range_start) %in% c(3,4,5,6,7),
-         year(fdatetime_range_start) != '2004') %>%
-  group_by(year = year(fdatetime_range_start),
-           yday = yday(fdatetime_range_start)) %>%
-  summarise(precip = mean(rast_acpcp, na.rm=TRUE),
-            temp2 = (mean(rast_air2m, na.rm=TRUE)-270),
-            temp0 = (mean(rast_airsfc, na.rm=TRUE)-270),
-            pressure = (mean(rast_prmsl, na.rm=TRUE)-100000),
-            wind = sqrt(mean(rast_uwnd, na.rm=TRUE)^2 + mean(rast_vwnd, na.rm=TRUE)^2)
-            )
-
-cov_means_daily_2012 <- tbl(con, in_schema("base","tbl_analysis_grid_cov_wx")) %>%
-  filter(cell %in% !!grid.sf$cell) %>%
-  filter(month(fdatetime_range_start) %in% c(3,4,5,6,7),
-         year(fdatetime_range_start) == '2012') %>%
-  group_by(year = year(fdatetime_range_start),
-           yday = yday(fdatetime_range_start),
-           cell) %>%
-  summarise(precip = mean(rast_acpcp, na.rm=TRUE),
-            temp2 = (mean(rast_air2m, na.rm=TRUE)-270),
-            temp_sfc = (mean(rast_airsfc, na.rm=TRUE)-270),
-            pressure = (mean(rast_prmsl, na.rm=TRUE)-100000),
[…]

### 2021-12-22 analysis/scripts/full_deployment_table.R

@@ -1,31 +1,0 @@
-View(dat.sf)
-
-library(DBI)
-library(odbc)
-library(pingr)
-stopifnot(
-"PEP Postgres Database Not Available; did you start VPN? ;)" =
-pingr::is_up("161.55.120.122", "5432")
-)
-
-con <- dbConnect(
-  odbc::odbc(),
-  dsn = "PostgreSQL pep",
-  uid = keyringr::get_kc_account("pgpep_londonj"),
-  pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
-)
-
-qry <- "SELECT d.speno, d.serial_num, d.ptt, d.field_effort_lku, effort_lead, meta_project
-FROM telem.tbl_tag_deployments d
-LEFT JOIN capture.tbl_event USING (speno)
-ORDER BY speno"
-
-res <- con %>% tbl(dplyr::sql(qry)) %>%
-  collect()
-
-dat.sf %>% as_tibble() %>%
-  group_by(speno,species,age,sex,deploy_dt) %>% count() %>%
-  left_join(res, by="speno") %>%
-  arrange(deploy_dt) %>%
-  write_csv("haulout_deployments.csv")
-

### 2021-12-22 analysis/scripts/get_data/Haulout_LoadAndProcessCov_20180926_SKH.R

@@ -1,30 +1,0 @@
-#library(icesealhaulout)
-library(RPostgreSQL)
-library(tidyverse)
-
-con <- RPostgreSQL::dbConnect(PostgreSQL(),
-                              dbname = Sys.getenv("pep_db"),
-                              host = Sys.getenv("pep_ip"),
-                              user = Sys.getenv("pep_admin"),
-                              rstudioapi::askForPassword(paste("Enter your DB password for user account: ", Sys.getenv("pep_admin"), sep = "")))
-
-#data("percent_locs")
-# load("//AKC0SS-N086/NMML_Users/Stacie.Hardy/Desktop/icesealhaulout/data/percent_locs.rda")
-# percent_locs %>% dplyr::ungroup()
-# percent_locs$fill_xy <- as.character(percent_locs$fill_xy)
-# percent_locs$id <- 1:nrow(percent_locs)
-# percent_locs <- percent_locs[, c(10, 1, 5, 6, 7, 3, 4, 8, 9)]
-# colnames(percent_locs)[[6]] <- "haulout_dt"
-# RPostgreSQL::dbSendQuery(con, "DROP TABLE IF EXISTS telem.res_iceseal_haulout")
-# sf::st_write_db(con, percent_locs, c("telem", "res_iceseal_haulout"), row.names = FALSE)
-# RPostgreSQL::dbSendQuery(con, "ALTER TABLE telem.res_iceseal_haulout RENAME COLUMN geometry TO geom")
-
-# Create index on haulout data
-dbSendQuery(con, "SELECT telem.fxn_iceseal_pred_idx();")
-RPostgreSQL::dbClearResult(RPostgreSQL::dbListResults(con)[[1]])
-
-# Create covariate tables
-RPostgreSQL::dbSendQuery(con, "SELECT telem.fxn_iceseal_haulout_cov();")
-RPostgreSQL::dbClearResult(RPostgreSQL::dbListResults(con)[[1]])
-
-RPostgreSQL::dbDisconnect(con)

### 2021-12-22 analysis/scripts/get_data/make_data.R

@@ -1,10 +1,0 @@
-library(here)
-source(here('R/get_locs_sf.R'))
-source(here('R/get_timeline_data.R'))
-source(here('R/create_db_input_data.R'))
-
-timeline_data = get_timeline_data()
-locs_sf = get_locs_sf()
-db_input_data = create_db_input_data(timeline_data, locs_sf)
-
-save(db_input_data, file = here::here("data/tbl_percent_locs.Rdata"))

### 2021-12-22 analysis/scripts/model.R

@@ -1,178 +1,0 @@
-library(tidyverse)
-library(here)
-library(lubridate)  #some date time functionality
-library(glmmLDTS)
-library(solaR)  #for getting hour of day in solar time
-library(splines)
-library(sf)
-library(DBI)
-library(rnaturalearth)
-library(purler)
-
-create_ar1_id <- function(h_dt) {
-  ar1_id <- tibble(
-    t_seq = seq(from = h_dt[1], to = h_dt[length(h_dt)], by = "hour",)
-  ) %>%
-    left_join(tibble(h_dt = h_dt, haulout_dt = h_dt), by = c("t_seq" = "h_dt")) %>%
-    mutate(x = is.na(haulout_dt),
-           rle = rlenc_id(x)) %>%
-    filter(!is.na(haulout_dt)) %>%
-    pull(rle)
-
-  return(ar1_id)
-}
-
-load(here::here('data/dat_sf.Rdata'))
-
-dat.sf <- dat.sf  %>%
-  mutate(dry = round(percent_dry/100)) %>%
-  mutate(hour_utc = lubridate::hour(haulout_dt),
-         yday = lubridate::yday(haulout_dt),
-         year = lubridate::year(haulout_dt)) %>%
-  mutate(coords_x = st_coordinates(st_transform(.,4326))[,1],
-         coords_y = st_coordinates(st_transform(.,4326))[,2],
-         solar_hour = solaR::local2Solar(haulout_dt,coords_x) %>% lubridate::hour(),
-         sin1 = sin(pi*solar_hour/12),
-         cos1 = cos(pi*solar_hour/12),
-         sin2 = sin(pi*solar_hour/6),
-         cos2 = cos(pi*solar_hour/6),
-         sin3 = sin(pi*solar_hour/4),
-         cos3 = sin(pi*solar_hour/4),
-         day = (yday-120)/10,
-         day2 = day^2,
-         day3 = day^3
-         ) %>%
-  mutate(precip = rast_acpcp,
-         temp2 = (rast_air2m-270)/27,
-         temp0 = (rast_airsfc-270)/27,
-         pressure = (rast_prmsl-100000)/10000,
-         wind = sqrt(rast_uwnd^2 + rast_vwnd^2)/10
-         ) %>%
-  mutate(northing = st_coordinates(.)[,2]/mean(st_coordinates(.)[,2]),
-         north_of_bering = case_when(
-           coords_y > 66.5 ~ TRUE,
-           TRUE ~ FALSE
[…]

### 2021-12-22 analysis/scripts/model_year.R

@@ -1,106 +1,0 @@
-library(tidyverse)
-library(here)
-library(lubridate)  #some date time functionality
-library(glmmLDTS)
-library(solaR)  #for getting hour of day in solar time
-library(splines)
-library(sf)
-library(DBI)
-library(rnaturalearth)
-library(purler)
-
-create_ar1_id <- function(h_dt) {
-  ar1_id <- tibble(
-    t_seq = seq(from = h_dt[1], to = h_dt[length(h_dt)], by = "hour",)
-  ) %>% left_join(tibble(h_dt = h_dt, haulout_dt = h_dt), by = c("t_seq" = "h_dt")) %>%
-    mutate(x = is.na(haulout_dt),
-           rle = rlenc_id(x)) %>%
-    filter(!is.na(haulout_dt)) %>%
-    pull(rle)
-
-  return(ar1_id)
-}
-
-load(here::here('data/dat_sf.Rdata'))
-
-dat.sf <- dat.sf  %>%
-  mutate(dry = round(percent_dry/100)) %>%
-  mutate(hour_utc = hour(haulout_dt),
-         yday = yday(haulout_dt),
-         year = year(haulout_dt)) %>%
-  mutate(coords_x = st_coordinates(st_transform(.,4326))[,1],
-         coords_y = st_coordinates(st_transform(.,4326))[,2],
-         solar_hour = solaR::local2Solar(haulout_dt,coords_x) %>% lubridate::hour(),
-         sin1 = sin(pi*solar_hour/12),
-         cos1 = cos(pi*solar_hour/12),
-         sin2 = sin(pi*solar_hour/6),
-         cos2 = cos(pi*solar_hour/6),
-         sin3 = sin(pi*solar_hour/4),
-         cos3 = sin(pi*solar_hour/4),
-         day = (yday-120)/10,
-         day2 = day^2,
-         day3 = day^3
-  ) %>%
-  mutate(precip = rast_acpcp,
-         temp2 = (rast_air2m-270)/27,
-         temp0 = (rast_airsfc-270)/27,
-         pressure = (rast_prmsl-100000)/10000,
-         wind = sqrt(rast_uwnd^2 + rast_vwnd^2)/10
-  ) %>%
-  mutate(northing = st_coordinates(.)[,2]/mean(st_coordinates(.)[,2]),
-         north_of_bering = case_when(
-           coords_y > 66.5 ~ TRUE,
-           TRUE ~ FALSE
-         )) %>%
[…]

### 2021-12-22 analysis/scripts/model_year_hf.R

@@ -1,106 +1,0 @@
-library(tidyverse)
-library(here)
-library(lubridate)  #some date time functionality
-library(glmmLDTS)
-library(solaR)  #for getting hour of day in solar time
-library(splines)
-library(sf)
-library(DBI)
-library(rnaturalearth)
-library(purler)
-
-create_ar1_id <- function(h_dt) {
-  ar1_id <- tibble(
-    t_seq = seq(from = h_dt[1], to = h_dt[length(h_dt)], by = "hour",)
-  ) %>% left_join(tibble(h_dt = h_dt, haulout_dt = h_dt), by = c("t_seq" = "h_dt")) %>%
-    mutate(x = is.na(haulout_dt),
-           rle = rlenc_id(x)) %>%
-    filter(!is.na(haulout_dt)) %>%
-    pull(rle)
-
-  return(ar1_id)
-}
-
-load(here::here('data/dat_sf.Rdata'))
-
-dat.sf <- dat.sf  %>%
-  mutate(dry = round(percent_dry/100)) %>%
-  mutate(hour_utc = lubridate::hour(haulout_dt),
-         yday = yday(haulout_dt),
-         year = lubridate::year(haulout_dt)) %>%
-  mutate(coords_x = st_coordinates(st_transform(.,4326))[,1],
-         coords_y = st_coordinates(st_transform(.,4326))[,2],
-         solar_hour = solaR::local2Solar(haulout_dt,coords_x) %>% lubridate::hour(),
-         sin1 = sin(pi*solar_hour/12),
-         cos1 = cos(pi*solar_hour/12),
-         sin2 = sin(pi*solar_hour/6),
-         cos2 = cos(pi*solar_hour/6),
-         sin3 = sin(pi*solar_hour/4),
-         cos3 = sin(pi*solar_hour/4),
-         day = (yday-120)/10,
-         day2 = day^2,
-         day3 = day^3
-  ) %>%
-  mutate(precip = rast_acpcp,
-         temp2 = (rast_air2m-270)/27,
-         temp0 = (rast_airsfc-270)/27,
-         pressure = (rast_prmsl-100000)/10000,
-         wind = sqrt(rast_uwnd^2 + rast_vwnd^2)/10
-  ) %>%
-  mutate(northing = st_coordinates(.)[,2]/mean(st_coordinates(.)[,2]),
-         north_of_bering = case_when(
-           coords_y > 66.5 ~ TRUE,
-           TRUE ~ FALSE
-         )) %>%
[…]

### 2021-12-22 analysis/scripts/model_year_pl.R

@@ -1,106 +1,0 @@
-library(tidyverse)
-library(here)
-library(lubridate)  #some date time functionality
-library(glmmLDTS)
-library(solaR)  #for getting hour of day in solar time
-library(splines)
-library(sf)
-library(DBI)
-library(rnaturalearth)
-library(purler)
-
-create_ar1_id <- function(h_dt) {
-  ar1_id <- tibble(
-    t_seq = seq(from = h_dt[1], to = h_dt[length(h_dt)], by = "hour",)
-  ) %>% left_join(tibble(h_dt = h_dt, haulout_dt = h_dt), by = c("t_seq" = "h_dt")) %>%
-    mutate(x = is.na(haulout_dt),
-           rle = rlenc_id(x)) %>%
-    filter(!is.na(haulout_dt)) %>%
-    pull(rle)
-
-  return(ar1_id)
-}
-
-load(here::here('data/dat_sf.Rdata'))
-
-dat.sf <- dat.sf  %>%
-  mutate(dry = round(percent_dry/100)) %>%
-  mutate(hour_utc = lubridate::hour(haulout_dt),
-         yday = yday(haulout_dt),
-         year = lubridate::year(haulout_dt)) %>%
-  mutate(coords_x = st_coordinates(st_transform(.,4326))[,1],
-         coords_y = st_coordinates(st_transform(.,4326))[,2],
-         solar_hour = solaR::local2Solar(haulout_dt,coords_x) %>% lubridate::hour(),
-         sin1 = sin(pi*solar_hour/12),
-         cos1 = cos(pi*solar_hour/12),
-         sin2 = sin(pi*solar_hour/6),
-         cos2 = cos(pi*solar_hour/6),
-         sin3 = sin(pi*solar_hour/4),
-         cos3 = sin(pi*solar_hour/4),
-         day = (yday-120)/10,
-         day2 = day^2,
-         day3 = day^3
-  ) %>%
-  mutate(precip = rast_acpcp,
-         temp2 = (rast_air2m-270)/27,
-         temp0 = (rast_airsfc-270)/27,
-         pressure = (rast_prmsl-100000)/10000,
-         wind = sqrt(rast_uwnd^2 + rast_vwnd^2)/10
-  ) %>%
-  mutate(northing = st_coordinates(.)[,2]/mean(st_coordinates(.)[,2]),
-         north_of_bering = case_when(
-           coords_y > 66.5 ~ TRUE,
-           TRUE ~ FALSE
-         )) %>%
[…]

### 2021-12-22 analysis/scripts/predict.R

@@ -1,385 +1,0 @@
-library(glmmLDTS)
-library(mgcv)
-library(dplyr)
-library(forcats)
-library(ggplot2)
-
-# get the ribbon seal fit object
-fit_ribbon <- readRDS(here::here("data/fit_ribbon.Rdata"))
-
-# function to create a prediction data frame for a given age_sex group
-create_newdata <- function(data, age_sex) {
-  df_list <- vector(mode = "list", length = length({{age_sex}} ))
-
-  for (a_s in {{age_sex}}) {
-  range_yday <- data %>% filter(age_sex == a_s ) %>%
-    summarize(start_day = min(yday),
-              end_day = max(yday))
-  start_day <- range_yday$start_day
-  end_day <- range_yday$end_day
-  n_days = (end_day - start_day) + 1
-
-  # for wx covariates we'll use a gam to get values by day/hour since wx is likely
-  # to vary w/in day over the season
-
-  gam.baro <-
-    gam(pressure ~ s(yday), data = data)
-  gam.temp <-
-    gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), data = data)
-  gam.wind <-
-    gam(wind ~ s(yday) + s(as.numeric(solar_hour)), data = data)
-  gam.precip <-
-    gam(precip ~ s(yday), data = data)
-
-  wx_new_data <- data.frame(
-    solar_hour = rep(0:23, each = n_days),
-    yday = rep(start_day:end_day, times = 24)
-  )
-
-  temp_pred <- predict(gam.temp, newdata = wx_new_data)
-  wind_pred <- predict(gam.wind, newdata = wx_new_data)
-  baro_pred <- predict(gam.baro, newdata = wx_new_data)
-  precip_pred <- predict(gam.precip, newdata = wx_new_data)
-
-  df <- data.frame(
-    age_sex = a_s,
-    solar_hour = rep(0:23, each = n_days),
-    yday = rep(start_day:end_day, times = 24),
-    northing = mean(data$northing),
-    temp2 = temp_pred,
-    wind = wind_pred,
-    pressure = baro_pred,
-    precip = precip_pred
-  ) %>%
-    mutate(
-      sin1 = sin(pi * solar_hour / 12),
-      cos1 = cos(pi * solar_hour / 12),
-      sin2 = sin(pi * solar_hour / 6),
-      cos2 = cos(pi * solar_hour / 6),
[…]

### 2021-12-22 analysis/scripts/predict_year.R

@@ -1,182 +1,0 @@
-library(tidyverse)
-library(lubridate)
-library(mgcv)
-library(glmmLDTS)
-
-# function to create a prediction data frame for a given age_sex group
-create_newdata_year <- function(data, age_sex) {
-  df_list <- vector(mode = "list", length = length({{age_sex}} ))
-  df_list_years <- vector(mode = "list", length = length(levels(data$year)))
-
-  for (a_s in {{age_sex}}) {
-  for (y in levels(data$year)) {
-    d <- data %>% filter(year == y, age_sex == a_s)
-    if(nrow(d) == 0) { next }
-    range_yday <- d %>%
-      summarize(start_day = min(yday),
-                end_day = max(yday))
-    if(range_yday$start_day > 152) {next}
-    if(range_yday$end_day < 121) {next}
-    print(range_yday)
-    start_day <- range_yday$start_day
-    end_day <- range_yday$end_day
-    n_days = (end_day - start_day) + 1
-
-    # for wx covariates we'll use a gam to get values by day/hour since wx is likely
-    # to vary w/in day over the season
-
-    gam.baro <-
-      gam(pressure ~ s(yday), data = data)
-    gam.temp <-
-      gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), data = data)
-    gam.wind <-
-      gam(wind ~ s(yday) + s(as.numeric(solar_hour)), data = data)
-    gam.precip <-
-      gam(precip ~ s(yday), data = data)
-
-    wx_new_data <- data.frame(
-      solar_hour = 12,
-      yday = start_day:end_day
-    )
-
-    temp_pred <- predict(gam.temp, newdata = wx_new_data)
-    wind_pred <- predict(gam.wind, newdata = wx_new_data)
-    baro_pred <- predict(gam.baro, newdata = wx_new_data)
-    precip_pred <- predict(gam.precip, newdata = wx_new_data)
-
-    df <- data.frame(
-      age_sex = a_s,
-      year = y,
-      solar_hour = 12,
-      yday = start_day:end_day,
-      northing = mean(data$northing),
-      temp2 = temp_pred,
-      wind = wind_pred,
-      pressure = baro_pred,
-      precip = precip_pred
-    ) %>%
-      mutate(
-        sin1 = sin(pi * solar_hour / 12),
[…]

### 2021-12-22 analysis/scripts/seaice_extent.R

@@ -1,41 +1,0 @@
-library(tidyverse)
-library(lubridate)
-library(glue)
-library(DBI)
-library(dbplyr)
-library(sf)
-
-load(here::here("data/dat_sf.Rdata"))
-
-xmin <- st_bbox(dat.sf)$xmin %>% as.integer()
-ymin <- st_bbox(dat.sf)$ymin %>% as.integer()
-xmax <- st_bbox(dat.sf)$xmax %>% as.integer()
-ymax <- st_bbox(dat.sf)$ymax %>% as.integer()
-epsg <- 3571
-
-con <- dbConnect(
-  odbc::odbc(),
-  dsn = "PostgreSQL pep",
-  uid = keyringr::get_kc_account("pgpep_londonj"),
-  pwd = keyringr::decrypt_kc_pw("pgpep_londonj"),
-  sslmode = "disable"
-)
-
-extent_list <- vector(mode = "list")
-
-for(y in 2005:2020) {
-  print(y)
-  start_date <- glue::glue('{y}-02-15')
-  end_date <- glue::glue('{y}-07-15')
-
-  sql<- glue_sql("
-           SELECT *
-  FROM environ.fxn_sea_ice_extent_by_date(
-    0.15,{xmin},{ymin},{xmax},{ymax},
-    {epsg},
-    {start_date},{end_date} )",
-  .con = con)
-  extent_list[[as.character(y)]] <- DBI::dbGetQuery(con, sql)
-}
-
-dbDisconnect(con)

### 2021-12-22 README.Rmd
ADD STATIC GRAPH IMAGE TO README
MORE README UPDATES
cleaned up and added a few additional sections

@@
 output: github_document
-always_allow_html: true
 ---
@@
 
+## Draft Manuscript Under Active Development
+
+Please note this manuscript is still in draft form and under active development.
+Changes to results, code, and the manuscript are likely and this **should not
+be cited without author permission**. We plan to provide a preprint to 
+[bioRxiv](https://www.biorxiv.org/) once internal NOAA reviews and final 
+revisions are completed.
+
 ## Contents
@@
     for manuscript. Includes code to reproduce the figures and tables
-    generated by the analysis. It also has a rendered version,
+    generated by the analysis. It also has a rendered MS Word version,
     `London_HauloutBehavior.docx`, suitable for reading (the code is replaced by figures
@@
   - [:file\_folder: figures](/analysis/figures): figures generated
-  - [:file\_folder: scripts](/analysis/scripts): R scripts in support of analysis.
   - [:file\_folder: templates](/analysis/templates): template files and scripts.
@@
 
-```{r, echo=FALSE}
-library(targets)
-targets::tar_glimpse()
-```
+![](./tar_glimpse.png)
 
-### NOAA README
+## Inspiration
 
-This repository is a scientific product and is not official communication of the National Oceanic and Atmospheric Administration, or the United States Department of Commerce.
-All NOAA GitHub project code is provided on an ‘as is’ basis and the user assumes responsibility for its use.
-Any claims against the Department of Commerce or Department of Commerce bureaus stemming from the use of this GitHub project will be governed by all applicable Federal law.
-Any reference to specific commercial products, processes, or services by service mark, trademark, manufacturer, or otherwise, does not constitute or imply their endorsement, recommendation or favoring by the Department of Commerce.
-The Department of Commerce seal and logo, or the seal and logo of a DOC bureau, shall not be used in any manner to imply endorsement of any commercial product or activity by DOC or the United States Government.
+I would like to provide special acknowledgement to Ben Marwick and Carl Boetinger.
+This research compendium borrows heavily from Ben's [{rrtools}](https://github.com/benmarwick/rrtools)
+package. Carl's example research compendiums ([https://github.com/cboettig/noise-phenomena](https://github.com/cboettig/noise-phenomena) and
+[https://github.com/cboettig/nonparametric-bayes](https://github.com/cboettig/nonparametric-bayes)) were
+also of great use when adapting the structure to meet my needs.

### 2021-12-22 README.Rmd
MORE README ENHANCEMENTS & EDITS

@@
 
-This repository contains the research compendium in support of the above titled paper.
-The manuscript is currently in preparation and the text, code, analysis, and results
-are under active development.
+This repository serves as the research compendium in support of the above titled paper.
+As the manuscript works its way through peer review and the publications process,
+additional reproducibility features (e.g. Dockerfile, MyBinder.org) and
+an R/Shiny application to explore results will be made available.
+
+Major releases of the research compendium are published and archived with Zenodo [https://doi.org/10.5281/zenodo.4638221](https://doi.org/10.5281/zenodo.4638221)
 

### 2021-12-23 analysis/paper/London_HauloutBeahvior.Rmd
ADDRESS PCONN INITAL COMMENTS to v1.5

@@
 the Bering and Chukchi seas mean direct scientific observation of seal behavior
-is impractical. Thus, bio-logging devices are, thus, especially useful tools for
+is impractical. Thus, bio-logging devices are especially useful tools for
 collecting key information on movement and haul-out behavior for these species.
@@
 
-While analyzing haul-out patterns clearly has utility for understanding natural
-history and ecology, some studies have focused, as we do here, specifically on
-quantifying the probability of seals being hauled out as a means of quantifying
-seals in the water that are not counted during aerial surveys. These studies
-typically use logistic regression-style analyses to estimate the time-specific
-probability of being hauled out based on 'wet/dry' data relayed by bio-loggers.
-In these models, haul-out probabilities were expressed as a function of
-predictive covariates, such as time-of-day, day-of-year, sex, age class, and
-environmental conditions (e.g., [@reder2003], [@bengtson2004], [@bengtson2005a],
-[@udevitz2009], [@verhoef2014a], and [@southwell2008]). However, sample sizes
-have often been insufficient to permit strong inference about life history
-and/or seasonal variation in haul-out probabilities. For instance, Bengtson and
-Cameron's [-@bengtson2004] study included 5 adult and 2 juvenile crabeater
-seals, while Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed
-seals in the Chukchi Sea. These studies were often further limited by logistical
-constraints on fieldwork and the attachment duration or operational life
-of bio-loggers. In this study, we address some of these limitations by deploying
-small bio-loggers designed for longer-term attachment on a subset of the study
+Knowledge of haul-out patterns is not only important for understanding natural
+history and ecology, but also for developing "availability" correction factors
+for aerial surveys. Specifically, researchers need to know the fraction of seals
+hauled out (versus in the water) when aerial surveys are conducted. Studies
+estimating availability correction factors for seals typically use logistic
+regression-style analyses to estimate the time-specific probability of being
+hauled out based on 'wet/dry' data relayed by bio-loggers. In these models,
+haul-out probabilities were expressed as a function of predictive covariates,
+such as time-of-day, day-of-year, sex, age class, and environmental conditions
+(e.g., [@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
+[@verhoef2014a], and [@southwell2008]). However, sample sizes have often been
+insufficient to permit strong inference about life history and/or seasonal
+variation in haul-out probabilities. For instance, Bengtson and Cameron's
+[-@bengtson2004] study included 5 adult and 2 juvenile crabeater seals, while
+Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed seals in the
+Chukchi Sea. These studies were often further limited by logistical constraints
+on fieldwork and the attachment duration or operational life of bio-loggers. In
+this study, we address some of these limitations by deploying small bio-loggers
+designed for longer-term attachment on rear flippers of a subset of the study
 individuals. These devices are designed to collect data through the molt period
@@
 behavior of 3 species of ice-associated seals in the Bering and Chukchi seas. We
-focused our effort on bearded, ribbon, and spotted seals and did not include
-ringed seals in the analysis. Ringed seals are unique in their use of snow lairs
-in the spring months, and they warrant a separate study that includes additional
-environmental covariates, such as the onset of snow melt and snow depth. Our
+focused our effort on bearded, ribbon, and spotted seals. Our
 goals were threefold. First, we wished to establish baseline estimates for the
@@
 
-Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
-bio-loggers deployed on bearded, ribbon, and spotted seals were subset to
-include only records from 1 March to 15 July between 
+Haul-out behavior data from bio-loggers deployed on 
+`r dplyr::n_distinct(deploy_table$speno)` bearded, ribbon, and spotted seals were
+subset to include only records from 1 March to 15 July between 
 `r min(lubridate::year(analysis_data$haulout_dt))` and 
@@
 ```{r tblDeploy, include=TRUE, tab.id="tblDeploy",tab.cap = "Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2020.
-Total seal hours represents the sum of available data across all seals."}
+Total seal hours represents the sum of available data across all seals.
+Because young-of-the-year are advanced to subadult on 1 March of the following year, some individual seals are represented in both columns in this table"}
 
@@
 
-Haul-out behavior data was recorded and transmitted via the Argos satellite
+Haul-out behavior data were recorded and transmitted via the Argos satellite
 network as hourly percent-dry timelines. For each hour of a day, the wet/dry
@@
 programmed in a manner to maximize data transmission during the spring pupping
-and molting period, however, hourly percent-dry data for days during a
-deployment were not successfully transmitted. This is due to a variety of
-factors including satellite coverage, tag availability (i.e. tags mounted to the
-rear flipper are not available to the surface while the seal is at sea), tag
-performance, duty cycling, and extra-terrestrial atmospheric interference.
+and molting period; however, hourly percent-dry data were not always
+successfully transmitted. This is due to a variety of factors including
+satellite coverage, tag availability (i.e. tags mounted to the rear flipper
+often do not transmit while at sea), tag performance, duty cycling, and
+extra-terrestrial atmospheric interference. Fortunately, missing records do not
+seem to greatly bias inference about haul-out probabilities [@conn2012b].
 
@@
 Of key interest in this study was the relationship between haul-out behavior and
-weather covariates that vary with space and time. We explored the use of a
-continuous-time correlated random walk [@devins.johnson2008] movement model to
-predict locations at specific times. However, the sparse nature of data from
+weather covariates that vary with time and animal location. We explored the use
+of a continuous-time correlated random walk [@devins.johnson2008] movement model
+to predict locations at specific times. However, the sparse nature of data from
 some bio-loggers, especially those mounted to the rear flipper, resulted in poor
@@
 was assigned an error radius based on either the categorical location quality (
-_3_=250m, _2_=500m, _1_=1500m, _0_=2500m, _A_=2500m, _B_=2500m) or, when
-available, the estimated error radius from the Argos Kalman filter algorithm.
-Location estimates from FastLoc GPS were all assigned an error radius of 50m.
-Any days where haul-out observations were present but location data were missing
-we used the last calculated weighted average daily location, and any days where
-the location intersected with land were removed from the data set. We recognized
-that bearded and spotted seals haul out on land. However, assessing the
-relationship between haul-out behavior and weather covariates and their
-availability for aerial surveys on land was outside the scope of this study.
-Additionally, any daily locations on land are likely more reflective of
-coordinate averaging and measurement error instead of actual use of coastal
-features. Figure \@ref(fig:dataMap) shows the spatial distribution weighted
-locations with available haul-out behavior data for each species across the
-study area.
+_3_=250m, _2_=500m, _1_=1500m, _0_=2500m [@lopez2013]; we chose 2500m for
+location classes _A_ and _B_) or, when available, the estimated error radius
+from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
+all assigned an error radius of 50m. Any days where haul-out observations were
+present but location data were missing we used the last calculated weighted
+average daily location, and any days where the location intersected with land
+were removed from the data set. We recognized that bearded and spotted seals
+haul out on land. However, assessing the relationship between haul-out behavior
+and weather covariates and their availability for aerial surveys on land was
+outside the scope of this study. Additionally, any daily locations on land are
+likely more reflective of coordinate averaging and measurement error instead of
+actual use of coastal features. Figure \@ref(fig:dataMap) shows the spatial
+distribution weighted locations with available haul-out behavior data for each
+species across the study area.
 
@@
 seal which do not provide at-sea locations, limiting our ability to fully
-evaluate habitat preferences related to sea-ice. It should be noted that the
-aerial survey counts are modeled with sea-ice and other habitat parameters. So,
-any final abundance estimates that combine models of survey counts and our model
-for proportion of seals hauled out on ice do account for the strong association
-between these species and sea-ice in the Bering and Chukchi seas. Lastly, our
-study was limited to the spring season when seal haul-out tendencies are
-strongly influenced by pupping, nursing, breeding behavior, and molt and these
-drivers are likely more influential than specific sea-ice concentration.
-Crawford et al [-@crawford2019] compared haul-out probability models for ringed
-seals and found those that only included season (and not sea-ice concentration)
-were the most parsimonious. Further, studies to evaluate habitat selection in
-relation sea-ice parameters and other environmental covariates are in progress.
-For these reasons, we have elected not to use sea ice concentration as a
-predictor for haul-out probability.
+evaluate fine-scale habitat preferences related to sea-ice. Lastly, our study
+was limited to the spring season when seal haul-out tendencies are strongly
+influenced by pupping, nursing, breeding behavior, and molt and these drivers
+are likely more influential than specific sea-ice concentration. Crawford et al
+[-@crawford2019] compared haul-out probability models for ringed seals and found
+those that only included season (and not sea-ice concentration) were the most
+parsimonious. For these reasons, we have elected not to use sea ice
+concentration as a predictor for haul-out probability in the present study
+(note, however, that habitat selection analyses incorporating sea-ice
+concentration are a focus of current and future research).
 
@@
 clearly violated if hourly responses are modeled. Any analysis that ignores
-temporal dependence in responses will thus have overstated precision
+temporal autocorrelation in responses will thus have overstated precision
 [@betts2006].
@@
 individual random effects representing heterogeneity in individual behavior. We
-used the {glmmLDTS} package [@verhoef2010a] to implement GLMPMs. We explored two
+used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
 different model formulations for our data, and owing to the large number of
@@
 probabilities for all species except bearded seals because of low sample size.
-While both ribbon and spotted seals exhibited a unimodal diel haul-out pattern
+Although ribbon and spotted seals exhibited a unimodal diel haul-out pattern
 generally centered around local solar noon, there were key differences across
 species, age, and sex that match expectations given what we know of their
-ecological behavior. Ribbon seals are not known to form triads and our model
-predicts a progression of increased haul-out behavior with females starting
-earlier in the season than males. Notably, female ribbon seals spend a large
-portion of the day in the water during the pupping period, aligning with the
-hypothesis that ribbon seal females continue foraging while nursing. Spotted
-seals are known to form triads during the breeding season [@burns1973] where a
-female and dependent pup are accompanied on the ice by a suitor male. The male
-waits for the female to wean the pup and enter estrus, and fends off any other
-potential suitor males. Triad formation results in both males and females
-spending a large portion of the day hauled out on ice and a protracted spring
-haul-out season for both sexes. Because mating occurs in the water, females may
-also be less inclined to interrupt their haul-out with foraging trips while
-still nursing the pup. We see this reflected in the predicted haul-out behavior,
-with both males and females exhibiting a broad distribution of time out of the
-water throughout the solar day and the season. In the case of both ribbon and
+ecological behavior. Spotted seals are known to form triads during the breeding
+season [@burns1973] where a female and dependent pup are accompanied on the ice
+by a suitor male. The male waits for the female to wean the pup and enter
+estrus, and fends off any other potential suitor males. Triad formation results
+in both males and females spending a large portion of the day hauled out on ice
+and a protracted spring haul-out season for both sexes. Because mating occurs in
+the water, females may also be less inclined to interrupt their haul-out with
+foraging trips while still nursing the pup. We see this reflected in the
+predicted haul-out behavior, with both males and females exhibiting a broad
+distribution of time out of the water throughout the solar day and the
+season.Ribbon seals are not known to form triads and our model predicts a
+progression of increased haul-out behavior with females starting earlier in the
+season than males. Notably, female ribbon seals spend a large portion of the day
+in the water during the pupping period, aligning with the hypothesis that ribbon
+seal females continue foraging while nursing.  In the case of both ribbon and
 spotted seals, subadults are the first to begin consistent haul-out behavior and
@@
 being important components of the model. Spotted seals were most affected by
-wind and barometric pressures. And, for bearded seals, the model indicated wind
+wind and barometric pressure. And, for bearded seals, the model indicated wind
 and temperature had the greatest impact. In general, and as might be expected,
@@
 provide an optimal situation for predator detection. Our results highlight the
-importance for haul-out behavior models (and subsequent calculations of
-availability during surveys) to incorporate weather covariates.
+importance of incorporating weather covariates when analyzing haul-out 
+behavior and calculating availability corrections for aerial surveys.
 
@@
 of abundance and trend. Those ramifications will only be exacerbated if climate
-variability amplifies those interannual differences.
+variability amplifies interannual differences.
 
@@
 0.60, respectively, under the current analysis framework. Our current estimates
-estimates of haul-out probability reflect increased sample sizes in terms of
+of haul-out probability reflect increased sample sizes in terms of
 number of animals, but also improvements to the way data were prepared prior to
@@
 deployment of bio-logging devices used in this study were often done in
-collaboration with Alaska Native seal hungers and their communities. We would
+collaboration with Alaska Native seal hunters and their communities. We would
 like to especially acknowledge the communities of Kotzebue, Koyuk, Utqiaġvik,
@@
 Slope Borough Department of Wildlife Management (NSB) were important
-contributions to the findings presented here. Both agencies collaborate
-closely with Alaska Native communities and the contributions of local
-hunters to the successful capture, deployment, and understanding of seal
-ecology are invaluable. Deployments in the western Bering Sea and Sea of
-Okhotsk were done in collaboration with Russian colleagues and North
-Pacific Wildlife.
+contributions to the findings presented here. Both agencies collaborate closely
+with Alaska Native communities and the contributions of local hunters to the
+successful capture, deployment, and understanding of seal ecology are
+invaluable. Deployments in the western Bering Sea were done in collaboration
+with Russian colleagues and North Pacific Wildlife.
 

### 2021-12-23 analysis/paper/London_HauloutBeahvior.Rmd
clear up text re: response variable
closes #19

@@
 In this study, we used 16 years of bio-logging data to investigate the haul-out
-behavior of 3 species of ice-associated seals in the Bering and Chukchi seas. We
-focused our effort on bearded, ribbon, and spotted seals. Our
-goals were threefold. First, we wished to establish baseline estimates for the
-chronology of haul-out behavior in the critical spring season for each species
-across different age and sex classes. Second, we refined estimates of
+behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
+Our goals were threefold. First, we wished to establish baseline estimates for
+the chronology of haul-out behavior in the critical spring season for each
+species across different age and sex classes. Second, we refined estimates of
 haul-out availability corrections for aerial surveys in order to improve
@@
 
+```{r binomialdata}
+
+portion_10 <- function(d) {
+  n <- nrow(d)
+  f <- d %>% dplyr::filter(percent_dry <= 10) %>% 
+    nrow()
+  p10 <- f/n
+  return(p10)
+}
+
+portion_90 <- function(d) {
+  n <- nrow(d)
+  f <- d %>% dplyr::filter(percent_dry >= 90) %>% 
+    nrow()
+  p90 <- f/n
+  return(p90)
+}
+
+```
+
 Haul-out behavior data were recorded and transmitted via the Argos satellite
@@
 whether haul-out patterns varied by year and to determine the annual timing of
-apparent peaks in haul-out behavior. For both models, the response variable was
-the transmitted hourly percent-dry value.
+apparent peaks in haul-out behavior. For both models, we assumed an hourly
+Bernoulli response (i.e., whether tags were mostly dry or mostly wet) where
+the linear predictor was modeled on the logit scale. This is consistent with
+previous approaches [@verhoef2014a, @london2012] and only
+`r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
+observations fell between 10% and 90% hourly percent dry.
 

### 2021-12-23 analysis/paper/London_HauloutBeahvior.Rmd

@@
   dplyr::mutate(yday = lubridate::yday(haulout_dt)) %>% 
-  dplyr::left_join(yday_groups) %>% 
-  dplyr::group_by(species, col_group, row_group) %>% 
+  dplyr::group_by(species, yday) %>% 
   dplyr::count()
@@
 
-```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of hourly percent-dry biologger data from 1 March (upper left) to 15 July (lower right) for each species.
+```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of hourly percent-dry biologger data from 1 March to 15 July for each species.
 Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020.
@@
 
-ggplot(data = seal_hours_tbl) + 
-  geom_point(aes(x = col_group, y = row_group, color = n, size = n), stroke = 0, shape = 15) +
-  scale_size(range = c(1,3.4), breaks = c(100,500,1000,1500)) +
-  scale_color_gradient(low = "#6baed6", high = "#08306b") +
-  scale_x_continuous(breaks = c(1,4,8,12,16),
+ggplot(data = seal_hours_tbl, aes(x=yday, y=n, fill=n)) +
+  geom_col(binwidth = 1) +
+  scale_fill_gradient(low = "#6baed6", high = "#08306b") +
+  scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
                      labels = c("March", "April", "May", "June", "July"),
                      expand = expansion(mult = c(0.01, 0.025))) +
-  guides(color = "none",
-        size = guide_legend(override.aes = list(color = "#2171b5"),
-                            title.position = "bottom", title.hjust = 0.5)
-        ) +
+  guides(fill = guide_legend(title.position = "bottom", title.hjust = 0.5)) +
   facet_grid(species ~ .) + 
-  theme(
+  theme(axis.text.x = element_text(hjust = 0),
         axis.text.y = element_blank(),
-        axis.text.x = element_text(hjust = 0),
         axis.title = element_blank(),
@@
         legend.position = "bottom") +
-  labs(color = "seal-hours",
-       size = "seal hours of bio-logger data") +
+  labs(fill = "seal hours of bio-logger data") +
   ggtitle("Density of haul-out behavior observations varies from March through July",
           subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")
+
 ```

### 2021-12-24 analysis/paper/London_HauloutBeahvior.Rmd
Improved Figure 1 and other updates

@@
 
-```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of hourly percent-dry biologger data from 1 March to 15 July for each species.
+```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
 Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020.
@@
         legend.position = "bottom") +
-  labs(fill = "seal hours of bio-logger data") +
+  labs(fill = "seal-hours of bio-logger data") +
   ggtitle("Density of haul-out behavior observations varies from March through July",
@@
 
-```{r beardedHOwx,fig.height = 5.5, include=FALSE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across the range of weather conditions encountered in the observed data.
+Visualizing the marginal effect of each weather covariate on haul-out
+probability is difficult in this analysis because of the collinearity
+between covariates as well as the temporal variation in covariate values
+during the study period. The relationship of each weather covariate with
+haul-out probability, averaged over the other weather conditions, is
+more variable than simply showing model coefficients would imply. Figure
+\@ref(fig:beardedHOwx) presents the predicted haul-out probability of
+bearded seals across the range of weather conditions encountered in the
+observed data. The collinearity and temporal variation are apparent in
+the visual patterns that emerge.
+
+```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across the range of weather conditions encountered in the observed data.
 Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
@@
 `r ribbon_type3$temp2_wind$Prob.F`) showed a moderately negative
-influence on haul-out probability.
-
-Visualizing the marginal effect of each weather covariate on haul-out
-probability is difficult in this analysis because of the collinearity
-between covariates as well as the temporal variation in covariate values
-during the study period. The relationship of each weather covariate with
-haul-out probability, averaged over the other weather conditions, is
-more variable than simply showing model coefficients would imply. Figure
+influence on haul-out probability. As with bearded seals, Figure
 \@ref(fig:ribbonHOwx) presents the predicted haul-out probability of
 ribbon seals across the range of weather conditions encountered in the
-observed data. The collinearity and temporal variation are apparent in
-the visual patterns that emerge.
+observed data. Because our ribbon seal model included age and sex class, we can
+visualize the different influences of weather covariates those classes. Of note,
+there is an indication that sub-adult ribbon seals are more likely to haul-out
+at lower temperatures. But, this may also be simply reflective of their
+tendency to haul-out earlier in the season compared to adults.
 
@@
 present and consistent across each of the weather covariates (Figure
-\@ref(fig:spottedHOwx)).
+\@ref(fig:spottedHOwx)). There is a consistent ranking of adult males being the most
+likely to haul out, followed by adult females, and, then, subadults.
 

### 2021-12-24 README.Rmd
add download link for rendered docx

@@
     generated by the analysis. It also has a rendered MS Word version,
-    `London_HauloutBehavior.docx`, suitable for reading (the code is replaced by figures
+    [London_HauloutBehavior.docx](https://github.com/jmlondon/berchukHaulout/raw/main/analysis/paper/London_HauloutBeahvior.docx), suitable for reading (the code is replaced by figures
     and tables in this file)

### 2021-12-24 analysis/paper/London_HauloutBeahvior.Rmd

@@
 
-```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
+```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
 Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020.
@@
 
-ggplot(data = seal_hours_tbl, aes(x=yday, y=n, fill=n)) +
-  geom_col(binwidth = 1) +
+# ggplot(data = seal_hours_tbl, aes(x=yday, y=n, fill=n)) +
+#   geom_col(binwidth = 1) +
+#   scale_fill_gradient(low = "#6baed6", high = "#08306b") +
+#   scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
+#                      labels = c("March", "April", "May", "June", "July"),
+#                      expand = expansion(mult = c(0.01, 0.025))) +
+#   guides(fill = guide_legend(title.position = "bottom", title.hjust = 0.5)) +
+#   facet_grid(species ~ .) + 
+#   theme(axis.text.x = element_text(hjust = 0),
+#         axis.text.y = element_blank(),
+#         axis.title = element_blank(),
+#         panel.grid = element_blank(),
+#         legend.position = "bottom") +
+#   labs(fill = "seal-hours of bio-logger data") +
+#   ggtitle("Density of haul-out behavior observations varies from March through July",
+#           subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")
+
+ggplot(data = seal_hours_tbl, aes(x = yday, y = 1, fill = n))+
+        geom_tile()+
         scale_fill_gradient(low = "#6baed6", high = "#08306b") +
         scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
-                     labels = c("March", "April", "May", "June", "July"),
-                     expand = expansion(mult = c(0.01, 0.025))) +
-  guides(fill = guide_legend(title.position = "bottom", title.hjust = 0.5)) +
+                           labels = c("01 March", "01 April", "01 May", "01 June", "01 July"),
+                           expand = expansion(mult = c(0, 0))) +
+        scale_y_continuous(expand = expansion(mult = c(0, 0))) +
+  guides(fill = guide_colorbar(title.position = "bottom", 
+                               title.hjust = 0.5, barwidth = 15,
+                               barheight = 0.5)) +
   facet_grid(species ~ .) + 

### 2022-01-05 analysis/paper/London_HauloutBeahvior.Rmd
REDESIGN & IMPROVE HO WX PLOTS

--- 2021-12-24 analysis/paper/London_HauloutBeahvior.Rmd
@@
 library(ragg)
+library(stickylabeller)
 
@@
 
-ggplot(data = seal_hours_tbl, aes(x=yday, y=n, fill=n)) +
-  geom_col(binwidth = 1) +
-  scale_fill_gradient(low = "#6baed6", high = "#08306b") +
+# bar chart instead of a strip
+#
+# ggplot(data = seal_hours_tbl, aes(x=yday, y=n, fill=n)) +
+#   geom_col(binwidth = 1) +
+#   scale_fill_gradient(low = "#6baed6", high = "#08306b") +
+#   scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
+#                      labels = c("March", "April", "May", "June", "July"),
+#                      expand = expansion(mult = c(0.01, 0.025))) +
+#   guides(fill = guide_legend(title.position = "bottom", title.hjust = 0.5)) +
+#   facet_grid(species ~ .) + 
+#   theme(axis.text.x = element_text(hjust = 0),
+#         axis.text.y = element_blank(),
+#         axis.title = element_blank(),
+#         panel.grid = element_blank(),
+#         legend.position = "bottom") +
+#   labs(fill = "seal-hours of bio-logger data") +
+#   ggtitle("Density of haul-out behavior observations varies from March through July",
+#           subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")
+
+ggplot(data = seal_hours_tbl, aes(x = yday, y = 1, fill = n))+
+        geom_tile()+
+        scale_fill_viridis_c(option="mako") +
         scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
-                     labels = c("March", "April", "May", "June", "July"),
-                     expand = expansion(mult = c(0.01, 0.025))) +
-  guides(fill = guide_legend(title.position = "bottom", title.hjust = 0.5)) +
+                           labels = c("01 March", "01 April", "01 May", "01 June", "01 July"),
+                           expand = expansion(mult = c(0, 0))) +
+        scale_y_continuous(expand = expansion(mult = c(0, 0))) +
+  guides(fill = guide_colorbar(title.position = "bottom", 
+                               title.hjust = 0.5, barwidth = 15,
+                               barheight = 0.5)) +
   facet_grid(species ~ .) + 
@@
 
+Visualizing the marginal or conditional effect of an individual weather
+covariate on haul-out probability is difficult in this analysis because of the
+collinearity between covariates as well as the spatial and temporal variation
+across such a large region. The relationship of each weather covariate with
+haul-out probability, averaged over the other weather conditions, is more
+variable than simply showing model coefficients would imply. To elucidate
+underlying patterns between the model's predicted haul-out probability and
+specific weather covariates, we explored a subset of predictions during three
+separate time windows. The same set of predictions used to create the haul-out
+probability surfaces was reduced to include only local solar hours between 06:00
+and 18:00 during three 7-day time windows. The time windows were centered on the
+day of maximum haul-out probability and 25 days before and 25 days after peak
+haul out. While this approach does not entirely remove patterns of collinearity
+from the visualization (e.g. temperature increases with day of year), by
+limiting our range of days and hours we gain some insight into how haul-out
+probability varies across different weather covariate values. Figures
+\@ref(fig:beardedHOwx), @\ref(fig:ribbonHOwx), and @\ref(fig:spottedHOwx)
+present the predicted haul-out probability with respect to age and sex class
+across the range of temperature and wind conditions encountered in the three
+time windows. The visualizations also include vertical lines representing 95%
+confidence intervals around the predicted haul-out probability to better
+communicate the variation in model uncertainty. Additionally, simple linear fits
+are shown to assist interpretation of overall patterns.
+
 # Results
@@
 
-Visualizing the marginal effect of each weather covariate on haul-out
-probability is difficult in this analysis because of the collinearity
-between covariates as well as the temporal variation in covariate values
-during the study period. The relationship of each weather covariate with
-haul-out probability, averaged over the other weather conditions, is
-more variable than simply showing model coefficients would imply. Figure
-\@ref(fig:beardedHOwx) presents the predicted haul-out probability of
-bearded seals across the range of weather conditions encountered in the
-observed data. The collinearity and temporal variation are apparent in
-the visual patterns that emerge.
-
 ```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across the range of weather conditions encountered in the observed data.
-Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
-Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}
-
-p_temp <- ggplot(plot_df, aes(temp2*27, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
+Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability."}
+
+plot_df <- bearded_newdata %>%
+  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+         month = lubridate::month(date,label = TRUE),
+         day = lubridate::day(date)) %>% 
+  filter(age_sex != "YOUNG OF YEAR",
+         between(solar_hour, 8,16)) %>% 
+  group_by(age_sex)
+
+plot_df_early <- plot_df %>% tidyr::nest() %>% 
+  rowwise() %>% 
+  mutate(peak_idx = which.max(data$ho_prob),
+         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
+  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
+  rowwise() %>% 
+  mutate(data = list(data %>% filter(between(yday,(peak_avg-25)-3, (peak_avg-25)+3))),
+         season = "early") %>% 
+  unnest(cols = c(data))
+
+plot_df_peak <- plot_df %>% tidyr::nest() %>% 
+  rowwise() %>% 
+  mutate(peak_idx = which.max(data$ho_prob),
+         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
+  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
+  rowwise() %>% 
+  mutate(data = list(data %>% filter(between(yday,(peak_avg)-3, (peak_avg)+3))),
+         season = "peak") %>% 
+  unnest(cols = c(data))
+
+plot_df_late <- plot_df %>% tidyr::nest() %>% 
+  rowwise() %>% 
+  mutate(peak_idx = which.max(data$ho_prob),
+         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
+  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
+  rowwise() %>% 
+  mutate(data = list(data %>% filter(between(yday,(peak_avg+25)-3, (peak_avg+25)+3))),
+         season = "late") %>% 
+  unnest(cols = c(data))
+
+plot_df <- bind_rows(list(plot_df_early, plot_df_peak, plot_df_late)) %>% 
+  mutate(date_start = case_when(
+    season == "peak" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 3, "%d %b"),
+    season == "early" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 28, "%d %b"),
+    season == "late" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 22, "%d %b")
+  )) %>% 
+  mutate(date_end = case_when(
+    season == "peak" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 3, "%d %b"),
+    season == "early" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 22, "%d %b"),
+    season == "late" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 28, "%d %b")
+  )) %>% 
+  mutate(season = forcats::fct_relevel(season, "early", "peak", "late"),
+         season = forcats::fct_recode(season, "peak haul out - 25 days" = "early",
+                             "peak haul out" = "peak",
+                             "peak haul out + 25 days" = "late"))
+
+p_temp <- ggplot(plot_df, 
+                 aes(temp2*27, ho_prob, color = age_sex, fill = age_sex)) +
+  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
+  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
+              show.legend = FALSE) +
+  scale_fill_d3() +
+  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1),
+                                      direction = "horizontal")) +
+  theme(legend.title = element_blank(),
+        legend.position = "none") +
+  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
+             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
   xlab("temperature (C)") + ylab("haul-out probability")
 
-p_wind <- ggplot(plot_df, aes(wind, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
-  xlab("wind") + ylab("")
-
-p_pressure <- ggplot(plot_df, aes(pressure, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
-  xlab("barometric pressure") + ylab("haul-out probability")
-
-p_precip <- ggplot(plot_df, aes(precip, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
-  xlab("precipitation") + ylab("")
+p_wind <- ggplot(plot_df, aes(wind*10, ho_prob, color = age_sex, fill = age_sex)) +
+  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
+  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
+              show.legend = FALSE) +
+  scale_fill_d3() +
+  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 0.75),
+                                      direction = "horizontal")) +
+  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
+             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
+  theme(legend.title = element_blank(),
+        legend.position = "bottom") +
+  xlab("total wind (m/s)") + ylab("haul-out probability")
 
@@
 
-((p_temp + p_wind) / (p_pressure + p_precip)) +
+p_temp / p_wind +
   plot_annotation(
-    title = 'Bearded seal haul-out behavior is less influeced by weather')
+    title = 'The influence of temperature and wind on predicted bearded seal \nhaul-out behavior varies within season',
+    subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
+    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate marginal patterns.')
 ```
@@
 ```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Variability in predicted haul-out probability of ribbon seals across the range of weather conditions encountered in the observed data.
-Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
-Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}
-
-p_temp <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(temp2*27, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  geom_smooth(se = FALSE) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  theme(legend.position = "none") +
+Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability."}
+
+plot_df <- ribbon_newdata %>%
+  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+         month = lubridate::month(date,label = TRUE),
+         day = lubridate::day(date)) %>% 
+  filter(age_sex != "YOUNG OF YEAR",
+         between(solar_hour, 8,16)) %>% 
+  group_by(age_sex)
+
+plot_df_early <- plot_df %>% tidyr::nest() %>% 
+  rowwise() %>% 
+  mutate(peak_idx = which.max(data$ho_prob),
+         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
+  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
+  rowwise() %>% 
+  mutate(data = list(data %>% filter(between(yday,(peak_avg-25)-3, (peak_avg-25)+3))),
+         season = "early") %>% 
+  unnest(cols = c(data))
+
+plot_df_peak <- plot_df %>% tidyr::nest() %>% 
+  rowwise() %>% 
+  mutate(peak_idx = which.max(data$ho_prob),
+         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
+  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
+  rowwise() %>% 
+  mutate(data = list(data %>% filter(between(yday,(peak_avg)-3, (peak_avg)+3))),
+         season = "peak") %>% 
+  unnest(cols = c(data))
+
+plot_df_late <- plot_df %>% tidyr::nest() %>% 
+  rowwise() %>% 
+  mutate(peak_idx = which.max(data$ho_prob),
+         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
+  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
+  rowwise() %>% 
+  mutate(data = list(data %>% filter(between(yday,(peak_avg+25)-3, (peak_avg+25)+3))),
+         season = "late") %>% 
+  unnest(cols = c(data))
+
+plot_df <- bind_rows(list(plot_df_early, plot_df_peak, plot_df_late)) %>% 
+  mutate(date_start = case_when(
+    season == "peak" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 3, "%d %b"),
+    season == "early" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 28, "%d %b"),
+    season == "late" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 22, "%d %b")
+  )) %>% 
+  mutate(date_end = case_when(
+    season == "peak" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 3, "%d %b"),
+    season == "early" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 22, "%d %b"),
+    season == "late" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 28, "%d %b")
+  )) %>% 
+  mutate(season = forcats::fct_relevel(season, "early", "peak", "late"),
+         season = forcats::fct_recode(season, "peak haul out - 25 days" = "early",
+                             "peak haul out" = "peak",
+                             "peak haul out + 25 days" = "late"))
+
+p_temp <- ggplot(plot_df, 
+                 aes(temp2*27, ho_prob, color = age_sex, fill = age_sex)) +
+  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
+  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
+              show.legend = FALSE) +
+  scale_fill_d3() +
+  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1),
+                                      direction = "horizontal")) +
+  theme(legend.title = element_blank(),
+        legend.position = "none") +
+  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
+             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
   xlab("temperature (C)") + ylab("haul-out probability")
 
-p_wind <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(wind, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) ),
-                 name = "") +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "top") +
-  xlab("wind") + ylab("")
-
-p_pressure <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(pressure, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
-  xlab("barometric pressure") + ylab("haul-out probability")
-
-p_precip <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(precip, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
-  xlab("precipitation") + ylab("")
+p_wind <- ggplot(plot_df, aes(wind*10, ho_prob, color = age_sex, fill = age_sex)) +
+  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
+  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
+              show.legend = FALSE) +
+  scale_fill_d3() +
+  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 0.75),
+                                      direction = "horizontal")) +
+  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
+             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
+  theme(legend.title = element_blank(),
+        legend.position = "bottom") +
+  xlab("total wind (m/s)") + ylab("haul-out probability")
 
@@
 
-p_temp + p_wind + p_pressure + p_precip +
+p_temp / p_wind +
   plot_annotation(
-    title = 'Ribbon seal haul-out behavior is influenced by weather')
+    title = 'The influence of temperature and wind on predicted ribbon seal \nhaul-out behavior varies within season',
+    subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
+    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate marginal patterns.')
 ```
@@
 ```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Variability in the predicted haul-out probability of spotted seals across the range of weather conditions encountered in the observed data.
-Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability.
-Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}
-
-library(ggsci)
-
-p_temp <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(temp2*27, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
+Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability."}
+
+plot_df <- spotted_newdata %>%
+  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+         month = lubridate::month(date,label = TRUE),
+         day = lubridate::day(date)) %>% 
+  filter(age_sex != "YOUNG OF YEAR",
+         between(solar_hour, 8,16)) %>% 
+  group_by(age_sex)
+
+plot_df_early <- plot_df %>% tidyr::nest() %>% 
+  rowwise() %>% 
+  mutate(peak_idx = which.max(data$ho_prob),
+         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
+  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
+  rowwise() %>% 
+  mutate(data = list(data %>% filter(between(yday,(peak_avg-25)-3, (peak_avg-25)+3))),
+         season = "early") %>% 
+  unnest(cols = c(data))
+
+plot_df_peak <- plot_df %>% tidyr::nest() %>% 
+  rowwise() %>% 
+  mutate(peak_idx = which.max(data$ho_prob),
+         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
+  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
+  rowwise() %>% 
+  mutate(data = list(data %>% filter(between(yday,(peak_avg)-3, (peak_avg)+3))),
+         season = "peak") %>% 
+  unnest(cols = c(data))
+
+plot_df_late <- plot_df %>% tidyr::nest() %>% 
+  rowwise() %>% 
+  mutate(peak_idx = which.max(data$ho_prob),
+         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
+  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
+  rowwise() %>% 
+  mutate(data = list(data %>% filter(between(yday,(peak_avg+25)-3, (peak_avg+25)+3))),
+         season = "late") %>% 
+  unnest(cols = c(data))
+
+plot_df <- bind_rows(list(plot_df_early, plot_df_peak, plot_df_late)) %>% 
+  mutate(date_start = case_when(
+    season == "peak" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 3, "%d %b"),
+    season == "early" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 28, "%d %b"),
+    season == "late" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 22, "%d %b")
+  )) %>% 
+  mutate(date_end = case_when(
+    season == "peak" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 3, "%d %b"),
+    season == "early" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 22, "%d %b"),
+    season == "late" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 28, "%d %b")
+  )) %>% 
+  mutate(season = forcats::fct_relevel(season, "early", "peak", "late"),
+         season = forcats::fct_recode(season, "peak haul out - 25 days" = "early",
+                             "peak haul out" = "peak",
+                             "peak haul out + 25 days" = "late"))
+
+p_temp <- ggplot(plot_df, 
+                 aes(temp2*27, ho_prob, color = age_sex, fill = age_sex)) +
+  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
+  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
+              show.legend = FALSE) +
+  scale_fill_d3() +
+  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1),
+                                      direction = "horizontal")) +
+  theme(legend.title = element_blank(),
+        legend.position = "none") +
+  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
+             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
   xlab("temperature (C)") + ylab("haul-out probability")
 
-p_wind <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(wind, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) ),
-                 name = "") +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "top") +
-  xlab("wind") + ylab("")
-
-p_pressure <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(pressure, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
-  xlab("barometric pressure") + ylab("haul-out probability")
-
-p_precip <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(precip, ho_prob, color = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.01, stroke = 0) +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1) )) +
-  geom_smooth(se = FALSE) +
-  theme(legend.position = "none") +
-  xlab("precipitation") + ylab("")
+p_wind <- ggplot(plot_df, aes(wind*10, ho_prob, color = age_sex, fill = age_sex)) +
+  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
+  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
+              show.legend = FALSE) +
+  scale_fill_d3() +
+  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 0.75),
+                                      direction = "horizontal")) +
+  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
+             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
+  theme(legend.title = element_blank(),
+        legend.position = "bottom") +
+  xlab("total wind (m/s)") + ylab("haul-out probability")
 
@@
 
-p_temp + p_wind + p_pressure + p_precip +
+p_temp / p_wind +
   plot_annotation(
-    title = 'Spotted seal haul-out behavior is influenced by weather')
+    title = 'The influence of temperature and wind on predicted spotted seal \nhaul-out behavior varies within season',
+    subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
+    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate marginal patterns.')
 ```

### 2022-01-05 R/create_newdata.R

@@ -14,13 +14,21 @@ create_newdata <- function(data, age_sex) {
     # to vary w/in day over the season
 
     gam.baro <-
-      gam(pressure ~ s(yday), data = data)
+      gam(pressure ~ s(yday), 
+          data = data, 
+          method = "REML")
     gam.temp <-
-      gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), data = data)
+      gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), 
+          data = data, 
+          method = "REML")
     gam.wind <-
-      gam(wind ~ s(yday) + s(as.numeric(solar_hour)), data = data)
+      gam(wind ~ s(yday) + s(as.numeric(solar_hour)), 
+          data = data, 
+          method = "REML")
     gam.precip <-
-      gam(precip ~ s(yday), data = data)
+      gam(precip ~ s(yday), 
+          data = data, 
+          method = "REML")
 
     wx_new_data <- data.frame(
       solar_hour = rep(0:23, each = n_days),

### 2022-01-05 analysis/paper/London_HauloutBeahvior.Rmd
CHANGE CAPTION & OTHER SMALL BITS

@@
 
-# bar chart instead of a strip
-#
-# ggplot(data = seal_hours_tbl, aes(x=yday, y=n, fill=n)) +
-#   geom_col(binwidth = 1) +
-#   scale_fill_gradient(low = "#6baed6", high = "#08306b") +
-#   scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
-#                      labels = c("March", "April", "May", "June", "July"),
-#                      expand = expansion(mult = c(0.01, 0.025))) +
-#   guides(fill = guide_legend(title.position = "bottom", title.hjust = 0.5)) +
-#   facet_grid(species ~ .) + 
-#   theme(axis.text.x = element_text(hjust = 0),
-#         axis.text.y = element_blank(),
-#         axis.title = element_blank(),
-#         panel.grid = element_blank(),
-#         legend.position = "bottom") +
-#   labs(fill = "seal-hours of bio-logger data") +
-#   ggtitle("Density of haul-out behavior observations varies from March through July",
-#           subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")
-
 ggplot(data = seal_hours_tbl, aes(x = yday, y = 1, fill = n))+
@@
 [@verhoef2010a] to evaluate model performance and important terms. We also
-produced predictions of haul-out behavior as a function of influential
+produced predictions of haul-out behavior as a function of three influential
 predictors (e.g. solar hour, day-of-year, age-sex). Weather covariates for these
@@
 across the study region. Such predictions were then used to develop haul-out
-probability surfaces, explore conditional effects of weather covariates, and
+probability surfaces (see figures \@ref(beardedHOCal), \@ref(ribbonHOCal), and
+\@ref(spottedHOCal)), explore conditional effects of weather covariates, and
 determine annual peaks in haul-out activity. The timing of peak haul-out
@@
 probability varies across different weather covariate values. Figures
-\@ref(fig:beardedHOwx), @\ref(fig:ribbonHOwx), and @\ref(fig:spottedHOwx)
+\@ref(fig:beardedHOwx), \@ref(fig:ribbonHOwx), and \@ref(fig:spottedHOwx)
 present the predicted haul-out probability with respect to age and sex class
@@
   theme(legend.title = element_blank(),
-        legend.position = "bottom") +
+        legend.position = "none") +
   xlab("total wind (m/s)") + ylab("haul-out probability")
@@
     subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
-    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate marginal patterns.')
+    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate underlying trends.')
 ```

### 2022-01-13 R/create_newdata.R
IMPLEMENT NEW WX COVAR PLOT DESIGN

@@ -84,10 +84,10 @@ create_newdata_margins <- function(data, age_sex,
       solar_hour = solar_hour,
       yday = yday,
       northing = mean(data$northing),
-      temp2 = if(term %in% c("temp2")) sample(unique(data[,"temp2"]),500) else mean(data[,"temp2"]),
-      wind = if(term %in% c("wind")) sample(unique(data[,"wind"]),500) else mean(data[,"wind"]),
-      pressure = if(term %in% c("pressure")) sample(unique(data[,"pressure"]),500) else mean(data[,"pressure"]),
-      precip = if(term %in% c("precip")) sample(unique(data[,"precip"]),500) else mean(data[,"precip"])
+      temp2 = if(term %in% c("temp2")) sample(unique(data[,"temp2"]),150) else mean(data[,"temp2"]),
+      wind = if(term %in% c("wind")) sample(unique(data[,"wind"]),150) else mean(data[,"wind"]),
+      pressure = if(term %in% c("pressure")) sample(unique(data[,"pressure"]),150) else mean(data[,"pressure"]),
+      precip = if(term %in% c("precip")) sample(unique(data[,"precip"]),150) else mean(data[,"precip"])
       ) %>%
       mutate(
         sin1 = sin(pi * solar_hour / 12),
@@ -144,15 +144,16 @@ create_ribbon_newdata <- function(fit_ribbon, margins = FALSE, solar_hour, yday,
     )
 }
 
-create_spotted_newdata <- function(fit_spotted, margins = FALSE, ...) {
-    if(margins) {
-      spotted_newdata <- create_newdata_margins(fit_spotted$dataset,
-                                               age_sex = levels(fit_spotted$dataset$age_sex),
-                                               terms)
-    } else {
-      spotted_newdata <- create_newdata(fit_spotted$dataset,
-                                       age_sex = levels(fit_spotted$dataset$age_sex))
-    }
+create_spotted_newdata <- function(fit_spotted, margins = FALSE, solar_hour, yday, term) {
+  if(margins) {
+    spotted_newdata <- create_newdata_margins(fit_spotted$dataset,
+                                             age_sex = levels(fit_spotted$dataset$age_sex),
+                                             solar_hour = solar_hour, yday = yday,
+                                             term)
+  } else {
+    spotted_newdata <- create_newdata(fit_spotted$dataset,
+                                     age_sex = levels(fit_spotted$dataset$age_sex))
+  }
 
   # create the model matrix
[…]

### 2022-01-13 R/wx_plots.R

@@ -1,0 +1,292 @@
+plot_ribbon_wx <- function(ribbon_fit_obj) {
+  
+  p_temp <- create_ribbon_newdata(ribbon_fit_obj, 
+                                  margins=TRUE, 
+                                  solar_hour = 12, yday = 135, 
+                                  term = "temp2") %>% 
+    mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+           month = lubridate::month(date,label = TRUE),
+           day = lubridate::day(date)) %>% 
+    filter(age_sex != "YOUNG OF YEAR") %>% 
+    group_by(age_sex) %>% 
+    
+    ggplot(aes(temp2*27, ho_prob, color = age_sex)) +
+    geom_pointrange(aes(ymin = lower95, ymax = upper95), 
+                    size = 0.25, alpha = 1, stroke = 0) +
+    scale_color_manual(values= met.brewer("Egypt",n=3),
+                       guide = guide_legend(override.aes = list(alpha = 1),
+                                            direction = "horizontal")) +
+    theme(legend.title = element_blank(),
+          legend.position = "none") +
+    xlab("temperature (C)") + ylab("haul-out probability")
+  
+  p_wind <- create_ribbon_newdata(ribbon_fit_obj, 
+                                  margins=TRUE, 
+                                  solar_hour = 12, yday = 135, 
+                                  term = "wind") %>%
+    mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+           month = lubridate::month(date,label = TRUE),
+           day = lubridate::day(date)) %>% 
+    filter(age_sex != "YOUNG OF YEAR") %>% 
+    group_by(age_sex) %>% 
+    
+    ggplot(aes(wind*10, ho_prob, color = age_sex)) +
+    geom_pointrange(aes(ymin = lower95, ymax = upper95), 
+                    size = 0.25, alpha = 1, stroke = 0) +
+    scale_color_manual(values=met.brewer("Egypt",n=3),
+                       guide = guide_legend(override.aes = list(alpha = 1),
+                                            direction = "horizontal")) +
+    theme(legend.title = element_blank(),
+          legend.position = "top") +
+    xlab("wind (m/s)") + ylab("haul-out probability")
+  
+  p_pressure <- create_ribbon_newdata(ribbon_fit_obj, margins=TRUE, 
+                                      solar_hour = 12, yday = 135, 
[…]

### 2022-01-13 _targets.R

@@ -25,7 +25,9 @@ tar_option_set(packages = c(
   "solaR",
   "splines",
   "purler",
-  "wcUtils"
+  "wcUtils",
+  "MetBrewer",
+  "patchwork"
   )
 )
 
@@ -70,6 +72,10 @@ list(
   tar_target(ribbon_newdata, create_ribbon_newdata(ribbon_fit)),
   tar_target(spotted_newdata, create_spotted_newdata(spotted_fit)),
   tar_target(bearded_newdata, create_bearded_newdata(bearded_fit)),
+  
+  tar_target(ribbon_wx_plot, plot_ribbon_wx(ribbon_fit)),
+  tar_target(spotted_wx_plot, plot_spotted_wx(spotted_fit)),
+  tar_target(bearded_wx_plot, plot_bearded_wx(bearded_fit)),
 
   tar_target(ribbon_newdata_year, ribbon_newdata_yr(ribbon_year_fit)),
   tar_target(spotted_newdata_year, spotted_newdata_yr(spotted_year_fit))

### 2022-01-13 analysis/paper/London_HauloutBeahvior.Rmd
COLOR PALETTE CHANGES
more use of metbrewer colorblind palettes and changes to better indicate overall reduced haul-out prob of bearded seals while still showing pattern

@@
 library(stickylabeller)
+library(MetBrewer)
 
@@
         geom_tile()+
-        scale_fill_viridis_c(option="mako") +
+        scale_fill_gradientn(colors = rev(met.brewer("Hokusai2"))) +
+        # scale_fill_viridis_c(option="mako") +
         scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
@@
   ) +
-  scale_fill_distiller(type = "seq", palette = "YlOrRd", direction = 1,
+  scale_fill_gradientn(colors = rev(met.brewer("Greek")),
                     trans = "log10", aesthetics = "fill",
@@
 
-```{r beardedHOCal, fig.height = 3.5, include=TRUE, fig.id = "beardedHOCal", fig.cap="Predicted haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes."}
+```{r beardedHOCal, fig.height = 4.5, include=TRUE, fig.id = "beardedHOCal", fig.cap="Predicted haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes."}
 
@@
   geom_tile(color= "white",size = 0) +
-  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
+  scale_fill_gradientn(colors = rev(met.brewer("Hiroshige")),
+                       name = "haul-out probability", aesthetics = "fill",
+                       limits = c(0,1), breaks = c(0.25, 0.50, 0.75),
                        guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                               barheight = 0.5, title.hjust = 0.5)
+  # scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
+  #                      guide = guide_colorbar(title.position = "bottom", barwidth = 15,
+  #                                             barheight = 0.5, title.hjust = 0.5)
   )
@@
 
-plot_df <- bearded_newdata %>%
-  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
-         month = lubridate::month(date,label = TRUE),
-         day = lubridate::day(date)) %>% 
-  filter(age_sex != "YOUNG OF YEAR",
-         between(solar_hour, 8,16)) %>% 
-  group_by(age_sex)
-
-plot_df_early <- plot_df %>% tidyr::nest() %>% 
-  rowwise() %>% 
-  mutate(peak_idx = which.max(data$ho_prob),
-         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
-  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
-  rowwise() %>% 
-  mutate(data = list(data %>% filter(between(yday,(peak_avg-25)-3, (peak_avg-25)+3))),
-         season = "early") %>% 
-  unnest(cols = c(data))
-
-plot_df_peak <- plot_df %>% tidyr::nest() %>% 
-  rowwise() %>% 
-  mutate(peak_idx = which.max(data$ho_prob),
-         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
-  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
-  rowwise() %>% 
-  mutate(data = list(data %>% filter(between(yday,(peak_avg)-3, (peak_avg)+3))),
-         season = "peak") %>% 
-  unnest(cols = c(data))
-
-plot_df_late <- plot_df %>% tidyr::nest() %>% 
-  rowwise() %>% 
-  mutate(peak_idx = which.max(data$ho_prob),
-         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
-  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
-  rowwise() %>% 
-  mutate(data = list(data %>% filter(between(yday,(peak_avg+25)-3, (peak_avg+25)+3))),
-         season = "late") %>% 
-  unnest(cols = c(data))
-
-plot_df <- bind_rows(list(plot_df_early, plot_df_peak, plot_df_late)) %>% 
-  mutate(date_start = case_when(
-    season == "peak" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 3, "%d %b"),
-    season == "early" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 28, "%d %b"),
-    season == "late" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 22, "%d %b")
-  )) %>% 
-  mutate(date_end = case_when(
-    season == "peak" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 3, "%d %b"),
-    season == "early" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 22, "%d %b"),
-    season == "late" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 28, "%d %b")
-  )) %>% 
-  mutate(season = forcats::fct_relevel(season, "early", "peak", "late"),
-         season = forcats::fct_recode(season, "peak haul out - 25 days" = "early",
-                             "peak haul out" = "peak",
-                             "peak haul out + 25 days" = "late"))
-
-p_temp <- ggplot(plot_df, 
-                 aes(temp2*27, ho_prob, color = age_sex, fill = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
-  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
-              show.legend = FALSE) +
-  scale_fill_d3() +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1),
-                                      direction = "horizontal")) +
-  theme(legend.title = element_blank(),
-        legend.position = "none") +
-  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
-             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
-  xlab("temperature (C)") + ylab("haul-out probability")
-
-p_wind <- ggplot(plot_df, aes(wind*10, ho_prob, color = age_sex, fill = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
-  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
-              show.legend = FALSE) +
-  scale_fill_d3() +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 0.75),
-                                      direction = "horizontal")) +
-  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
-             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
-  theme(legend.title = element_blank(),
-        legend.position = "none") +
-  xlab("total wind (m/s)") + ylab("haul-out probability")
-
-library(patchwork)
-
-p_temp / p_wind +
-  plot_annotation(
-    title = 'The influence of temperature and wind on predicted bearded seal \nhaul-out behavior varies within season',
-    subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
-    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate underlying trends.')
+targets::tar_load(bearded_wx_plot, store = here::here("_targets"))
+
+bearded_wx_plot
 ```
@@
   geom_tile(color = "white",size = 0) +
-  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
+  scale_fill_gradientn(colors = rev(met.brewer("Hiroshige")),
+                       name = "haul-out probability", aesthetics = "fill",
+                       limits = c(0,1), breaks = c(0.25, 0.50, 0.75),
                        guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                               barheight = 0.5, title.hjust = 0.5)
+  # scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
+  #                      guide = guide_colorbar(title.position = "bottom", barwidth = 15,
+  #                                             barheight = 0.5, title.hjust = 0.5)
   )
@@
 
-plot_df <- ribbon_newdata %>%
-  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
-         month = lubridate::month(date,label = TRUE),
-         day = lubridate::day(date)) %>% 
-  filter(age_sex != "YOUNG OF YEAR",
-         between(solar_hour, 8,16)) %>% 
-  group_by(age_sex)
-
-plot_df_early <- plot_df %>% tidyr::nest() %>% 
-  rowwise() %>% 
-  mutate(peak_idx = which.max(data$ho_prob),
-         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
-  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
-  rowwise() %>% 
-  mutate(data = list(data %>% filter(between(yday,(peak_avg-25)-3, (peak_avg-25)+3))),
-         season = "early") %>% 
-  unnest(cols = c(data))
-
-plot_df_peak <- plot_df %>% tidyr::nest() %>% 
-  rowwise() %>% 
-  mutate(peak_idx = which.max(data$ho_prob),
-         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
-  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
-  rowwise() %>% 
-  mutate(data = list(data %>% filter(between(yday,(peak_avg)-3, (peak_avg)+3))),
-         season = "peak") %>% 
-  unnest(cols = c(data))
-
-plot_df_late <- plot_df %>% tidyr::nest() %>% 
-  rowwise() %>% 
-  mutate(peak_idx = which.max(data$ho_prob),
-         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
-  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
-  rowwise() %>% 
-  mutate(data = list(data %>% filter(between(yday,(peak_avg+25)-3, (peak_avg+25)+3))),
-         season = "late") %>% 
-  unnest(cols = c(data))
-
-plot_df <- bind_rows(list(plot_df_early, plot_df_peak, plot_df_late)) %>% 
-  mutate(date_start = case_when(
-    season == "peak" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 3, "%d %b"),
-    season == "early" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 28, "%d %b"),
-    season == "late" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 22, "%d %b")
-  )) %>% 
-  mutate(date_end = case_when(
-    season == "peak" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 3, "%d %b"),
-    season == "early" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 22, "%d %b"),
-    season == "late" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 28, "%d %b")
-  )) %>% 
-  mutate(season = forcats::fct_relevel(season, "early", "peak", "late"),
-         season = forcats::fct_recode(season, "peak haul out - 25 days" = "early",
-                             "peak haul out" = "peak",
-                             "peak haul out + 25 days" = "late"))
-
-p_temp <- ggplot(plot_df, 
-                 aes(temp2*27, ho_prob, color = age_sex, fill = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
-  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
-              show.legend = FALSE) +
-  scale_fill_d3() +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1),
-                                      direction = "horizontal")) +
-  theme(legend.title = element_blank(),
-        legend.position = "none") +
-  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
-             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
-  xlab("temperature (C)") + ylab("haul-out probability")
-
-p_wind <- ggplot(plot_df, aes(wind*10, ho_prob, color = age_sex, fill = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
-  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
-              show.legend = FALSE) +
-  scale_fill_d3() +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 0.75),
-                                      direction = "horizontal")) +
-  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
-             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
-  theme(legend.title = element_blank(),
-        legend.position = "bottom") +
-  xlab("total wind (m/s)") + ylab("haul-out probability")
-
-library(patchwork)
-
-p_temp / p_wind +
-  plot_annotation(
-    title = 'The influence of temperature and wind on predicted ribbon seal \nhaul-out behavior varies within season',
-    subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
-    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate underlying trends.')
+targets::tar_load(ribbon_wx_plot, store = here::here("_targets"))
+
+ribbon_wx_plot
 ```
@@
 
-plot_df <- spotted_newdata %>%
-  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
-         month = lubridate::month(date,label = TRUE),
-         day = lubridate::day(date)) %>% 
-  filter(age_sex != "YOUNG OF YEAR",
-         between(solar_hour, 8,16)) %>% 
-  group_by(age_sex)
-
-plot_df_early <- plot_df %>% tidyr::nest() %>% 
-  rowwise() %>% 
-  mutate(peak_idx = which.max(data$ho_prob),
-         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
-  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
-  rowwise() %>% 
-  mutate(data = list(data %>% filter(between(yday,(peak_avg-25)-3, (peak_avg-25)+3))),
-         season = "early") %>% 
-  unnest(cols = c(data))
-
-plot_df_peak <- plot_df %>% tidyr::nest() %>% 
-  rowwise() %>% 
-  mutate(peak_idx = which.max(data$ho_prob),
-         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
-  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
-  rowwise() %>% 
-  mutate(data = list(data %>% filter(between(yday,(peak_avg)-3, (peak_avg)+3))),
-         season = "peak") %>% 
-  unnest(cols = c(data))
-
-plot_df_late <- plot_df %>% tidyr::nest() %>% 
-  rowwise() %>% 
-  mutate(peak_idx = which.max(data$ho_prob),
-         peak_day = data$yday[peak_idx]) %>% ungroup() %>% 
-  mutate(peak_avg = as.integer(mean(peak_day))) %>% 
-  rowwise() %>% 
-  mutate(data = list(data %>% filter(between(yday,(peak_avg+25)-3, (peak_avg+25)+3))),
-         season = "late") %>% 
-  unnest(cols = c(data))
-
-plot_df <- bind_rows(list(plot_df_early, plot_df_peak, plot_df_late)) %>% 
-  mutate(date_start = case_when(
-    season == "peak" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 3, "%d %b"),
-    season == "early" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 28, "%d %b"),
-    season == "late" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 22, "%d %b")
-  )) %>% 
-  mutate(date_end = case_when(
-    season == "peak" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 3, "%d %b"),
-    season == "early" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") - 22, "%d %b"),
-    season == "late" ~ format(lubridate::as_date(peak_avg, origin = "2015-01-01") + 28, "%d %b")
-  )) %>% 
-  mutate(season = forcats::fct_relevel(season, "early", "peak", "late"),
-         season = forcats::fct_recode(season, "peak haul out - 25 days" = "early",
-                             "peak haul out" = "peak",
-                             "peak haul out + 25 days" = "late"))
-
-p_temp <- ggplot(plot_df, 
-                 aes(temp2*27, ho_prob, color = age_sex, fill = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
-  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
-              show.legend = FALSE) +
-  scale_fill_d3() +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 1),
-                                      direction = "horizontal")) +
-  theme(legend.title = element_blank(),
-        legend.position = "none") +
-  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
-             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
-  xlab("temperature (C)") + ylab("haul-out probability")
-
-p_wind <- ggplot(plot_df, aes(wind*10, ho_prob, color = age_sex, fill = age_sex)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
-  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
-              show.legend = FALSE) +
-  scale_fill_d3() +
-  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 0.75),
-                                      direction = "horizontal")) +
-  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
-             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
-  theme(legend.title = element_blank(),
-        legend.position = "bottom") +
-  xlab("total wind (m/s)") + ylab("haul-out probability")
-
-library(patchwork)
-
-p_temp / p_wind +
-  plot_annotation(
-    title = 'The influence of temperature and wind on predicted spotted seal \nhaul-out behavior varies within season',
-    subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
-    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate underlying trends.')
+targets::tar_load(spotted_wx_plot, store = here::here("_targets"))
+
+spotted_wx_plot
 ```

### 2022-01-13 analysis/paper/London_HauloutBeahvior.Rmd
WEATHER COVARIATE PLOTS/CAPTIONS FIXED

@@
   geom_tile(color= "white",size=0) + 
-  scale_fill_distiller(palette = "YlGnBu",
+  scale_fill_gradientn(colors = rev(met.brewer("Hiroshige")),
                        name = "hourly percent dry", aesthetics = "fill",
@@
 haul-out probability, averaged over the other weather conditions, is more
-variable than simply showing model coefficients would imply. To elucidate
-underlying patterns between the model's predicted haul-out probability and
-specific weather covariates, we explored a subset of predictions during three
-separate time windows. The same set of predictions used to create the haul-out
-probability surfaces was reduced to include only local solar hours between 06:00
-and 18:00 during three 7-day time windows. The time windows were centered on the
-day of maximum haul-out probability and 25 days before and 25 days after peak
-haul out. While this approach does not entirely remove patterns of collinearity
-from the visualization (e.g. temperature increases with day of year), by
-limiting our range of days and hours we gain some insight into how haul-out
-probability varies across different weather covariate values. Figures
-\@ref(fig:beardedHOwx), \@ref(fig:ribbonHOwx), and \@ref(fig:spottedHOwx)
-present the predicted haul-out probability with respect to age and sex class
-across the range of temperature and wind conditions encountered in the three
-time windows. The visualizations also include vertical lines representing 95%
-confidence intervals around the predicted haul-out probability to better
-communicate the variation in model uncertainty. Additionally, simple linear fits
-are shown to assist interpretation of overall patterns.
+variable than simply showing model coefficients would imply. That said,
+important insights can be gained from plots of marginal effects. To create these
+plots (figures \@ref(fig:beardedHOwx), \@ref(fig:ribbonHOwx), and
+\@ref(fig:spottedHOwx)), we predicted haul-out probability across the full range
+of weather covariate values while fixing hour of the day at local solar noon and
+day-of-year at the average day of peak haul-out for each species. The
+visualizations also include vertical lines representing 95% confidence intervals
+around the predicted haul-out probability to better communicate the variation in
+model uncertainty.
 
@@
 Age and sex class were not included in the model for bearded seals due to our
-lower sample size for adult and young-of-year age classes. As such,
-results are shown for all ages. Bearded seals had a bi-modal
-distribution of haul-out probability across the day (Figure
+lower sample size for adult and young-of-year age classes. As such, results are
+shown for all ages. Overall, bearded seals were less likely to haul out and had
+a bi-modal distribution of haul-out probability across the day (Figure
 \@ref(fig:beardedHOCal)). In addition to a peak around local solar noon, the
 bearded seal model predicts additional haul-out activity around local midnight.
-The haul-out behavior is also more protracted throughout the spring season
-compared to ribbon and spotted seals.
+In concert with the lower magnitude of haul-out probability, bearded seal
+haul-out behavior is also more protracted throughout the spring season compared
+to ribbon and spotted seals.
 
@@
 
-```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across the range of weather conditions encountered in the observed data.
+```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
+Hour of the day was fixed at local solar noon and day-of-year held at the day of peak haul-out for bearded seals.
 Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability."}
@@
 
-```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Variability in predicted haul-out probability of ribbon seals across the range of weather conditions encountered in the observed data.
+```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification.
+Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for ribbon seals.
 Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability."}
@@
 
-```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Variability in the predicted haul-out probability of spotted seals across the range of weather conditions encountered in the observed data.
+```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
+Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for spotted seals.
 Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability."}

### 2022-01-16 analysis/paper/London_HauloutBeahvior.Rmd
INITIAL CONVERSION TO PDF / PEERJ

@@
 ---
-title: "Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas"
+title: Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas
+preprint: false
 author: 
-  - Josh M. London:
+  - name: Josh M. London
     email: josh.london@noaa.gov
-      institute: [MML]
-      correspondence: true
-  - Paul B. Conn:
-      institute: [MML]
-      correspondence: false
-  - Stacie K. Hardy:
-      institute: [MML]
-      correspondence: false
-  - Erin L. Richmond:
-      institute: [MML]
-      correspondence: false
-  - Jay M. Ver Hoef:
-      institute: [MML]
-      correspondence: false
-  - Michael F. Cameron:
-      institute: [MML]
-      correspondence: false
-  - Justin Crawford:
-      institute: [ADFG]
-      correspondence: false
-  - Andrew L. Von Duyke:
-      institute: [NSB]
-      correspondence: false
-  - Lori T. Quakenbush:
-      institute: [ADFG]
-      correspondence: false
-  - Peter L. Boveng:
-      institute: [MML]
-      correspondence: false
-institute:
-  - MML: Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
-  - ADFG: Arctic Marine Mammals Program, Alaska Department of Fish and Game, Fairbanks, Alaska, USA
-  - NSB: Department of Wildlife Management, North Slope Borough, Utqiaġvik, Alaska, USA
-output: 
-  officedown::rdocx_document:
-    reference_docx: ../templates/jml_template.docx
-    pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-    page_size:
-      width: 8.5
-      height: 11
-      orient: "portrait"
-    page_margins:
-      bottom: 1.25
-      top: 1.25
-      right: 1.0
-      left: 0.5
-      header: 0.5
-      footer: 0.5
-      gutter: 0.5
-bibliography: references.json
-link-citations: true
-csl: ../templates/proceedings-of-the-royal-society-b.csl
-always_allow_html: true
----
-
-## Abstract
-
+    affiliation: 1
+    corresponding: true
+  - name: Paul B. Conn
+    affiliation: 1
+  - name: Stacie K. Hardy
+    affiliation: 1
+  - name: Erin L. Richmond
+    affiliation: 1
+  - name: Jay M. Ver Hoef
+    affiliation: 1
+  - name: Michael F. Cameron
+    affiliation: 1
+  - name: Justin Crawford
+    affiliation: 2
+  - name: Andrew L. Von Duyke
+    affiliation: 3
+  - name: Lori T. Quakenbush
+    affiliation: 2
+  - name: Peter L. Boveng
+    affiliation: 1
+affiliation:
+  - code: 1
+    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
+  - code: 2
+    address: Arctic Marine Mammals Program, Alaska Department of Fish and Game, Fairbanks, Alaska, USA
+  - code: 3
+    address: Department of Wildlife Management, North Slope Borough, Utqiaġvik, Alaska, USA
+abstract: >
   Ice-associated seals rely on sea ice for a variety of activities, including
@@
   aerial surveys.
-
-keywords: availability; generalized linear mixed pseudo-model; haul-out
-behavior; phenology; Phocidae; sea ice
+output:
+  bookdown::pdf_book:
+    base_format: rticles::peerj_article # for using bookdown features like \@ref()
+header-includes:
+  - \usepackage [T1]{fontenc}
+  - \usepackage {newpxtext}
+  - \usepackage {newpxmath}
+  - \usepackage [semibold]{sourcesanspro}
+  - \usepackage {sectsty}
+  - \allsectionsfont {\sffamily}
+  - \usepackage {hyperref}
+  - \hypersetup {colorlinks, linkcolor={red!50!black}, citecolor={blue!50!black},urlcolor={blue!80!black}}
+  - \linespread{1.1}
+bibliography: references.json
+csl: peerj.csl
+---
 
@@
 
-knitr::opts_chunk$set(
-  tab.cap.style = "Table Caption",
-  tab.cap.pre = "Table ",
-  tab.cap.sep = ": "
-  )
-
-knitr::opts_chunk$set(
-  fig.cap.style = "Image Caption",
-  fig.cap.pre = "Figure ",
-  fig.cap.sep = ": "
-  )
+# knitr::opts_chunk$set(
+#   tab.cap.style = "Table Caption",
+#   tab.cap.pre = "Table ",
+#   tab.cap.sep = ": "
+#   )
+# 
+# knitr::opts_chunk$set(
+#   fig.cap.style = "Image Caption",
+#   fig.cap.pre = "Figure ",
+#   fig.cap.sep = ": "
+#   )
 
@@
 
-# Introduction
+# Introduction {-}
 
@@
 
-# Methods
-
-## Data collection
+# Methods {-}
+
+## Data collection {-}
 
@@
 
-```{r tblDeploy, include=TRUE, tab.id="tblDeploy",tab.cap = "Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2020.
+```{r tblDeploy, include=TRUE, tab.cap = "Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2020.
 Total seal hours represents the sum of available data across all seals.
@@
   merge_h(part = "header") %>% 
-  align(align = "center", part = "header") %>% 
-  font(fontname = "Source Serif Pro Light", part = "all") %>% 
-  fontsize(size = 10) %>% 
+  align(align = "left", part = "header") %>% 
+  # font(fontname = "Source Serif Pro Light", part = "all") %>% 
+  fontsize(size = 10, part = "all") %>% 
   autofit()
@@
 
-```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
+```{r dataCal, include=TRUE, fig.height = 4, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
 Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020.
@@
 
-## Explanatory variables
+## Explanatory variables {-}
 
@@
 
-## Haul-out modeling
+## Haul-out modeling {-}
 
@@
 $$ 
-where $H_t$ gives the effect for solar hour $t$ and $α_{i}$ are
+where $H_t$ gives the effect for solar hour $t$ and $\alpha_{i}$ are
 estimated parameters (regression coefficients).
@@
 
-# Results
+# Results {-}
 
@@
 
-## Bearded Seals
+## Bearded Seals {-}
 
@@
 
-## Ribbon Seals
+## Ribbon Seals {-}
 
@@
 
-## Spotted Seals
+## Spotted Seals {-}
 
@@
 
-## Annual variation in haul-out
+## Annual variation in haul-out timing {-}
 
@@
 
-# Discussion
+# Discussion {-}
 
@@
 
-# Author Contributions
+# Author Contributions {-}
 
@@
 
-# Data Availability
+# Data Availability {-}
 
@@
 
-# Acknowledgements
+# Acknowledgements {-}
 
@@
 
-# References
+# References {-}
 

### 2022-01-17 analysis/paper/London_HauloutBeahvior.Rmd

@@
   - \usepackage [semibold]{sourcesanspro}
-  - \usepackage {sectsty}
-  - \allsectionsfont {\sffamily}
   - \usepackage {hyperref}
   - \hypersetup {colorlinks, linkcolor={red!50!black}, citecolor={blue!50!black},urlcolor={blue!80!black}}
-  - \linespread{1.1}
+  - \usepackage{setspace}
+  - \onehalfspacing
 bibliography: references.json
@@
 library(flextable)
+library(gt)
 library(patchwork)
@@
   fig.path = "../figures/",
-  dpi = 400,
-  fig.width = 6.5,
+  dpi = 300,
   out.width = "100%"
@@
 within regions and throughout their full ranges. Additionally, trends in
-abundance from large scale surveys of these species are unknown, so it is difficult to assess the
-effect, if any, declines in sea ice habitat have had, or will have, on seal
-densities.
+abundance from large scale surveys of these species are unknown, so it is
+difficult to assess the effect, if any, declines in sea ice habitat have had, or
+will have, on seal densities.
 
@@
 bio-logging data to demonstrate greater haul-out propensity in juvenile
-crabeater seals (*Lobodon carcinophaga*) than adults, with highest probabilities
+crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
 in February and at times close to solar noon. In the Arctic, Von Duyke et al.
@@
   align(align = "left", part = "header") %>% 
-  # font(fontname = "Source Serif Pro Light", part = "all") %>% 
+  align(i = 1, align = "center", part = "header") %>% 
   fontsize(size = 10, part = "all") %>% 
-  autofit()
+  width(width = 1.35) %>% 
+  width(j = ~ Species, width = 0.9) %>% 
+  width(j = ~ Sex, width = 0.4)
+
 ```
@@
   flextable() %>% 
-  font(fontname = "Source Serif Pro Light", part = "all") %>% 
-  fontsize(size = 10) %>% 
+  fontsize(size = 10, part = "all") %>% 
   hline(i=8) %>% 
-  set_table_properties(width = .95, layout = "autofit")
+  width(width = 1.0) %>% 
+  width(j = ~ Description, width = 2.5)
 ```
@@
 
-```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
+```{r beardedHOwx, fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
 Hour of the day was fixed at local solar noon and day-of-year held at the day of peak haul-out for bearded seals.
-Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability."}
+Transparent vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
 
@@
 
-```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOCal", fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model.
+```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model.
 Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males.
@@
 
-```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification.
+```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification.
 Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for ribbon seals.
-Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability."}
+Transparent vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
 
@@
 
-```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.id = "spottedHOCal", fig.cap="Predicted haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model.
+```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model.
 Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males.
@@
 
-```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
+```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
 Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for spotted seals.
-Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability."}
+Transparent vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
 
@@
 
-```{r annualHO, fig.height = 5.5, include=TRUE, fig.id = "annualHO", fig.cap="Annual variability in the timing of peak haul-out probability for ribbon and spotted seals across 10 years.
+```{r annualHO, fig.height = 5.5, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability for ribbon and spotted seals across 10 years.
 Predictions are shown for local solar noon and under smoothed weather conditions.
@@
 
-```{r maxextentHO, fig.height = 5.5, include=FALSE, fig.id = "maxextent", fig.cap="Annual timing of peak haul-out behavior analyzed against the yearly maximum area of sea ice extent in the study area.
+```{r maxextentHO, fig.height = 5.5, include=FALSE, fig.cap="Annual timing of peak haul-out behavior analyzed against the yearly maximum area of sea ice extent in the study area.
 While spotted seal adult females and subadults show a negative trend line, none of the relationships were determined to be significant."}
@@
 
-**Josh M. London**: investigation, conceptualization, methodology,
+* **Josh M. London**: investigation, conceptualization, methodology,
 formal analysis, validation, software, writing: original draft, writing:
 review and editing, visualization, and data curation
-
-**Paul B. Conn**: conceptualization, methodology, formal analysis,
+* **Paul B. Conn**: conceptualization, methodology, formal analysis,
 software, validation, writing: original draft, writing: review and
 editing
-
-**Stacie K. Hardy**: investigation, data curation, methodology,
+* **Stacie K. Hardy**: investigation, data curation, methodology,
 validation, writing: review and editing
-
-**Erin L. Richmond**: data curation, investigation, methodology,
+* **Erin L. Richmond**: data curation, investigation, methodology,
 validation, writing: review and editing
-
-**Jay M. Ver Hoef**: conceptualization, methodology, software, writing:
+* **Jay M. Ver Hoef**: conceptualization, methodology, software, writing:
 review and editing
-
-**Michael F. Cameron**: investigation, project administration, writing:
+* **Michael F. Cameron**: investigation, project administration, writing:
 review and editing
-
-**Justin Crawford**: investigation, methodology, validation, data curation, 
+* **Justin Crawford**: investigation, methodology, validation, data curation, 
 writing: review and editing
-
-**Lori T. Quakenbush**: investigation, methodology, supervision, project
+* **Lori T. Quakenbush**: investigation, methodology, supervision, project
 administration, writing: review and editing
-
-**Andrew L. Von Duyke**: investigation, methodology, validation, data curation,
+* **Andrew L. Von Duyke**: investigation, methodology, validation, data curation,
 writing: review and editing
-
-**Peter L. Boveng**: investigation, conceptualization, supervision,
+* **Peter L. Boveng**: investigation, conceptualization, supervision,
 project administration, writing: review and editing

### 2022-01-17 analysis/paper/London_HauloutBeahvior.Rmd
CHANGE FONT TO Times (via newtxtext)
ADDRESS LORI Q FINAL COMMENTS
lori's comments were largely related to references. still need to address improvements to description of weather covariate response as in #26

@@
 header-includes:
+  - \usepackage {newtxtext,newtxmath}
+  - \usepackage [semibold]{sourcesanspro}
   - \usepackage [T1]{fontenc}
-  - \usepackage {newpxtext}
-  - \usepackage {newpxmath}
-  - \usepackage [semibold]{sourcesanspro}
   - \usepackage {hyperref}
@@
 in the Bering and Chukchi seas as a platform for important life history
-functions, such as pupping, nursing, breeding, and molting [@boveng2009;
+functions, such as pupping, nursing, breeding behavior, and molting [@boveng2009;
 @boveng2013a; @cameron2010; @kelly2010]. Limited data and large knowledge gaps
@@
 Ringed seals (_Phoca hispida_) --- not included in this study --- bask on
-sea-ice but also within snow lairs built during the spring. The remoteness of
-the Bering and Chukchi seas mean direct scientific observation of seal behavior
-is impractical. Thus, bio-logging devices are especially useful tools for
-collecting key information on movement and haul-out behavior for these species.
+sea-ice but also within snow lairs. The remoteness of the Bering and Chukchi
+seas mean direct scientific observation of seal behavior is impractical. Thus,
+bio-logging devices are especially useful tools for collecting key information
+on movement and haul-out behavior for these species.
 
@@
 changes between diurnal and nocturnal haul-out behavior of ringed seals
-previously described by Kelly et al. [-@kelly2010] using VHF radio tags and
+previously described by Kelly and Quakenbush [-@kelly1990c] using VHF radio tags and
 direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
@@
 meteorological variables that have been shown to influence haul-out behavior of
-walruses [@reder2003; @udevitz2009]. Third, we assessed the annual variability
+walruses [@udevitz2009]. Third, we assessed the annual variability
 in haul-out timing and possible linkage to changes in the extent of seasonal sea
@@
 capture events occurred from May to October, generally between the coastal
-communities of Scammon Bay, Alaska, in the Bering Sea, and Utqiaġvik, Alaska,
-near the junction of the Chukchi and Beaufort seas, from 2005 to 2020. We refer
-readers to the primary literature for detailed capture and bio-logger attachment
-methods (Supplemental Material, Table).
+communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska in the
+Chukchi Sea, and Nuiqsut in the Beaufort Sea from 2005 to 2020. We refer readers
+to the primary literature for detailed capture and bio-logger attachment methods
+(Supplemental Material, Table).
 
@@
 bias the hourly percent-dry values reported by the bio-logger. For this study,
-in cases where both bio-logger types were deployed, hourly percent dry
+in cases where both bio-logger types were deployed, hourly percent-dry
 observations from the flipper tag were preferred.
@@
 sexually immature _subadults_, and mature _adults_) were estimated at the time
-of deployment by various combinations of length, claw growth ridges [@burns1969;
-@burns1983; @mclaren1958a], and pelage characteristics for some species. In the
-case of ribbon seals, subadults often have less distinct ribbons than adults.
-Bearded seal subadults will often have a spotted pattern in the pelage that is
-not seen in adults. Spotted seal pelage cannot be used to reliably discern
-adults from subadults. Despite these challenges, we feel the age classifications
-used in this analysis are useful in testing for age-related effects on haul-out
+of deployment by various combinations of length, claw growth ridges
+[@mclaren1958a], and pelage characteristics for some species. In the case of
+ribbon seals, subadults often have less distinct ribbons than adults. Bearded
+seal subadults will often have a spotted pattern in the pelage that is not seen
+in adults. Spotted seal pelage cannot be used to reliably discern adults from
+subadults. Despite these challenges, we feel the age classifications used in
+this analysis are useful in testing for age-related effects on haul-out
 behavior. Seals determined to be less than one year were classified as
@@
 hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). On board the
-bio-logger, hourly percent-dry data were rounded to the nearest 10 percent
-inclusive of 0 and 100 percent and additional values at 3 and 98 percent. This
+bio-logger, hourly percent-dry data were rounded to the nearest 10%
+inclusive of 0% and 100% and additional values at 3% and 98%. This
 compression resulted in additional data transmission as each message consisted
@@
 
-```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, and May.
+```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployed on a ribbon seal over two years during the months of March, April, and May.
 Areas of the plot indicated with an 'X' represent missing data that were not successfully transmitted from the tag."}
@@
   ggtitle("Observation records are incomplete because not all data are transmitted",
-          subtitle = "percent-dry data for a single ribbon seal across two years") +
+          subtitle = "percent-dry actogram for a single ribbon seal across two years") +
   xlab("day of month") + ylab("hour (UTC)") 
@@
 weather covariates that vary with time and animal location. We explored the use
-of a continuous-time correlated random walk [@devins.johnson2008] movement model
+of a continuous-time correlated random walk [@johnson2008a] movement model
 to predict locations at specific times. However, the sparse nature of data from
@@
 behavior along a latitudinal gradient. We omitted the latitudinal effect from
-ribbon and spotted seal models since, during the spring, these species are most
+ribbon and spotted seal models because, during the spring, these species are most
 prevalent near the southern ice edge in the Bering Sea [@conn2014a].
@@
 Maximum spring sea ice extent was simply the largest daily count of grid cells
-with greater than 15 percent concentration for each year.
+with greater than 15% concentration for each year.
 
@@
   "Precip", "Continuous", "NARR", 
-  "convective precipitation (kg/m2)",
+  "convective precipitation (kg/m^2^)",
   "Pressure", "Continuous", "NARR", 
@@
 `r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
-observations fell between 10% and 90% hourly percent dry.
+observations fell between 10% and 90% hourly percent-dry.
 
@@
 
-```{r beardedHOCal, fig.height = 4.5, include=TRUE, fig.id = "beardedHOCal", fig.cap="Predicted haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes."}
+```{r beardedHOCal, fig.height = 4.5, include=TRUE, fig.cap="Predicted haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes."}
 
@@
 some years where the pattern shown is informed by a small number of
-individuals that may not represent population level processes.
+individuals that may not represent population level patterns.
 
@@
 Only those groups (age-sex + year) that included observations from more than one seal are shown.
-Additionally, any groups where the underlying data started after 1 June or ended before 1 May are not included."}
+Additionally, any groups where data were only available after 1 June or before 1 May are not included."}
 
@@
 ecological behavior. Spotted seals are known to form triads during the breeding
-season [@burns1973] where a female and dependent pup are accompanied on the ice
-by a suitor male. The male waits for the female to wean the pup and enter
+season where a female and dependent pup are accompanied on the ice by a suitor
+male [@frost2018]. The male waits for the female to wean the pup and enter
 estrus, and fends off any other potential suitor males. Triad formation results
 in both males and females spending a large portion of the day hauled out on ice
-and a protracted spring haul-out season for both sexes. Because mating occurs in
-the water, females may also be less inclined to interrupt their haul-out with
-foraging trips while still nursing the pup. We see this reflected in the
-predicted haul-out behavior, with both males and females exhibiting a broad
-distribution of time out of the water throughout the solar day and the
-season.Ribbon seals are not known to form triads and our model predicts a
-progression of increased haul-out behavior with females starting earlier in the
-season than males. Notably, female ribbon seals spend a large portion of the day
-in the water during the pupping period, aligning with the hypothesis that ribbon
-seal females continue foraging while nursing.  In the case of both ribbon and
-spotted seals, subadults are the first to begin consistent haul-out behavior and
-follow a typical phocid pattern where subadults molt first as they do not have
-any reproductive constraints.
+and a protracted spring haul-out season for both sexes. While females are still
+nursing the pup and not yet in estrus they may be less inclined to interrupt
+their haul-out and enter the water where the suitor male would attempt mating.
+We see this reflected in the predicted haul-out behavior, with both males and
+females exhibiting a broad distribution of time out of the water throughout the
+solar day and the season. Ribbon seals are not known to form triads and our
+model predicts a progression of increased haul-out behavior with females
+starting earlier in the season than males. Notably, female ribbon seals spend a
+large portion of the day in the water during the pupping period, aligning with
+the hypothesis that ribbon seal females continue foraging while nursing.
+Subadult ribbon and spotted seals begin consistent haul-out behavior earlier and
+follow a pattern seen in other phocids where yearlings and subadults molt first
+followed by adult females and males [@reder2003, @thompson1987, @kirkman2003].
 
@@
 which provides energetic benefits. Low winds and precipitation may also
-provide an optimal situation for predator detection. Our results highlight the
+provide better predator detection conditions. Our results highlight the
 importance of incorporating weather covariates when analyzing haul-out 
@@
 minimal threshold in the spatial extent or density of sea ice will have a
-meaningful impact on the timing of peak haul-out behavior -- if there is no sea
-ice, seals will not haul out. Additionally, while from an ecological perspective
-the haul-out behavior appears consistent, the interannual differences in timing
-and magnitude are large enough to have important ramifications on calculations
-of abundance and trend. Those ramifications will only be exacerbated if climate
-variability amplifies interannual differences.
+meaningful impact on the timing of peak haul-out behavior. Additionally, while
+from an ecological perspective the haul-out behavior appears consistent, the
+interannual differences in timing and magnitude are large enough to have
+important ramifications on calculations of abundance and trend. Those
+ramifications will only be exacerbated if climate variability amplifies
+interannual differences.
 
@@
 different than those previously reported for these species, especially for
-bearded seals. For instance, Ver Hoef et al. [-@verhoef2014a)] and Conn et al.
+bearded seals. For instance, Ver Hoef et al. [-@verhoef2014a] and Conn et al.
 [-@conn2014a] used haul-out correction factors with maximums of 0.66 for bearded
@@
 seals. Ringed seals are also present in the Bering and Chukchi seas but exhibit
-a unique complicating factor. Adult ringed seals build subnivean lairs in the
-snow on top of the sea ice, where they haul out and pup until snow melt causes
-their lairs to collapse [@kelly1990]. Thus, the wet-dry sensor on a bio-logger
+a unique complicating factor. Adult ringed seals build subnivean lairs under the
+snow on top of the sea ice, where they haul out and where females rear pups until
+conditions are good for basking [@frost2004]. Thus, the wet-dry sensor on a bio-logger
 could indicate that an animal is hauled out, but if it is within a lair, it is
@@
 deployment of bio-logging devices used in this study were often done in
-collaboration with Alaska Native seal hunters and their communities. We would
-like to especially acknowledge the communities of Kotzebue, Koyuk, Utqiaġvik,
-and Ulguniq (Wainwright) and the following individuals: James Adams, Jeff
-Barger, David Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris, Grover
-Harris, Lee Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin,
-John Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil
-Naylor Jr., Virgil Naylor Sr., Dan Savetilik, Allen Stone, and Randy Toshavik
-from Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Billy Adams, Mary-Ellen
-Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard Kittick,
-Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik, Shawn Oktollik,
-Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross Schaeffer, Bob Shears, and
-Joe Skin from the North Slope Borough, Alaska;
+collaboration with Alaska Native seal hunters near their communities. We would
+like to especially acknowledge the communities of Kotzebue, Koyuk, Nuiqsut,
+Utqiaġvik, and Ulguniq (Wainwright) and the following individuals: James Adams,
+Jeff Barger, David Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris,
+Grover Harris, Lee Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry
+Goodwin, John Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor,
+Virgil Naylor Jr., Virgil Naylor Sr., Dan Savetilik, Allen Stone, and Randy
+Toshavik from Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Billy Adams,
+Mary-Ellen Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman,
+Howard Kittick, Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik,
+Shawn Oktollik, Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross Schaeffer,
+Bob Shears, and Joe Skin from the North Slope Borough, Alaska;
 

### 2022-01-18 analysis/paper/London_HauloutBeahvior.Rmd
JUSTIN C FINAL COMMENTS ADDRESSED

@@
     affiliation: 1
-  - name: Justin Crawford
+  - name: Justin A. Crawford
     affiliation: 2
@@
   breeding, molting, pupping, and resting. In the Arctic, many of these activities
-  occur in spring (April-June) as sea ice begins to melt and retreat northward.
+  occur in spring (April -- June) as sea ice begins to melt and retreat northward.
   Rapid acceleration of climate change in Arctic ecosystems is therefore of
@@
 
-Haul-out behavior data from bio-loggers deployed on 
-`r dplyr::n_distinct(deploy_table$speno)` bearded, ribbon, and spotted seals were
+```{r deploystats}
+species_n <- deploy_table %>% 
+  group_by(species) %>% 
+  distinct(speno) %>% 
+  count(species)
+```
+
+Haul-out behavior data from 
+`r species_n %>% .$n %>% sum()` bio-loggers deployed on 
+`r species_n %>% filter(species == "Bearded seal") %>% .$n` bearded, 
+`r species_n %>% filter(species == "Ribbon seal") %>% .$n` ribbon, and 
+`r species_n %>% filter(species == "Spotted seal") %>% .$n` spotted seals were
 subset to include only records from 1 March to 15 July between 
@@
 ```{r tblDeploy, include=TRUE, tab.cap = "Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2020.
-Total seal hours represents the sum of available data across all seals.
+Total seal hours represents the sum of hourly observations across all seals used in the analysis.
 Because young-of-the-year are advanced to subadult on 1 March of the following year, some individual seals are represented in both columns in this table"}
@@
 the seal is hauled out. As such, end times of each deployment were identified by
-examining bio-logger locations and dive behavior to determine when bio-loggers
-ceased providing data consistent with seal behavior. Data outside of the
-deployment start and end times were discarded prior to analysis. After
-approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
-reported implausible hourly percent-dry data. All data after the first instance
-of unrealistic values were censored from this analysis. Figure
-\@ref(fig:dataCal) shows the distribution of all haul-out data
-across the study season for each species. Observations for ribbon and spotted
-seals are concentrated in the months of May and June due to the timing of
-deployment (April and May) and the timing of molt (May and June). During molt,
-seals (and their attached bio-loggers) spend more time out of the water and more
-data are transmitted. Molt timing also impacts when many deployments end as any
+examining bio-logger locations, percent-dry records, and dive behavior (if
+available) to determine when bio-loggers ceased providing data consistent with
+seal behavior. Data outside of the deployment start and end times were discarded
+prior to analysis. After approximately 9 months, 7 devices deployed on the rear
+flipper of bearded seals reported implausible hourly percent-dry data. All data
+after the first instance of unrealistic values were censored from this analysis.
+Figure \@ref(fig:dataCal) shows the distribution of all haul-out data across the
+study season for each species. Observations for ribbon and spotted seals are
+concentrated in the months of May and June due to the timing of deployment
+(April and May) and the timing of molt (May and June). During molt, seals (and
+their attached bio-loggers) spend more time out of the water and more data are
+transmitted. Molt timing also impacts when many deployments end as any
 bio-loggers adhered to the hair will fall off. Bearded seal deployments are more
@@
 seal deployments start later in the summer and by May bio-loggers have either
-fallen off or the batteries are depleted.
+fallen off or their batteries are depleted.
 
@@
 actual use of coastal features. Figure \@ref(fig:dataMap) shows the spatial
-distribution weighted locations with available haul-out behavior data for each
+distribution of weighted locations with available haul-out behavior data for each
 species across the study area.
@@
 
-```{r dataMap, fig.height = 3.75, fig.id="dataMap", include=TRUE, fig.cap="Spatial distribution of haul-out data during the months of March through July 15 for each of the three species.
-Data were collated across all years between 2005 and 2020."}
+(ref:mapCaption) Spatial distribution of haul-out data during the months of March through July 15 for each of the three species.
+Data were collated across all years between 2005 and 2020.
+Each hexagonal cell represents an area of 50 km^2^
+
+```{r dataMap, fig.height = 3.75, include=TRUE, fig.cap="(ref:mapCaption)"}
 
@@
   theme(legend.position = "bottom") +
-  ggtitle("Spatial distribution of haul-out behavior data")
+  ggtitle("Spatial distribution of haul-out behavior data") +
+  labs(caption = bquote("One hexagonal cell represents 50"~km^2))
 ```
@@
 using the {solaR} package [@perpinan2012] within the R statistical environment
-[@rcoreteam2021] using the daily locations. We included several weather
-variables shown to affect haul-out behavior in other Arctic pinnipeds
-[@reder2003; @udevitz2009; @perry2017]. In particular, we linked the weighted
-average daily locations from our bio-loggers to key weather values from the
-North American Regional Reanalysis (NARR) model produced by the National Centers
-for Environmental Prediction [@mesinger2006]. The NARR model assimilates
+[@rcoreteam2021] using the daily locations. We also linked the weighted average
+daily locations from our bio-loggers to key weather values from the North
+American Regional Reanalysis (NARR) model produced by the National Centers for
+Environmental Prediction [@mesinger2006]. The NARR model assimilates
 observational data to produce a long-term picture of weather over North America.
@@
 over the Bering and Chukchi seas at 3-hr intervals based on the native grid
-resolution of 32 km (1024 sq. km). The following meteorological variables were
-interpolated and assigned to seal locations using a bilinear method: 1) air
-temperature at 2m above the earth's surface, 2) wind consisting of northerly and
-easterly vector components converted to wind speed using the Euclidean norm, 3)
-barometric pressure at sea level, and 4) precipitation (Table
-\@ref(tab:tblCovariates)).
+resolution of 32 km (1024 sq. km). The following meteorological variables are
+known to affect haul-out behavior in other Arctic pinnipeds [@reder2003;
+@udevitz2009; @perry2017] and were interpolated and assigned to daily seal locations
+using a bilinear method: 1) air temperature at 2m above the earth's surface, 2)
+wind consisting of northerly and easterly vector components converted to wind
+speed using the Euclidean norm, 3) barometric pressure at sea level, and 4)
+precipitation (Table \@ref(tab:tblCovariates)).
 
@@
 
+(ref:kgm2) kg/m^2^
+
 ```{r tblCovariates, tab.id = "tblCovariates", include=TRUE, tab.cap = "Explanatory covariates used in analyses of binary haul-out records for bearded, spotted, and ribbon seals.
@@
   "Precip", "Continuous", "NARR", 
-  "convective precipitation (kg/m^2^)",
+  "convective precipitation (ref:kgm2)",
   "Pressure", "Continuous", "NARR", 
@@
 across the study region. Such predictions were then used to develop haul-out
-probability surfaces (see figures \@ref(beardedHOCal), \@ref(ribbonHOCal), and
-\@ref(spottedHOCal)), explore conditional effects of weather covariates, and
+probability surfaces (see figures \@ref(fig:beardedHOCal), \@ref(fig:ribbonHOCal), and
+\@ref(fig:spottedHOCal)), explore conditional effects of weather covariates, and
 determine annual peaks in haul-out activity. The timing of peak haul-out
@@
 
-```{r beardedHOCal, fig.height = 4.5, include=TRUE, fig.cap="Predicted haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes."}
+```{r beardedHOCal, fig.height = 4.5, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes.
+Hour of the day was fixed at local solar noon."}
 
@@
 
-Ribbon seals show a pattern of gradually increasing haul-out probability
+Ribbon seals exhibited a pattern of gradually increasing haul-out probability
 in April that peaks in late May for subadults and in early June for
@@
 
-```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model.
+```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted hourly haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model.
 Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males.
-'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}
+'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes.
+Hour of the day was fixed at local solar noon."}
 
@@
 
-```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model.
+```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted hourly haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model.
 Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males.
-'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}
+'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes.
+Hour of the day was fixed at local solar noon."}
 
@@
 generally centered around local solar noon, there were key differences across
-species, age, and sex that match expectations given what we know of their
+species, age, and sex that match our understandinf of their
 ecological behavior. Spotted seals are known to form triads during the breeding
@@
 nursing the pup and not yet in estrus they may be less inclined to interrupt
-their haul-out and enter the water where the suitor male would attempt mating.
+their haul out and enter the water where the suitor male would attempt mating.
 We see this reflected in the predicted haul-out behavior, with both males and
@@
 [@hamilton2018, @perry2017]. Ribbon seals seemed to be the most influenced by
-weather, with wind, temperature, barometric pressure, and precipitation all
+weather, with wind, temperature, and barometric pressure all
 being important components of the model. Spotted seals were most affected by
@@
 minimal threshold in the spatial extent or density of sea ice will have a
-meaningful impact on the timing of peak haul-out behavior. Additionally, while
-from an ecological perspective the haul-out behavior appears consistent, the
-interannual differences in timing and magnitude are large enough to have
-important ramifications on calculations of abundance and trend. Those
+meaningful impact on the timing of peak haul-out behavior --- if there is no sea
+ice, seals will not haul out or be forced to use terrestrial haul-out sites.
+Additionally, while from an ecological perspective the haul-out behavior appears
+consistent, the interannual differences in timing and magnitude are large enough
+to have important ramifications on calculations of abundance and trend. Those
 ramifications will only be exacerbated if climate variability amplifies
@@
 collaboration with Alaska Native seal hunters near their communities. We would
-like to especially acknowledge the communities of Kotzebue, Koyuk, Nuiqsut,
-Utqiaġvik, and Ulguniq (Wainwright) and the following individuals: James Adams,
-Jeff Barger, David Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris,
-Grover Harris, Lee Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry
-Goodwin, John Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor,
-Virgil Naylor Jr., Virgil Naylor Sr., Dan Savetilik, Allen Stone, and Randy
-Toshavik from Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Billy Adams,
+like to especially acknowledge the communities of Kotzebue, Koyuk, Nome,
+Nuiqsut, Scammon Bay, St. Michael, Utqiaġvik, and Ulguniq (Wainwright) and the
+following individuals: James Adams, Jeff Barger, David Barr, Wendell Booth,
+Cyrus Harris, Nereus 'Doc' Harris, Grover Harris, Lee Harris, Tom Jones, Frank
+Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin, Pearl Goodwin, Willie
+Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr., Virgil Naylor Sr., Dan
+Savetilik, Ross Schaeffer, Allen Stone, and Randy Toshavik from Kotzebue,
+Alaska; Merlin Henry from Koyuk, Alaska; Tom Gray from Nome; Vernon Long and
+Richard Tukle from Nuiqsuit, Alaska; Morgan Simon, River Simon, and Al Smith
+from Scammon Bay, Alaska; Alex Niksik Jr. from St. Michael, Alaska; Billy Adams,
 Mary-Ellen Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman,
 Howard Kittick, Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik,
-Shawn Oktollik, Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross Schaeffer,
-Bob Shears, and Joe Skin from the North Slope Borough, Alaska;
-
-We are grateful for the assistance in catching and sampling seals by
-James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir
-Burkanov, Cynthia Christman, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
-Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey
-Goldstein, Jeff Harris, Markus Horning, John Jansen, Shawn Johnson,
-Charles Littnan, Lloyd Lowry, Erin Moreland, Mark Nelson, Lorrie Rea, Gay Sheffield,
-Brent Stewart, and Dave Withrow. We also appreciate the commitment to
-science and safety by all officers and crew of the NOAA Ship Oscar
-Dyson, the NOAA Ship MacArthur II, and the RV Thomas G. Thompson.
+Shawn Oktollik, Stacey Osborn, Fred Rexford, Chuck Schaeffer, Bob Shears, and
+Joe Skin from the North Slope Borough, Alaska.
+
+We are grateful for the assistance in catching and sampling seals by Ryan Adam,
+James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov,
+Cynthia Christman, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb Fauquier, Shannon
+Fitzgerald, Kathy Frost, Scott Gende, Tracey Goldstein, Jeff Harris, Markus
+Horning, John Jansen, Shawn Johnson, Charles Littnan, Lloyd Lowry, Erin
+Moreland, Mark Nelson, Justin Olnes, Lorrie Rea, Gay Sheffield, Brent Stewart,
+and Dave Withrow. We also appreciate the commitment to science and safety by all
+officers and crew of the NOAA Ship Oscar Dyson, the NOAA Ship MacArthur II, and
+the RV Thomas G. Thompson.
 

### 2022-01-18 analysis/paper/London_HauloutBeahvior.Rmd
UPDATE ACKNOWLEDGEMENTS

@@
   focused on bearded (_Erignathus barbatus_), ribbon (_Histriophoca fasciata_),
-  and spotted seals (_Phoca largha_)) in the Bering and Chukchi seas. Because
+  and spotted seals (_Phoca largha_) in the Bering and Chukchi seas. Because
   ringed seals (_Phoca hispida_) are unique in their use of snow lairs, they were
@@
 ice-associated seals that depend on spring and early summer sea ice (March-June)
-in the Bering and Chukchi seas as a platform for important life history
+in the Bering, Chukchi seas as a platform for important life history
 functions, such as pupping, nursing, breeding behavior, and molting [@boveng2009;
@@
 behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
-Our goals were threefold. First, we wished to establish baseline estimates for
+Our goals were threefold. First, we established baseline estimates for
 the chronology of haul-out behavior in the critical spring season for each
@@
 ```
+
 Data used in our study came from bio-loggers deployed on bearded, ribbon, and
@@
 Field identification of age class can be inexact, particularly when discerning
-subadults from adults. Sex and age class (non-dependent _young-of-the-year_,
+subadults from adults. Sex as well as age class (non-dependent _young-of-the-year_,
 sexually immature _subadults_, and mature _adults_) were estimated at the time
@@
 Tags that fall off due to molt, attachment failure, or seal mortality and remain
-on ice or land can still send data to satellites; i.e., the bio-logger will be
-on ice or land and dry, therefore, it will record and transmit data suggesting
+on ice or land can still send data to satellites; i.e., a detached bio-logger 
+that is dry (either on ice or land) will record and transmit data suggesting
 the seal is hauled out. As such, end times of each deployment were identified by
@@
 from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
-all assigned an error radius of 50m. Any days where haul-out observations were
+all assigned an error radius of 50m. For any days where haul-out observations were
 present but location data were missing we used the last calculated weighted
-average daily location, and any days where the location intersected with land
+average daily location. Any days where the location intersected with land
 were removed from the data set. We recognized that bearded and spotted seals
 haul out on land. However, assessing the relationship between haul-out behavior
-and weather covariates and their availability for aerial surveys on land was
+and weather covariates and seals' availability for aerial surveys on land was
 outside the scope of this study. Additionally, any daily locations on land are
@@
 over the Bering and Chukchi seas at 3-hr intervals based on the native grid
-resolution of 32 km (1024 sq. km). The following meteorological variables are
+resolution of 32km (1024km^2^). The following meteorological variables are
 known to affect haul-out behavior in other Arctic pinnipeds [@reder2003;
@@
 parsimonious. For these reasons, we have elected not to use sea ice
-concentration as a predictor for haul-out probability in the present study
-(note, however, that habitat selection analyses incorporating sea-ice
-concentration are a focus of current and future research).
+concentration as a predictor for haul-out probability in the present study.
 
@@
   "Temp", "Continuous", "NARR",
-  "air temperatures at 2m above the earth’s surface",
+  "air temperature (C) at 2m above the earth’s surface",
   "Wind", "Continuous", "NARR", 
@@
 records, we fit separate models to bearded, ribbon, and spotted seal data sets.
-In our first model formulation, for each species, we fitted a year-independent
+In our first model formulation and for each species, we fitted a year-independent
 model that predicted average haul-out behavior as a function of demographic,
@@
 haul-out probability, averaged over the other weather conditions, is more
-variable than simply showing model coefficients would imply. That said,
+variable than model coefficients would imply. That said,
 important insights can be gained from plots of marginal effects. To create these
@@
 
-When exploring the influence of weather, Bearded seal haul-out probability appears most influenced by wind
+When exploring the influence of weather, bearded seal haul-out probability appeared most influenced by wind
 ($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$
@@
 observed data. Because our ribbon seal model included age and sex class, we can
-visualize the different influences of weather covariates those classes. Of note,
-there is an indication that sub-adult ribbon seals are more likely to haul-out
-at lower temperatures. But, this may also be simply reflective of their
-tendency to haul-out earlier in the season compared to adults.
+visualize the different influences of weather covariates on those classes.
 
@@
 and sampling variability. While we did remove any years where only one
-deployment in a species + age-sex group was present, there are still
+deployment in a species + age:sex group was present, there are still
 some years where the pattern shown is informed by a small number of
@@
 
-```{r annualHO, fig.height = 5.5, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability for ribbon and spotted seals across 10 years.
+```{r annualHO, fig.height = 5.5, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability (colored markers) for ribbon and spotted seals across 13 years.
 Predictions are shown for local solar noon and under smoothed weather conditions.
-Only those groups (age-sex + year) that included observations from more than one seal are shown.
+Only those groups (age:sex + year) that included observations from more than one seal are shown.
 Additionally, any groups where data were only available after 1 June or before 1 May are not included."}
@@
 photoperiod, ribbon and spotted seals are dependent upon the presence of sea ice
-for pupping and molt. We didn't find any support in our models for a
+for pupping and molt. We did not find any support in our models for a
 relationship between the timing of peaks in haul-out behavior and the amount of
 yearly maximum sea ice. This could indicate that, while the extent of spring sea
-ice in the Bering sea varied widely during our study period, seals were still
+ice in the Bering Sea varied widely during our study period, seals were still
 able to locate sea ice and haul out. We should expect, however, that some
@@
 Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr., Virgil Naylor Sr., Dan
-Savetilik, Ross Schaeffer, Allen Stone, and Randy Toshavik from Kotzebue,
-Alaska; Merlin Henry from Koyuk, Alaska; Tom Gray from Nome; Vernon Long and
-Richard Tukle from Nuiqsuit, Alaska; Morgan Simon, River Simon, and Al Smith
-from Scammon Bay, Alaska; Alex Niksik Jr. from St. Michael, Alaska; Billy Adams,
-Mary-Ellen Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman,
-Howard Kittick, Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik,
-Shawn Oktollik, Stacey Osborn, Fred Rexford, Chuck Schaeffer, Bob Shears, and
-Joe Skin from the North Slope Borough, Alaska.
+Savetilik, Chuck Shaeffer, Ross Schaeffer, Allen Stone, and Randy Toshavik from
+Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Tom Gray from Nome; Vernon
+Long and Richard Tukle from Nuiqsuit, Alaska; Morgan Simon, River Simon, and Al
+Smith from Scammon Bay, Alaska; Alex Niksik Jr. from St. Michael, Alaska; Billy
+Adams, James Aiken, Tim Aiken, Howard Kittick, Gilbert Leavitt, Isaac Leavitt,
+Joe Skin from Utqiaġvik, Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn
+Oktollik, Stacey Osborn, and Fred Rexford from Ulguniq, Alaska.
 
@@
 James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov,
-Cynthia Christman, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb Fauquier, Shannon
-Fitzgerald, Kathy Frost, Scott Gende, Tracey Goldstein, Jeff Harris, Markus
-Horning, John Jansen, Shawn Johnson, Charles Littnan, Lloyd Lowry, Erin
-Moreland, Mark Nelson, Justin Olnes, Lorrie Rea, Gay Sheffield, Brent Stewart,
-and Dave Withrow. We also appreciate the commitment to science and safety by all
+Cynthia Christman, Sarah Coburn, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
+Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey Goldstein, Jeff
+Harris, Jason Herreman, Markus Horning, John Jansen, Shawn Johnson, Charles
+Littnan, Lloyd Lowry, Brett McClintock, Erin Moreland, Mark Nelson, Justin
+Olnes, Lorrie Rea, Bob Shears, Gay Sheffield, Brent Stewart, Dave Withrow, and
+Heather Ziel. We also appreciate the commitment to science and safety by all
 officers and crew of the NOAA Ship Oscar Dyson, the NOAA Ship MacArthur II, and

### 2022-01-26 README.Rmd
UPDATE ACKNOWLEDGEMENTS

--- 2021-12-22 README.Rmd
@@
     for manuscript. Includes code to reproduce the figures and tables
-    generated by the analysis. It also has a rendered MS Word version,
-    `London_HauloutBehavior.docx`, suitable for reading (the code is replaced by figures
+    generated by the analysis. It also has a rendered as a PDF,
+    [London_HauloutBehavior.pdf](https://github.com/jmlondon/berchukHaulout/blob/main/analysis/paper/London_HauloutBeahvior.pdf), suitable for reading (the code is replaced by figures
     and tables in this file)
@@
 
+## Acknowledgements
+
+We recognize that the species and ecosystems studied in this paper are within
+the traditional range of the Inpuiaq and Yupik people. And, we would like to
+acknowledge and thank them as the original stewards of these ecosystems. The
+deployment of bio-logging devices used in this study were often done in
+collaboration with Alaska Native seal hunters near their communities. We would
+like to especially acknowledge the communities of Kotzebue, Koyuk, Nome,
+Nuiqsut, Scammon Bay, St. Michael, Utqiaġvik, and Ulguniq (Wainwright) and the
+following individuals: James Adams, Jeff Barger, David Barr, Wendell Booth,
+Cyrus Harris, Nereus 'Doc' Harris, Grover Harris, Lee Harris, Tom Jones, Frank
+Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin, Pearl Goodwin, Willie
+Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr., Virgil Naylor Sr., Dan
+Savetilik, Chuck Schaeffer, Ross Schaeffer, Allen Stone, and Randy Toshavik from
+Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Tom Gray from Nome; Vernon
+Long and Richard Tukle from Nuiqsuit, Alaska; Morgan Simon, River Simon, and Al
+Smith from Scammon Bay, Alaska; Alex Niksik Jr. from St. Michael, Alaska; Billy
+Adams, James Aiken, Tim Aiken, Howard Kittick, Gilbert Leavitt, Isaac Leavitt,
+J.R. Leavitt, and Joe Skin from Utqiaġvik, Alaska; Mary Ellen Ahmaogak, Enoch
+Oktollik, Shawn Oktollik, Stacey Osborn, and Fred Rexford from Ulguniq, Alaska.
+
+We are grateful for the assistance in catching and sampling seals by Ryan Adam,
+James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov,
+Cynthia Christman, Sarah Coburn, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
+Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey Goldstein, Jeff
+Harris, Jason Herreman, Markus Horning, John Jansen, Shawn Johnson, Charles
+Littnan, Lloyd Lowry, Brett McClintock, Erin Moreland, Mark Nelson, Justin
+Olnes, Lorrie Rea, Bob Shears, Gay Sheffield, Brent Stewart, Dave Withrow, and
+Heather Ziel. We also appreciate the commitment to science and safety by all
+officers and crew of the NOAA Ship Oscar Dyson, the NOAA Ship MacArthur II, and
+the RV Thomas G. Thompson.
+
+Telemetry data from the Alaska Department of Fish and Game (ADFG) and the North
+Slope Borough Department of Wildlife Management (NSB) were important
+contributions to the findings presented here. Both agencies collaborate closely
+with Alaska Native communities and the contributions of local hunters to the
+successful capture, deployment, and understanding of seal ecology are
+invaluable. Deployments in the western Bering Sea were done in collaboration
+with Russian colleagues and North Pacific Wildlife.
+
 ## Inspiration

### 2022-01-26 analysis/paper/London_HauloutBeahvior.Rmd
UPDATE ACKNOWLEDGEMENTS

@@
 Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr., Virgil Naylor Sr., Dan
-Savetilik, Chuck Shaeffer, Ross Schaeffer, Allen Stone, and Randy Toshavik from
+Savetilik, Chuck Schaeffer, Ross Schaeffer, Allen Stone, and Randy Toshavik from
 Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Tom Gray from Nome; Vernon
@@
 Adams, James Aiken, Tim Aiken, Howard Kittick, Gilbert Leavitt, Isaac Leavitt,
-Joe Skin from Utqiaġvik, Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn
+J.R. Leavitt, and Joe Skin from Utqiaġvik, Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn
 Oktollik, Stacey Osborn, and Fred Rexford from Ulguniq, Alaska.

### 2022-01-26 README.Rmd
PLB COMMENTS ADDRESSED

@@
 
-We recognize that the species and ecosystems studied in this paper are within
-the traditional range of the Inpuiaq and Yupik people. And, we would like to
-acknowledge and thank them as the original stewards of these ecosystems. The
-deployment of bio-logging devices used in this study were often done in
-collaboration with Alaska Native seal hunters near their communities. We would
-like to especially acknowledge the communities of Kotzebue, Koyuk, Nome,
-Nuiqsut, Scammon Bay, St. Michael, Utqiaġvik, and Ulguniq (Wainwright) and the
-following individuals: James Adams, Jeff Barger, David Barr, Wendell Booth,
-Cyrus Harris, Nereus 'Doc' Harris, Grover Harris, Lee Harris, Tom Jones, Frank
-Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin, Pearl Goodwin, Willie
-Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr., Virgil Naylor Sr., Dan
-Savetilik, Chuck Schaeffer, Ross Schaeffer, Allen Stone, and Randy Toshavik from
-Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Tom Gray from Nome; Vernon
-Long and Richard Tukle from Nuiqsuit, Alaska; Morgan Simon, River Simon, and Al
-Smith from Scammon Bay, Alaska; Alex Niksik Jr. from St. Michael, Alaska; Billy
-Adams, James Aiken, Tim Aiken, Howard Kittick, Gilbert Leavitt, Isaac Leavitt,
-J.R. Leavitt, and Joe Skin from Utqiaġvik, Alaska; Mary Ellen Ahmaogak, Enoch
-Oktollik, Shawn Oktollik, Stacey Osborn, and Fred Rexford from Ulguniq, Alaska.
+We recognize that the species and ecosystems we studied are within the ancestral
+and present-day environs of the Inupiat and Yup'ik people who, through many
+uncredited contributions of traditional knowledge, provided early western
+naturalists and scientists with much of what gets described as the ‘basic
+biology’ of Arctic seals. The deployment of bio-logging devices used in this
+study were often done in collaboration with Alaska Native seal hunters near
+their communities. We would like to especially acknowledge the communities of
+Kotzebue, Koyuk, Nome, Nuiqsut, Scammon Bay, St. Michael, Utqiaġvik, and Ulguniq
+(Wainwright) and the following individuals: James Adams, Jeff Barger, David
+Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris, Grover Harris, Lee
+Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin,
+Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr.,
+Virgil Naylor Sr., Dan Savetilik, Chuck Schaeffer, Ross Schaeffer, Allen Stone,
+and Randy Toshavik from Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Tom
+Gray from Nome; Vernon Long and Richard Tukle from Nuiqsuit, Alaska; Morgan
+Simon, River Simon, and Al Smith from Scammon Bay, Alaska; Alex Niksik Jr. from
+St. Michael, Alaska; Billy Adams, James Aiken, Tim Aiken, Howard Kittick,
+Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, and Joe Skin from Utqiaġvik,
+Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, and
+Fred Rexford from Ulguniq, Alaska.
 

### 2022-01-26 analysis/paper/London_HauloutBeahvior.Rmd
PLB COMMENTS ADDRESSED

@@
 ice-associated seals that depend on spring and early summer sea ice (March-June)
-in the Bering, Chukchi seas as a platform for important life history
-functions, such as pupping, nursing, breeding behavior, and molting [@boveng2009;
+in the Bering, Chukchi seas as a platform for important life history functions,
+such as pupping, nursing, breeding behavior, and molting [@boveng2009;
 @boveng2013a; @cameron2010; @kelly2010]. Limited data and large knowledge gaps
@@
 within regions and throughout their full ranges. Additionally, trends in
-abundance from large scale surveys of these species are unknown, so it is
-difficult to assess the effect, if any, declines in sea ice habitat have had, or
-will have, on seal densities.
+abundance of these species are unknown, so it is difficult to assess the effect,
+if any, declines in sea ice habitat have had, or will have, on seal demography.
 
@@
 Ringed seals (_Phoca hispida_) --- not included in this study --- bask on
-sea-ice but also within snow lairs. The remoteness of the Bering and Chukchi
-seas mean direct scientific observation of seal behavior is impractical. Thus,
-bio-logging devices are especially useful tools for collecting key information
-on movement and haul-out behavior for these species.
-
-Bio-logging records of time spent out of the water provide valuable data for
-identifying covariates that explain variation in haul-out behavior. For
+sea-ice but also within snow lairs during winter and spring.
+
+The remoteness of the Bering and Chukchi seas means direct scientific observation
+of seal behavior is impractical. Thus, bio-logging devices are especially useful
+tools for collecting key information on movement and haul-out behavior for these
+species. Bio-logging records of time spent out of the water provide valuable
+data for identifying covariates that explain variation in haul-out behavior. For
 instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
@@
 changes between diurnal and nocturnal haul-out behavior of ringed seals
-previously described by Kelly and Quakenbush [-@kelly1990c] using VHF radio tags and
-direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
+previously described by Kelly and Quakenbush [-@kelly1990c] using VHF radio tags
+and direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
 propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
@@
 Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed seals in the
-Chukchi Sea. These studies were often further limited by logistical constraints
-on fieldwork and the attachment duration or operational life of bio-loggers. In
-this study, we address some of these limitations by deploying small bio-loggers
-designed for longer-term attachment on rear flippers of a subset of the study
-individuals. These devices are designed to collect data through the molt period
-(when those adhered to the hair would fall off) and, in some situations, provide
-multiple years of data.
+Chukchi and Beaufort seas. These studies were often further limited by
+logistical constraints on fieldwork and the attachment duration or operational
+life of bio-loggers. In this study, we address some of these limitations by
+deploying small bio-loggers designed for longer-term attachment on rear flippers
+of a subset of the study individuals. These devices are designed to collect data
+through the molt period (when those adhered to the hair would fall off) and, in
+some situations, provide multiple years of data.
 
@@
 behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
-Our goals were threefold. First, we established baseline estimates for
+Our goals were threefold. First, we wished to establish baseline estimates for
 the chronology of haul-out behavior in the critical spring season for each
-species across different age and sex classes. Second, we refined estimates of
-haul-out availability corrections for aerial surveys in order to improve
-estimates of seal abundance. Previously estimated availability correction
-factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
-for variables such as the time-of-day and day-of-year, but did not investigate
-meteorological variables that have been shown to influence haul-out behavior of
-walruses [@udevitz2009]. Third, we assessed the annual variability
-in haul-out timing and possible linkage to changes in the extent of seasonal sea
-ice between 2005 and 2020. Our work extends the scope of previous haul-out
-analyses, includes the influence of meteorological variability, and investigates
-the potential impact of a changing icescape on the behavior of these species.
+species across different age and sex classes. Second, we sought to refine
+estimates of haul-out availability corrections for aerial surveys in order to
+improve estimates of seal abundance. Previously estimated availability
+correction factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a])
+accounted for variables such as the time-of-day and day-of-year, but did not
+investigate meteorological variables that have been shown to influence haul-out
+behavior of walruses [@udevitz2009]. Third, we aimed to assess the annual
+variability in haul-out timing and possible linkage to changes in the extent of
+seasonal sea ice between 2005 and 2020. Our work extends the scope of previous
+haul-out analyses, includes the influence of meteorological variability, and
+investigates the potential impact of a changing icescape on the behavior of
+these species.
 
@@
 Data used in our study came from bio-loggers deployed on bearded, ribbon, and
-spotted seals in the Bering, Chukchi, and Beaufort seas by multiple
+spotted seals in the Bering, Chukchi, and western Beaufort seas by multiple
 organizations as part of collaborative investigations from 2005 through 2020.
-Seals were captured using nets and bio-loggers attached during ship-based
-capture trips and from coastal Alaska communities. Ship-based capture events
+Seals were captured using nets and bio-loggers were attached during studies based
+in coastal Alaska communities or on research ships. Ship-based capture events
 occurred during spring near the southern ice edge in the Bering Sea. Land-based
@@
 'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
-Washington, USA) and either adhered to the hair on the seal or attached to the
+Washington, USA) and either adhered to the hair on the seal or attached through the
 rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
@@
 of haul-out data they provide. Additionally, bio-loggers attached to the head or
-dorsal region are often dry while the seal is at the surface and can slightly
-bias the hourly percent-dry values reported by the bio-logger. For this study,
-in cases where both bio-logger types were deployed, hourly percent-dry
-observations from the flipper tag were preferred.
-
-Field identification of age class can be inexact, particularly when discerning
+dorsal region are often dry while the seal is floating at the surface, inducing
+a slight positive bias in the hourly percent-dry values reported by the
+bio-logger. For this study, in cases where both bio-logger types were deployed,
+hourly percent-dry observations from the flipper tag were preferred.
+
+Field identification of age class can be inexact, particularly when differentiating
 subadults from adults. Sex as well as age class (non-dependent _young-of-the-year_,
@@
 young-of-the-year. Subadults are those seals likely greater than one year of age
-but not yet sexually mature. Adults are likely sexually mature and older than
-four years. For those bio-loggers deployed on young-of-the-year and transmitting
-into the next year, the age class was advanced to subadult on 1 March of the
-following year. Table \@ref(tab:tblDeploy) provides a summary of these
-deployments and data received from them.
+but not yet sexually mature. Adults are individuals that are likely sexually
+mature and older than four years. For those bio-loggers deployed on
+young-of-the-year and transmitting into the next year, the age class was
+advanced to subadult on 1 March of the following year. Table
+\@ref(tab:tblDeploy) provides a summary of these deployments and data received
+from them.
 
@@
 Tags that fall off due to molt, attachment failure, or seal mortality and remain
-on ice or land can still send data to satellites; i.e., a detached bio-logger 
+on ice or land may continue to send data to satellites; i.e., a detached bio-logger 
 that is dry (either on ice or land) will record and transmit data suggesting
-the seal is hauled out. As such, end times of each deployment were identified by
+the seal is hauled out. Therefore, end times of each deployment were identified by
 examining bio-logger locations, percent-dry records, and dive behavior (if
@@
 transmitted. Molt timing also impacts when many deployments end as any
-bio-loggers adhered to the hair will fall off. Bearded seal deployments are more
-evenly spread with fewer observations in May and June. The majority of bearded
-seal deployments start later in the summer and by May bio-loggers have either
-fallen off or their batteries are depleted.
+bio-loggers adhered to the hair will fall off. Relative to the other species in
+the study, there were fewer deployments of bio-loggers on bearded seals. This
+resulted in fewer data observations overall and noticeably lower in numbers May
+and June. The majority of bearded seal deployments start later in the summer and
+by May bio-loggers have either fallen off or their batteries are depleted.
 
@@
 The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out.
-Additionally, many bio-logger deployments started in April and May."}
+Additionally, many bio-logger deployments started in April and May.
+The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study."}
 
@@
 network as hourly percent-dry timelines. For each hour of a day, the wet/dry
-sensor was polled by the tag firmware every few seconds and a percent of the
-hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). On board the
+sensor was polled by the tag firmware every few seconds and the percent of the
+hour in the dry state was calculated (Figure \@ref(fig:examplePlot)). On board the
 bio-logger, hourly percent-dry data were rounded to the nearest 10%
@@
 compression resulted in additional data transmission as each message consisted
-of two complete 24-hour records. Bio-loggers were, generally, deployed and
+of two complete 24-hour records. Bio-loggers were, typically, deployed and
 programmed in a manner to maximize data transmission during the spring pupping
-and molting period; however, hourly percent-dry data were not always
+and molting period, though hourly percent-dry data were not always
 successfully transmitted. This is due to a variety of factors including
-satellite coverage, tag availability (i.e. tags mounted to the rear flipper
+satellite coverage, tag availability (e.g., tags mounted to the rear flipper
 often do not transmit while at sea), tag performance, duty cycling, and
 extra-terrestrial atmospheric interference. Fortunately, missing records do not
-seem to greatly bias inference about haul-out probabilities [@conn2012b].
+substantially bias inference about haul-out probabilities [@conn2012b].
 
 ```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployed on a ribbon seal over two years during the months of March, April, and May.
-Areas of the plot indicated with an 'X' represent missing data that were not successfully transmitted from the tag."}
+Areas of the plot indicated with an 'X' represent missing data that were not successfully recieved from the tag."}
 
@@
 Of key interest in this study was the relationship between haul-out behavior and
-weather covariates that vary with time and animal location. We explored the use
+weather covariates that vary with time and seal location. We explored the use
 of a continuous-time correlated random walk [@johnson2008a] movement model
@@
 from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
-all assigned an error radius of 50m. For any days where haul-out observations were
-present but location data were missing we used the last calculated weighted
-average daily location. Any days where the location intersected with land
-were removed from the data set. We recognized that bearded and spotted seals
-haul out on land. However, assessing the relationship between haul-out behavior
-and weather covariates and seals' availability for aerial surveys on land was
-outside the scope of this study. Additionally, any daily locations on land are
-likely more reflective of coordinate averaging and measurement error instead of
-actual use of coastal features. Figure \@ref(fig:dataMap) shows the spatial
-distribution of weighted locations with available haul-out behavior data for each
-species across the study area.
+all assigned an error radius of 50m. On days when haul-out observations were
+present but location data were missing we used the seal's last calculated weighted
+average daily location; days when the location intersected with land
+were removed from the seal's record. We recognized that bearded and spotted
+seals haul out on land. However, assessing the relationship between haul-out
+behavior and weather covariates and seals' availability for aerial surveys on
+land was outside the scope of this study. Additionally, any daily locations on
+land are likely more reflective of coordinate averaging and measurement error,
+rather than actual use of coastal features. Figure \@ref(fig:dataMap) shows the
+spatial distribution of weighted locations with available haul-out behavior data
+used for analysis of each species across the study area.
 
@@
 Environmental Prediction [@mesinger2006]. The NARR model assimilates
-observational data to produce a long-term picture of weather over North America.
-Numerous weather variables are made available across the region 8 times daily.
-For this study, NARR weather values were subset to the extent of our study area
-over the Bering and Chukchi seas at 3-hr intervals based on the native grid
-resolution of 32km (1024km^2^). The following meteorological variables are
-known to affect haul-out behavior in other Arctic pinnipeds [@reder2003;
-@udevitz2009; @perry2017] and were interpolated and assigned to daily seal locations
-using a bilinear method: 1) air temperature at 2m above the earth's surface, 2)
-wind consisting of northerly and easterly vector components converted to wind
-speed using the Euclidean norm, 3) barometric pressure at sea level, and 4)
-precipitation (Table \@ref(tab:tblCovariates)).
+observational data to produce a long-term picture of weather over North America
+and portions of the surrounding seas. Weather variables are made
+available across the region 8 times daily. For this study, NARR weather values
+were subset to the extent of our study area over the Bering and Chukchi seas at
+3-hr intervals based on the native grid resolution of 32km (1024km^2^). The
+following meteorological variables are known to affect haul-out behavior in
+other Arctic pinnipeds [@reder2003;
+@udevitz2009; @perry2017] and were interpolated and assigned to daily seal
+locations using a bilinear method: 1) air temperature at 2m above the earth's
+surface, 2) wind consisting of northerly and easterly vector components
+converted to wind speed using the Euclidean norm, 3) barometric pressure at sea
+level, and 4) precipitation (Table \@ref(tab:tblCovariates)).
 
@@
 (and an interaction with day-of-year), since bearded seals occupy a substantial
-range and we were interested in possible differences in the timing of haul-out
-behavior along a latitudinal gradient. We omitted the latitudinal effect from
-ribbon and spotted seal models because, during the spring, these species are most
-prevalent near the southern ice edge in the Bering Sea [@conn2014a].
+range during the spring and we were interested in possible differences in the
+timing of haul-out behavior along a latitudinal gradient. We omitted the
+latitudinal effect from ribbon and spotted seal models because, during the
+spring, these species are most prevalent near the southern ice edge in the
+Bering Sea [@conn2014a].
 
@@
 We assessed whether the annual variation in maximum spring sea ice extent in the
-Bering Sea influences the seasonal peak of seal haul-out behavior. In
+Bering Sea influences the seasonal peak of seal haul-out probability. In
 particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
@@
 previous approaches [@verhoef2014a, @london2012] and only
-`r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
+`r (1 - round((portion_10(analysis_data) + portion_90(analysis_data))) * 100)`% of our
 observations fell between 10% and 90% hourly percent-dry.
@@
 also represents hour-of-day with 6 parameters, which is a considerable reduction
-when compared to a 24 parameter variable, especially when fitting models with
+when compared to a 24-parameter variable, especially when fitting models including
 interactions between hour-of-day and other variables (e.g., age-sex class,
@@
 across the study region. Such predictions were then used to develop haul-out
-probability surfaces (see figures \@ref(fig:beardedHOCal), \@ref(fig:ribbonHOCal), and
-\@ref(fig:spottedHOCal)), explore conditional effects of weather covariates, and
+probability surfaces, explore conditional effects of weather covariates, and
 determine annual peaks in haul-out activity. The timing of peak haul-out
@@
 important insights can be gained from plots of marginal effects. To create these
-plots (figures \@ref(fig:beardedHOwx), \@ref(fig:ribbonHOwx), and
-\@ref(fig:spottedHOwx)), we predicted haul-out probability across the full range
+plots, we predicted haul-out probability across the full range
 of weather covariate values while fixing hour of the day at local solar noon and
@@
 lower sample size for adult and young-of-year age classes. As such, results are
-shown for all ages. Overall, bearded seals were less likely to haul out and had
-a bi-modal distribution of haul-out probability across the day (Figure
-\@ref(fig:beardedHOCal)). In addition to a peak around local solar noon, the
-bearded seal model predicts additional haul-out activity around local midnight.
-In concert with the lower magnitude of haul-out probability, bearded seal
-haul-out behavior is also more protracted throughout the spring season compared
-to ribbon and spotted seals.
+shown for all ages (Figure \@ref(fig:beardedHOCal)). Overall, bearded seals were
+less likely to haul out and had a bi-modal distribution of haul-out probability
+across the day. In addition to a peak around local solar noon, the bearded seal
+model predicts additional haul-out activity around local midnight. In concert
+with the lower magnitude of haul-out probability, bearded seal haul-out behavior
+is also more protracted throughout the spring season compared to ribbon and
+spotted seals.
 
@@
 
-When exploring the influence of weather, bearded seal haul-out probability appeared most influenced by wind
+When exploring the influence of weather, bearded seal haul-out probability
+appeared most influenced by wind
 ($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$
@@
 = `r bearded_type3$temp2$F.value`; $p$ =
-`r bearded_type3$temp2$Prob.F`). Additionally, wind chill was also a
-significant influence
+`r bearded_type3$temp2$Prob.F`) (Figure \@ref(fig:beardedHOwx)). 
+Additionally, wind chill was also a significant influence
 ($F_{`r paste(bearded_type3$temp2_wind$Num.df, bearded_type3$temp2_wind$Den.df, sep=",")`}$
@@
 
-```{r beardedHOwx, fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
+```{r beardedHOwx, fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
 Hour of the day was fixed at local solar noon and day-of-year held at the day of peak haul-out for bearded seals.
-Transparent vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
+Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
 
@@
 adults (Figure \@ref(fig:ribbonHOCal)). The behavior is clearly centered
-around local solar noon and expands to other hours later in the season
+around local solar noon and expands to other hours later in the spring
 as seals enter their molting period. Subadults showed an earlier start
@@
 
-```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification.
+```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification.
 Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for ribbon seals.
-Transparent vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
+Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
 
@@
 
-Spotted seal haul-out behavior appears most strongly influenced by
+Spotted seal haul-out behavior was most strongly influenced by
 temperature
@@
 precipitation were not as influential as the other covariates.
-Differences in the magnitude of response between the age-sex classes are
+Differences in the magnitude of response between the age-sex classes were
 present and consistent across each of the weather covariates (Figure
-\@ref(fig:spottedHOwx)). There is a consistent ranking of adult males being the most
+\@ref(fig:spottedHOwx)). There was a consistent ranking of adult males being the most
 likely to haul out, followed by adult females, and, then, subadults.
 
-```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
+```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
 Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for spotted seals.
-Transparent vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
+Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
 
@@
 
-The second set of models that included annual variation in haul-out
-patterns uncovered significant contributions for linear and quadratic
+The second set of models, which included annual variation in haul-out
+patterns, uncovered significant contributions for linear and quadratic
 interactions between day and year in both ribbon seals (day:year,
@@
 (Figure \@ref(fig:annualHO)). It is important to note that predicted
-variation in annual haul-out patterns likely reflects both process error
+variation in annual haul-out patterns likely reflected both process error
 and sampling variability. While we did remove any years where only one
-deployment in a species + age:sex group was present, there are still
-some years where the pattern shown is informed by a small number of
-individuals that may not represent population level patterns.
+deployment in a species + age:sex group was present, there were still
+some years where the pattern shown was informed by a small number of
+individuals that may not represent population-level patterns.
 
@@
 The timing of peak haul-out probability for ribbon and spotted seals
-appeared to have a limited to no relationship with the amount of
+appeared to have little to no relationship with the amount of
 yearly maximum sea ice extent. Adult female and subadult spotted seals
@@
 declining again. This pattern aligns well with what has been previously
-documented and confirms our haul-out data are likely capturing population-level
-behavioral patterns. Seals preferentially haul out on ice shortly after solar
-noon, which allows seals to maximize absorption of solar radiation thought to
-facilitate the molting process [@feltz1966]. Interestingly, bearded seals appear
-to have two peaks in haul-out activity within a day, one shortly after solar
-noon, and one centered near solar midnight. This, of course, could be an
-artifact of our limited sample size for bearded seal deployments across all age
-classes. However, a similar bi-modal pattern has been seen in ringed seals
+documented qualitatively [@boveng2009; @boveng2013; @cameron2010] and confirms
+our haul-out data are likely quantifying population-level behavioral patterns.
+Seals preferentially haul out on ice shortly after solar noon, which allows
+seals to maximize absorption of solar radiation thought to facilitate the
+molting process [@feltz1966]. Interestingly, bearded seals appear to have two
+peaks in haul-out activity within a day, one shortly after solar noon, and one
+centered near solar midnight. This, of course, could be an artifact of our
+limited sample size for bearded seal deployments across all age classes.
+However, a similar bi-modal pattern has been seen in ringed seals
 [@vonduyke2020] and suggests that bearded and ringed seals may be operating
@@
 factors such as predation by polar bears (which is rare for ribbon and spotted
-seals in the Bering Sea) may also explain differing haul-out patterns. Bearded
-seals also showed a more protracted season of haul-out behavior compared to
-ribbon and spotted seals. This aligns with findings from Thometz et al.
-[@thometz2021] who observed a mean molting period of 119&pm;2 days and a
-relatively stable resting metabolic rate during that time. While ribbon seals
-were not considered in the same study, spotted and ringed seals underwent molt
-periods of 33&pm;4 and 28&pm;6 days and exhibited increased resting metabolic
-rates.
-
-Unlike previous analyses of seal haul-out behavior (e.g. [@verhoef2010a],
-[@conn2014a]), we also investigated the influence of sex-age class on haul-out
-probabilities for all species except bearded seals because of low sample size.
-Although ribbon and spotted seals exhibited a unimodal diel haul-out pattern
-generally centered around local solar noon, there were key differences across
-species, age, and sex that match our understandinf of their
-ecological behavior. Spotted seals are known to form triads during the breeding
-season where a female and dependent pup are accompanied on the ice by a suitor
-male [@frost2018]. The male waits for the female to wean the pup and enter
-estrus, and fends off any other potential suitor males. Triad formation results
-in both males and females spending a large portion of the day hauled out on ice
-and a protracted spring haul-out season for both sexes. While females are still
-nursing the pup and not yet in estrus they may be less inclined to interrupt
-their haul out and enter the water where the suitor male would attempt mating.
-We see this reflected in the predicted haul-out behavior, with both males and
-females exhibiting a broad distribution of time out of the water throughout the
-solar day and the season. Ribbon seals are not known to form triads and our
-model predicts a progression of increased haul-out behavior with females
-starting earlier in the season than males. Notably, female ribbon seals spend a
-large portion of the day in the water during the pupping period, aligning with
-the hypothesis that ribbon seal females continue foraging while nursing.
-Subadult ribbon and spotted seals begin consistent haul-out behavior earlier and
-follow a pattern seen in other phocids where yearlings and subadults molt first
-followed by adult females and males [@reder2003, @thompson1987, @kirkman2003].
+seals in the Bering Sea) may also explain differing haul-out patterns. The
+change in haul-out behavior during the season was less pronounced in bearded
+seals compared to ribbon and spotted seals. This aligns with findings from
+Thometz et al. [@thometz2021] who observed a mean molting period of 119&pm;2
+days and a relatively stable resting metabolic rate during that time. While
+ribbon seals were not considered in that same study, spotted and ringed seals
+underwent molt periods of just 33&pm;4 and 28&pm;6 days and had increased
+resting metabolic rates.
+
+Unlike previous analyses of seal haul-out behavior in spotted and ribbon seals
+(e.g. [@verhoef2010a], [@conn2014a]), we also investigated the influence of
+sex-age class on haul-out probabilities for all species (not for bearded seals
+because of low sample size). Although ribbon and spotted seals exhibited a
+unimodal diel haul-out pattern generally centered around local solar noon, there
+were key differences across species, age, and sex that match our understanding
+from natural history descriptions of their ecological behavior. Spotted seals
+are known to form triads during the breeding season where a female and dependent
+pup are accompanied on the ice by a suitor male [@frost2018]. The male waits for
+the female to wean the pup and enter estrus, and fends off any other potential
+suitor males. Triad formation results in both males and females spending a large
+portion of the day hauled out on ice and a protracted spring haul-out season for
+both sexes. While females are still nursing the pup and not yet in estrus they
+may be less inclined to interrupt their haul out and enter the water where the
+suitor male would attempt mating. We see this reflected in the predicted
+haul-out patterns, with both males and females exhibiting a broad distribution
+of time out of the water throughout the solar day and the season. Ribbon seals
+are not known to form triads and our model predicts a progression of increased
+haul-out behavior with females starting earlier in the season than males.
+Notably, female ribbon seals spend a large portion of the day in the water
+during the pupping period, aligning with the hypothesis that ribbon seal females
+continue foraging while nursing. Subadult ribbon and spotted seals begin
+elevated haul-out behavior earlier in the spring and follow a pattern seen in
+other phocids where yearlings and subadults molt first followed by adult females
+and males [@reder2003, @thompson1987, @kirkman2003].
 
@@
 [-@udevitz2009]) and a few select studies of ice-associated seals
-[@hamilton2018, @perry2017]. Ribbon seals seemed to be the most influenced by
-weather, with wind, temperature, and barometric pressure all
-being important components of the model. Spotted seals were most affected by
-wind and barometric pressure. And, for bearded seals, the model indicated wind
-and temperature had the greatest impact. In general, and as might be expected,
-seals were more likely to haul-out when daily temperatures were warmer, winds
-speeds were lower, barometric pressure was higher, and precipitation was lower.
-These weather conditions are general indicators of increased solar radiation
-which provides energetic benefits. Low winds and precipitation may also
-provide better predator detection conditions. Our results highlight the
-importance of incorporating weather covariates when analyzing haul-out 
-behavior and calculating availability corrections for aerial surveys.
+[@hamilton2018, @perry2017]. In our study, ribbon seals seemed to be the most
+influenced by weather, with wind, temperature, and barometric pressure all being
+important components of the model. Spotted seals were most affected by wind and
+barometric pressure. And, for bearded seals, the model indicated wind and
+temperature had the greatest impact. In general, and as might be expected, seals
+were more likely to haul-out when daily temperatures were warmer, winds speeds
+were lower, barometric pressure was higher, and precipitation was lower. Those
+weather conditions are general indicators of increased solar radiation and lower
+convective heat loss, both of which provides energetic benefits. Low winds and
+precipitation could also enhance predator detection. Our results highlight the
+importance of incorporating weather covariates when analyzing haul-out behavior
+and calculating availability corrections for aerial surveys.
 
@@
 ice in the Bering Sea varied widely during our study period, seals were still
-able to locate sea ice and haul out. We should expect, however, that some
-minimal threshold in the spatial extent or density of sea ice will have a
-meaningful impact on the timing of peak haul-out behavior --- if there is no sea
-ice, seals will not haul out or be forced to use terrestrial haul-out sites.
-Additionally, while from an ecological perspective the haul-out behavior appears
-consistent, the interannual differences in timing and magnitude are large enough
-to have important ramifications on calculations of abundance and trend. Those
-ramifications will only be exacerbated if climate variability amplifies
-interannual differences.
+able to locate sea ice and haul out. Only a small portion of our data was from
+2018 – 2019, years of extreme low spring ice extent in the Bering Sea that
+appeared to impact body condition of ribbon and spotted seals [@boveng2020a], so we may
+currently lack sufficient contrast in ice extent to characterize an effect on
+haul-out probability. We should expect, however, that some minimal threshold in
+the spatial extent or density of sea ice will have a meaningful impact on the
+timing of peak haul-out behavior --- if there is no sea ice, seals will not haul
+out or be forced to use terrestrial haul-out sites which were not part of the
+evolution of their normal behaviors. Additionally, while from an
+ecological perspective the haul-out behavior appears consistent, the interannual
+differences in timing and magnitude are large enough to have important
+ramifications on calculations of abundance and trend. Those ramifications will
+only be exacerbated if climate variability amplifies interannual differences.
 
@@
 seals hauled out on ice based on these factors. We recommend that future
-abundance analyses employ availability models that account for these factors. For
+abundance analyses employ availability models that account for them. For
 instance, it is relatively straightforward to obtain weather reanalysis products
@@
 factors, particularly because research teams would likely need to tag seals and
-conduct aerial surveys concurrently which requires considerably more personnel
+conduct aerial surveys concurrently, requiring considerably more personnel
 and money. One possible suggestion is to estimate a "shift" parameter within
@@
 observed over time. Regardless, researchers should anticipate there being some
-unmodeled heterogeneity in availability probability present in abundance
-estimates obtained from aerial surveys. This may make trend detection difficult,
+unmodeled heterogeneity in availability when using aerial surveys to estimate
+Arctic seal abundance. This may require consideration in trend estimation,
 as one will not know if moderate differences in abundance estimates are
@@
 of haul-out probability reflect increased sample sizes in terms of
-number of animals, but also improvements to the way data were prepared prior to
+number of individuals, but also improvements to the way data were prepared prior to
 analysis.
@@
 
-We recognize that the species and ecosystems studied in this paper are within
-the traditional range of the Inpuiaq and Yupik people. And, we would like to
-acknowledge and thank them as the original stewards of these ecosystems. The
-deployment of bio-logging devices used in this study were often done in
-collaboration with Alaska Native seal hunters near their communities. We would
-like to especially acknowledge the communities of Kotzebue, Koyuk, Nome,
-Nuiqsut, Scammon Bay, St. Michael, Utqiaġvik, and Ulguniq (Wainwright) and the
-following individuals: James Adams, Jeff Barger, David Barr, Wendell Booth,
-Cyrus Harris, Nereus 'Doc' Harris, Grover Harris, Lee Harris, Tom Jones, Frank
-Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin, Pearl Goodwin, Willie
-Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr., Virgil Naylor Sr., Dan
-Savetilik, Chuck Schaeffer, Ross Schaeffer, Allen Stone, and Randy Toshavik from
-Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Tom Gray from Nome; Vernon
-Long and Richard Tukle from Nuiqsuit, Alaska; Morgan Simon, River Simon, and Al
-Smith from Scammon Bay, Alaska; Alex Niksik Jr. from St. Michael, Alaska; Billy
-Adams, James Aiken, Tim Aiken, Howard Kittick, Gilbert Leavitt, Isaac Leavitt,
-J.R. Leavitt, and Joe Skin from Utqiaġvik, Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn
-Oktollik, Stacey Osborn, and Fred Rexford from Ulguniq, Alaska.
+We recognize that the species and ecosystems we studied are within the ancestral
+and present-day environs of the Inpuiat and Yup'ik people who, through many
+uncredited contributions of traditional knowledge, provided early western
+naturalists and scientists with much of what gets described as the ‘basic
+biology’ of Arctic seals. The deployment of bio-logging devices used in this
+study were often done in collaboration with Alaska Native seal hunters near
+their communities. We would like to especially acknowledge the communities of
+Kotzebue, Koyuk, Nome, Nuiqsut, Scammon Bay, St. Michael, Utqiaġvik, and Ulguniq
+(Wainwright) and the following individuals: James Adams, Jeff Barger, David
+Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris, Grover Harris, Lee
+Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin,
+Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr.,
+Virgil Naylor Sr., Dan Savetilik, Chuck Schaeffer, Ross Schaeffer, Allen Stone,
+and Randy Toshavik from Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Tom
+Gray from Nome; Vernon Long and Richard Tukle from Nuiqsuit, Alaska; Morgan
+Simon, River Simon, and Al Smith from Scammon Bay, Alaska; Alex Niksik Jr. from
+St. Michael, Alaska; Billy Adams, James Aiken, Tim Aiken, Howard Kittick,
+Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, and Joe Skin from Utqiaġvik,
+Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, and
+Fred Rexford from Ulguniq, Alaska.
 
@@
 Slope Borough Department of Wildlife Management (NSB) were important
-contributions to the findings presented here. Both agencies collaborate closely
-with Alaska Native communities and the contributions of local hunters to the
-successful capture, deployment, and understanding of seal ecology are
-invaluable. Deployments in the western Bering Sea were done in collaboration
-with Russian colleagues and North Pacific Wildlife.
+contributions to the findings presented here. Deployments in the western Bering
+Sea were done in collaboration with Russian colleagues and North Pacific
+Wildlife.
 

### 2022-01-26 R/wx_plots.R

@@ -77,7 +77,7 @@ plot_ribbon_wx <- function(ribbon_fit_obj) {
                                             direction = "horizontal")) +
     theme(legend.title = element_blank(),
           legend.position = "none") +
-    xlab("accumulated precip (mm/hr)") + ylab("haul-out probability")
+    xlab("precip (mm/hr)") + ylab("haul-out probability")
   
   apply_consistent_y_lims <- function(this_plot){
     num_plots <- length(this_plot$layers)
@@ -176,7 +176,7 @@ plot_spotted_wx <- function(spotted_fit_obj) {
                                             direction = "horizontal")) +
     theme(legend.title = element_blank(),
           legend.position = "none") +
-    xlab("accumulated precip (mm/hr)") + ylab("haul-out probability")
+    xlab("precip (mm/hr)") + ylab("haul-out probability")
   
   apply_consistent_y_lims <- function(this_plot){
     num_plots <- length(this_plot$layers)
@@ -271,7 +271,7 @@ plot_bearded_wx <- function(bearded_fit_obj) {
                                             direction = "horizontal")) +
     theme(legend.title = element_blank(),
           legend.position = "none") +
-    xlab("accumulated precip (mm/hr)") + ylab("haul-out probability")
+    xlab("precip (mm/hr)") + ylab("haul-out probability")
   
   apply_consistent_y_lims <- function(this_plot){
     num_plots <- length(this_plot$layers)

### 2022-03-07 R/script_deployment-data-for-suppl-table1.R
Write data sheet for suppl table 1
Supplemental table 1 is based on suggestion from #7 and needs some additional data from collaborators. This script assembles the data writes to a Google Sheet where co-authors can append additional info.

@@ -1,0 +1,35 @@
+library(dbplyr)
+nsb_captures <- tbl(con, in_schema("telem","nsb_capture")) %>% 
+  select(speno, agency) %>% 
+  group_by(speno, agency) %>% 
+  summarise() %>% 
+  collect() %>% 
+  mutate(source = "database")
+mml_deployments <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>% 
+  select(speno, agency = effort_lead) %>% 
+  group_by(speno, agency) %>% 
+  summarise() %>% 
+  collect() %>% 
+  mutate(source = "database")
+
+targets::tar_load(adfg_deployments, store = here::here("_targets"))
+
+adfg_deployments <- adfg_deployments %>% 
+  group_by(speno) %>% 
+  summarise() %>% 
+  mutate(agency = "ADFG", source = "file")
+
+deploy_details <- mml_deployments %>% 
+  bind_rows(nsb_captures) %>% 
+  bind_rows(adfg_deployments %>% select(speno,agency,source))
+
+analysis_data %>% 
+  sf::st_drop_geometry() %>% 
+  select(speno, species, sex, age, deploy_dt, haulout_dt) %>%
+  left_join(deploy_details, by = "speno") %>% 
+  mutate(deploy_year = lubridate::year(deploy_dt)) %>% 
+  group_by(agency, speno, species, deploy_year) %>% 
+  summarise(n_seals = n_distinct(speno), across(sex:age, ~paste(unique(.), collapse = ", "))) %>% 
+  select(-n_seals) %>% 
+  arrange(agency, species, deploy_year, speno) %>% 
+  sheet_write()

### 2022-03-07 analysis/paper/London_HauloutBeahvior.Rmd
add section for Supplemental Material
apply edits from Jim Lee (AFSC review)

@@
   - code: 1
-    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
+    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Marine Fisheries Service, NOAA, Seattle, Washington, USA
   - code: 2
@@
 time bearded seals spent hauled out progressively increased through spring and
-summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
-increased gradually starting in March and peaking in May and June for bearded,
+summer, and Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
+increased gradually starting in March and peaked in May and June for bearded,
 ribbon, and spotted seals.
@@
 to the primary literature for detailed capture and bio-logger attachment methods
-(Supplemental Material, Table).
+(see Supplemental Material, Table 1).
 
@@
 was assigned an error radius based on either the categorical location quality (
-_3_=250m, _2_=500m, _1_=1500m, _0_=2500m [@lopez2013]; we chose 2500m for
+_3_ = 250 m, _2_ = 500 m, _1_ = 1500 m, _0_ = 2500 m [@lopez2013]; we chose 2500 m for
 location classes _A_ and _B_) or, when available, the estimated error radius
 from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
-all assigned an error radius of 50m. On days when haul-out observations were
+all assigned an error radius of 50 m. On days when haul-out observations were
 present but location data were missing we used the seal's last calculated weighted
@@
 were subset to the extent of our study area over the Bering and Chukchi seas at
-3-hr intervals based on the native grid resolution of 32km (1024km^2^). The
+3-hr intervals based on the native grid resolution of 32 km (1024 km^2^). The
 following meteorological variables are known to affect haul-out behavior in
@@
 @udevitz2009; @perry2017] and were interpolated and assigned to daily seal
-locations using a bilinear method: 1) air temperature at 2m above the earth's
+locations using a bilinear method: 1) air temperature at 2 m above the Earth's
 surface, 2) wind consisting of northerly and easterly vector components
@@
 influenced by pupping, nursing, breeding behavior, and molt and these drivers
-are likely more influential than specific sea-ice concentration. Crawford et al
+are likely more influential than specific sea-ice concentration. Crawford et al.
 [-@crawford2019] compared haul-out probability models for ringed seals and found
@@
 important components of the model. Spotted seals were most affected by wind and
-barometric pressure. And, for bearded seals, the model indicated wind and
+barometric pressure. For bearded seals, the model indicated wind and
 temperature had the greatest impact. In general, and as might be expected, seals
@@
 
-# Acknowledgements {-}
+# Acknowledgments {-}
 
@@
 Heather Ziel. We also appreciate the commitment to science and safety by all
-officers and crew of the NOAA Ship Oscar Dyson, the NOAA Ship MacArthur II, and
-the RV Thomas G. Thompson.
+officers and crew of the NOAA ship _Oscar Dyson_, the NOAA ship _MacArthur II_, and
+the RV _Thomas G. Thompson_.
 

### 2022-03-29 README.Rmd

@@
 
-1. Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
+1. Marine Mammal Laboratory, Alaska Fisheries Science Center, National Marine Fisheries Service, NOAA, Seattle, Washington, USA
 2. Arctic Marine Mammals Program, Alaska Department of Fish and Game, Fairbanks, Alaska, USA
@@
 additional reproducibility features (e.g. Dockerfile, MyBinder.org) and
-an R/Shiny application to explore results will be made available.
+an R/Shiny application to explore results will be made developed and made available.
 
@@
     for manuscript. Includes code to reproduce the figures and tables
-    generated by the analysis. It also has a rendered as a PDF,
+    generated by the analysis. It also has a rendered PDF,
     [London_HauloutBehavior.pdf](https://github.com/jmlondon/berchukHaulout/blob/main/analysis/paper/London_HauloutBeahvior.pdf), suitable for reading (the code is replaced by figures
@@
   - [:file\_folder: data](/data): data used for model fits and final model output
+  - [:file\folder: data_raw](/data_raw): data provided by partners from ADFG and NSB
 

### 2022-04-07 analysis/paper/London_HauloutBeahvior.Rmd

@@
   - \onehalfspacing
+  - \usepackage{pdflscape}
+  - \newcommand{\blandscape}{\begin{landscape}}
+  - \newcommand{\elandscape}{\end{landscape}}
 bibliography: references.json
@@
 :::
+
+\blandscape
+# Supplemental Material {-}
+
+# S1. Supplemental Table of Bio-logger Deployment Details
+
+The following table provides additional details about the timing, location, and
+agencies responsible for the bio-logger deployments used in this study.
+
+```{r, include=TRUE}
+library(gt)
+library(googlesheets4)
+library(glue)
+
+deploy_tbl <- read_sheet("1MpCiyAKWLBXufwnzW8QTnb2SsTkrKSuPlWdpWMnqoto")
+
+deploy_tbl %>% 
+  dplyr::select(species, agency, deploy_year, permit, location, speno, sex, age, publications) %>% 
+  dplyr::group_by(species, agency, deploy_year, location, publications) %>% 
+  dplyr::mutate(agency = 
+                  case_when(
+                    agency %in% c('ADFG') ~ "Alaska Dept. Fish & Game",
+                    agency %in% c('NSB') ~ "North Slope Burough",
+                    agency %in% c('PEP') ~ "NOAA Alaska Fisheries Sci. Ctr",
+                    agency %in% c('NPWC') ~ "N. Pacific Wildlife Consortium"
+                  )) %>% 
+  dplyr::summarise(permits = toString(unique(permit)),
+                   n_seals = n(),
+                   sex = toString(unique(sex)),
+                   age_class = tolower(toString(unique(age)))) %>% 
+  dplyr::arrange(species, agency, deploy_year) %>% 
+  gt(
+    rowname_col = "species",
+    groupname_col = "agency"
+  ) %>% 
+  cols_label(
+    deploy_year = "Year Deployed",
+    location = "Location",
+    publications = "Publication(s)",
+    permits = "Permits",
+    n_seals = "No. Seals",
+    sex = "Sex",
+    age_class = "Age Class"
+  ) %>% 
+  tab_style(
+    style = list(
+      cell_text(style = "italic")
+    ),
+    locations = cells_body(
+      columns = publications
+    )
+  ) %>% 
+  cols_align(
+    align = "center",
+    columns = everything()
+  ) %>% 
+  cols_align(
+    align = "right",
+    columns = age_class
+  ) %>% 
+  tab_header(
+    title = "Data Contributions From Collaborators and Additional Deployment Details"
+  )
+
+```
+\elandscape

### 2022-04-07 analysis/paper/London_HauloutBeahvior.Rmd
confirm inclusion of Nuiqsut; closes #24

@@
 organizations as part of collaborative investigations from 2005 through 2020.
-Seals were captured using nets and bio-loggers were attached during studies based
-in coastal Alaska communities or on research ships. Ship-based capture events
-occurred during spring near the southern ice edge in the Bering Sea. Land-based
-capture events occurred from May to October, generally between the coastal
-communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska in the
-Chukchi Sea, and Nuiqsut in the Beaufort Sea from 2005 to 2020. We refer readers
-to the primary literature for detailed capture and bio-logger attachment methods
-(see Supplemental Material, Table 1).
+Seals were captured using nets and bio-loggers were attached during studies
+based in coastal Alaska communities or on research ships. Ship-based capture
+events occurred during spring near the southern ice edge in the Bering Sea.
+Land-based capture events occurred between 2005 and 2020 from May to October,
+generally between the coastal communities of Scammon Bay, Alaska in the Bering
+Sea, Utqiaġvik, Alaska in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. We
+refer readers to the primary literature for detailed capture and bio-logger
+attachment methods (see Supplemental Material, S1).
 

### 2022-04-07 analysis/paper/London_HauloutBeahvior.Rmd
minor improvements to Data Availability

@@
 
-This manuscript was developed as a reproducible research compendium. Data and
+This manuscript was developed as a reproducible research compendium. All data and
 code are available on GitHub (https://github.com/jmlondon/berchukHaulout) and
 major versions archived at Zenodo (https://doi.org/10.5281/zenodo.4638221).
-Original data sources for telemetry are archived at the United States Animal
-Telemetry Network.
+Original data sources for telemetry are archived
+at the United States Animal Telemetry Network, archived at Movebank, or
+associated with other published manuscripts (see supplemental material S1).
 

### 2022-04-08 analysis/paper/London_HauloutBeahvior.Rmd
remove reference to trend line; closes #18

@@
 yearly maximum sea ice extent. Adult female and subadult spotted seals
-show a negative trend line, but there is no indication that the observed
+show a negative relationship, but there is no indication that the observed
 trend is meaningful (spotted seal adult female: $R^{2}$ =
@@
 a unique complicating factor. Adult ringed seals build subnivean lairs under the
-snow on top of the sea ice, where they haul out and where females rear pups until
-conditions are good for basking [@frost2004]. Thus, the wet-dry sensor on a bio-logger
-could indicate that an animal is hauled out, but if it is within a lair, it is
-not available to be detected during an aerial survey. We hope to address
-availability of ringed seals using data from satellite tags, replicate survey
-tracks, and auxiliary information about snow depth and timing of melt in a
-future study.
+snow on top of the sea ice, where they haul out and where females rear pups
+until conditions are good for basking [@frost2004]. Thus, the wet-dry sensor on
+a bio-logger could indicate that an animal is hauled out, but if it is within a
+lair, it is not available to be detected during an aerial survey. We hope to
+address availability of ringed seals using data from satellite tags, replicate
+survey tracks, and auxiliary information about snow depth and timing of melt in
+a future study.
 

### 2022-04-08 analysis/paper/London_HauloutBeahvior.Rmd
edits for verb tense; closes #29

--- 2022-04-07 analysis/paper/London_HauloutBeahvior.Rmd
@@
 logistical constraints on fieldwork and the attachment duration or operational
-life of bio-loggers. In this study, we address some of these limitations by
+life of bio-loggers. In this study, we addressed some of these limitations by
 deploying small bio-loggers designed for longer-term attachment on rear flippers
@@
 Figure \@ref(fig:dataCal) shows the distribution of all haul-out data across the
-study season for each species. Observations for ribbon and spotted seals are
+study season for each species. Observations for ribbon and spotted seals were
 concentrated in the months of May and June due to the timing of deployment
@@
 resulted in fewer data observations overall and noticeably lower in numbers May
-and June. The majority of bearded seal deployments start later in the summer and
-by May bio-loggers have either fallen off or their batteries are depleted.
+and June. The majority of bearded seal deployments started later in the summer and,
+by May, bio-loggers had either fallen off or their batteries were depleted.
 
@@
 land was outside the scope of this study. Additionally, any daily locations on
-land are likely more reflective of coordinate averaging and measurement error,
+land were likely more reflective of coordinate averaging and measurement error,
 rather than actual use of coastal features. Figure \@ref(fig:dataMap) shows the
@@
 
-Models for all species include the following effects: day-of-year, solar hour,
+Models for all species included the following effects: day-of-year, solar hour,
 temperature, wind speed, barometric pressure, precipitation, wind chill
@@
 could assess the local concentration of sea ice during these events (see
-[@olnes2020]). However, a major focus of this study is to develop models
+[@olnes2020]). However, a major focus of this study was to develop models
 applicable for aerial survey correction factors and using sea ice as a covariate
@@
 the deployments consisted of a single device attached to the rear flipper of the
-seal which do not provide at-sea locations, limiting our ability to fully
+seal which did not provide at-sea locations, limiting our ability to fully
 evaluate fine-scale habitat preferences related to sea-ice. Lastly, our study
@@
 parsimonious. For these reasons, we have elected not to use sea ice
-concentration as a predictor for haul-out probability in the present study.
-
-We assessed whether the annual variation in maximum spring sea ice extent in the
-Bering Sea influences the seasonal peak of seal haul-out probability. In
+concentration _as a predictor for haul-out probability_ in the present study.
+
+We did assess whether the annual variation in maximum spring sea ice extent in the
+Bering Sea influenced the seasonal peak of seal haul-out probability. In
 particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
@@
 Visualizing the marginal or conditional effect of an individual weather
-covariate on haul-out probability is difficult in this analysis because of the
+covariate on haul-out probability was difficult in this analysis because of the
 collinearity between covariates as well as the spatial and temporal variation
 across such a large region. The relationship of each weather covariate with
-haul-out probability, averaged over the other weather conditions, is more
+haul-out probability, averaged over the other weather conditions, was more
 variable than model coefficients would imply. That said,
@@
 across the day. In addition to a peak around local solar noon, the bearded seal
-model predicts additional haul-out activity around local midnight. In concert
+model predicted additional haul-out activity around local midnight. In concert
 with the lower magnitude of haul-out probability, bearded seal haul-out behavior
-is also more protracted throughout the spring season compared to ribbon and
+was also more protracted throughout the spring season compared to ribbon and
 spotted seals.
@@
 in April that peaks in late May for subadults and in early June for
-adults (Figure \@ref(fig:ribbonHOCal)). The behavior is clearly centered
-around local solar noon and expands to other hours later in the spring
-as seals enter their molting period. Subadults showed an earlier start
+adults (Figure \@ref(fig:ribbonHOCal)). The behavior was clearly centered
+around local solar noon and expanded to other hours later in the spring
+as seals entered their molting period. Subadults showed an earlier start
 and more intense haul-out activity in April and May. The
 young-of-the-year records begin after weaning and the model predictions
-demonstrate the ontogeny of in-water activities (e.g. diving, foraging)
-in May. Adult females have a more protracted haul-out season compared to
-males, and more time spent resting in June and July. The model suggests
-adult male ribbon seals complete their molt by the end of June. Lastly,
-for adults of both sexes there is some indication for a shift to a
+demonstrated the ontogeny of in-water activities (e.g. diving, foraging)
+in May. Adult females had a more protracted haul-out season compared to
+males, and more time was spent resting in June and July. The model suggested
+adult male ribbon seals completed their molt by the end of June. Lastly,
+for adults of both sexes there was some indication for a shift to a
 crepuscular haul-out pattern in late June and July.
@@
 Compared to ribbon seals, spotted seals showed a longer spring haul-out season
-that is less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)).
-Adults of both sexes spend considerable time in April, May and June hauled out.
-In contrast to ribbon seals, adult spotted seal males have a more protracted
+that was less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)).
+Adults of both sexes spent considerable time in April, May and June hauled out.
+In contrast to ribbon seals, adult spotted seal males had a more protracted
 haul-out season compared to females, and more time out of the water in June and
 July (Figures \@ref(fig:ribbonHOCal) and \@ref(fig:spottedHOCal)). As with
-ribbon seals, the young-of-the-year records begin after weaning and the model
-predictions demonstrate the ontogeny of in-water activities (e.g. diving,
+ribbon seals, the young-of-the-year records began after weaning and the model
+predictions demonstrated the ontogeny of in-water activities (e.g. diving,
 foraging) in May.
@@
 yearly maximum sea ice extent. Adult female and subadult spotted seals
-show a negative trend line, but there is no indication that the observed
+show a negative relationship, but there was no indication that the observed
 trend is meaningful (spotted seal adult female: $R^{2}$ =

### 2023-02-02 analysis/paper/London_HauloutBeahvior.Rmd
change title
some edits for reviewer 1

@@
 ---
-title: Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas
+title: Spring haul-out behavior of seals in the Bering and Chukchi seas
 preprint: false
@@
   concern as the quantity and quality of suitable habitat is forecast to decrease.
-  In this paper, we use data collected from satellite-linked bio-loggers deployed
-  between 2005 and 2020 to investigate the seasonal timing and environmental
+  Improved estimates of seal population abundance are needed to properly monitor
+  the impacts of these changes over time. In this paper, we use hourly
+  percent-dry data from satellite-linked bio-loggers deployed
+  between 2005 and 2020 to quantify the proportion of seals hauled out on ice,
+  which is needed to accurately estimate abundance from aerial survey counts of
+  ice-associated seals (i.e., to correct for the proportion of animals that
+  are in the water while surveys are conducted). In addition to providing
+  essential data for survey 'availability' calculations, our analysis also provides 
+  insights into the seasonal timing and environmental
   factors affecting haul-out behavior by ice-associated seals. We specifically
@@
   and spotted seals (_Phoca largha_) in the Bering and Chukchi seas. Because
-  ringed seals (_Phoca hispida_) are unique in their use of snow lairs, they were
-  not included and a separate analysis is warranted. In addition to providing
-  baseline data on phenology, these data also allow us to quantify 'availability',
-  which is needed to accurately estimate abundance from aerial survey counts of
-  these seals basking on ice (i.e., to correct for the proportion of animals that
-  are in the water while surveys are conducted). Using generalized linear mixed
+  ringed seals (_Phoca hispida_) can be out of the water but hidden from view in
+  snow lairs, they were not included in this analysis. Using generalized linear mixed
   pseudo-models to properly account for temporal autocorrelation, we fit models
@@
   speed, barometric pressure, temperature, precipitation) to examine their ability
-  to explain variation in hourly haul-out probability. We found evidence for
+  to explain variation in haul-out probability. We found evidence for
   strong diurnal and within-season patterns in haul-out behavior, as well as
@@
 ice extent, volume, and seasonal presence [@kwok2018;
-@meier2014; @overland2021; @wang2017]. These changes have ripple effects on
+@meier2014; @overland2021; @wang2017]. These changes have tangible effects on
 Arctic organisms, ecosystems, and the human communities who live in the region
@@
 ice-associated seals that depend on spring and early summer sea ice (March-June)
-in the Bering, Chukchi seas as a platform for important life history functions,
+in the Bering and Chukchi seas as a platform for important life history functions,
 such as pupping, nursing, breeding behavior, and molting [@boveng2009;
@@
 
-For ice-associated seals, haul-out behavior is tightly linked with sea ice.
+How ice-assoicated seals use sea ice as a haul-out platform varies between species.
 Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
@@
 (_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
-ice as a resting platform during the spring and early summer [@frost2018].
-Ringed seals (_Phoca hispida_) --- not included in this study --- bask on
+ice as a haul-out platform during the spring and early summer [@frost2018].
+Ringed seals (_Phoca hispida_) --- not included in this study --- haul out on
 sea-ice but also within snow lairs during winter and spring.
@@
 species. Bio-logging records of time spent out of the water provide valuable
-data for identifying covariates that explain variation in haul-out behavior. For
-instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
-bio-logging data to demonstrate greater haul-out propensity in juvenile
-crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
-in February and at times close to solar noon. In the Arctic, Von Duyke et al.
+data for identifying covariates that explain variation in haul-out behavior. 
+For instance,Von Duyke et al.
 [-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
@@
 increased gradually starting in March and peaked in May and June for bearded,
-ribbon, and spotted seals.
+ribbon, and spotted seals. Such analyses have not been limited to the Arctic. In 
+the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
+bio-logging data to demonstrate greater haul-out propensity in juvenile
+crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
+in February and at times close to solar noon.
 
@@
 [@verhoef2014a], and [@southwell2008]). However, sample sizes have often been
-insufficient to permit strong inference about life history and/or seasonal
+insufficient to permit strong inference about demographic and/or seasonal
 variation in haul-out probabilities. For instance, Bengtson and Cameron's
@@
 
-Ultimately, knowledge of trends in phenology and abundance (or life history
-surrogates such as survival and recruitment) will be necessary to make credible
-quantitative predictions about the effects of climate disruptions on the
-abundance and distribution of Arctic seal populations. Before we can construct a
-trend, however, we first require a baseline. Several studies have contributed
+Statutory requirements under, for instance, the Marine Mammal Protection Act
+(MMPA) and the Endangered Species Act (ESA) for timely estimates of population 
+abundance and trend mean improved aerial survey effort must be paired with
+improved knowledge of haul-out behavior. Several studies have contributed
 estimates of the distribution and abundance of ice-associated seal species in
@@
 affecting the observation of seals on ice. These included availability (only
-seals basking on ice were available to be counted), detection probability
+seals on ice were available to be counted), detection probability
 (observers or automated detection systems may have missed some seals on ice),
@@
 [@conn2014a; @verhoef2014a]. Refining these inferences will improve the
-accuracy of abundance estimates in the Arctic.
+accuracy of abundance estimates and, hopefully, allow credible
+predictions about the effects of climate disruptions on the abundance and
+distribution of Arctic seal populations.
 

### 2023-02-02 analysis/paper/reviewer_response.md
initial reviewer responses

---
editor_options: 
  markdown: 
    wrap: 72
---

We would like to thank each of the reviewers for taking their time to
read and offer suggestions for improving the quality and readability of
the manuscript. Here, we respond directly to each of the suggested edits
and comments. These responses are organized by reviewer.

## Reviewer 1 (A. Fishbach)

> In their general comments, the reviewer mentioned that they would like
> to see a more direct framing of the problem and to see a specific
> example of how the modeled availability would have an impact on
> population abundance estimates.

We address this by ...

> **Title**\
> The reviewer suggests removing *and aerial survey detectability* from
> the title of the manuscript and suggests adding the seasonal time
> period our study focused on.

We have changed the title to: Spring haul-out behavior of seals in the
Bering and Chukchi seas

> **Abstract**\
> *Can you make a statement in your abstract that use of modeled haulout
> behavior using covariates available from regional reananalysis data
> improved the confidence intervals on availability estimates and that
> this would improve aerial survey abundance estimation efforts?*

We don't feel comfortable making a precision comparison, simply because
there are too many differences between the analyses, including (1) the
data sets are different (the present study has substantially more
records), and (2) the model structures are different (the present study
has more parameters). As such, it will be difficult to ascribe a
specific reason to precision increases or decreases.

> *Please use direct language. You are not making findings on how seals
> rely on sea ice, rather you are improving methods to estimate their
> availability for improved aerial survey population estimation efforts.
> Here you are focused on modeling ice sea availability to aerial survey
> so that aerial survey counts may be appropriately interpreted. Please
> be clear that you are considering ice seal behaviors that may be
> telemtered as "dry out of water behaviors" that may be interpreted as
> behaviors that place the seal on top of the ice and available for
> observation by an aerial survey.*

The opening sentence of the abstract is a statement regarding the
reliance of ice-associated seals on sea-ice for a range of important
life history behaviors. We think this statement is important
introduction for readers who are not familiar with ice-associated seals.
We do agree with the reviewer's suggestion for more direct language and
have made edits to the beginning of the abstract with this in mind.

> **Abstract**
>
> *Did you use previously published data, or did you collect it?  Be
> specific using an active voice so that readers may understand your
> effort.*
>
> *For better flow, please indicate the behavioral data logging time
> scale here when you describe the data collection, not later when you
> speak to the analytical construct.*

We have edited this to simply state *data from satellite-linked
bio-loggers* ... we have also altered this sentence to specifically
mention the hourly timescale for the bio-logger data

> *The term "basking" implies a subset of activities that ice seals may
> engage in when hauled out on sea ice and not an activity that can be
> directly telemetered using your data logger.  Yet, what you are
> describing is "haul-out behavior".  Please use this term here.*

edited to remove use of 'basking'

> *Can you help your naive reader understand why you consider these
> covariates to be of interest and why others were not considered.*

The main body of the paper includes this discussion. Given the limited
space and purpose of the abstract we feel it best to simply list the
covariates of interest

> **Introduction**
>
> *Please use specific terms and not metaphors so that second-language
> English readers may better understand your writing.*

We have edited 'ripple effects' to 'tangible effects'

> *What about human communities that don't live in the region, but that
> send large LNG tankers and other vessel through the region?  The loss
> of sea ice has effects on those communities and activities too.*

There is a seemingly endless list of impacts of loss of sea ice to human
communities (and ecosystems) beyond the Arctic region to the rest of the
world. However, for this sentence, we wanted to keep the reader focused
on the Arctic region.

> *Please use direct language you mean biologically important
> functions?*

The term 'life history function' is a commonly used, direct term that
refers to biologically important functions. To ensure the reader is
clearly aware of which behaviors/functions we are referring to we
provide specific examples within the sentence.

> *What does "readiness" mean in the context of a marine mammal?*

This is a good catch and we have edited the sentence to, now, read:

"Knowledge about evolutionary constraints on the timing of reproductive
and molting behavior is generally lacking, so it is difficult to predict
**how or if** ice-associated seal species **might adapt** to future
changes (e.g., by adjusting pupping or molting schedules to earlier
dates or different locales)"

> *Perhaps you mean to offer a topic sentence, stating "How
> ice-assoicated seals use sea ice as a haulout platform varies betweens
> species".*

We have adopted this suggestion

> There is so much to state about ice-associated seals in the Pacific
> Arctic that this reference to Antarctic work seems out of place here.

We have moved this reference to later in the paragraph so the first
references are about Arctic work. While the example is from the
Antarctic, it is an important, early example of analysis using
bio-logger data to estimate haul-out behavior.

> *Do you mean age-class? The term life-history implies knowledge of
> previous behavioral efforts or occupancy of regions of interest.  
> Here you are specifically interested in age-class.*

We have edited to use the term 'demographic'

### 2023-02-03 analysis/paper/London_HauloutBeahvior.Rmd
additional reviewer #1 edits

@@
 library(ragg)
-library(stickylabeller)
 library(MetBrewer)
@@
 ice-associated seals that depend on spring and early summer sea ice (March-June)
-in the Bering and Chukchi seas as a platform for important life history functions,
-such as pupping, nursing, breeding behavior, and molting [@boveng2009;
-@boveng2013a; @cameron2010; @kelly2010]. Limited data and large knowledge gaps
-complicate predictions about the ultimate effects of changes in sea ice on the
-behavior, health, abundance, and distribution of these seals. To date, indices
-of seal health sampled during periods of declining sea ice differ regionally
-[@crawford2015; @harwood2020]. Knowledge about evolutionary constraints on the
-timing of reproductive and molting behavior is generally lacking, so it is
-difficult to predict the readiness with which ice-associated seal species will
-be able to adapt to future changes (e.g., by adjusting pupping or molting
-schedules to earlier dates or different locales). This is further complicated by
-the spatio-temporal variation in the phenology of these life history events
-within regions and throughout their full ranges. Additionally, trends in
-abundance of these species are unknown, so it is difficult to assess the effect,
-if any, declines in sea ice habitat have had, or will have, on seal demography.
+in the Bering and Chukchi seas as a platform for important life history
+functions, such as pupping, nursing, breeding behavior, and molting
+[@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data and large
+knowledge gaps complicate predictions about the ultimate effects of changes in
+sea ice on the behavior, health, abundance, and distribution of these seals. To
+date, indices of seal health sampled during periods of declining sea ice differ
+regionally [@crawford2015; @harwood2020]. Knowledge about evolutionary
+constraints on the timing of reproductive and molting behavior is generally
+lacking, so it is difficult to predict how or if ice-associated seal species
+adapt to future changes (e.g., by adjusting pupping or molting schedules to
+earlier dates or different locales). This is further complicated by the
+spatio-temporal variation in the phenology of these life history events within
+regions and throughout their full ranges. Additionally, trends in abundance of
+these species are unknown, so it is difficult to assess the effect, if any,
+declines in sea ice habitat have had, or will have, on seal demography.
+
+Statutory requirements (e.g., United States Marine Mammal Protection Act (MMPA),
+United States Endangered Species Act (ESA)) for timely estimates of population
+abundance and trend of these species mean improved aerial survey effort is
+needed -- and, must be paired with improved knowledge of haul-out behavior to
+ensure accurate estimates. Several studies have contributed estimates of the
+distribution and abundance of ice-associated seal species in the Arctic using
+aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]). Such
+abundance studies were conducted over very large areas and estimation of
+absolute abundance required making inference about numerous issues affecting the
+observation of seals on ice. These included availability (only seals on ice were
+available to be counted), detection probability (observers or automated
+detection systems may have missed some seals on ice), species misclassification,
+and possible disturbance of seals by aircraft [@conn2014a; @verhoef2014a].
+Refining these inferences will improve the accuracy of abundance estimates and,
+hopefully, allow credible predictions about the effects of climate disruptions
+on the abundance and distribution of Arctic seal populations.
 
@@
 
-The remoteness of the Bering and Chukchi seas means direct scientific observation
-of seal behavior is impractical. Thus, bio-logging devices are especially useful
-tools for collecting key information on movement and haul-out behavior for these
-species. Bio-logging records of time spent out of the water provide valuable
-data for identifying covariates that explain variation in haul-out behavior. 
-For instance,Von Duyke et al.
-[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
-changes between diurnal and nocturnal haul-out behavior of ringed seals
-previously described by Kelly and Quakenbush [-@kelly1990c] using VHF radio tags
-and direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
-propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
-[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
-larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
-time bearded seals spent hauled out progressively increased through spring and
-summer, and Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
-increased gradually starting in March and peaked in May and June for bearded,
-ribbon, and spotted seals. Such analyses have not been limited to the Arctic. In 
-the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
-bio-logging data to demonstrate greater haul-out propensity in juvenile
-crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
-in February and at times close to solar noon.
+The remoteness of the Bering and Chukchi seas means direct scientific
+observation of seal behavior is impractical. Thus, bio-logging devices are
+especially useful tools for collecting key information on movement and haul-out
+behavior for these species. Bio-logging records of time spent out of the water
+provide valuable data for identifying covariates that explain variation in
+haul-out behavior. For instance,Von Duyke et al. [-@vonduyke2020] used
+satellite-linked bio-loggers to corroborate seasonal changes between diurnal and
+nocturnal haul-out behavior of ringed seals previously described by Kelly and
+Quakenbush [-@kelly1990c] using VHF radio tags and direct observation. Bengtson
+et al. [-@bengtson2005a] documented a higher propensity for ringed seal basking
+near solar noon, as did Ver Hoef et al. [-@verhoef2014a] in an analysis of
+bearded, ribbon, and spotted seals using much larger sample sizes. Olnes et al.
+[-@olnes2020] showed that the proportion of time bearded seals spent hauled out
+progressively increased through spring and summer, and Ver Hoef et al.
+[-@verhoef2014a] found haul-out probabilities increased gradually starting in
+March and peaked in May and June for bearded, ribbon, and spotted seals. Such
+analyses have not been limited to the Arctic. In the Antarctic, Bengtson and
+Cameron [-@bengtson2004] relied on bio-logging data to demonstrate greater
+haul-out propensity in juvenile crabeater seals (_Lobodon carcinophaga_) than
+adults, with highest probabilities in February and at times close to solar noon.
 
@@
 
-Statutory requirements under, for instance, the Marine Mammal Protection Act
-(MMPA) and the Endangered Species Act (ESA) for timely estimates of population 
-abundance and trend mean improved aerial survey effort must be paired with
-improved knowledge of haul-out behavior. Several studies have contributed
-estimates of the distribution and abundance of ice-associated seal species in
-the Arctic using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
-[@verhoef2014a]). Such abundance studies were conducted over very large areas and
-estimation of absolute abundance required making inference about numerous issues
-affecting the observation of seals on ice. These included availability (only
-seals on ice were available to be counted), detection probability
-(observers or automated detection systems may have missed some seals on ice),
-species misclassification, and possible disturbance of seals by aircraft
-[@conn2014a; @verhoef2014a]. Refining these inferences will improve the
-accuracy of abundance estimates and, hopefully, allow credible
-predictions about the effects of climate disruptions on the abundance and
-distribution of Arctic seal populations.
-
 In this study, we used 16 years of bio-logging data to investigate the haul-out
@@
 haul-out analyses, includes the influence of meteorological variability, and
-investigates the potential impact of a changing icescape on the behavior of
+investigates the potential impact of changing sea-ice extent on the behavior of
 these species.
@@
 
-Data used in our study came from bio-loggers deployed on bearded, ribbon, and
-spotted seals in the Bering, Chukchi, and western Beaufort seas by multiple
-organizations as part of collaborative investigations from 2005 through 2020.
-Seals were captured using nets and bio-loggers were attached during studies
-based in coastal Alaska communities or on research ships. Ship-based capture
-events occurred during spring near the southern ice edge in the Bering Sea.
+For this study we used haul-out behavior data and location estimates from
+bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
+Chukchi, and western Beaufort seas by multiple organizations as part of
+collaborative investigations from 2005 through 2020. Seals were captured using
+nets and bio-loggers were attached during studies based in coastal Alaska
+communities or on research ships. Ship-based capture events occurred during
+spring near the southern ice edge in the Bering Sea between 2005 and 2018.
 Land-based capture events occurred between 2005 and 2020 from May to October,
@@
 
-Haul-out behavior data from 
+We subset haul-out behavior data from 
 `r species_n %>% .$n %>% sum()` bio-loggers deployed on 
@@
 `r species_n %>% filter(species == "Ribbon seal") %>% .$n` ribbon, and 
-`r species_n %>% filter(species == "Spotted seal") %>% .$n` spotted seals were
-subset to include only records from 1 March to 15 July between 
+`r species_n %>% filter(species == "Spotted seal") %>% .$n` spotted seals 
+to include only records from 1 March to 15 July between 
 `r min(lubridate::year(analysis_data$haulout_dt))` and 
@@
 'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
-Washington, USA) and either adhered to the hair on the seal or attached through the
-rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
-or head provide some benefits over flipper mounted devices (e.g. increased
-satellite transmittal rates, locations at sea) but these are lost during the
-following annual molt, which, depending on deployment date, limits the duration
-of haul-out data they provide. Additionally, bio-loggers attached to the head or
-dorsal region are often dry while the seal is floating at the surface, inducing
-a slight positive bias in the hourly percent-dry values reported by the
-bio-logger. For this study, in cases where both bio-logger types were deployed,
-hourly percent-dry observations from the flipper tag were preferred.
+Washington, USA) and we either adhered them to the hair on the seal or attached
+them through the rear flipper inter-digital webbing. The use of bio-loggers
+adhered to the back or head provides some benefits over flipper mounted devices
+(e.g. increased satellite transmittal rates, locations at sea) but these are
+lost during the following annual molt, which, depending on deployment date,
+limits the duration of haul-out data they provide. Additionally, bio-loggers
+attached to the head or dorsal region are often dry while the seal is floating
+at the surface, inducing a slight positive bias in the hourly percent-dry values
+reported by the bio-logger. For this study, in cases where both bio-logger types
+were deployed, we preferred hourly percent-dry observations from the flipper tag.>
 

### 2023-02-03 analysis/paper/London_HauloutBeahvior.Rmd
change data plot to geom_col
this was suggested by reviewer #1.

@@
 
-ggplot(data = seal_hours_tbl, aes(x = yday, y = 1, fill = n))+
-        geom_tile()+
+ggplot(data = seal_hours_tbl, aes(x = yday, y = n, fill = n))+
+  geom_col() +
         scale_fill_gradientn(colors = rev(met.brewer("Hokusai2"))) +

### 2023-02-04 analysis/paper/reviewer_response.md
additional edits re: rev #1

@@
 
-The term 'life history function' is a commonly used, direct term that
-refers to biologically important functions. To ensure the reader is
-clearly aware of which behaviors/functions we are referring to we
-provide specific examples within the sentence.
+The term 'life history function' is a commonly used, direct term.
+
+Here is a definition from Oxford:
+
+"The life history of an organism is its pattern of survival and reproduction,
+along with the traits that directly affect survival and the timing or amount of
+reproduction. Rates of survival and reproduction can be estimated across age
+classes, or across different stages in organisms with complex life cycles. Life
+history traits include growth rate; age and size at sexual maturity; the
+temporal pattern or schedule of reproduction; the number, size, and sex ratio of
+offspring; the distribution of intrinsic or extrinsic mortality rates (e.g.,
+patterns of senescence); and patterns of dormancy and dispersal."
+
+To ensure the reader is clearly aware of which life-history behaviors/functions
+we are referring to we provide specific examples within the sentence.
 
@@
 We have edited to use the term 'demographic'
+
+> Please help your reader understand why are you doing your analysis and not 
+> offer a wide and jargon rich inventory of concepts that may be considered.
+
+> Please keep your introduction focused on establishing the knowledge gap that 
+> you are seeking to fill.   Please do not lead your reader to the numerous other 
+> knowledge gaps that are not being addressed by your current study.
+
+We have removed the sentence in question
+
+> Please use the term "hauled-out".  Neither the biologger nor the aerial 
+> survey is able to distinguish among the various seal behaviors that result in 
+> them being hauled out and available for view by an aerial survey.
+
+noted, and the terms 'hauled out', 'to haul out', etc is used consistently
+throughout the manuscript
+
+> You have indicate that you would consider sea ice extent.  Here you introduce 
+> the concept of icescape.  Please introduce this concept earlier and indicate 
+> how you will incorporate this into your analysis.
+
+changed `icescape` to `sea-ice extent`
+
+> Please use an active voice for clarity.
+
+This sentence has been changed to:
+
+"For this study we used haul-out behavior data and location estimates from
+bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
+Chukchi, and western Beaufort seas by multiple organizations as part of 
+collaborative investigations from 2005 through 2020."
+
+> You summarize the years during which land-based studies applied bio-loggers.  
+> Please do the same for the offshore deployments.
+
+This sentence now reads:
+
+"Ship-based capture events occurred during spring near the southern ice edge in
+the Bering Sea between 2005 and 2018."
+
+> Please provide the geographic context of your tag deployments and data 
+> collection using a plot similar to that used by Jay et al. 
+> (2012:  10.3354/meps10057 figure 2).  Please consider using a tabulation that 
+> enables readers to understand the geographic context of the deployments 
+> similar to that used by Jay et al. 2012, table 1.
+
+We feel the spatial distribution of the actual transmitted data -- compared to
+the deployment location -- is most informative and relevant for the audience's
+understanding of how our data are distributed spatially and how well data
+represents the range of each species. We provide this visually in Figure 3 and
+adding an additional layer to that figure would to show deployments would be
+an additional complication. To ensure the audience as a clear understanding
+of our study area, we have added an additional figure with a map of the Bering
+and Chukchi sea region that indicates generally where deployments started. We
+have also added a more detailed figure to the supplemental section with each
+location specifically marked.
+
+> Please use active voice for all actions that you took in your data collection, 
+> preparation and analysis.
+
+Noted and addressed
+
+> Please move this to the discussion.  This is a summary of your data findings, 
+> not a description of your methods.
+
+We can see the reviewer's point. However, we do not feel this topic is a summary 
+of our data findings in this paper. Instead, our view
+is that this information provides the context and rationale for our
+methodological choice regarding bio-logger type and attachment methods. It is
+also important regarding the analytic methods -- preference for flipper tag
+haul-out data over those from a tag adhered to the dorsal/head. If we were to 
+move these important points to the discussion, we worry those points would be 
+left disconnected.
+
+> This is a discussion of your methods.  This is not your methods.  Please state 
+> how the data was collected in your Methods section and discuss how data 
+> collection have been uncertain or biased in your discussion section.
+
+> Please move to your discussion.
+
+We appreciate this suggestion from the reviewer and we have moved large
+portions of this paragraph to the discussion. However, some information
+regarding age/sex class determination is left here so the reader has context
+for Table 1.
+
+> Please indicate how this was determined during data collection.
+
+We have edited this sentence some to clearly communicate that field 
+determination of reproductive status (and, thus ideal delineation of
+sub-adult and adult age class) was not possible.
+
+> Please allow your readers the opportunity to understand the geographic 
+> context of these deployments through use of symbols that maybe used to 
+> cross-reference the deployment locations in a map that indicates deployment 
+> locations along with filtered movement tracks.
+
+As previously addressed, there are now two figures available that provide
+geographic context of the deployments. Figure 3 shows the geographic
+distribution of the data from the deployments. The paper is not presenting or
+discussing movement ecology and, therefore, we do not present filtered
+movement tracks. We keep the focus on haul-out behavior data.
+
+> Please consider bar plot similar to that used by Jay et al. 
+> (2012: 10.3354/meps10057 figure 3), replacing the region with the seal species.
+
+This figure has been updated with this suggestion in mind
+
+> Please state clearly how you determined the end of a deployment.  Here you 
+> indicate various elements of the haulout chronology, movement and dive 
+> chronology that may have been used, however not reader could replicate your 
+> methods base don this text.
+
+We've added the following text to the paragraph:
+
+"For example, a data record that ends with several consecutive
+days (~10+ days) of 100% dry observations and with locations indicating the tag
+was on land would be truncated to not include the final stretch of 100% dry
+observations. The vast majority of deployments end with the device detaching in the 
+water and the deployment end date is obvious. There is no perfect algorithm for
+identifying deployment end dates and each deployment in question must be considered 
+separately. While not perfect, we are confident our reliance on expert opinion and
+examination of multiple data streams provides the best option."
+
+> Deployment activity is known to disrupt normal pinniped behavior.  Did you 
+> censure some period of data collection during and after the deployment to 
+> ensure that that disruptions from the deployment activity did not bias your 
+> analysis?
+
+For these phocid species, we have no empirical data (or even anecdotal 
+observational data) on how captures may or may not disrupt 'normal' behavior.
+This is likely to be influenced by the capture method, how long the animal
+was restrained, whether the animal was with a dependent pup, whether the
+animal was molting, environmental variables such as wind and temperature, etc
+etc. Given all these unknowns, we have not applied a censure to the beginning of
+our deployment data. We can't 100% ensure this choice doesn't result in some
+level of bias. However, given the large sample size and duration of deployments
+we expect any bias imposed is negligible.
+
+> This is a result.
+> Please indicate the method that you used to distinguish "plausible" from 
+> "implausible" telemetered seal behavior.  In your Results section, you may 
+> indicate how method was able to identify occasions of implausible seal behavior 
+> and any patterns that arose relative to tag model or attachment.
+
+This sentence has been moved to the results section. Addtionally, we've provided
+a brief explanation of implausible. For these devices, the percent-dry timeline
+reported 100% dry for several weeks followed by an additional period of time
+that was more typical of expected bearded seal behavior. During the extended dry
+time the locations indicated movement of the tag that would not be simply
+drifting ice. We also observed increasing transmission rates. The working (i.e.
+unconfirmed guess) is that the wet/dry sensor surface became obscured by some
+corrosive element such that when in the water conductivity was prevented and,
+thus, recorded as dry.
+
+> Please move to the Results section.
+> Please move the discussion of how molt timing affects termination of tag 
+> deployments to the discussion section.
+
+Figure 1 (original) has been moved to the results section along with additional
+descriptive text. We have also added mention to the Discussion section.

### 2023-02-04 analysis/paper/London_HauloutBeahvior.Rmd
additional edits re: rev #1

@@
 
-# knitr::opts_chunk$set(
-#   tab.cap.style = "Table Caption",
-#   tab.cap.pre = "Table ",
-#   tab.cap.sep = ": "
-#   )
-# 
-# knitr::opts_chunk$set(
-#   fig.cap.style = "Image Caption",
-#   fig.cap.pre = "Figure ",
-#   fig.cap.sep = ": "
-#   )
-
 options(scipen = 3, digits = 3)
@@
 
-Field identification of age class can be inexact, particularly when differentiating
-subadults from adults. Sex as well as age class (non-dependent _young-of-the-year_,
-sexually immature _subadults_, and mature _adults_) were estimated at the time
-of deployment by various combinations of length, claw growth ridges
-[@mclaren1958a], and pelage characteristics for some species. In the case of
-ribbon seals, subadults often have less distinct ribbons than adults. Bearded
-seal subadults will often have a spotted pattern in the pelage that is not seen
-in adults. Spotted seal pelage cannot be used to reliably discern adults from
-subadults. Despite these challenges, we feel the age classifications used in
-this analysis are useful in testing for age-related effects on haul-out
-behavior. Seals determined to be less than one year were classified as
-young-of-the-year. Subadults are those seals likely greater than one year of age
-but not yet sexually mature. Adults are individuals that are likely sexually
-mature and older than four years. For those bio-loggers deployed on
+Sex as well as age class (non-dependent _young-of-the-year_, sexually immature
+_subadults_, and mature _adults_) were estimated at the time of deployment by
+various combinations of length, claw growth ridges [@mclaren1958a], and pelage
+characteristics for some species. Seals determined to be less than one year were
+classified as young-of-the-year. For those bio-loggers deployed on
 young-of-the-year and transmitting into the next year, the age class was
-advanced to subadult on 1 March of the following year. Table
-\@ref(tab:tblDeploy) provides a summary of these deployments and data received
-from them.
+advanced to subadult on 1 March of the following year. Subadults are those seals
+likely greater than one year of age but less than four years. Adults are
+individuals that are likely older than four years. Table \@ref(tab:tblDeploy)
+provides a summary of these deployments and data received from them.
 
@@
 Tags that fall off due to molt, attachment failure, or seal mortality and remain
-on ice or land may continue to send data to satellites; i.e., a detached bio-logger 
-that is dry (either on ice or land) will record and transmit data suggesting
-the seal is hauled out. Therefore, end times of each deployment were identified by
-examining bio-logger locations, percent-dry records, and dive behavior (if
-available) to determine when bio-loggers ceased providing data consistent with
-seal behavior. Data outside of the deployment start and end times were discarded
-prior to analysis. After approximately 9 months, 7 devices deployed on the rear
-flipper of bearded seals reported implausible hourly percent-dry data. All data
-after the first instance of unrealistic values were censored from this analysis.
-Figure \@ref(fig:dataCal) shows the distribution of all haul-out data across the
-study season for each species. Observations for ribbon and spotted seals were
-concentrated in the months of May and June due to the timing of deployment
-(April and May) and the timing of molt (May and June). During molt, seals (and
-their attached bio-loggers) spend more time out of the water and more data are
-transmitted. Molt timing also impacts when many deployments end as any
-bio-loggers adhered to the hair will fall off. Relative to the other species in
-the study, there were fewer deployments of bio-loggers on bearded seals. This
-resulted in fewer data observations overall and noticeably lower in numbers May
-and June. The majority of bearded seal deployments started later in the summer and,
-by May, bio-loggers had either fallen off or their batteries were depleted.
-
-```{r dataCal, include=TRUE, fig.height = 4, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
-Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020.
-The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out.
-Additionally, many bio-logger deployments started in April and May.
-The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study."}
-
-ggplot(data = seal_hours_tbl, aes(x = yday, y = n, fill = n))+
-  geom_col() +
-        scale_fill_gradientn(colors = rev(met.brewer("Hokusai2"))) +
-        # scale_fill_viridis_c(option="mako") +
-        scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
-                           labels = c("01 March", "01 April", "01 May", "01 June", "01 July"),
-                           expand = expansion(mult = c(0, 0))) +
-        scale_y_continuous(expand = expansion(mult = c(0, 0))) +
-  guides(fill = guide_colorbar(title.position = "bottom", 
-                               title.hjust = 0.5, barwidth = 15,
-                               barheight = 0.5)) +
-  facet_grid(species ~ .) + 
-  theme(axis.text.x = element_text(hjust = 0),
-        axis.text.y = element_blank(),
-        axis.title = element_blank(),
-        panel.grid = element_blank(),
-        legend.position = "bottom") +
-  labs(fill = "seal-hours of bio-logger data") +
-  ggtitle("Density of haul-out behavior observations varies from March through July",
-          subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")
-
-```
-
-```{r binomialdata}
-
-portion_10 <- function(d) {
-  n <- nrow(d)
-  f <- d %>% dplyr::filter(percent_dry <= 10) %>% 
-    nrow()
-  p10 <- f/n
-  return(p10)
-}
-
-portion_90 <- function(d) {
-  n <- nrow(d)
-  f <- d %>% dplyr::filter(percent_dry >= 90) %>% 
-    nrow()
-  p90 <- f/n
-  return(p90)
-}
-
-```
+on ice or land may continue to send data to satellites; i.e., a detached
+bio-logger that is dry (either on ice or land) will record and transmit data
+suggesting the seal is hauled out. Therefore, end times of each deployment were
+identified by examining bio-logger locations, percent-dry records, and dive
+behavior (if available) to determine when bio-loggers ceased providing data
+consistent with seal behavior. For example, a data record that ends with several
+consecutive days (~10+ days) of 100% dry observations and with locations
+indicating the tag was on land would be truncated to not include the final
+stretch of 100% dry observations. The vast majority of deployments end with the
+device detaching in the water and the deployment end date is obvious. There is
+no perfect algorithm for identifying deployment end dates and each deployment in
+question must be considered separately. While not perfect, we are confident our
+reliance on expert opinion and examination of multiple data streams provides the
+best option. Data outside of the deployment start and end times were discarded
+prior to analysis.
 
@@
 
+```{r binomialdata}
+
+portion_10 <- function(d) {
+  n <- nrow(d)
+  f <- d %>% dplyr::filter(percent_dry <= 10) %>% 
+    nrow()
+  p10 <- f/n
+  return(p10)
+}
+
+portion_90 <- function(d) {
+  n <- nrow(d)
+  f <- d %>% dplyr::filter(percent_dry >= 90) %>% 
+    nrow()
+  p90 <- f/n
+  return(p90)
+}
+
+```
+
 To properly account for temporal dependence within a computationally tractable
@@
 
+Figure \@ref(fig:dataCal) shows the distribution of all
+haul-out data across the study season for each species. Observations for ribbon
+and spotted seals were concentrated in the months of May and June due to the
+timing of deployment (April and May) and the timing of molt (May and June).
+During molt, seals (and their attached bio-loggers) spend more time out of the
+water and more data are transmitted. Molt timing also impacts when many
+deployments end as any bio-loggers adhered to the hair will fall off. Relative
+to the other species in the study, there were fewer deployments of bio-loggers
+on bearded seals. This resulted in fewer data observations overall and
+noticeably lower in numbers May and June. The majority of bearded seal
+deployments started later in the summer and, by May, bio-loggers had either
+fallen off or their batteries were depleted.
+
+```{r dataCal, include=TRUE, fig.height = 4, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
+Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020.
+The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out.
+Additionally, many bio-logger deployments started in April and May.
+The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study."}
+
+ggplot(data = seal_hours_tbl, aes(x = yday, y = n, fill = n))+
+  geom_col() +
+        scale_fill_gradientn(colors = met.brewer("VanGogh3")) +
+        # scale_fill_viridis_c(option="mako") +
+        scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
+                           labels = c("01 March", "01 April", "01 May", "01 June", "01 July"),
+                           expand = expansion(mult = c(0, 0))) +
+        scale_y_continuous(expand = expansion(mult = c(0, 0))) +
+  guides(fill = guide_colorbar(title.position = "bottom", 
+                               title.hjust = 0.5, barwidth = 15,
+                               barheight = 0.5)) +
+  facet_grid(species ~ .) + 
+  theme(axis.text.x = element_text(hjust = 0),
+        axis.text.y = element_blank(),
+        axis.title = element_blank(),
+        panel.grid = element_blank(),
+        legend.position = "bottom") +
+  labs(fill = "seal-hours of bio-logger data") +
+  ggtitle("Density of haul-out behavior observations varies from March through July",
+          subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")
+
+```
+
 Models omitting year effects suggested that day-of-year, solar hour, age-sex
@@
 lower sample size for adult and young-of-year age classes. As such, results are
-shown for all ages (Figure \@ref(fig:beardedHOCal)). Overall, bearded seals were
-less likely to haul out and had a bi-modal distribution of haul-out probability
-across the day. In addition to a peak around local solar noon, the bearded seal
-model predicted additional haul-out activity around local midnight. In concert
-with the lower magnitude of haul-out probability, bearded seal haul-out behavior
-was also more protracted throughout the spring season compared to ribbon and
-spotted seals.
+shown for all ages (Figure \@ref(fig:beardedHOCal)). Additionally, after
+approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
+reported implausible hourly percent-dry data (100% dry for several weeks but
+indicative of movement and increasing transmission rates (see @boveng2013)). All
+data after the first instance of unrealistic values were censored from this
+analysis. Overall, bearded seals were less likely to haul out and had a bi-modal
+distribution of haul-out probability across the day. In addition to a peak
+around local solar noon, the bearded seal model predicted additional haul-out
+activity around local midnight. In concert with the lower magnitude of haul-out
+probability, bearded seal haul-out behavior was also more protracted throughout
+the spring season compared to ribbon and spotted seals.
 
@@
 (e.g. [@verhoef2010a], [@conn2014a]), we also investigated the influence of
-sex-age class on haul-out probabilities for all species (not for bearded seals
-because of low sample size). Although ribbon and spotted seals exhibited a
+sex-age class on haul-out probabilities for both species (not for bearded seals
+because of low sample size). Field identification of age class can be inexact,
+particularly when differentiating subadults from adults. In the case of ribbon
+seals, subadults often have less distinct ribbons than adults. Spotted seal
+pelage cannot be used to reliably discern adults from subadults. Despite these
+challenges, we feel the age classifications used in this analysis are useful in
+testing for age-related effects on haul-out behavior.
+
+Although ribbon and spotted seals exhibited a
 unimodal diel haul-out pattern generally centered around local solar noon, there

### 2023-02-07 analysis/paper/reviewer_response.md
more edits

@@
 
-"The life history of an organism is its pattern of survival and reproduction,
-along with the traits that directly affect survival and the timing or amount of
-reproduction. Rates of survival and reproduction can be estimated across age
-classes, or across different stages in organisms with complex life cycles. Life
-history traits include growth rate; age and size at sexual maturity; the
-temporal pattern or schedule of reproduction; the number, size, and sex ratio of
-offspring; the distribution of intrinsic or extrinsic mortality rates (e.g.,
-patterns of senescence); and patterns of dormancy and dispersal."
+"The life history of an organism is its pattern of survival and
+reproduction, along with the traits that directly affect survival and
+the timing or amount of reproduction. Rates of survival and reproduction
+can be estimated across age classes, or across different stages in
+organisms with complex life cycles. Life history traits include growth
+rate; age and size at sexual maturity; the temporal pattern or schedule
+of reproduction; the number, size, and sex ratio of offspring; the
+distribution of intrinsic or extrinsic mortality rates (e.g., patterns
+of senescence); and patterns of dormancy and dispersal."
 
-To ensure the reader is clearly aware of which life-history behaviors/functions
-we are referring to we provide specific examples within the sentence.
+To ensure the reader is clearly aware of which life-history
+behaviors/functions we are referring to we provide specific examples
+within the sentence.
 
@@
 
-> Please help your reader understand why are you doing your analysis and not 
-> offer a wide and jargon rich inventory of concepts that may be considered.
+> Please help your reader understand why are you doing your analysis and
+> not offer a wide and jargon rich inventory of concepts that may be
+> considered.
 
-> Please keep your introduction focused on establishing the knowledge gap that 
-> you are seeking to fill.   Please do not lead your reader to the numerous other 
-> knowledge gaps that are not being addressed by your current study.
+> Please keep your introduction focused on establishing the knowledge
+> gap that you are seeking to fill. Please do not lead your reader to
+> the numerous other knowledge gaps that are not being addressed by your
+> current study.
 
@@
 > Please use the term "hauled-out". Neither the biologger nor the aerial
-> survey is able to distinguish among the various seal behaviors that result in 
-> them being hauled out and available for view by an aerial survey.
+> survey is able to distinguish among the various seal behaviors that
+> result in them being hauled out and available for view by an aerial
+> survey.
 
-noted, and the terms 'hauled out', 'to haul out', etc is used consistently
-throughout the manuscript
+noted, and the terms 'hauled out', 'to haul out', etc is used
+consistently throughout the manuscript
 
-> You have indicate that you would consider sea ice extent.  Here you introduce 
-> the concept of icescape.  Please introduce this concept earlier and indicate 
-> how you will incorporate this into your analysis.
+> You have indicate that you would consider sea ice extent. Here you
+> introduce the concept of icescape. Please introduce this concept
+> earlier and indicate how you will incorporate this into your analysis.
 
@@
 
-"For this study we used haul-out behavior data and location estimates from
-bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
-Chukchi, and western Beaufort seas by multiple organizations as part of 
-collaborative investigations from 2005 through 2020."
+"For this study we used haul-out behavior data and location estimates
+from bio-loggers deployed on bearded, ribbon, and spotted seals in the
+Bering, Chukchi, and western Beaufort seas by multiple organizations as
+part of collaborative investigations from 2005 through 2020."
 
-> You summarize the years during which land-based studies applied bio-loggers.  
+> You summarize the years during which land-based studies applied
+> bio-loggers.\
 > Please do the same for the offshore deployments.
@@
 
-"Ship-based capture events occurred during spring near the southern ice edge in
-the Bering Sea between 2005 and 2018."
+"Ship-based capture events occurred during spring near the southern ice
+edge in the Bering Sea between 2005 and 2018."
 
 > Please provide the geographic context of your tag deployments and data
-> collection using a plot similar to that used by Jay et al. 
-> (2012:  10.3354/meps10057 figure 2).  Please consider using a tabulation that 
-> enables readers to understand the geographic context of the deployments 
-> similar to that used by Jay et al. 2012, table 1.
+> collection using a plot similar to that used by Jay et al. (2012:
+> 10.3354/meps10057 figure 2). Please consider using a tabulation that
+> enables readers to understand the geographic context of the
+> deployments similar to that used by Jay et al. 2012, table 1.
 
-We feel the spatial distribution of the actual transmitted data -- compared to
-the deployment location -- is most informative and relevant for the audience's
-understanding of how our data are distributed spatially and how well data
-represents the range of each species. We provide this visually in Figure 3 and
-adding an additional layer to that figure would to show deployments would be
-an additional complication. To ensure the audience as a clear understanding
-of our study area, we have added an additional figure with a map of the Bering
-and Chukchi sea region that indicates generally where deployments started. We
-have also added a more detailed figure to the supplemental section with each
+We feel the spatial distribution of the actual transmitted data --
+compared to the deployment location -- is most informative and relevant
+for the audience's understanding of how our data are distributed
+spatially and how well data represents the range of each species. We
+provide this visually in Figure 3 and adding an additional layer to that
+figure would to show deployments would be an additional complication. To
+ensure the audience as a clear understanding of our study area, we have
+added an additional figure with a map of the Bering and Chukchi sea
+region that indicates generally where deployments started. We have also
+added a more detailed figure to the supplemental section with each
 location specifically marked.
 
-> Please use active voice for all actions that you took in your data collection, 
-> preparation and analysis.
+> Please use active voice for all actions that you took in your data
+> collection, preparation and analysis.
 
@@
 
-> Please move this to the discussion.  This is a summary of your data findings, 
-> not a description of your methods.
+> Please move this to the discussion. This is a summary of your data
+> findings, not a description of your methods.
 
-We can see the reviewer's point. However, we do not feel this topic is a summary 
-of our data findings in this paper. Instead, our view
-is that this information provides the context and rationale for our
-methodological choice regarding bio-logger type and attachment methods. It is
-also important regarding the analytic methods -- preference for flipper tag
-haul-out data over those from a tag adhered to the dorsal/head. If we were to 
-move these important points to the discussion, we worry those points would be 
-left disconnected.
+We can see the reviewer's point. However, we do not feel this topic is a
+summary of our data findings in this paper. Instead, our view is that
+this information provides the context and rationale for our
+methodological choice regarding bio-logger type and attachment methods.
+It is also important regarding the analytic methods -- preference for
+flipper tag haul-out data over those from a tag adhered to the
+dorsal/head. If we were to move these important points to the
+discussion, we worry those points would be left disconnected.
 
-> This is a discussion of your methods.  This is not your methods.  Please state 
-> how the data was collected in your Methods section and discuss how data 
-> collection have been uncertain or biased in your discussion section.
+> This is a discussion of your methods. This is not your methods. Please
+> state how the data was collected in your Methods section and discuss
+> how data collection have been uncertain or biased in your discussion
+> section.
 
@@
 portions of this paragraph to the discussion. However, some information
-regarding age/sex class determination is left here so the reader has context
-for Table 1.
+regarding age/sex class determination is left here so the reader has
+context for Table 1.
 
@@
 > context of these deployments through use of symbols that maybe used to
-> cross-reference the deployment locations in a map that indicates deployment 
-> locations along with filtered movement tracks.
+> cross-reference the deployment locations in a map that indicates
+> deployment locations along with filtered movement tracks.
 
-As previously addressed, there are now two figures available that provide
-geographic context of the deployments. Figure 3 shows the geographic
-distribution of the data from the deployments. The paper is not presenting or
-discussing movement ecology and, therefore, we do not present filtered
-movement tracks. We keep the focus on haul-out behavior data.
+As previously addressed, there are now two figures available that
+provide geographic context of the deployments. Figure 3 shows the
+geographic distribution of the data from the deployments. The paper is
+not presenting or discussing movement ecology and, therefore, we do not
+present filtered movement tracks. We keep the focus on haul-out behavior
+data.
 
-> Please consider bar plot similar to that used by Jay et al. 
-> (2012: 10.3354/meps10057 figure 3), replacing the region with the seal species.
+> Please consider bar plot similar to that used by Jay et al. (2012:
+> 10.3354/meps10057 figure 3), replacing the region with the seal
+> species.
 
@@
 
-> Please state clearly how you determined the end of a deployment.  Here you 
-> indicate various elements of the haulout chronology, movement and dive 
-> chronology that may have been used, however not reader could replicate your 
-> methods base don this text.
+> Please state clearly how you determined the end of a deployment. Here
+> you indicate various elements of the haulout chronology, movement and
+> dive chronology that may have been used, however not reader could
+> replicate your methods base don this text.
 
@@
 
-"For example, a data record that ends with several consecutive
-days (~10+ days) of 100% dry observations and with locations indicating the tag
-was on land would be truncated to not include the final stretch of 100% dry
-observations. The vast majority of deployments end with the device detaching in the 
-water and the deployment end date is obvious. There is no perfect algorithm for
-identifying deployment end dates and each deployment in question must be considered 
-separately. While not perfect, we are confident our reliance on expert opinion and
-examination of multiple data streams provides the best option."
+"For example, a data record that ends with several consecutive days
+(\~10+ days) of 100% dry observations and with locations indicating the
+tag was on land would be truncated to not include the final stretch of
+100% dry observations. The vast majority of deployments end with the
+device detaching in the water and the deployment end date is obvious.
+There is no perfect algorithm for identifying deployment end dates and
+each deployment in question must be considered separately. While not
+perfect, we are confident our reliance on expert opinion and examination
+of multiple data streams provides the best option."
 
-> Deployment activity is known to disrupt normal pinniped behavior.  Did you 
-> censure some period of data collection during and after the deployment to 
-> ensure that that disruptions from the deployment activity did not bias your 
-> analysis?
+> Deployment activity is known to disrupt normal pinniped behavior. Did
+> you censure some period of data collection during and after the
+> deployment to ensure that that disruptions from the deployment
+> activity did not bias your analysis?
 
 For these phocid species, we have no empirical data (or even anecdotal
-observational data) on how captures may or may not disrupt 'normal' behavior.
-This is likely to be influenced by the capture method, how long the animal
-was restrained, whether the animal was with a dependent pup, whether the
-animal was molting, environmental variables such as wind and temperature, etc
-etc. Given all these unknowns, we have not applied a censure to the beginning of
-our deployment data. We can't 100% ensure this choice doesn't result in some
-level of bias. However, given the large sample size and duration of deployments
-we expect any bias imposed is negligible.
+observational data) on how captures may or may not disrupt 'normal'
+behavior. This is likely to be influenced by the capture method, how
+long the animal was restrained, whether the animal was with a dependent
+pup, whether the animal was molting, environmental variables such as
+wind and temperature, etc etc. Given all these unknowns, we have not
+applied a censure to the beginning of our deployment data. We can't 100%
+ensure this choice doesn't result in some level of bias. However, given
+the large sample size and duration of deployments we expect any bias
+imposed is negligible.
 
-> This is a result.
-> Please indicate the method that you used to distinguish "plausible" from 
-> "implausible" telemetered seal behavior.  In your Results section, you may 
-> indicate how method was able to identify occasions of implausible seal behavior 
-> and any patterns that arose relative to tag model or attachment.
+> This is a result. Please indicate the method that you used to
+> distinguish "plausible" from "implausible" telemetered seal behavior.
+> In your Results section, you may indicate how method was able to
+> identify occasions of implausible seal behavior and any patterns that
+> arose relative to tag model or attachment.
 
-This sentence has been moved to the results section. Addtionally, we've provided
-a brief explanation of implausible. For these devices, the percent-dry timeline
-reported 100% dry for several weeks followed by an additional period of time
-that was more typical of expected bearded seal behavior. During the extended dry
-time the locations indicated movement of the tag that would not be simply
-drifting ice. We also observed increasing transmission rates. The working (i.e.
-unconfirmed guess) is that the wet/dry sensor surface became obscured by some
-corrosive element such that when in the water conductivity was prevented and,
-thus, recorded as dry.
+This sentence has been moved to the results section. Additionally, we've
+provided a brief explanation of implausible along with a reference for
+more details. For these devices, the percent-dry timeline reported 100%
+dry for several weeks followed by an additional period of time that was
+more typical of expected bearded seal behavior. During the extended dry
+time the locations indicated movement of the tag that would not be
+simply drifting ice. We also observed increasing transmission rates. The
+working (i.e. unconfirmed guess) is that the wet/dry sensor surface
+became obscured by some corrosive element such that when in the water
+conductivity was prevented and, thus, recorded as dry.
 
-> Please move to the Results section.
-> Please move the discussion of how molt timing affects termination of tag 
-> deployments to the discussion section.
+> Please move to the Results section. Please move the discussion of how
+> molt timing affects termination of tag deployments to the discussion
+> section.
 
-Figure 1 (original) has been moved to the results section along with additional
-descriptive text. We have also added mention to the Discussion section.
+This figure has been moved to the results section along with additional
+descriptive text. We have left the discussion regarding molt timing and
+termination of the data record here to better communicate the point to
+the reader and prevent an isolated mention in the discussion that has no
+anchor.
+
+> Thus, the a successful transmission of data was required at no greater
+> than 48 h intervals.
+
+We're not sure how to interpret this comment. The main point of the
+sentence is to explain the rationale/benefits of transmitting the
+limited selection of percent dry values. We have expanded the text to
+provide more details regarding the caching of messages and data
+transmission.
+
+> Do you mean "atmospheric" magnetic interference.\
+> Presumably "extra-terrestrial atmospheric" events (such as events on
+> Venus or Mars) played no role in the transmission of the data signal
+> to the Argos receivers in orbit around Earth.
+
+We were using a less strict definition of extra-terrestrial to refer to
+interference that wasn't Earth/surface-based. We've edited to
+'atmospheric interference'
+
+> This note belongs in the Discussion, not the Results.
+
+We feel this note is important to acknowledge in the methods because it
+speaks to the rationale for our methodological choices (bio-logger type,
+programming choices, analytical choices).
+
+> Do you mean that each cell "encompasses" and area of 50 km\^2?
+
+We have edited to the reviewer's suggested word 'encompasses'
+
+> Please use the active voice.
+
+Thank you for catching this. We have edited this sentence to active
+voice
+
+> For most tags, your daily locations are from the Argos system, not
+> from your data loggers. However, you included some FastLoc tag
+> deployments. Please re-write for clarity of what these location are.
+
+The previous paragraph describes how we determined daily locations. We
+see the reviewers point re: the data not technicaly coming from our data
+loggers. We have edited the sentence for additional clarity.
+
+> What is a "key" weather variable? Please write for clarity.
+
+We've edited this sentence removed the word 'key'
+
+> This is a procedural detail. Other readers may replicate your method
+> without needing to do this. Please omit this detail.
+
+> Readers can replicate your methods without the details on converting
+> the NARR wind vectors into wind speed.
+
+We're not sure why the reviewer would recommend omitting a 'procedural
+detail' as part of our methods. We feel it is important to clearly
+communicated our procedure for processing the large NARR data set and
+that we used the native grid size of 32km. Regarding the calculation of
+wind vectors, we include this detail because many readers, like us
+initially, may not be well versed in how wind vectors components are
+converted to wind speed.
+
+> Please use an active voice: "For all seal species we considered the
+> following variables when modeling the hourly haulout behavior..."
+
+We have edited this sentence with the reviewer's suggestion
+
+> Using the term "major focus" implies that there may be numerous minor
+> foci. Can you not state that this is the goal of the study? Ancillary
+> natural history insights can be discussed incidental to this, but are
+> not driving the research. The statutory requirements of the MMPA and
+> ESA are driving your study. These Federal laws require robust methods
+> to monitor population status. Your study seeks to build this method.
+> Discussion of natural history belong in your Discussion section and
+> not in the framing of your study. Please help the reader focus on why
+> the Federal government is doing this study.
+
+> Understanding pinniped haulout behavior is essential to designing and
+> interpreting population estimation aerial surveys. Understanding how
+> they use sea ice during a time when sea ice availability is rapidly
+> changing is also important, but is ancillary to the paper's focus.
+
+We have removed the term 'major focus' and, as the reviewer suggests,
+simply state that this is the goal of the study. We have also added
+mention of the statutory requirements to paragraph 2 of the Introduction
+section and expanded the text to acknowledge the importance of
+understanding how seals use sea ice during this period of change but,
+ultimately, it's ancillary to the focus of this paper.
+
+As for providing the reader focus on why the Federal government is doing
+this study: this is a collaborative study across multiple organizations
+and the analysis/purpose is not limited to the Federal government. We've
+incorporated the reviewers suggests to expand discussion of the
+statutory requirements of the MMPA and ESA but don't feel we need to
+single out the motivations for the Federal government.
+
+> Seals can just as happily be on foraging trips within the pack and
+> marginal sea ice as in the open water, depending on the species.
+> Suggesting that foraging trips are associated with ice-free conditions
+> is misleading.
+
+We have removed 'e.g., foraging trips' from this sentence so as not to
+mislead the reader.
+
+> This is your primary reason for not including sea ice geography in
+> your analysis: you simply lacked geolocation of much of your haulout
+> behavior data.
+>
+> Please help the reader understand how coarse your tracking data was
+> for the flipper tagged animals and what proportion of each species was
+> tracked with these flipper tags. Please also help the reader
+> understand how the tracking data relates geographically to regions
+> that may be subject to aerial survey efforts (i.e. are your tracking
+> data collected from across the regions that may be surveyed during
+> range-wide spring aerial surveys?).
+
+We have re-organized this paragraph to, hopefully, better communicate our
+choice not to explore sea-ice related covariates. As the reviewer suggests,
+we have limited geolocation data for at-sea periods of our haul-out behavior
+data -- we know the seal was in the water but don't have any estimates of
+location when only a flipper tag is attached. By leading with this statement,
+we hope that communicates our primary rationale. But, the other rationales
+are also important and we still include them.
+
+We have also provided some statistics regarding the percentage of the
+percent-dry timeline data that is not associated with a corresponding location
+estimate because the seal was in the water and no estimates were available from
+flipper mounted SPOT tags.
+
+> Please help the reader understand why you have not included a basic
+> metric of sea ice geography, such as distance from marginal ice edge
+> (from the NIC's daily Marginal Ice Zone charts), which is a metric
+> that would be available for all data collected during an aerial
+> survey.
+
+> It is a bit misleading to indicate that the NARR variables were
+> sourced from NARR alone. Their values were interpolated onto a
+> geogrpahic location that was widely variable in precision and that
+> precision relates to the both the tag attachment (glued on or flipper
+> attached) and the seal behavior (in water, hauled out).
+>
+> Indicating that your "Northing" variable was from the "Bio-logger" is
+> also misleading because your locations was determined by either the
+> Doppler inflection point time, satellite orbitography and a model of
+> the Earth (the Argos relocation method) or by reprocessing of GNSS
+> signals that were logged and then transmitted via the Argos system as
+> a data message. Really, neither of these were determined by the
+> "Bio-logger", rather they were determined by a very complex system of
+> geolocation.
+
+> Your readers may be better served by not including this "Source"
+> field, given that you describe how you obtained or estimated these
+> variable values in your methods text. Offering an over-simplified
+> "Source" in the table obscures your actual methods.
+
+We appreciate the reviewer's thoughts on this and agree with that removing
+the 'Source' column better serves the readers.

### 2023-02-07 analysis/paper/London_HauloutBeahvior.Rmd
more edits

@@
 needed -- and, must be paired with improved knowledge of haul-out behavior to
-ensure accurate estimates. Several studies have contributed estimates of the
-distribution and abundance of ice-associated seal species in the Arctic using
-aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]). Such
-abundance studies were conducted over very large areas and estimation of
-absolute abundance required making inference about numerous issues affecting the
-observation of seals on ice. These included availability (only seals on ice were
-available to be counted), detection probability (observers or automated
-detection systems may have missed some seals on ice), species misclassification,
-and possible disturbance of seals by aircraft [@conn2014a; @verhoef2014a].
-Refining these inferences will improve the accuracy of abundance estimates and,
-hopefully, allow credible predictions about the effects of climate disruptions
-on the abundance and distribution of Arctic seal populations.
+ensure appropriate survey design, robust methods, and accurate estimates.
+Several studies have contributed estimates of the distribution and abundance of
+ice-associated seal species in the Arctic using aerial surveys (e.g.,
+[@bengtson2005a], [@conn2014a], and [@verhoef2014a]). Such abundance studies
+were conducted over very large areas and estimation of absolute abundance
+required making inference about numerous issues affecting the observation of
+seals on ice. These included availability (only seals on ice were available to
+be counted), detection probability (observers or automated detection systems may
+have missed some seals on ice), species misclassification, and possible
+disturbance of seals by aircraft [@conn2014a; @verhoef2014a]. Refining these
+inferences will improve the accuracy of abundance estimates and, hopefully,
+allow credible predictions about the effects of climate disruptions on the
+abundance and distribution of Arctic seal populations.
 
@@
 reported by the bio-logger. For this study, in cases where both bio-logger types
-were deployed, we preferred hourly percent-dry observations from the flipper tag.>
+were deployed, we preferred hourly percent-dry observations from the flipper tag.
 
@@
 sensor was polled by the tag firmware every few seconds and the percent of the
-hour in the dry state was calculated (Figure \@ref(fig:examplePlot)). On board the
-bio-logger, hourly percent-dry data were rounded to the nearest 10%
-inclusive of 0% and 100% and additional values at 3% and 98%. This
+hour in the dry state was calculated (Figure \@ref(fig:examplePlot)). On board
+the bio-logger, hourly percent-dry calculations were rounded to the nearest 10%
+inclusive of 0% and 100% along with additional values at 3% and 98%. This
 compression resulted in additional data transmission as each message consisted
-of two complete 24-hour records. Bio-loggers were, typically, deployed and
-programmed in a manner to maximize data transmission during the spring pupping
-and molting period, though hourly percent-dry data were not always
-successfully transmitted. This is due to a variety of factors including
+of two complete 24-hour records. Memory capacity allowed caching of percent-dry
+records for several weeks or months and each message was transmitted several
+times to ensure reception at the satellite. Bio-loggers were, typically,
+deployed and programmed in a manner to maximize data transmission during the
+spring pupping and molting period, though hourly percent-dry data were not
+always successfully transmitted. This is due to a variety of factors including
 satellite coverage, tag availability (e.g., tags mounted to the rear flipper
 often do not transmit while at sea), tag performance, duty cycling, and
-extra-terrestrial atmospheric interference. Fortunately, missing records do not
-substantially bias inference about haul-out probabilities [@conn2012b].
+atmospheric interference. Fortunately, missing records do not substantially bias
+inference about haul-out probabilities [@conn2012b].
 
@@
 average daily location; days when the location intersected with land
-were removed from the seal's record. We recognized that bearded and spotted
+were removed from the seal's record. We recognize that bearded and spotted
 seals haul out on land. However, assessing the relationship between haul-out
@@
 variation in haul-out probabilities. These included day-of-year (for seasonal
-effects) and local solar hour (for diurnal effects). Solar hour was calculated
-using the {solaR} package [@perpinan2012] within the R statistical environment
-[@rcoreteam2021] using the daily locations. We also linked the weighted average
-daily locations from our bio-loggers to key weather values from the North
-American Regional Reanalysis (NARR) model produced by the National Centers for
-Environmental Prediction [@mesinger2006]. The NARR model assimilates
+effects) and local solar hour (for diurnal effects). We calculated local solar
+hour using the {solaR} package [@perpinan2012] within the R statistical
+environment [@rcoreteam2021] based on the weighted daily average locations. We
+also linked the weighted average daily locations to weather values from the
+North American Regional Reanalysis (NARR) model produced by the National Centers
+for Environmental Prediction [@mesinger2006]. The NARR model assimilates
 observational data to produce a long-term picture of weather over North America
-and portions of the surrounding seas. Weather variables are made
-available across the region 8 times daily. For this study, NARR weather values
-were subset to the extent of our study area over the Bering and Chukchi seas at
-3-hr intervals based on the native grid resolution of 32 km (1024 km^2^). The
+and portions of the surrounding seas. Weather variables are made available
+across the region 8 times daily. For this study, NARR weather values were subset
+to the extent of our study area over the Bering and Chukchi seas at 3-hr
+intervals based on the native grid resolution of 32 km (1024 km^2^). The
 following meteorological variables are known to affect haul-out behavior in
-other Arctic pinnipeds [@reder2003;
-@udevitz2009; @perry2017] and were interpolated and assigned to daily seal
-locations using a bilinear method: 1) air temperature at 2 m above the Earth's
-surface, 2) wind consisting of northerly and easterly vector components
-converted to wind speed using the Euclidean norm, 3) barometric pressure at sea
-level, and 4) precipitation (Table \@ref(tab:tblCovariates)).
-
-Models for all species included the following effects: day-of-year, solar hour,
+other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017] and were
+interpolated and assigned to daily seal locations using a bilinear method: 1)
+air temperature at 2 m above the Earth's surface, 2) wind consisting of
+northerly and easterly vector components converted to wind speed using the
+Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation (Table
+\@ref(tab:tblCovariates)).
+
+For all seal species, we considered the following 
+variables when modeling the hourly haul-out behavior: day-of-year, solar hour,
 temperature, wind speed, barometric pressure, precipitation, wind chill
@@
 could assess the local concentration of sea ice during these events (see
-[@olnes2020]). However, a major focus of this study was to develop models
-applicable for aerial survey correction factors and using sea ice as a covariate
-would almost certainly bias haul-out predictions towards those seals that are on
-or near ice and therefore more likely to haul out. Since aerial surveys can only
-detect seals on ice, abundance estimates would be missing a correction for those
-seals that are away from ice (e.g., on foraging trips). Additionally, many of
-the deployments consisted of a single device attached to the rear flipper of the
-seal which did not provide at-sea locations, limiting our ability to fully
-evaluate fine-scale habitat preferences related to sea-ice. Lastly, our study
-was limited to the spring season when seal haul-out tendencies are strongly
-influenced by pupping, nursing, breeding behavior, and molt and these drivers
-are likely more influential than specific sea-ice concentration. Crawford et al.
-[-@crawford2019] compared haul-out probability models for ringed seals and found
-those that only included season (and not sea-ice concentration) were the most
-parsimonious. For these reasons, we have elected not to use sea ice
-concentration _as a predictor for haul-out probability_ in the present study.
+[@olnes2020]). This, however, expands the scope of our analysis into the realm
+of habitat selection and many of our deployments consisted of a single device
+attached to the rear flipper of the seal which did not provide at-sea locations,
+limiting our ability to fully evaluate fine-scale habitat preferences related to
+sea-ice. Insight into how seals use and interact with sea-ice during an extended
+period when the availability and characteristics of sea-ice is rapidly changing
+is important but ancillary to the focus of this analysis. The focus of this
+study was to develop models applicable for aerial survey correction factors and
+using sea ice as a covariate would almost certainly bias haul-out predictions
+towards those seals that are on or near ice and therefore more likely to haul
+out. Since aerial surveys can only detect seals on ice, abundance estimates
+would be missing a correction for those seals that are away from ice. Lastly,
+our study was limited to the spring season when seal haul-out behaviors are
+strongly influenced by pupping, nursing, breeding behavior, and molt and these
+drivers are likely more influential than specific sea-ice concentration.
+Crawford et al. [-@crawford2019] compared haul-out probability models for ringed
+seals and found those that only included season (and not sea-ice concentration)
+were the most parsimonious. For these reasons, we have elected not to use sea
+ice concentration _as a predictor for haul-out probability_ in the present
+study.
 
@@
 covariate_tbl %>% 
+  select(-Source) %>% 
   flextable() %>% 

### 2023-02-27 analysis/paper/London_HauloutBeahvior.Rmd
additional responses to reviews
quite a bit in this commit. really focused on rearranging the location for some of the text, adding deployment location figure, and expanding the discussin re: not including sea ice as a covariate.

@@
 header-includes:
-  - \usepackage {newtxtext,newtxmath}
+  - \usepackage{stix2}
+  - \usepackage {booktabs}
+  - \usepackage {setspace}
+  - \onehalfspacing
+  - \usepackage [T1]{fontenc}
   - \usepackage [semibold]{sourcesanspro}
-  - \usepackage [T1]{fontenc}
   - \usepackage {hyperref}
   - \hypersetup {colorlinks, linkcolor={red!50!black}, citecolor={blue!50!black},urlcolor={blue!80!black}}
-  - \usepackage{setspace}
-  - \onehalfspacing
-  - \usepackage{pdflscape}
-  - \newcommand{\blandscape}{\begin{landscape}}
-  - \newcommand{\elandscape}{\end{landscape}}
+  - \usepackage {pdflscape}
+  - \newcommand {\blandscape}{\begin{landscape}}
+  - \newcommand {\elandscape}{\end{landscape}}
+  - \usepackage {caption}
+  - \captionsetup{font={normalsize,onehalfspacing}}
 bibliography: references.json
@@
   dpi = 300,
-  out.width = "100%"
+  fig.width = 6,
+  fig.asp = 0.618,
+  out.width = "100%",
+  tab.topcaption = TRUE
 )
@@
 ice-associated seal species in the Arctic using aerial surveys (e.g.,
-[@bengtson2005a], [@conn2014a], and [@verhoef2014a]). Such abundance studies
-were conducted over very large areas and estimation of absolute abundance
-required making inference about numerous issues affecting the observation of
-seals on ice. These included availability (only seals on ice were available to
+[@bengtson2005a], [@conn2014a], and [@verhoef2014a]) and more recent efforts
+have significantly expanded on previous survey effort. Such abundance studies
+are conducted over very large areas and estimation of absolute abundance
+requires making inference about numerous issues affecting the observation of
+seals on ice. These include availability (only seals on ice are available to
 be counted), detection probability (observers or automated detection systems may
-have missed some seals on ice), species misclassification, and possible
+miss some seals on ice), species misclassification, and possible
 disturbance of seals by aircraft [@conn2014a; @verhoef2014a]. Refining these
@@
 collaborative investigations from 2005 through 2020. Seals were captured using
-nets and bio-loggers were attached during studies based in coastal Alaska
-communities or on research ships. Ship-based capture events occurred during
-spring near the southern ice edge in the Bering Sea between 2005 and 2018.
-Land-based capture events occurred between 2005 and 2020 from May to October,
-generally between the coastal communities of Scammon Bay, Alaska in the Bering
-Sea, Utqiaġvik, Alaska in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. We
-refer readers to the primary literature for detailed capture and bio-logger
-attachment methods (see Supplemental Material, S1).
+nets and bio-loggers were attached during studies based in coastal communities
+or on research ships. Ship-based capture events occurred during spring near the
+southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
+events occurred between 2005 and 2020 from May to October, generally between the
+coastal communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska
+in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. Data from additional
+deployments along the Kamchatka peninsula in the western Bering Sea are also
+included. We refer readers to Figure \@ref(fig:deployplot) and the primary literature
+for detailed capture and bio-logger attachment methods (see Supplemental
+Material, S1).
+
+(ref:deployMapCaption) Initial deployment areas for each seal species (bearded, ribbon, and spotted seals) in the study between 2005 and 2020.
+Deployments were initiated across a range of months but only data from 1 March to 15 July were included in the analysis.
+See Figure \@ref(fig:dataMap) for the spatial distribution of observed data used in the study.
+
+```{r deployplot, include=TRUE, fig.asp = 0.45, fig.cap = "(ref:deployMapCaption)"}
+targets::tar_load(deploy_details, store = here::here("_targets"))
+
+deploy_sf <- deploy_details %>% 
+  sf::st_as_sf(coords = c('longitude','latitude'), crs = 4326) %>% 
+  sf::st_transform(3571) %>% 
+  dplyr::group_by(species,location, region) %>% 
+  dplyr::group_modify(~ concaveman::concaveman(., concavity = 4),.keep=TRUE) %>% 
+  sf::st_sf(crs = 3571) %>% 
+  sf::st_buffer(40*1000)
+
+world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
+  sf::st_transform(st_crs(deploy_sf))
+
+species_colors <- setNames(met.brewer("Isfahan2",n=3,direction=-1,override.order = TRUE),
+                           c("Bearded seal","Ribbon seal","Spotted seal"))
+species_labels <- c("Bearded seal" = paste0("<span style = 'color:",
+                                            species_colors['Bearded seal'],
+                                            ";'>Bearded seal</span>"),
+                    "Ribbon seal" = paste0("<span style = 'color:",
+                                           species_colors['Ribbon seal'],
+                                           ";'>Ribbon seal</span>"),
+                    "Spotted seal" = paste0("<span style = 'color:",
+                                            species_colors['Spotted seal'],
+                                            ";'>Spotted seal</span>")
+) %>% as_labeller()
+
+p <- ggplot() +
+  layer_spatial(deploy_sf, aes(fill=species),
+                linewidth=0) +
+  annotation_spatial(world,
+                     linewidth=0,
+                     fill = "grey80") +
+  facet_grid(cols = vars(species), labeller = species_labels,
+             switch="x") +
+  scale_fill_manual(values = met.brewer("Isfahan2",n=3,direction=-1,override.order = TRUE),
+                    aesthetics = "fill", 
+                    name = "species") + 
+  ggtitle("Areas of initial deployment location",
+          subtitle = 
+                 paste0("Bio-loggers were initially deployed on ",
+                "<span style = 'color:",species_colors['Bearded seal'],";'>bearded, </span> ", 
+                "<span style = 'color:",species_colors['Ribbon seal'],";'>ribbon, </span> ",  
+                "and ", 
+                "<span style = 'color:",species_colors['Spotted seal'],";'>spotted seals</span> ", 
+                "across multiple regions within the Bering and Chukchi seas.")) +
+  theme(legend.position = "none",
+        strip.background = element_blank(),
+        strip.placement = "outside",
+        strip.text.x = element_markdown(face="bold"),
+        plot.subtitle = element_textbox_simple(margin = ggplot2::margin(b=12)))
+
+p
+```
 
@@
   align(i = 1, align = "center", part = "header") %>% 
-  fontsize(size = 10, part = "all") %>% 
+  fontsize(size = 9, part = "all") %>% 
   width(width = 1.35) %>% 
@@
 land were likely more reflective of coordinate averaging and measurement error,
-rather than actual use of coastal features. Figure \@ref(fig:dataMap) shows the
-spatial distribution of weighted locations with available haul-out behavior data
-used for analysis of each species across the study area.
-
-```{r locs-count-mapdata}
-haulout_locs <- analysis_data %>% 
-  filter(lubridate::hour(haulout_dt) == 12)
-
-hexgrid <- sf::st_make_grid(st_bbox(haulout_locs) %>% st_as_sfc(), cellsize = 50*1000,
-                            what = "polygons", square = FALSE) 
-hexgrid <- st_sf(index = 1:length(lengths(hexgrid)), hexgrid)
-
-hexbin <- st_join(haulout_locs, hexgrid, join = st_intersects)
-
-locs_count <- hexgrid %>%
-  left_join(
-    count(hexbin, index, species) %>%
-      as_tibble() %>%
-      dplyr::select(species, index, ct=n)
-  ) %>% drop_na()
-```
-
-(ref:mapCaption) Spatial distribution of haul-out data during the months of March through July 15 for each of the three species.
-Data were collated across all years between 2005 and 2020.
-Each hexagonal cell represents an area of 50 km^2^
-
-```{r dataMap, fig.height = 3.75, include=TRUE, fig.cap="(ref:mapCaption)"}
-
-world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
-  sf::st_transform(st_crs(locs_count))
-
-ggplot() +
-  annotation_spatial(world, 
-                     size = 0, 
-                     fill = "grey70") +
-  layer_spatial(
-    data = locs_count, size = 0.125,
-    aes(fill = ct)
-  ) +
-  scale_fill_gradientn(colors = rev(met.brewer("Greek")),
-                    trans = "log10", aesthetics = "fill",
-    name = "number of locations",
-    guide = guide_colorbar(title.position = "bottom", barwidth = 15, 
-                           barheight = 0.5, title.hjust = 0.5)) + 
-  facet_grid(~ species) +
-  theme(legend.position = "bottom") +
-  ggtitle("Spatial distribution of haul-out behavior data") +
-  labs(caption = bquote("One hexagonal cell represents 50"~km^2))
-```
+rather than actual use of coastal features.
 
@@
 northerly and easterly vector components converted to wind speed using the
-Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation (Table
-\@ref(tab:tblCovariates)).
+Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation 
+(Table \@ref(tab:tblCovariates)).
 
@@
 
-Notably missing from this list of explanatory variables is any spatial-temporal
-representation of sea ice concentration, area, or extent. This may seem
-counterintuitive when modeling the haul-out behavior of seal species with such a
-close association to sea ice; seals haul out in the presence of sea ice, and we
-could assess the local concentration of sea ice during these events (see
-[@olnes2020]). This, however, expands the scope of our analysis into the realm
-of habitat selection and many of our deployments consisted of a single device
-attached to the rear flipper of the seal which did not provide at-sea locations,
-limiting our ability to fully evaluate fine-scale habitat preferences related to
-sea-ice. Insight into how seals use and interact with sea-ice during an extended
-period when the availability and characteristics of sea-ice is rapidly changing
-is important but ancillary to the focus of this analysis. The focus of this
-study was to develop models applicable for aerial survey correction factors and
-using sea ice as a covariate would almost certainly bias haul-out predictions
-towards those seals that are on or near ice and therefore more likely to haul
-out. Since aerial surveys can only detect seals on ice, abundance estimates
-would be missing a correction for those seals that are away from ice. Lastly,
-our study was limited to the spring season when seal haul-out behaviors are
-strongly influenced by pupping, nursing, breeding behavior, and molt and these
-drivers are likely more influential than specific sea-ice concentration.
-Crawford et al. [-@crawford2019] compared haul-out probability models for ringed
-seals and found those that only included season (and not sea-ice concentration)
-were the most parsimonious. For these reasons, we have elected not to use sea
-ice concentration _as a predictor for haul-out probability_ in the present
-study.
-
 We did assess whether the annual variation in maximum spring sea ice extent in the
@@
 
-```{r tblCovariates, tab.id = "tblCovariates", include=TRUE, tab.cap = "Explanatory covariates used in analyses of binary haul-out records for bearded, spotted, and ribbon seals.
+```{r tblCovariates, include=TRUE, tab.cap = "Explanatory covariates used in analyses of binary haul-out records for bearded, spotted, and ribbon seals.
 Note that we also considered select interactions (see article text) between these primary covariates.
-For instance, wind chill was represented by the interaction _temperature:wind_."}
+For instance, wind chill was represented by the interaction temperature:wind."}
 
@@
   flextable() %>% 
-  fontsize(size = 10, part = "all") %>% 
+  fontsize(size = 9, part = "all") %>% 
   hline(i=8) %>% 
@@
 
-Figure \@ref(fig:dataCal) shows the distribution of all
+Figure \@ref(fig:dataMap) shows the
+spatial distribution of weighted locations with available haul-out behavior data
+used for analysis of each species across the study area.
+
+```{r locs-count-mapdata}
+haulout_locs <- analysis_data %>% 
+  filter(lubridate::hour(haulout_dt) == 12)
+
+hexgrid <- sf::st_make_grid(st_bbox(haulout_locs) %>% st_as_sfc(), cellsize = 50*1000,
+                            what = "polygons", square = FALSE) 
+hexgrid <- st_sf(index = 1:length(lengths(hexgrid)), hexgrid)
+
+hexbin <- st_join(haulout_locs, hexgrid, join = st_intersects)
+
+locs_count <- hexgrid %>%
+  left_join(
+    count(hexbin, index, species) %>%
+      as_tibble() %>%
+      dplyr::select(species, index, ct=n)
+  ) %>% drop_na()
+```
+
+(ref:mapCaption) Spatial distribution of haul-out data during the months of March through July 15 for each of the three species.
+Data were collated across all years between 2005 and 2020.
+Each hexagonal cell represents an area of 50 km^2^
+
+```{r dataMap, include=TRUE, fig.cap="(ref:mapCaption)"}
+
+world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
+  sf::st_transform(st_crs(locs_count))
+
+ggplot() +
+  annotation_spatial(world, 
+                     size = 0, 
+                     fill = "grey80") +
+  layer_spatial(
+    data = locs_count, size = 0.125,
+    aes(fill = ct)
+  ) +
+  scale_fill_gradientn(colors = rev(met.brewer("Greek")),
+                    trans = "log10", aesthetics = "fill",
+    name = "number of locations",
+    guide = guide_colorbar(title.position = "bottom", barwidth = 15, 
+                           barheight = 0.5, title.hjust = 0.5)) + 
+  facet_grid(~ species) +
+  theme(legend.position = "bottom") +
+  ggtitle("Spatial distribution of haul-out behavior data") +
+  labs(caption = bquote("One hexagonal cell represents 50"~km^2))
+```
+
+Figure \@ref(fig:dataCal) shows the temporal distribution of all
 haul-out data across the study season for each species. Observations for ribbon
@@
 
-```{r dataCal, include=TRUE, fig.height = 4, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
+```{r dataCal, include=TRUE, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
 Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020.
@@
   geom_col() +
-        scale_fill_gradientn(colors = met.brewer("VanGogh3")) +
-        # scale_fill_viridis_c(option="mako") +
+        scale_fill_gradientn(colors = rev(met.brewer("Greek"))) +
         scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
@@
 
-```{r beardedHOCal, fig.height = 4.5, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes.
+```{r beardedHOCal, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes.
 Hour of the day was fixed at local solar noon."}
@@
 
-```{r beardedHOwx, fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
+```{r beardedHOwx, fig.asp = 1, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
 Hour of the day was fixed at local solar noon and day-of-year held at the day of peak haul-out for bearded seals.
@@
 
-```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted hourly haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model.
+```{r ribbonHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model.
 Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males.
@@
 
-```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification.
+```{r ribbonHOwx, fig.asp = 0.9, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification.
 Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for ribbon seals.
@@
 
-```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted hourly haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model.
+```{r spottedHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model.
 Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males.
@@
 
-```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
+```{r spottedHOwx,fig.asp = 0.9, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
 Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for spotted seals.
@@
 
-```{r annualHO, fig.height = 5.5, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability (colored markers) for ribbon and spotted seals across 13 years.
+```{r annualHO, fig.asp = 0.75, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability (colored markers) for ribbon and spotted seals across 13 years.
 Predictions are shown for local solar noon and under smoothed weather conditions.
@@
 and calculating availability corrections for aerial surveys.
+
+Notably missing from our list of explanatory variables is any spatial-temporal
+representation of sea ice concentration, area, or extent. This may seem
+counterintuitive when modeling the haul-out behavior of seal species with such a
+close association to sea ice; seals haul out in the presence of sea ice, and we
+could assess the local concentration of sea ice during these events (see
+[@olnes2020]). This, however, expands the scope of our analysis into the realm
+of habitat selection and many of our deployments consisted of a single device
+attached to the rear flipper of the seal which did not provide at-sea locations,
+limiting our ability to fully evaluate fine-scale habitat preferences related to
+sea-ice. Insight into how seals use and interact with sea-ice during an extended
+period when the availability and characteristics of sea-ice is rapidly changing
+is important but ancillary to the focus of this analysis. The focus of this
+study was to develop models applicable for aerial survey correction factors and
+using sea ice as a covariate would almost certainly bias haul-out predictions
+towards those seals that are on or near ice and therefore more likely to haul
+out. Since aerial surveys can only detect seals on ice, abundance estimates
+would be missing a correction for those seals that are away from ice. Consider
+the case of employing haul-out probabilities as an aerial survey correction in a
+population inhabiting two areas: an ice-free region and a region with sea ice.
+Further, assume that half of the population is in each area on average, and that
+the probability of hauling out is 1.0 in the ice covered region and 0.0 in the
+region without ice.  Denoting the total population as $N$, let us examine what
+happens when we use (1) population level availability, and (2) ice-specific
+availability as a correction factor.  We'll further assume, for sake of this
+example, that detection probability of seals on ice is 1.0, and that the
+entirety of the ice covered region is censused. In case (1), our aerial survey
+count is $C = 0.5 \cdot N$, with a population-level availability probability of
+$\hat{a} = 0.5$. A Horvitz-Thompson-type estimator for abundance is simply 
+$\hat{N} = C \div \hat{a}$, which has expectation 
+$\mathbb{E}(\hat{a}) = \frac{0.5 \cdot N}{0.5} = N$, as desired.  Now
+consider case (2).  In this case, the probability of hauling out is 1.0 for
+seals in the ice-covered area, so we have 
+$\mathbb{E}(\hat{N}) = \frac{0.5 \cdot N}{1.0} = 0.5N$. That
+is, we bias abundance because we are not accounting for seals that are away from
+ice, and therefore have a zero chance of hauling out. The same logic holds for
+any other covariate, such as distance from the ice edge, that has different
+values in the surveyed and unsurveyed habitat of the seals (assuming distance
+from ice edge is coded with opposite signs inside and outside the pack ice).
+Lastly, our study was limited to the spring season when seal haul-out behaviors
+are strongly influenced by pupping, nursing, breeding behavior, and molt and
+these drivers are likely more influential than specific sea-ice concentration.
+Crawford et al. [-@crawford2019] compared haul-out probability models for ringed
+seals and found those that only included season (and not sea-ice concentration)
+were the most parsimonious. For these reasons, we have elected not to use sea
+ice concentration _as a predictor for haul-out probability_ in the present
+study.
 

### 2023-02-27 R/create_deploy_details.R

@@ -1,0 +1,3 @@
+create_deploy_details <- function(deploy_details_file) {
+  deploy_tbl <- readr::read_csv(deploy_details_file)
+}

### 2023-02-27 _targets.R

@@ -9,7 +9,9 @@ sapply(files, source)
 # Set target-specific options such as packages.
 tar_option_set(packages = c(
   "tidyverse",
+  "ggplot2",
   "sf",
+  "ggspatial",
   "lubridate",
   "odbc",
   "RPostgres",
@@ -27,7 +29,8 @@ tar_option_set(packages = c(
   "purler",
   "wcUtils",
   "MetBrewer",
-  "patchwork"
+  "patchwork",
+  "concaveman"
   )
 )
 
@@ -42,6 +45,8 @@ list(
   tar_target(adfg_timelines_file1, "data_raw/adfg/ADFG_SealTimelineData3.csv",format = "file"),
   tar_target(adfg_timelines_file2, "data_raw/adfg/ADFG_SealTimelineData_NewRecords.csv", format = "file"),
   tar_target(adfg_timelines, adfg_clean_tl(adfg_timelines_file1, adfg_timelines_file2, adfg_deployments)),
+  tar_target(deploy_details_file, "data_raw/deploy_details.csv",format = "file"),
+  tar_target(deploy_details, create_deploy_details(deploy_details_file)),
 
   tar_target(nsb_deployments, get_nsb_deployments()),
   tar_target(nsb_locations, get_nsb_locs(nsb_deployments)),

### 2023-02-27 deploy_plot.R

@@ -1,0 +1,56 @@
+library(googlesheets4)
+library(ggplot2)
+library(rnaturalearth)
+library(ggspatial)
+library(dplyr)
+library(dbplyr)
+library(odbc)
+
+deploy_tbl <- read_sheet("1MpCiyAKWLBXufwnzW8QTnb2SsTkrKSuPlWdpWMnqoto")
+
+con <- dbConnect(
+  odbc(),
+  dsn = "PostgreSQL pep",
+  uid = keyringr::get_kc_account("pgpep_londonj"),
+  pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
+)
+
+deploy_db <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>% 
+  dplyr::select(speno, deploy_lat, deploy_long, capture_lat, capture_long) %>% 
+  dplyr::mutate(deploy_lat = if_else(is.na(deploy_lat), capture_lat, deploy_lat),
+                deploy_long = if_else(is.na(deploy_long), capture_long, deploy_long)) %>% 
+  collect()
+
+dbDisconnect(con)
+
+deploy_tbl <- deploy_tbl %>% 
+  # dplyr::select(speno, species, location, region, latitude, longitude) %>% 
+  left_join(deploy_db, by = "speno") %>% 
+  mutate(latitude = coalesce(latitude,deploy_lat),
+         longitude = coalesce(longitude, deploy_long))
+
+deploy_sf <- deploy_tbl %>% 
+  sf::st_as_sf(coords = c('longitude','latitude'), crs = 4326) %>% 
+  st_transform(3571) %>% 
+  group_by(species,location, region) %>% 
+  group_modify(~concaveman(., concavity = 4),.keep=TRUE) %>% 
+  st_sf(crs = 3571) %>% 
+  st_buffer(40*1000) 
+  
+
+
+world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
+  sf::st_transform(st_crs(deploy_sf))
+
+ggplot() +
+  layer_spatial(deploy_sf, aes(fill=species),
+                linewidth=0) +
+  annotation_spatial(world,
+                     linewidth=0,
+                     fill = "grey70") +
+  facet_grid(cols = vars(species)) +
+  scale_fill_manual(values = rev(met.brewer("Isfahan2")),
+                       aesthetics = "fill",
+                       name = "species") + 
+  theme(legend.position = "bottom")
+

### 2023-03-01 deploy_plot.R
temp files no longer needed

@@ -1,56 +1,0 @@
-library(googlesheets4)
-library(ggplot2)
-library(rnaturalearth)
-library(ggspatial)
-library(dplyr)
-library(dbplyr)
-library(odbc)
-
-deploy_tbl <- read_sheet("1MpCiyAKWLBXufwnzW8QTnb2SsTkrKSuPlWdpWMnqoto")
-
-con <- dbConnect(
-  odbc(),
-  dsn = "PostgreSQL pep",
-  uid = keyringr::get_kc_account("pgpep_londonj"),
-  pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
-)
-
-deploy_db <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>% 
-  dplyr::select(speno, deploy_lat, deploy_long, capture_lat, capture_long) %>% 
-  dplyr::mutate(deploy_lat = if_else(is.na(deploy_lat), capture_lat, deploy_lat),
-                deploy_long = if_else(is.na(deploy_long), capture_long, deploy_long)) %>% 
-  collect()
-
-dbDisconnect(con)
-
-deploy_tbl <- deploy_tbl %>% 
-  # dplyr::select(speno, species, location, region, latitude, longitude) %>% 
-  left_join(deploy_db, by = "speno") %>% 
-  mutate(latitude = coalesce(latitude,deploy_lat),
-         longitude = coalesce(longitude, deploy_long))
-
-deploy_sf <- deploy_tbl %>% 
-  sf::st_as_sf(coords = c('longitude','latitude'), crs = 4326) %>% 
-  st_transform(3571) %>% 
-  group_by(species,location, region) %>% 
-  group_modify(~concaveman(., concavity = 4),.keep=TRUE) %>% 
-  st_sf(crs = 3571) %>% 
-  st_buffer(40*1000) 
-  
-
-
-world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
-  sf::st_transform(st_crs(deploy_sf))
-
-ggplot() +
-  layer_spatial(deploy_sf, aes(fill=species),
-                linewidth=0) +
-  annotation_spatial(world,
-                     linewidth=0,
-                     fill = "grey70") +
-  facet_grid(cols = vars(species)) +
-  scale_fill_manual(values = rev(met.brewer("Isfahan2")),
-                       aesthetics = "fill",
-                       name = "species") + 
-  theme(legend.position = "bottom")
-

### 2023-03-01 data_raw/adfg/adfg_tidydata.Rmd
troubleshoot/fix labeling of SPOT/SPLASH
there were two deployments ("EB16KKA-110597","PL17SCM-158422") for which the timeline data incorrectly labeled both tags as SPOT. This should fix this issue.

@@
 ```{r}
-adfg_deployments <- readr::read_csv('ADFG_SealDeploymentTable3.csv')
-adfg_deployments_new <- readr::read_csv('ADFG_SealDeploymentTable_NewRecords.csv')
+adfg_deployments <- readr::read_csv(here::here('data_raw/adfg/ADFG_SealDeploymentTable3.csv'))
+adfg_deployments_new <- readr::read_csv(here::here('data_raw/adfg/ADFG_SealDeploymentTable_NewRecords.csv'))
 ```
@@
 
-adfg_deployments <- readr::read_csv('ADFG_SealDeploymentTable3.csv',
+adfg_deployments <- readr::read_csv(here::here('data_raw/adfg/ADFG_SealDeploymentTable3.csv'),
                                     col_types = adfg_deploy_cols) %>% 
@@
 
-adfg_deployments_new <- readr::read_csv('ADFG_SealDeploymentTable_NewRecords.csv',
+adfg_deployments_new <- readr::read_csv(here::here('data_raw/adfg/ADFG_SealDeploymentTable_NewRecords.csv'),
                                     col_types = adfg_deploy_cols) %>% 
@@
 ```{r}
-adfg_locations <- readr::read_csv('ADFG_SealLocationData3.csv')
-adfg_locations_new <- readr::read_csv('ADFG_SealLocationData_NewRecords.csv')
+adfg_locations <- readr::read_csv(here::here('data_raw/adfg/ADFG_SealLocationData3.csv'))
+adfg_locations_new <- readr::read_csv(here::here('data_raw/adfg/ADFG_SealLocationData_NewRecords.csv'))
 ```
@@
 
-adfg_locations <- readr::read_csv('ADFG_SealLocationData3.csv',
+adfg_locations <- readr::read_csv(here::here('data_raw/adfg/ADFG_SealLocationData3.csv'),
                                   col_types = adfg_locs_cols) %>% 
@@
 
-adfg_locations_new <- readr::read_csv('ADFG_SealLocationData_NewRecords.csv',
+adfg_locations_new <- readr::read_csv(here::here('data_raw/adfg/ADFG_SealLocationData_NewRecords.csv'),
                                   col_types = adfg_locs_cols) %>% 
@@
   HistType = readr::col_character(),
+  GMTDate = readr::col_character(),
   MonthNumb = readr::col_character(),
@@
                                   col_types = adfg_timeline_cols) %>% 
-  dplyr::mutate(GMTDate = as.Date(GMTDate, origin = "1899-12-30", 
-                                     tz = "UTC") %>% 
-                  lubridate::as_datetime(),
+  dplyr::mutate(GMTDate = lubridate::parse_date_time(GMTDate,orders = "mdy"),
                 Ptt = as.integer(Ptt),

### 2023-03-01 R/adfg_clean.R

@@ -162,6 +162,11 @@ adfg_clean_tl <- function(file1,file2, adfg_deployments) {
                   hist_type = HistType,
                   timeline_start_dt = GMTDate) %>%
     dplyr::rename_all(tolower) %>%
+    dplyr::mutate(tag_family = case_when(
+      ptt %in% c(110597, 158422) ~ "SPLASH",
+      ptt %in% c(167946, 137526) ~ "SPOT",
+      TRUE ~ tag_family
+    )) %>% 
     dplyr::left_join(adfg_deployments, by = c("speno","tag_family")) %>%
     mutate(species_code = case_when(
       species == "bearded" ~ "EB",

### 2023-03-01 R/create_deploy_details.R

@@ -1,3 +1,24 @@
 create_deploy_details <- function(deploy_details_file) {
   deploy_tbl <- readr::read_csv(deploy_details_file)
+  
+  con <- dbConnect(
+    odbc(),
+    dsn = "PostgreSQL pep",
+    uid = keyringr::get_kc_account("pgpep_londonj"),
+    pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
+  )
+  
+  deploy_db <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>% 
+    dplyr::select(speno, deploy_lat, deploy_long, capture_lat, capture_long) %>% 
+    dplyr::mutate(deploy_lat = if_else(is.na(deploy_lat), capture_lat, deploy_lat),
+                  deploy_long = if_else(is.na(deploy_long), capture_long, deploy_long)) %>% 
+    collect()
+  
+  dbDisconnect(con)
+  
+  deploy_tbl <- deploy_tbl %>% 
+    # dplyr::select(speno, species, location, region, latitude, longitude) %>% 
+    left_join(deploy_db, by = "speno") %>% 
+    mutate(latitude = coalesce(latitude,deploy_lat),
+           longitude = coalesce(longitude, deploy_long))
 }

### 2023-03-01 analysis/paper/reviewer_response.md
updated figures and response to reviewer

@@
 Antarctic, it is an important, early example of analysis using
-bio-logger data to estimate haul-out behavior.
+bio-logger data to estimate haul-out behavior in seals.
 
@@
 added an additional figure with a map of the Bering and Chukchi sea
-region that indicates generally where deployments started. We have also
-added a more detailed figure to the supplemental section with each
-location specifically marked.
+region that indicates generally where deployments were initiated.
 
@@
 simply drifting ice. We also observed increasing transmission rates. The
-working (i.e. unconfirmed guess) is that the wet/dry sensor surface
+working hypothesis (i.e. unconfirmed guess) is that the wet/dry sensor surface
 became obscured by some corrosive element such that when in the water
@@
 
-We've edited this sentence removed the word 'key'
+We've edited this sentence and removed the word 'key'
 
@@
 detail' as part of our methods. We feel it is important to clearly
-communicated our procedure for processing the large NARR data set and
+communicate our procedure for processing the large NARR data set and
 that we used the native grid size of 32km. Regarding the calculation of
@@
 
+We have also added a figure depicting the estimated species range for bearded,
+ribbon, and spotted seals along with general representation of recent aerial
+survey effort. This should better communicate to the reader how the spatial
+distribution of our haul-out behavior overlaps with the species range and
+survey effort.
+
 > Seals can just as happily be on foraging trips within the pack and
@@
 we hope that communicates our primary rationale. But, the other rationales
-are also important and we still include them.
+are also important and we still include them. We have also moved this section
+to the 'Discussion' in response to other reviewer comments.
 
-We have also provided some statistics regarding the percentage of the
+We have provided some statistics regarding the percentage of the
 percent-dry timeline data that is not associated with a corresponding location
@@
 flipper mounted SPOT tags.
+
+We have also added a figure depicting the estimated species range for bearded,
+ribbon, and spotted seals along with general representation of recent aerial
+survey effort. This should better communicate to the reader how the spatial
+distribution of our haul-out behavior overlaps with the species range and
+survey effort.
 

### 2023-03-01 analysis/paper/London_HauloutBeahvior.Rmd
updated figures and response to reviewer

@@
 
-How ice-assoicated seals use sea ice as a haul-out platform varies between species.
+How ice-associated seals use sea ice as a haul-out platform varies between species.
 Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
@@
 (ref:deployMapCaption) Initial deployment areas for each seal species (bearded, ribbon, and spotted seals) in the study between 2005 and 2020.
+Colored regions shown are minimum concave polygons buffered by 60 km for enhanced visibility.
+See Supplemental Material S1 for specific details.
 Deployments were initiated across a range of months but only data from 1 March to 15 July were included in the analysis.
@@
   sf::st_sf(crs = 3571) %>% 
-  sf::st_buffer(40*1000)
+  sf::st_buffer(60*1000)
 
@@
                  paste0("Bio-loggers were initially deployed on ",
-                "<span style = 'color:",species_colors['Bearded seal'],";'>bearded, </span> ", 
-                "<span style = 'color:",species_colors['Ribbon seal'],";'>ribbon, </span> ",  
+                "<span style = 'color:",species_colors['Bearded seal'],";'>**bearded**, </span> ", 
+                "<span style = 'color:",species_colors['Ribbon seal'],";'>**ribbon**, </span> ",  
                 "and ", 
-                "<span style = 'color:",species_colors['Spotted seal'],";'>spotted seals</span> ", 
-                "across multiple regions within the Bering and Chukchi seas.")) +
+                "<span style = 'color:",species_colors['Spotted seal'],";'>**spotted**</span> seals ", 
+                "across multiple regions within the Bering, Chukchi, and western Beaufort seas.")) +
   theme(legend.position = "none",

### 2023-04-26 R/adfg_clean.R
re-run analysis; troubleshoot data anomalies
after some updates to telemetry database, a few minor coding improvements, etc. I re-ran the analysis and, initially, had some errors that caused me to realize observations were present outside of the deployment period. Those issues were fixed and, now, models fit without error

@@ -48,7 +48,7 @@ adfg_clean_deploy <- function(file1,file2) {
     dplyr::mutate(species = tolower(species),
                   age = tolower(age),
                   age = case_when(
-                    age %in% c("1","3") ~ "SUBADULT",
+                    age %in% c("1","3") ~ "subadult",
                     TRUE ~ age
                   ),
                   sex = tolower(sex)) %>%

### 2023-04-26 R/get_locs_sf.R

@@ -26,6 +26,7 @@ get_locs_sf <- function(adfg_locations, nsb_locations) {
     left_join(deployments_db, by = 'deployid') %>%
     left_join(spenos_db, by = 'speno') %>%
     filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
+    filter(between(locs_dt,deploy_dt,end_dt)) %>%
     filter(lubridate::month(locs_dt) %in% c(3,4,5,6,7)) %>%
     mutate(unique_day =
              glue::glue("{lubridate::year(locs_dt)}",

### 2023-04-26 R/get_timeline_data.R

@@ -23,6 +23,7 @@ get_timeline_data <- function(adfg_timelines, nsb_timelines) {
     left_join(spenos_db, by = 'speno') %>%
     filter(!deployid %in% c("PL2017_9001_16U2112")) %>%
     filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
+    filter(between(timeline_start_dt,deploy_dt,end_dt)) %>% 
     collect() %>%
     filter(lubridate::month(timeline_start_dt) %in% c(3,4,5,6,7)) %>%
     mutate(unique_day =

### 2023-04-26 _targets.R

@@ -34,6 +34,12 @@ tar_option_set(packages = c(
   )
 )
 
+# tar_make_clustermq() configuration (okay to leave alone):
+options(clustermq.scheduler = "multicore")
+
+# tar_make_future() configuration (okay to leave alone):
+future::plan(future.callr::callr, workers = 3)
+
 # End this file with a list of target objects.
 list(
   tar_target(adfg_deploy_file1, "data_raw/adfg/ADFG_SealDeploymentTable3.csv",format = "file"),

### 2023-06-30 R/create_db_input_data.R
switch to Rpostgres pkg; update connect

@@ -52,16 +52,17 @@ create_source_data <- function(locs_sf, timeline_data) {
     rename(haulout_dt = timeline_start_dt) %>%
     dplyr::select(speno,species,age,sex,haulout_dt,percent_dry,n_tags,fill_xy)
 
-  stopifnot(
-    "PEP Postgres Database Not Available; did you start VPN? ;)" =
-      is_up("161.55.120.122", "5432")
-  )
 
-  con <- dbConnect(RPostgres::Postgres(),dbname = 'pep',
-                   host = '161.55.120.122',
-                   port = 5432,
-                   user = keyringr::get_kc_account("pgpep_sa"),
-                   password = keyringr::decrypt_kc_pw("pgpep_sa"))
+  tryCatch({
+    con <- dbConnect(RPostgres::Postgres(),
+                     dbname = 'pep', 
+                     host = Sys.getenv('PEP_PG_IP'),
+                     user = keyringr::get_kc_account("pgpep_londonj"),
+                     password = keyringr::decrypt_kc_pw("pgpep_londonj"))
+  },
+  error = function(cond) {
+    print("Unable to connect to Database.")
+  })
   on.exit(dbDisconnect(con))
 
   st_write(obj = tbl_percent_locs,

### 2023-06-30 R/create_deploy_details.R

@@ -1,12 +1,17 @@
 create_deploy_details <- function(deploy_details_file) {
   deploy_tbl <- readr::read_csv(deploy_details_file)
   
-  con <- dbConnect(
-    odbc(),
-    dsn = "PostgreSQL pep",
-    uid = keyringr::get_kc_account("pgpep_londonj"),
-    pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
-  )
+  tryCatch({
+    con <- dbConnect(RPostgres::Postgres(),
+                      dbname = 'pep', 
+                      host = Sys.getenv('PEP_PG_IP'),
+                      user = keyringr::get_kc_account("pgpep_londonj"),
+                      password = keyringr::decrypt_kc_pw("pgpep_londonj"))
+  },
+  error = function(cond) {
+    print("Unable to connect to Database.")
+  })
+  on.exit(dbDisconnect(con))
   
   deploy_db <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>% 
     dplyr::select(speno, deploy_lat, deploy_long, capture_lat, capture_long) %>% 

### 2023-06-30 R/create_grid.R

@@ -1,15 +1,17 @@
 create_grid_sf <- function(data_sf) {
-  stopifnot(
-    "PEP Postgres Database Not Available; did you start VPN? ;)" =
-      is_up("161.55.120.122", "5432")
-  )
 
-  con <- dbConnect(RPostgres::Postgres(),dbname = 'pep',
-                   host = '161.55.120.122',
-                   port = 5432,
-                   user = keyringr::get_kc_account("pgpep_londonj"),
-                   password = keyringr::decrypt_kc_pw("pgpep_londonj"))
-  on.exit(dbDisconnect(con))
+  tryCatch({
+    con <- dbConnect(RPostgres::Postgres(),
+                     dbname = 'pep', 
+                     host = Sys.getenv('PEP_PG_IP'),
+                     user = keyringr::get_kc_account("pgpep_londonj"),
+                     password = keyringr::decrypt_kc_pw("pgpep_londonj"))
+  },
+  error = function(cond) {
+    print("Unable to connect to Database.")
+  })
+  on.exit(odbc::dbDisconnect(con))
+  
   sf::st_read(con,
               query = "SELECT * FROM base.geo_analysis_grid") %>%
     st_filter(st_bbox(data_sf) %>% st_as_sfc())

### 2023-06-30 R/get_locs_sf.R
update filter to remove active tags

@@ -1,16 +1,16 @@
 get_locs_sf <- function(adfg_locations, nsb_locations) {
 
-  stopifnot(
-    "PEP Postgres Database Not Available; did you start VPN? ;)" =
-      pingr::is_up("161.55.120.122", "5432")
-  )
-
-  con <- dbConnect(
-    odbc(),
-    dsn = "PostgreSQL pep",
-    uid = keyringr::get_kc_account("pgpep_londonj"),
-    pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
-  )
+  tryCatch({
+    con <- dbConnect(RPostgres::Postgres(),
+                      dbname = 'pep', 
+                      host = Sys.getenv('PEP_PG_IP'),
+                      user = keyringr::get_kc_account("pgpep_londonj"),
+                      password = keyringr::decrypt_kc_pw("pgpep_londonj"))
+  },
+  error = function(cond) {
+    print("Unable to connect to Database.")
+  })
+  on.exit(dbDisconnect(con))
 
   deployments_db <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>%
     dplyr::select(speno, deployid, tag_family, deploy_dt, end_dt) %>%
@@ -19,14 +19,13 @@ get_locs_sf <- function(adfg_locations, nsb_locations) {
   spenos_db <- tbl(con, in_schema("capture","for_telem")) %>% collect()
 
   locs_qry <- "SELECT deployid, ptt, type, error_radius, locs_dt, quality, geom as geometry
-              FROM telem.geo_wc_locs_qa;"
+              FROM telem.geo_wc_locs_qa WHERE qa_status != 'tag_actively_transmitting';"
 
   locs_sf <- read_sf(con, query = locs_qry) %>%
     filter(!deployid %in% c("PL2017_9001_16U2112")) %>%
     left_join(deployments_db, by = 'deployid') %>%
     left_join(spenos_db, by = 'speno') %>%
     filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
-    filter(between(locs_dt,deploy_dt,end_dt)) %>%
     filter(lubridate::month(locs_dt) %in% c(3,4,5,6,7)) %>%
     mutate(unique_day =
              glue::glue("{lubridate::year(locs_dt)}",

### 2023-06-30 R/get_timeline_data.R

@@ -1,18 +1,18 @@
 get_timeline_data <- function(adfg_timelines, nsb_timelines) {
-
-  stopifnot(
-    "PEP Postgres Database Not Available; did you start VPN? ;)" =
-      pingr::is_up("161.55.120.122", "5432")
-  )
-
-  con <- dbConnect(
-    odbc(),
-    dsn = "PostgreSQL pep",
-    uid = keyringr::get_kc_account("pgpep_londonj"),
-    pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
-  )
+  
+  tryCatch({
+    con <- dbConnect(RPostgres::Postgres(),
+                     dbname = 'pep', 
+                     host = Sys.getenv('PEP_PG_IP'),
+                     user = keyringr::get_kc_account("pgpep_londonj"),
+                     password = keyringr::decrypt_kc_pw("pgpep_londonj"))
+  },
+  error = function(cond) {
+    print("Unable to connect to Database.")
+  })
 
   timeline_db <- tbl(con, in_schema("telem","tbl_wc_histos_timeline_qa")) %>%
+    dplyr::filter(qa_status != 'tag_actively_transmitting') %>%
     dplyr::select(deployid,timeline_start_dt, percent_dry)
   deployments_db <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>%
     dplyr::select(speno, deployid, tag_family, deploy_dt, end_dt)
@@ -23,7 +23,6 @@ get_timeline_data <- function(adfg_timelines, nsb_timelines) {
     left_join(spenos_db, by = 'speno') %>%
     filter(!deployid %in% c("PL2017_9001_16U2112")) %>%
     filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
-    filter(between(timeline_start_dt,deploy_dt,end_dt)) %>% 
     collect() %>%
     filter(lubridate::month(timeline_start_dt) %in% c(3,4,5,6,7)) %>%
     mutate(unique_day =

### 2023-06-30 analysis/paper/London_HauloutBeahvior.Rmd
re-run analysis; troubleshoot data anomalies
after some updates to telemetry database, a few minor coding improvements, etc. I re-ran the analysis and, initially, had some errors that caused me to realize observations were present outside of the deployment period. Those issues were fixed and, now, models fit without error
update suppl table of deployments

@@
 ```{r seal-hours}
-create_yday_groups <- function(analysis_data) {
-  d <- analysis_data %>%
-    tibble::as_tibble() %>%
-    dplyr::mutate(yday = lubridate::yday(haulout_dt)) %>%
-    dplyr::select(yday) %>%
-    dplyr::distinct() %>%
-    dplyr::arrange(yday)
-
-  d <- split(d$yday, ceiling(seq_along(d$yday) / 8)) %>%
-    dplyr::as_tibble() %>%
-    tidyr::pivot_longer(
-      cols = 1:18,
-      names_to = "col_group",
-      values_to = "yday",
-      names_transform = list(col_group = as.integer),
-      values_transform = list(yday = as.integer)
-    ) %>%
-    dplyr::arrange(yday) %>% 
-    transform(row_group = letters[1:8]) %>% 
-    dplyr::filter(col_group < 18)
-  return(d)
-}
-
-yday_groups <- create_yday_groups(analysis_data)
+# create_yday_groups <- function(analysis_data) {
+#   d <- analysis_data %>%
+#     tibble::as_tibble() %>%
+#     dplyr::mutate(yday = lubridate::yday(haulout_dt)) %>%
+#     dplyr::select(yday) %>%
+#     dplyr::distinct() %>%
+#     dplyr::arrange(yday)
+#   
+#   d <- split(d$yday, ceiling(seq_along(d$yday) / 8)) %>%
+#     dplyr::as_tibble() %>%
+#     tidyr::pivot_longer(
+#       cols = 1:15,
+#       names_to = "col_group",
+#       values_to = "yday",
+#       names_transform = list(col_group = as.integer),
+#       values_transform = list(yday = as.integer)
+#     ) %>%
+#     dplyr::arrange(yday) %>% 
+#     transform(row_group = letters[1:8])
+#   return(d)
+# }
+# 
+# yday_groups <- create_yday_groups(analysis_data)
 
@@
 review and editing
-* **Justin Crawford**: investigation, methodology, validation, data curation, 
+* **Justin A. Crawford**: investigation, methodology, validation, data curation, 
 writing: review and editing
@@
 The following table provides additional details about the timing, location, and
-agencies responsible for the bio-logger deployments used in this study.
+institutions responsible for the bio-logger deployments used in this study.
 
@@
 
+options(
+  gargle_oauth_cache = here::here(".secrets"),
+  gargle_oauth_email = TRUE
+)
+
+gs4_auth()
+
 deploy_tbl <- read_sheet("1MpCiyAKWLBXufwnzW8QTnb2SsTkrKSuPlWdpWMnqoto")
 
-deploy_tbl %>% 
+deploy_tbl <- deploy_tbl %>% 
   dplyr::select(species, agency, deploy_year, permit, location, speno, sex, age, publications) %>% 
@@
                   case_when(
-                    agency %in% c('ADFG') ~ "Alaska Dept. Fish & Game",
-                    agency %in% c('NSB') ~ "North Slope Burough",
-                    agency %in% c('PEP') ~ "NOAA Alaska Fisheries Sci. Ctr",
-                    agency %in% c('NPWC') ~ "N. Pacific Wildlife Consortium"
+                    agency %in% c('ADFG') ~ "ADFG",
+                    agency %in% c('NSB') ~ "NSB",
+                    agency %in% c('PEP') ~ "NMFS",
+                    agency %in% c('NPWC') ~ "NPWC"
                   )) %>% 
@@
   gt(
-    rowname_col = "species",
-    groupname_col = "agency"
+    groupname_col = "species"
   ) %>% 
   cols_label(
+    agency = "Institution",
     deploy_year = "Year Deployed",
@@
     title = "Data Contributions From Collaborators and Additional Deployment Details"
-  )
+  ) %>%
+  tab_footnote(footnote = "ADF&G = Alaska Department of Fish and Game",
+               locations = cells_body(columns = agency, 
+                                      rows = agency == "ADFG")) %>%
+  tab_footnote(footnote = "NSB = North Slope Borough",
+               locations = cells_body(columns = agency, 
+                                      rows = agency == "NSB")) %>%
+  tab_footnote(footnote = "NMFS = NOAA National Marine Fisheries Service",
+               locations = cells_body(columns = agency, 
+                                      rows = agency == "NMFS")) %>%
+  tab_footnote(footnote = "NPWC = North Pacific Wildlife Consortium",
+               locations = cells_body(columns = agency, 
+                                      rows = agency == "NPWC"))
 

### 2023-09-06 analysis/paper/London_HauloutBeahvior.Rmd
finalize model specification & data
did some exploration of various model specifications (what if day^4, what if less interactions, etc) and still feel this is the best specification

@@
 
-For all seal species, we considered the following 
-variables when modeling the hourly haul-out behavior: day-of-year, solar hour,
-temperature, wind speed, barometric pressure, precipitation, wind chill
-(represented by a *wind:temperature* interaction [@udevitz2009]), and
-day-of-year and time-of-day (solar hour) interactions to permit diurnal patterns
-to change from March to July. Ribbon and spotted seal models included age-sex
-class and interactions between day-of-year and age-sex class, but we omitted
-these from bearded seal models due to poor representation of age-sex classes
-(Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
-(and an interaction with day-of-year), since bearded seals occupy a substantial
-range during the spring and we were interested in possible differences in the
-timing of haul-out behavior along a latitudinal gradient. We omitted the
-latitudinal effect from ribbon and spotted seal models because, during the
-spring, these species are most prevalent near the southern ice edge in the
-Bering Sea [@conn2014a].
+For all seal species, we considered the following variables when modeling the
+hourly haul-out behavior: day-of-year, solar hour, temperature, wind speed,
+barometric pressure, precipitation, and wind chill (represented by a
+*wind:temperature* interaction [@udevitz2009]). Ribbon and spotted seal models
+included age-sex class and interactions between day-of-year and age-sex class,
+but we omitted these from bearded seal models due to poor representation of
+age-sex classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a
+latitudinal effect (and an interaction with day-of-year), since bearded seals
+occupy a substantial range during the spring and we were interested in possible
+differences in the timing of haul-out behavior along a latitudinal gradient. We
+omitted the latitudinal effect from ribbon and spotted seal models because,
+during the spring, these species are most prevalent near the southern ice edge
+in the Bering Sea [@conn2014a].
 
@@
 
-We followed Ver Hoef et al. [-@verhoef2014a] in using linear, quadratic, and
-cubic effects of day-of-year to represent temporal changes in behavior. However,
-unlike previous models for harbor seals [@london2012] and ice-associated seals
-[@verhoef2014a], which treated hour-of-day as a 24-level categorical variable to
-capture diurnal cycles, we adopted a continuous formulation based on Fourier
-series that provides a flexible model while preserving the inherent circularity
-needed for time-of-day effects (i.e., hour 0 should be equal to hour 24). It
-also represents hour-of-day with 6 parameters, which is a considerable reduction
-when compared to a 24-parameter variable, especially when fitting models including
-interactions between hour-of-day and other variables (e.g., age-sex class,
-day-of-year). According to this approach, we used the following specification
-for hour-of-day effects:
+We followed Ver Hoef et al. [-@verhoef2014a] in using linear, quadratic, and cubic
+effects of day-of-year to represent temporal changes
+in behavior during the season. However, unlike previous models for harbor seals
+[@london2012] and ice-associated seals [@verhoef2014a], which treated
+hour-of-day as a 24-level categorical variable to capture diurnal cycles, we
+adopted a continuous formulation based on Fourier series that provides a
+flexible model while preserving the inherent circularity needed for time-of-day
+effects (i.e., hour 0 should be equal to hour 24). It also represents
+hour-of-day with 6 parameters, which is a considerable reduction when compared
+to a 24-parameter variable. According to this approach, we used the following
+specification for hour-of-day effects:
 
@@
   gargle_oauth_cache = here::here(".secrets"),
-  gargle_oauth_email = TRUE
+  gargle_oauth_email = "josh.london@noaa.gov"
 )

### 2023-09-06 R/create_model_input.R

@@ -14,12 +14,13 @@ create_model_input <- function(dat.sf) {
          sin3 = sin(pi*solar_hour/4),
          cos3 = sin(pi*solar_hour/4),
          day = (yday-120)/10,
-         day2 = day^2,
-         day3 = day^3
+         day_2 = day^2,
+         day_3 = day^3,
+         day_4 = day^4
   ) %>%
   mutate(precip = rast_acpcp,
-         temp2 = (rast_air2m-270)/27,
-         temp0 = (rast_airsfc-270)/27,
+         temp2m = (rast_air2m-273.15)/27,
+         temp0 = (rast_airsfc-273.15)/27,
          pressure = (rast_prmsl-100000)/10000,
          wind = sqrt(rast_uwnd^2 + rast_vwnd^2)/10
   ) %>%

### 2023-09-06 R/create_newdata.R

@@ -18,7 +18,7 @@ create_newdata <- function(data, age_sex) {
           data = data, 
           method = "REML")
     gam.temp <-
-      gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), 
+      gam(temp2m ~ s(yday) + s(as.numeric(solar_hour)), 
           data = data, 
           method = "REML")
     gam.wind <-
@@ -45,7 +45,7 @@ create_newdata <- function(data, age_sex) {
       solar_hour = rep(0:23, each = n_days),
       yday = rep(start_day:end_day, times = 24),
       northing = mean(data$northing),
-      temp2 = temp_pred,
+      temp2m = temp_pred,
       wind = wind_pred,
       pressure = baro_pred,
       precip = precip_pred
@@ -59,8 +59,9 @@ create_newdata <- function(data, age_sex) {
         cos3 = sin(pi * solar_hour / 4),
       ) %>%
       mutate(day = (yday - 120) / 10,
-             day2 = day ^ 2,
-             day3 = day ^ 3)
+             day_2 = day ^ 2,
+             day_3 = day ^ 3,
+             day_4 = day ^ 4)
     df_list[[a_s]] <- df
   }
   if(length({{age_sex}}) > 1) {
@@ -84,7 +85,7 @@ create_newdata_margins <- function(data, age_sex,
       solar_hour = solar_hour,
       yday = yday,
       northing = mean(data$northing),
-      temp2 = if(term %in% c("temp2")) sample(unique(data[,"temp2"]),150) else mean(data[,"temp2"]),
+      temp2m = if(term %in% c("temp2m")) sample(unique(data[,"temp2m"]),150) else mean(data[,"temp2m"]),
       wind = if(term %in% c("wind")) sample(unique(data[,"wind"]),150) else mean(data[,"wind"]),
       pressure = if(term %in% c("pressure")) sample(unique(data[,"pressure"]),150) else mean(data[,"pressure"]),
       precip = if(term %in% c("precip")) sample(unique(data[,"precip"]),150) else mean(data[,"precip"])
@@ -98,8 +99,9 @@ create_newdata_margins <- function(data, age_sex,
         cos3 = sin(pi * solar_hour / 4),
       ) %>%
       mutate(day = (yday - 120) / 10,
-             day2 = day ^ 2,
-             day3 = day ^ 3)
[…]

### 2023-09-06 R/fit_bearded_models.R

@@ -7,15 +7,14 @@ create_bearded_data <- function(dat.sf) {
 }
 
 fit_bearded <- function (HO_bearded) {
-  glmmLDTS(fixed.formula = dry ~ sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + day + day2+ day3 + northing + temp2 + wind + pressure + precip + wind*temp2 +
-             sin1*day + cos1*day + sin2*day + cos2*day + sin3*day + cos3*day +
-             sin1*day2 + cos1*day2 + sin2*day2 + cos2*day2 + sin3*day2 + cos3*day2 +
-             northing*day + northing*day2,
+  glmmLDTS(fixed.formula = dry ~  
+             sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
+             day + day_2 + day_3 + day_4 + 
+             temp2m*wind + pressure + precip + northing +
+             northing:day + northing:day_2 + northing:day_3 + northing_day_4,
            random.formula = dry ~ speno,
            data = HO_bearded,
            EstMeth = "REML",
            timecol = "time_vec",
-           #ridge.reg = "global",
-           #lambda = 0.5,
            group.vec = "ar1_id")
 }

### 2023-09-06 R/fit_ribbon_models.R

@@ -8,52 +8,58 @@ create_ribbon_data <- function(dat.sf) {
 
 fit_ribbon <- function(HO_ribbon) {
   glmmLDTS(
-    fixed.formula = dry ~ age_sex + sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + day +
-      day2 + day3 + temp2 + wind + pressure + precip + wind * temp2 +
-      sin1 * day + cos1 * day + sin2 * day + cos2 * day + sin3 *
-      day + cos3 * day +
-      sin1 * day2 + cos1 * day2 + sin2 * day2 + cos2 *
-      day2 + sin3 * day2 + cos3 * day2 +
-      age_sex:day + age_sex:day2 + age_sex:day3,
+    fixed.formula = dry ~ age_sex + 
+      sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
+      day + day_2 + day_3 +  
+      temp2m*wind + pressure + precip + 
+      age_sex:day + age_sex:day_2 + age_sex:day_3 + 
+      sin1:day + sin1:day_2 + sin1:day_3 +
+      cos1:day + cos1:day_2 + cos1:day_3 +
+      sin2:day + sin2:day_2 + sin2:day_3 +
+      cos2:day + cos2:day_2 + cos2:day_3 +
+      sin3:day + sin3:day_2 + sin3:day_3 +
+      cos3:day + cos3:day_2 + cos3:day_3,
     random.formula = dry ~ speno,
     data = HO_ribbon,
     EstMeth = "REML",
     timecol = "time_vec",
-    #ridge.reg = "global",
-    #lambda = 0.5,
     group.vec = "ar1_id"
   )
 }
 
 
-fit_ribbon_noagesex <- function(HO_ribbon) {
-  glmmLDTS(
-    fixed.formula = dry ~ sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + day + day2 +
-      day3 + temp2 + wind + pressure + precip + wind * temp2 +
-      sin1 * day + cos1 * day + sin2 * day + cos2 *
-      day + sin3 * day + cos3 * day +
-      sin1 * day2 + cos1 * day2 + sin2 * day2 + cos2 *
-      day2 + sin3 * day2 + cos3 * day2,
-    random.formula = dry ~ speno,
-    data = HO_ribbon,
-    EstMeth = "REML",
-    timecol = "time_vec",
[…]

### 2023-09-06 R/fit_spotted_models.R

@@ -7,87 +7,50 @@ create_spotted_data <- function(dat.sf) {
 }
 
 fit_spotted <- function(HO_spotted) {
-  glmmLDTS(fixed.formula = dry ~ age_sex + sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + day + day2+ day3 + temp2 + wind + pressure + precip + wind*temp2 +
-             sin1*day + cos1*day + sin2*day + cos2*day + sin3*day + cos3*day +
-             sin1*day2 + cos1*day2 + sin2*day2 + cos2*day2 + sin3*day2 + cos3*day2 +
-             age_sex:day + age_sex:day2 + age_sex:day3,
+  glmmLDTS(fixed.formula = dry ~ age_sex + 
+             sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
+             day + day_2 + day_3 +  
+             temp2m*wind + pressure + precip + 
+             age_sex:day + age_sex:day_2 + age_sex:day_3 + 
+             sin1:day + sin1:day_2 + sin1:day_3 +
+             cos1:day + cos1:day_2 + cos1:day_3 +
+             sin2:day + sin2:day_2 + sin2:day_3 +
+             cos2:day + cos2:day_2 + cos2:day_3 +
+             sin3:day + sin3:day_2 + sin3:day_3 +
+             cos3:day + cos3:day_2 + cos3:day_3,
            random.formula = dry ~ speno,
            data = HO_spotted,
            EstMeth = "REML",
            timecol = "time_vec",
-           #ridge.reg = "global",
-           #lambda = 0.5,
            group.vec = "ar1_id")
 }
 
-fit_spotted_ssrd <- function(HO_spotted) {
-  glmmLDTS(fixed.formula = dry ~ age_sex + day + day2+ day3 + temp2 + wind + pressure + precip + wind*temp2 +
-             era5_ssrd_watts + era5_ssrd_watts:day + era5_ssrd_watts:day2 +
-             age_sex:day + age_sex:day2 + age_sex:day3,
-           random.formula = dry ~ speno,
-           data = HO_spotted,
-           EstMeth = "REML",
-           timecol = "time_vec",
-           #ridge.reg = "global",
-           #lambda = 0.5,
-           group.vec = "ar1_id")
-}
-
-fit_spotted_bam <- function(HO_spotted) {
-  mgcv::bam(
[…]

### 2023-09-06 R/wx_plots.R

@@ -3,14 +3,14 @@ plot_ribbon_wx <- function(ribbon_fit_obj) {
   p_temp <- create_ribbon_newdata(ribbon_fit_obj, 
                                   margins=TRUE, 
                                   solar_hour = 12, yday = 135, 
-                                  term = "temp2") %>% 
+                                  term = "temp2m") %>% 
     mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
            month = lubridate::month(date,label = TRUE),
            day = lubridate::day(date)) %>% 
     filter(age_sex != "YOUNG OF YEAR") %>% 
     group_by(age_sex) %>% 
     
-    ggplot(aes(temp2*27, ho_prob, color = age_sex)) +
+    ggplot(aes(temp2m*27, ho_prob, color = age_sex)) +
     geom_pointrange(aes(ymin = lower95, ymax = upper95), 
                     size = 0.25, alpha = 1, stroke = 0) +
     scale_color_manual(values= met.brewer("Egypt",n=3),
@@ -102,14 +102,14 @@ plot_spotted_wx <- function(spotted_fit_obj) {
   p_temp <- create_spotted_newdata(spotted_fit_obj, 
                                    margins=TRUE, 
                                    solar_hour = 12, yday = 135, 
-                                   term = "temp2") %>% 
+                                   term = "temp2m") %>% 
     mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
            month = lubridate::month(date,label = TRUE),
            day = lubridate::day(date)) %>% 
     filter(age_sex != "YOUNG OF YEAR") %>% 
     group_by(age_sex) %>% 
     
-    ggplot(aes(temp2*27, ho_prob, color = age_sex)) +
+    ggplot(aes(temp2m*27, ho_prob, color = age_sex)) +
     geom_pointrange(aes(ymin = lower95, ymax = upper95), 
                     size = 0.25, alpha = 1, stroke = 0) +
     scale_color_manual(values= met.brewer("Egypt",n=3),
@@ -201,13 +201,13 @@ plot_bearded_wx <- function(bearded_fit_obj) {
   p_temp <- create_bearded_newdata(bearded_fit_obj, 
                                   margins=TRUE, 
                                   solar_hour = 12, yday = 135, 
-                                  term = "temp2") %>% 
+                                  term = "temp2m") %>% 
[…]

### 2023-09-06 _targets.R

@@ -71,13 +71,9 @@ list(
 
   tar_target(ribbon_model_data, create_ribbon_data(model_data)),
   tar_target(ribbon_fit, fit_ribbon(ribbon_model_data)),
-  tar_target(ribbon_noagesex_fit, fit_ribbon_noagesex(ribbon_model_data)),
   tar_target(ribbon_year_fit, fit_ribbon_year(ribbon_model_data)),
   tar_target(spotted_model_data, create_spotted_data(model_data)),
   tar_target(spotted_fit, fit_spotted(spotted_model_data)),
-  tar_target(spotted_fit_bam, fit_spotted_bam(spotted_model_data)),
-  tar_target(spotted_fit_bam_discrete, fit_spotted_bam_discrete(spotted_model_data)),
-  tar_target(spotted_noagesex_fit, fit_spotted_noagesex(spotted_model_data)),
   tar_target(spotted_year_fit, fit_spotted_year(spotted_model_data)),
   tar_target(bearded_model_data, create_bearded_data(model_data)),
   tar_target(bearded_fit, fit_bearded(bearded_model_data)),

### 2023-09-11 analysis/paper/London_HauloutBeahvior.Rmd
additional rationale for no sea-ice covariate
just some additional editing of this section.
clarify haul-out records were WC standard
expand why separate species models

@@
 
-Haul-out behavior data were recorded and transmitted via the Argos satellite
+Haul-out behavior data were recorded in a manner standard across Wildlife Computers
+bio-loggers and transmitted via the Argos satellite
 network as hourly percent-dry timelines. For each hour of a day, the wet/dry
@@
 
-To properly account for temporal dependence within a computationally tractable
-framework, we used generalized linear mixed pseudo-models (GLMPMs;
+To properly account for temporal dependence and to take advantage of computational
+efficiency, we used generalized linear mixed pseudo-models (GLMPMs;
 [@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
@@
 used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
-different model formulations for our data, and owing to the large number of
-records, we fit separate models to bearded, ribbon, and spotted seal data sets.
+different model formulations for our data and we fit separate models to bearded, 
+ribbon, and spotted seal data sets as we expected differing behavior by species. 
+Separate models for each species were also
+needed because a single, very large data set proved computationally intractable.
 In our first model formulation and for each species, we fitted a year-independent
@@
 out. Since aerial surveys can only detect seals on ice, abundance estimates
-would be missing a correction for those seals that are away from ice. Consider
+would be missing a correction for those seals that are away from ice.
+
+Consider
 the case of employing haul-out probabilities as an aerial survey correction in a
@@
 from ice edge is coded with opposite signs inside and outside the pack ice).
+
 Lastly, our study was limited to the spring season when seal haul-out behaviors
@@
 
+deploy_tbl
+
 ```

### 2023-09-15 R/fit_bearded_models.R
remove year:day_3 from model spec
computationally ustable

@@ -7,7 +7,7 @@ create_bearded_data <- function(dat.sf) {
 }
 
 fit_bearded <- function (HO_bearded) {
-  glmmLDTS(fixed.formula = dry ~ age_sex + 
+  glmmLDTS(fixed.formula = dry ~
              sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
              day + day_2 + day_3 +  
              temp2m*wind + pressure + precip + 

### 2023-09-15 R/fit_ribbon_models.R

@@ -54,9 +54,9 @@ fit_ribbon_year <- function(HO_ribbon) {
              sin2:day + sin2:day_2 + sin2:day_3 +
              cos2:day + cos2:day_2 + cos2:day_3 +
              sin3:day + sin3:day_2 + sin3:day_3 +
-             cos3:day + cos3:day_2 + cos3:day_3,
+             cos3:day + cos3:day_2 + cos3:day_3 +
            age_sex:day + age_sex:day_2 + age_sex:day_3 +  
-             year:day + year:day_2 + year:day_3,
+             year:day + year:day_2,
            random.formula = dry ~ speno,
            data = HO_ribbon,
            EstMeth = "REML",

### 2023-09-15 R/fit_spotted_models.R

@@ -48,9 +48,9 @@ fit_spotted_year <- function(HO_spotted) {
              sin2:day + sin2:day_2 + sin2:day_3 +
              cos2:day + cos2:day_2 + cos2:day_3 +
              sin3:day + sin3:day_2 + sin3:day_3 +
-             cos3:day + cos3:day_2 + cos3:day_3,
+             cos3:day + cos3:day_2 + cos3:day_3 +
              age_sex:day + age_sex:day_2 + age_sex:day_3 +  
-             year:day + year:day_2 + year:day_3,
+             year:day + year:day_2,
            random.formula = dry ~ speno,
            data = HO_spotted,
            EstMeth = "REML",

### 2023-09-15 analysis/paper/Letter_Response_to_Reviewers.qmd
initial repsone to reviewers quarto letter

---
title: " "
subtitle: Maunscript 2022:04:72978
format: 
  pdf:
    include-in-header:
      text: |
        \usepackage[dvipsnames]{xcolor}
filters:
   - latex-environment
environments: [author-response]
editor: visual
---

```{=tex}
\newenvironment{author-response}{\color{Bittersweet}}

```
Josh M. London, PhD\newline
7600 Sand Point Way NE\newline Seattle, Washington, 98115\newline josh.london\@noaa.gov\newline

**Manuscript 2022:04:72978**

\hfill \break

Dr. Prange,

First, please accept my apologies for the significant delay in returning revisions on this manuscript. A number of life and work priorities interfered with a more timely response schedule and I sincerely appreciate the flexibility *PeerJ* extends to authors. We are commited to a quick turnaround on any future revisions or tasks associated with this manuscript and we look forward to a final publication in *PeerJ*.

We feel, while many of the comments and suggestions from reviewers were minor, they deserved appropriate time, consideration, and exploration. We hope our responses here reflect that consideration and we are grateful to the reviewers for their thoughtful comments. Lastly, we also used this opportunity to improve the quality of the underlying telemetry data set and expand the reproducibility of the analysis to meet or exceed expectations for open science, data availability, and code availability.

Below, please find detailed responses to each of the reviewer comments. In addition to this letter, several of the suggestions are documented as specific issues within the GitHub repository and the full suite of updates are all contained and documented within a single pull request. We have also provided a pdf document that indicates all changes between the revised manuscript and the submitted manuscript.

Thank you, again, for your time and consideration. And, additionally, thank you to the reviewers for their time. It is much appreciated and we feel the manuscript is a better product as a result.

Regards

Josh M London, PhD

\pagebreak

# Response to Reviewers

Here, we respond directly to each of the suggested edits and comments. These responses are organized by reviewer. All responses are prefaced with **\[author response\]** and line numbers in the original submission are used as reference. \hfill \break

\hrule

## **Reviewer 1 (A. Fishbach)**

> *The authors have put substantial efforts into setting the context for their analysis, citing relevant studies. I would like to see a more direct approach in the framing of their problem: the improvements of abundance estimates required by the Marine Mammal Protection Act and Endangered Species Act. I would like to see a specific example of how the modeled availability would have affect population abundance estimates by apply the availability model to aerial survey data and providing an adjusted abundance estimate.*

::: author-response
**\[author response\]** We address this in the manuscript by introducing the importance of the statutory requirements of the MMPA and ESA early in the second paragraph of the introduction. We hope this frames the important link in the mind of the reader between the work described here and the need for improved abundance estimates.

Regarding the desire for specific examples of how the modeled availability would have impacted population abundance estimates from aerial survey data, we don't have the ability to provide such an example because the aerial survey data are not yet published and available. Such an example and discussion might be better suited for subsequent papers that focus on presentation of the aerial survey data and the abundance estimation process.
:::

> *The findings are appear robust and intuitive. The authors need to improve the presentation of the findings to improve clarity and help the reader understand how they addressed the knowledge gap.*

::: author-response
**\[author response\]** The reviewer provided numerous comments and suggestions annotated throughout the PDF that address the concern of presentation and clarity. In the next section we provide responses to those annotations as best as possible
:::

### **Reviewer 1 (A. Fishbach) - PDF Annotated Comments & Suggested Edits**

> ***Title:** The reviewer suggests removing and aerial survey detectability from the title of the manuscript and suggests adding the seasonal time period our study focused on.*

::: author-response
**\[author response\]** We have changed the title to: Spring haul-out behavior of seals in the Bering and Chukchi seas
:::

> ***Abstract:** Can you make a statement in your abstract that use of modeled haulout behavior using covariates available from regional reananalysis data improved the confidence intervals on availability estimates and that this would improve aerial survey abundance estimation efforts?*

::: author-response
**\[author response\]** We don't feel comfortable making a precision comparison, simply because there are too many differences between the analyses, including (1) the data sets are different (the present study has substantially more records), and (2) the model structures are different (the present study has more parameters). As such, it will be difficult to ascribe a specific reason to precision increases or decreases.
:::

> *Please use direct language. You are not making findings on how seals rely on sea ice, rather you are improving methods to estimate their availability for improved aerial survey population estimation efforts. Here you are focused on modeling ice sea availability to aerial survey so that aerial survey counts may be appropriately interpreted. Please be clear that you are considering ice seal behaviors that may be telemtered as "dry out of water behaviors" that may be interpreted as behaviors that place the seal on top of the ice and available for observation by an aerial survey.*

::: author-response
**\[author response\]** The opening sentence of the abstract is a statement regarding the reliance of ice-associated seals on sea-ice for a range of important life history behaviors. We think this statement is important introduction for readers who are not familiar with ice-associated seals. We do agree with the reviewer's suggestion for more direct language and have made edits to the beginning of the abstract with this in mind.
:::

> ***Abstract:** Did you use previously published data, or did you collect it?  Be specific using an active voice so that readers may understand your effort.*
>
> *For better flow, please indicate the behavioral data logging time scale here when you describe the data collection, not later when you speak to the analytical construct.*

::: author-response
**\[author response\]** We have edited this to simply state *data from satellite-linked bio-loggers* ... we have also altered this sentence to specifically mention the hourly timescale for the bio-logger data
:::

> *The term "basking" implies a subset of activities that ice seals may engage in when hauled out on sea ice and not an activity that can be directly telemetered using your data logger.  Yet, what you are describing is "haul-out behavior".  Please use this term here.*

::: author-response
**\[author response\]** edited to remove use of 'basking'
:::

> *Can you help your naive reader understand why you consider these covariates to be of interest and why others were not considered.*

::: author-response
**\[author response\]** The main body of the paper includes this discussion. Given the limited space and purpose of the abstract we feel it best to simply list the covariates of interest
:::

> ***Line 41:** Please use specific terms and not metaphors so that second-language English readers may better understand your writing.*

::: author-response
**\[author response\]** We have edited 'ripple effects' to 'tangible effects'
:::

> ***Line 41:** What about human communities that don't live in the region, but that send large LNG tankers and other vessel through the region?  The loss of sea ice has effects on those communities and activities too.*

::: author-response
**\[author response\]** There is a seemingly endless list of impacts of loss of sea ice to human communities (and ecosystems) beyond the Arctic region to the rest of the world. However, for this sentence, we wanted to keep the reader focused on the Arctic region.
:::

> ***Line 44:** Please use direct language you mean biologically important functions?*

::: author-response
**\[author response\]** The term 'life history function' is a commonly used, direct term.

Here is a definition from Oxford:

"The life history of an organism is its pattern of survival and reproduction, along with the traits that directly affect survival and the timing or amount of reproduction. Rates of survival and reproduction can be estimated across age classes, or across different stages in organisms with complex life cycles. Life history traits include growth rate; age and size at sexual maturity; the temporal pattern or schedule of reproduction; the number, size, and sex ratio of offspring; the distribution of intrinsic or extrinsic mortality rates (e.g., patterns of senescence); and patterns of dormancy and dispersal."

To ensure the reader is clearly aware of which life-history behaviors/functions we are referring to we provide specific examples within the sentence.
:::

> ***Line 50:** What does "readiness" mean in the context of a marine mammal?*

::: author-response
**\[author response\]** This is a good catch and we have edited the sentence to, now, read:

"Knowledge about evolutionary constraints on the timing of reproductive and molting behavior is generally lacking, so it is difficult to predict how or if ice-associated seal species might adapt to future changes (e.g., by adjusting pupping or molting schedules to earlier dates or different locales)"
:::

> ***Line 56:** Perhaps you mean to offer a topic sentence, stating "How ice-assoicated seals use sea ice as a haulout platform varies betweens species".*

::: author-response
**\[author response\]** We have adopted this suggestion
:::

> ***Line 66-68:** There is so much to state about ice-associated seals in the Pacific Arctic that this reference to Antarctic work seems out of place here.*

::: author-response
**\[author response\]** We have moved this reference to later in the paragraph so the first references are about Arctic work. While the example is from the Antarctic, it is an important, early example of analysis using bio-logger data to estimate haul-out behavior in seals.
:::

> ***Line 77:** suggested edit*

::: author-response
**\[author response\]** We have adopted the suggested edit
:::

> ***Line 84:** Do you mean age-class? The term life-history implies knowledge of previous behavioral efforts or occupancy of regions of interest.   Here you are specifically interested in age-class.*

::: author-response
**\[author response\]** We have edited to use the term 'demographic'
:::

> ***Line 92:** Please help your reader understand why are you doing your analysis and not offer a wide and jargon rich inventory of concepts that may be considered.*

> ***Line 93-94:** Please keep your introduction focused on establishing the knowledge gap that you are seeking to fill. Please do not lead your reader to the numerous other knowledge gaps that are not being addressed by your current study.*

::: author-response
**\[author response\]** We have removed the sentence in question
:::

> ***Line 99:** Please use the term "hauled-out". Neither the biologger nor the aerial survey is able to distinguish among the various seal behaviors that result in them being hauled out and available for view by an aerial survey.*

::: author-response
**\[author response\]** noted, and the terms "hauled out", "to haul out", etc is used consistently throughout the manuscript
:::

> ***Line 114:** You have indicate that you would consider sea ice extent. Here you introduce the concept of icescape. Please introduce this concept earlier and indicate how you will incorporate this into your analysis.*

::: author-response
**\[author response\]** changed "icescape" to "sea-ice extent"
:::

> ***Line 117:** Please use an active voice for clarity.*

::: author-response
**\[author response\]** This sentence has been changed to:

"For this study we used haul-out behavior data and location estimates from bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering, Chukchi, and western Beaufort seas by multiple organizations as part of collaborative investigations from 2005 through 2020."
:::

> ***Line 120:** You summarize the years during which land-based studies applied bio-loggers.\
> Please do the same for the offshore deployments.*

::: author-response
**\[author response\]** This sentence now reads:

"Ship-based capture events occurred during spring near the southern ice edge in the Bering Sea between 2005 and 2018."
:::

> ***Line 124:** Please provide the geographic context of your tag deployments and data collection using a plot similar to that used by Jay et al. (2012: 10.3354/meps10057 figure 2). Please consider using a tabulation that enables readers to understand the geographic context of the deployments similar to that used by Jay et al. 2012, table 1.*

::: author-response
**\[author response\]** We feel the spatial distribution of the actual transmitted data -- compared to the deployment location -- is most informative and relevant for the audience's understanding of how our data are distributed spatially and how well data represents the range of each species. We provide this visually in Figure 3 and adding an additional layer to that figure to show deployments would be an additional complication. To ensure the audience has a clear understanding of our study area, we have added an additional figure with a map of the Bering and Chukchi sea region that indicates generally where deployments were initiated.
:::

> ***Line 126:** Please use active voice for all actions that you took in your data collection, preparation and analysis.*

::: author-response
**\[author response\]** Noted and addressed
:::

> ***Line 129-135:** Please move this to the discussion. This is a summary of your data findings, not a description of your methods.*

::: author-response
**\[author response\]** We can see the reviewer's point. However, we do not feel this topic is a summary of our data findings in this paper. Instead, our view is that this information provides the context and rationale for our methodological choice regarding bio-logger type and attachment methods. It is also important regarding the analytic methods -- preference for flipper tag haul-out data over those from a tag adhered to the dorsal/head. If we were to move these important points to the discussion, we worry those points would be left disconnected.
:::

> ***Line 136:** This is a discussion of your methods. This is not your methods. Please state how the data was collected in your Methods section and discuss how data collection have been uncertain or biased in your discussion section.*

> ***Line 141-143:** Please move to your discussion.*

::: author-response
**\[author response\]** We appreciate this suggestion from the reviewer and we have moved large portions of this paragraph to the discussion. However, some information regarding age/sex class determination is left here so the reader has context for Table 1.
:::

> ***Line 144:** Please indicate how this was determined during data collection.*

::: author-response
**\[author response\]** We have edited this sentence some to clearly communicate that field determination of reproductive status (and, thus ideal delineation of sub-adult and adult age class) was not possible.
:::

> ***Line 147-148:** Please allow your readers the opportunity to understand the geographic context of these deployments through use of symbols that maybe used to cross-reference the deployment locations in a map that indicates deployment locations along with filtered movement tracks.*

::: author-response
**\[author response\]** As previously addressed, there are now two figures available that provide geographic context of the deployments. Figure 3 shows the geographic distribution of the data from the deployments. The paper is not presenting or discussing movement ecology and, therefore, we do not present filtered movement tracks. We keep the focus on haul-out behavior data.
:::

> ***Figure 1:** Please consider bar plot similar to that used by Jay et al. (2012: 10.3354/meps10057 figure 3), replacing the region with the seal species.*

::: author-response
**\[author response\]** This figure has been updated with this suggestion in mind
:::

> ***Line 151:** Please state clearly how you determined the end of a deployment. Here you indicate various elements of the haulout chronology, movement and dive chronology that may have been used, however not reader could replicate your methods base don this text.*

::: author-response
**\[author response\]** We've added the following text to the paragraph:

"For example, a data record that ends with several consecutive days (\~10+ days) of 100% dry observations and with locations indicating the tag was on land would be truncated to not include the final stretch of 100% dry observations. The vast majority of deployments end with the device detaching in the water and the deployment end date is obvious. There is no perfect algorithm for identifying deployment end dates and each deployment in question must be considered separately. While not perfect, we are confident our reliance on expert opinion and examination of multiple data streams provides the best option."
:::

> ***Line 152:** Deployment activity is known to disrupt normal pinniped behavior. Did you censure some period of data collection during and after the deployment to ensure that that disruptions from the deployment activity did not bias your analysis?*

::: author-response
**\[author response\]** For these phocid species, we have no empirical data (or even anecdotal observational data) on how captures may or may not disrupt 'normal' behavior. This is likely to be influenced by the capture method, how long the animal was restrained, whether the animal was with a dependent pup, whether the animal was molting, environmental variables such as wind and temperature, etc etc. Given all these unknowns, we have not applied a censure to the beginning of our deployment data. We can't 100% ensure this choice doesn't result in some level of bias. However, given the large sample size and duration of deployments we expect any bias imposed is negligible.
:::

> ***Line 154:** This is a result. Please indicate the method that you used to distinguish "plausible" from "implausible" telemetered seal behavior. In your Results section, you may indicate how method was able to identify occasions of implausible seal behavior and any patterns that arose relative to tag model or attachment.*

::: author-response
**\[author response\]** This sentence has been moved to the results section. Additionally, we've provided a brief explanation of implausible along with a reference for more details. For these devices, the percent-dry timeline reported 100% dry for several weeks followed by an additional period of time that was more typical of expected bearded seal behavior. During the extended dry time the locations indicated movement of the tag that would not be simply drifting ice. We also observed increasing transmission rates. The working hypothesis (i.e. unconfirmed guess) is that the wet/dry sensor surface became obscured by some corrosive element such that when in the water conductivity was prevented and, thus, recorded as dry.
:::

> ***Line 156-164:** Please move to the Results section. Please move the discussion of how molt timing affects termination of tag deployments to the discussion section.*

::: author-response
**\[author response\]** This figure has been moved to the results section along with additional descriptive text. We have left the discussion regarding molt timing and termination of the data record here to better communicate the point to the reader and prevent an isolated mention in the discussion that has no anchor.
:::

> ***Line 169-170:** Thus, the a successful transmission of data was required at no greater than 48 h intervals.*

::: author-response
**\[author response\]** We're not sure how to interpret this comment. The main point of the sentence is to explain the rationale/benefits of transmitting the limited selection of percent dry values. We have expanded the text to provide more details regarding the caching of messages and data transmission.
:::

> ***Line 174:** Do you mean "atmospheric" magnetic interference.\
> Presumably "extra-terrestrial atmospheric" events (such as events on Venus or Mars) played no role in the transmission of the data signal to the Argos receivers in orbit around Earth.*

::: author-response
**\[author response\]** We were using a less strict definition of extra-terrestrial to refer to interference that wasn't Earth/surface-based. We've edited to 'atmospheric interference'
:::

> ***Line 174-175:** This note belongs in the Discussion, not the Results.*

::: author-response
**\[author response\]** We feel this note is important to acknowledge in the methods because it speaks to the rationale for our methodological choices (bio-logger type, programming choices, analytical choices).
:::

> ***Figure 3:** Do you mean that each cell "encompasses" and area of 50 km\^2?*

::: author-response
**\[author response\]** We have edited to the reviewer's suggested word 'encompasses'
:::

> ***Line 197:** Please use the active voice.*

::: author-response
**\[author response\]** Thank you for catching this. We have edited this sentence to active voice
:::

> ***Line 198-199:** For most tags, your daily locations are from the Argos system, not from your data loggers. However, you included some FastLoc tag deployments. Please re-write for clarity of what these location are.*

::: author-response
**\[author response\]** The previous paragraph describes how we determined daily locations. We see the reviewers point re: the data not technicaly coming from our data loggers. We have edited the sentence for additional clarity.
:::

> ***Line 199:** What is a "key" weather variable? Please write for clarity.*

::: author-response
**\[author response\]** We've edited this sentence and removed the word 'key'
:::

> ***Line 203:** This is a procedural detail. Other readers may replicate your method without needing to do this. Please omit this detail.*

> ***Line 207-208:** Readers can replicate your methods without the details on converting the NARR wind vectors into wind speed.*

::: author-response
**\[author response\]** We're not sure why the reviewer would recommend omitting a 'procedural detail' as part of our methods. We feel it is important to clearly communicate our procedure for processing the large NARR data set and that we used the native grid size of 32km. Regarding the calculation of wind vectors, we include this detail because many readers, like us initially, may not be well versed in how wind vectors components are converted to wind speed.
:::

> ***Line 210:** Please use an active voice: "For all seal species we considered the following variables when modeling the hourly haulout behavior..."*

::: author-response
**\[author response\]** We have edited this sentence with the reviewer's suggestion
:::

> ***Line 224:** Using the term "major focus" implies that there may be numerous minor foci. Can you not state that this is the goal of the study? Ancillary natural history insights can be discussed incidental to this, but are not driving the research. The statutory requirements of the MMPA and ESA are driving your study. These Federal laws require robust methods to monitor population status. Your study seeks to build this method. Discussion of natural history belong in your Discussion section and not in the framing of your study. Please help the reader focus on why the Federal government is doing this study.*

> *Understanding pinniped haulout behavior is essential to designing and interpreting population estimation aerial surveys. Understanding how they use sea ice during a time when sea ice availability is rapidly changing is also important, but is ancillary to the paper's focus.*

::: author-response
**\[author response\]** We have removed the term 'major focus' and, as the reviewer suggests, simply state that this is the goal of the study. We have also added mention of the statutory requirements to paragraph 2 of the Introduction section and expanded the text to acknowledge the importance of understanding how seals use sea ice during this period of change but, ultimately, it's ancillary to the focus of this paper.

As for providing the reader focus on why the Federal government is doing this study: this is a collaborative study across multiple organizations and the analysis/purpose is not limited to the Federal government. We've incorporated the reviewers suggests to expand discussion of the statutory requirements of the MMPA and ESA but don't feel we need to single out the motivations for the Federal government.

We have also added a figure depicting the estimated species range for bearded, ribbon, and spotted seals along with general representation of recent aerial survey effort. This should better communicate to the reader how the spatial distribution of our haul-out behavior overlaps with the species range and survey effort.
:::

> ***Line 227:** Seals can just as happily be on foraging trips within the pack and marginal sea ice as in the open water, depending on the species. Suggesting that foraging trips are associated with ice-free conditions is misleading.*

::: author-response
**\[author response\]** We have removed 'e.g., foraging trips' from this sentence so as not to mislead the reader.
:::

> ***Line 229-230:** This is your primary reason for not including sea ice geography in your analysis: you simply lacked geolocation of much of your haulout behavior data.*
>
> *Please help the reader understand how coarse your tracking data was for the flipper tagged animals and what proportion of each species was tracked with these flipper tags. Please also help the reader understand how the tracking data relates geographically to regions that may be subject to aerial survey efforts (i.e. are your tracking data collected from across the regions that may be surveyed during range-wide spring aerial surveys?).*

::: author-response
**\[author response\]** We have re-organized this paragraph to, hopefully, better communicate our choice not to explore sea-ice related covariates. As the reviewer suggests, we have limited geolocation data for at-sea periods of our haul-out behavior data -- we know the seal was in the water but don't have any estimates of location when only a flipper tag is attached. By leading with this statement, we hope that communicates our primary rationale. But, the other rationales are also important and we still include them. We have also moved this section to the 'Discussion' in response to other reviewer comments.

We have provided some statistics regarding the percentage of the percent-dry timeline data that is not associated with a corresponding location estimate because the seal was in the water and no estimates were available from flipper mounted SPOT tags.

We have also added a figure depicting the estimated species range for bearded, ribbon, and spotted seals along with general representation of recent aerial survey effort. This should better communicate to the reader how the spatial distribution of our haul-out behavior overlaps with the species range and survey effort.
:::

> ***Line 229-230:** Please help the reader understand why you have not included a basic metric of sea ice geography, such as distance from marginal ice edge (from the NIC's daily Marginal Ice Zone charts), which is a metric that would be available for all data collected during an aerial survey.*

::: author-response
**\[author response\]** We have revised the text in this section to try to provide a clearer understanding of our reasoning. Also, note this has been moved to the Discussion section.

The text now reads:

Notably missing from our list of explanatory variables is any spatial-temporal representation of sea ice concentration, area, or extent. This may seem counterintuitive when modeling the haul-out behavior of seal species with such a close association to sea ice; seals haul out in the presence of sea ice, and we could assess the local concentration of sea ice during these events (see Olnes 2020). This, however, expands the scope of our analysis into the realm of habitat selection and many of our deployments consisted of a single device attached to the rear flipper of the seal which did not provide at-sea locations, limiting our ability to fully evaluate fine-scale habitat preferences related to sea-ice. Insight into how seals use and interact with sea-ice during an extended period when the availability and characteristics of sea-ice is rapidly changing is important but ancillary to the focus of this analysis. The focus of this study was to develop models applicable for aerial survey correction factors and using sea ice as a covariate would almost certainly bias haul-out predictions towards those seals that are on or near ice and therefore more likely to haul out. Since aerial surveys can only detect seals on ice, abundance estimates would be missing a correction for those seals that are away from ice.

Consider the case of employing haul-out probabilities as an aerial survey correction in a population inhabiting two areas: an ice-free region and a region with sea ice. Further, assume that half of the population is in each area on average, and that the probability of hauling out is 1.0 in the ice covered region and 0.0 in the region without ice. Denoting the total population as $N$, let us examine what happens when we use (1) population level availability, and (2) ice-specific availability as a correction factor. We'll further assume, for sake of this example, that detection probability of seals on ice is 1.0, and that the entirety of the ice covered region is censused. In case (1), our aerial survey count is $C = 0.5 \cdot N$, with a population-level availability probability of $\hat{a} = 0.5$. A Horvitz-Thompson-type estimator for abundance is simply $\hat{N} = C \div \hat{a}$, which has expectation $\mathbb{E}(\hat{a}) = \frac{0.5 \cdot N}{0.5} = N$, as desired. Now consider case (2). In this case, the probability of hauling out is 1.0 for seals in the ice-covered area, so we have $\mathbb{E}(\hat{N}) = \frac{0.5 \cdot N}{1.0} = 0.5N$. That is, we bias abundance because we are not accounting for seals that are away from ice, and therefore have a zero chance of hauling out. The same logic holds for any other covariate, such as distance from the ice edge, that has different values in the surveyed and unsurveyed habitat of the seals (assuming distance from ice edge is coded with opposite signs inside and outside the pack ice).

Lastly, our study was limited to the spring season when seal haul-out behaviors are strongly influenced by pupping, nursing, breeding behavior, and molt and these drivers are likely more influential than specific sea-ice concentration. Crawford et al. (2019) compared haul-out probability models for ringed seals and found those that only included season (and not sea-ice concentration) were the most parsimonious. For these reasons, we have elected not to use sea ice concentration *as a predictor for haul-out probability* in the present study.
:::

> ***Table 2:** It is a bit misleading to indicate that the NARR variables were sourced from NARR alone. Their values were interpolated onto a geogrpahic location that was widely variable in precision and that precision relates to the both the tag attachment (glued on or flipper attached) and the seal behavior (in water, hauled out).*
>
> *Indicating that your "Northing" variable was from the "Bio-logger" is also misleading because your locations was determined by either the Doppler inflection point time, satellite orbitography and a model of the Earth (the Argos relocation method) or by reprocessing of GNSS signals that were logged and then transmitted via the Argos system as a data message. Really, neither of these were determined by the "Bio-logger", rather they were determined by a very complex system of geolocation.*

> *Your readers may be better served by not including this "Source" field, given that you describe how you obtained or estimated these variable values in your methods text. Offering an over-simplified "Source" in the table obscures your actual methods.*

::: author-response
**\[author response\]** We appreciate the reviewer's thoughts on this and agree with that removing the 'Source' column better serves the readers.
:::

> ***Line 244:** Neither your telemetry data nor aerial surveys can distinguish "basking" as a behavioral state.  You are measuring "hauled out" behavior and aerial surveys observe "hauled out" seals.  I encourage you to use these terms and not mix in ideas of "basking", "nursing", "breeding", "taking refuge from marine predators" etc.*

::: author-response
**\[author response\]** As previously mentioned, we've addressed this and consistently use "hauled out" and not "basking"
:::

> ***Line 254:** It is nice that you have sufficient data to do this. However, the reason for having separate models for each species is because you expect different behaviors from each specifies and because you expect that aerial efforts will be able to distinguish sea species and be able to apply an availability correction factor for each species. Please correct me if I am wrong, else, please keep your reader focused on the value that you are bringing to the analysis here in the Methods section and the good fortune that you have in the sample size, which should be noted in the Results and Discussion section.*

::: author-response
**\[author response\]** We have rephrased this sentence to emphasize that we fit each species separately because a) we expected differing behaviors across species and b) fitting a single model was not possible with such a large data set (even with the advantages provided by the glmmLDTS package).
:::

> ***Line 263:** Please clarify what is intended by the notation "only -99% or our observations".*

::: author-response
**\[author response\]** This is an error in the code that has been fixed.
:::

> ***Line 263:** Also, please consider using consistent terms for "behavior records".  Perhaps it could be fair to term your telemetered haulout status as an "observation", however, it seems more intuitive to reference this a "record"*

::: author-response
**\[author response\]** We have made edits throughout the manuscript to use consistent terms for 'behavior records'
:::

> ***Line 263:** Please help your reader understand why you have not adopted use of a classified variable for each telemetered behavior interval whereby it could be either "wet" or "dry" based on a threshold proportion of readings reported as "dry".  This has two advantages: (1) were this enabled through processing on-board the tags, it would greatly increase your data compression and collection; (2) it may enable greater analytical power in your modeling effort.*

::: author-response
**\[author response\]** We provide fairly clear and detailed information on the haul-out behavior records, how they are recorded and transmitted in the methods section. We have further edited this to make sure it is clear that these are standard settings within the Wildlife Computers tag platform -- we do not have direct control over the details for on-board processing or transmission. As stated in the text, for this analysis, the linear predictor was modeled on the logit scale as a Bernoulli response where tags are either 'wet' or 'dry' for a given hour based on whether the tag was greater than 50% dry. This is a classified variable as suggested. The difference is that we conducted this classification post-transmission instead of on-board the tag. We feel the transmission of various percent-dry values for a given hour provide more flexibility in the types of analyses and insights that can be gained from deployments.
:::

> ***Line 265-272:** Please seek input from a competent biometrician. I am not able to offer any guidance on this other than to muse why you are pushing to estimate six parameters for a time of day covariate when your seals are not known to keep track of time.*
>
> *It would be reasonable to assume that seals respond to insolation or the angular height of the sun in its daily arc.  Please help your reader understand why you have sought to model a complex representation of hourly time when you could have pulled insolation from NARR or calculated sun angular altitude (perhaps as a linear proportion of the day's maximum sun angular altitude) from standard astronomical functions.*
>
> *Perhaps your motivation is to provide plots that appear intuitive with hours on the y-axis, day of year on the x-axis and haulout proportion as a "heat map".  However beautiful these plot are, I wonder if your goal for improving precision and reducing bias when estimating availability during an aerial survey would not be better served by  the alternative and simpler variables.*

::: author-response
**\[author response\]** We appreciate the suggestion and find the possibility of using insolation values from reanalysis products as a covariate intriguing. As the reviewer suggests, such an approach could provide a less complicated and more parsimonious model than the approach we have taken with the Fourier series as a cyclic representation of hour-of day. We have taken the time to do some additional analysis and methodological exploration and provide additional details in Supplement 2. While promising, we feel the approach taken in the analysis is the most appropriate because it distinguishes daily bi-modal (e.g. crepuscular) patterns that might be important but would not be detected by insolation alone (i.e. insolation at dawn or dusk would be similar values but occur at different times of day that might be ecologically important). Additionally, as the reviewer points out, results from the analysis and haul-out predictions are more intuitive and human interpretable when using hour-of-day.
:::

> ***Line 278:** Please help your reader understand what is meant by your term "horizontal shift".  It is clear that you have a graphical representation in your mind that has not been shared with your reader. Please rewrite to avoid "jargon".*

> ***Line 285:** Please help your reader understand what could be "important" about these terms. Please write for clarity, avoiding jargon.*

> ***Figure 4:** Please use standard conventions for plot titles.  Your title deserves a statement in your Results or Discussion section.  It does not follow the expectations for a publication figure title.*

::: author-response
**\[author response\]** The plot titles were developed with strong consideration for improving scientific communication over approaches that are more typically used in publications. They were developed with guidance from modern approaches to data visualization that recognize the plot title is often the very first thing the reader sees. Therefore it can serve as an effective communication tool to help the reader interpret the results.

That said, we recognize that the approach may depart too much from the standard expectations and have made adjustments to each of the titles with the feedback in mind. We hope the new titles provide a balanced approach.
:::

> ***Figure 4:** Please scale your color ramp to enable better visualization of the subject: limit the color ramp to the modeled values (which appear to be from zero to \~0.5).*

::: author-response
**\[author response\]** The scale used is consistent across all three species to maintain comparability. If we were to adjust the scale as suggested it would make comparisons across species difficult
:::

> ***Figure 4:** Please switch to a color ramp that is suitable for readers with color perception disabilities such as the viridis color ramps.*

::: author-response
**\[author response\]** The color ramp used is the 'Hiroshige' color ramp from the MetBrewer R package and has been tested and confirmed to be colorblind friendly. We initially used the viridis color palette for these plots but the Hiroshige colors provided additional visual detail we feel is important.
:::

> ***Figure 4:** This is unclear. Your plot indicates a y-axis of local solar hour which varies from 0 to 24.  This is clearly not fixed at local noon.*

::: author-response
**\[author response\]** This was an error and has been fixed. We thank the reviewer for recognizing this and highlighting the mistake.
:::

> ***Line 323:** Please use standard English.  "Ontongeny" is generally understood to describe the physical development of an individual organism.  Here you are referencing the onset of of behaviors within the life of an organism.  It may benefit your readers to use a term that is more direct, such as "development".*

::: author-response
**\[author response\]** We have replaced 'ontongeny' with 'development' as suggested
:::

> ***Line 326-327:** Given that Ribbon seals occupy a high latitude region during this time and given that there is scarcely any twilight during this season, please consider using a term other than "crepuscular".*

::: author-response
**\[author response\]** Subsequent model runs (with improved QA/QC on telemetry data) have reduced the indication of any crepuscular or bi-modal haul-out behavior of ribbon seals and we no longer feel it is worth mentioning in the paper so it has been removed.
:::

> ***Line 331-332:** This is a fair summary of your modeling efforts.  However, it raises the questions as to what these seals are responding to.  Do they sense air pressure and make a decision on their haulout behavior or are low pressure conditions associated with low insolation and they are responding to that?  Given that you can readily extract insolation from the NARR, please consider use of insolation as a covariate that may more directly relate to seal behavior.*

::: author-response
**\[author response\]** See previous responses regarding the potential use of 'insolation' from NARR or other reanalysis products.
:::

> ***Figure 5:** Please use standard figure titling and not use a statement that should below in a PowerPoint slide or in your Results and Discussion section.*
>
> ***Figure 6:** Please use standard conventions for plot titles.  Your title deserves a statement in your Results or Discussion section.  It does not follow the expectations for a publication figure title.*

::: author-response
**\[author response\]** See previous response regarding our choice in plot titles and subsequent edits
:::

> ***Figure 8:** Please label your plots using standard capitalization and words, instead of all capitalized abbreviations.*

::: author-response
**\[author response\]** We have made edits based on this suggestion
:::

> ***Line 360-362:** This belongs in the discussion, not results.*

::: author-response
**\[author response\]** The suggestion to move the mention that there "were still some years where the pattern shown was informed by a small number of individuals that may not represent population-level patterns" to the discussion is noted and changes have been made.
:::

> ***Line 371-372:** Please indicate the season (or months of the year) in this statement.*

::: author-response
**\[author response\]** Noted and edited to include mention of the spring season
:::

> ***Line 375-376:** Do you mean to state that "haulout behavior modeled from telemetered haulout behavior records may represent population-level behavior patterns."*

::: author-response
**\[author response\]** Noted and edits have been made
:::

> ***Line 377-378:** Feltz and Fay (1966) offer (an amazingly cool) functional explanation of the thermal constraints on hair follicle growth.  Whether seals during specific study times are optimizing hair follicle growth is unknown.*

::: author-response
**\[author response\]** We have edited this statement for additional clarity
:::

> ***Line 418-419:** Yes, and you have had access to direct estimates of insolation from the NARR.  Can you incorporate that and find a tighter model of haulout behavior.  This would be an improvement for design and interpretation of aerial survey efforts.*

::: author-response
**\[author response\]** See previous comments regarding the potential use of insolation values from reanalysis products
:::

> ***Line 420-421:** You may state this if you first offered an estimate of haulout behavior and confidence intervals on that behavior state, then compared that with your estimated haulout behavior modeled with the use of co-variates.  Once you perform this, you may state that inclusion of weather co-variates improved haulout state estimates.  However, you have not done this.  If you wish to state this, please do this.*

::: author-response
**\[author response\]** Here we simply state that 'our results highlight the importance of incorporating weather covariates' and that importance is indicated with the reported 'type III' F statistics for the weather covariate terms in the results section. We are clear in our methods that the use of pseudo-likelihoods in the analysis prevent us from conducting typical model selection with AIC or similar as suggested. We have moved a separate discussion of comparison with previous analysis results here to provide additional comparative context.
:::

> ***Line 425:** This term suggests some evaluation of a threshold above which haulout behavior shifts. Please write for clarity and for what you have presented in your results.*

::: author-response
**\[author response\]** We have edited this sentence for improved clarity
:::

> ***Line 455:** Please be clear as to why you anticipate greater unmodeled heterogeneity of availability for Arctic seals?  Here you have considered Arctic seals, have you not?  Or are you using a different definition of the Arctic other than the ice bearing waters occupied by ice-associated seals?*

::: author-response
**\[author response\]** We have edited this statement for additional clarity. The interpreted emphasis on 'Arctic seals' was not intended. Instead, the key point is that there is likely unmodeled heterogeneity related to the seasonal distribution and annual timing of peak haul-out behavior. This is an acknowledgement that our model is not a complete picture or perfect representation of seal haul-out behavior. In an environment like the Arctic that is undergoing dramatic change and inter-annual variation in climate and sea-ice distribution, this is important to keep in mind.
:::

> ***Line 459-464:** Here you provide the discussion required to make the statement that inclusion of weather and other covariates results in a different estimate of haulout proportion.  Can you extend this to indicate an improvement in the confidence interval on this estimated proportion?*

::: author-response
**\[author response\]** Noted and this section has been moved to that discussion
:::

> ***Line 465-466:** Please be specific. Here you linked the telemetered haulout behavior to available covariate data then modeled haulout behavior.*

::: author-response
**\[author response\]** Noted and additional specifics have been added to this statement
:::

> ***Line 467-473:** This can be stated in the introduction so that you can clearly indicate the framing of your study. "We modeled availability from telemetered haulout behavior records of seals that haulout on ice where they may be viewed by aerial surveys (list the species considered).*

::: author-response
**\[author response\]** This section has been moved to the introduction.
:::

## Reviewer 2

> *One suggestion I would share with the authors is the addition of a limitations section in the discussion. The limitations were outlined throughout the methods, but a succinct accounting of these limitations in the discussion would allow readers who do not want to carefully dissect all of the modeling in the methods to be able to better interpret the robustness of the results and conclusions.*

::: author-response
**\[author response\]** We thank the reviewer for this question and have added a section about limitations to the Discussion section
:::

> *I still had one remaining question on why the authors couldn't run the models with and without the sea ice concentration as a predictor. The comparative results might be interesting (I understand why the correction factor might be biased\...but given that this is a baseline study, it would have been an interesting comparison to see whether it was or wasn't a decent predictor for the different seal species).*

::: author-response
**\[author response\]** We think we've addressed this in edits and comments to reviewer #1. We think the inclusion of sea ice as a predictor is better suited for a habitat use/habitat preference study and we hope to pursue that in future analyses.
:::

> *The authors need to explain more how they are attributing statistical significance. For example, in lines 344-345, they ascribe strong influence of temperature with a p-value of 0.064.\
> \
> Not to be one who is fixated on a p-value of 0.05 denoting significance, I was a bit surprised to see the same authors, in lines 364-366, to state that there is 'no indication that the observed trend is meaningful'for spotted seal adult females having an R2 of 0.767, p=0.022. If the extent of sea ice explains nearly 80% of the variance in peak haul-out probability, how is it that the author's call this a 'trend'? This requires more explanation. As does the statement in line 427-428.*

::: author-response
**\[author response\]** We appreciate the reviewer highlighting this confusion and apologize for it. These were typos or mis-coded values that snuck into the section. We have reviewed and revised all of them to ensure they are correct and that we are consistent in our attribution of significance.
:::

> *I found 463-464 confusing. This is an important part of the discussion (using your framework to assess previous results from another study), but it could be more clearly written. The sentence, 'Applying models that ignore age, sex, and year effects\...under the current analysis framework' is very confusing and doesn't seem to follow the argument of the paper easily. Could this be reworded so that the full intent of the paragraph is communicated?*

::: author-response
**\[author response\]** Reviewer #1 also commented on this section and offered suggestions for editing. With both comments in mind, we have made edits to the text and re-arranged the location so that this is more closely linked with the discussion regarding the importance of weather covariates in the model.
:::

> *I appreciate the acknowledgement that it is difficult to know whether differences in abundance estimates are attributable to changes in abundance or changes in haul-out behavior and the potential proposed solution (lines 450-458) Has anyone done a smaller-scale study that would address this inherent problem?*

## Reviewer 3 (M. Carter)

> *I have only a few minor comments, mostly where the text would benefit from extra clarity:\
> *
>
> ***L43:** Word missing? ...Bearing \[and\] Chukchi Seas...*\

::: author-response
**\[author response\]** noted and fixed
:::

> **L63:** It may be useful to highlight somewhere that matching seal haulout behaviour to environmental conditions with any degree of certainty requires location data as well as wet-dry data, and that modern bio-logging devices can collect both data streams simultaneously.\

::: author-response
**\[author response\]** thank you for this suggestion and we have edited this paragraph to include mention regarding the ability of modern bio-logging devices to collect both behavioral and location data streams which allows inclusion of environmental covariates.
:::

> **L82-83:** All citations given here relate to ice-associated seals, but these techniques are also relevant to seals that haul-out on land. Lonergan et al. (2013) Aquatic Conservation: Marine and Freshwater Ecosystems 23: 135-144 is a key example from non-polar regions.\

::: author-response
**\[author response\]** we have expanded the list of references to include Lonergan et al as well as a few other studies from non ice-associated seals
:::

> **L110:** Clarify why meteorological factors affecting walruses may be relevant to the study species.\

::: author-response
**\[author response\]** thank you for the suggestion. We've edited the sentence to, now, read:

"Previously estimated availability correction factors accounted for variables such as time-of-day and day-of-year but did not investigate the impact of meteorological variables. Such variables have been shown to influence walrus haul-out behavior and we expect weather conditions to also influence seal haul-out behavior and including them within the model framework will benefit our estimates of seal availability during aerial surveys."
:::

> **L120-124:** A map of the study area with deployment areas indicated would be useful for those readers not familiar with the local geography.\

::: author-response
**\[author response\]** Reviewer 1 had a similar suggestion and we now include such a figure.
:::

> **L126-128:** More information is required on the location data type (e.g. GPS or ARGOS?) and data retrieval (e.g. archival loggers or ARGOS-relay?).\

::: author-response
**\[author response\]** we have added the following sentence:

"All data were transmitted by the deployed instruments via the Argos satellite network and location data were either derived from Argos transmissions or transmitted FastLoc GPS data."
:::

> **L133-135:** Please clarify, were there any cases where only head/back-mounted tags were deployed? If so, how did you deal with the potential for false haul-out records due to prolonged surface activity?\

::: author-response
**\[author response\]** we've clarified the language in this paragraph so it's clear that deployments consisted of head/back tag only, flipper tag only, and combined. Examination of the haul-out records hasn't indicated there's a propensity for these tags to record extended dry periods when surface resting at sea for the head/back tags.
:::

> **L135:** At this stage you have not yet explained what type of data each device collects, so we do not know what the hourly percent-dry observations from the flipper tag were preferred to.\

::: author-response
**\[author response\]** thank you for noting this. we have moved the later paragraph (starting line 165) up so it is before this section.
:::

> **L139:** Which species?\

::: author-response
**\[author response\]** we have removed the 'for some species' from this sentence
:::

> **L136-144:** This could be clearer. I assume the "some species" part refers only to pelage, but it appears at first read to refer to the whole list of metrics. I would suggest re-wording this section to improve clarity regarding which factors were used for which species and with what degree of certainty age class could be determined.\

::: author-response
**\[author response\]** as noted in the previous response, we've removed the 'for some species' and think this improves the clarity of this section. Unfortunately, we don't have any way to quantify a degree of certainty for age class determination across the various field teams responsible for field identification.
:::

> **L146:** How many YOY were advanced to subadult?\

::: author-response
**\[author response\]** we've added this metric
:::

> **L188:** On land as well as ice?\

::: author-response
**\[author response\]** we have edited this sentence to state 'also haul out on land'
:::

> **L194:** It is not clear to me if all tags for a given species were deployed in the same region, or across multiple regions (a map of the study area with deployment regions per species would help here). Given regional differences in phenology mentioned in the Introduction (L52-53), did you consider whether region should be included as a covariate?\

::: author-response
**\[author response\]** we have added a figure/map as suggested
:::

> **Figs 4, 6, 8:** How is that hour of the day is fixed at local solar noon if predictions are shown for all hours? Also, is it possible to show uncertainty (lower and upper CIs) for these figures, perhaps in Supplementary Materials?\

::: author-response
**\[author response\]** Reviewer 1 also noted the incorrect statement that hour of day was fixed at local solar noon and this has been corrected. We have taken the reviewer's suggestion and added supplemental materials with plots (this time fixed at solar noon) that indicated the lower/upper CIs.
:::

> **L316:** Use of "additionally" negates the need for "also".\

::: author-response
**\[author response\]** noted and removed 'also'
:::

> **L341:** Word missing? ...and \[spent\] more time out of the water...\

::: author-response
**\[author response\]** noted and we've made the suggested correction
:::

> **L387-390:** It is not clear what species was studied by Thometz et al, and thus maybe a citation is needed for the information relating to ribbon and spotted seals.\

::: author-response
**\[author response\]** we've added a clarification that the Thometz et al study and data presented was for bearded seals.
:::

> **L393:** Use "both species" instead of "all species" to avoid confusion.\

::: author-response
**\[author response\]** we've edited to specifically state 'ribbon and spotted seals' to avoid any confusion
:::

> \
> The paper could benefit from a brief concluding paragraph summarising the importance of the work in the context of climate change and population trends, as outlined in the Introduction.

::: author-response
**\[author response\]** thank you for this suggestion. we've added a concluding paragraph that addresses this comment.
:::

### 2023-09-15 analysis/paper/Letter_Response_to_Reviewers.qmd
minor edits

@@
 ::: author-response
-**\[author response\]** We were using a less strict definition of extra-terrestrial to refer to interference that wasn't Earth/surface-based. We've edited to 'atmospheric interference'
+**\[author response\]** We've edited to 'atmospheric interference'
 :::
@@
 ::: author-response
-**\[author response\]** We feel this note is important to acknowledge in the methods because it speaks to the rationale for our methodological choices (bio-logger type, programming choices, analytical choices).
+**\[author response\]** We feel this note is important to acknowledge in the methods because it speaks to the rationale for our methodological choices (bio-logger type, programming choices, analytic choices).
 :::
@@
 **\[author response\]** The previous paragraph describes how we determined daily locations.
-We see the reviewers point re: the data not technicaly coming from our data loggers.
+We see the reviewers point re: the data not technically coming from our data loggers.
 We have edited the sentence for additional clarity.

### 2023-09-15 analysis/paper/London_HauloutBeahvior.Rmd
new figure/updates for deployment locs
addresses #50

@@
 
-theme_set(theme_minimal(base_family = 'Source Sans Pro'))
+theme_set(theme_minimal(base_family = 'Source Sans Pro',
+                        base_size = 9))
 theme_replace(plot.title.position = "plot")
@@
 (ref:deployMapCaption) Initial deployment areas for each seal species (bearded, ribbon, and spotted seals) in the study between 2005 and 2020.
-Colored regions shown are minimum concave polygons buffered by 60 km for enhanced visibility.
-See Supplemental Material S1 for specific details.
+Regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility.
+The larger region (shaded yellow) indicates the spatial extent of aerial survey effort to date in US waters of the Bering and Chukchi seas.
 Deployments were initiated across a range of months but only data from 1 March to 15 July were included in the analysis.
-See Figure \@ref(fig:dataMap) for the spatial distribution of observed data used in the study.
+See Figure \@ref(fig:dataMap) for the spatial distribution of observed data used in the study and Supplemental Materials 1 for additional deployment details.
 
@@
 targets::tar_load(deploy_details, store = here::here("_targets"))
+targets::tar_load(survey_area, store = here::here("_targets"))
 
@@
 world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
+  sf::st_transform(st_crs(deploy_sf))
+
+survey_area <- survey_area %>% 
   sf::st_transform(st_crs(deploy_sf))
@@
 
+label_coords_bering <- st_sfc(st_point(c(178.5,57)), crs = 4326) %>%
+  st_transform(st_crs(deploy_sf))
+label_coords_chukchi <- st_sfc(st_point(c(-176,73)), crs = 4326) %>%
+  st_transform(st_crs(deploy_sf))
+label_coords_beaufort <- st_sfc(st_point(c(-146.5,72.25)), crs = 4326) %>%
+  st_transform(st_crs(deploy_sf))
+
 p <- ggplot() +
+  layer_spatial(survey_area, color = "#ddc000", 
+                fill = "#ddc000", 
+                alpha = 0.15,
+                linewidth = 0.5) +
   layer_spatial(deploy_sf, aes(fill=species),
@@
                      fill = "grey80") +
+  geom_sf_text(data = label_coords_bering, aes(label = "Bering sea"), size = 1.75, color = "grey40") +
+  geom_sf_text(data = label_coords_chukchi, aes(label = "Chuckchi \nsea"), size = 1.75, color = "grey40") +
+  geom_sf_text(data = label_coords_beaufort, aes(label = "Beaufort \nsea"), size = 1.75, color = "grey40") +
   facet_grid(cols = vars(species), labeller = species_labels,
@@
                     name = "species") + 
-  ggtitle("Areas of initial deployment location",
+  ggtitle("Initial bio-logger deployment locations by species",
           subtitle = 
                  paste0("Bio-loggers were initially deployed on ",
-                "<span style = 'color:",species_colors['Bearded seal'],";'>**bearded**, </span> ", 
-                "<span style = 'color:",species_colors['Ribbon seal'],";'>**ribbon**, </span> ",  
+                "<span style = 'color:",species_colors['Bearded seal'],";'>bearded,</span> ", 
+                "<span style = 'color:",species_colors['Ribbon seal'],";'>ribbon,</span> ",  
                 "and ", 
-                "<span style = 'color:",species_colors['Spotted seal'],";'>**spotted**</span> seals ", 
+                "<span style = 'color:",species_colors['Spotted seal'],";'>spotted</span> seals ", 
                 "across multiple regions within the Bering, Chukchi, and western Beaufort seas.")) +
+  labs(caption = "shaded yellow region indicates spatial extent of aerial survey effort in US waters") +
   theme(legend.position = "none",
+        axis.title = element_blank(),
         strip.background = element_blank(),
         strip.placement = "outside",
-        strip.text.x = element_markdown(face="bold"),
+        strip.text.x = element_markdown(face = "bold"),
         plot.subtitle = element_textbox_simple(margin = ggplot2::margin(b=12)))

### 2023-09-15 R/get_survey_area.R

@@ -1,0 +1,32 @@
+get_survey_area <- function() {
+  
+  tryCatch({
+    con <- dbConnect(RPostgres::Postgres(),
+                     dbname = 'pep', 
+                     host = Sys.getenv('PEP_PG_IP'),
+                     user = keyringr::get_kc_account("pgpep_londonj"),
+                     password = keyringr::decrypt_kc_pw("pgpep_londonj"))
+  },
+  error = function(cond) {
+    print("Unable to connect to Database.")
+  })
+  on.exit(dbDisconnect(con))
+  
+  boss_lines_qry <- "SELECT * FROM surv_boss.geo_tracks_by_effort;"
+  chess_lines_qry <- "SELECT * FROM surv_chess.geo_track_by_effort;"
+  
+  boss_area_sf <- read_sf(con, query = boss_lines_qry) |> 
+    dplyr::filter(effort == "On", st_length(geom) > units::set_units(0,m)) |> 
+    concaveman::concaveman(concavity = 4) |> 
+    st_simplify(dTolerance = 25000, preserveTopology = TRUE)
+  
+  chess_area_sf <- read_sf(con, query = chess_lines_qry) |> 
+    dplyr::filter(effort_type == "ON", st_length(geom) > units::set_units(0,m)) |> 
+    concaveman::concaveman(concavity = 6) |> 
+    st_simplify(dTolerance = 10000, preserveTopology = TRUE)
+
+  survey_area <- st_union(boss_area_sf, chess_area_sf)
+  
+  return(survey_area)
+}
+  

### 2023-09-15 _targets.R

@@ -56,7 +56,9 @@ list(
   tar_target(nsb_deployments, get_nsb_deployments()),
   tar_target(nsb_locations, get_nsb_locs(nsb_deployments)),
   tar_target(nsb_timelines, get_nsb_timelines(nsb_deployments)),
-
+  
+  tar_target(survey_area, get_survey_area()),
+  
   tar_target(locs_sf, get_locs_sf(adfg_locations, nsb_locations)),
   tar_target(timeline_data, get_timeline_data(adfg_timelines, nsb_timelines)),
 

### 2023-09-16 analysis/paper/London_HauloutBeahvior.Rmd
improve title/subtitle/caption
addresses #50

Also removes source col from covariate table

@@
 ```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployed on a ribbon seal over two years during the months of March, April, and May.
-Areas of the plot indicated with an 'X' represent missing data that were not successfully recieved from the tag."}
+This actogram plot represents the transmitted hourly percent-dry values.
+Areas indicated with an 'X' represent missing data that were not successfully recieved from the tag."}
 
@@
   theme(panel.spacing = unit(0,"line")) +
-  ggtitle("Observation records are incomplete because not all data are transmitted",
-          subtitle = "percent-dry actogram for a single ribbon seal across two years") +
+  ggtitle("Example percent-dry actogram from bio-logger data",
+          subtitle = str_wrap("Observation records shown are from a single ribbon seal across a two year deployment. Note that not all records are received from the device.", width = 85)) +
   xlab("day of month") + ylab("hour (UTC)") 
@@
 covariate_tbl <- tibble::tribble(
-  ~Covariate, ~Type, ~Source, ~Description,
-  "Age-sex class",   "Categorical", "Field Assessment", 
-  "young-of-the-year, subadult, adult male and adult female",
-  "Hour", "Continuous; Fourier basis", "Bio-logger", 
-  "local solar hour using 6 variables of a Fourier-series basis",
-  "Day",   "Continuous", "Bio-logger", 
-  "linear, quadratic, and cubic effects of day-of-year",
-  "Precip", "Continuous", "NARR", 
-  "convective precipitation (ref:kgm2)",
-  "Pressure", "Continuous", "NARR", 
-  "atmospheric pressure at sea level (kPa)",
-  "Temp", "Continuous", "NARR",
-  "air temperature (C) at 2m above the earth’s surface",
-  "Wind", "Continuous", "NARR", 
-  "northerly and easterly vector components for wind converted into a single wind speed via the Euclidean norm",
-  "Northing", "Continuous", "Bio-logger","latitude divided by the mean latitude across all locations (for bearded seals only)",
-  "Year", "Continuous", "Bio-logger", "For the set of models examining inter-annual variation in sea ice use, we fitted models with the addition of year by day-of-year interactions." 
+  ~Covariate, ~Type, ~Description,
+  "Age-sex class","Categorical","young-of-the-year, subadult, adult male and adult female",
+  "Hour", "Continuous; Fourier basis", "local solar hour using 6 variables of a Fourier-series basis",
+  "Day",   "Continuous", "linear, quadratic, and cubic effects of day-of-year",
+  "Precip", "Continuous", "convective precipitation (ref:kgm2) (NARR)",
+  "Pressure", "Continuous", "atmospheric pressure at sea level (kPa) (NARR)",
+  "Temp", "Continuous", "air temperature (C) at 2m above the earth’s surface (NARR)",
+  "Wind", "Continuous",
+  "northerly and easterly vector components for wind (NARR) converted into a single wind speed via the Euclidean norm",
+  "Northing", "Continuous", "latitude divided by the mean latitude across all locations (for bearded seals only)",
+  "Year", "Continuous", "For the set of models examining inter-annual variation in sea ice use, we fitted models with the addition of year by day-of-year interactions." 
 )

### 2023-09-16 analysis/paper/London_HauloutBeahvior.Rmd
improve spatial map & data distribution plots
updated the titles, subtitle, and caption to address #50

@@
 
-(ref:mapCaption) Spatial distribution of haul-out data during the months of March through July 15 for each of the three species.
+(ref:mapCaption) Spatial distribution of haul-out records during the months of March through July 15 for each of the three species.
 Data were collated across all years between 2005 and 2020.
@@
   theme(legend.position = "bottom") +
-  ggtitle("Spatial distribution of haul-out behavior data") +
+  ggtitle("Spatial distribution of haul-out behavior records") +
   labs(caption = bquote("One hexagonal cell represents 50"~km^2))
@@
   labs(fill = "seal-hours of bio-logger data") +
-  ggtitle("Density of haul-out behavior observations varies from March through July",
-          subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")
+  ggtitle("Density of haul-out behavior observations by species",
+          subtitle = str_wrap("data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020) from March through July"))
 

### 2023-09-16 analysis/paper/London_HauloutBeahvior.Rmd
updates to plot titles/subtitles
haul-out heatmap figures updated to address #50

@@
 
-```{r beardedHOCal, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes.
-Hour of the day was fixed at local solar noon."}
+```{r beardedHOCal, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes."}
 
@@
   xlab("day of month") + ylab("local solar hour") +
-  ggtitle("Bearded seals rest on sea ice around solar noon and solar midnight")
+  ggtitle("Bearded seal predicted haul-out probability",
+          subtitle = str_wrap("Periods of highest haul-out probability for bearded seals appear to be centered around solar noon and midnight"))
 p
@@
   xlab("day of month") + ylab("local solar hour") +
-  ggtitle("Ribbon seals rest on sea ice in late spring and around solar noon.")
+  ggtitle("Ribbon seal predicted haul-out probability.",
+          subtitle = str_wrap("Periods of highest haul-out probability for ribbon seals appear centered around solar noon and age-sex classes show differing seasonal patterns."))
 p
@@
   xlab("day of month") + ylab("local solar hour") +
-  ggtitle("Spotted seals rest on sea ice in late spring and around solar noon.")
+  ggtitle("Spotted seal predicted haul-out probability.",
+          subtitle = str_wrap("Periods of highest haul-out probability for spotted seals appear centered around solar noon and age-sex classes show differing seasonal patterns."))
 p

### 2023-09-25 analysis/paper/London_HauloutBeahvior.Rmd
major updates to most plots, titles, captions
should fully address #50 and also #40. Issue #28 was also addressed by fixing the wx marginal plots to May 15.

--- 2023-09-16 analysis/paper/London_HauloutBeahvior.Rmd
@@
 
-theme_set(theme_minimal(base_family = 'Source Sans Pro',
+theme_set(theme_minimal(base_family = 'Spline Sans',
                         base_size = 9))
-theme_replace(plot.title.position = "plot")
+theme_update(
+  plot.title.position = "plot",
+  # left-align title
+  plot.caption.position = "plot",
+  # right-align caption
+  plot.title = element_text(face = "bold"),
+  # larger, bold title
+  plot.subtitle = element_textbox_simple(
+    color = "grey30",
+    margin = ggplot2::margin(t = 0, b = 12, r = 60)
+  ),
+  plot.caption = element_text(color = "grey30"),
+  # change color of caption
+  panel.grid.minor = element_blank() # no minor grid lines
+)
+
+# use a named vector for explicit matching
+species_colors <- c(
+  'Spotted seal' = "#52BCA3",
+  'Bearded seal' = "#E58606",
+  'Ribbon seal' = "#5D69B1"
+)
+
+age_sex_labels <- c(
+  "ADULT.F" = "adult female",
+  "ADULT.M" = "adult male",
+  "SUBADULT" = "subadult",
+  "YOUNG OF YEAR" = "young of year",
+  "ALL AGES" = "all ages"
+)
+
 ```
@@
 (ref:deployMapCaption) Initial deployment areas for each seal species (bearded, ribbon, and spotted seals) in the study between 2005 and 2020.
-Regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility.
-The larger region (shaded yellow) indicates the spatial extent of aerial survey effort to date in US waters of the Bering and Chukchi seas.
+Solid regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility.
+The larger, shaded region indicates the spatial extent of aerial survey effort to date in US waters of the Bering and Chukchi seas.
 Deployments were initiated across a range of months but only data from 1 March to 15 July were included in the analysis.
@@
 
-species_colors <- setNames(met.brewer("Isfahan2",n=3,direction=-1,override.order = TRUE),
-                           c("Bearded seal","Ribbon seal","Spotted seal"))
 species_labels <- c("Bearded seal" = paste0("<span style = 'color:",
@@
 p <- ggplot() +
-  layer_spatial(survey_area, color = "#ddc000", 
-                fill = "#ddc000", 
+  layer_spatial(survey_area, color = "#A26E7C", 
+                fill = "#A26E7C", 
                 alpha = 0.15,
@@
                      fill = "grey80") +
-  geom_sf_text(data = label_coords_bering, aes(label = "Bering sea"), size = 1.75, color = "grey40") +
-  geom_sf_text(data = label_coords_chukchi, aes(label = "Chuckchi \nsea"), size = 1.75, color = "grey40") +
-  geom_sf_text(data = label_coords_beaufort, aes(label = "Beaufort \nsea"), size = 1.75, color = "grey40") +
+  geom_sf_text(data = label_coords_bering, aes(label = "Bering sea"), size = 1.65, color = "grey30") +
+  geom_sf_text(data = label_coords_chukchi, 
+               aes(label = "Chuckchi \nsea"), size = 1.65, lineheight = .85, color = "grey30") +
+  geom_sf_text(data = label_coords_beaufort, 
+               aes(label = "Beaufort \nsea"), size = 1.65, lineheight = .85, color = "grey30") +
   facet_grid(cols = vars(species), labeller = species_labels,
              switch="x") +
-  scale_fill_manual(values = met.brewer("Isfahan2",n=3,direction=-1,override.order = TRUE),
+  scale_fill_manual(values = species_colors,
                     aesthetics = "fill", 
                     name = "species") + 
-  ggtitle("Initial bio-logger deployment locations by species",
+  labs(title = "Initial bio-logger deployment locations by species",
           subtitle = 
@@
                 "<span style = 'color:",species_colors['Spotted seal'],";'>spotted</span> seals ", 
-                "across multiple regions within the Bering, Chukchi, and western Beaufort seas.")) +
-  labs(caption = "shaded yellow region indicates spatial extent of aerial survey effort in US waters") +
+                "across multiple regions within the Bering, Chukchi, and western Beaufort seas."),
+       caption = "outlined region indicates spatial extent of aerial survey effort in US waters") +
   theme(legend.position = "none",
@@
         strip.text.x = element_markdown(face = "bold"),
-        plot.subtitle = element_textbox_simple(margin = ggplot2::margin(b=12)))
+        plot.caption = element_markdown(color = "#A26E7C")
+  )
 
@@
 This actogram plot represents the transmitted hourly percent-dry values.
-Areas indicated with an 'X' represent missing data that were not successfully recieved from the tag."}
+Not all data during a deployment are reliably transmitted from the bio-logger and since data are transmitted in 24-hour chunks occasional missing days are present in the record.
+Here, missing data that were not successfully recieved from the device are represented as empty rectangles."}
 
@@
 p <- ggplot(d,aes(day,hour,fill=percent_dry))+
-  geom_tile(color= "white",size=0) + 
-  scale_fill_gradientn(colors = rev(met.brewer("Hiroshige")),
+  geom_tile(linewidth = 0.1, color = "grey75") + 
+  scico::scale_fill_scico(palette = "nuuk", direction = 1,
                        name = "hourly percent dry", aesthetics = "fill",
@@
                            barheight = 0.5, title.hjust = 0.5)
+    ) +
+  geom_tile(data = missing_hours, fill = NA, color = "grey30", linewidth = 0.2) +
+  coord_cartesian(clip = "off")
+
+label_df <- data.frame(year = 2016, month = "Apr",day = 1, hour = 12,
+                       percent_dry = 0,
+                       lab = "**Missing Data**<br>not all data are reliably received from the bio-logger") |> 
+  mutate(month = forcats::fct_relevel(month,c("Mar","Apr","May","Jun")))
+
+curve_df <- data.frame(year = 2016, 
+                       month = "Apr", 
+                       day = 1, day_end = 2, 
+                       hour = 8, hour_end = 0,
+                       percent_dry = 0) |> 
+  mutate(month = forcats::fct_relevel(month,c("Mar","Apr","May","Jun")))
+
+p <- p +
+  geom_textbox(
+    data = label_df,
+    mapping = aes(day, hour, label = lab),
+    width = grid::unit(0.65, "npc"), # 65% of plot panel width
+    halign = 0, valign = 0.5,
+    nudge_y = 2,
+    size = 2.75, box.size = NA,
+    lineheight = .9,
+    fill = NA, 
+    label.color = NA,
+  ) +
+  geom_curve(
+    data = curve_df,
+    mapping = aes(x = day, y = hour, xend = day_end, yend = hour_end),
+    curvature = 0
   )
@@
 p <- p + scale_x_continuous(breaks = c(5,15,25), expand = c(0,0)) +
-  scale_y_continuous(breaks = c(4,12,20))
+  scale_y_continuous(breaks = c(4,12,20), labels = c("04:00","12:00","20:00"))
+
 p <- p + theme(legend.position = "bottom") +
@@
   ggtitle("Example percent-dry actogram from bio-logger data",
-          subtitle = str_wrap("Observation records shown are from a single ribbon seal across a two year deployment. Note that not all records are received from the device.", width = 85)) +
+          subtitle = "Observation records shown are from a single ribbon seal across a two year deployment.") +
   xlab("day of month") + ylab("hour (UTC)")
-p <- p + 
-  geom_point(aes(x = day, y = hour), data = missing_hours, 
-             color = "black", shape = 4, size=0.8)
+
 p
@@
 covariate_tbl %>% 
-  select(-Source) %>% 
   flextable() %>% 
@@
 (ref:mapCaption) Spatial distribution of haul-out records during the months of March through July 15 for each of the three species.
-Data were collated across all years between 2005 and 2020.
-Each hexagonal cell represents an area of 50 km^2^
+Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model.
+For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km^2^
 
@@
   annotation_spatial(world,
-                     size = 0, 
+                     linewidth = 0,
                      fill = "grey80") +
   layer_spatial(
-    data = locs_count, size = 0.125,
-    aes(fill = ct)
+    data = locs_count,
+    aes(fill = ct),
+    linewidth = 0.1,
+    color = "grey75"
   ) +
-  scale_fill_gradientn(colors = rev(met.brewer("Greek")),
-                    trans = "log10", aesthetics = "fill",
+  scico::scale_fill_scico(
+    palette = "imola",
+    trans = "log10",
+    aesthetics = "fill",
     name = "number of locations",
-    guide = guide_colorbar(title.position = "bottom", barwidth = 15, 
-                           barheight = 0.5, title.hjust = 0.5)) + 
-  facet_grid(~ species) +
-  theme(legend.position = "bottom") +
-  ggtitle("Spatial distribution of haul-out behavior records") +
-  labs(caption = bquote("One hexagonal cell represents 50"~km^2))
+    guide = guide_colorbar(
+      title.position = "bottom",
+      barwidth = 15,
+      barheight = 0.5,
+      title.hjust = 0.5
+    )
+  ) +
+  facet_grid( ~ species) +
+  labs(
+    title = "Spatial distribution of haul-out behavior records",
+    subtitle = "Location estimates were linked by time with haul-out records. 
+    Data are shown collated across 2005-2020 between 01 March-15 July
+    for each of the three species.",
+    caption = bquote("One hexagonal cell represents 50" ~ km ^ 2)
+  ) +
+  theme(
+    legend.position = "bottom"
+  )
 ```
@@
   geom_col() +
-        scale_fill_gradientn(colors = rev(met.brewer("Greek"))) +
+        scico::scale_fill_scico(palette = "oslo", direction = -1) +
         scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
@@
   facet_grid(species ~ .) + 
+  labs(fill = "seal-hours of bio-logger data",
+       title = "Seasonal density of haul-out behavior observations by species",
+       subtitle = "Data are grouped by day-of-year and presented in seal-hours collated across 
+       years (2005-2020) from March through July.") +
   theme(axis.text.x = element_text(hjust = 0),
@@
         panel.grid = element_blank(),
-        legend.position = "bottom") +
-  labs(fill = "seal-hours of bio-logger data") +
-  ggtitle("Density of haul-out behavior observations by species",
-          subtitle = str_wrap("data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020) from March through July"))
+        legend.position = "bottom",
+        plot.subtitle = element_textbox_simple(
+          color = "grey30",
+          margin = ggplot2::margin(t = 0, b = 0, r = 60)
+        )
+  )
 
@@
 
-p <- ggplot(plot_df,aes(day,solar_hour,fill = ho_prob))+
-  geom_tile(color= "white",size = 0) +
-  scale_fill_gradientn(colors = rev(met.brewer("Hiroshige")),
-                       name = "haul-out probability", aesthetics = "fill",
+p <- ggplot(plot_df, aes(day,solar_hour,fill = ho_prob, color = ho_prob)) +
+  geom_tile(linewidth = 0.15) +
+  scico::scale_fill_scico(palette = "navia", direction = 1,
+                          begin = 0.1, end = 1,
+                       name = "haul-out probability", 
                        limits = c(0,1), breaks = c(0.25, 0.50, 0.75),
@@
                                               barheight = 0.5, title.hjust = 0.5)
-  # scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
-  #                      guide = guide_colorbar(title.position = "bottom", barwidth = 15,
-  #                                             barheight = 0.5, title.hjust = 0.5)
-  )
-p <- p + facet_grid(age_sex~month)
+  ) +
+  scico::scale_color_scico(palette = "navia", direction = 1,
+                           begin = 0.1, end = 1,
+                           limits = c(0,1)) +
+  guides(col = FALSE)
+
+p <- p + facet_grid(age_sex~month,
+                    labeller = labeller(
+                      age_sex = age_sex_labels
+                    ))
 p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
-  scale_y_continuous(breaks = c(4,12,20))
-p <- p + theme(legend.position = "bottom") +
-  theme(strip.background = element_rect(colour="white")) +
-  theme(axis.ticks=element_blank()) +
+  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20"))
+p <- p + theme(legend.position = "bottom",
+               strip.background = element_rect(colour="white"),
+               axis.ticks=element_blank()) + 
   xlab("day of month") + ylab("local solar hour") +
-  ggtitle("Bearded seals rest on sea ice around solar noon and solar midnight")
+  labs(title = "Bearded seal predicted haul-out probability",
+       subtitle = "Our model predicts lower haul-out probability for bearded seals overall 
+       compared to ribbon and spotted seals. Highest probability is associated with solar noon and midnight.")
 p
@@
 ```{r beardedHOwx, fig.asp = 1, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
-Hour of the day was fixed at local solar noon and day-of-year held at the day of peak haul-out for bearded seals.
+Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
 Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
@@
 ```{r ribbonHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model.
-Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males.
-'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes.
-Hour of the day was fixed at local solar noon."}
+There is an apparent seasonal progression with subadults hauling out earlier in the season followed by adult males and, then, adult females.
+Predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities."}
 
@@
 
-p <- ggplot(plot_df, aes(day,solar_hour,fill = ho_prob)) +
-  geom_tile(color = "white",size = 0) +
-  scale_fill_gradientn(colors = rev(met.brewer("Hiroshige")),
-                       name = "haul-out probability", aesthetics = "fill",
+p <- ggplot(plot_df, aes(day,solar_hour,fill = ho_prob, color = ho_prob)) +
+  geom_tile(linewidth = 0.15) +
+  scico::scale_fill_scico(palette = "navia", direction = 1,
+                          begin = 0.1, end = 1,
+                       name = "haul-out probability", 
                        limits = c(0,1), breaks = c(0.25, 0.50, 0.75),
@@
                                               barheight = 0.5, title.hjust = 0.5)
-  # scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
-  #                      guide = guide_colorbar(title.position = "bottom", barwidth = 15,
-  #                                             barheight = 0.5, title.hjust = 0.5)
-  )
-p <- p + facet_grid(age_sex~month)
+  ) +
+  scico::scale_color_scico(palette = "navia", direction = 1,
+                           begin = 0.1, end = 1,
+                           limits = c(0,1)) +
+  guides(col = FALSE)
+
+p <- p + facet_grid(age_sex~month,
+                    labeller = labeller(
+                      age_sex = age_sex_labels
+                    ))
 p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
-  scale_y_continuous(breaks = c(4,12,20))
-p <- p + theme(legend.position = "bottom") +
-  theme(strip.background = element_rect(colour = "white")) +
-  theme(axis.ticks = element_blank()) +
+  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20"))
+p <- p + theme(legend.position = "bottom",
+               strip.background = element_rect(colour = "white"),
+               axis.ticks = element_blank()) +
   xlab("day of month") + ylab("local solar hour") +
-  ggtitle("Ribbon seals rest on sea ice in late spring and around solar noon.")
+  ggtitle("Ribbon seal predicted haul-out probability.",
+          subtitle = "Our model predicts increased haul-out probability 
+          for ribbon seals centered on solar noon and a sequential 
+          seasonal pattern linked to age and sex classifications.")
 p
@@
 ```{r ribbonHOwx, fig.asp = 0.9, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification.
-Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for ribbon seals.
+Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
 Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
@@
 ```{r spottedHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model.
-Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males.
-'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes.
-Hour of the day was fixed at local solar noon."}
+Compared to ribbon seals, the period of increased haul-out behavior is broader and there is more alignment in the seasonal timing of haul-out behavior across age and sex classifications.
+This likely reflects the triad behavior in spotted seals when suitor males haul out with nursing females.
+As with ribbon seals, redictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities."}
 
@@
 
-p <- ggplot(plot_df,aes(day,solar_hour,fill=ho_prob))+
-  geom_tile(color= "white",size = 0) +
-  scale_fill_gradientn(colors = rev(met.brewer("Hiroshige")),
-                       name = "haul-out probability", aesthetics = "fill",
+p <- ggplot(plot_df, aes(day,solar_hour,fill = ho_prob, color = ho_prob)) +
+  geom_tile(linewidth = 0.15) +
+  scico::scale_fill_scico(palette = "navia", direction = 1,
+                          begin = 0.1, end = 1,
+                       name = "haul-out probability", 
                        limits = c(0,1), breaks = c(0.25, 0.50, 0.75),
@@
                                               barheight = 0.5, title.hjust = 0.5)
-  # scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
-  #                      guide = guide_colorbar(title.position = "bottom", barwidth = 15,
-  #                                             barheight = 0.5, title.hjust = 0.5)
-  )
-p <- p + facet_grid(age_sex~month)
+  ) +
+  scico::scale_color_scico(palette = "navia", direction = 1,
+                           begin = 0.1, end = 1,
+                           limits = c(0,1)) +
+  guides(col = FALSE)
+
+p <- p + facet_grid(age_sex~month,
+                    labeller = labeller(
+                      age_sex = age_sex_labels
+                    ))
 p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
-  scale_y_continuous(breaks = c(4,12,20))
+  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20"))
 p <- p + theme(legend.position = "bottom") +
@@
   xlab("day of month") + ylab("local solar hour") +
-  ggtitle("Spotted seals rest on sea ice in late spring and around solar noon.")
+  ggtitle("Spotted seal predicted haul-out probability.",
+          subtitle = "Our model predicts increased haul-out probability 
+          for spotted seals centered on solar noon with general alignment in the
+          seasonal timing across age and sex classifications.")
 p
@@
 ```{r spottedHOwx,fig.asp = 0.9, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
-Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for spotted seals.
+Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
 Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
@@
 
-pal <- c(RColorBrewer::brewer.pal(12, "Paired"), "black")
-
-p <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(yday, ho_prob)) +
-  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.25, 
-                  stroke = 0, color = "grey80") +
-  geom_pointrange(data = peak_ho_df,
-                  mapping = aes(yday, ho_prob, ymin = 0, ymax = ho_prob, color = year),
-                  shape = 15) +
+pal <- c(rcartocolor::carto_pal(12,'Bold')[1:11],
+         rcartocolor::carto_pal(12,'Antique')[1:11])
+
+p <-
+  ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(yday, ho_prob)) +
+  geom_pointrange(
+    aes(ymin = lower95, ymax = upper95),
+    alpha = 0.25,
+    stroke = 0,
+    color = "grey80"
+  ) +
+  geom_pointrange(
+    data = peak_ho_df,
+    mapping = aes(
+      yday,
+      ho_prob,
+      ymin = 0,
+      ymax = ho_prob,
+      color = year
+    ),
+    shape = 15
+  ) +
   scale_color_discrete(type = pal) +
-  scale_x_continuous(breaks = c(74,105,135,166),
-                     labels = c("15-Mar", "15-Apr", "15-May", "15-Jun")) +
-  facet_grid(age_sex~species) +
-  theme(legend.position = "bottom") +
-  xlab("") + ylab("haul-out probability") +
-  ggtitle("Annual variability in the timing of peak haul-out probability")
+  scale_x_continuous(
+    breaks = c(74, 105, 135, 166),
+    labels = c("15-Mar", "15-Apr", "15-May", "15-Jun")
+  ) +
+  facet_grid(age_sex ~ species) +
+  ylab("haul-out probability") +
+  labs(title = "Annual variability in the timing of peak haul-out probability",
+       subtitle = "Predictions for ribbon and spotted seals are shown for local 
+       solar noon and under smoothed weather conditions") +
+  theme(legend.position = "bottom",
+        axis.title.x = element_blank())
+
   p

### 2023-09-25 R/wx_plots.R

@@ -1,4 +1,22 @@
-plot_ribbon_wx <- function(ribbon_fit_obj) {
+plot_ribbon_wx <- function(ribbon_fit_obj,age_sex_colors,age_sex_labels) {
+  
+  theme_set(theme_minimal(base_family = 'Spline Sans',
+                          base_size = 9))
+  theme_update(
+    plot.title.position = "plot",
+    # left-align title
+    plot.caption.position = "plot",
+    # right-align caption
+    plot.title = element_text(face = "bold"),
+    # larger, bold title
+    plot.subtitle = element_textbox_simple(
+      color = "grey30",
+      margin = ggplot2::margin(t = 0, b = 12, r = 60)
+    ),
+    plot.caption = element_text(color = "grey30"),
+    # change color of caption
+    panel.grid.minor = element_blank() # no minor grid lines
+  )
   
   p_temp <- create_ribbon_newdata(ribbon_fit_obj, 
                                   margins=TRUE, 
@@ -13,7 +31,8 @@ plot_ribbon_wx <- function(ribbon_fit_obj) {
     ggplot(aes(temp2m*27, ho_prob, color = age_sex)) +
     geom_pointrange(aes(ymin = lower95, ymax = upper95), 
                     size = 0.25, alpha = 1, stroke = 0) +
-    scale_color_manual(values= met.brewer("Egypt",n=3),
+    scale_color_manual(values= age_sex_colors,
+                       labels = age_sex_labels,
                        guide = guide_legend(override.aes = list(alpha = 1),
                                             direction = "horizontal")) +
     theme(legend.title = element_blank(),
@@ -33,7 +52,8 @@ plot_ribbon_wx <- function(ribbon_fit_obj) {
     ggplot(aes(wind*10, ho_prob, color = age_sex)) +
     geom_pointrange(aes(ymin = lower95, ymax = upper95), 
                     size = 0.25, alpha = 1, stroke = 0) +
-    scale_color_manual(values=met.brewer("Egypt",n=3),
+    scale_color_manual(values= age_sex_colors,
+                       labels = age_sex_labels,
                        guide = guide_legend(override.aes = list(alpha = 1),
                                             direction = "horizontal")) +
     theme(legend.title = element_blank(),
@@ -52,7 +72,8 @@ plot_ribbon_wx <- function(ribbon_fit_obj) {
     ggplot(aes(((pressure * 10000) + 100000)/1000, ho_prob, color = age_sex)) +
     geom_pointrange(aes(ymin = lower95, ymax = upper95), 
[…]

### 2023-09-25 _targets.R

@@ -12,6 +12,7 @@ tar_option_set(packages = c(
   "ggplot2",
   "sf",
   "ggspatial",
+  "ggtext",
   "lubridate",
   "RPostgres",
   "DBI",
@@ -84,9 +85,12 @@ list(
   tar_target(spotted_newdata, create_spotted_newdata(spotted_fit)),
   tar_target(bearded_newdata, create_bearded_newdata(bearded_fit)),
   
-  tar_target(ribbon_wx_plot, plot_ribbon_wx(ribbon_fit)),
-  tar_target(spotted_wx_plot, plot_spotted_wx(spotted_fit)),
-  tar_target(bearded_wx_plot, plot_bearded_wx(bearded_fit)),
+  tar_target(age_sex_colors, get_age_sex_colors()),
+  tar_target(age_sex_labels, get_age_sex_labels()),
+  
+  tar_target(ribbon_wx_plot, plot_ribbon_wx(ribbon_fit, age_sex_colors, age_sex_labels)),
+  tar_target(spotted_wx_plot, plot_spotted_wx(spotted_fit, age_sex_colors, age_sex_labels)),
+  tar_target(bearded_wx_plot, plot_bearded_wx(bearded_fit, age_sex_colors, age_sex_labels)),
 
   tar_target(ribbon_newdata_year, ribbon_newdata_yr(ribbon_year_fit)),
   tar_target(spotted_newdata_year, spotted_newdata_yr(spotted_year_fit))

### 2023-09-29 analysis/paper/London_HauloutBeahvior.Rmd
update sea-ice extent regression analysis

@@
 ```{r}
-load(here::here('data/seaice_extent.Rdata'))
-
-extent_max <- extent_df %>% 
-  group_by(year) %>% 
-  slice_max(area_km2_epsg3571) %>% 
-  ungroup() %>% 
+tar_load(sea_ice_extent)
+extent_max <- sea_ice_extent %>%
+  dplyr::bind_rows() %>%
+  dplyr::mutate(year = lubridate::year(fdate) %>% forcats::as_factor(),
+                yday = lubridate::yday(fdate)) %>% 
+  dplyr::filter(area_km2_epsg3571 > 345609.6) %>%
+  slice_max(area_km2_epsg3571, by=year) %>% 
   rename(max_extent_km2 = area_km2_epsg3571,
@@
 
-extent_lookup <- extent_df %>% 
-  dplyr::select(fdate, area_km2_epsg3571) %>% 
+extent_lookup <- sea_ice_extent %>%
+  dplyr::bind_rows() %>% 
   rename(extent_km2 = area_km2_epsg3571)
@@
 appeared to have little to no relationship with the amount of
-yearly maximum sea ice extent. Adult female and subadult spotted seals
+yearly maximum sea ice extent in the study area. Adult female and subadult spotted seals
 show a negative relationship, but there was no indication that the observed
@@
 
-ggplot(data = peak_ho_df, aes(y = yday, x = max_extent_km2, color = species)) + 
+ggplot(data = peak_ho_df, aes(y = max_extent_km2, x = yday, color = species)) + 
   geom_smooth(method = "lm", alpha = 0.2, aes(fill = species)) +
@@
   facet_grid(age_sex~.) +
-  scale_y_continuous(breaks = c(125,135,145, 156),
+  scale_x_continuous(breaks = c(125,135,145, 156),
                      labels = c("05-May", "15-May", "25-May", "05-Jun")) +
@@
         legend.justification = "right") +
-  labs(y = "date of peak haul-out", x = "maximum sea ice extent (km<sup>2</sup> / 1000)",
-       title = "The connection between yearly maximum sea ice extent and the \ntiming of peak haul-out probability is limited") +
-  theme(axis.title.x = element_markdown())
+  labs(x = "date of peak haul-out", y = "maximum sea ice extent (km<sup>2</sup> / 1000)",
+       title = "The relationship between yearly maximum sea ice extent and the \nannual timing of peak haul-out probability") +
+  theme(axis.title.y = element_markdown())
 

### 2023-09-29 R/sea_ice_extent.R

@@ -1,0 +1,53 @@
+qry_sea_ice_extent <- function(con,
+                               xmin,ymin,xmax,ymax,
+                               epsg,
+                               start_date,
+                               end_date) {
+  qry <- glue::glue(
+    "SELECT * FROM environ.fxn_sea_ice_extent_by_date(0.15, --concentration
+                                              {round(xmin)}, --xmin
+                                              {round(ymin)}, --ymin
+                                              {round(xmax)}, --xmax
+                                              {round(ymax)}, --ymax
+                                              {epsg}, --epsg of the bounding box
+                                              '{start_date}', --start_date
+                                              '{end_date}' -- end_date
+      )"
+  )
+  RPostgres::dbGetQuery(conn = con, 
+                        statement = qry)
+}
+
+get_sea_ice_extent <- function(min_year, max_year, bbox, epsg) {
+  tryCatch({
+    con <- dbConnect(RPostgres::Postgres(),
+                      dbname = 'pep', 
+                      host = Sys.getenv('PEP_PG_IP'),
+                      user = keyringr::get_kc_account("pgpep_londonj"),
+                      password = keyringr::decrypt_kc_pw("pgpep_londonj"))
+  },
+  error = function(cond) {
+    print("Unable to connect to Database.")
+  })
+  on.exit(dbDisconnect(con))
+  
+  res <- vector("list", length(min_year:max_year))
+  names(res) <- paste0('y',min_year:max_year)
+  
+  epsg <- strsplit(epsg,":")[[1]][2]
+  xmin <- bbox[1]
+  ymin <- bbox[2]
+  xmax <- bbox[3]
+  ymax <- bbox[4]
+  
+  for(y in names(res)) {
+    start_date <- paste0(substr(y,2,5),'-','03-01')
+    end_date <- paste0(substr(y,2,5),'-','07-31')
+    res[[y]] <- qry_sea_ice_extent(con,
+                                 xmin,ymin,xmax,ymax,
+                                 epsg,
+                                 start_date,end_date)
+  }
+  
+  return(res)
+}

### 2023-09-29 _targets.R

@@ -66,6 +66,8 @@ list(
   tar_target(source_data, create_source_data(locs_sf, timeline_data)),
 
   tar_target(analysis_data, create_data_sf(locs_sf, source_data)),
+  tar_target(sea_ice_extent, get_sea_ice_extent(2005,2022,st_bbox(analysis_data),
+                                                st_crs(analysis_data,parameters=TRUE)$srid)),
   tar_target(analysis_data_ssrd, add_ssrd(analysis_data)),
   # tar_target(grid, create_grid_sf(analysis_data_ssrd)),
   tar_target(deploy_table, create_deploy_tbl(analysis_data_ssrd)),

### 2023-09-29 R/sea_ice_extent.R
set date range for sea-ice extent 15Feb-15Jul

@@ -41,8 +41,8 @@ get_sea_ice_extent <- function(min_year, max_year, bbox, epsg) {
   ymax <- bbox[4]
   
   for(y in names(res)) {
-    start_date <- paste0(substr(y,2,5),'-','03-01')
-    end_date <- paste0(substr(y,2,5),'-','07-31')
+    start_date <- paste0(substr(y,2,5),'-','02-15')
+    end_date <- paste0(substr(y,2,5),'-','07-15')
     res[[y]] <- qry_sea_ice_extent(con,
                                  xmin,ymin,xmax,ymax,
                                  epsg,

### 2023-09-29 analysis/paper/London_HauloutBeahvior.Rmd
add square font-awesome to fig1 caption

@@
   ),
-  plot.caption = element_text(color = "grey30"),
+  plot.caption = element_markdown(color = "grey30"),
   # change color of caption
@@
                 alpha = 0.15,
-                linewidth = 0.5) +
+                linewidth = 0.35) +
   layer_spatial(deploy_sf, aes(fill=species),
@@
                 "across multiple regions within the Bering, Chukchi, and western Beaufort seas."),
-       caption = "outlined region indicates spatial extent of aerial survey effort in US waters") +
+       caption = "<span style = 'color: #A26E7C;'><span style = 'font-family: \"Font Awesome 6 Free Regular\";'>&#xf0c8;</span> spatial extent of aerial survey effort in US waters</span>") +
   theme(legend.position = "none",
@@
         strip.placement = "outside",
-        strip.text.x = element_markdown(face = "bold"),
-        plot.caption = element_markdown(color = "#A26E7C")
+        strip.text.x = element_markdown(face = "bold")
   )
@@
 ```{r}
-tar_load(sea_ice_extent)
+targets::tar_load(sea_ice_extent, store = here::here("_targets"))
 extent_max <- sea_ice_extent %>%
@@
 
-The timing of peak haul-out probability for ribbon and spotted seals
-appeared to have little to no relationship with the amount of
-yearly maximum sea ice extent in the study area. Adult female and subadult spotted seals
-show a negative relationship, but there was no indication that the observed
-trend is meaningful (spotted seal adult female: $R^{2}$ =
-`r pl_adultF$r.squared`, $p$ = `r pl_adultF$p.value`; spotted seal
-subadults: $R^{2}$ = `r pl_subadult$r.squared`, $p$ =
-`r pl_subadult$p.value`). For ribbon seals and adult male spotted seals,
+The annual timing of peak haul-out probability for ribbon seals and adult
+male spotted seals appeared to have
+no relationship with the amount of yearly maximum sea ice extent in the
+study area. For ribbon seals and adult male spotted seals,
 $p$-values were substantially larger than 0.05 (ribbon seal adult
@@
 adult males: $R^{2}$ = `r pl_adultM$r.squared`, $p$ =
-`r pl_adultM$p.value`).
+`r pl_adultM$p.value`). Adult female and subadult spotted seals both
+showed a negative trend (peak haul-out occurs later in years with less
+sea ice) with a significant relationship for adult female spotted seals 
+(spotted seal adult female: $R^{2}$ =
+`r pl_adultF$r.squared`, $p$ = `r pl_adultF$p.value`; spotted seal
+subadults: $R^{2}$ = `r pl_subadult$r.squared`, $p$ =
+`r pl_subadult$p.value`).
 

### 2023-10-02 analysis/paper/London_HauloutBeahvior.Rmd
more perfecting of figures/captions etc

@@
   percent-dry data from satellite-linked bio-loggers deployed
-  between 2005 and 2020 to quantify the proportion of seals hauled out on ice,
+  between 2005 and 2021 to quantify the proportion of seals hauled out on ice,
   which is needed to accurately estimate abundance from aerial survey counts of
@@
   supported, indicating that the timing and magnitude of haul-out behavior varied
-  among years. However, we did not find evidence that haul-out timing was linked
+  among years. However, we did not find broad evidence that haul-out timing was linked
   to annual sea ice extent. Our analysis emphasizes the importance of accounting
@@
 Chukchi, and western Beaufort seas by multiple organizations as part of
-collaborative investigations from 2005 through 2020. Seals were captured using
+collaborative investigations from 2005 through 2021. Seals were captured using
 nets and bio-loggers were attached during studies based in coastal communities
@@
 behavior was further used to regress against the annual maximum sea ice extent
-in the study region.
+in the study region. Predictions before 15 March and after 30 June were not
+included in visualizations or other evaluations to avoid spurious model
+predictions at the edge of the data range.
 
@@
 of weather covariate values while fixing hour of the day at local solar noon and
-day-of-year at the average day of peak haul-out for each species. The
+day-of-year at 15 May. The
 visualizations also include vertical lines representing 95% confidence intervals
@@
 
-```{r beardedHOCal, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes."}
+```{r beardedHOCal, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 30 June for all age and sex classes combined.
+Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range."}
 
@@
   mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
-         month = lubridate::month(date,label=TRUE),
+         month = lubridate::month(date,label = TRUE),
          day = lubridate::day(date)) %>%
-  filter(month != "Aug")
+  filter(!month %in% c("Jul","Aug"),
+         lubridate::yday(date) >73)
 
@@
 p <- p + facet_grid(age_sex~month,
+                    scales = "free_x",
+                    space = "free_x",
                     labeller = labeller(
@@
                     ))
-p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
-  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20"))
-p <- p + theme(legend.position = "bottom",
+p <- p + scale_x_continuous(breaks = c(5,10,15,20,25)) +
+  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20")) +
+  coord_cartesian(expand=FALSE)
+p <- p + theme(panel.spacing = unit(0.1, "lines"),
+               legend.position = "bottom",
                strip.background = element_rect(colour="white"),
@@
 young-of-the-year records begin after weaning and the model predictions
-demonstrated the ontogeny of in-water activities (e.g. diving, foraging)
+demonstrated the ontogeny of in-water activities (e.g. swimming, foraging)
 in May. Adult females had a more protracted haul-out season compared to
-males, and more time was spent resting in June and July. The model suggested
-adult male ribbon seals completed their molt by the end of June. Lastly,
-for adults of both sexes there was some indication for a shift to a
-crepuscular haul-out pattern in late June and July.
-
-```{r ribbonHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model.
+males, and more time was spent hauled out in June compared to adult males and
+subadults.
+
+```{r ribbonHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
 There is an apparent seasonal progression with subadults hauling out earlier in the season followed by adult males and, then, adult females.
-Predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities."}
+Predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities.
+Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range."}
 
@@
          day = lubridate::day(date)) %>%
-  filter(month != "Aug")
+  filter(!month %in% c("Jul","Aug"),
+         lubridate::yday(date) >73)
 
@@
                     ))
-p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
-  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20"))
-p <- p + theme(legend.position = "bottom",
-               strip.background = element_rect(colour = "white"),
-               axis.ticks = element_blank()) +
+p <- p + scale_x_continuous(breaks = c(5,10,15,20,25)) +
+  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20")) +
+  coord_cartesian(expand=FALSE)
+p <- p + theme(panel.spacing = unit(0.1, "lines"),
+               legend.position = "bottom",
+               strip.background = element_rect(colour="white"),
+               axis.ticks=element_blank()) +
   xlab("day of month") + ylab("local solar hour") +
@@
 In contrast to ribbon seals, adult spotted seal males had a more protracted
-haul-out season compared to females, and more time out of the water in June and
-July (Figures \@ref(fig:ribbonHOCal) and \@ref(fig:spottedHOCal)). As with
+haul-out season compared to females, and more time out of the water in June 
+(Figures \@ref(fig:ribbonHOCal) and \@ref(fig:spottedHOCal)). As with
 ribbon seals, the young-of-the-year records began after weaning and the model
-predictions demonstrated the ontogeny of in-water activities (e.g. diving,
+predictions demonstrated the development of in-water activities (e.g. swimming,
 foraging) in May.
 
-```{r spottedHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model.
+```{r spottedHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of spotted seals from 1 March to 30 June for each age and sex class used in the model.
 Compared to ribbon seals, the period of increased haul-out behavior is broader and there is more alignment in the seasonal timing of haul-out behavior across age and sex classifications.
 This likely reflects the triad behavior in spotted seals when suitor males haul out with nursing females.
-As with ribbon seals, redictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities."}
+As with ribbon seals, redictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities.
+Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range."}
 
@@
                     ))
-p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
-  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20"))
-p <- p + theme(legend.position = "bottom") +
-  theme(strip.background = element_rect(colour="white")) +
-  theme(axis.ticks=element_blank()) +
+p <- p + scale_x_continuous(breaks = c(5,10,15,20,25)) +
+  scale_y_continuous(breaks = c(4,12,20), labels = c("04","12", "20")) +
+  coord_cartesian(expand=FALSE)
+p <- p + theme(panel.spacing = unit(0.1, "lines"),
+               legend.position = "bottom",
+               strip.background = element_rect(colour="white"),
+               axis.ticks=element_blank()) +
   xlab("day of month") + ylab("local solar hour") +
@@
 
-```{r annualHO, fig.asp = 0.75, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability (colored markers) for ribbon and spotted seals across 13 years.
-Predictions are shown for local solar noon and under smoothed weather conditions.
+```{r annualHO, fig.asp = 0.75, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability (colored markers) for ribbon and spotted seals across 14 years.
+The full seasonal range of annual predictions along with uncertainty are shown as grey lines.
+Covariates were fixed at local solar noon and under smoothed weather conditions.
 Only those groups (age:sex + year) that included observations from more than one seal are shown.
@@
   ) +
-  facet_grid(age_sex ~ species) +
+  scale_y_continuous(breaks = c(0.2,0.4,0.6,0.8), labels = c("0.2","0.4", "0.6", "0.8")) +
+  guides(col = guide_legend(ncol = 7)) +
+  facet_grid(age_sex ~ species,
+             labeller = labeller(
+                      age_sex = age_sex_labels
+                    )) +
   ylab("haul-out probability") +
   labs(title = "Annual variability in the timing of peak haul-out probability",
-       subtitle = "Predictions for ribbon and spotted seals are shown for local 
-       solar noon and under smoothed weather conditions") +
+       subtitle = "Predictions of maximum annual haul-out probability for ribbon and spotted seals are indicated by the colored squares with grey lines showing the individual seasonal trajectories and uncertainty.",
+       caption = "predictions are shown for local solar noon and under smoothed weather conditions") +
   theme(legend.position = "bottom",

### 2023-10-03 analysis/paper/London_HauloutBeahvior.Rmd
update data availability

@@
   - \captionsetup{font={normalsize,onehalfspacing}}
+  - \newcommand{\beginsupplement}{\setcounter{table}{0}\renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0}\renewcommand{\thefigure}{S\arabic{figure}}}
 bibliography: references.json
@@
 for detailed capture and bio-logger attachment methods (see Supplemental
-Material, S1).
+Material, [S1](#s1)).
 
@@
 Deployments were initiated across a range of months but only data from 1 March to 15 July were included in the analysis.
-See Figure \@ref(fig:dataMap) for the spatial distribution of observed data used in the study and Supplemental Materials 1 for additional deployment details.
+See Figure \@ref(fig:dataMap) for the spatial distribution of observed data used in the study and Supplemental Materials \@ref(tab:deploymentTable) for additional deployment details.
 
@@
 
+\newpage
+\beginsupplement
 \blandscape
@@
 
-# S1. Supplemental Table of Bio-logger Deployment Details
-
-The following table provides additional details about the timing, location, and
-institutions responsible for the bio-logger deployments used in this study.
-
-```{r, include=TRUE}
+## Additional Bio-logger Deployment Details
+
+```{r deploymentTable, include=TRUE, tab.cap="The following table provides additional details about the timing, location, and institutions responsible for the bio-logger deployments used in this study along with research permits and any associated publications."}
 library(gt)
@@
   dplyr::select(species, agency, deploy_year, permit, location, speno, sex, age, publications) %>% 
+  dplyr::mutate(deploy_year = forcats::as_factor(deploy_year)) %>%
   dplyr::group_by(species, agency, deploy_year, location, publications) %>% 
@@
                    age_class = tolower(toString(unique(age)))) %>% 
-  dplyr::arrange(species, agency, deploy_year) %>% 
-  gt(
-    groupname_col = "species"
-  ) %>% 
-  cols_label(
+  dplyr::arrange(species, agency, deploy_year)
+
+  deploy_tbl <- as_grouped_data(deploy_tbl,groups="species") %>% 
+  flextable::as_flextable(hide_grouplabel = TRUE) %>% 
+    width(j = c('agency','deploy_year','location','publications','permits',
+                'n_seals', 'sex', 'age_class'), 
+          width = c(1, 0.8, 1.5, 1.5, 0.8, 0.8, 0.8, 1.5)) %>%
+    align(j = 'n_seals', align = 'left', part = 'all') %>%
+  set_header_labels(
     agency = "Institution",
@@
     sex = "Sex",
-    age_class = "Age Class"
+    age_class = "Age_Class"
   ) %>%
-  tab_style(
-    style = list(
-      cell_text(style = "italic")
-    ),
-    locations = cells_body(
-      columns = publications
-    )
-  ) %>% 
-  cols_align(
-    align = "center",
-    columns = everything()
-  ) %>% 
-  cols_align(
-    align = "right",
-    columns = age_class
-  ) %>% 
-  tab_header(
-    title = "Data Contributions From Collaborators and Additional Deployment Details"
-  ) %>%
-  tab_footnote(footnote = "ADF&G = Alaska Department of Fish and Game",
-               locations = cells_body(columns = agency, 
-                                      rows = agency == "ADFG")) %>%
-  tab_footnote(footnote = "NSB = North Slope Borough",
-               locations = cells_body(columns = agency, 
-                                      rows = agency == "NSB")) %>%
-  tab_footnote(footnote = "NMFS = NOAA National Marine Fisheries Service",
-               locations = cells_body(columns = agency, 
-                                      rows = agency == "NMFS")) %>%
-  tab_footnote(footnote = "NPWC = North Pacific Wildlife Consortium",
-               locations = cells_body(columns = agency, 
-                                      rows = agency == "NPWC"))
+  bold(j = 1, i = ~ !is.na(species), bold = TRUE, part = "body") %>%
+  bold(part = "header", bold = TRUE) %>%
+  add_footer_lines(as_paragraph(
+    as_b("ADFG"),"=Alaska Department of Fish and Game; ", as_b("NSB"),"=North Slope Borough; ", as_b("NMFS"),"=NOAA National Marine Fisheries Service; ", as_b("NPWC"),"=North Pacific Wildlife Consortium"
+  )) %>%
+    fontsize(size = 10, part ="header") %>%
+    fontsize(size = 9, part = "body") %>%
+    fontsize(size = 8, part = "footer") %>%
+  set_table_properties(layout = "fixed",
+                       opts_pdf = list('caption_repeat' = FALSE))
 
@@
 \elandscape
+
+## Supplemental Figures Showing Confidence Intervals Associated with Predictions
+
+The following series of figures show the seasonal variability in haul-out
+probability and the associated confidence intervals for bearded, ribbon, and
+spotted seals.
+
+```{r beardedPredSE,fig.asp = 0.5, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for bearded seals.
+Model predictions are shown for local solar noon with smoothed weather covariates.
+Age and sex classes are combined into a single 'all ages' category."}
+plot_df <- bearded_newdata %>%
+  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+         month = lubridate::month(date,label = TRUE),
+         day = lubridate::day(date)) %>%
+  filter(!month %in% c("Jul","Aug"),
+         lubridate::yday(date) >73,
+         solar_hour == 12)
+
+bearded_predse <- ggplot(plot_df, aes(date,ho_prob)) +
+  geom_ribbon(aes(ymin=lower95,ymax=upper95), 
+              fill = species_colors['Bearded seal'], alpha = 0.3) +
+  geom_line(aes(date,ho_prob),
+            color = species_colors['Bearded seal']) +
+  facet_grid(age_sex ~ .,
+             scales = "free_x",
+                    space = "free_x",
+                    labeller = labeller(
+                      age_sex = age_sex_labels
+                    )) +
+  scale_y_continuous(breaks = c(0.2,0.4,0.6,0.8), labels = c("0.2","0.4", "0.6", "0.8"), limits = c(0,1)) +
+  xlab("day of year") +
+  ylab("haul-out probability") +
+  labs(title = "Bearded seal predicted haul-out probability",
+       subtitle = "predictions and associated 95% confidence intervals are shown for local solar noon and with smoothed weather covariates") +
+  theme(legend.position = "none")
+
+bearded_predse
+```
+
+```{r ribbonPredSE,fig.asp = 0.6, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for ribbon seals.
+Model predictions are shown for local solar noon with smoothed weather covariates.
+Age and sex classes are separated to allow comparisons."}
+plot_df <- ribbon_newdata %>%
+  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+         month = lubridate::month(date,label = TRUE),
+         day = lubridate::day(date)) %>%
+  filter(!month %in% c("Jul","Aug"),
+         lubridate::yday(date) >73,
+         solar_hour == 12)
+
+ribbon_predse <- ggplot(plot_df, aes(date,ho_prob)) +
+  geom_ribbon(aes(ymin=lower95,ymax=upper95), 
+              fill = species_colors['Ribbon seal'], alpha = 0.3) +
+  geom_line(aes(date,ho_prob),
+            color = species_colors['Ribbon seal']) +
+  facet_grid(age_sex ~ .,
+             scales = "free_x",
+                    space = "free_x",
+                    labeller = labeller(
+                      age_sex = age_sex_labels
+                    )) +
+  scale_y_continuous(breaks = c(0.2,0.4,0.6,0.8), labels = c("0.2","0.4", "0.6", "0.8"), limits = c(0,1)) +
+  xlab("day of year") +
+  ylab("haul-out probability") +
+  labs(title = "Ribbon seal predicted haul-out probability",
+       subtitle = "predictions and associated 95% confidence intervals are shown for local solar noon and with smoothed weather covariates") +
+  theme(legend.position = "none")
+
+ribbon_predse
+```
+
+```{r spottedPredSE,fig.asp = 0.6, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for spotted seals.
+Model predictions are shown for local solar noon with smoothed weather covariates.
+Age and sex classes are separated to allow comparisons."}
+plot_df <- spotted_newdata %>%
+  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+         month = lubridate::month(date,label = TRUE),
+         day = lubridate::day(date)) %>%
+  filter(!month %in% c("Jul","Aug"),
+         lubridate::yday(date) >73,
+         solar_hour == 12)
+
+spotted_predse <- ggplot(plot_df, aes(date,ho_prob)) +
+  geom_ribbon(aes(ymin=lower95,ymax=upper95), 
+              fill = species_colors['Spotted seal'], alpha = 0.3) +
+  geom_line(aes(date,ho_prob),
+            color = species_colors['Spotted seal']) +
+  facet_grid(age_sex ~ .,
+             scales = "free_x",
+                    space = "free_x",
+                    labeller = labeller(
+                      age_sex = age_sex_labels
+                    )) +
+  scale_y_continuous(breaks = c(0.2,0.4,0.6,0.8), labels = c("0.2","0.4", "0.6", "0.8"), limits = c(0,1)) +
+  xlab("day of year") +
+  ylab("haul-out probability") +
+  labs(title = "Spotted seal predicted haul-out probability",
+       subtitle = "predictions and associated 95% confidence intervals are shown for local solar noon and with smoothed weather covariates") +
+  theme(legend.position = "none")
+
+spotted_predse
+```

### 2023-10-04 analysis/paper/London_HauloutBeahvior.Rmd
fix superscript error and add refs to sup figs
superscript type error in map caption for km^2 fixed; also added references to the new supplemental plots within main text.

@@
     for each of the three species.",
-    caption = bquote("One hexagonal cell represents 50" ~ km ^ 2)
+    caption = "One hexagonal cell represents 50 km^2^"
   ) +
@@
 lower sample size for adult and young-of-year age classes. As such, results are
-shown for all ages (Figure \@ref(fig:beardedHOCal)). Additionally, after
+shown for all ages (Figure \@ref(fig:beardedHOCal); see also 
+@\ref(fig:beardedPredSE)). Additionally, after
 approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
@@
 in April that peaks in late May for subadults and in early June for
-adults (Figure \@ref(fig:ribbonHOCal)). The behavior was clearly centered
+adults (Figure \@ref(fig:ribbonHOCal); see also 
+@\ref(fig:ribbonPredSE)). The behavior was clearly centered
 around local solar noon and expanded to other hours later in the spring
@@
   xlab("day of month") + ylab("local solar hour") +
-  ggtitle("Ribbon seal predicted haul-out probability.",
+  ggtitle("Ribbon seal predicted haul-out probability",
           subtitle = "Our model predicts increased haul-out probability 
@@
 Compared to ribbon seals, spotted seals showed a longer spring haul-out season
-that was less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)).
-Adults of both sexes spent considerable time in April, May and June hauled out.
-In contrast to ribbon seals, adult spotted seal males had a more protracted
-haul-out season compared to females, and more time out of the water in June 
-(Figures \@ref(fig:ribbonHOCal) and \@ref(fig:spottedHOCal)). As with
-ribbon seals, the young-of-the-year records began after weaning and the model
-predictions demonstrated the development of in-water activities (e.g. swimming,
-foraging) in May.
+that was less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal);
+see also @\ref(fig:spottedPredSE)). Adults of both sexes spent considerable time
+in April, May and June hauled out. In contrast to ribbon seals, adult spotted
+seal males had a more protracted haul-out season compared to females, and more
+time out of the water in June (Figures \@ref(fig:ribbonHOCal) and
+\@ref(fig:spottedHOCal)). As with ribbon seals, the young-of-the-year records
+began after weaning and the model predictions demonstrated the development of
+in-water activities (e.g. swimming, foraging) in May.
 
@@
   xlab("day of month") + ylab("local solar hour") +
-  ggtitle("Spotted seal predicted haul-out probability.",
+  ggtitle("Spotted seal predicted haul-out probability",
           subtitle = "Our model predicts increased haul-out probability 
@@
 This manuscript was developed as a reproducible research compendium. All data and
-code are available on GitHub (https://github.com/jmlondon/berchukHaulout) and
+code are available on GitHub (https://github.com/noaa-afsc/berchukseals-haulout) and
 major versions archived at Zenodo (https://doi.org/10.5281/zenodo.4638221).
@@
 at the United States Animal Telemetry Network, archived at Movebank, or
-associated with other published manuscripts (see supplemental material S1).
+associated with other published manuscripts (see supplemental 
+material \@ref(tab:deploymentTable)). Collated and cleaned data products
+needed to replicate the analysis are available in a separate GitHub
+repository (https://github.com/noaa-afsc/berchukseals-data) and archived at 
+Zenodo. Similarly, all model fits are available on 
+GitHub (https://github.com/noaa-afsc/berchukseals-fits) and archived at Zenodo.
 
@@
 
-The following series of figures show the seasonal variability in haul-out
-probability and the associated confidence intervals for bearded, ribbon, and
-spotted seals.
+The following series of figures (\@ref(fig:beardedPredSE), \@ref(fig:ribbonPredSE),
+and \@ref(fig:spottedPredSE)) show the seasonal variability in predicted haul-out
+probability and the associated 95% confidence intervals for bearded, ribbon, and
+spotted seals. The predictions shown are based on the same data used in
+\@ref(fig:beardedHOCal), \@ref(fig:ribbonHOCal), and \@ref(fig:spottedHOCal) but
+selected for three to local solar hours (07:00, 12:00, and 17:00) so the 
+confidence intervals can also be shown and comparisons can be made.
+
+```{r}
+local_solar_labels <- c(
+  '7' = '07:00',
+  '12' = '12:00',
+  '17' = '17:00'
+)
+```
 
 ```{r beardedPredSE,fig.asp = 0.5, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for bearded seals.
-Model predictions are shown for local solar noon with smoothed weather covariates.
+Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00) with smoothed weather covariates.
 Age and sex classes are combined into a single 'all ages' category."}
@@
          lubridate::yday(date) >73,
-         solar_hour == 12)
+         solar_hour %in% c(7,12,17))
 
@@
             color = species_colors['Bearded seal']) +
-  facet_grid(age_sex ~ .,
+  facet_grid(age_sex ~ solar_hour,
              scales = "free_x",
@@
                     labeller = labeller(
-                      age_sex = age_sex_labels
+                      age_sex = age_sex_labels,
+                      solar_hour = local_solar_labels
                     )) +
@@
   ylab("haul-out probability") +
-  labs(title = "Bearded seal predicted haul-out probability",
-       subtitle = "predictions and associated 95% confidence intervals are shown for local solar noon and with smoothed weather covariates") +
+  labs(title = "Bearded seal predicted haul-out probability with uncertainty",
+       subtitle = "predictions and associated 95% confidence intervals are shown for three **local solar** times and with smoothed weather covariates") +
   theme(legend.position = "none")
@@
 
-```{r ribbonPredSE,fig.asp = 0.6, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for ribbon seals.
-Model predictions are shown for local solar noon with smoothed weather covariates.
+```{r ribbonPredSE,fig.asp = 0.7, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for ribbon seals.
+Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00) with smoothed weather covariates.
 Age and sex classes are separated to allow comparisons."}
@@
             color = species_colors['Ribbon seal']) +
-  facet_grid(age_sex ~ .,
+  facet_grid(age_sex ~ solar_hour,
              scales = "free_x",
@@
   ylab("haul-out probability") +
-  labs(title = "Ribbon seal predicted haul-out probability",
-       subtitle = "predictions and associated 95% confidence intervals are shown for local solar noon and with smoothed weather covariates") +
+  labs(title = "Ribbon seal predicted haul-out probability with uncertainty",
+       subtitle = "predictions and associated 95% confidence intervals are shown for three **local solar** times and with smoothed weather covariates") +
   theme(legend.position = "none")
@@
 
-```{r spottedPredSE,fig.asp = 0.6, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for spotted seals.
-Model predictions are shown for local solar noon with smoothed weather covariates.
+```{r spottedPredSE,fig.asp = 0.7, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for spotted seals.
+Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00) with smoothed weather covariates.
 Age and sex classes are separated to allow comparisons."}
@@
             color = species_colors['Spotted seal']) +
-  facet_grid(age_sex ~ .,
+  facet_grid(age_sex ~ solar_hour,
              scales = "free_x",
@@
   ylab("haul-out probability") +
-  labs(title = "Spotted seal predicted haul-out probability",
-       subtitle = "predictions and associated 95% confidence intervals are shown for local solar noon and with smoothed weather covariates") +
+  labs(title = "Spotted seal predicted haul-out probability with uncertainty",
+       subtitle = "predictions and associated 95% confidence intervals are shown for three **local solar** times and with smoothed weather covariates") +
   theme(legend.position = "none")

### 2023-10-17 analysis/paper/London_HauloutBeahvior.Rmd
solar radiation supplemental + fits

@@
 library(MetBrewer)
+library(terra)
 
@@
   merge_h(part = "header") %>%
-  align(align = "left", part = "header") %>% 
-  align(i = 1, align = "center", part = "header") %>% 
+  flextable::align(align = "left", part = "header") %>% 
+  flextable::align(i = 1, align = "center", part = "header") %>% 
   fontsize(size = 9, part = "all") %>% 
-  width(width = 1.35) %>% 
-  width(j = ~ Species, width = 0.9) %>% 
-  width(j = ~ Sex, width = 0.4)
+  flextable::width(width = 1.35) %>% 
+  flextable::width(j = ~ Species, width = 0.9) %>% 
+  flextable::width(j = ~ Sex, width = 0.4)
 
@@
   hline(i=8) %>% 
-  width(width = 1.0) %>% 
-  width(j = ~ Description, width = 2.5)
+  flextable::width(width = 1.0) %>% 
+  flextable::width(j = ~ Description, width = 2.5)
 ```
@@
   flextable::as_flextable(hide_grouplabel = TRUE) %>% 
-    width(j = c('agency','deploy_year','location','publications','permits',
+    flextable::width(j = c('agency','deploy_year','location','publications','permits',
                 'n_seals', 'sex', 'age_class'), 
           width = c(1, 0.8, 1.5, 1.5, 0.8, 0.8, 0.8, 1.5)) %>%
-    align(j = 'n_seals', align = 'left', part = 'all') %>%
+    flextable::align(j = 'n_seals', align = 'left', part = 'all') %>%
   set_header_labels(
@@
 ```
+
+## Exploring Insolation (Solar Radiation) as a Model Covariate
+
+### Introduction
+
+During the peer review process for this manuscript, Anthony Fishbach suggested
+the possibility of using predicted insolation (or solar radiation) values from
+the reanalysis model as a more direct and, potentially, more informative
+predictor of the daily haul-out cycle in seals compared to time of day. The
+notion being that seals are directly responding to changes in solar radiation
+throughout the day and not what time of day it is (i.e. seals don't have
+watches). Additionally, given the energetic benefits of increased solar
+radiation it could be more informative as we would expect seals might have a
+higher haul-out probability on sunnier days and for there to be geographic
+variability in haul-out behavior associated with geographical differences in
+insolation. This approach has an additional benefit of being more parsimonious
+compared to our use of the Fourier series or other approaches to represent
+hour-of-day in the model (e.g. 24 factors for each hour).
+
+Because of these reasons, we spent some time considering and exploring this
+possibility for our model and the analysis presented in this manuscript. A key
+drawback to reliance on solar insolation, in our minds, is that we would lose
+insight regarding potential crepuscular patterns -- insolation does not 
+differentiate between dusk or dawn. Bi-modal patterns have been
+previously observed in ringed seals and our results in this analysis show some indication of
+increased haul-out probability during dawn and dusk periods for bearded seals.
+For other phocid speices, increased haul-out probability before solar noon or after
+solar noon has been observed. Importantly, understanding these relationships
+between haul-out probability and hour-of-day can have important ramifications
+on aerial survey study design -- a key focus of this paper.  
+Another hesitation we had was that insolation estimates from reanalysis models have
+not been previously used as a model covariate within a published study of 
+pinniped haul-out behavior. Thus, for this analysis, we chose to keep our
+original approach and rely on the Fourier series to capture any hour-of-day
+effects.
+
+That said, we think the idea of insolation as a model covariate in pinniped
+haul-out models is intriguing and worth further exploration. The current
+availability and increased accessibility to detailed climate reanalysis
+products that include insolation is exciting and we encrourage future, more
+detailed exploration of this as a component in pinniped haul-out analysis. To
+provide some inspiration, we present some initial efforts and model 
+comparisons.
+
+### Methods
+
+In this manuscript, we rely on the NARR reanalysis model as the source for our
+weather covariates. However, since our initiation of this analysis, the ERA5
+reanalysis model (https://doi.org/10.24381/cds.adbb2d47) has become one of the
+go-to standards for global climate reanalysis and provides an increased temporal
+resolution to hourly (compared to the 3-hour resolution of NARR). The global
+coverage of ERA5 provides additional flexibility in that the area of interest is
+not limited to North America. The ERA5 model provides a number of solar
+radiation parameters and it is important to evaluate and understand each of
+these estimates in order to select the one that is likely most relevant to
+seals. Here, we use the 'surface short-wave (solar) radiation downwards.' This
+parameter is described as _the amount of solar radiation (also known as shortwave radiation)
+that reaches a horizontal plane at the surface of the Earth and comprises both
+direct and diffuse solar radiation. To a reasonably good approximation, this
+parameter is the model equivalent of what would be measured by a pyranometer (an
+instrument used for measuring solar radiation) at the surface_ 
+(https://codes.ecmwf.int/grib/param-db/?id=169). Thus, this is the
+parameter which most closely represents the amount of solar radiation likely
+felt by a seal hauled out of the water.
+
+ERA5 data is available via the Copernicus climate data store API which can be
+queried with the CDS-API Python package 
+(https://cds.climate.copernicus.eu/api-how-to). The R code provided here
+documents the download of the _surface_solar_radiation_downwards_ parameter for
+our study area of interest and years of interest. The `reticulate` R package
+(https://CRAN.R-project.org/package=reticulate) allows interaction with Python.
+Additionally, note, extra steps are required to download data on either side
+of the 180 anti-meridian.
+
+```{r cdsapi, eval=FALSE, include=TRUE, echo = TRUE}
+library(tidyverse)
+library(reticulate)
+library(sf)
+library(terra)
+
+#import python CDS-API
+cdsapi <- import('cdsapi')
+#for this step there must exist the file .cdsapirc
+server = cdsapi$Client() #start the connection
+
+get_era5 <- function(y) {
+  #we create the query
+  query <- r_to_py(
+    list(
+      variable = "surface_solar_radiation_downwards",
+      product_type = "reanalysis",
+      area = "75/152/47/180", # North, West, South, East
+      year = y,
+      month = str_pad(2:7, 2, "left", "0"),
+      day = str_pad(1:31, 2, "left", "0"),
+      time = str_c(0:23, "00", sep = ":") %>% str_pad(5, "left", "0"),
+      format = "netcdf"
+    )
+  )
+  #query to get the ncdf
+  server$retrieve("reanalysis-era5-single-levels",
+                  query,
+                  paste0("era5_ssrd_", y, "_left.nc"))
+
+  query <- r_to_py(
+    list(
+      variable = "surface_solar_radiation_downwards",
+      product_type = "reanalysis",
+      area = "75/-180/47/-142", # North, West, South, East
+      year = y,
+      month = str_pad(2:7, 2, "left", "0"),
+      day = str_pad(1:31, 2, "left", "0"),
+      time = str_c(0:23, "00", sep = ":") %>% str_pad(5, "left", "0"),
+      format = "netcdf"
+    )
+  )
+  #query to get the ncdf
+  server$retrieve("reanalysis-era5-single-levels",
+                  query,
+                  paste0("era5_ssrd_", y, "_right.nc"))
+
+}
+
+years <- as.character(2005:2021)
+for(i in 1:length(years)) {
+  get_era5(years[i])
+}
+
+```
+
+To explore performance of our solar radiation parameter within a haul-out model
+we replace the various Fourier series parameters in our model from the manuscript
+with the ERA5 _surface solar radiation downwards_ (`era_ssrd`) parameter. As
+with other reanalysis values (from NARR) in the manuscript, the `era-ssrd`
+values are matched in time and space to the seal haul-out observation data. We
+use the full hourly temporal resolution from ERA5. The glmmLDTS framework used
+in the paper does not allow for model comparisons with AIC because of
+the reliance on pseudo-likelihood. The `bam` function within the `mgcv` package
+provides a quick model fitting option that also allows us to do some model
+comparison with AIC. This is sufficient for the general demonstration and
+exploartion purposes here
+but future research should consider a range of model fitting frameworks and
+approaches that might be appropriate.
+
+```{r load-model-data, include=FALSE}
+library(targets)
+tar_load(spotted_model_data,store = here::here("_targets"))
+tar_load(ribbon_model_data,store = here::here("_targets"))
+tar_load(bearded_model_data, store = here::here("_targets"))
+```
+
+The model specification below was used to specify an `mgcv::bam()` model that
+matches the formula used in the manuscript for spotted seals. The `s(speno, bs = "re")` 
+term is the smooth term for the random effect. All other predictors are
+the same.
+
+```{r bam-m1-spotted, include=TRUE, echo = TRUE}
+m1_spotted <- mgcv::bam(
+  dry ~ age_sex + s(speno, bs = "re") + 
+    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
+    poly(day, 3, raw=TRUE) + 
+    sin1:poly(day, 3, raw=TRUE) +
+    cos1:poly(day, 3, raw=TRUE) +
+    sin2:poly(day, 3, raw=TRUE) +
+    cos2:poly(day, 3, raw=TRUE) +
+    sin3:poly(day, 3, raw=TRUE) +
+    cos3:poly(day, 3, raw=TRUE) +
+    wind*temp2m + pressure + precip + 
+    age_sex:poly(day, 4, raw=TRUE),
+  data = spotted_model_data,
+  family = binomial,
+  discrete = TRUE)
+```
+
+```{r lag1-spotted, include=FALSE}
+lag1_spotted <- acf(resid(m1_spotted), plot=FALSE)[1][[1]]
+```
+
+Note, the specification for _m1_spotted_ here does not include
+any AR1 structure for temporal autocorrelation. To include this, we need to
+provide a value for $\rho$ (or _rho_). We can examine the autocorrelation within
+the model and use the lag-1 value for $\rho$ .The value for lag-1 
+autocorrelation is `r formatC(as.numeric(lag1_spotted))` which is
+rather high but not surprising. We can, now, update our model specification with
+a value for $\rho$ as well as the `A1.start` argument with specifies (TRUE/FALSE) the
+start point of each block.
+
+```{r bam-m2-spotted, include=TRUE, echo = TRUE}
+m2_spotted <- mgcv::bam(
+  dry ~ age_sex + s(speno, bs = "re") + 
+    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
+    poly(day, 3, raw=TRUE) + 
+    sin1:poly(day, 3, raw=TRUE) +
+    cos1:poly(day, 3, raw=TRUE) +
+    sin2:poly(day, 3, raw=TRUE) +
+    cos2:poly(day, 3, raw=TRUE) +
+    sin3:poly(day, 3, raw=TRUE) +
+    cos3:poly(day, 3, raw=TRUE) +
+    wind*temp2m + pressure + precip + 
+    age_sex:poly(day, 3, raw=TRUE),
+  data = spotted_model_data,
+  family = binomial,
+  AR.start = ar1_start,
+  rho = lag1_spotted,
+  discrete = TRUE)
+```
+
+The model specification for exploring the use of solar radiation is
+specified similarly but without all of the Fourier series parameters and
+interactions.
+
+```{r bam-ssrd-spotted, include=TRUE, echo = TRUE}
+m2_ssrd_spotted <- mgcv::bam(
+  dry ~ age_sex + s(speno, bs = "re") + 
+    era5_ssrd +
+    poly(day, 3, raw=TRUE) + 
+    era5_ssrd:poly(day, 3, raw=TRUE) +
+    wind*temp2m + pressure + precip + 
+    age_sex:poly(day, 3, raw=TRUE),
+  data = spotted_model_data,
+  family = binomial,
+  AR.start = ar1_start,
+  rho = lag1_spotted,
+  discrete = TRUE)
+```
+
+Lastly, the two models were compared with AIC to evaluate whether the
+reduction in degrees of freedom with fewer terms in the solar radiation
+model was matched with improved explanatory power in the model fit. While the
+model and code specified above is for spotted seals, the same approach was
+repeated for bearded and ribbon seals.
+
+```{r,bam-ssrd-ribbon-bearded}
+m1_ribbon <- mgcv::bam(
+  dry ~ age_sex + s(speno, bs = "re") + 
+    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
+    poly(day, 3, raw=TRUE) + 
+    sin1:poly(day, 3, raw=TRUE) +
+    cos1:poly(day, 3, raw=TRUE) +
+    sin2:poly(day, 3, raw=TRUE) +
+    cos2:poly(day, 3, raw=TRUE) +
+    sin3:poly(day, 3, raw=TRUE) +
+    cos3:poly(day, 3, raw=TRUE) +
+    wind*temp2m + pressure + precip + 
+    age_sex:poly(day, 4, raw=TRUE),
+  data = ribbon_model_data,
+  family = binomial,
+  discrete = TRUE)
+
+lag1_ribbon <- acf(resid(m1_ribbon), plot=FALSE)[1][[1]]
+
+m2_ribbon <- mgcv::bam(
+  dry ~ age_sex + s(speno, bs = "re") + 
+    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
+    poly(day, 3, raw=TRUE) + 
+    sin1:poly(day, 3, raw=TRUE) +
+    cos1:poly(day, 3, raw=TRUE) +
+    sin2:poly(day, 3, raw=TRUE) +
+    cos2:poly(day, 3, raw=TRUE) +
+    sin3:poly(day, 3, raw=TRUE) +
+    cos3:poly(day, 3, raw=TRUE) +
+    wind*temp2m + pressure + precip + 
+    age_sex:poly(day, 3, raw=TRUE),
+  data = ribbon_model_data,
+  family = binomial,
+  AR.start = ar1_start,
+  rho = lag1_ribbon,
+  discrete = TRUE)
+
+m2_ssrd_ribbon <- mgcv::bam(
+  dry ~ age_sex + s(speno, bs = "re") + 
+    era5_ssrd +
+    poly(day, 3, raw=TRUE) + 
+    era5_ssrd:poly(day, 3, raw=TRUE) +
+    wind*temp2m + pressure + precip + 
+    age_sex:poly(day, 3, raw=TRUE),
+  data = ribbon_model_data,
+  family = binomial,
+  AR.start = ar1_start,
+  rho = lag1_ribbon,
+  discrete = TRUE)
+
+m1_bearded <- mgcv::bam(
+  dry ~ s(speno, bs = "re") + 
+    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
+    poly(day, 3, raw=TRUE) + 
+    sin1:poly(day, 3, raw=TRUE) +
+    cos1:poly(day, 3, raw=TRUE) +
+    sin2:poly(day, 3, raw=TRUE) +
+    cos2:poly(day, 3, raw=TRUE) +
+    sin3:poly(day, 3, raw=TRUE) +
+    cos3:poly(day, 3, raw=TRUE) +
+    wind*temp2m + pressure + precip,
+  data = bearded_model_data,
+  family = binomial,
+  discrete = TRUE)
+
+lag1_bearded <- acf(resid(m1_bearded), plot=FALSE)[1][[1]]
+
+m2_bearded <- mgcv::bam(
+  dry ~ s(speno, bs = "re") + 
+    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
+    poly(day, 3, raw=TRUE) + 
+    sin1:poly(day, 3, raw=TRUE) +
+    cos1:poly(day, 3, raw=TRUE) +
+    sin2:poly(day, 3, raw=TRUE) +
+    cos2:poly(day, 3, raw=TRUE) +
+    sin3:poly(day, 3, raw=TRUE) +
+    cos3:poly(day, 3, raw=TRUE) +
+    wind*temp2m + pressure + precip,
+  data = ribbon_model_data,
+  family = binomial,
+  AR.start = ar1_start,
+  rho = lag1_bearded,
+  discrete = TRUE)
+
+m2_ssrd_bearded <- mgcv::bam(
+  dry ~ s(speno, bs = "re") + 
+    era5_ssrd +
+    poly(day, 3, raw=TRUE) + 
+    era5_ssrd:poly(day, 3, raw=TRUE) +
+    wind*temp2m + pressure + precip,
+  data = bearded_model_data,
+  family = binomial,
+  AR.start = ar1_start,
+  rho = lag1_bearded,
+  discrete = TRUE)
+```
+
+## Results
+
+To evaluate whether the solar radiation parameter matches our expectations and
+compares well with hour of the day, we can visualize the variability of the 
+`era5_ssrd` values within
+our study area as they relate to hour of the day. The unimodal distribution
+is centered around the middle of the solar day with peak solar radiation
+coinciding with 13:00 local solar. This suggests solar radiation could be an
+informative covariate for capturing unimodal diel patterns in haul-out
+behavior.
+
+```{r, include=FALSE}
+ssrd_dataset <- readRDS(file = here::here('data/ssrd_dataset.rds'))
+
+pt_ul = st_point(c(160, 70))
+pt_br = st_point(c(179, 60))
+poly_left <- st_bbox(st_sfc(pt_ul, pt_br,crs = 4326)) |> st_as_sfc()
+
+pt_ul = st_point(c(-160, 70))
+pt_br = st_point(c(-179, 60))
+poly_right <- st_bbox(st_sfc(pt_ul, pt_br,crs = 4326)) |> st_as_sfc()
+
+# how many random points do we want to generate
+n_pts <- 5000
+
+pts1 <- st_sample(poly_left,size = n_pts/2)
+pts2 <- st_sample(poly_right, size = n_pts/2)
+
+pts <- c(pts1,pts2)
+
+# Set the seed for reproducibility
+set.seed(123)
+
+# Generate random hourly timestamps
+get_random_times <- function(start_t, end_t, months, n) {
+  tibble(
+  time = seq(as.POSIXct(start_t, tz = "UTC"), as.POSIXct(end_t, tz = "UTC"), by = "hours")
+)  |> 
+  filter(month(time) %in% c(months))  |> 
+    filter(month(time) != 2 & mday(time) != 1) |> 
+  sample_n(n) %>%
+  pull(time)
+}
+
+random_times <- get_random_times(start_t = "2005-02-01", end_t = "2021-07-30",
+                                 months = c(2,3,4,5,6,7), n = n_pts)
+
+# create a dataset of points with those random times
+pts_sf <- st_sf(time=random_times,pts)
+pts_vec <- terra::vect(pts_sf)
+
+# list to store for-loop results
+pts_ssrd_list <- vector(mode = "list", length = length(years))
+
+years <- 2005:2021
+
+for (i in 1:length(years)) {
+  r <- terra::subset(ssrd_dataset, lubridate::year(time(ssrd_dataset)) == years[i])
+  v <- terra::subset(pts_vec, lubridate::year(pts_vec$time) == years[i])
+  tm <- match(v$time,time(r))
+  pts_ssrd_list[[i]] <- terra::extract(r,v,layer=tm, bind=TRUE)
+}
+
+library(solaR)
+
+pts_ssrd_sf <- vect(pts_ssrd_list) |> st_as_sf()
+
+pts_ssrd_sf <- pts_ssrd_sf |> 
+  mutate(lon = st_coordinates(geometry)[,1]) |> 
+  mutate(solar_hour = lubridate::hour(solaR::local2Solar(time,lon)))
+```
+
+```{r era5-ssrd-plot, fig.asp = 0.7, include=TRUE, fig.cap="Downward surface solar radiation estimates from the ERA5 climate reanalysis for 5000 random points within the study area between 2005 and 2021.
+Solar radiation values are presented in Watts per meter^2 and the smoothed line highlights the strong diel pattern."}
+ssrd_solarhour_plot <- ggplot(data = pts_ssrd_sf, aes(x=solar_hour, y=value)) + 
+  geom_point(size = 0.5, alpha = 0.2, color = "#edd9a3") + 
+  geom_smooth(se=TRUE, linewidth = 0.2, color = "#4b2991") +
+  # geom_vline(xintercept = 13, linetype="dashed", 
+  #               linewidth = 1) +
+  scale_x_continuous(breaks = c(3,6,9,12,15,18,21),
+                     labels = c("03:00","06:00","09:00","12:00","15:00","18:00",
+                                "21:00")) +
+  xlab("Local Solar Hour") +
+  ylab("Solar Radiation") +
+  labs(title = "Diel Pattern of Solar Radiation Values from ERA5 Reanalysis",
+       subtitle = "A random sample of hourly ERA5 downward surface solar radiation values for random points within the study area confirms the expected unimodal relationship with local solar hour.",
+       caption = "solar radiation values are presented in Watts per meter^2")
+
+ssrd_solarhour_plot
+```
+
+The spotted seal model matching the specification from the manuscript resulted in 
+`r AIC(m2_spotted,m2_ssrd_spotted)['m2_spotted','df']` degrees of freedom and an AIC value of
+`r AIC(m2_spotted,m2_ssrd_spotted)['m2_spotted','AIC']`. The model with solar radiation resulted in 
+`r AIC(m2_spotted,m2_ssrd_spotted)['m2_ssrd_spotted','df']` degrees of freedom and an AIC value of
+`r AIC(m2_spotted,m2_ssrd_spotted)['m2_ssrd_spotted','AIC']`. Despite the additional terms, the
+model with the Fourier series representation of hour of day resulted in a 
+lower AIC value and was still preferred for spotted seals. The ribbon seal model matching the specification from the manuscript resulted in 
+`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ribbon','df']` degrees of freedom and an AIC value of
+`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ribbon','AIC']`. The model with solar radiation resulted in 
+`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ssrd_ribbon','df']` degrees of freedom and an AIC value of
+`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ssrd_ribbon','AIC']`. Despite the additional terms, the
+model with the Fourier series representation of hour of day resulted in a 
+lower AIC value and was still preferred for bearded seals. The bearded seal model matching the specification from the manuscript resulted in 
+`r AIC(m2_bearded,m2_ssrd_bearded)['m2_bearded','df']` degrees of freedom and an AIC value of
+`r AIC(m2_bearded,m2_ssrd_bearded)['m2_bearded','AIC']`. The model with solar radiation resulted in 
+`r AIC(m2_bearded,m2_ssrd_bearded)['m2_ssrd_bearded','df']` degrees of freedom and an AIC value of
+`r AIC(m2_bearded,m2_ssrd_bearded)['m2_ssrd_bearded','AIC']`. Despite the additional terms, the
+model with the Fourier series representation of hour of day resulted in a 
+lower AIC value and was still preferred for bearded seals.

### 2023-11-08 analysis/paper/Letter_Response_to_Reviewers.qmd
solar radiation supplemental + fits
major update to qmd and figures

@@
 
+J
 ```
-Josh M. London, PhD\newline
+osh M. London, PhD\newline
 7600 Sand Point Way NE\newline Seattle, Washington, 98115\newline josh.london\@noaa.gov\newline
@@
 **\[author response\]** The reviewer provided numerous comments and suggestions annotated throughout the PDF that address the concern of presentation and clarity.
-In the next section we provide responses to those annotations as best as possible
+In the next section we provide responses to those annotations
 :::
@@
 ::: author-response
-**\[author response\]** We don't feel comfortable making a precision comparison, simply because there are too many differences between the analyses, including (1) the data sets are different (the present study has substantially more records), and (2) the model structures are different (the present study has more parameters).
+**\[author response\]** We don't feel comfortable making a precision comparison with previous abundance estimates, simply because there are too many differences between the analyses, including (1) the data sets are different (the present study has substantially more records), and (2) the model structures are different (the present study has more parameters).
 As such, it will be difficult to ascribe a specific reason to precision increases or decreases.
@@
 **\[author response\]** The opening sentence of the abstract is a statement regarding the reliance of ice-associated seals on sea-ice for a range of important life history behaviors.
-We think this statement is important introduction for readers who are not familiar with ice-associated seals.
+We think this statement is an important introduction for readers who are not familiar with ice-associated seals.
 We do agree with the reviewer's suggestion for more direct language and have made edits to the beginning of the abstract with this in mind.
@@
 We provide this visually in Figure 3 and adding an additional layer to that figure to show deployments would be an additional complication.
-To ensure the audience has a clear understanding of our study area, we have added an additional figure with a map of the Bering and Chukchi sea region that indicates generally where deployments were initiated.
+To ensure the audience has a clear understanding of our study area, we have added an additional figure (now, Figure 1) with a map of the Bering and Chukchi sea region that indicates generally where deployments were initiated.
+In response to other reviewer comments this figure also indicates the spatial range of aerial survey effort to date.
 :::
@@
 As for providing the reader focus on why the Federal government is doing this study: this is a collaborative study across multiple organizations and the analysis/purpose is not limited to the Federal government.
-We've incorporated the reviewers suggests to expand discussion of the statutory requirements of the MMPA and ESA but don't feel we need to single out the motivations for the Federal government.
+We've incorporated the reviewers suggestions to expand discussion of the statutory requirements of the MMPA and ESA but don't feel we need to single out the motivations for the Federal government.
 
-We have also added a figure depicting the estimated species range for bearded, ribbon, and spotted seals along with general representation of recent aerial survey effort.
+We have also added additional general representation of recent aerial survey effort to the new Figure 1.
 This should better communicate to the reader how the spatial distribution of our haul-out behavior overlaps with the species range and survey effort.
@@
 **\[author response\]** We have re-organized this paragraph to, hopefully, better communicate our choice not to explore sea-ice related covariates.
-As the reviewer suggests, we have limited geolocation data for at-sea periods of our haul-out behavior data -- we know the seal was in the water but don't have any estimates of location when only a flipper tag is attached.
+As the reviewer suggests, we have limited geo-location data for at-sea periods of our haul-out behavior data -- we know the seal was in the water but don't have any estimates of location when only a flipper tag is attached.
 By leading with this statement, we hope that communicates our primary rationale.
@@
 
-We have also added a figure depicting the estimated species range for bearded, ribbon, and spotted seals along with general representation of recent aerial survey effort.
+We have also added a figure 1 depicting a spatial representation of recent aerial survey effort (which reflects estimated species range for bearded, ribbon, and spotted seals in US waters).
 This should better communicate to the reader how the spatial distribution of our haul-out behavior overlaps with the species range and survey effort.
@@
 We have taken the time to do some additional analysis and methodological exploration and provide additional details in Supplement 2.
-While promising, we feel the approach taken in the analysis is the most appropriate because it distinguishes daily bi-modal (e.g.
+Note, our initial exploration of this approach resulted in lower AIC values for the Fourier series approach presented in the analysis.
+While promising and worth even further exploration, we feel the approach taken in the analysis is the most appropriate because it distinguishes daily bi-modal (e.g.
 crepuscular) patterns that might be important but would not be detected by insolation alone (i.e.
@@
 **\[author response\]** The plot titles were developed with strong consideration for improving scientific communication over approaches that are more typically used in publications.
-They were developed with guidance from modern approaches to data visualization that recognize the plot title is often the very first thing the reader sees.
+They were developed with guidance from modern approaches to data visualization that recognize the plot title is often the very first thing the reader sees and that data visualizations are often viewed and interpreted as standalone figures.
 Therefore it can serve as an effective communication tool to help the reader interpret the results.
 
-That said, we recognize that the approach may depart too much from the standard expectations and have made adjustments to each of the titles with the feedback in mind.
-We hope the new titles provide a balanced approach.
+That said, we recognize that the initial approach could still be improved and have made adjustments to each of the titles with the feedback in mind. We hope the new titles provide a balanced approach.
 :::
@@
 ::: author-response
-**\[author response\]** The color ramp used is the 'Hiroshige' color ramp from the MetBrewer R package and has been tested and confirmed to be colorblind friendly.
-We initially used the viridis color palette for these plots but the Hiroshige colors provided additional visual detail we feel is important.
+**\[author response\]** The color ramp used has been updated to use palettes from the R package 'scico' have been tested and confirmed to be colorblind friendly and appropriately balanced for scientific interpretation.
+We initially used the viridis color palette for these plots but the 'scico' palette colors provided additional visual detail we feel is important.
 :::

### 2023-11-08 analysis/paper/London_HauloutBeahvior.Rmd
major update to qmd and figures

@@
   percent-dry data from satellite-linked bio-loggers deployed
-  between 2005 and 2021 to quantify the proportion of seals hauled out on ice,
-  which is needed to accurately estimate abundance from aerial survey counts of
+  between 2005 and 2021 to quantify the proportion of seals hauled out on ice.
+  This information is needed to accurately estimate abundance from aerial survey counts of
   ice-associated seals (i.e., to correct for the proportion of animals that
@@
   to explain variation in haul-out probability. We found evidence for
-  strong diurnal and within-season patterns in haul-out behavior, as well as
+  strong diel and within-season patterns in haul-out behavior, as well as
   strong weather effects (particularly wind and temperature). In general, seals
@@
 ice extent, volume, and seasonal presence [@kwok2018;
-@meier2014; @overland2021; @wang2017]. These changes have tangible effects on
+@meier2014; @overland2021; @wang2018]. These changes have tangible effects on
 Arctic organisms, ecosystems, and the human communities who live in the region
@@
 ice-associated seal species in the Arctic using aerial surveys (e.g.,
-[@bengtson2005a], [@conn2014a], and [@verhoef2014a]) and more recent efforts
+[@bengtson2005a], [@conn2013b], and [@verhoef2014a]) and more recent efforts
 have significantly expanded on previous survey effort. Such abundance studies
@@
 miss some seals on ice), species misclassification, and possible
-disturbance of seals by aircraft [@conn2014a; @verhoef2014a]. Refining these
+disturbance of seals by aircraft [@conn2013b; @verhoef2014a]. Refining these
 inferences will improve the accuracy of abundance estimates and, hopefully,
@@
 [-@verhoef2014a] found haul-out probabilities increased gradually starting in
-March and peaked in May and June for bearded, ribbon, and spotted seals. Such
+March and peaked in May and June for bearded, ribbon, and spotted seals. Recent
+Such
 analyses have not been limited to the Arctic. In the Antarctic, Bengtson and
@@
 (e.g., [@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
-[@verhoef2014a], and [@southwell2008]). However, sample sizes have often been
+[@verhoef2014a], [@southwell2008], and [@niemi2023]). However, sample sizes have often been
 insufficient to permit strong inference about demographic and/or seasonal
@@
 improve estimates of seal abundance. Previously estimated availability
-correction factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a])
+correction factors (e.g., [@bengtson2005a], [@conn2013b], and [@verhoef2014a])
 accounted for variables such as the time-of-day and day-of-year, but did not
@@
 atmospheric interference. Fortunately, missing records do not substantially bias
-inference about haul-out probabilities [@conn2012b].
+inference about haul-out probabilities [@conn2012c].
 
@@
 North American Regional Reanalysis (NARR) model produced by the National Centers
-for Environmental Prediction [@mesinger2006]. The NARR model assimilates
+for Environmental Prediction [@mesinger2006a]. The NARR model assimilates
 observational data to produce a long-term picture of weather over North America
@@
 during the spring, these species are most prevalent near the southern ice edge
-in the Bering Sea [@conn2014a].
+in the Bering Sea [@conn2013b].
 
@@
 shown for all ages (Figure \@ref(fig:beardedHOCal); see also 
-@\ref(fig:beardedPredSE)). Additionally, after
+\@ref(fig:beardedPredSE)). Additionally, after
 approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
@@
 When exploring the influence of weather, bearded seal haul-out probability
-appeared most influenced by wind
+was strongly affected by wind
 ($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$
@@
 and temperature
-($F_{`r paste(bearded_type3$temp2$Num.df, bearded_type3$temp2$Den.df, sep=",")`}$
-= `r bearded_type3$temp2$F.value`; $p$ =
-`r bearded_type3$temp2$Prob.F`) (Figure \@ref(fig:beardedHOwx)). 
-Additionally, wind chill was also a significant influence
-($F_{`r paste(bearded_type3$temp2_wind$Num.df, bearded_type3$temp2_wind$Den.df, sep=",")`}$
-= `r bearded_type3$temp2_wind$F.value`; $p$ =
-`r bearded_type3$temp2_wind$Prob.F`). Any influence of barometric
-pressure or precipitation was less apparent.
+($F_{`r paste(bearded_type3$temp2m$Num.df, bearded_type3$temp2m$Den.df, sep=",")`}$
+= `r bearded_type3$temp2m$F.value`; $p$ =
+`r bearded_type3$temp2m$Prob.F`) with much higher haul-out probability during 
+periods of higher temperatures and low wind speeds (Figure \@ref(fig:beardedHOwx)). 
+Not surprisingly, wind chill 
+($F_{`r paste(bearded_type3$temp2m_wind$Num.df, bearded_type3$temp2m_wind$Den.df, sep=",")`}$
+= `r bearded_type3$temp2m_wind$F.value`; $p$ =
+`r bearded_type3$temp2m_wind$Prob.F`) was also important. Barometric
+pressure ($F_{`r paste(bearded_type3$pressure$Num.df, bearded_type3$pressure$Den.df, sep=",")`}$
+= `r bearded_type3$pressure$F.value`; $p$ =
+`r bearded_type3$pressure$Prob.F`) was also significant factor although less 
+apparent (Figure \@ref(fig:beardedHOwx)). Any effect of precipitation 
+was not a significant influence on haul-out probability 
+($F_{`r paste(bearded_type3$precip$Num.df, bearded_type3$precip$Den.df, sep=",")`}$
+= `r bearded_type3$precip$F.value`; $p$ =
+`r bearded_type3$precip$Prob.F`).
 
@@
 
-The haul-out probability for ribbon seals was strongly influenced by
-each of the weather covariates included within the model except
-precipitation. The impacts of temperature
-($F_{`r paste(ribbon_type3$temp2$Num.df, ribbon_type3$temp2$Den.df, sep=",")`}$
-= `r ribbon_type3$temp2$F.value`; $p$ = `r ribbon_type3$temp2$Prob.F`),
-wind
+The haul-out probability for ribbon seals was mostly influenced by
+temperature
+($F_{`r paste(ribbon_type3$temp2m$Num.df, ribbon_type3$temp2m$Den.df, sep=",")`}$
+= `r ribbon_type3$temp2m$F.value`; $p$ = `r ribbon_type3$temp2m$Prob.F`) and wind
 ($F_{`r paste(ribbon_type3$wind$Num.df, ribbon_type3$wind$Den.df, sep=",")`}$
-= `r ribbon_type3$wind$F.value`; $p$ = `r ribbon_type3$wind$Prob.F`),
-and barometric pressure
+= `r ribbon_type3$wind$F.value`; $p$ = `r ribbon_type3$wind$Prob.F`) with barometric pressure
 ($F_{`r paste(ribbon_type3$pressure$Num.df, ribbon_type3$pressure$Den.df, sep=",")`}$
 = `r ribbon_type3$pressure$F.value`; $p$ =
-`r ribbon_type3$pressure$Prob.F`) were especially noticeable as ribbon
-seals were less likely to haul out at higher winds and lower pressure
+`r ribbon_type3$pressure$Prob.F`) having a milder impact.
+Ribbon seals were less likely to haul out at higher winds and lower pressure
 values and more likely to haul out when temperatures were
-relatively warm. Wind chill
-($F_{`r paste(ribbon_type3$temp2_wind$Num.df, ribbon_type3$temp2_wind$Den.df, sep=",")`}$
-= `r ribbon_type3$temp2_wind$F.value`; $p$ =
-`r ribbon_type3$temp2_wind$Prob.F`) showed a moderately negative
-influence on haul-out probability. As with bearded seals, Figure
-\@ref(fig:ribbonHOwx) presents the predicted haul-out probability of
-ribbon seals across the range of weather conditions encountered in the
-observed data. Because our ribbon seal model included age and sex class, we can
-visualize the different influences of weather covariates on those classes.
+relatively warm (Figure \@ref(fig:ribbonHOwx)). Neither wind chill
+($F_{`r paste(ribbon_type3$temp2m_wind$Num.df, ribbon_type3$temp2m_wind$Den.df, sep=",")`}$
+= `r ribbon_type3$temp2m_wind$F.value`; $p$ =
+`r ribbon_type3$temp2m_wind$Prob.F`) nor precipitation 
+($F_{`r paste(ribbon_type3$precip$Num.df, ribbon_type3$precip$Den.df, sep=",")`}$
+= `r ribbon_type3$precip$F.value`; $p$ =
+`r ribbon_type3$precip$Prob.F`) were a significant
+influence on haul-out probability. Compared with bearded seals, the effect of
+weather covariates on the predicted haul-out probability for
+ribbon seals was less striking. Because our ribbon seal model included age and sex class, we can
+visualize the different influences of weather covariates on those classes and see
+that subadults separate out from adult males and females (Figure \@ref(fig:ribbonHOwx)).
 
@@
 that was less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal);
-see also @\ref(fig:spottedPredSE)). Adults of both sexes spent considerable time
+see also \@ref(fig:spottedPredSE)). Adults of both sexes spent considerable time
 in April, May and June hauled out. In contrast to ribbon seals, adult spotted
@@
 
-Spotted seal haul-out behavior was most strongly influenced by
-temperature
-($F_{`r paste(spotted_type3$temp2$Num.df, spotted_type3$temp2$Den.df, sep=",")`}$
-= `r spotted_type3$temp2$F.value`; $p$ = `r spotted_type3$temp2$Prob.F`)
-and wind
+Spotted seal haul-out behavior was less affected by the weather covariates 
+compared to ribbon and bearded seals but their influence on the model was still
+significant in some cases.
+Temperature
+($F_{`r paste(spotted_type3$temp2m$Num.df, spotted_type3$temp2m$Den.df, sep=",")`}$
+= `r spotted_type3$temp2m$F.value`; $p$ = `r spotted_type3$temp2m$Prob.F`), wind
 ($F_{`r paste(spotted_type3$wind$Num.df, spotted_type3$wind$Den.df, sep=",")`}$
-= `r spotted_type3$wind$F.value`; $p$ = `r spotted_type3$wind$Prob.F`)
-with barometric pressure having a moderate influence
+= `r spotted_type3$wind$F.value`; $p$ = `r spotted_type3$wind$Prob.F`),
+and barometric pressure
 ($F_{`r paste(spotted_type3$pressure$Num.df, spotted_type3$pressure$Den.df, sep=",")`}$
 = `r spotted_type3$pressure$F.value`; $p$ =
-`r spotted_type3$pressure$Prob.F`). Spotted seals were less likely to
+`r spotted_type3$pressure$Prob.F`) were all significant. Spotted seals were less likely to
 haul out at higher winds and more likely to be on the ice when
-temperatures were relatively warm. Wind chill (*temperature:wind*) and
-precipitation were not as influential as the other covariates.
+temperatures were relatively warm. Wind chill 
+($F_{`r paste(spotted_type3$temp2m_wind$Num.df, spotted_type3$temp2m_wind$Den.df, sep=",")`}$
+= `r spotted_type3$temp2m_wind$F.value`; $p$ =
+`r spotted_type3$temp2m_wind$Prob.F`) and
+precipitation 
+($F_{`r paste(spotted_type3$precip$Num.df, spotted_type3$precip$Den.df, sep=",")`}$
+= `r spotted_type3$precip$F.value`; $p$ =
+`r spotted_type3$precip$Prob.F`)
+were not as influential as the other covariates.
 Differences in the magnitude of response between the age-sex classes were
@@
 \@ref(fig:spottedHOwx)). There was a consistent ranking of adult males being the most
-likely to haul out, followed by adult females, and, then, subadults.
+likely to haul out, followed by adult females, and, then, subadults. This differs
+from ribbon seals which showed more overlap between adult males and adult females
+and that subadults were most likely to haul out across the presented range of weather
+covariates.
 
@@
 Unlike previous analyses of seal haul-out behavior in spotted and ribbon seals
-(e.g. [@verhoef2010a], [@conn2014a]), we also investigated the influence of
+(e.g. [@verhoef2010a], [@conn2013b]), we also investigated the influence of
 sex-age class on haul-out probabilities for both species (not for bearded seals
@@
 photoperiod, ribbon and spotted seals are dependent upon the presence of sea ice
-for pupping and molt. We did not find any support in our models for a
+for pupping and molt. We did not find large support in our models for a
 relationship between the timing of peaks in haul-out behavior and the amount of
@@
 2018 – 2019, years of extreme low spring ice extent in the Bering Sea that
-appeared to impact body condition of ribbon and spotted seals [@boveng2020a], so we may
+appeared to impact body condition of ribbon and spotted seals [@boveng2020], so we may
 currently lack sufficient contrast in ice extent to characterize an effect on
@@
 Previous attempts to estimate the abundance of phocid seals from aerial survey
-data in the Bering and Chukchi seas (e.g. [@bengtson2005a], [@conn2014a],
+data in the Bering and Chukchi seas (e.g. [@bengtson2005a], [@conn2013b],
 [@verhoef2014a]) have used estimated haul-out probabilities to correct for the
@@
 bearded seals. For instance, Ver Hoef et al. [-@verhoef2014a] and Conn et al.
-[-@conn2014a] used haul-out correction factors with maximums of 0.66 for bearded
-seals, 0.62 for ribbon seals, and 0.54 for spotted seals, where maximums
-corresponded to times near solar noon in mid-late May. Applying models that
-ignore age, sex, and year effects, these probabilities were 0.38, 0.72, and
-0.60, respectively, under the current analysis framework. Our current estimates
+[-@conn2013b] used haul-out correction factors with maximums of 0.66 for bearded
+seals, 0.62 for ribbon seals, and 0.54 for spotted seals, where these maximums
+corresponded to times near solar noon in mid-late May. Our current estimates
 of haul-out probability reflect increased sample sizes in terms of
-number of individuals, but also improvements to the way data were prepared prior to
-analysis.
+number of individuals, inclusion of weather covariates, and improvements to the 
+way data were prepared prior to analysis.
 
@@
 predictor of the daily haul-out cycle in seals compared to time of day. The
-notion being that seals are directly responding to changes in solar radiation
+notion being that seals are, likely, directly responding to changes in solar radiation
 throughout the day and not what time of day it is (i.e. seals don't have
@@
 
-Because of these reasons, we spent some time considering and exploring this
+Because of these reasons, we considered and explored this
 possibility for our model and the analysis presented in this manuscript. A key
-drawback to reliance on solar insolation, in our minds, is that we would lose
-insight regarding potential crepuscular patterns -- insolation does not 
+drawback to reliance on solar radiation, in our minds, is that we would lose
+insight regarding potential diel patterns -- solar radiation does not 
 differentiate between dusk or dawn. Bi-modal patterns have been
@@
   dry ~ age_sex + s(speno, bs = "re") + 
-    era5_ssrd +
+    era5_ssrd_watts +
     poly(day, 3, raw=TRUE) + 
-    era5_ssrd:poly(day, 3, raw=TRUE) +
+    era5_ssrd_watts:poly(day, 3, raw=TRUE) +
     wind*temp2m + pressure + precip + 
@@
   dry ~ s(speno, bs = "re") + 
-    era5_ssrd +
+    era5_ssrd_watts +
     poly(day, 3, raw=TRUE) + 
-    era5_ssrd:poly(day, 3, raw=TRUE) +
+    era5_ssrd_watts:poly(day, 3, raw=TRUE) +
     wind*temp2m + pressure + precip,
@@
 
-```{r, include=FALSE}
-ssrd_dataset <- readRDS(file = here::here('data/ssrd_dataset.rds'))
+```{r random-ssrd-pts, include=FALSE, eval=FALSE}
+years <- 2005:2021
+
+raster_merge <- vector(mode = "list", length = length(years))
+
+for (i in 1:length(years)) {
+  left_file <-
+    here::here('data_raw',
+               'era5_ssrd',
+               paste0('era5_ssrd_', years[i], '_left.nc'))
+  right_file <-
+    here::here('data_raw',
+               'era5_ssrd',
+               paste0('era5_ssrd_', years[i], '_right.nc'))
+  r1 <- terra::rast(left_file)
+  r2 <- terra::rast(right_file)
+  raster_merge[[i]] <- terra::merge(r1, r2)
+}
+
+ssrd_dataset <- terra::rast(raster_merge)
 
@@
 
-```{r era5-ssrd-plot, fig.asp = 0.7, include=TRUE, fig.cap="Downward surface solar radiation estimates from the ERA5 climate reanalysis for 5000 random points within the study area between 2005 and 2021.
-Solar radiation values are presented in Watts per meter^2 and the smoothed line highlights the strong diel pattern."}
+```{r era5-ssrd-plot, fig.asp = 0.75, include=TRUE, fig.cap="Downward surface solar radiation estimates from the ERA5 climate reanalysis for 5000 random points within the study area between 2005 and 2021.
+Solar radiation values are presented in Watts per square-meter and the smoothed line highlights the strong diel pattern."}
+
+pts_ssrd_sf <- readRDS(here::here('data/random_ssrd_pts.rds'))
+
 ssrd_solarhour_plot <- ggplot(data = pts_ssrd_sf, aes(x=solar_hour, y=value)) + 
@@
   labs(title = "Diel Pattern of Solar Radiation Values from ERA5 Reanalysis",
-       subtitle = "A random sample of hourly ERA5 downward surface solar radiation values for random points within the study area confirms the expected unimodal relationship with local solar hour.",
+       subtitle = "A random sample of hourly ERA5 downward surface solar radiation values for points within the study area confirms a unimodal relationship with local solar hour.",
        caption = "solar radiation values are presented in Watts per meter^2")

### 2023-11-08 scripts/deploy_plot.R

@@ -1,0 +1,61 @@
+library(googlesheets4)
+library(ggplot2)
+library(rnaturalearth)
+library(ggspatial)
+library(dplyr)
+library(dbplyr)
+library(odbc)
+
+deploy_tbl <- read_sheet("1MpCiyAKWLBXufwnzW8QTnb2SsTkrKSuPlWdpWMnqoto")
+
+tryCatch({
+  con <- dbConnect(RPostgres::Postgres(),
+                    dbname = 'pep', 
+                    host = Sys.getenv('PEP_PG_IP'),
+                    user = keyringr::get_kc_account("pgpep_londonj"),
+                    password = keyringr::decrypt_kc_pw("pgpep_londonj"))
+},
+error = function(cond) {
+  print("Unable to connect to Database.")
+})
+on.exit(dbDisconnect(con))
+
+deploy_db <- tbl(con, in_schema("telem","tbl_tag_deployments")) %>% 
+  dplyr::select(speno, deploy_lat, deploy_long, capture_lat, capture_long) %>% 
+  dplyr::mutate(deploy_lat = if_else(is.na(deploy_lat), capture_lat, deploy_lat),
+                deploy_long = if_else(is.na(deploy_long), capture_long, deploy_long)) %>% 
+  collect()
+
+dbDisconnect(con)
+
+deploy_tbl <- deploy_tbl %>% 
+  # dplyr::select(speno, species, location, region, latitude, longitude) %>% 
+  left_join(deploy_db, by = "speno") %>% 
+  mutate(latitude = coalesce(latitude,deploy_lat),
+         longitude = coalesce(longitude, deploy_long))
+
+deploy_sf <- deploy_tbl %>% 
+  sf::st_as_sf(coords = c('longitude','latitude'), crs = 4326) %>% 
+  st_transform(3571) %>% 
+  group_by(species,location, region) %>% 
+  group_modify(~concaveman(., concavity = 4),.keep=TRUE) %>% 
+  st_sf(crs = 3571) %>% 
+  st_buffer(40*1000) 
+  
+
+
+world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
+  sf::st_transform(st_crs(deploy_sf))
+
+ggplot() +
+  layer_spatial(deploy_sf, aes(fill=species),
+                linewidth=0) +
+  annotation_spatial(world,
+                     linewidth=0,
+                     fill = "grey70") +
+  facet_grid(cols = vars(species)) +
[…]

### 2023-11-08 scripts/ssrd_pred_heathmap_plots.R

@@ -1,0 +1,55 @@
+r <- ribbon_newdata |> 
+  group_by(age_sex,day,yday) |> 
+  summarize(across(c(northing,temp2m,wind,pressure,precip), 
+                   ~ mean(.x, na.rm = TRUE)))
+
+ssrd_breaks <- ribbon_model_data$era5_ssrd_watts |> 
+  quantile()
+
+r <- r |> expand_grid(tibble(era5_ssrd_watts = ssrd_breaks,
+       ssrd_quants = factor(names(ssrd_breaks),
+                            levels = c("0%","25%","50%","75%","100%"))))
+
+spenos <- unique(ribbon_model_data$speno) %>% 
+  as.character()
+
+r <- r %>% 
+  mutate(speno = sample(spenos,1))
+
+ribbon_ssrd_predict <- bind_cols(r,predict(m2_ssrd_ribbon,
+                               newdata = r,
+                               type = "response",
+                               se.fit = TRUE,
+                               exclude = "s(speno)")
+)
+
+plot_df <- ribbon_ssrd_predict %>%
+  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+         month = lubridate::month(date,label=TRUE),
+         day = lubridate::day(date)) %>%
+  filter(!month %in% c("Jul","Aug"))
+
+p2 <- ggplot(plot_df, aes(day, ssrd_quants, fill = fit)) +
+  geom_tile(color = "white", linewidth = 0) +
+  scale_fill_viridis_c(
+    name = "haul-out probability",
+    aesthetics = "fill",
+    limits = c(0, 1),
+    breaks = c(0.25, 0.50, 0.75),
+    guide = guide_colorbar(
+      title.position = "bottom",
+      barwidth = 15,
+      barheight = 0.5,
+      title.hjust = 0.5
+    )
+  )
+
+p2 <- p2 + facet_grid(age_sex~month, scales = "free_x")
+p2 <- p2 + scale_x_continuous(breaks = c(5,10,15,20,25)) +
+  coord_cartesian(expand=FALSE)
+p2 <- p2 + theme_minimal() +
+  theme(panel.spacing = unit(0.1, "lines")) +
+  theme(legend.position = "bottom") +
+  theme(strip.background = element_rect(colour="white")) +
+  theme(axis.ticks=element_blank()) +
+  xlab("day of month") + ylab("ssrd quantiles")

### 2024-01-29 analysis/paper/London_HauloutBeahvior.Rmd
update SMK name
citation fixes and updates

@@
     affiliation: 1
-  - name: Stacie K. Hardy
+  - name: Stacie M. Koslovsky
     affiliation: 1
@@
 ice-associated seal species in the Arctic using aerial surveys (e.g.,
-[@bengtson2005a], [@conn2013b], and [@verhoef2014a]) and more recent efforts
+@bengtson2005a, @conn2014a, and @verhoef2014a) and more recent efforts
 have significantly expanded on previous survey effort. Such abundance studies
@@
 miss some seals on ice), species misclassification, and possible
-disturbance of seals by aircraft [@conn2013b; @verhoef2014a]. Refining these
+disturbance of seals by aircraft [@conn2014a; @verhoef2014a]. Refining these
 inferences will improve the accuracy of abundance estimates and, hopefully,
@@
 such as time-of-day, day-of-year, sex, age class, and environmental conditions
-(e.g., [@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
-[@verhoef2014a], [@southwell2008], and [@niemi2023]). However, sample sizes have often been
+(e.g., @reder2003, @bengtson2004, @bengtson2005a, @udevitz2009, @lonergan2013,
+@verhoef2014a, @southwell2008, and @niemi2023). However, sample sizes have often been
 insufficient to permit strong inference about demographic and/or seasonal
@@
 improve estimates of seal abundance. Previously estimated availability
-correction factors (e.g., [@bengtson2005a], [@conn2013b], and [@verhoef2014a])
+correction factors (e.g., @bengtson2005a, @conn2014a, and @verhoef2014a)
 accounted for variables such as the time-of-day and day-of-year, but did not
@@
 Previous attempts to estimate the abundance of phocid seals from aerial survey
-data in the Bering and Chukchi seas (e.g. [@bengtson2005a], [@conn2013b],
-[@verhoef2014a]) have used estimated haul-out probabilities to correct for the
+data in the Bering and Chukchi seas (e.g. @bengtson2005a, @conn2014a,
 proportion of animals that are in the water and thus unavailable to be counted.
@@
 bearded seals. For instance, Ver Hoef et al. [-@verhoef2014a] and Conn et al.
-[-@conn2013b] used haul-out correction factors with maximums of 0.66 for bearded
+[-@conn2014a] used haul-out correction factors with maximums of 0.66 for bearded
 seals, 0.62 for ribbon seals, and 0.54 for spotted seals, where these maximums

### 2024-01-29 analysis/paper/London_HauloutBeahvior.Rmd
updates to Intro text

@@
 accounted for variables such as the time-of-day and day-of-year, but did not
-investigate meteorological variables that have been shown to influence haul-out
-behavior of walruses [@udevitz2009]. Third, we aimed to assess the annual
-variability in haul-out timing and possible linkage to changes in the extent of
-seasonal sea ice between 2005 and 2020. Our work extends the scope of previous
-haul-out analyses, includes the influence of meteorological variability, and
-investigates the potential impact of changing sea-ice extent on the behavior of
-these species.
+investigate the impact of meteorological variables. Such variables have been
+shown to influence walrus haul-out behavior [@udevitz2009] and we expect weather
+conditions to also influence seal haul-out behavior and including them within
+the model framework will benefit our estimates of seal availability during
+aerial surveys. Third, we aimed to assess the annual variability in haul-out
+timing and possible linkage to changes in the extent of seasonal sea ice between
+2005 and 2020. Our work extends the scope of previous haul-out analyses,
+includes the influence of meteorological variability, and investigates the
+potential impact of changing sea-ice extent on the behavior of these species.
 

### 2024-01-29 analysis/paper/London_HauloutBeahvior.Rmd
improve the deployment location map/figure
includes removing the figure title and increasing the ocean basin label font size

@@
 
-(ref:deployMapCaption) Initial deployment areas for each seal species (bearded, ribbon, and spotted seals) in the study between 2005 and 2020.
-Solid regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility.
-The larger, shaded region indicates the spatial extent of aerial survey effort to date in US waters of the Bering and Chukchi seas.
+(ref:deployMapCaption) **Initial bio-logger deployment areas for each seal species (bearded, ribbon, and spotted seals) in the study between 2005 and 2020 across multiple regions within the Bering, Chukchi, and western Beaufort seas.** \linebreak Solid regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility.
+The larger, shaded region indicates the spatial extent of aerial survey effort to date in the Bering, Chukchi, and Beaufort seas.
 Deployments were initiated across a range of months but only data from 1 March to 15 July were included in the analysis.
@@
 
-```{r deployplot, include=TRUE, fig.asp = 0.45, fig.cap = "(ref:deployMapCaption)"}
+```{r deployplot, include=TRUE, fig.asp = 0.4, fig.cap = "(ref:deployMapCaption)"}
 targets::tar_load(deploy_details, store = here::here("_targets"))
@@
                                             species_colors['Bearded seal'],
-                                            ";'>Bearded seal</span>"),
+                                            ";'>Bearded seal<br />Deployments</span>"),
                     "Ribbon seal" = paste0("<span style = 'color:",
                                            species_colors['Ribbon seal'],
-                                           ";'>Ribbon seal</span>"),
+                                           ";'>Ribbon seal<br />Deployments</span>"),
                     "Spotted seal" = paste0("<span style = 'color:",
                                             species_colors['Spotted seal'],
-                                            ";'>Spotted seal</span>")
+                                            ";'>Spotted seal<br />Deployments</span>")
 ) %>% as_labeller()
@@
                      fill = "grey80") +
-  geom_sf_text(data = label_coords_bering, aes(label = "Bering sea"), size = 1.65, color = "grey30") +
+  geom_sf_text(data = label_coords_bering, aes(label = "Bering sea"), size = 2, color = "grey30") +
   geom_sf_text(data = label_coords_chukchi, 
-               aes(label = "Chuckchi \nsea"), size = 1.65, lineheight = .85, color = "grey30") +
+               aes(label = "Chuckchi \nsea"), size = 2, lineheight = .85, color = "grey30") +
   geom_sf_text(data = label_coords_beaufort, 
-               aes(label = "Beaufort \nsea"), size = 1.65, lineheight = .85, color = "grey30") +
+               aes(label = "Beaufort \nsea"), size = 2, lineheight = .85, color = "grey30") +
   facet_grid(cols = vars(species), labeller = species_labels,
@@
                     name = "species") + 
-  labs(title = "Initial bio-logger deployment locations by species",
-          subtitle = 
-                 paste0("Bio-loggers were initially deployed on ",
-                "<span style = 'color:",species_colors['Bearded seal'],";'>bearded,</span> ", 
-                "<span style = 'color:",species_colors['Ribbon seal'],";'>ribbon,</span> ",  
-                "and ", 
-                "<span style = 'color:",species_colors['Spotted seal'],";'>spotted</span> seals ", 
-                "across multiple regions within the Bering, Chukchi, and western Beaufort seas."),
-       caption = "<span style = 'color: #A26E7C;'><span style = 'font-family: \"Font Awesome 6 Free Regular\";'>&#xf0c8;</span> spatial extent of aerial survey effort in US waters</span>") +
+  labs(caption = "<span style = 'color: #A26E7C;'><span style = 'font-family: \"Font Awesome 6 Free Regular\";'>&#xf0c8;</span> spatial extent of aerial survey effort in US waters</span>") +
   theme(legend.position = "none",

### 2024-01-29 analysis/paper/London_HauloutBeahvior.Rmd
improvements to methods text

@@
 included. We refer readers to Figure \@ref(fig:deployplot) and the primary literature
-for detailed capture and bio-logger attachment methods (see Supplemental
+for detailed capture and bio-logger attachment methods (see also Supplemental
 Material, [S1](#s1)).
@@
 'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
-Washington, USA) and we either adhered them to the hair on the seal or attached
-them through the rear flipper inter-digital webbing. The use of bio-loggers
-adhered to the back or head provides some benefits over flipper mounted devices
-(e.g. increased satellite transmittal rates, locations at sea) but these are
-lost during the following annual molt, which, depending on deployment date,
-limits the duration of haul-out data they provide. Additionally, bio-loggers
-attached to the head or dorsal region are often dry while the seal is floating
-at the surface, inducing a slight positive bias in the hourly percent-dry values
-reported by the bio-logger. For this study, in cases where both bio-logger types
-were deployed, we preferred hourly percent-dry observations from the flipper tag.
+Washington, USA). Deployments consisted of either a single 'SPLASH' device, a
+single 'SPOT' device, or both types. Devices were either adhered to the hair
+on the seal or attached through the rear flipper inter-digital webbing. The use
+of bio-loggers adhered to the back or head provides some benefits over flipper
+mounted devices (e.g. increased satellite transmittal rates, locations at sea)
+but these are lost during the following annual molt, which, depending on
+deployment date, limits the duration of haul-out data they provide.
+Additionally, bio-loggers attached to the head or dorsal region are often dry
+while the seal is floating at the surface, inducing a slight positive bias in
+the hourly percent-dry values reported by the bio-logger. For this study, in
+cases where both bio-logger types were deployed, we preferred hourly percent-dry
+observations from the flipper tag. All data were transmitted by the deployed
+instruments via the Argos satellite network and location data were either
+derived from Argos transmissions or transmitted FastLoc GPS data.
 
@@
 various combinations of length, claw growth ridges [@mclaren1958a], and pelage
-characteristics for some species. Seals determined to be less than one year were
+characteristics. Seals determined to be less than one year were
 classified as young-of-the-year. For those bio-loggers deployed on
@@
 
-Tags that fall off due to molt, attachment failure, or seal mortality and remain
-on ice or land may continue to send data to satellites; i.e., a detached
-bio-logger that is dry (either on ice or land) will record and transmit data
-suggesting the seal is hauled out. Therefore, end times of each deployment were
-identified by examining bio-logger locations, percent-dry records, and dive
-behavior (if available) to determine when bio-loggers ceased providing data
-consistent with seal behavior. For example, a data record that ends with several
-consecutive days (~10+ days) of 100% dry observations and with locations
-indicating the tag was on land would be truncated to not include the final
-stretch of 100% dry observations. The vast majority of deployments end with the
-device detaching in the water and the deployment end date is obvious. There is
-no perfect algorithm for identifying deployment end dates and each deployment in
-question must be considered separately. While not perfect, we are confident our
-reliance on expert opinion and examination of multiple data streams provides the
-best option. Data outside of the deployment start and end times were discarded
-prior to analysis.
-
 Haul-out behavior data were recorded in a manner standard across Wildlife Computers
@@
 Here, missing data that were not successfully recieved from the device are represented as empty rectangles."}
+Tags that fall off due to molt, attachment failure, or seal mortality and remain
+on ice or land may continue to send data to satellites; i.e., a detached
+bio-logger that is dry (either on ice or land) will record and transmit data
+suggesting the seal is hauled out. Therefore, end times of each deployment were
+identified by examining bio-logger locations, percent-dry records, and dive
+behavior (if available) to determine when bio-loggers ceased providing data
+consistent with seal behavior. For example, a data record that ends with several
+consecutive days (~10+ days) of 100% dry observations and with locations
+indicating the tag was on land would be truncated to not include the final
+stretch of 100% dry observations. The vast majority of deployments end with the
+device detaching in the water and the deployment end date is obvious. There is
+no perfect algorithm for identifying deployment end dates and each deployment in
+question must be considered separately. While not perfect, we are confident our
+reliance on expert opinion and examination of multiple data streams provides the
+best option. Data outside of the deployment start and end times were discarded
+prior to analysis.
 

### 2024-01-29 analysis/paper/London_HauloutBeahvior.Rmd
update deployment table

@@
 
-```{r tblDeploy, include=TRUE, tab.cap = "Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2020.
-Total seal hours represents the sum of hourly observations across all seals used in the analysis.
-Because young-of-the-year are advanced to subadult on 1 March of the following year, some individual seals are represented in both columns in this table"}
+(ref:deployTblCaption) **Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2021.** \linebreak Total seal hours represents the sum of hourly observations across all seals used in the analysis.
+Because young-of-the-year are advanced to subadult on 1 March of the following year, some individual seals are represented in both columns in this table
+
+```{r tblDeploy, include=TRUE, tab.cap = "(ref:deployTblCaption)"}
 

### 2024-01-29 analysis/paper/London_HauloutBeahvior.Rmd
improvements to example actogram figure

@@
 
-```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployed on a ribbon seal over two years during the months of March, April, and May.
-This actogram plot represents the transmitted hourly percent-dry values.
-Not all data during a deployment are reliably transmitted from the bio-logger and since data are transmitted in 24-hour chunks occasional missing days are present in the record.
-Here, missing data that were not successfully recieved from the device are represented as empty rectangles."}
 Tags that fall off due to molt, attachment failure, or seal mortality and remain
@@
 prior to analysis.
+
+(ref:examplePlotCaption) **Example percent-dry actogram from bio-logger data**
+\linebreak Haul-out behavior observations recorded by a bio-logger deployed on a
+ribbon seal over two years during the months of March, April, and May. This
+actogram plot represents the transmitted hourly percent-dry values. Not all data
+during a deployment are reliably transmitted from the bio-logger and since data
+are transmitted in 24-hour chunks occasional missing days are present in the
+record. Here, missing data that were not successfully recieved from the device
+are represented as empty rectangles.
+
+```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="(ref:examplePlotCaption)"}
 
@@
   theme(panel.spacing = unit(0,"line")) +
-  ggtitle("Example percent-dry actogram from bio-logger data",
-          subtitle = "Observation records shown are from a single ribbon seal across a two year deployment.") +
   xlab("day of month") + ylab("hour (UTC)")

### 2024-01-29 analysis/paper/London_HauloutBeahvior.Rmd
more methods edits

@@
 Of key interest in this study was the relationship between haul-out behavior and
-weather covariates that vary with time and seal location. We explored the use
+weather covariates that vary with time and seal location. The use of modern
+bio-loggers that record and transmit behavioral data while simultaneously
+providing location estimates was key to this objective. We explored the use
 of a continuous-time correlated random walk [@johnson2008a] movement model
@@
 during the spring, these species are most prevalent near the southern ice edge
-in the Bering Sea [@conn2013b].
+in the Bering Sea [@conn2014a].
 

### 2024-01-29 analysis/paper/London_HauloutBeahvior.Rmd
edit table caption for covariate table

@@
 
-```{r tblCovariates, include=TRUE, tab.cap = "Explanatory covariates used in analyses of binary haul-out records for bearded, spotted, and ribbon seals.
-Note that we also considered select interactions (see article text) between these primary covariates.
-For instance, wind chill was represented by the interaction temperature:wind."}
+(ref:tblCovariatesCaption) **Explanatory covariates used in analyses of binary haul-out records for bearded, spotted, and ribbon seals.** \linebreak Note that we also considered select interactions (see article text) between these primary covariates.
+For instance, wind chill was represented by the interaction temperature:wind.
+
+```{r tblCovariates, include=TRUE, tab.cap = "(ref:tblCovariatesCaption)"}
 

### 2024-01-29 analysis/paper/London_HauloutBeahvior.Rmd
update caption for spatial data map/figure
update to spatial distribution figure caption

@@
 
-Figure \@ref(fig:dataMap) shows the
-spatial distribution of weighted locations with available haul-out behavior data
-used for analysis of each species across the study area.
+Figure \@ref(fig:dataMap) shows the spatial distribution of weighted locations
+with available haul-out behavior data used for analysis of each species across
+the study area. Figure \@ref(fig:dataCal) shows the temporal distribution of all
+haul-out data across the study season for each species. Observations for ribbon
+and spotted seals were concentrated in the months of May and June due to the
+timing of deployment (April and May) and the timing of molt (May and June).
+During molt, seals (and their attached bio-loggers) spend more time out of the
+water and more data are transmitted. Molt timing also impacts when many
+deployments end as any bio-loggers adhered to the hair will fall off. Relative
+to the other species in the study, there were fewer deployments of bio-loggers
+on bearded seals. This resulted in fewer data observations overall and
+noticeably lower in numbers May and June. The majority of bearded seal
+deployments started later in the summer and, by May, bio-loggers had either
+fallen off or their batteries were depleted.
 
@@
 
-(ref:mapCaption) Spatial distribution of haul-out records during the months of March through July 15 for each of the three species.
-Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model.
+(ref:mapCaption) **Spatial distribution of haul-out records from 1 March through July 15 for each of the three species.** \linebreak Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model.
 For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km^2^
 
-```{r dataMap, include=TRUE, fig.cap="(ref:mapCaption)"}
+```{r dataMap, include=TRUE, fig.asp = 0.5, fig.cap="(ref:mapCaption)"}
 
@@
   labs(
-    title = "Spatial distribution of haul-out behavior records",
-    subtitle = "Location estimates were linked by time with haul-out records. 
-    Data are shown collated across 2005-2020 between 01 March-15 July
-    for each of the three species.",
     caption = "One hexagonal cell represents 50 km^2^"

### 2024-01-29 analysis/paper/London_HauloutBeahvior.Rmd
updates to data density plot

@@
 
-Figure \@ref(fig:dataCal) shows the temporal distribution of all
-haul-out data across the study season for each species. Observations for ribbon
-and spotted seals were concentrated in the months of May and June due to the
-timing of deployment (April and May) and the timing of molt (May and June).
-During molt, seals (and their attached bio-loggers) spend more time out of the
-water and more data are transmitted. Molt timing also impacts when many
-deployments end as any bio-loggers adhered to the hair will fall off. Relative
-to the other species in the study, there were fewer deployments of bio-loggers
-on bearded seals. This resulted in fewer data observations overall and
-noticeably lower in numbers May and June. The majority of bearded seal
-deployments started later in the summer and, by May, bio-loggers had either
-fallen off or their batteries were depleted.
-
-```{r dataCal, include=TRUE, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
+(ref:dataDistributionCaption) **Seasonal density of haul-out behavior observations by species** \linebreak Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
 Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020.
+Density values are represented as a color gradient and bar height.
 The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out.
 Additionally, many bio-logger deployments started in April and May.
-The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study."}
+The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study.
+
+```{r dataCal, include=TRUE, fig.cap = "(ref:dataDistributionCaption)"}
 
@@
                                barheight = 0.5)) +
-  facet_grid(species ~ .) + 
-  labs(fill = "seal-hours of bio-logger data",
-       title = "Seasonal density of haul-out behavior observations by species",
-       subtitle = "Data are grouped by day-of-year and presented in seal-hours collated across 
-       years (2005-2020) from March through July.") +
+  facet_wrap(species ~ ., nrow=3) + 
+  labs(fill = "seal-hours of bio-logger data") +
   theme(axis.text.x = element_text(hjust = 0),
+        strip.text.x = element_text(hjust = 0, margin=margin(l=15)),
         axis.text.y = element_blank(),

### 2024-01-29 analysis/paper/London_HauloutBeahvior.Rmd
remove titles from haul-out prediction plots

@@
 
-```{r beardedHOCal, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 30 June for all age and sex classes combined.
-Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range."}
+(ref:beardedHOCalCaption) **Bearded seal predicted haul-out probability** \linebreak Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 30 June for all age and sex classes combined.
+Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range.
+Note, our model predicts lower haul-out probability for bearded seals overall compared to ribbon and spotted seals and the color scales are not directly comparable.
+
+```{r beardedHOCal, include=TRUE, fig.cap="(ref:beardedHOCalCaption)"}
 
@@
   scico::scale_fill_scico(palette = "navia", direction = 1,
-                          begin = 0.1, end = 1,
-                       name = "haul-out probability", 
-                       limits = c(0,1), breaks = c(0.25, 0.50, 0.75),
-                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
-                                              barheight = 0.5, title.hjust = 0.5)
+                       name = "Bearded seal haul-out probability", 
+                       guide = guide_colorbar(title.position = "top", barwidth = 15,
+                                              barheight = 0.5, title.hjust = 0)
   ) +
-  scico::scale_color_scico(palette = "navia", direction = 1,
-                           begin = 0.1, end = 1,
-                           limits = c(0,1)) +
+  scico::scale_color_scico(palette = "navia", direction = 1) +
   guides(col = FALSE)
@@
 p <- p + theme(panel.spacing = unit(0.1, "lines"),
-               legend.position = "bottom",
+               legend.position = "top", legend.justification='left',
+               legend.text.align = 0,
                strip.background = element_rect(colour="white"),
                axis.ticks=element_blank()) +
-  xlab("day of month") + ylab("local solar hour") +
-  labs(title = "Bearded seal predicted haul-out probability",
-       subtitle = "Our model predicts lower haul-out probability for bearded seals overall 
-       compared to ribbon and spotted seals. Highest probability is associated with solar noon and midnight.")
+  xlab("day of month") + ylab("local solar hour") 
 p
@@
 
-```{r ribbonHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
+(ref:ribbonHOCalCaption) **Ribbon seal predicted haul-out probability** \linebreak Predicted hourly haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
 There is an apparent seasonal progression with subadults hauling out earlier in the season followed by adult males and, then, adult females.
 Predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities.
-Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range."}
+Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range.
+
+```{r ribbonHOCal, fig.asp = 0.9, include=TRUE, fig.cap="(ref:ribbonHOCalCaption)"}
 
@@
                           begin = 0.1, end = 1,
-                       name = "haul-out probability", 
+                       name = "Ribon seal haul-out probability", 
                        limits = c(0,1), breaks = c(0.25, 0.50, 0.75),
-                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
-                                              barheight = 0.5, title.hjust = 0.5)
+                       guide = guide_colorbar(title.position = "top", barwidth = 15,
+                                              barheight = 0.5, title.hjust = 0)
   ) +
@@
 p <- p + theme(panel.spacing = unit(0.1, "lines"),
-               legend.position = "bottom",
+               legend.position = "top", legend.justification='left',
+               legend.text.align = 0,
                strip.background = element_rect(colour="white"),
                axis.ticks=element_blank()) +
-  xlab("day of month") + ylab("local solar hour") +
-  ggtitle("Ribbon seal predicted haul-out probability",
-          subtitle = "Our model predicts increased haul-out probability 
-          for ribbon seals centered on solar noon and a sequential 
-          seasonal pattern linked to age and sex classifications.")
+  xlab("day of month") + ylab("local solar hour") 
 p
@@
 
-```{r spottedHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of spotted seals from 1 March to 30 June for each age and sex class used in the model.
+(ref:spottedHOCalCaption) **Spotted seal predicted haul-out probability** \linebreak Predicted hourly haul-out probability of spotted seals from 1 March to 30 June for each age and sex class used in the model.
 Compared to ribbon seals, the period of increased haul-out behavior is broader and there is more alignment in the seasonal timing of haul-out behavior across age and sex classifications.
@@
 As with ribbon seals, redictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities.
-Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range."}
+Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range.
+
+```{r spottedHOCal, fig.asp = 0.9, include=TRUE, fig.cap="(ref:spottedHOCalCaption)"}
 
@@
                           begin = 0.1, end = 1,
-                       name = "haul-out probability", 
+                       name = "Spotted seal haul-out probability", 
                        limits = c(0,1), breaks = c(0.25, 0.50, 0.75),
-                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
-                                              barheight = 0.5, title.hjust = 0.5)
+                       guide = guide_colorbar(title.position = "top", barwidth = 15,
+                                              barheight = 0.5, title.hjust = 0)
   ) +
@@
 p <- p + theme(panel.spacing = unit(0.1, "lines"),
-               legend.position = "bottom",
+               legend.position = "top", legend.justification='left',
+               legend.text.align = 0,
                strip.background = element_rect(colour="white"),
                axis.ticks=element_blank()) +
-  xlab("day of month") + ylab("local solar hour") +
-  ggtitle("Spotted seal predicted haul-out probability",
-          subtitle = "Our model predicts increased haul-out probability 
-          for spotted seals centered on solar noon with general alignment in the
-          seasonal timing across age and sex classifications.")
+  xlab("day of month") + ylab("local solar hour") 
 p

### 2024-02-16 R/get_survey_area.R
expand aerial survey area to include russian

@@ -14,18 +14,56 @@ get_survey_area <- function() {
   
   boss_lines_qry <- "SELECT * FROM surv_boss.geo_tracks_by_effort;"
   chess_lines_qry <- "SELECT * FROM surv_chess.geo_track_by_effort;"
+  jobss_lines_qry <- "SELECT * FROM surv_jobss.geo_track_by_effort;"
+  
+  ru_area_2012 <- read_sf(here::here('data_raw/ru_tracks/ru_tracks_2012/Routes_2012.shp')) |> 
+    sf::st_transform(3571) |> 
+    concaveman::concaveman(concavity = 8) |> 
+    st_simplify(dTolerance = 10000, preserveTopology = TRUE)
+  
+  ru_area_apr_2013 <- read_sf(here::here('data_raw/ru_tracks/ru_tracks_Apr_2013/ru_tracks_Apr_2013.shp')) |> 
+    sf::st_transform(3571) |> 
+    concaveman::concaveman(concavity = 8) |> 
+    sf::st_zm() |> 
+    st_simplify(dTolerance = 10000, preserveTopology = TRUE)
+  
+  ru_area_may_2013 <- read_sf(here::here('data_raw/ru_tracks/ru_tracks_May_2013/ru_tracks_May_2013.shp')) |> 
+    sf::st_transform(3571) |> 
+    concaveman::concaveman(concavity = 10) |> 
+    sf::st_zm() |> 
+    st_simplify(dTolerance = 10000, preserveTopology = TRUE)
+  
+  ru_area_2016 <- read_sf(here::here('data_raw/ru_tracks/ru_tracks_2016/2016_Russian_tracks.shp')) |> 
+    sf::st_transform(3571) |> 
+    concaveman::concaveman(concavity = 8) |> 
+    st_simplify(dTolerance = 10000, preserveTopology = TRUE)
   
   boss_area_sf <- read_sf(con, query = boss_lines_qry) |> 
     dplyr::filter(effort == "On", st_length(geom) > units::set_units(0,m)) |> 
+    sf::st_transform(3571) |> 
     concaveman::concaveman(concavity = 4) |> 
     st_simplify(dTolerance = 25000, preserveTopology = TRUE)
   
   chess_area_sf <- read_sf(con, query = chess_lines_qry) |> 
     dplyr::filter(effort_type == "ON", st_length(geom) > units::set_units(0,m)) |> 
+    sf::st_transform(3571) |> 
[…]

### 2024-02-16 analysis/paper/London_HauloutBeahvior.Rmd
expand survey area; improve wx covariate plots

@@
   plot.title = element_text(face = "bold"),
+  legend.title = element_text(face="bold"),
   # larger, bold title
@@
 on sea ice and are mostly pelagic outside the spring pupping, breeding, and
-molting season [@boveng2018a]. While spotted (_Phoca largha_) and bearded
+molting season [@boveng2018a]. Although spotted (_Phoca largha_) and bearded
 (_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
@@
   st_transform(st_crs(deploy_sf))
-label_coords_chukchi <- st_sfc(st_point(c(-176,73)), crs = 4326) %>%
+label_coords_chukchi <- st_sfc(st_point(c(-175,74.5)), crs = 4326) %>%
   st_transform(st_crs(deploy_sf))
-label_coords_beaufort <- st_sfc(st_point(c(-146.5,72.25)), crs = 4326) %>%
+label_coords_beaufort <- st_sfc(st_point(c(-150,72.25)), crs = 4326) %>%
   st_transform(st_crs(deploy_sf))
@@
 p <- ggplot() +
-  layer_spatial(survey_area, color = "#A26E7C", 
+  annotation_spatial(survey_area, color = "#A26E7C", 
                 fill = "#A26E7C", 
@@
                aes(label = "Beaufort \nsea"), size = 2, lineheight = .85, color = "grey30") +
-  facet_grid(cols = vars(species), labeller = species_labels,
-             switch="x") +
+  facet_grid(cols = vars(species), labeller = species_labels) +
   scale_fill_manual(values = species_colors,
@@
                     name = "species") + 
-  labs(caption = "<span style = 'color: #A26E7C;'><span style = 'font-family: \"Font Awesome 6 Free Regular\";'>&#xf0c8;</span> spatial extent of aerial survey effort in US waters</span>") +
+  labs(caption = "<span style = 'color: #A26E7C;'><span style = 'font-family: \"Font Awesome 6 Free Regular\";'>&#xf0c8;</span> spatial extent of aerial survey effort in US and Russian waters</span>") +
   theme(legend.position = "none",
@@
   geom_tile(linewidth = 0.1, color = "grey75") + 
-  scico::scale_fill_scico(palette = "nuuk", direction = 1,
-                       name = "hourly percent dry", aesthetics = "fill",
-    guide = guide_colorbar(title.position = "bottom", barwidth = 15,
-                           barheight = 0.5, title.hjust = 0.5)
+  scico::scale_fill_scico(palette = "nuuk", direction = 1,labels = scales::percent_format(scale=1),
+                       name = "Example hourly percent dry data from a ribbon seal deployment", aesthetics = "fill",
+    guide = guide_colorbar(title.position = "top", barwidth = 15,
+                           barheight = 0.5, title.hjust = 0)
     ) +
@@
 
-p <- p + theme(legend.position = "bottom") +
+p <- p + theme(legend.position = "top", legend.justification = "left") +
   theme(strip.background = element_rect(colour="white")) +
@@
     aesthetics = "fill",
-    name = "number of locations",
+    name = "number of haul-out records",
     guide = guide_colorbar(
@@
         scale_y_continuous(expand = expansion(mult = c(0, 0))) +
-  guides(fill = guide_colorbar(title.position = "bottom", 
-                               title.hjust = 0.5, barwidth = 15,
+  guides(fill = guide_colorbar(title.position = "top", 
+                               title.hjust = 0, barwidth = 15,
                                barheight = 0.5)) +
   facet_wrap(species ~ ., nrow=3) + 
-  labs(fill = "seal-hours of bio-logger data") +
+  labs(fill = "Seal-hours of haul-out behavior data") +
   theme(axis.text.x = element_text(hjust = 0),
-        strip.text.x = element_text(hjust = 0, margin=margin(l=15)),
+        strip.text.x = element_text(hjust = 1, margin=margin(r=15)),
         axis.text.y = element_blank(),
@@
         panel.grid = element_blank(),
-        legend.position = "bottom",
-        plot.subtitle = element_textbox_simple(
-          color = "grey30",
-          margin = ggplot2::margin(t = 0, b = 0, r = 60)
-        )
+         legend.position = "top", legend.justification='left'
   )
@@
 
-```{r beardedHOwx, fig.asp = 1, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
+(ref:beardedHOwxCaption) **Influence of weather covariates on bearded seal haul-out probability** \linebreak Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
 Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
-Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
+Vertical lines represent the 95% confidence interval around the predicted haul-out probability at observed weather values.
+
+```{r beardedHOwx, fig.asp = 1, include=TRUE, fig.cap="(ref:beardedHOwxCaption)"}
 
@@
 
-```{r ribbonHOwx, fig.asp = 0.9, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification.
+(ref:ribbonHOwxCaption) **Influence of weather covariates on ribbon seal haul-out probability** \linebreak Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classifications.
 Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
-Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
+Vertical lines represent the 95% confidence interval around the predicted haul-out probability at observed weather values.
+
+```{r ribbonHOwx, fig.asp = 0.9, include=TRUE, fig.cap="(ref:ribbonHOwxCaption)"}
 
@@
 
-```{r spottedHOwx,fig.asp = 0.9, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
+(ref:spottedHOwxCaption) **Influence of weather covariates on spotted seal haul-out probability** \linebreak Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
 Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
-Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}
+Vertical lines represent the 95% confidence interval around the predicted haul-out probability.
+
+```{r spottedHOwx,fig.asp = 0.9, include=TRUE, fig.cap="(ref:spottedHOwxCaption)"}
 
@@
 `r ribbon_year_type3$day_year$Prob.F`; day^2^:year,
-$F_{`r paste(ribbon_year_type3$day2_year$Num.df, ribbon_year_type3$day2_year$Den.df, sep=",")`}$
-= `r ribbon_year_type3$day2_year$F.value`; $p$ =
-`r ribbon_year_type3$day2_year$Prob.F`) and spotted seals (day:year,
+$F_{`r paste(ribbon_year_type3$day_2_year$Num.df, ribbon_year_type3$day_2_year$Den.df, sep=",")`}$
+= `r ribbon_year_type3$day_2_year$F.value`; $p$ =
+`r ribbon_year_type3$day_2_year$Prob.F`) and spotted seals (day:year,
 $F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$
@@
 `r spotted_year_type3$day_year$Prob.F`; day^2^:year,
-$F_{`r paste(spotted_year_type3$day2_year$Num.df, spotted_year_type3$day2_year$Den.df, sep=",")`}$
-= `r spotted_year_type3$day2_year$F.value`; $p$ =
-`r spotted_year_type3$day2_year$Prob.F`). Predicted distributions of
+$F_{`r paste(spotted_year_type3$day_2_year$Num.df, spotted_year_type3$day_2_year$Den.df, sep=",")`}$
+= `r spotted_year_type3$day_2_year$F.value`; $p$ =
+`r spotted_year_type3$day_2_year$Prob.F`). Predicted distributions of
 haul-out activity were largely unimodal, but varied some among and
@@
 
-```{r annualHO, fig.asp = 0.75, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability (colored markers) for ribbon and spotted seals across 14 years.
-The full seasonal range of annual predictions along with uncertainty are shown as grey lines.
+(ref:annualHOcaption) **Annual variability in the timing of peak haul-out probability for ribbon and spotted seals.** \linebreak Spring haul-out probability (colored markers) for ribbon and spotted seals across 14 years are shown.
+The full seasonal range of annual predictions along with uncertainty are presented as grey lines.
 Covariates were fixed at local solar noon and under smoothed weather conditions.
 Only those groups (age:sex + year) that included observations from more than one seal are shown.
-Additionally, any groups where data were only available after 1 June or before 1 May are not included."}
+Additionally, any groups where data were only available after 1 June or before 1 May are not included.
+
+```{r annualHO, fig.asp = 0.75, include=TRUE, fig.cap="(ref:annualHOcaption)"}
 
@@
   scale_y_continuous(breaks = c(0.2,0.4,0.6,0.8), labels = c("0.2","0.4", "0.6", "0.8")) +
-  guides(col = guide_legend(ncol = 7)) +
+  guides(col = guide_legend(ncol = 7, title.position = "top",
+                            title.hjust = 0,
+                            title = "Annual timing of peak haul-out probability")) +
   facet_grid(age_sex ~ species,
@@
   ylab("haul-out probability") +
-  labs(title = "Annual variability in the timing of peak haul-out probability",
-       subtitle = "Predictions of maximum annual haul-out probability for ribbon and spotted seals are indicated by the colored squares with grey lines showing the individual seasonal trajectories and uncertainty.",
-       caption = "predictions are shown for local solar noon and under smoothed weather conditions") +
-  theme(legend.position = "bottom",
+  labs(caption = "predictions are shown for local solar noon and under smoothed weather conditions") +
+  theme(legend.position = "top",
+        legend.justification='left',
         axis.title.x = element_blank())
@@
 Unlike previous analyses of seal haul-out behavior in spotted and ribbon seals
-(e.g. [@verhoef2010a], [@conn2013b]), we also investigated the influence of
+(e.g. [@verhoef2010a], [@conn2014a]), we also investigated the influence of
 sex-age class on haul-out probabilities for both species (not for bearded seals
@@
 weather conditions are general indicators of increased solar radiation and lower
-convective heat loss, both of which provides energetic benefits. Low winds and
+convective heat loss, both of which provides energetic benefits (see additonal
+discussion in Supplemental Material [S3][#s3] regarding the potential use of
+solar radiation directly). Low winds and
 precipitation could also enhance predator detection. Our results highlight the
@@
 data in the Bering and Chukchi seas (e.g. @bengtson2005a, @conn2014a,
+@verhoef2014a) have used estimated haul-out probabilities to correct for the
 proportion of animals that are in the water and thus unavailable to be counted.
@@
 
+```{r ho-probs, include = FALSE}
+bearded_ho_prob <- bearded_newdata |> 
+  filter(yday == 141, solar_hour == 12) |> 
+  select(ho_prob,lower95,upper95)
+ribbon_ho_prob <- ribbon_newdata |> 
+  filter(yday == 141, solar_hour == 12) |> 
+  select(age_sex,ho_prob,lower95,upper95)
+spotted_ho_prob <- spotted_newdata |> 
+  filter(yday == 141, solar_hour == 12) |> 
+  select(age_sex,ho_prob,lower95,upper95)
+```
+
 Predictions of absolute haul-out probability in this paper were somewhat
@@
 seals, 0.62 for ribbon seals, and 0.54 for spotted seals, where these maximums
-corresponded to times near solar noon in mid-late May. Our current estimates
+corresponded to times near local solar noon in mid-late May. Results here
+suggest a haul-out probability on May 20 at local solar noon of `r bearded_ho_prob$ho_prob[[1]]` 
+(95% CI: `r bearded_ho_prob$lower95[[1]]` -- `r bearded_ho_prob$upper95[[1]]`)
+for bearded seals across all age and sex classes, 
+`r ribbon_ho_prob$ho_prob[[2]]` 
+(95% CI: `r ribbon_ho_prob$lower95[[2]]` -- `r ribbon_ho_prob$upper95[[2]]`) for
+adult male ribbon seals, `r ribbon_ho_prob$ho_prob[[1]]` 
+(95% CI: `r ribbon_ho_prob$lower95[[1]]` -- `r ribbon_ho_prob$upper95[[1]]`) for
+adult female ribbon seals, `r spotted_ho_prob$ho_prob[[2]]` 
+(95% CI: `r spotted_ho_prob$lower95[[2]]` -- `r spotted_ho_prob$upper95[[2]]`) for
+adult male spotted seals, and `r spotted_ho_prob$ho_prob[[1]]` 
+(95% CI: `r spotted_ho_prob$lower95[[1]]` -- `r spotted_ho_prob$upper95[[1]]`) for
+adult female spotted seals. Our estimates
 of haul-out probability reflect increased sample sizes in terms of
 number of individuals, inclusion of weather covariates, and improvements to the 
-way data were prepared prior to analysis.
+way data were prepared prior to analysis and should be the basis for any
+future estimates of seal abundance from aerial surveys in the Bering and 
+Chukchi seas.
 
@@
 a future study.
+
+Our analysis emphasizes the importance of accounting for seasonal and temporal
+variation in haul-out behavior, as well as associated environmental covariates,
+when interpreting the number of seals counted in aerial surveys. The rapid
+acceleration of climate change in Arctic ecosystems is already occurring and is
+forecast to decrease the quantity and quality of suitable habitat for
+ice-associated seals that rely on sea ice for a variety of activities. Improved
+estimates of seal population abundance from aerial surveys are needed to
+properly monitor the impacts of these changes on seal populations over time.
+Those monitoring surveys will need to be paired with continued investigation and
+assessment of seal haul-out behavior to allow credible predictions about the
+effects of climate disruptions on the abundance and distribution of Arctic seal
+populations.
 
@@
 material \@ref(tab:deploymentTable)). Collated and cleaned data products
-needed to replicate the analysis are available in a separate GitHub
-repository (https://github.com/noaa-afsc/berchukseals-data) and archived at 
-Zenodo. Similarly, all model fits are available on 
-GitHub (https://github.com/noaa-afsc/berchukseals-fits) and archived at Zenodo.
+needed to replicate the analysis along with the results of all model fits are 
+also available and versioned as an R package on 
+GitHub (https://github.com/noaa-afsc/berchuksealsHauloutFits) and archived at 
+Zenodo.
 
@@
   labs(title = "Bearded seal predicted haul-out probability with uncertainty",
-       subtitle = "predictions and associated 95% confidence intervals are shown for three **local solar** times and with smoothed weather covariates") +
+       subtitle = "Predictions and associated 95% confidence intervals are shown for three **local solar** times and with smoothed weather covariates") +
   theme(legend.position = "none")
@@
   labs(title = "Ribbon seal predicted haul-out probability with uncertainty",
-       subtitle = "predictions and associated 95% confidence intervals are shown for three **local solar** times and with smoothed weather covariates") +
+       subtitle = "Predictions and associated 95% confidence intervals are shown for three **local solar** times and with smoothed weather covariates") +
   theme(legend.position = "none")
@@
   labs(title = "Spotted seal predicted haul-out probability with uncertainty",
-       subtitle = "predictions and associated 95% confidence intervals are shown for three **local solar** times and with smoothed weather covariates") +
+       subtitle = "Predictions and associated 95% confidence intervals are shown for three **local solar** times and with smoothed weather covariates") +
   theme(legend.position = "none")
@@
 previously observed in ringed seals and our results in this analysis show some indication of
-increased haul-out probability during dawn and dusk periods for bearded seals.
-For other phocid speices, increased haul-out probability before solar noon or after
+increased haul-out probability during dawn compared to dusk periods for bearded seals
+and some age and sex classes for ribbon and spotted seals.
+For other phocid species, increased haul-out probability before solar noon or after
 solar noon has been observed. Importantly, understanding these relationships
@@
 on aerial survey study design -- a key focus of this paper.  
-Another hesitation we had was that insolation estimates from reanalysis models have
+Another hesitation we had was that solar radiation estimates from reanalysis models have
 not been previously used as a model covariate within a published study of 
@@
 
-That said, we think the idea of insolation as a model covariate in pinniped
+That said, we think the idea of solar radiation as a model covariate in pinniped
 haul-out models is intriguing and worth further exploration. The current
 availability and increased accessibility to detailed climate reanalysis
-products that include insolation is exciting and we encrourage future, more
+products that include solar radiation is exciting and we encourage future, more
 detailed exploration of this as a component in pinniped haul-out analysis. To
-provide some inspiration, we present some initial efforts and model 
-comparisons.
+provide some inspiration, we present some initial efforts and examples for 
+comparison.
 
@@
 not limited to North America. The ERA5 model provides a number of solar
-radiation parameters and it is important to evaluate and understand each of
-these estimates in order to select the one that is likely most relevant to
-seals. Here, we use the 'surface short-wave (solar) radiation downwards.' This
-parameter is described as _the amount of solar radiation (also known as shortwave radiation)
+radiation parameters and it was important to evaluate and understand each of
+these estimates in order to select the one that was likely most relevant to
+seals. Here, we used the 'surface short-wave (solar) radiation downwards'
+parameter. This
+parameter is described as _"the amount of solar radiation (also known as shortwave radiation)
 that reaches a horizontal plane at the surface of the Earth and comprises both
@@
 parameter is the model equivalent of what would be measured by a pyranometer (an
-instrument used for measuring solar radiation) at the surface_ 
+instrument used for measuring solar radiation) at the surface"_ 
 (https://codes.ecmwf.int/grib/param-db/?id=169). Thus, this is the
-parameter which most closely represents the amount of solar radiation likely
+value which most closely represents the amount of solar radiation likely
 felt by a seal hauled out of the water.
@@
 our study area of interest and years of interest. The `reticulate` R package
-(https://CRAN.R-project.org/package=reticulate) allows interaction with Python.
+(https://CRAN.R-project.org/package=reticulate) allowed interaction with Python.
 Additionally, note, extra steps are required to download data on either side
@@
 To explore performance of our solar radiation parameter within a haul-out model
-we replace the various Fourier series parameters in our model from the manuscript
-with the ERA5 _surface solar radiation downwards_ (`era_ssrd`) parameter. As
-with other reanalysis values (from NARR) in the manuscript, the `era-ssrd`
-values are matched in time and space to the seal haul-out observation data. We
+we replaced the various Fourier series parameters in our model from the manuscript
+with the ERA5 _surface solar radiation downwards_ (`era_ssrd_watts`) parameter. As
+with other reanalysis values (from NARR) in the manuscript, the `era-ssrd-watts`
+values are matched in time and space to the seal haul-out observation data; we
 use the full hourly temporal resolution from ERA5. The glmmLDTS framework used
 in the paper does not allow for model comparisons with AIC because of
-the reliance on pseudo-likelihood. The `bam` function within the `mgcv` package
-provides a quick model fitting option that also allows us to do some model
-comparison with AIC. This is sufficient for the general demonstration and
-exploartion purposes here
+the reliance on pseudo-likelihood. The `bam()` function within the `mgcv` package
+provides a quick model fitting option that also allowed us to do some model
+comparison with AIC. This approach was sufficient for the general demonstration and
+exploration purposes here
 but future research should consider a range of model fitting frameworks and
-approaches that might be appropriate.
+approaches that might be more appropriate.
 
@@
 The model specification below was used to specify an `mgcv::bam()` model that
-matches the formula used in the manuscript for spotted seals. The `s(speno, bs = "re")` 
-term is the smooth term for the random effect. All other predictors are
+matched the formula used in the manuscript for ribbon seals. The `s(speno, bs = "re")` 
+term is the smooth term for the random effect. All other predictors were
 the same.
 
-```{r bam-m1-spotted, include=TRUE, echo = TRUE}
+```{r bam-m1-ribbon, include=TRUE, echo = TRUE}
+m1_ribbon <- mgcv::bam(
+  dry ~ age_sex + s(speno, bs = "re") + 
+    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
+    poly(day, 3, raw=TRUE) + 
+    sin1:poly(day, 3, raw=TRUE) +
+    cos1:poly(day, 3, raw=TRUE) +
+    sin2:poly(day, 3, raw=TRUE) +
+    cos2:poly(day, 3, raw=TRUE) +
+    sin3:poly(day, 3, raw=TRUE) +
+    cos3:poly(day, 3, raw=TRUE) +
+    wind*temp2m + pressure + precip + 
+    age_sex:poly(day, 4, raw=TRUE),
+  data = ribbon_model_data,
+  family = binomial,
+  discrete = TRUE)
+```
+
+```{r lag1-ribbon, include=FALSE}
+lag1_ribbon <- acf(resid(m1_ribbon), plot=FALSE)[1][[1]]
+```
+
+Note, the specification for _m1_ribbon_ here does not include
+any AR1 structure for temporal autocorrelation. To include this, we needed to
+provide a value for $\rho$ (or _rho_). We examined the autocorrelation within
+the model and used the lag-1 value for $\rho$ .The value for lag-1 
+autocorrelation was `r formatC(as.numeric(lag1_ribbon))` which is
+rather high but not surprising. We then updated our model specification with
+a value for $\rho$ as well as the `A1.start` argument which specifies (as either 
+__TRUE__ or __FALSE__) the start point of each block.
+
+```{r bam-m2-ribbon, include=TRUE, echo = TRUE}
+m2_ribbon <- mgcv::bam(
+  dry ~ age_sex + s(speno, bs = "re") + 
+    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
+    poly(day, 3, raw=TRUE) + 
+    sin1:poly(day, 3, raw=TRUE) +
+    cos1:poly(day, 3, raw=TRUE) +
+    sin2:poly(day, 3, raw=TRUE) +
+    cos2:poly(day, 3, raw=TRUE) +
+    sin3:poly(day, 3, raw=TRUE) +
+    cos3:poly(day, 3, raw=TRUE) +
+    wind*temp2m + pressure + precip + 
+    age_sex:poly(day, 3, raw=TRUE),
+  data = ribbon_model_data,
+  family = binomial,
+  AR.start = ar1_start,
+  rho = lag1_ribbon,
+  discrete = TRUE)
+```
+
+The model specification for exploring the use of solar radiation was
+specified similarly but without all of the Fourier series parameters and
+interactions.
+
+```{r bam-ssrd-ribbon, include=TRUE, echo = TRUE}
+m2_ssrd_ribbon <- mgcv::bam(
+  dry ~ age_sex + s(speno, bs = "re") + 
+    era5_ssrd_watts +
+    poly(day, 3, raw=TRUE) + 
+    era5_ssrd_watts:poly(day, 3, raw=TRUE) +
+    wind*temp2m + pressure + precip + 
+    age_sex:poly(day, 3, raw=TRUE),
+  data = ribbon_model_data,
+  family = binomial,
+  AR.start = ar1_start,
+  rho = lag1_ribbon,
+  discrete = TRUE)
+```
+
+The two models were compared with AIC to evaluate whether the
+reduction in degrees of freedom with fewer terms in the solar radiation
+model was matched with improved explanatory power in the model fit. While the
+model and code specified above is for ribbon seals, the same approach was
+repeated for bearded and spotted seals.
+
+```{r,bam-ssrd-spotted-bearded}
 m1_spotted <- mgcv::bam(
@@
   discrete = TRUE)
-```
-
-```{r lag1-spotted, include=FALSE}
+
 lag1_spotted <- acf(resid(m1_spotted), plot=FALSE)[1][[1]]
-```
-
-Note, the specification for _m1_spotted_ here does not include
-any AR1 structure for temporal autocorrelation. To include this, we need to
-provide a value for $\rho$ (or _rho_). We can examine the autocorrelation within
-the model and use the lag-1 value for $\rho$ .The value for lag-1 
-autocorrelation is `r formatC(as.numeric(lag1_spotted))` which is
-rather high but not surprising. We can, now, update our model specification with
-a value for $\rho$ as well as the `A1.start` argument with specifies (TRUE/FALSE) the
-start point of each block.
-
-```{r bam-m2-spotted, include=TRUE, echo = TRUE}
+
 m2_spotted <- mgcv::bam(
@@
   discrete = TRUE)
-```
-
-The model specification for exploring the use of solar radiation is
-specified similarly but without all of the Fourier series parameters and
-interactions.
-
-```{r bam-ssrd-spotted, include=TRUE, echo = TRUE}
+
 m2_ssrd_spotted <- mgcv::bam(
@@
   rho = lag1_spotted,
-  discrete = TRUE)
-```
-
-Lastly, the two models were compared with AIC to evaluate whether the
-reduction in degrees of freedom with fewer terms in the solar radiation
-model was matched with improved explanatory power in the model fit. While the
-model and code specified above is for spotted seals, the same approach was
-repeated for bearded and ribbon seals.
-
-```{r,bam-ssrd-ribbon-bearded}
-m1_ribbon <- mgcv::bam(
-  dry ~ age_sex + s(speno, bs = "re") + 
-    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
-    poly(day, 3, raw=TRUE) + 
-    sin1:poly(day, 3, raw=TRUE) +
-    cos1:poly(day, 3, raw=TRUE) +
-    sin2:poly(day, 3, raw=TRUE) +
-    cos2:poly(day, 3, raw=TRUE) +
-    sin3:poly(day, 3, raw=TRUE) +
-    cos3:poly(day, 3, raw=TRUE) +
-    wind*temp2m + pressure + precip + 
-    age_sex:poly(day, 4, raw=TRUE),
-  data = ribbon_model_data,
-  family = binomial,
-  discrete = TRUE)
-
-lag1_ribbon <- acf(resid(m1_ribbon), plot=FALSE)[1][[1]]
-
-m2_ribbon <- mgcv::bam(
-  dry ~ age_sex + s(speno, bs = "re") + 
-    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
-    poly(day, 3, raw=TRUE) + 
-    sin1:poly(day, 3, raw=TRUE) +
-    cos1:poly(day, 3, raw=TRUE) +
-    sin2:poly(day, 3, raw=TRUE) +
-    cos2:poly(day, 3, raw=TRUE) +
-    sin3:poly(day, 3, raw=TRUE) +
-    cos3:poly(day, 3, raw=TRUE) +
-    wind*temp2m + pressure + precip + 
-    age_sex:poly(day, 3, raw=TRUE),
-  data = ribbon_model_data,
-  family = binomial,
-  AR.start = ar1_start,
-  rho = lag1_ribbon,
-  discrete = TRUE)
-
-m2_ssrd_ribbon <- mgcv::bam(
-  dry ~ age_sex + s(speno, bs = "re") + 
-    era5_ssrd_watts +
-    poly(day, 3, raw=TRUE) + 
-    era5_ssrd_watts:poly(day, 3, raw=TRUE) +
-    wind*temp2m + pressure + precip + 
-    age_sex:poly(day, 3, raw=TRUE),
-  data = ribbon_model_data,
-  family = binomial,
-  AR.start = ar1_start,
-  rho = lag1_ribbon,
   discrete = TRUE)
@@
 
-## Results
-
-To evaluate whether the solar radiation parameter matches our expectations and
-compares well with hour of the day, we can visualize the variability of the 
+A similar approach to that presented in this manuscript for prediction was 
+employed with solar radiation values in lieu of hour of day. For prediction
+values, quantiles (5% increments) of the observed range of ERA5 solar radiation values were
+used with 100% representing the maximum observed solar radiation value. This 
+allowed similar data visualizations and easier comparisons to those
+predictions in the manuscript that include hour of day.
+
+### Results
+
+To evaluate whether the solar radiation parameter matched our expectations and
+compared well with hour of the day, we visualized the variability of the 
 `era5_ssrd` values within
-our study area as they relate to hour of the day. The unimodal distribution
+our study area as they relate to hour of the day (\@ref(fig:era5-ssrd-plot)). 
+The unimodal distribution
 is centered around the middle of the solar day with peak solar radiation
@@
 ssrd_solarhour_plot <- ggplot(data = pts_ssrd_sf, aes(x=solar_hour, y=value)) + 
-  geom_point(size = 0.5, alpha = 0.2, color = "#edd9a3") + 
+  geom_point(size = 0.5, alpha = 0.3, color = "#edd9a3") + 
   geom_smooth(se=TRUE, linewidth = 0.2, color = "#4b2991") +
-  # geom_vline(xintercept = 13, linetype="dashed", 
-  #               linewidth = 1) +
   scale_x_continuous(breaks = c(3,6,9,12,15,18,21),
@@
                                 "21:00")) +
+  scale_y_continuous(labels = scales::label_scientific()) +
   xlab("Local Solar Hour") +
@@
 
-The spotted seal model matching the specification from the manuscript resulted in 
+The bearded seal model matching the specification from the manuscript resulted in 
+`r AIC(m2_bearded,m2_ssrd_bearded)['m2_bearded','df']` degrees of freedom and an AIC value of
+`r AIC(m2_bearded,m2_ssrd_bearded)['m2_bearded','AIC']`. The model with solar radiation resulted in 
+`r AIC(m2_bearded,m2_ssrd_bearded)['m2_ssrd_bearded','df']` degrees of freedom and an AIC value of
+`r AIC(m2_bearded,m2_ssrd_bearded)['m2_ssrd_bearded','AIC']`. The ribbon seal model matching the specification from the manuscript resulted in 
+`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ribbon','df']` degrees of freedom and an AIC value of
+`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ribbon','AIC']`. The model with solar radiation resulted in 
+`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ssrd_ribbon','df']` degrees of freedom and an AIC value of
+`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ssrd_ribbon','AIC']`. The spotted seal model matching the specification from the manuscript resulted in 
 `r AIC(m2_spotted,m2_ssrd_spotted)['m2_spotted','df']` degrees of freedom and an AIC value of
@@
 `r AIC(m2_spotted,m2_ssrd_spotted)['m2_ssrd_spotted','AIC']`. Despite the additional terms, the
-model with the Fourier series representation of hour of day resulted in a 
-lower AIC value and was still preferred for spotted seals. The ribbon seal model matching the specification from the manuscript resulted in 
-`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ribbon','df']` degrees of freedom and an AIC value of
-`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ribbon','AIC']`. The model with solar radiation resulted in 
-`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ssrd_ribbon','df']` degrees of freedom and an AIC value of
-`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ssrd_ribbon','AIC']`. Despite the additional terms, the
-model with the Fourier series representation of hour of day resulted in a 
-lower AIC value and was still preferred for bearded seals. The bearded seal model matching the specification from the manuscript resulted in 
-`r AIC(m2_bearded,m2_ssrd_bearded)['m2_bearded','df']` degrees of freedom and an AIC value of
-`r AIC(m2_bearded,m2_ssrd_bearded)['m2_bearded','AIC']`. The model with solar radiation resulted in 
-`r AIC(m2_bearded,m2_ssrd_bearded)['m2_ssrd_bearded','df']` degrees of freedom and an AIC value of
-`r AIC(m2_bearded,m2_ssrd_bearded)['m2_ssrd_bearded','AIC']`. Despite the additional terms, the
-model with the Fourier series representation of hour of day resulted in a 
-lower AIC value and was still preferred for bearded seals.
+models with the Fourier series representation of hour of day resulted in a 
+lower AIC value and were still preferred models for each of the species.
+
+Predictions from the model fits and visualization of those predictions were
+produced for each species but, here, we only present visualizations from 
+ribbon seals as an example (Figure \@ref(fig:ssrd-prediction-heatmap) and Figure
+\@ref(fig:ssrd-ribbonPredSE)). Similar seasonal patterns previously observed were
+still apparent with subadults hauling out earlier in the season followed by 
+adult males and, then, adult females. The observed relationship with hour of
+day and the centering of peak haul-out probability around solar noon was
+reflected in these predictions as a one-sided distribution with maximum solar
+radiation having the highest haul-out probability and minimal solar radiation
+the least. The seasonal distribution of haul-out probability along with 95%
+confidence intervals also provided comparable insights (see figures 
+\@ref(fig:ribbonPredSE) and \@ref(fig:ssrd-ribbonPredSE)). That said, subtle
+differences in the shape and extent of confidence limits were present.
+
+```{r ssrd-prediction-heatmap, include = TRUE, fig.asp=0.9, fig.cap="Predicted haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
+In this model, solar radiation was used in lieu of hour of day.
+The apparent seasonal progression with subadults hauling out earlier in the season followed by adult males and, then, adult females is still notable although maybe not as clear.
+Predictions for young of the year still show their transition from newly weaned pups resting on the ice to more in-water activities.
+The overall pattern is in agreement with a one-sided view of Figure 7 where maximum solar radiation is equivalent to local solar noon."}
+r <- ribbon_newdata |> 
+  group_by(age_sex,day,yday) |> 
+  summarize(across(c(northing,temp2m,wind,pressure,precip), 
+                   ~ mean(.x, na.rm = TRUE)))
+
+ssrd_breaks <- ribbon_model_data$era5_ssrd_watts |> 
+  quantile(probs = seq(0,1,by=0.05))
+
+r <- r |> expand_grid(tibble(era5_ssrd_watts = ssrd_breaks,
+       ssrd_quants = factor(names(ssrd_breaks),
+                            levels = names(ssrd_breaks) )))
+
+spenos <- unique(ribbon_model_data$speno) %>% 
+  as.character()
+
+r <- r %>% 
+  mutate(speno = sample(spenos,1))
+
+ribbon_ssrd_predict <- bind_cols(r,predict(m2_ssrd_ribbon,
+                               newdata = r,
+                               type = "response",
+                               se.fit = TRUE,
+                               exclude = "s(speno)")
+)
+
+plot_df <- ribbon_ssrd_predict %>%
+  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+         month = lubridate::month(date,label = TRUE),
+         day = lubridate::day(date)) %>%
+  filter(!month %in% c("Jul","Aug"),
+         lubridate::yday(date) >73)
+
+p <- ggplot(plot_df, aes(day, ssrd_quants, fill = fit, color = fit)) +
+  geom_tile(linewidth = 0.15) +
+  scico::scale_fill_scico(palette = "navia", direction = 1,
+                          begin = 0.1, end = 1,
+                       name = "haul-out probability", 
+                       limits = c(0,1), breaks = c(0.25, 0.50, 0.75),
+                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
+                                              barheight = 0.5, title.hjust = 0.5)
+  ) +
+  scico::scale_color_scico(palette = "navia", direction = 1,
+                           begin = 0.1, end = 1,
+                           limits = c(0,1)) +
+  guides(col = FALSE)
+
+p <- p + facet_grid(age_sex~month,
+                    scales = "free_x",
+                    space = "free_x",
+                    labeller = labeller(
+                      age_sex = age_sex_labels
+                    ))
+p <- p + scale_x_continuous(breaks = c(5,10,15,20,25)) +
+  scale_y_discrete(breaks = c("20%","60%","95%"), labels = c("15%","50%","100%")) +
+  coord_cartesian(expand=FALSE)
+p <- p + theme(panel.spacing.x = unit(0.1, "lines"),
+               panel.spacing.y = unit(0.2, "lines"),
+               legend.position = "bottom",
+               strip.background = element_rect(colour="white"),
+               axis.ticks=element_blank()) +
+  xlab("day of month") + ylab("% of max solar radiation") +
+  ggtitle("Solar radiation as a predictor of ribbon seal haul-out probability",
+          subtitle = "Increased haul-out probability at higher levels of solar
+          radiation is consistent with maximum predicted haul-out at solar noon")
+p
+```
+
+```{r ssrd-ribbonPredSE,fig.asp = 0.7, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for ribbon seals.
+In this model predictions are shown for low, medium, and high values of solar radiation (as percentages of the maximum value observed) in lieu of local solar hour.
+There's general agreement in the overall seasonal patterns between the two approaches but sublte differences in shape and extent of the confidence limits were present (see Figure S2 for comparisons)."}
+
+ssrd_facet_labels <- c(
+  '35%' = 'Low Solar Radiation (35% of max value)',
+  '65%' = 'Medium Solar Radiation (65% of max value)',
+  '100%' = 'High Solar Radiation (100% of max value)'
+) |> as_tibble(rownames = 'ssrd_facet') |> 
+  mutate(value = forcats::fct(value,levels = value))
+
+plot_df <- ribbon_ssrd_predict %>%
+  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+         month = lubridate::month(date,label = TRUE),
+         day = lubridate::day(date)) %>%
+  filter(!month %in% c("Jul","Aug"),
+         lubridate::yday(date) >73,
+         ssrd_quants %in% c("35%","65%","100%")) |> 
+  inner_join(ssrd_facet_labels, by = join_by(ssrd_quants == ssrd_facet)) |> 
+  rename(ssrd_facet_label = value)
+
+ssrd_ribbon_predse <- ggplot(plot_df, aes(date,fit)) +
+  geom_ribbon(aes(ymin=fit-(1.96*se.fit),ymax=fit + (1.96*se.fit)), 
+              fill = species_colors['Ribbon seal'], alpha = 0.3) +
+  geom_line(aes(date,fit),
+            color = species_colors['Ribbon seal']) +
+  facet_grid(age_sex ~ ssrd_facet_label,
+             scales = "free_x",
+                    space = "free_x",
+                    labeller = labeller(
+                      age_sex = age_sex_labels,
+                      ssrd_facet_label = label_wrap_gen(23)
+                    )) +
+  scale_y_continuous(breaks = c(0.2,0.4,0.6,0.8), labels = c("0.2","0.4", "0.6", "0.8"), limits = c(0,1)) +
+  xlab("day of year") +
+  ylab("haul-out probability") +
+  labs(title = "Solar radiation as a predictor of haul-out probability (with uncertainty)",
+       subtitle = "Predictions and associated 95% confidence intervals for three **solar radiation** quantile categories show similar seasonal patterns to those based on local solar hour") +
+  theme(legend.position = "none")
+
+ssrd_ribbon_predse
+```
+
+### Discussion
+
+<!-- - ear5 solar radiation has potential as an informative covariate that is a more -->
+<!-- directly linked with seal behavior and physiology -->
+<!-- - the unimodal distribution and correlation with hour of day also demonstrates -->
+<!-- it could be considered a replacement for hour of day; though, with increased -->
+<!-- variability given the impact of cloud cover and other weather parameters -->
+<!-- - despite the reduced number of parameters, AIC comparison for all species -->
+<!-- still preferred the model with local solar hour.  -->
+<!-- - there is some a slight skewness present in the predicted distributions with -->
+<!-- higher predictions in the morning compared to the late afternoon; this isn't -->
+<!-- captured with solar radiation alone -->
+
+Solar radiation has potential as an informative
+covariate in pinniped haul-out models that can be directly linked to seal
+physiology and expected behavioral changes. The ERA5's _surface solar radiation downwards_
+values aligned with hour of day and maximum values
+occurred at or just after local solar noon. This highlighted the informative
+potential for this approach. However, despite an overall reduction in the
+total number of parameters and degrees of freedom, AIC comparison still favored
+the models for each species that included hour of day as a Fourier series. This
+analysis was not intended to be a full comparison -- we simply want to 
+demonstrate the potential and inspire further investigation -- but, there are
+three possibilities that might explain the preference for hour of day.
+
+First,
+there are a broad range of solar radiation values represented for each hour of the
+day. Cloud cover, fog, and precipitation all reduce downward solar radiation at
+the surface and we might expect this to impact haul-out probability. However, the photoperiod and the timing of sunrise and sunset are
+not impacted by weather and seals may be responding to these signals more than
+the amount of solar radiation. Additionally, this study spans a range of 
+physiological cycles and energetic needs and higher solar radiation may not be
+a consistently driving influence on seals. Increased energy from the sun may
+be important during molt but less so during pupping and breeding periods. Second,
+the timing and duration of haul-out behavior may also be influenced by diel
+patterns in weather (e.g. lower winds in the morning) or ecosystem dynamics (e.g.
+prey availability) that lead to a skewness in the distribution of haul-out 
+behavior that wouldn't be reliably captured by solar radiation values. Third,
+this effort is only an initial effort to explore the use of solar radiation in
+pinniped haul-out models. A more in depth and rigorous exploration of this topic
+might discover an approach that results in a more parsimonious and preferred 
+model formulation.
+
+Again, we want to acknowledge Anthony Fisbach for the suggestion during the
+peer review process. We think this is an excellent example of the peer review
+process working to improve the quality of our manuscript and advance the
+scientific process. We hope others will take our example and expand on it
+within future analyses.

### 2024-02-16 R/wx_plots.R

@@ -29,14 +29,22 @@ plot_ribbon_wx <- function(ribbon_fit_obj,age_sex_colors,age_sex_labels) {
     group_by(age_sex) %>% 
     
     ggplot(aes(temp2m*27, ho_prob, color = age_sex)) +
+    geom_ribbon(aes(ymin = lower95, ymax = upper95, 
+                    fill = age_sex), linetype = 0, alpha = 0.2,
+                show.legend = FALSE) +
     geom_pointrange(aes(ymin = lower95, ymax = upper95), 
                     size = 0.25, alpha = 1, stroke = 0) +
     scale_color_manual(values= age_sex_colors,
                        labels = age_sex_labels,
+                       name = "Ribbon seal",
                        guide = guide_legend(override.aes = list(alpha = 1),
+                                            title.position = "top",
+                                            title.hjust = 0,
                                             direction = "horizontal")) +
-    theme(legend.title = element_blank(),
-          legend.position = "none") +
+    scale_fill_manual(values = age_sex_colors,
+                      labels = age_sex_labels) +
+    theme(legend.position = "top",
+          legend.justification = "left") +
     xlab("temperature (C)") + ylab("haul-out probability")
   
   p_wind <- create_ribbon_newdata(ribbon_fit_obj, 
@@ -50,14 +58,19 @@ plot_ribbon_wx <- function(ribbon_fit_obj,age_sex_colors,age_sex_labels) {
     group_by(age_sex) %>% 
     
     ggplot(aes(wind*10, ho_prob, color = age_sex)) +
+    geom_ribbon(aes(ymin = lower95, ymax = upper95, 
+                    fill = age_sex), linetype = 0, alpha = 0.2,
+                show.legend = FALSE) +
     geom_pointrange(aes(ymin = lower95, ymax = upper95), 
                     size = 0.25, alpha = 1, stroke = 0) +
     scale_color_manual(values= age_sex_colors,
                        labels = age_sex_labels,
                        guide = guide_legend(override.aes = list(alpha = 1),
                                             direction = "horizontal")) +
+    scale_fill_manual(values = age_sex_colors,
+                      labels = age_sex_labels) +
     theme(legend.title = element_blank(),
-          legend.position = "top") +
+          legend.position = "none") +
     xlab("wind (m/s)") + ylab("haul-out probability")
   
   p_pressure <- create_ribbon_newdata(ribbon_fit_obj, margins=TRUE, 
[…]

### 2024-02-17 analysis/paper/London_HauloutBeahvior.Rmd
pconn revision comments addressed

@@
 [-@verhoef2014a] found haul-out probabilities increased gradually starting in
-March and peaked in May and June for bearded, ribbon, and spotted seals. Recent
-Such
+March and peaked in May and June for bearded, ribbon, and spotted seals. Such
 analyses have not been limited to the Arctic. In the Antarctic, Bengtson and
@@
 
-Haul-out behavior data were recorded in a manner standard across Wildlife Computers
-bio-loggers and transmitted via the Argos satellite
-network as hourly percent-dry timelines. For each hour of a day, the wet/dry
-sensor was polled by the tag firmware every few seconds and the percent of the
-hour in the dry state was calculated (Figure \@ref(fig:examplePlot)). On board
-the bio-logger, hourly percent-dry calculations were rounded to the nearest 10%
-inclusive of 0% and 100% along with additional values at 3% and 98%. This
-compression resulted in additional data transmission as each message consisted
-of two complete 24-hour records. Memory capacity allowed caching of percent-dry
-records for several weeks or months and each message was transmitted several
-times to ensure reception at the satellite. Bio-loggers were, typically,
-deployed and programmed in a manner to maximize data transmission during the
-spring pupping and molting period, though hourly percent-dry data were not
-always successfully transmitted. This is due to a variety of factors including
-satellite coverage, tag availability (e.g., tags mounted to the rear flipper
-often do not transmit while at sea), tag performance, duty cycling, and
-atmospheric interference. Fortunately, missing records do not substantially bias
-inference about haul-out probabilities [@conn2012c].
+Haul-out behavior data were recorded in a manner standard across Wildlife
+Computers bio-loggers and transmitted via the Argos satellite network as hourly
+percent-dry timelines. For each hour of a day, the wet/dry sensor was polled by
+the tag firmware every few seconds and the percent of the hour in the dry state
+was calculated (Figure \@ref(fig:examplePlot)). On board the bio-logger, hourly
+percent-dry calculations were rounded to the nearest 10% inclusive of 0% and
+100% along with additional values at 3% and 98%. This compression resulted in
+additional data transmission as each message consisted of two complete 24-hour
+records. Memory capacity allowed caching of percent-dry records for several
+weeks or months and each message was transmitted several times to ensure
+reception at the satellite. Bio-loggers were deployed and programmed in a manner
+to maximize data transmission during the spring pupping and molting period,
+though hourly percent-dry data were not always successfully transmitted. This is
+due to a variety of factors including satellite coverage, tag availability
+(e.g., tags mounted to the rear flipper often do not transmit while at sea), tag
+performance, duty cycling, and atmospheric interference. Fortunately, missing
+records do not substantially bias inference about haul-out probabilities
+[@conn2012c].
 
@@
 across the region 8 times daily. For this study, NARR weather values were subset
-to the extent of our study area over the Bering and Chukchi seas at 3-hr
+to the extent of our study area over the Bering, Chukchi, and Beaufort seas at 3-hr
 intervals based on the native grid resolution of 32 km (1024 km^2^). The
@@
 previous approaches [@verhoef2014a, @london2012] and only
-`r (1 - round((portion_10(analysis_data) + portion_90(analysis_data))) * 100)`% of our
+`r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
 observations fell between 10% and 90% hourly percent-dry.
@@
 the study area. Figure \@ref(fig:dataCal) shows the temporal distribution of all
-haul-out data across the study season for each species. Observations for ribbon
+haul-out data across the study season for each species. Observations of ribbon
 and spotted seals were concentrated in the months of May and June due to the
@@
 adults (Figure \@ref(fig:ribbonHOCal); see also 
-@\ref(fig:ribbonPredSE)). The behavior was clearly centered
+\@ref(fig:ribbonPredSE)). The behavior was clearly centered
 around local solar noon and expanded to other hours later in the spring
@@
 young-of-the-year records begin after weaning and the model predictions
-demonstrated the ontogeny of in-water activities (e.g. swimming, foraging)
+demonstrated the development of in-water activities (e.g. swimming, foraging)
 in May. Adult females had a more protracted haul-out season compared to
@@
 (e.g. [@verhoef2010a], [@conn2014a]), we also investigated the influence of
-sex-age class on haul-out probabilities for both species (not for bearded seals
+sex-age class on haul-out probabilities of both species (but not for bearded seals
 because of low sample size). Field identification of age class can be inexact,
@@
 other phocids where yearlings and subadults molt first followed by adult females
-and males [@reder2003, @thompson1987, @kirkman2003].
+and males [@reder2003; @thompson1987; @kirkman2003].
 
@@
 [-@udevitz2009]) and a few select studies of ice-associated seals
-[@hamilton2018, @perry2017]. In our study, ribbon seals seemed to be the most
+[@hamilton2018; @perry2017]. In our study, ribbon seals seemed to be the most
 influenced by weather, with wind, temperature, and barometric pressure all being
 important components of the model. Spotted seals were most affected by wind and
-barometric pressure. For bearded seals, the model indicated wind and
-temperature had the greatest impact. In general, and as might be expected, seals
-were more likely to haul-out when daily temperatures were warmer, winds speeds
-were lower, barometric pressure was higher, and precipitation was lower. Those
-weather conditions are general indicators of increased solar radiation and lower
-convective heat loss, both of which provides energetic benefits (see additonal
-discussion in Supplemental Material [S3][#s3] regarding the potential use of
-solar radiation directly). Low winds and
-precipitation could also enhance predator detection. Our results highlight the
-importance of incorporating weather covariates when analyzing haul-out behavior
-and calculating availability corrections for aerial surveys.
+barometric pressure. For bearded seals, the model indicated wind and temperature
+had the greatest impact. In general, and as might be expected, seals were more
+likely to haul-out when daily temperatures were warmer, winds speeds were lower,
+barometric pressure was higher, and precipitation was lower. Those weather
+conditions are general indicators of increased solar radiation and lower
+convective heat loss, both of which provides energetic benefits (see additional
+discussion in Supplemental Material [Exploring Insolation (Solar Radiation) as a
+Model Covariate] regarding the potential use of solar radiation directly). Low
+winds and precipitation could also enhance predator detection. Our results
+highlight the importance of incorporating weather covariates when analyzing
+haul-out behavior and calculating availability corrections for aerial surveys.
 
@@
 $\hat{a} = 0.5$. A Horvitz-Thompson-type estimator for abundance is simply 
-$\hat{N} = C \div \hat{a}$, which has expectation 
+$\hat{N} = \frac{C}{\hat{a}}$, which has expectation 
 $\mathbb{E}(\hat{a}) = \frac{0.5 \cdot N}{0.5} = N$, as desired.  Now
@@
 paper, we have shown that there can be considerable differences in the number of
-seals hauled out on ice based on these factors. We recommend that future
+seals hauled out on ice based on these factors and subsequent analyses have shown
+the potential for considerable bias in abundance estimates if such covariates
+are unaccounted for [@conn2023]. We recommend that future
 abundance analyses employ availability models that account for them. For
@@
 survey tracks, and auxiliary information about snow depth and timing of melt in
-a future study.
+a future study (see, e.g., @lindsay2021).
 

### 2024-02-17 analysis/paper/London_HauloutBeahvior.Rmd
ver hoef and crawford comments addressed

@@
   Ice-associated seals rely on sea ice for a variety of activities, including
-  breeding, molting, pupping, and resting. In the Arctic, many of these activities
-  occur in spring (April -- June) as sea ice begins to melt and retreat northward.
-  Rapid acceleration of climate change in Arctic ecosystems is therefore of
-  concern as the quantity and quality of suitable habitat is forecast to decrease.
-  Improved estimates of seal population abundance are needed to properly monitor
-  the impacts of these changes over time. In this paper, we use hourly
-  percent-dry data from satellite-linked bio-loggers deployed
-  between 2005 and 2021 to quantify the proportion of seals hauled out on ice.
-  This information is needed to accurately estimate abundance from aerial survey counts of
-  ice-associated seals (i.e., to correct for the proportion of animals that
-  are in the water while surveys are conducted). In addition to providing
-  essential data for survey 'availability' calculations, our analysis also provides 
-  insights into the seasonal timing and environmental
+  pupping, breeding, molting, and resting. In the Arctic, many of these
+  activities occur in spring (April -- June) as sea ice begins to melt and
+  retreat northward. Rapid acceleration of climate change in Arctic ecosystems
+  is therefore of concern as the quantity and quality of suitable habitat is
+  forecast to decrease. Improved estimates of seal population abundance are
+  needed to properly monitor the impacts of these changes over time. In this
+  paper, we use hourly percent-dry data from satellite-linked bio-loggers
+  deployed between 2005 and 2021 to quantify the proportion of seals hauled out
+  on ice. This information is needed to accurately estimate abundance from
+  aerial survey counts of ice-associated seals (i.e., to correct for the
+  proportion of animals that are in the water while surveys are conducted). In
+  addition to providing essential data for survey 'availability' calculations,
+  our analysis also provides insights into the seasonal timing and environmental
   factors affecting haul-out behavior by ice-associated seals. We specifically
@@
   ringed seals (_Phoca hispida_) can be out of the water but hidden from view in
-  snow lairs, they were not included in this analysis. Using generalized linear mixed
-  pseudo-models to properly account for temporal autocorrelation, we fit models
-  with covariates of interest (e.g., day-of-year, solar hour, age-sex class, wind
-  speed, barometric pressure, temperature, precipitation) to examine their ability
-  to explain variation in haul-out probability. We found evidence for
-  strong diel and within-season patterns in haul-out behavior, as well as
-  strong weather effects (particularly wind and temperature). In general, seals
-  were more likely to haul out on ice in the middle of the day and when wind speed
-  was low and temperatures were higher. Haul-out probability increased through
-  March and April, peaking in May and early June before declining again. The
-  timing and frequency of haul-out events also varied based on species and age-sex
-  class. For ribbon and spotted seals, models with year effects were highly
-  supported, indicating that the timing and magnitude of haul-out behavior varied
-  among years. However, we did not find broad evidence that haul-out timing was linked
-  to annual sea ice extent. Our analysis emphasizes the importance of accounting
-  for seasonal and temporal variation in haul-out behavior, as well as associated
-  environmental covariates, when interpreting the number of seals counted in
-  aerial surveys.
+  snow lairs analysis of their 'availability' to surveys requires special
+  consideration; therefore,they were not included in this analysis. Using
+  generalized linear mixed pseudo-models to properly account for temporal
+  autocorrelation, we fit models with covariates of interest (e.g., day-of-year,
+  solar hour, age-sex class, wind speed, barometric pressure, temperature,
+  precipitation) to examine their ability to explain variation in haul-out
+  probability. We found evidence for strong diel and within-season patterns in
+  haul-out behavior, as well as strong weather effects (particularly wind and
+  temperature). In general, seals were more likely to haul out on ice in the
+  middle of the day and when wind speed was low and temperatures were higher.
+  Haul-out probability increased through March and April, peaking in May and
+  early June before declining again. The timing and frequency of haul-out events
+  also varied based on species and age-sex class. For ribbon and spotted seals,
+  models with year effects were highly supported, indicating that the timing and
+  magnitude of haul-out behavior varied among years. However, we did not find
+  broad evidence that haul-out timing was linked to annual sea ice extent. Our
+  analysis emphasizes the importance of accounting for seasonal and temporal
+  variation in haul-out behavior, as well as associated environmental
+  covariates, when interpreting the number of seals counted in aerial surveys.
 output:
@@
 United States Endangered Species Act (ESA)) for timely estimates of population
-abundance and trend of these species mean improved aerial survey effort is
-needed -- and, must be paired with improved knowledge of haul-out behavior to
-ensure appropriate survey design, robust methods, and accurate estimates.
-Several studies have contributed estimates of the distribution and abundance of
-ice-associated seal species in the Arctic using aerial surveys (e.g.,
-@bengtson2005a, @conn2014a, and @verhoef2014a) and more recent efforts
-have significantly expanded on previous survey effort. Such abundance studies
-are conducted over very large areas and estimation of absolute abundance
-requires making inference about numerous issues affecting the observation of
-seals on ice. These include availability (only seals on ice are available to
-be counted), detection probability (observers or automated detection systems may
-miss some seals on ice), species misclassification, and possible
-disturbance of seals by aircraft [@conn2014a; @verhoef2014a]. Refining these
-inferences will improve the accuracy of abundance estimates and, hopefully,
-allow credible predictions about the effects of climate disruptions on the
-abundance and distribution of Arctic seal populations.
-
-How ice-associated seals use sea ice as a haul-out platform varies between species.
-Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
-on sea ice and are mostly pelagic outside the spring pupping, breeding, and
-molting season [@boveng2018a]. Although spotted (_Phoca largha_) and bearded
-(_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
-ice as a haul-out platform during the spring and early summer [@frost2018].
-Ringed seals (_Phoca hispida_) --- not included in this study --- haul out on
-sea-ice but also within snow lairs during winter and spring.
+abundance and trend mean improved aerial survey effort is needed for these
+species. Those survey efforts must also be paired with improved knowledge of
+haul-out behavior to ensure appropriate survey design, robust methods, and
+accurate estimates. Several studies have contributed estimates of the
+distribution and abundance of ice-associated seal species in the Arctic using
+aerial surveys (e.g.,
+@bengtson2005a, @conn2014a, and @verhoef2014a) and more recent efforts have
+significantly expanded on previous survey effort. Such abundance studies are
+conducted over very large areas and estimation of absolute abundance requires
+making inference about numerous issues affecting the observation of seals on
+ice. These include availability (only seals on ice are available to be counted),
+detection probability (observers or automated detection systems may miss some
+seals on ice), species misclassification, and possible disturbance of seals by
+aircraft [@conn2014a; @verhoef2014a]. Refining these inferences will improve the
+accuracy of abundance estimates and, hopefully, allow credible predictions about
+the effects of climate disruptions on the abundance and distribution of Arctic
+seal populations.
+
+How ice-associated seals use sea ice as a haul-out platform varies between
+species. Ribbon seals (_Histriophoca fasciata_) haul out of the water almost
+exclusively on sea ice and are mostly pelagic outside the spring pupping,
+breeding, and molting season [@boveng2018a]. Although spotted (_Phoca largha_)
+and bearded (_Erignathus barbatus_) seals occasionally rest on coastal land,
+they primarily use sea ice as a haul-out platform during the spring and early
+summer [@frost2018]. Ringed seals (_Phoca hispida_) --- not included in this
+study --- haul out on sea-ice but primarily use snow lairs on sea-ice during
+winter and spring.
 
@@
 accounted for variables such as the time-of-day and day-of-year, but did not
-investigate the impact of meteorological variables. Such variables have been
+investigate the impact of weather variables. Such variables have been
 shown to influence walrus haul-out behavior [@udevitz2009] and we expect weather
@@
 2005 and 2020. Our work extends the scope of previous haul-out analyses,
-includes the influence of meteorological variability, and investigates the
+includes the influence of weather variability, and investigates the
 potential impact of changing sea-ice extent on the behavior of these species.
@@
 coastal communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska
-in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. Data from additional
+in the Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental
+Material, [S1](#s1)). Data from additional
 deployments along the Kamchatka peninsula in the western Bering Sea are also
@@
 
-(ref:deployMapCaption) **Initial bio-logger deployment areas for each seal species (bearded, ribbon, and spotted seals) in the study between 2005 and 2020 across multiple regions within the Bering, Chukchi, and western Beaufort seas.** \linebreak Solid regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility.
+(ref:deployMapCaption) **Initial bio-logger deployment areas for bearded, ribbon, and spotted seals in the study between 2005 and 2020 across multiple regions within the Bering, Chukchi, and western Beaufort seas.** \linebreak Solid regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility.
 The larger, shaded region indicates the spatial extent of aerial survey effort to date in the Bering, Chukchi, and Beaufort seas.
-Deployments were initiated across a range of months but only data from 1 March to 15 July were included in the analysis.
+Deployments were initiated across a range of months but only haul-out data from 1 March to 15 July were included in the analysis.
 See Figure \@ref(fig:dataMap) for the spatial distribution of observed data used in the study and Supplemental Materials \@ref(tab:deploymentTable) for additional deployment details.
@@
 
-We subset haul-out behavior data from 
+We created a subset of haul-out behavior data from 
 `r species_n %>% .$n %>% sum()` bio-loggers deployed on 
@@
 Washington, USA). Deployments consisted of either a single 'SPLASH' device, a
-single 'SPOT' device, or both types. Devices were either adhered to the hair
-on the seal or attached through the rear flipper inter-digital webbing. The use
-of bio-loggers adhered to the back or head provides some benefits over flipper
+single 'SPOT' device, or both types. Devices were either adhered to the hair on
+the seal or attached through the rear flipper inter-digital webbing. The use of
+bio-loggers adhered to the back or head provides some benefits over flipper
 mounted devices (e.g. increased satellite transmittal rates, locations at sea)
-but these are lost during the following annual molt, which, depending on
-deployment date, limits the duration of haul-out data they provide.
-Additionally, bio-loggers attached to the head or dorsal region are often dry
-while the seal is floating at the surface, inducing a slight positive bias in
-the hourly percent-dry values reported by the bio-logger. For this study, in
-cases where both bio-logger types were deployed, we preferred hourly percent-dry
-observations from the flipper tag. All data were transmitted by the deployed
-instruments via the Argos satellite network and location data were either
-derived from Argos transmissions or transmitted FastLoc GPS data.
-
-Sex as well as age class (non-dependent _young-of-the-year_, sexually immature
+but these fall off during the following annual molt, which, depending on
+deployment date, limits the duration of haul-out data they provide especially
+during the focus months of our study. Additionally, bio-loggers attached to the
+head or dorsal region are often dry while the seal is floating at the surface,
+inducing a slight positive bias in the hourly percent-dry values reported by the
+bio-logger. For this study, in cases where both bio-logger types were deployed,
+we preferred hourly percent-dry observations from the flipper tag. All data were
+transmitted by the deployed instruments via the Argos satellite network and
+location data were either derived from Argos transmissions or transmitted
+FastLoc GPS data.
+
+Sex and age class (non-dependent _young-of-the-year_, sexually immature
 _subadults_, and mature _adults_) were estimated at the time of deployment by
 various combinations of length, claw growth ridges [@mclaren1958a], and pelage
-characteristics. Seals determined to be less than one year were
-classified as young-of-the-year. For those bio-loggers deployed on
-young-of-the-year and transmitting into the next year, the age class was
-advanced to subadult on 1 March of the following year. Subadults are those seals
-likely greater than one year of age but less than four years. Adults are
-individuals that are likely older than four years. Table \@ref(tab:tblDeploy)
-provides a summary of these deployments and data received from them.
+characteristics. Seals determined to be less than one year were classified as
+young-of-the-year. For those bio-loggers deployed on young-of-the-year and
+transmitting into the next year, the age class was advanced to subadult on 1
+March of the following year -- the assumed anniversary of their birth. Subadults
+are those seals likely greater than one year of age but less than four years.
+Adults are individuals that are likely older than four years. Table
+\@ref(tab:tblDeploy) provides a summary of these deployments and data received
+from them.
 
@@
 
-Tags that fall off due to molt, attachment failure, or seal mortality and remain
-on ice or land may continue to send data to satellites; i.e., a detached
+Tags that fell off due to molt, attachment failure, or seal mortality and remained
+on ice or land may have continued to send data to satellites; i.e., a detached
 bio-logger that is dry (either on ice or land) will record and transmit data
@@
 behavior (if available) to determine when bio-loggers ceased providing data
-consistent with seal behavior. For example, a data record that ends with several
+consistent with seal behavior. For example, a data record that ended with several
 consecutive days (~10+ days) of 100% dry observations and with locations
-indicating the tag was on land would be truncated to not include the final
-stretch of 100% dry observations. The vast majority of deployments end with the
-device detaching in the water and the deployment end date is obvious. There is
+indicating the tag was on land were truncated to the final
+stretch of 100% dry observations. The vast majority of deployments ended with the
+device detaching in the water and the deployment end date was obvious. There is
 no perfect algorithm for identifying deployment end dates and each deployment in
-question must be considered separately. While not perfect, we are confident our
-reliance on expert opinion and examination of multiple data streams provides the
+question was considered separately. While not perfect, we are confident our
+reliance on expert opinion and examination of multiple data streams provided the
 best option. Data outside of the deployment start and end times were discarded
@@
 are transmitted in 24-hour chunks occasional missing days are present in the
-record. Here, missing data that were not successfully recieved from the device
+record. Here, missing data that were not successfully received from the device
 are represented as empty rectangles.
@@
 and portions of the surrounding seas. Weather variables are made available
-across the region 8 times daily. For this study, NARR weather values were subset
+across the region 8 times daily. For this study, NARR weather values were limited
 to the extent of our study area over the Bering, Chukchi, and Beaufort seas at 3-hr
 intervals based on the native grid resolution of 32 km (1024 km^2^). The
-following meteorological variables are known to affect haul-out behavior in
+following weather variables are known to affect haul-out behavior in
 other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017] and were
@@
 
-We did assess whether the annual variation in maximum spring sea ice extent in the
-Bering Sea influenced the seasonal peak of seal haul-out probability. In
-particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
-SSM/I-SSMIS Passive Microwave Dataset, Version 1 [@cavalieri1996] to calculate
-maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
-study area with greater than 15% concentration were counted daily to get the total sea
-ice extent for each day between 15 February and 15 July across all years.
-Maximum spring sea ice extent was simply the largest daily count of grid cells
-with greater than 15% concentration for each year.
-
 (ref:kgm2) kg/m^2^
@@
 Haul-out records for seals are often characterized by sequential hours spent
-basking on ice alternating with long periods in the water (Figure
+hauled out on ice alternating with long periods in the water (Figure
 \@ref(fig:examplePlot)). Commonly used statistical models for binary data (e.g.
@@
 
-To properly account for temporal dependence and to take advantage of computational
-efficiency, we used generalized linear mixed pseudo-models (GLMPMs;
-[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
-covariate predictors, (2) temporally autocorrelated random effects, and (3)
-individual random effects representing heterogeneity in individual behavior. We
-used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
-different model formulations for our data and we fit separate models to bearded, 
-ribbon, and spotted seal data sets as we expected differing behavior by species. 
-Separate models for each species were also
-needed because a single, very large data set proved computationally intractable.
-In our first model formulation and for each species, we fitted a year-independent
-model that predicted average haul-out behavior as a function of demographic,
-environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
-seals (which had considerably more data than bearded seals), we fitted models
-that included all the effects from the first model, but also permitted annual
-variation in haul-out timing. This second class of models was used to examine
-whether haul-out patterns varied by year and to determine the annual timing of
-apparent peaks in haul-out behavior. For both models, we assumed an hourly
-Bernoulli response (i.e., whether tags were mostly dry or mostly wet) where
-the linear predictor was modeled on the logit scale. This is consistent with
-previous approaches [@verhoef2014a, @london2012] and only
+To properly account for temporal dependence and to take advantage of
+computational efficiency, we used generalized linear mixed pseudo-models
+(GLMPMs; [@verhoef2010a]) to model variation in haul-out behavior as a function
+of (1) covariate predictors, (2) temporally autocorrelated random effects, and
+(3) individual random effects representing heterogeneity in individual behavior.
+We used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored
+two different model formulations for our data and we fit separate models to
+bearded, ribbon, and spotted seal data sets as we expected differing behavior by
+species. Separate models for each species were also needed because a single,
+very large data set proved computationally intractable. In our first model
+formulation and for each species, we fitted a year-independent model that
+predicted average haul-out behavior as a function of demographic, weather,
+seasonal, and diurnal effects. Second, for ribbon and spotted seals (which had
+considerably more data than bearded seals), we fitted models that included all
+the effects from the first model, but also permitted annual variation in
+haul-out timing. This second set of models was used to examine whether
+haul-out patterns varied by year and to determine the annual timing of apparent
+peaks in haul-out behavior. For both models, we assumed an hourly Bernoulli
+response (a binary 0-1 response dependent upon whether the tag was greater than 
+50% dry for a given hour) where the linear predictor was 
+modeled on the logit scale. This is
+consistent with previous approaches [@verhoef2014a, @london2012] and only
 `r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
@@
 Visualizing the marginal or conditional effect of an individual weather
-covariate on haul-out probability was difficult in this analysis because of the
-collinearity between covariates as well as the spatial and temporal variation
-across such a large region. The relationship of each weather covariate with
-haul-out probability, averaged over the other weather conditions, was more
-variable than model coefficients would imply. That said,
-important insights can be gained from plots of marginal effects. To create these
-plots, we predicted haul-out probability across the full range
-of weather covariate values while fixing hour of the day at local solar noon and
-day-of-year at 15 May. The
-visualizations also include vertical lines representing 95% confidence intervals
-around the predicted haul-out probability to better communicate the variation in
-model uncertainty.
+covariate (where all other weather covariates are being held at mean values) on
+haul-out probability was difficult in this analysis because of the collinearity
+between covariates as well as the spatial and temporal variation across such a
+large region. The relationship of each weather covariate with haul-out
+probability, averaged over the other weather conditions, was more variable than
+model coefficients would imply. That said, important insights can be gained from
+plots of marginal effects. To create these plots, we predicted haul-out
+probability across the full range of each weather covariate while fixing hour
+of the day at local solar noon and day-of-year at 15 May. For the other weather
+covariate values, we chose not to use a fixed mean value because we expect weather
+to vary within day over the season (e.g. the temperature at solar noon will
+gradually increase from March through June). To account for this, we fit a simple generalized
+additive model for each weather covariate with smooth terms for day-of-year and 
+solar hour. We used predicted values from the generalized additive model in lieu 
+of holding other weather covariates at a fixed mean value which would not capture
+seasonal change. The visualizations
+also include vertical lines representing 95% confidence intervals around the
+predicted haul-out probability to better communicate the variation in model
+uncertainty.
+
+We also assessed whether the annual variation in maximum spring sea ice extent in the
+Bering Sea influenced the seasonal peak of seal haul-out probability. In
+particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
+SSM/I-SSMIS Passive Microwave Dataset, Version 1 [@cavalieri1996] to calculate
+maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
+study area with greater than 15% concentration were counted daily to get the total sea
+ice extent for each day between 15 February and 15 July across all years.
+Maximum spring sea ice extent was simply the largest daily count of grid cells
+with greater than 15% concentration for each year. A separate regression model,
+built on the results of the haul-out model, was used to evaluate the relationship
+between the annual computed peak haul-out day (as the response) with the
+maximum sea ice extent (as the predictor).
 
@@
 on bearded seals. This resulted in fewer data observations overall and
-noticeably lower in numbers May and June. The majority of bearded seal
-deployments started later in the summer and, by May, bio-loggers had either
-fallen off or their batteries were depleted.
+noticeably lower in numbers May and June. The majority of deployments on bearded 
+seals occurred in August and September and, by May, bio-loggers had either
+fallen off or their batteries were depleted. Therefore, observations for bearded 
+seals were concentrated in March (Figure \@ref(fig:dataCal)).
 
@@
 
-(ref:mapCaption) **Spatial distribution of haul-out records from 1 March through July 15 for each of the three species.** \linebreak Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model.
+(ref:mapCaption) **Spatial distribution of haul-out records from 1 March through July 15 for bearded, ribbon, and spotted seal.** \linebreak Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model.
 For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km^2^
@@
 
-(ref:dataDistributionCaption) **Seasonal density of haul-out behavior observations by species** \linebreak Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
+(ref:dataDistributionCaption) **Seasonal distribution of haul-out behavior observations by species** \linebreak Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
 Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020.
-Density values are represented as a color gradient and bar height.
-The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out.
+Seal-hours of data are represented as a color gradient and bar height.
+The higher amount of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out.
 Additionally, many bio-logger deployments started in April and May.
-The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study.
+The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study and the fact that the majority of deployments on bearded seals occurred in August and September; therefore most bio-loggers deployed had exhausted their batteries.
 
@@
 
-Models omitting year effects suggested that day-of-year, solar hour, age-sex
-class, temperature, and wind substantially influenced haul-out behavior of all
-three species, with $F$ tests producing $p$-values less than 0.05 for variables
-embodying these effects and/or their interactions. Haul-out probabilities
-typically increased throughout March and April, reaching a peak in May and early
-June before declining again. Diurnal patterns were present, with maximum
-haul-out behavior centered around local solar noon.
+Across all three seal species, generally, models omitting year effects suggested
+that day-of-year, solar hour, age-sex class, temperature, and wind substantially
+influenced haul-out behavior of all three species, with $F$ tests producing
+$p$-values less than 0.05 for variables embodying these effects and/or their
+interactions. Haul-out probabilities typically increased throughout March and
+April, reaching a peak in May and early June before declining again. Diurnal
+patterns were present, with maximum haul-out behavior centered around local
+solar noon.
 
@@
 shown for all ages (Figure \@ref(fig:beardedHOCal); see also
-\@ref(fig:beardedPredSE)). Additionally, after
-approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
-reported implausible hourly percent-dry data (100% dry for several weeks but
-indicative of movement and increasing transmission rates (see @boveng2013)). All
-data after the first instance of unrealistic values were censored from this
-analysis. Overall, bearded seals were less likely to haul out and had a bi-modal
-distribution of haul-out probability across the day. In addition to a peak
-around local solar noon, the bearded seal model predicted additional haul-out
-activity around local midnight. In concert with the lower magnitude of haul-out
-probability, bearded seal haul-out behavior was also more protracted throughout
-the spring season compared to ribbon and spotted seals.
+\@ref(fig:beardedPredSE)). Additionally, after approximately 9 months, 7 devices
+deployed on the rear flipper of bearded seals reported implausible hourly
+percent-dry data (100% dry for several weeks but indicative of movement and
+increasing transmission rates (see @boveng2013)). All data after the first
+instance of unrealistic values were censored from this analysis. In addition to
+a peak around local solar noon, the bearded seal model predicted additional
+haul-out activity around local midnight. In concert with the lower magnitude of
+haul-out probability, bearded seal haul-out behavior was also more protracted
+throughout the spring season compared to ribbon (Figure \@ref(fig:ribbonHOCal))
+and spotted seals (Figure \@ref(fig:spottedHOCal)). Overall, bearded seals were
+less likely to haul out and had a bi-modal distribution of haul-out probability
+across the day.
 
@@
 
-(ref:beardedHOCalCaption) **Bearded seal predicted haul-out probability** \linebreak Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 30 June for all age and sex classes combined.
+(ref:beardedHOCalCaption) **Bearded seal predicted haul-out probability** \linebreak Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 15 March to 30 June for all age and sex classes combined.
 Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range.
+Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
 Note, our model predicts lower haul-out probability for bearded seals overall compared to ribbon and spotted seals and the color scales are not directly comparable.
@@
 
-(ref:beardedHOwxCaption) **Influence of weather covariates on bearded seal haul-out probability** \linebreak Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
+(ref:beardedHOwxCaption) **Influence of weather covariates on bearded seal haul-out probability** \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
 Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
-Vertical lines represent the 95% confidence interval around the predicted haul-out probability at observed weather values.
+The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
+While the marginal effect of the covariate is continuous, points and vertical lines representing the 95% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.
 
@@
 
-Ribbon seals exhibited a pattern of gradually increasing haul-out probability
-in April that peaks in late May for subadults and in early June for
-adults (Figure \@ref(fig:ribbonHOCal); see also 
-\@ref(fig:ribbonPredSE)). The behavior was clearly centered
-around local solar noon and expanded to other hours later in the spring
-as seals entered their molting period. Subadults showed an earlier start
-and more intense haul-out activity in April and May. The
-young-of-the-year records begin after weaning and the model predictions
-demonstrated the development of in-water activities (e.g. swimming, foraging)
-in May. Adult females had a more protracted haul-out season compared to
-males, and more time was spent hauled out in June compared to adult males and
-subadults.
+Ribbon seals exhibited a pattern of gradually increasing haul-out probability in
+April that peaks in late May for subadults and in early June for adults (Figure
+\@ref(fig:ribbonHOCal); see also \@ref(fig:ribbonPredSE)). There is an apparent
+seasonal progression with subadults hauling out earlier in the season followed
+by adult males and, then, adult females. Haul-out behavior was clearly centered
+around local solar noon and expanded to other hours later in the spring as seals
+entered their molting period. Subadults showed an earlier start and more intense
+haul-out activity in April and May. The young-of-the-year records began after
+weaning and the model predictions seemed to indicate development of in-water
+activities (e.g. swimming, foraging) in May. Adult females had a more protracted
+haul-out season compared to males, and more time was spent hauled out in June
+compared to adult males and subadults.
 
 (ref:ribbonHOCalCaption) **Ribbon seal predicted haul-out probability** \linebreak Predicted hourly haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
-There is an apparent seasonal progression with subadults hauling out earlier in the season followed by adult males and, then, adult females.
+Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
 Predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities.
@@
 `r ribbon_type3$pressure$Prob.F`) having a milder impact.
-Ribbon seals were less likely to haul out at higher winds and lower pressure
-values and more likely to haul out when temperatures were
-relatively warm (Figure \@ref(fig:ribbonHOwx)). Neither wind chill
+Ribbon seals were more likely to haul out when temperatures were
+relatively warm and less likely to haul out at higher winds and lower pressure
+values (Figure \@ref(fig:ribbonHOwx)). Neither wind chill
 ($F_{`r paste(ribbon_type3$temp2m_wind$Num.df, ribbon_type3$temp2m_wind$Den.df, sep=",")`}$
@@
 visualize the different influences of weather covariates on those classes and see
-that subadults separate out from adult males and females (Figure \@ref(fig:ribbonHOwx)).
-
-(ref:ribbonHOwxCaption) **Influence of weather covariates on ribbon seal haul-out probability** \linebreak Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classifications.
+that subadults differ from adult males and females (Figure \@ref(fig:ribbonHOwx)).
+
+(ref:ribbonHOwxCaption) **Influence of weather covariates on ribbon seal haul-out probability** \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classifications.
 Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
-Vertical lines represent the 95% confidence interval around the predicted haul-out probability at observed weather values.
+The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
+While the marginal effect of the covariate is continuous, points and vertical lines representing the 95% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.
 
@@
 see also \@ref(fig:spottedPredSE)). Adults of both sexes spent considerable time
-in April, May and June hauled out. In contrast to ribbon seals, adult spotted
-seal males had a more protracted haul-out season compared to females, and more
-time out of the water in June (Figures \@ref(fig:ribbonHOCal) and
-\@ref(fig:spottedHOCal)). As with ribbon seals, the young-of-the-year records
-began after weaning and the model predictions demonstrated the development of
-in-water activities (e.g. swimming, foraging) in May.
+in April, May and June hauled out. Adult spotted seal males had a more
+protracted haul-out season compared to females, and more time out of the water
+in June (Figure \@ref(fig:spottedHOCal)). This likely reflects the triad
+behavior in spotted seals when suitor males haul out with nursing females. As
+with ribbon seals, the young-of-the-year records began after weaning and the
+model predictions reflected development of in-water activities (e.g.
+swimming, foraging) in May.
 
 (ref:spottedHOCalCaption) **Spotted seal predicted haul-out probability** \linebreak Predicted hourly haul-out probability of spotted seals from 1 March to 30 June for each age and sex class used in the model.
-Compared to ribbon seals, the period of increased haul-out behavior is broader and there is more alignment in the seasonal timing of haul-out behavior across age and sex classifications.
-This likely reflects the triad behavior in spotted seals when suitor males haul out with nursing females.
-As with ribbon seals, redictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities.
+Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
+As with ribbon seals, predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities.
 Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range.
@@
 = `r spotted_type3$pressure$F.value`; $p$ =
-`r spotted_type3$pressure$Prob.F`) were all significant. Spotted seals were less likely to
-haul out at higher winds and more likely to be on the ice when
-temperatures were relatively warm. Wind chill 
+`r spotted_type3$pressure$Prob.F`) were all significant. Spotted seals were more likely to be on the ice when
+temperatures were relatively warm and less likely to
+haul out at higher wind speeds. Wind chill 
 ($F_{`r paste(spotted_type3$temp2m_wind$Num.df, spotted_type3$temp2m_wind$Den.df, sep=",")`}$
@@
 
-(ref:spottedHOwxCaption) **Influence of weather covariates on spotted seal haul-out probability** \linebreak Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
+(ref:spottedHOwxCaption) **Influence of weather covariates on spotted seal haul-out probability** \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
 Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
-Vertical lines represent the 95% confidence interval around the predicted haul-out probability.
+The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season..
+While the marginal effect of the covariate is continuous, points and vertical lines representing the 95% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.
 
@@
 patterns, uncovered significant contributions for linear and quadratic
-interactions between day and year in both ribbon seals (day:year,
+interactions between day and year for only spotted seals (day:year,
+$F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$
+= `r spotted_year_type3$day_year$F.value`; $p$ =
+`r spotted_year_type3$day_year$Prob.F`; day^2^:year,
+$F_{`r paste(spotted_year_type3$day_2_year$Num.df, spotted_year_type3$day_2_year$Den.df, sep=",")`}$
+= `r spotted_year_type3$day_2_year$F.value`; $p$ =
+`r spotted_year_type3$day_2_year$Prob.F`). Ribbon seals showed no significant
+contribution for interactions between day and year (day:year,
 $F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$
@@
 = `r ribbon_year_type3$day_2_year$F.value`; $p$ =
-`r ribbon_year_type3$day_2_year$Prob.F`) and spotted seals (day:year,
-$F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$
-= `r spotted_year_type3$day_year$F.value`; $p$ =
-`r spotted_year_type3$day_year$Prob.F`; day^2^:year,
-$F_{`r paste(spotted_year_type3$day_2_year$Num.df, spotted_year_type3$day_2_year$Den.df, sep=",")`}$
-= `r spotted_year_type3$day_2_year$F.value`; $p$ =
-`r spotted_year_type3$day_2_year$Prob.F`). Predicted distributions of
+`r ribbon_year_type3$day_2_year$Prob.F`). Predicted distributions of
 haul-out activity were largely unimodal, but varied some among and
@@
   bind_rows(peak_ho_df2) %>% ungroup() %>% 
-  mutate(year = forcats::fct_relevel(.$year, sort))
+  mutate(year = forcats::fct_relevel(.$year, sort)) %>%
+  mutate(year_shape = if_else(as.numeric(paste(year)) < 2015,"a","b"))
 
 pal <- c(rcartocolor::carto_pal(12,'Bold')[1:11],
-         rcartocolor::carto_pal(12,'Antique')[1:11])
+         rcartocolor::carto_pal(12,'Antique')[1:3])
+names(pal) <- as.character(sort(unique(peak_ho_df$year)))
+
+shapes <- c(rep(15,7), rep(19,7))
+names(shapes) <- as.character(sort(unique(peak_ho_df$year)))
 
@@
       ymax = ho_prob,
-      color = year
-    ),
-    shape = 15
+      color = year,
+      shape = year
+    )
   ) +
-  scale_color_discrete(type = pal) +
+  scale_color_manual(name = "Annual timing of peak haul-out probability",
+                     values = pal) +
   scale_x_continuous(
@@
   scale_y_continuous(breaks = c(0.2,0.4,0.6,0.8), labels = c("0.2","0.4", "0.6", "0.8")) +
+  scale_shape_manual(name = "Annual timing of peak haul-out probability",
+                     values = shapes) +
   guides(col = guide_legend(ncol = 7, title.position = "top",
-                            title.hjust = 0,
-                            title = "Annual timing of peak haul-out probability")) +
+                            title.hjust = 0)) +
   facet_grid(age_sex ~ species,
@@
   ylab("haul-out probability") +
-  labs(caption = "predictions are shown for local solar noon and under smoothed weather conditions") +
+  labs(caption = "predictions are shown for local solar noon and under mean weather conditions") +
   theme(legend.position = "top",
@@
 We modeled data from bio-loggers deployed on bearded, ribbon, and spotted seals
-to examine factors affecting haul-out behavior on sea ice in the Bering and
-Chukchi seas. Our analysis shows all three species of seal haul out
+to examine factors affecting their haul-out behavior on sea ice in the Bering
+and Chukchi seas. Our analysis shows all three species of seal haul out
 progressively more through the spring and peak near mid-May to early June before
@@
 our haul-out data are likely quantifying population-level behavioral patterns.
-Seals preferentially haul out on ice shortly after solar noon, which allows
-seals to maximize absorption of solar radiation thought to facilitate the
-molting process [@feltz1966]. Interestingly, bearded seals appear to have two
-peaks in haul-out activity within a day, one shortly after solar noon, and one
-centered near solar midnight. This, of course, could be an artifact of our
-limited sample size for bearded seal deployments across all age classes.
-However, a similar bi-modal pattern has been seen in ringed seals
+Beginning in spring, seals preferentially haul out on ice shortly after solar
+noon, which allows seals to maximize absorption of solar radiation thought to
+facilitate the molting process [@feltz1966]. Interestingly, bearded seals appear
+to have two peaks in haul-out activity within a day, one shortly after solar
+noon, and one centered near solar midnight. This, of course, could be an
+artifact of our limited sample size for bearded seal deployments across all age
+classes. However, a similar bi-modal pattern has been seen in ringed seals
 [@vonduyke2020] and suggests that bearded and ringed seals may be operating
 under different constraints than ribbon and spotted seals. Bearded and ringed
-seals are distributed across higher latitudes and the extended daylight hours
-may allow more flexibility in optimizing resting periods with foraging. Other
-factors such as predation by polar bears (which is rare for ribbon and spotted
-seals in the Bering Sea) may also explain differing haul-out patterns. The
-change in haul-out behavior during the season was less pronounced in bearded
-seals compared to ribbon and spotted seals. This aligns with findings from
-Thometz et al. [@thometz2021] who observed a mean molting period of 119&pm;2
-days and a relatively stable resting metabolic rate during that time. While
-ribbon seals were not considered in that same study, spotted and ringed seals
-underwent molt periods of just 33&pm;4 and 28&pm;6 days and had increased
-resting metabolic rates.
+seals are distributed across higher latitudes higher latitudes that experience
+extended daylight hours during spring which may allow more flexibility in
+alternating resting and foraging events. Other factors such as predation by
+polar bears (which is rare for ribbon and spotted seals in the Bering Sea) may
+also explain differing haul-out patterns. The change in haul-out behavior during
+the season was less pronounced in bearded seals compared to ribbon and spotted
+seals. This aligns with findings from Thometz et al. [@thometz2021] who observed
+a mean molting period of 119&pm;2 days and a relatively stable resting metabolic
+rate for bearded seals during that time. While ribbon seals were not considered
+in that study, spotted and ringed seals underwent molt periods of just 33&pm;4
+and 28&pm;6 days and had increased resting metabolic rates.
 
@@
 [-@udevitz2009]) and a few select studies of ice-associated seals
-[@hamilton2018; @perry2017]. In our study, ribbon seals seemed to be the most
+[@hamilton2018; @perry2017]. In our study, ribbon seal haul-out behavior seemed to be the most
 influenced by weather, with wind, temperature, and barometric pressure all being
@@
 
-Notably missing from our list of explanatory variables is any spatial-temporal
-representation of sea ice concentration, area, or extent. This may seem
-counterintuitive when modeling the haul-out behavior of seal species with such a
-close association to sea ice; seals haul out in the presence of sea ice, and we
-could assess the local concentration of sea ice during these events (see
-[@olnes2020]). This, however, expands the scope of our analysis into the realm
-of habitat selection and many of our deployments consisted of a single device
-attached to the rear flipper of the seal which did not provide at-sea locations,
-limiting our ability to fully evaluate fine-scale habitat preferences related to
-sea-ice. Insight into how seals use and interact with sea-ice during an extended
-period when the availability and characteristics of sea-ice is rapidly changing
-is important but ancillary to the focus of this analysis. The focus of this
-study was to develop models applicable for aerial survey correction factors and
-using sea ice as a covariate would almost certainly bias haul-out predictions
-towards those seals that are on or near ice and therefore more likely to haul
-out. Since aerial surveys can only detect seals on ice, abundance estimates
-would be missing a correction for those seals that are away from ice.
-
-Consider
-the case of employing haul-out probabilities as an aerial survey correction in a
-population inhabiting two areas: an ice-free region and a region with sea ice.
-Further, assume that half of the population is in each area on average, and that
-the probability of hauling out is 1.0 in the ice covered region and 0.0 in the
-region without ice.  Denoting the total population as $N$, let us examine what
-happens when we use (1) population level availability, and (2) ice-specific
-availability as a correction factor.  We'll further assume, for sake of this
-example, that detection probability of seals on ice is 1.0, and that the
-entirety of the ice covered region is censused. In case (1), our aerial survey
-count is $C = 0.5 \cdot N$, with a population-level availability probability of
-$\hat{a} = 0.5$. A Horvitz-Thompson-type estimator for abundance is simply 
-$\hat{N} = \frac{C}{\hat{a}}$, which has expectation 
-$\mathbb{E}(\hat{a}) = \frac{0.5 \cdot N}{0.5} = N$, as desired.  Now
-consider case (2).  In this case, the probability of hauling out is 1.0 for
-seals in the ice-covered area, so we have 
-$\mathbb{E}(\hat{N}) = \frac{0.5 \cdot N}{1.0} = 0.5N$. That
-is, we bias abundance because we are not accounting for seals that are away from
-ice, and therefore have a zero chance of hauling out. The same logic holds for
-any other covariate, such as distance from the ice edge, that has different
-values in the surveyed and unsurveyed habitat of the seals (assuming distance
-from ice edge is coded with opposite signs inside and outside the pack ice).
-
-Lastly, our study was limited to the spring season when seal haul-out behaviors
-are strongly influenced by pupping, nursing, breeding behavior, and molt and
-these drivers are likely more influential than specific sea-ice concentration.
+Notably missing from our list of haul-out model explanatory variables is any
+spatial-temporal representation of sea ice concentration, area, or extent. This
+may seem counterintuitive when modeling the haul-out behavior of seal species
+with such a close association to sea ice; seals haul out in the presence of sea
+ice, and we could assess the local concentration of sea ice during these events
+(see [@olnes2020]). This, however, expands the scope of our analysis into the
+realm of habitat selection and many of our deployments consisted of a single
+device attached to the rear flipper of the seal which meant we only received
+locations when seals were hauled out on sea-ice, limiting our ability to fully
+evaluate fine-scale habitat preferences related to sea-ice. Insight into how
+seals use and interact with sea-ice during an extended period when the
+availability and characteristics of sea-ice is rapidly changing is important
+[@breed2018; @cameron2018] but ancillary to the focus of this analysis -- and,
+in the end, not possible given key limitations of our data. Additionally, our
+study was limited to the spring season when seal haul-out behaviors are strongly
+influenced by pupping, nursing, breeding behavior, and molt. These drivers are
+likely more influential on haul-out behavior than sea-ice concentration.
 Crawford et al. [-@crawford2019] compared haul-out probability models for ringed
 seals and found those that only included season (and not sea-ice concentration)
-were the most parsimonious. For these reasons, we have elected not to use sea
-ice concentration _as a predictor for haul-out probability_ in the present
+were the most parsimonious. For all of these reasons, we have elected not to use
+sea ice concentration _as a predictor for haul-out probability_ in the present
 study.
@@
 currently lack sufficient contrast in ice extent to characterize an effect on
-haul-out probability. We should expect, however, that some minimal threshold in
+the timing of peak haul-out probability. We should expect, however, that some minimal threshold in
 the spatial extent or density of sea ice will have a meaningful impact on the
-timing of peak haul-out behavior --- if there is no sea ice, seals will not haul
-out or be forced to use terrestrial haul-out sites which were not part of the
+timing of peak haul-out behavior --- if there is no sea-ice, seals will not haul
+out or they will be forced to use terrestrial haul-out sites which were not part of the
 evolution of their normal behaviors. Additionally, while from an
@@
 day-of-year and time-of-day, they have not accounted for variability among
-years, in weather conditions, and in the age-sex class of the sample. In this
-paper, we have shown that there can be considerable differences in the number of
-seals hauled out on ice based on these factors and subsequent analyses have shown
-the potential for considerable bias in abundance estimates if such covariates
-are unaccounted for [@conn2023]. We recommend that future
-abundance analyses employ availability models that account for them. For
-instance, it is relatively straightforward to obtain weather reanalysis products
-for times and locations that are surveyed and to construct a relevant correction
-factor based on predictions of GLMPMs. The most challenging element in
-developing availability correction factors is with annual variability. It can be
-difficult to get a sufficient sample size to estimate year-specific correction
-factors, particularly because research teams would likely need to tag seals and
-conduct aerial surveys concurrently, requiring considerably more personnel
-and money. One possible suggestion is to estimate a "shift" parameter within
-models for aerial survey counts that allow the peak of haul-out distributions to
-be adjusted earlier or later in the year based on the frequency of counts
-observed over time. Regardless, researchers should anticipate there being some
-unmodeled heterogeneity in availability when using aerial surveys to estimate
-Arctic seal abundance. This may require consideration in trend estimation,
-as one will not know if moderate differences in abundance estimates are
+years, weather conditions, or in the age-sex class of the sample. In this paper,
+we have shown that there can be considerable differences in the haul-out
+probability of seals on ice based on these factors and subsequent analyses have
+shown the potential for considerable bias in abundance estimates if such
+covariates are unaccounted for [@conn2023]. We recommend that future abundance
+analyses employ availability models that account for them. For instance, it is
+relatively straightforward to obtain weather reanalysis products for times and
+locations that are surveyed and to construct a relevant correction factor based
+on predictions of GLMPMs. The most challenging element in developing
+availability correction factors is with annual variability. It can be difficult
+to get a sufficient sample size to estimate year-specific correction factors,
+particularly because research teams would likely need to deploy bio-loggers on
+seals and conduct aerial surveys concurrently, requiring considerably more
+personnel and money. One possible suggestion is to include year as a random
+effect within models for aerial survey counts such that, without specific
+knowledge of any particular year, the among-year variance is included in the
+modeled standard errors. Regardless of the specific approach, future estimates
+of Arctic seal abundance will require specific consideration of annual
+variability and changes in the timing of peak haul-out behavior when estimating
+trends, as one will not know if moderate differences in abundance estimates are
 attributable to changes in abundance or changes in haul-out behavior.
@@
 Our analysis emphasizes the importance of accounting for seasonal and temporal
-variation in haul-out behavior, as well as associated environmental covariates,
+variation in haul-out behavior, as well as associated weather covariates,
 when interpreting the number of seals counted in aerial surveys. The rapid
@@
 forecast to decrease the quantity and quality of suitable habitat for
-ice-associated seals that rely on sea ice for a variety of activities. Improved
+ice-associated seals that rely on sea-ice for a variety of activities. Improved
 estimates of seal population abundance from aerial surveys are needed to
@@
 Funding to ADFG for tagging seals was provided by the Bureau of Ocean Energy
-Management (No. M13PC0015) and the Office of Naval Research (No.
-N00014-16-1-3019). ADFG and NSB field work was covered by Research Permits Nos.
-358-1585, 358-1787, 15324, and 20466 and by ADF&G IACUC permits Nos. 06-16,
-09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27, 0027-2018-29, 0027-2019-041.
+Management (No. M13PC0015 for work in Kotzebue, Alaska in 2009) and the Office
+of Naval Research (No. N00014-16-1-3019). ADFG and NSB field work was covered by
+Research Permits Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF&G IACUC
+permits Nos. 06-16, 09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27,
+0027-2018-29, 0027-2019-041.
 
@@
 
-```{r deploymentTable, include=TRUE, tab.cap="The following table provides additional details about the timing, location, and institutions responsible for the bio-logger deployments used in this study along with research permits and any associated publications."}
+```{r deploymentTable, include=TRUE, tab.cap="The  timing, location, and institutions responsible for the bio-logger deployments used in this study along with research permits and any associated publications."}
 library(gt)
@@
     sex = "Sex",
-    age_class = "Age_Class"
+    age_class = "Age Class"
   ) %>%
@@
 
-During the peer review process for this manuscript, Anthony Fishbach suggested
+During the peer review process for this manuscript, Anthony Fischbach suggested
 the possibility of using predicted insolation (or solar radiation) values from

### 2024-02-17 analysis/paper/London_HauloutBeahvior.Rmd
lori q comments addressed

@@
   and spotted seals (_Phoca largha_) in the Bering and Chukchi seas. Because
-  ringed seals (_Phoca hispida_) can be out of the water but hidden from view in
+  ringed seals (_Phoca (pusa) hispida_) can be out of the water but hidden from view in
   snow lairs analysis of their 'availability' to surveys requires special
@@
 
-The global climate disruption is causing considerable reduction in Arctic sea
+Global climate disruption is causing considerable reduction in Arctic sea
 ice extent, volume, and seasonal presence [@kwok2018;
 @meier2014; @overland2021; @wang2018]. These changes have tangible effects on
-Arctic organisms, ecosystems, and the human communities who live in the region
+Arctic organisms, ecosystems, and the people who live in the region
 [@huntington2020]. Such disruptions are a particular cause of concern for the
@@
 lacking, so it is difficult to predict how or if ice-associated seal species
-adapt to future changes (e.g., by adjusting pupping or molting schedules to
+will adjust to future changes (e.g., by adjusting pupping or molting schedules to
 earlier dates or different locales). This is further complicated by the
@@
 
-How ice-associated seals use sea ice as a haul-out platform varies between
+How ice-associated seals use sea ice as a haul-out platform varies among
 species. Ribbon seals (_Histriophoca fasciata_) haul out of the water almost
@@
 they primarily use sea ice as a haul-out platform during the spring and early
-summer [@frost2018]. Ringed seals (_Phoca hispida_) --- not included in this
+summer [@frost2018]. Ringed seals (_Phoca (pusa) hispida_) --- not included in this
 study --- haul out on sea-ice but primarily use snow lairs on sea-ice during
@@
 events occurred between 2005 and 2020 from May to October, generally between the
-coastal communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska
+Alaska coastal communities of Scammon Bay in the Bering Sea, Utqiaġvik
 in the Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental
@@
 
-(ref:deployTblCaption) **Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2021.** \linebreak Total seal hours represents the sum of hourly observations across all seals used in the analysis.
+(ref:deployTblCaption) **Summary of bio-logger data by seal species and age classification from 1 March to 15 July 2005-2021.** \linebreak Seal hours represents the sum of hourly observations across all seals used in the analysis.
 Because young-of-the-year are advanced to subadult on 1 March of the following year, some individual seals are represented in both columns in this table
@@
 actogram plot represents the transmitted hourly percent-dry values. Not all data
-during a deployment are reliably transmitted from the bio-logger and since data
+during a deployment are reliably transmitted from the bio-logger and because data
 are transmitted in 24-hour chunks occasional missing days are present in the
@@
 to the extent of our study area over the Bering, Chukchi, and Beaufort seas at 3-hr
-intervals based on the native grid resolution of 32 km (1024 km^2^). The
+intervals based on the original grid resolution of 32 km (1024 km^2^). The
 following weather variables are known to affect haul-out behavior in
@@
 age-sex classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a
-latitudinal effect (and an interaction with day-of-year), since bearded seals
+latitudinal effect (and an interaction with day-of-year), because bearded seals
 occupy a substantial range during the spring and we were interested in possible
@@
 modeled on the logit scale. This is
-consistent with previous approaches [@verhoef2014a, @london2012] and only
+consistent with previous approaches [@verhoef2014a; @london2012] and only
 `r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
@@
 biology’ of Arctic seals. The deployment of bio-logging devices used in this
-study were often done in collaboration with Alaska Native seal hunters near
+study were often done in collaboration with Alaska Native seal hunters and the
+approval of
 their communities. We would like to especially acknowledge the communities of
@@
 Virgil Naylor Sr., Dan Savetilik, Chuck Schaeffer, Ross Schaeffer, Allen Stone,
-and Randy Toshavik from Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Tom
-Gray from Nome; Vernon Long and Richard Tukle from Nuiqsuit, Alaska; Morgan
-Simon, River Simon, and Al Smith from Scammon Bay, Alaska; Alex Niksik Jr. from
-St. Michael, Alaska; Billy Adams, James Aiken, Tim Aiken, Howard Kittick,
+and Randy Toshavik from Kotzebue; Merlin Henry from Koyuk; Tom
+Gray from Nome; Vernon Long and Richard Tukle from Nuiqsuit; Morgan
+Simon, River Simon, and Al Smith from Scammon Bay; Alex Niksik Jr. from
+St. Michael; Billy Adams, James Aiken, Tim Aiken, Howard Kittick,
 Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, and Joe Skin from Utqiaġvik,
 Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, and
-Fred Rexford from Ulguniq, Alaska.
+Fred Rexford from Ulguniq.
 
@@
 
-Telemetry data from the Alaska Department of Fish and Game (ADFG) and the North
+Telemetry data from the Alaska Department of Fish and Game (ADF&G) and the North
 Slope Borough Department of Wildlife Management (NSB) were important
@@
 Fisheries Science Center Institutional Animal Care and Use Committee (IACUC). ).
-Funding to ADFG for tagging seals was provided by the Bureau of Ocean Energy
+Funding to ADF&G for tagging seals was provided by the Bureau of Ocean Energy
 Management (No. M13PC0015 for work in Kotzebue, Alaska in 2009) and the Office
-of Naval Research (No. N00014-16-1-3019). ADFG and NSB field work was covered by
+of Naval Research (No. N00014-16-1-3019). ADF&G and NSB field work was covered by
 Research Permits Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF&G IACUC
@@
 \@ref(fig:beardedHOCal), \@ref(fig:ribbonHOCal), and \@ref(fig:spottedHOCal) but
-selected for three to local solar hours (07:00, 12:00, and 17:00) so the 
+selected for three local solar hours (07:00, 12:00, and 17:00) so the 
 confidence intervals can also be shown and comparisons can be made.
@@
 notion being that seals are, likely, directly responding to changes in solar radiation
-throughout the day and not what time of day it is (i.e. seals don't have
+throughout the day and not what time of day it is (i.e. seals don't have human
 watches). Additionally, given the energetic benefits of increased solar

### 2024-02-18 analysis/paper/Letter_Response_to_Reviewers.qmd
final commit for revision

@@
 
-J
 ```
-osh M. London, PhD\newline
-7600 Sand Point Way NE\newline Seattle, Washington, 98115\newline josh.london\@noaa.gov\newline
+
+Josh M. London, PhD\newline
+7600 Sand Point Way NE\newline Seattle, Washington,
+98115\newline josh.london\@noaa.gov\newline
 
@@
 
-First, please accept my apologies for the significant delay in returning revisions on this manuscript.
-A number of life and work priorities interfered with a more timely response schedule and I sincerely appreciate the flexibility *PeerJ* extends to authors.
-We are commited to a quick turnaround on any future revisions or tasks associated with this manuscript and we look forward to a final publication in *PeerJ*.
-
-We feel, while many of the comments and suggestions from reviewers were minor, they deserved appropriate time, consideration, and exploration.
-We hope our responses here reflect that consideration and we are grateful to the reviewers for their thoughtful comments.
-Lastly, we also used this opportunity to improve the quality of the underlying telemetry data set and expand the reproducibility of the analysis to meet or exceed expectations for open science, data availability, and code availability.
-
-Below, please find detailed responses to each of the reviewer comments.
-In addition to this letter, several of the suggestions are documented as specific issues within the GitHub repository and the full suite of updates are all contained and documented within a single pull request.
-We have also provided a pdf document that indicates all changes between the revised manuscript and the submitted manuscript.
-
-Thank you, again, for your time and consideration. And, additionally, thank you to the reviewers for their time. It is much appreciated and we feel the manuscript is a better product as a result.
+First, please accept my apologies for the significant delay in returning
+revisions on this manuscript. A number of life and work priorities interfered
+with a more timely response schedule and I sincerely appreciate the flexibility
+*PeerJ* extends to authors. We are commited to a quick turnaround on any future
+revisions or tasks associated with this manuscript and we look forward to a
+final publication in *PeerJ*.
+
+While many of the comments and suggestions from reviewers were minor, we feel
+they deserved appropriate time, consideration, and exploration. We hope our
+responses here reflect that consideration and we are grateful to the reviewers
+for their thoughtful comments. Lastly, we also used this opportunity to improve
+the quality of the underlying telemetry data set and expand the reproducibility
+of the analysis to meet or exceed expectations for open science, data
+availability, and code availability.
+
+Below, please find detailed responses to each of the reviewer comments. In
+addition to this letter, several of the suggestions are documented as specific
+issues within the GitHub repository and the full suite of updates are all
+contained and documented within a single pull request. We have also provided a
+pdf document that indicates all changes between the revised manuscript and the
+submitted manuscript.
+
+Thank you, again, for your time and consideration. And, additionally, thank you
+to the reviewers for their time. It is much appreciated and we feel the
+manuscript is a better product as a result.
 
@@
 
-Here, we respond directly to each of the suggested edits and comments.
-These responses are organized by reviewer.
-All responses are prefaced with **\[author response\]** and line numbers in the original submission are used as reference.
+Here, we respond directly to each of the suggested edits and comments. These
+responses are organized by reviewer. All responses are prefaced with **\[author
+response\]** and line numbers in the original submission are used as reference.
 \hfill \break
@@
 
-## **Reviewer 1 (A. Fishbach)**
-
-> *The authors have put substantial efforts into setting the context for their analysis, citing relevant studies.
-I would like to see a more direct approach in the framing of their problem: the improvements of abundance estimates required by the Marine Mammal Protection Act and Endangered Species Act.
-I would like to see a specific example of how the modeled availability would have affect population abundance estimates by apply the availability model to aerial survey data and providing an adjusted abundance estimate.*
-
-::: author-response
-**\[author response\]** We address this in the manuscript by introducing the importance of the statutory requirements of the MMPA and ESA early in the second paragraph of the introduction.
-We hope this frames the important link in the mind of the reader between the work described here and the need for improved abundance estimates.
-
-Regarding the desire for specific examples of how the modeled availability would have impacted population abundance estimates from aerial survey data, we don't have the ability to provide such an example because the aerial survey data are not yet published and available.
-Such an example and discussion might be better suited for subsequent papers that focus on presentation of the aerial survey data and the abundance estimation process.
-:::
-
-> *The findings are appear robust and intuitive. The authors need to improve the presentation of the findings to improve clarity and help the reader understand how they addressed the knowledge gap.*
-
-::: author-response
-**\[author response\]** The reviewer provided numerous comments and suggestions annotated throughout the PDF that address the concern of presentation and clarity.
-In the next section we provide responses to those annotations
-:::
-
-### **Reviewer 1 (A. Fishbach) - PDF Annotated Comments & Suggested Edits**
-
-> ***Title:** The reviewer suggests removing and aerial survey detectability from the title of the manuscript and suggests adding the seasonal time period our study focused on.*
-
-::: author-response
-**\[author response\]** We have changed the title to: Spring haul-out behavior of seals in the Bering and Chukchi seas
-:::
-
-> ***Abstract:** Can you make a statement in your abstract that use of modeled haulout behavior using covariates available from regional reananalysis data improved the confidence intervals on availability estimates and that this would improve aerial survey abundance estimation efforts?*
-
-::: author-response
-**\[author response\]** We don't feel comfortable making a precision comparison with previous abundance estimates, simply because there are too many differences between the analyses, including (1) the data sets are different (the present study has substantially more records), and (2) the model structures are different (the present study has more parameters).
-As such, it will be difficult to ascribe a specific reason to precision increases or decreases.
-:::
-
-> *Please use direct language.
-You are not making findings on how seals rely on sea ice, rather you are improving methods to estimate their availability for improved aerial survey population estimation efforts.
-Here you are focused on modeling ice sea availability to aerial survey so that aerial survey counts may be appropriately interpreted.
-Please be clear that you are considering ice seal behaviors that may be telemtered as "dry out of water behaviors" that may be interpreted as behaviors that place the seal on top of the ice and available for observation by an aerial survey.*
-
-::: author-response
-**\[author response\]** The opening sentence of the abstract is a statement regarding the reliance of ice-associated seals on sea-ice for a range of important life history behaviors.
-We think this statement is an important introduction for readers who are not familiar with ice-associated seals.
-We do agree with the reviewer's suggestion for more direct language and have made edits to the beginning of the abstract with this in mind.
-:::
-
-> ***Abstract:** Did you use previously published data, or did you collect it?  Be specific using an active voice so that readers may understand your effort.*
+## **Reviewer 1 (A. Fischbach)**
+
+> *The authors have put substantial efforts into setting the context for their
+> analysis, citing relevant studies. I would like to see a more direct approach
+> in the framing of their problem: the improvements of abundance estimates
+> required by the Marine Mammal Protection Act and Endangered Species Act. I
+> would like to see a specific example of how the modeled availability would
+> have affect population abundance estimates by apply the availability model to
+> aerial survey data and providing an adjusted abundance estimate.*
+
+::: author-response
+**\[author response\]** We address this in the manuscript by introducing the
+importance of the statutory requirements of the MMPA and ESA early in the second
+paragraph of the introduction. We hope this frames the important link in the
+mind of the reader between the work described here and the need for improved
+abundance estimates.
+
+Regarding the desire for specific examples of how the modeled availability would
+have impacted population abundance estimates from aerial survey data, we don’t
+have the ability to provide a full example because the aerial survey data are
+not yet published and available. However, we note that Conn and Trukhanova
+(2023, Marine Mammal Science 39:648-661) used some of our preliminary estimates
+to assess potential aerial survey bias in using telemetered samples that are
+biased towards younger age-sex classes. Their analysis is based on bias(N) =
+bias(1/a) where a is availability (i.e., proportion hauled out).
+:::
+
+> *The findings are appear robust and intuitive. The authors need to improve the
+> presentation of the findings to improve clarity and help the reader understand
+> how they addressed the knowledge gap.*
+
+::: author-response
+**\[author response\]** The reviewer provided numerous comments and suggestions
+annotated throughout the PDF that address the concern of presentation and
+clarity. In the next section we provide responses to those annotations
+:::
+
+### **Reviewer 1 (A. Fischbach) - PDF Annotated Comments & Suggested Edits**
+
+> ***Title:** The reviewer suggests removing and aerial survey detectability
+> from the title of the manuscript and suggests adding the seasonal time period
+> our study focused on.*
+
+::: author-response
+**\[author response\]** We have changed the title to: Spring haul-out behavior
+of seals in the Bering and Chukchi seas
+:::
+
+> ***Abstract:** Can you make a statement in your abstract that use of modeled
+> haulout behavior using covariates available from regional reananalysis data
+> improved the confidence intervals on availability estimates and that this
+> would improve aerial survey abundance estimation efforts?*
+
+::: author-response
+**\[author response\]** Unfortunately, we can't make this statement because 
+inclusion of additional covariates (such as weather products) in haul-out models
+only serves to increase standard errors in predictions.  The benefit of
+including weather covariates is to more accurately explain variation in survey
+counts; it is thus more relevant to reducing bias in aerial survey estimates
+than it is to increasing precision.
+:::
+
+> *Please use direct language. You are not making findings on how seals rely on
+> sea ice, rather you are improving methods to estimate their availability for
+> improved aerial survey population estimation efforts. Here you are focused on
+> modeling ice sea availability to aerial survey so that aerial survey counts
+> may be appropriately interpreted. Please be clear that you are considering ice
+> seal behaviors that may be telemtered as "dry out of water behaviors" that may
+> be interpreted as behaviors that place the seal on top of the ice and
+> available for observation by an aerial survey.*
+
+::: author-response
+**\[author response\]** The opening sentence of the abstract is a statement
+regarding the reliance of ice-associated seals on sea-ice for a range of
+important life history behaviors. We think this statement is an important
+introduction for readers who are not familiar with ice-associated seals. We do
+agree with the reviewer's suggestion for more direct language and have made
+edits to the beginning of the abstract with this in mind.
+:::
+
+> ***Abstract:** Did you use previously published data, or did you collect it? 
+> Be specific using an active voice so that readers may understand your effort.*
 >
-> *For better flow, please indicate the behavioral data logging time scale here when you describe the data collection, not later when you speak to the analytical construct.*
-
-::: author-response
-**\[author response\]** We have edited this to simply state *data from satellite-linked bio-loggers* ...
-we have also altered this sentence to specifically mention the hourly timescale for the bio-logger data
-:::
-
-> *The term "basking" implies a subset of activities that ice seals may engage in when hauled out on sea ice and not an activity that can be directly telemetered using your data logger.  Yet, what you are describing is "haul-out behavior".  Please use this term here.*
+> *For better flow, please indicate the behavioral data logging time scale here
+> when you describe the data collection, not later when you speak to the
+> analytical construct.*
+
+::: author-response
+**\[author response\]** We have edited this to simply state *data from
+satellite-linked bio-loggers* ... we have also altered this sentence to
+specifically mention the hourly timescale for the bio-logger data
+:::
+
+> *The term "basking" implies a subset of activities that ice seals may engage
+> in when hauled out on sea ice and not an activity that can be directly
+> telemetered using your data logger.  Yet, what you are describing is "haul-out
+> behavior".  Please use this term here.*
 
@@
 
-> *Can you help your naive reader understand why you consider these covariates to be of interest and why others were not considered.*
-
-::: author-response
-**\[author response\]** The main body of the paper includes this discussion. Given the limited space and purpose of the abstract we feel it best to simply list the covariates of interest
-:::
-
-> ***Line 41:** Please use specific terms and not metaphors so that second-language English readers may better understand your writing.*
+> *Can you help your naive reader understand why you consider these covariates
+> to be of interest and why others were not considered.*
+
+::: author-response
+**\[author response\]** The main body of the paper includes this discussion.
+Given the limited space and purpose of the abstract we feel it best to simply
+list the covariates of interest
+:::
+
+> ***Line 41:** Please use specific terms and not metaphors so that
+> second-language English readers may better understand your writing.*
 
@@
 
-> ***Line 41:** What about human communities that don't live in the region, but that send large LNG tankers and other vessel through the region?  The loss of sea ice has effects on those communities and activities too.*
-
-::: author-response
-**\[author response\]** There is a seemingly endless list of impacts of loss of sea ice to human communities (and ecosystems) beyond the Arctic region to the rest of the world.
-However, for this sentence, we wanted to keep the reader focused on the Arctic region.
-:::
-
-> ***Line 44:** Please use direct language you mean biologically important functions?*
-
-::: author-response
-**\[author response\]** The term 'life history function' is a commonly used, direct term.
+> ***Line 41:** What about human communities that don't live in the region, but
+> that send large LNG tankers and other vessel through the region?  The loss of
+> sea ice has effects on those communities and activities too.*
+
+::: author-response
+**\[author response\]** Clearly there are a large number impacts of loss of
+sea ice to human communities (and ecosystems) beyond the Arctic region to the
+rest of the world. However, for this sentence, we wanted to keep the reader
+focused on the Arctic region.
+:::
+
+> ***Line 44:** Please use direct language you mean biologically important
+> functions?*
+
+::: author-response
+**\[author response\]** The term 'life history function' is a commonly used,
+direct term.
 
@@
 
-"The life history of an organism is its pattern of survival and reproduction, along with the traits that directly affect survival and the timing or amount of reproduction.
-Rates of survival and reproduction can be estimated across age classes, or across different stages in organisms with complex life cycles.
-Life history traits include growth rate; age and size at sexual maturity; the temporal pattern or schedule of reproduction; the number, size, and sex ratio of offspring; the distribution of intrinsic or extrinsic mortality rates (e.g., patterns of senescence); and patterns of dormancy and dispersal."
-
-To ensure the reader is clearly aware of which life-history behaviors/functions we are referring to we provide specific examples within the sentence.
+"The life history of an organism is its pattern of survival and reproduction,
+along with the traits that directly affect survival and the timing or amount of
+reproduction. Rates of survival and reproduction can be estimated across age
+classes, or across different stages in organisms with complex life cycles. Life
+history traits include growth rate; age and size at sexual maturity; the
+temporal pattern or schedule of reproduction; the number, size, and sex ratio of
+offspring; the distribution of intrinsic or extrinsic mortality rates (e.g.,
+patterns of senescence); and patterns of dormancy and dispersal."
+
+To ensure the reader is clearly aware of which life-history behaviors/functions
+we are referring to we provide specific examples within the sentence.
 :::
@@
 ::: author-response
-**\[author response\]** This is a good catch and we have edited the sentence to, now, read:
-
-"Knowledge about evolutionary constraints on the timing of reproductive and molting behavior is generally lacking, so it is difficult to predict how or if ice-associated seal species might adapt to future changes (e.g., by adjusting pupping or molting schedules to earlier dates or different locales)"
-:::
-
-> ***Line 56:** Perhaps you mean to offer a topic sentence, stating "How ice-assoicated seals use sea ice as a haulout platform varies betweens species".*
+**\[author response\]** This is a good catch and we have edited the sentence to,
+now, read:
+
+"Knowledge about evolutionary constraints on the timing of reproductive and
+molting behavior is generally lacking, so it is difficult to predict how or if
+ice-associated seal species will adjust to future changes (e.g., by adjusting
+pupping or molting schedules to earlier dates or different locales)"
+:::
+
+> ***Line 56:** Perhaps you mean to offer a topic sentence, stating "How
+> ice-assoicated seals use sea ice as a haulout platform varies betweens
+> species".*
 
@@
 
-> ***Line 66-68:** There is so much to state about ice-associated seals in the Pacific Arctic that this reference to Antarctic work seems out of place here.*
-
-::: author-response
-**\[author response\]** We have moved this reference to later in the paragraph so the first references are about Arctic work.
-While the example is from the Antarctic, it is an important, early example of analysis using bio-logger data to estimate haul-out behavior in seals.
+> ***Line 66-68:** There is so much to state about ice-associated seals in the
+> Pacific Arctic that this reference to Antarctic work seems out of place here.*
+
+::: author-response
+**\[author response\]** We have moved this reference to later in the paragraph
+so the first references are about Arctic work. While the example is from the
+Antarctic, it is an important, early example of analysis using bio-logger data
+to estimate haul-out behavior in seals.
 :::
@@
 
-> ***Line 84:** Do you mean age-class? The term life-history implies knowledge of previous behavioral efforts or occupancy of regions of interest.   Here you are specifically interested in age-class.*
+> ***Line 84:** Do you mean age-class? The term life-history implies knowledge
+> of previous behavioral efforts or occupancy of regions of interest.   Here you
+> are specifically interested in age-class.*
 
@@
 
-> ***Line 92:** Please help your reader understand why are you doing your analysis and not offer a wide and jargon rich inventory of concepts that may be considered.*
-
-> ***Line 93-94:** Please keep your introduction focused on establishing the knowledge gap that you are seeking to fill.
-Please do not lead your reader to the numerous other knowledge gaps that are not being addressed by your current study.*
+> ***Line 92:** Please help your reader understand why are you doing your
+> analysis and not offer a wide and jargon rich inventory of concepts that may
+> be considered.*
+
+> ***Line 93-94:** Please keep your introduction focused on establishing the
+> knowledge gap that you are seeking to fill. Please do not lead your reader to
+> the numerous other knowledge gaps that are not being addressed by your current
+> study.*
 
@@
 
-> ***Line 99:** Please use the term "hauled-out".
-Neither the biologger nor the aerial survey is able to distinguish among the various seal behaviors that result in them being hauled out and available for view by an aerial survey.*
-
-::: author-response
-**\[author response\]** noted, and the terms "hauled out", "to haul out", etc is used consistently throughout the manuscript
-:::
-
-> ***Line 114:** You have indicate that you would consider sea ice extent.
-Here you introduce the concept of icescape.
-Please introduce this concept earlier and indicate how you will incorporate this into your analysis.*
+> ***Line 99:** Please use the term "hauled-out". Neither the biologger nor the
+> aerial survey is able to distinguish among the various seal behaviors that
+> result in them being hauled out and available for view by an aerial survey.*
+
+::: author-response
+**\[author response\]** noted, and the terms "hauled out", "to haul out", etc is
+used consistently throughout the manuscript
+:::
+
+> ***Line 114:** You have indicate that you would consider sea ice extent. Here
+> you introduce the concept of icescape. Please introduce this concept earlier
+> and indicate how you will incorporate this into your analysis.*
 
@@
 
-"For this study we used haul-out behavior data and location estimates from bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering, Chukchi, and western Beaufort seas by multiple organizations as part of collaborative investigations from 2005 through 2020."
-:::
-
-> ***Line 120:** You summarize the years during which land-based studies applied bio-loggers.\
+"For this study we used haul-out behavior data and location estimates from
+bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
+Chukchi, and western Beaufort seas by multiple organizations as part of
+collaborative investigations from 2005 through 2020."
+:::
+
+> ***Line 120:** You summarize the years during which land-based studies applied
+> bio-loggers.\
 > Please do the same for the offshore deployments.*
@@
 
-"Ship-based capture events occurred during spring near the southern ice edge in the Bering Sea between 2005 and 2018."
-:::
-
-> ***Line 124:** Please provide the geographic context of your tag deployments and data collection using a plot similar to that used by Jay et al.
-(2012: 10.3354/meps10057 figure 2).
-Please consider using a tabulation that enables readers to understand the geographic context of the deployments similar to that used by Jay et al.
-2012, table 1.*
-
-::: author-response
-**\[author response\]** We feel the spatial distribution of the actual transmitted data -- compared to the deployment location -- is most informative and relevant for the audience's understanding of how our data are distributed spatially and how well data represents the range of each species.
-We provide this visually in Figure 3 and adding an additional layer to that figure to show deployments would be an additional complication.
-To ensure the audience has a clear understanding of our study area, we have added an additional figure (now, Figure 1) with a map of the Bering and Chukchi sea region that indicates generally where deployments were initiated.
-In response to other reviewer comments this figure also indicates the spatial range of aerial survey effort to date.
-:::
-
-> ***Line 126:** Please use active voice for all actions that you took in your data collection, preparation and analysis.*
+"Ship-based capture events occurred during spring near the southern ice edge in
+the Bering Sea between 2005 and 2018."
+:::
+
+> ***Line 124:** Please provide the geographic context of your tag deployments
+> and data collection using a plot similar to that used by Jay et al. (2012:
+> 10.3354/meps10057 figure 2). Please consider using a tabulation that enables
+> readers to understand the geographic context of the deployments similar to
+> that used by Jay et al. 2012, table 1.*
+
+::: author-response
+**\[author response\]** We feel the spatial distribution of the actual
+transmitted data -- compared to the deployment location -- is most informative
+and relevant for the audience's understanding of how our data are distributed
+spatially and how well data represents the range of each species. We provide
+this visually in Figure 3 and adding an additional layer to that figure to show
+deployments would be an additional complication. To ensure the audience has a
+clear understanding of our study area, we have added an additional figure (now,
+Figure 1) with a map of the Bering and Chukchi sea region that indicates
+generally where deployments were initiated. In response to other reviewer
+comments this figure also indicates the spatial range of aerial survey effort to
+date.
+:::
+
+> ***Line 126:** Please use active voice for all actions that you took in your
+> data collection, preparation and analysis.*
 
@@
 
-> ***Line 129-135:** Please move this to the discussion. This is a summary of your data findings, not a description of your methods.*
-
-::: author-response
-**\[author response\]** We can see the reviewer's point.
-However, we do not feel this topic is a summary of our data findings in this paper.
-Instead, our view is that this information provides the context and rationale for our methodological choice regarding bio-logger type and attachment methods.
-It is also important regarding the analytic methods -- preference for flipper tag haul-out data over those from a tag adhered to the dorsal/head.
-If we were to move these important points to the discussion, we worry those points would be left disconnected.
-:::
-
-> ***Line 136:** This is a discussion of your methods.
-This is not your methods.
-Please state how the data was collected in your Methods section and discuss how data collection have been uncertain or biased in your discussion section.*
+> ***Line 129-135:** Please move this to the discussion. This is a summary of
+> your data findings, not a description of your methods.*
+
+::: author-response
+**\[author response\]** We can see the reviewer's point. However, we do not feel
+this topic is a summary of our data findings in this paper. Instead, our view is
+that this information provides the context and rationale for our methodological
+choice regarding bio-logger type and attachment methods. It is also important
+regarding the analytic methods -- preference for flipper tag haul-out data over
+those from a tag adhered to the dorsal/head. If we were to move these important
+points to the discussion, we worry those points would be left disconnected.
+:::
+
+> ***Line 136:** This is a discussion of your methods. This is not your methods.
+> Please state how the data was collected in your Methods section and discuss
+> how data collection have been uncertain or biased in your discussion section.*
 
@@
 ::: author-response
-**\[author response\]** We appreciate this suggestion from the reviewer and we have moved large portions of this paragraph to the discussion.
-However, some information regarding age/sex class determination is left here so the reader has context for Table 1.
-:::
-
-> ***Line 144:** Please indicate how this was determined during data collection.*
-
-::: author-response
-**\[author response\]** We have edited this sentence some to clearly communicate that field determination of reproductive status (and, thus ideal delineation of sub-adult and adult age class) was not possible.
-:::
-
-> ***Line 147-148:** Please allow your readers the opportunity to understand the geographic context of these deployments through use of symbols that maybe used to cross-reference the deployment locations in a map that indicates deployment locations along with filtered movement tracks.*
-
-::: author-response
-**\[author response\]** As previously addressed, there are now two figures available that provide geographic context of the deployments.
-Figure 3 shows the geographic distribution of the data from the deployments.
-The paper is not presenting or discussing movement ecology and, therefore, we do not present filtered movement tracks.
-We keep the focus on haul-out behavior data.
-:::
-
-> ***Figure 1:** Please consider bar plot similar to that used by Jay et al. (2012: 10.3354/meps10057 figure 3), replacing the region with the seal species.*
-
-::: author-response
-**\[author response\]** This figure has been updated with this suggestion in mind
-:::
-
-> ***Line 151:** Please state clearly how you determined the end of a deployment.
-Here you indicate various elements of the haulout chronology, movement and dive chronology that may have been used, however not reader could replicate your methods base don this text.*
+**\[author response\]** We appreciate this suggestion from the reviewer and we
+have moved large portions of this paragraph to the discussion. However, some
+information regarding age/sex class determination is left here so the reader has
+context for Table 1.
+:::
+
+> ***Line 144:** Please indicate how this was determined during data
+> collection.*
+
+::: author-response
+**\[author response\]** We have edited this sentence some to clearly communicate
+that field determination of reproductive status (and, thus ideal delineation of
+sub-adult and adult age class) was not possible.
+:::
+
+> ***Line 147-148:** Please allow your readers the opportunity to understand the
+> geographic context of these deployments through use of symbols that maybe used
+> to cross-reference the deployment locations in a map that indicates deployment
+> locations along with filtered movement tracks.*
+
+::: author-response
+**\[author response\]** As previously addressed, there are now two figures
+available that provide geographic context of the deployments. Figure 3 shows the
+geographic distribution of the data from the deployments. The paper is not
+presenting or discussing movement ecology and, therefore, we do not present
+filtered movement tracks. We keep the focus on haul-out behavior data.
+:::
+
+> ***Figure 1:** Please consider bar plot similar to that used by Jay et al.
+> (2012: 10.3354/meps10057 figure 3), replacing the region with the seal
+> species.*
+
+::: author-response
+**\[author response\]** This figure has been updated with this suggestion in
+mind
+:::
+
+> ***Line 151:** Please state clearly how you determined the end of a
+> deployment. Here you indicate various elements of the haulout chronology,
+> movement and dive chronology that may have been used, however not reader could
+> replicate your methods base don this text.*
 
@@
 
-"For example, a data record that ends with several consecutive days (\~10+ days) of 100% dry observations and with locations indicating the tag was on land would be truncated to not include the final stretch of 100% dry observations.
-The vast majority of deployments end with the device detaching in the water and the deployment end date is obvious.
-There is no perfect algorithm for identifying deployment end dates and each deployment in question must be considered separately.
-While not perfect, we are confident our reliance on expert opinion and examination of multiple data streams provides the best option."
-:::
-
-> ***Line 152:** Deployment activity is known to disrupt normal pinniped behavior.
-Did you censure some period of data collection during and after the deployment to ensure that that disruptions from the deployment activity did not bias your analysis?*
-
-::: author-response
-**\[author response\]** For these phocid species, we have no empirical data (or even anecdotal observational data) on how captures may or may not disrupt 'normal' behavior.
-This is likely to be influenced by the capture method, how long the animal was restrained, whether the animal was with a dependent pup, whether the animal was molting, environmental variables such as wind and temperature, etc etc.
-Given all these unknowns, we have not applied a censure to the beginning of our deployment data.
-We can't 100% ensure this choice doesn't result in some level of bias.
-However, given the large sample size and duration of deployments we expect any bias imposed is negligible.
-:::
-
-> ***Line 154:** This is a result.
-Please indicate the method that you used to distinguish "plausible" from "implausible" telemetered seal behavior.
-In your Results section, you may indicate how method was able to identify occasions of implausible seal behavior and any patterns that arose relative to tag model or attachment.*
+"For example, a data record that ends with several consecutive days (\~10+ days)
+of 100% dry observations and with locations indicating the tag was on land would
+be truncated to the final stretch of 100% dry observations. The vast
+majority of deployments end with the device detaching in the water and the
+deployment end date is obvious. There is no perfect algorithm for identifying
+deployment end dates and each deployment in question must be considered
+separately. While not perfect, we are confident our reliance on expert judgement
+and examination of multiple data streams provides the best option."
+:::
+
+> ***Line 152:** Deployment activity is known to disrupt normal pinniped
+> behavior. Did you censure some period of data collection during and after the
+> deployment to ensure that that disruptions from the deployment activity did
+> not bias your analysis?*
+
+::: author-response
+**\[author response\]** For these phocid species, we have no empirical data (or
+even anecdotal observational data) on how captures may or may not disrupt
+'normal' behavior. This is likely to be influenced by the capture method, how
+long the animal was restrained, whether the animal was with a dependent pup,
+whether the animal was molting, environmental variables such as wind and
+temperature, etc etc. Given all these unknowns, we have not applied a censure to
+the beginning of our deployment data. We can't 100% ensure this choice doesn't
+result in some level of bias. However, given the large sample size and duration
+of deployments we expect any bias imposed is negligible.
+:::
+
+> ***Line 154:** This is a result. Please indicate the method that you used to
+> distinguish "plausible" from "implausible" telemetered seal behavior. In your
+> Results section, you may indicate how method was able to identify occasions of
+> implausible seal behavior and any patterns that arose relative to tag model or
+> attachment.*
 
@@
 **\[author response\]** This sentence has been moved to the results section.
-Additionally, we've provided a brief explanation of implausible along with a reference for more details.
-For these devices, the percent-dry timeline reported 100% dry for several weeks followed by an additional period of time that was more typical of expected bearded seal behavior.
-During the extended dry time the locations indicated movement of the tag that would not be simply drifting ice.
-We also observed increasing transmission rates.
-The working hypothesis (i.e.
-unconfirmed guess) is that the wet/dry sensor surface became obscured by some corrosive element such that when in the water conductivity was prevented and, thus, recorded as dry.
-:::
-
-> ***Line 156-164:** Please move to the Results section. Please move the discussion of how molt timing affects termination of tag deployments to the discussion section.*
-
-::: author-response
-**\[author response\]** This figure has been moved to the results section along with additional descriptive text.
-We have left the discussion regarding molt timing and termination of the data record here to better communicate the point to the reader and prevent an isolated mention in the discussion that has no anchor.
-:::
-
-> ***Line 169-170:** Thus, the a successful transmission of data was required at no greater than 48 h intervals.*
-
-::: author-response
-**\[author response\]** We're not sure how to interpret this comment.
-The main point of the sentence is to explain the rationale/benefits of transmitting the limited selection of percent dry values.
-We have expanded the text to provide more details regarding the caching of messages and data transmission.
+Additionally, we've provided a brief explanation of implausible along with a
+reference for more details. For these devices, the percent-dry timeline reported
+100% dry for several weeks followed by an additional period of time that was
+more typical of expected bearded seal behavior. During the extended dry time the
+locations indicated movement of the tag that would not be simply drifting ice.
+We also observed increasing transmission rates. The working hypothesis (i.e.
+unconfirmed guess) is that the wet/dry sensor surface became obscured by some
+corrosive element such that when in the water conductivity was prevented and,
+thus, recorded as dry.
+:::
+
+> ***Line 156-164:** Please move to the Results section. Please move the
+> discussion of how molt timing affects termination of tag deployments to the
+> discussion section.*
+
+::: author-response
+**\[author response\]** This figure has been moved to the results section along
+with additional descriptive text. We have left the discussion regarding molt
+timing and termination of the data record here to better communicate the point
+to the reader and prevent an isolated mention in the discussion that has no
+anchor.
+:::
+
+> ***Line 169-170:** Thus, the a successful transmission of data was required at
+> no greater than 48 h intervals.*
+
+::: author-response
+**\[author response\]** We're not sure how to interpret this comment. The main
+point of the sentence is to explain the rationale/benefits of transmitting the
+limited selection of percent dry values. We have expanded the text to provide
+more details regarding the caching of messages and data transmission.
 :::
@@
 > ***Line 174:** Do you mean "atmospheric" magnetic interference.\
-> Presumably "extra-terrestrial atmospheric" events (such as events on Venus or Mars) played no role in the transmission of the data signal to the Argos receivers in orbit around Earth.*
+> Presumably "extra-terrestrial atmospheric" events (such as events on Venus or
+> Mars) played no role in the transmission of the data signal to the Argos
+> receivers in orbit around Earth.*
 
@@
 ::: author-response
-**\[author response\]** We feel this note is important to acknowledge in the methods because it speaks to the rationale for our methodological choices (bio-logger type, programming choices, analytic choices).
+**\[author response\]** We feel this note is important to acknowledge in the
+methods because it speaks to the rationale for our methodological choices
+(bio-logger type, programming choices, analytic choices).
 :::
@@
 ::: author-response
-**\[author response\]** We have edited to the reviewer's suggested word 'encompasses'
+**\[author response\]** We have edited to the reviewer's suggested word
+'encompasses'
 :::
@@
 ::: author-response
-**\[author response\]** Thank you for catching this. We have edited this sentence to active voice
-:::
-
-> ***Line 198-199:** For most tags, your daily locations are from the Argos system, not from your data loggers.
-However, you included some FastLoc tag deployments.
-Please re-write for clarity of what these location are.*
-
-::: author-response
-**\[author response\]** The previous paragraph describes how we determined daily locations.
-We see the reviewers point re: the data not technically coming from our data loggers.
-We have edited the sentence for additional clarity.
+**\[author response\]** Thank you for catching this. We have edited this
+sentence to active voice
+:::
+
+> ***Line 198-199:** For most tags, your daily locations are from the Argos
+> system, not from your data loggers. However, you included some FastLoc tag
+> deployments. Please re-write for clarity of what these location are.*
+
+::: author-response
+**\[author response\]** The previous paragraph describes how we determined daily
+locations. We see the reviewers point re: the data not technically coming from
+our data loggers. We have edited the sentence for additional clarity.
 :::
@@
 
-> ***Line 203:** This is a procedural detail. Other readers may replicate your method without needing to do this. Please omit this detail.*
-
-> ***Line 207-208:** Readers can replicate your methods without the details on converting the NARR wind vectors into wind speed.*
-
-::: author-response
-**\[author response\]** We're not sure why the reviewer would recommend omitting a 'procedural detail' as part of our methods.
-We feel it is important to clearly communicate our procedure for processing the large NARR data set and that we used the native grid size of 32km.
-Regarding the calculation of wind vectors, we include this detail because many readers, like us initially, may not be well versed in how wind vectors components are converted to wind speed.
-:::
-
-> ***Line 210:** Please use an active voice: "For all seal species we considered the following variables when modeling the hourly haulout behavior..."*
-
-::: author-response
-**\[author response\]** We have edited this sentence with the reviewer's suggestion
-:::
-
-> ***Line 224:** Using the term "major focus" implies that there may be numerous minor foci.
-Can you not state that this is the goal of the study?
-Ancillary natural history insights can be discussed incidental to this, but are not driving the research.
-The statutory requirements of the MMPA and ESA are driving your study.
-These Federal laws require robust methods to monitor population status.
-Your study seeks to build this method.
-Discussion of natural history belong in your Discussion section and not in the framing of your study.
-Please help the reader focus on why the Federal government is doing this study.*
-
-> *Understanding pinniped haulout behavior is essential to designing and interpreting population estimation aerial surveys.
-Understanding how they use sea ice during a time when sea ice availability is rapidly changing is also important, but is ancillary to the paper's focus.*
-
-::: author-response
-**\[author response\]** We have removed the term 'major focus' and, as the reviewer suggests, simply state that this is the goal of the study.
-We have also added mention of the statutory requirements to paragraph 2 of the Introduction section and expanded the text to acknowledge the importance of understanding how seals use sea ice during this period of change but, ultimately, it's ancillary to the focus of this paper.
-
-As for providing the reader focus on why the Federal government is doing this study: this is a collaborative study across multiple organizations and the analysis/purpose is not limited to the Federal government.
-We've incorporated the reviewers suggestions to expand discussion of the statutory requirements of the MMPA and ESA but don't feel we need to single out the motivations for the Federal government.
-
-We have also added additional general representation of recent aerial survey effort to the new Figure 1.
-This should better communicate to the reader how the spatial distribution of our haul-out behavior overlaps with the species range and survey effort.
-:::
-
-> ***Line 227:** Seals can just as happily be on foraging trips within the pack and marginal sea ice as in the open water, depending on the species.
-Suggesting that foraging trips are associated with ice-free conditions is misleading.*
-
-::: author-response
-**\[author response\]** We have removed 'e.g., foraging trips' from this sentence so as not to mislead the reader.
-:::
-
-> ***Line 229-230:** This is your primary reason for not including sea ice geography in your analysis: you simply lacked geolocation of much of your haulout behavior data.*
+> ***Line 203:** This is a procedural detail. Other readers may replicate your
+> method without needing to do this. Please omit this detail.*
+
+> ***Line 207-208:** Readers can replicate your methods without the details on
+> converting the NARR wind vectors into wind speed.*
+
+::: author-response
+**\[author response\]** We're not sure why the reviewer would recommend omitting
+a 'procedural detail' as part of our methods. We feel it is important to clearly
+communicate our procedure for processing the large NARR data set and that we
+used the original grid size of 32km. Regarding the calculation of wind vectors, we
+include this detail because many readers, like us initially, may not be well
+versed in how wind vectors components are converted to wind speed.
+:::
+
+> ***Line 210:** Please use an active voice: "For all seal species we considered
+> the following variables when modeling the hourly haulout behavior..."*
+
+::: author-response
+**\[author response\]** We have adopted the reviewer's suggestion
+:::
+
+> ***Line 224:** Using the term "major focus" implies that there may be numerous
+> minor foci. Can you not state that this is the goal of the study? Ancillary
+> natural history insights can be discussed incidental to this, but are not
+> driving the research. The statutory requirements of the MMPA and ESA are
+> driving your study. These Federal laws require robust methods to monitor
+> population status. Your study seeks to build this method. Discussion of
+> natural history belong in your Discussion section and not in the framing of
+> your study. Please help the reader focus on why the Federal government is
+> doing this study.*
+
+> *Understanding pinniped haulout behavior is essential to designing and
+> interpreting population estimation aerial surveys. Understanding how they use
+> sea ice during a time when sea ice availability is rapidly changing is also
+> important, but is ancillary to the paper's focus.*
+
+::: author-response
+**\[author response\]** We have removed the term 'major focus' and, as the
+reviewer suggests, simply state that this is the goal of the study. We have also
+added mention of the statutory requirements to paragraph 2 of the Introduction
+section and expanded the text to acknowledge the importance of understanding how
+seals use sea ice during this period of change but, ultimately, it's ancillary
+to the focus of this paper.
+
+As for providing the reader focus on why the Federal government is doing this
+study: this is a collaborative study across multiple organizations and the
+analysis/purpose is not limited to the Federal government. We've incorporated
+the reviewers suggestions to expand discussion of the statutory requirements of
+the MMPA and ESA but don't feel we need to single out the motivations for the
+Federal government.
+
+We have also added additional general representation of recent aerial survey
+effort to the new Figure 1. This should better communicate to the reader how the
+spatial distribution of our haul-out behavior overlaps with the species range
+and survey effort.
+:::
+
+> ***Line 227:** Seals can just as happily be on foraging trips within the pack
+> and marginal sea ice as in the open water, depending on the species.
+> Suggesting that foraging trips are associated with ice-free conditions is
+> misleading.*
+
+::: author-response
+**\[author response\]** We have removed 'e.g., foraging trips' from this
+sentence so as not to mislead the reader.
+:::
+
+> ***Line 229-230:** This is your primary reason for not including sea ice
+> geography in your analysis: you simply lacked geolocation of much of your
+> haulout behavior data.*
 >
-> *Please help the reader understand how coarse your tracking data was for the flipper tagged animals and what proportion of each species was tracked with these flipper tags.
-Please also help the reader understand how the tracking data relates geographically to regions that may be subject to aerial survey efforts (i.e.
-are your tracking data collected from across the regions that may be surveyed during range-wide spring aerial surveys?).*
-
-::: author-response
-**\[author response\]** We have re-organized this paragraph to, hopefully, better communicate our choice not to explore sea-ice related covariates.
-As the reviewer suggests, we have limited geo-location data for at-sea periods of our haul-out behavior data -- we know the seal was in the water but don't have any estimates of location when only a flipper tag is attached.
-By leading with this statement, we hope that communicates our primary rationale.
-But, the other rationales are also important and we still include them.
-We have also moved this section to the 'Discussion' in response to other reviewer comments.
-
-We have provided some statistics regarding the percentage of the percent-dry timeline data that is not associated with a corresponding location estimate because the seal was in the water and no estimates were available from flipper mounted SPOT tags.
-
-We have also added a figure 1 depicting a spatial representation of recent aerial survey effort (which reflects estimated species range for bearded, ribbon, and spotted seals in US waters).
-This should better communicate to the reader how the spatial distribution of our haul-out behavior overlaps with the species range and survey effort.
-:::
-
-> ***Line 229-230:** Please help the reader understand why you have not included a basic metric of sea ice geography, such as distance from marginal ice edge (from the NIC's daily Marginal Ice Zone charts), which is a metric that would be available for all data collected during an aerial survey.*
-
-::: author-response
-**\[author response\]** We have revised the text in this section to try to provide a clearer understanding of our reasoning. Also, note this has been moved to the Discussion section.
+> *Please help the reader understand how coarse your tracking data was for the
+> flipper tagged animals and what proportion of each species was tracked with
+> these flipper tags. Please also help the reader understand how the tracking
+> data relates geographically to regions that may be subject to aerial survey
+> efforts (i.e. are your tracking data collected from across the regions that
+> may be surveyed during range-wide spring aerial surveys?).*
+
+::: author-response
+**\[author response\]** We have re-organized this paragraph to, hopefully,
+better communicate our choice not to explore sea-ice related covariates. As the
+reviewer suggests, we have limited geo-location data for in-water periods of our
+haul-out behavior data -- we know the seal was in the water but don't have any
+estimates of location when only a flipper tag is attached. By leading with this
+statement, we hope that communicates our primary rationale. But, the other
+rationales are also important and we still include them. We have also moved this
+section to the 'Discussion' in response to other reviewer comments.
+
+We have also added a figure 1 depicting a spatial representation of recent
+aerial survey effort (which reflects estimated species range for bearded,
+ribbon, and spotted seals in US waters). This should better communicate to the
+reader how the spatial distribution of our haul-out behavior overlaps with the
+species range and survey effort.
+:::
+
+> ***Line 229-230:** Please help the reader understand why you have not included
+> a basic metric of sea ice geography, such as distance from marginal ice edge
+> (from the NIC's daily Marginal Ice Zone charts), which is a metric that would
+> be available for all data collected during an aerial survey.*
+
+::: author-response
+**\[author response\]** We have revised the text in this section to try to
+provide a clearer understanding of our reasoning. Also, note this has been moved
+to the Discussion section.
 
@@
 
-Notably missing from our list of explanatory variables is any spatial-temporal representation of sea ice concentration, area, or extent.
-This may seem counterintuitive when modeling the haul-out behavior of seal species with such a close association to sea ice; seals haul out in the presence of sea ice, and we could assess the local concentration of sea ice during these events (see Olnes 2020).
-This, however, expands the scope of our analysis into the realm of habitat selection and many of our deployments consisted of a single device attached to the rear flipper of the seal which did not provide at-sea locations, limiting our ability to fully evaluate fine-scale habitat preferences related to sea-ice.
-Insight into how seals use and interact with sea-ice during an extended period when the availability and characteristics of sea-ice is rapidly changing is important but ancillary to the focus of this analysis.
-The focus of this study was to develop models applicable for aerial survey correction factors and using sea ice as a covariate would almost certainly bias haul-out predictions towards those seals that are on or near ice and therefore more likely to haul out.
-Since aerial surveys can only detect seals on ice, abundance estimates would be missing a correction for those seals that are away from ice.
-
-Consider the case of employing haul-out probabilities as an aerial survey correction in a population inhabiting two areas: an ice-free region and a region with sea ice.
-Further, assume that half of the population is in each area on average, and that the probability of hauling out is 1.0 in the ice covered region and 0.0 in the region without ice.
-Denoting the total population as $N$, let us examine what happens when we use (1) population level availability, and (2) ice-specific availability as a correction factor.
-We'll further assume, for sake of this example, that detection probability of seals on ice is 1.0, and that the entirety of the ice covered region is censused.
-In case (1), our aerial survey count is $C = 0.5 \cdot N$, with a population-level availability probability of $\hat{a} = 0.5$.
-A Horvitz-Thompson-type estimator for abundance is simply $\hat{N} = C \div \hat{a}$, which has expectation $\mathbb{E}(\hat{a}) = \frac{0.5 \cdot N}{0.5} = N$, as desired.
-Now consider case (2).
-In this case, the probability of hauling out is 1.0 for seals in the ice-covered area, so we have $\mathbb{E}(\hat{N}) = \frac{0.5 \cdot N}{1.0} = 0.5N$.
-That is, we bias abundance because we are not accounting for seals that are away from ice, and therefore have a zero chance of hauling out.
-The same logic holds for any other covariate, such as distance from the ice edge, that has different values in the surveyed and unsurveyed habitat of the seals (assuming distance from ice edge is coded with opposite signs inside and outside the pack ice).
-
-Lastly, our study was limited to the spring season when seal haul-out behaviors are strongly influenced by pupping, nursing, breeding behavior, and molt and these drivers are likely more influential than specific sea-ice concentration.
-Crawford et al.
-(2019) compared haul-out probability models for ringed seals and found those that only included season (and not sea-ice concentration) were the most parsimonious.
-For these reasons, we have elected not to use sea ice concentration *as a predictor for haul-out probability* in the present study.
-:::
-
-> ***Table 2:** It is a bit misleading to indicate that the NARR variables were sourced from NARR alone.
-Their values were interpolated onto a geogrpahic location that was widely variable in precision and that precision relates to the both the tag attachment (glued on or flipper attached) and the seal behavior (in water, hauled out).*
+Notably missing from our list of haul-out model explanatory variables is any
+spatial-temporal representation of sea ice concentration, area, or extent. This
+may seem counterintuitive when modeling the haul-out behavior of seal species
+with such a close association to sea ice; seals haul out in the presence of sea
+ice, and we could assess the local concentration of sea ice during these events
+(see [@olnes2020]). This, however, expands the scope of our analysis into the
+realm of habitat selection and many of our deployments consisted of a single
+device attached to the rear flipper of the seal which meant we only received
+locations when seals were hauled out on sea-ice, limiting our ability to fully
+evaluate fine-scale habitat preferences related to sea-ice. Insight into how
+seals use and interact with sea-ice during an extended period when the
+availability and characteristics of sea-ice is rapidly changing is important
+[@breed2018; @cameron2018] but ancillary to the focus of this analysis -- and,
+in the end, not possible given key limitations of our data. Additionally, our
+study was limited to the spring season when seal haul-out behaviors are strongly
+influenced by pupping, nursing, breeding behavior, and molt. These drivers are
+likely more influential on haul-out behavior than sea-ice concentration.
+Crawford et al. [-@crawford2019] compared haul-out probability models for ringed
+seals and found those that only included season (and not sea-ice concentration)
+were the most parsimonious. For all of these reasons, we have elected not to use
+sea ice concentration _as a predictor for haul-out probability_ in the present
+study.
+:::
+
+> ***Table 2:** It is a bit misleading to indicate that the NARR variables were
+> sourced from NARR alone. Their values were interpolated onto a geogrpahic
+> location that was widely variable in precision and that precision relates to
+> the both the tag attachment (glued on or flipper attached) and the seal
+> behavior (in water, hauled out).*
 >
-> *Indicating that your "Northing" variable was from the "Bio-logger" is also misleading because your locations was determined by either the Doppler inflection point time, satellite orbitography and a model of the Earth (the Argos relocation method) or by reprocessing of GNSS signals that were logged and then transmitted via the Argos system as a data message.
-Really, neither of these were determined by the "Bio-logger", rather they were determined by a very complex system of geolocation.*
-
-> *Your readers may be better served by not including this "Source" field, given that you describe how you obtained or estimated these variable values in your methods text.
-Offering an over-simplified "Source" in the table obscures your actual methods.*
-
-::: author-response
-**\[author response\]** We appreciate the reviewer's thoughts on this and agree with that removing the 'Source' column better serves the readers.
-:::
-
-> ***Line 244:** Neither your telemetry data nor aerial surveys can distinguish "basking" as a behavioral state.  You are measuring "hauled out" behavior and aerial surveys observe "hauled out" seals.  I encourage you to use these terms and not mix in ideas of "basking", "nursing", "breeding", "taking refuge from marine predators" etc.*
-
-::: author-response
-**\[author response\]** As previously mentioned, we've addressed this and consistently use "hauled out" and not "basking"
-:::
-
-> ***Line 254:** It is nice that you have sufficient data to do this.
-However, the reason for having separate models for each species is because you expect different behaviors from each specifies and because you expect that aerial efforts will be able to distinguish sea species and be able to apply an availability correction factor for each species.
-Please correct me if I am wrong, else, please keep your reader focused on the value that you are bringing to the analysis here in the Methods section and the good fortune that you have in the sample size, which should be noted in the Results and Discussion section.*
-
-::: author-response
-**\[author response\]** We have rephrased this sentence to emphasize that we fit each species separately because a) we expected differing behaviors across species and b) fitting a single model was not possible with such a large data set (even with the advantages provided by the glmmLDTS package).
-:::
-
-> ***Line 263:** Please clarify what is intended by the notation "only -99% or our observations".*
+> *Indicating that your "Northing" variable was from the "Bio-logger" is also
+> misleading because your locations was determined by either the Doppler
+> inflection point time, satellite orbitography and a model of the Earth (the
+> Argos relocation method) or by reprocessing of GNSS signals that were logged
+> and then transmitted via the Argos system as a data message. Really, neither
+> of these were determined by the "Bio-logger", rather they were determined by a
+> very complex system of geolocation.*
+
+> *Your readers may be better served by not including this "Source" field, given
+> that you describe how you obtained or estimated these variable values in your
+> methods text. Offering an over-simplified "Source" in the table obscures your
+> actual methods.*
+
+::: author-response
+**\[author response\]** We appreciate the reviewer's thoughts on this and agree
+with that removing the 'Source' column better serves the readers.
+:::
+
+> ***Line 244:** Neither your telemetry data nor aerial surveys can distinguish
+> "basking" as a behavioral state. You are measuring "hauled out" behavior and
+> aerial surveys observe "hauled out" seals. I encourage you to use these terms
+> and not mix in ideas of "basking", "nursing", "breeding", "taking refuge from
+> marine predators" etc.*
+
+::: author-response
+**\[author response\]** As previously mentioned, we've addressed this and
+consistently use "hauled out" and not "basking"
+:::
+
+> ***Line 254:** It is nice that you have sufficient data to do this. However,
+> the reason for having separate models for each species is because you expect
+> different behaviors from each specifies and because you expect that aerial
+> efforts will be able to distinguish sea species and be able to apply an
+> availability correction factor for each species. Please correct me if I am
+> wrong, else, please keep your reader focused on the value that you are
+> bringing to the analysis here in the Methods section and the good fortune that
+> you have in the sample size, which should be noted in the Results and
+> Discussion section.*
+
+::: author-response
+**\[author response\]** We have rephrased this sentence to emphasize that we fit
+each species separately because a) we expected differing behaviors across
+species and b) fitting a single model was not possible with such a large data
+set (even with the advantages provided by the glmmLDTS package).
+:::
+
+> ***Line 263:** Please clarify what is intended by the notation "only -99% or
+> our observations".*
 
@@
 
-> ***Line 263:** Also, please consider using consistent terms for "behavior records".  Perhaps it could be fair to term your telemetered haulout status as an "observation", however, it seems more intuitive to reference this a "record"*
-
-::: author-response
-**\[author response\]** We have made edits throughout the manuscript to use consistent terms for 'behavior records'
-:::
-
-> ***Line 263:** Please help your reader understand why you have not adopted use of a classified variable for each telemetered behavior interval whereby it could be either "wet" or "dry" based on a threshold proportion of readings reported as "dry".  This has two advantages: (1) were this enabled through processing on-board the tags, it would greatly increase your data compression and collection; (2) it may enable greater analytical power in your modeling effort.*
-
-::: author-response
-**\[author response\]** We provide fairly clear and detailed information on the haul-out behavior records, how they are recorded and transmitted in the methods section.
-We have further edited this to make sure it is clear that these are standard settings within the Wildlife Computers tag platform -- we do not have direct control over the details for on-board processing or transmission.
-As stated in the text, for this analysis, the linear predictor was modeled on the logit scale as a Bernoulli response where tags are either 'wet' or 'dry' for a given hour based on whether the tag was greater than 50% dry.
-This is a classified variable as suggested.
-The difference is that we conducted this classification post-transmission instead of on-board the tag.
-We feel the transmission of various percent-dry values for a given hour provide more flexibility in the types of analyses and insights that can be gained from deployments.
-:::
-
-> ***Line 265-272:** Please seek input from a competent biometrician.
-I am not able to offer any guidance on this other than to muse why you are pushing to estimate six parameters for a time of day covariate when your seals are not known to keep track of time.*
+> ***Line 263:** Also, please consider using consistent terms for "behavior
+> records".  Perhaps it could be fair to term your telemetered haulout status as
+> an "observation", however, it seems more intuitive to reference this a
+> "record"*
+
+::: author-response
+**\[author response\]** We have made edits throughout the manuscript to use
+consistent terms for 'behavior records'
+:::
+
+> ***Line 263:** Please help your reader understand why you have not adopted use
+> of a classified variable for each telemetered behavior interval whereby it
+> could be either "wet" or "dry" based on a threshold proportion of readings
+> reported as "dry".  This has two advantages: (1) were this enabled through
+> processing on-board the tags, it would greatly increase your data compression
+> and collection; (2) it may enable greater analytical power in your modeling
+> effort.*
+
+::: author-response
+**\[author response\]** We provide fairly clear and detailed information on the
+haul-out behavior records, how they are recorded and transmitted in the methods
+section. We have further edited this to make sure it is clear that these are
+standard settings within the Wildlife Computers tag platform -- we do not have
+direct control over the details for on-board processing or transmission. As
+stated in the text, for this analysis, the linear predictor was modeled on the
+logit scale as a Bernoulli response where tags are either 'wet' or 'dry' for a
+given hour based on whether the tag was greater than 50% dry. This is a
+classified variable as suggested. The difference is that we conducted this
+classification post-transmission instead of on-board the tag. We feel the
+transmission of various percent-dry values for a given hour provide more
+flexibility in the types of analyses and insights that can be gained from
+deployments.
+:::
+
+> ***Line 265-272:** Please seek input from a competent biometrician. I am not
+> able to offer any guidance on this other than to muse why you are pushing to
+> estimate six parameters for a time of day covariate when your seals are not
+> known to keep track of time.*
 >
-> *It would be reasonable to assume that seals respond to insolation or the angular height of the sun in its daily arc.  Please help your reader understand why you have sought to model a complex representation of hourly time when you could have pulled insolation from NARR or calculated sun angular altitude (perhaps as a linear proportion of the day's maximum sun angular altitude) from standard astronomical functions.*
+> *It would be reasonable to assume that seals respond to insolation or the
+> angular height of the sun in its daily arc.  Please help your reader
+> understand why you have sought to model a complex representation of hourly
+> time when you could have pulled insolation from NARR or calculated sun angular
+> altitude (perhaps as a linear proportion of the day's maximum sun angular
+> altitude) from standard astronomical functions.*
 >
-> *Perhaps your motivation is to provide plots that appear intuitive with hours on the y-axis, day of year on the x-axis and haulout proportion as a "heat map".  However beautiful these plot are, I wonder if your goal for improving precision and reducing bias when estimating availability during an aerial survey would not be better served by  the alternative and simpler variables.*
-
-::: author-response
-**\[author response\]** We appreciate the suggestion and find the possibility of using insolation values from reanalysis products as a covariate intriguing.
-As the reviewer suggests, such an approach could provide a less complicated and more parsimonious model than the approach we have taken with the Fourier series as a cyclic representation of hour-of day.
-We have taken the time to do some additional analysis and methodological exploration and provide additional details in Supplement 2.
-Note, our initial exploration of this approach resulted in lower AIC values for the Fourier series approach presented in the analysis.
-While promising and worth even further exploration, we feel the approach taken in the analysis is the most appropriate because it distinguishes daily bi-modal (e.g.
-crepuscular) patterns that might be important but would not be detected by insolation alone (i.e.
-insolation at dawn or dusk would be similar values but occur at different times of day that might be ecologically important).
-Additionally, as the reviewer points out, results from the analysis and haul-out predictions are more intuitive and human interpretable when using hour-of-day.
-:::
-
-> ***Line 278:** Please help your reader understand what is meant by your term "horizontal shift".  It is clear that you have a graphical representation in your mind that has not been shared with your reader.
-Please rewrite to avoid "jargon".*
-
-> ***Line 285:** Please help your reader understand what could be "important" about these terms. Please write for clarity, avoiding jargon.*
-
-> ***Figure 4:** Please use standard conventions for plot titles.  Your title deserves a statement in your Results or Discussion section.  It does not follow the expectations for a publication figure title.*
-
-::: author-response
-**\[author response\]** The plot titles were developed with strong consideration for improving scientific communication over approaches that are more typically used in publications.
-They were developed with guidance from modern approaches to data visualization that recognize the plot title is often the very first thing the reader sees and that data visualizations are often viewed and interpreted as standalone figures.
-Therefore it can serve as an effective communication tool to help the reader interpret the results.
-
-That said, we recognize that the initial approach could still be improved and have made adjustments to each of the titles with the feedback in mind. We hope the new titles provide a balanced approach.
-:::
-
-> ***Figure 4:** Please scale your color ramp to enable better visualization of the subject: limit the color ramp to the modeled values (which appear to be from zero to \~0.5).*
-
-::: author-response
-**\[author response\]** The scale used is consistent across all three species to maintain comparability. If we were to adjust the scale as suggested it would make comparisons across species difficult
-:::
-
-> ***Figure 4:** Please switch to a color ramp that is suitable for readers with color perception disabilities such as the viridis color ramps.*
-
-::: author-response
-**\[author response\]** The color ramp used has been updated to use palettes from the R package 'scico' have been tested and confirmed to be colorblind friendly and appropriately balanced for scientific interpretation.
-We initially used the viridis color palette for these plots but the 'scico' palette colors provided additional visual detail we feel is important.
-:::
-
-> ***Figure 4:** This is unclear. Your plot indicates a y-axis of local solar hour which varies from 0 to 24.  This is clearly not fixed at local noon.*
-
-::: author-response
-**\[author response\]** This was an error and has been fixed. We thank the reviewer for recognizing this and highlighting the mistake.
-:::
-
-> ***Line 323:** Please use standard English.  "Ontongeny" is generally understood to describe the physical development of an individual organism.  Here you are referencing the onset of of behaviors within the life of an organism.  It may benefit your readers to use a term that is more direct, such as "development".*
-
-::: author-response
-**\[author response\]** We have replaced 'ontongeny' with 'development' as suggested
-:::
-
-> ***Line 326-327:** Given that Ribbon seals occupy a high latitude region during this time and given that there is scarcely any twilight during this season, please consider using a term other than "crepuscular".*
-
-::: author-response
-**\[author response\]** Subsequent model runs (with improved QA/QC on telemetry data) have reduced the indication of any crepuscular or bi-modal haul-out behavior of ribbon seals and we no longer feel it is worth mentioning in the paper so it has been removed.
-:::
-
-> ***Line 331-332:** This is a fair summary of your modeling efforts.  However, it raises the questions as to what these seals are responding to.  Do they sense air pressure and make a decision on their haulout behavior or are low pressure conditions associated with low insolation and they are responding to that?  Given that you can readily extract insolation from the NARR, please consider use of insolation as a covariate that may more directly relate to seal behavior.*
-
-::: author-response
-**\[author response\]** See previous responses regarding the potential use of 'insolation' from NARR or other reanalysis products.
-:::
-
-> ***Figure 5:** Please use standard figure titling and not use a statement that should below in a PowerPoint slide or in your Results and Discussion section.*
+> *Perhaps your motivation is to provide plots that appear intuitive with hours
+> on the y-axis, day of year on the x-axis and haulout proportion as a "heat
+> map".  However beautiful these plot are, I wonder if your goal for improving
+> precision and reducing bias when estimating availability during an aerial
+> survey would not be better served by  the alternative and simpler variables.*
+
+::: author-response
+**\[author response\]** We appreciate the suggestion and find the possibility of
+using insolation values from reanalysis products as a covariate intriguing. As
+the reviewer suggests, such an approach could provide a less complicated and
+more parsimonious model than the approach we have taken with the Fourier series
+as a cyclic representation of hour-of day. We have taken the time to do some
+additional analysis and methodological exploration and provide additional
+details in Supplement 2. Note, our initial exploration of this approach resulted
+in lower AIC values for the Fourier series approach presented in the analysis.
+While promising and worth even further exploration, we feel the approach taken
+in the analysis is the most appropriate because it distinguishes daily bi-modal
+(e.g. crepuscular) patterns that might be important but would not be detected by
+insolation alone (i.e. insolation at dawn or dusk would be similar values but
+occur at different times of day that might be ecologically important).
+Additionally, as the reviewer points out, results from the analysis and haul-out
+predictions are more intuitive and applicable to aerial survey methods when 
+using hour-of-day.
+:::
+
+> ***Line 278:** Please help your reader understand what is meant by your term
+> "horizontal shift".  It is clear that you have a graphical representation in
+> your mind that has not been shared with your reader. Please rewrite to avoid
+> "jargon".*
+
+> ***Line 285:** Please help your reader understand what could be "important"
+> about these terms. Please write for clarity, avoiding jargon.*
+
+> ***Figure 4:** Please use standard conventions for plot titles.  Your title
+> deserves a statement in your Results or Discussion section.  It does not
+> follow the expectations for a publication figure title.*
+
+::: author-response
+**\[author response\]** We have removed plot titles and edited the figure
+captions to include the title information to be consistent with the journal
+styling and requirements. We've also revised many of the
+figure legends to better communicate the information to the reader and ensure
+if any figures are presented out of context (e.g. without the figure caption) in
+other venues the information is clear.
+:::
+
+> ***Figure 4:** Please scale your color ramp to enable better visualization of
+> the subject: limit the color ramp to the modeled values (which appear to be
+> from zero to \~0.5).*
+
+::: author-response
+**\[author response\]** We have adjusted the color ramp such that the pattern
+for bearded seals is more visible. We've also noted this change/difference in
+the figure caption so the reader is aware.
+:::
+
+> ***Figure 4:** Please switch to a color ramp that is suitable for readers with
+> color perception disabilities such as the viridis color ramps.*
+
+::: author-response
+**\[author response\]** The color ramp used has been updated to use palettes
+from the R package 'scico' have been tested and confirmed to be colorblind
+friendly and appropriately balanced for scientific interpretation. We initially
+used the viridis color palette for these plots but the 'scico' palette colors
+provided additional visual detail we feel is important.
+:::
+
+> ***Figure 4:** This is unclear. Your plot indicates a y-axis of local solar
+> hour which varies from 0 to 24. This is clearly not fixed at local noon.*
+
+::: author-response
+**\[author response\]** This was an error and has been fixed. We thank the
+reviewer for recognizing this and highlighting the mistake.
+:::
+
+> ***Line 323:** Please use standard English. "Ontongeny" is generally
+> understood to describe the physical development of an individual organism.
+> Here you are referencing the onset of of behaviors within the life of an
+> organism. It may benefit your readers to use a term that is more direct, such
+> as "development".*
+
+::: author-response
+**\[author response\]** We have replaced 'ontongeny' with 'development' as
+suggested
+:::
+
+> ***Line 326-327:** Given that Ribbon seals occupy a high latitude region
+> during this time and given that there is scarcely any twilight during this
+> season, please consider using a term other than "crepuscular".*
+
+::: author-response
+**\[author response\]** Subsequent model runs (with improved QA/QC on telemetry
+data) have reduced the indication of any crepuscular or bi-modal haul-out
+behavior of ribbon seals and we no longer feel it is worth mentioning in the
+paper so it has been removed.
+:::
+
+> ***Line 331-332:** This is a fair summary of your modeling efforts. However,
+> it raises the questions as to what these seals are responding to. Do they
+> sense air pressure and make a decision on their haulout behavior or are low
+> pressure conditions associated with low insolation and they are responding to
+> that?  Given that you can readily extract insolation from the NARR, please
+> consider use of insolation as a covariate that may more directly relate to
+> seal behavior.*
+
+::: author-response
+**\[author response\]** See previous responses regarding the potential use of
+'insolation' from NARR or other reanalysis products.
+:::
+
+> ***Figure 5:** Please use standard figure titling and not use a statement that
+> should below in a PowerPoint slide or in your Results and Discussion section.*
 >
-> ***Figure 6:** Please use standard conventions for plot titles.  Your title deserves a statement in your Results or Discussion section.  It does not follow the expectations for a publication figure title.*
-
-::: author-response
-**\[author response\]** See previous response regarding our choice in plot titles and subsequent edits
-:::
-
-> ***Figure 8:** Please label your plots using standard capitalization and words, instead of all capitalized abbreviations.*
+> ***Figure 6:** Please use standard conventions for plot titles. Your title
+> deserves a statement in your Results or Discussion section. It does not
+> follow the expectations for a publication figure title.*
+
+::: author-response
+**\[author response\]** See previous response regarding our choice in plot
+titles and subsequent edits
+:::
+
+> ***Figure 8:** Please label your plots using standard capitalization and
+> words, instead of all capitalized abbreviations.*
 
@@
 ::: author-response
-**\[author response\]** The suggestion to move the mention that there "were still some years where the pattern shown was informed by a small number of individuals that may not represent population-level patterns" to the discussion is noted and changes have been made.
-:::
-
-> ***Line 371-372:** Please indicate the season (or months of the year) in this statement.*
+**\[author response\]** The suggestion to move the mention that there "were
+still some years where the pattern shown was informed by a small number of
+individuals that may not represent population-level patterns" to the discussion
+is noted and changes have been made.
+:::
+
+> ***Line 371-372:** Please indicate the season (or months of the year) in this
+> statement.*
 
@@
 
-> ***Line 375-376:** Do you mean to state that "haulout behavior modeled from telemetered haulout behavior records may represent population-level behavior patterns."*
-
-::: author-response
-**\[author response\]** Noted and edits have been made
-:::
-
-> ***Line 377-378:** Feltz and Fay (1966) offer (an amazingly cool) functional explanation of the thermal constraints on hair follicle growth.  Whether seals during specific study times are optimizing hair follicle growth is unknown.*
+> ***Line 375-376:** Do you mean to state that "haulout behavior modeled from
+> telemetered haulout behavior records may represent population-level behavior
+> patterns."*
+
+::: author-response
+**\[author response\]** Noted and edited accordingly
+:::
+
+> ***Line 377-378:** Feltz and Fay (1966) offer (an amazingly cool) functional
+> explanation of the thermal constraints on hair follicle growth.  Whether seals
+> during specific study times are optimizing hair follicle growth is unknown.*
 
@@
 
-> ***Line 418-419:** Yes, and you have had access to direct estimates of insolation from the NARR.  Can you incorporate that and find a tighter model of haulout behavior.  This would be an improvement for design and interpretation of aerial survey efforts.*
-
-::: author-response
-**\[author response\]** See previous comments regarding the potential use of insolation values from reanalysis products
-:::
-
-> ***Line 420-421:** You may state this if you first offered an estimate of haulout behavior and confidence intervals on that behavior state, then compared that with your estimated haulout behavior modeled with the use of co-variates.  Once you perform this, you may state that inclusion of weather co-variates improved haulout state estimates.  However, you have not done this.  If you wish to state this, please do this.*
-
-::: author-response
-**\[author response\]** Here we simply state that 'our results highlight the importance of incorporating weather covariates' and that importance is indicated with the reported 'type III' F statistics for the weather covariate terms in the results section.
-We are clear in our methods that the use of pseudo-likelihoods in the analysis prevent us from conducting typical model selection with AIC or similar as suggested.
-We have moved a separate discussion of comparison with previous analysis results here to provide additional comparative context.
-:::
-
-> ***Line 425:** This term suggests some evaluation of a threshold above which haulout behavior shifts. Please write for clarity and for what you have presented in your results.*
+> ***Line 418-419:** Yes, and you have had access to direct estimates of
+> insolation from the NARR.  Can you incorporate that and find a tighter model
+> of haulout behavior.  This would be an improvement for design and
+> interpretation of aerial survey efforts.*
+
+::: author-response
+**\[author response\]** See previous comments regarding the potential use of
+insolation values from reanalysis products
+:::
+
+> ***Line 420-421:** You may state this if you first offered an estimate of
+> haulout behavior and confidence intervals on that behavior state, then
+> compared that with your estimated haulout behavior modeled with the use of
+> co-variates. Once you perform this, you may state that inclusion of weather
+> co-variates improved haulout state estimates. However, you have not done
+> this. If you wish to state this, please do this.*
+
+::: author-response
+**\[author response\]** Here we simply state that 'our results highlight the
+importance of incorporating weather covariates' and that importance is indicated
+with the reported 'type III' F statistics for the weather covariate terms in the
+results section. We are clear in our methods that the use of pseudo-likelihoods
+in the analysis prevent us from conducting typical model selection with AIC or
+similar as suggested. We have moved a separate discussion of comparison with
+previous analysis results here to provide additional comparative context.
+:::
+
+> ***Line 425:** This term suggests some evaluation of a threshold above which
+> haulout behavior shifts. Please write for clarity and for what you have
+> presented in your results.*
 
@@
 
-> ***Line 455:** Please be clear as to why you anticipate greater unmodeled heterogeneity of availability for Arctic seals?  Here you have considered Arctic seals, have you not?  Or are you using a different definition of the Arctic other than the ice bearing waters occupied by ice-associated seals?*
-
-::: author-response
-**\[author response\]** We have edited this statement for additional clarity.
-The interpreted emphasis on 'Arctic seals' was not intended.
-Instead, the key point is that there is likely unmodeled heterogeneity related to the seasonal distribution and annual timing of peak haul-out behavior.
-This is an acknowledgement that our model is not a complete picture or perfect representation of seal haul-out behavior.
-In an environment like the Arctic that is undergoing dramatic change and inter-annual variation in climate and sea-ice distribution, this is important to keep in mind.
-:::
-
-> ***Line 459-464:** Here you provide the discussion required to make the statement that inclusion of weather and other covariates results in a different estimate of haulout proportion.  Can you extend this to indicate an improvement in the confidence interval on this estimated proportion?*
+> ***Line 455:** Please be clear as to why you anticipate greater unmodeled
+> heterogeneity of availability for Arctic seals? Here you have considered
+> Arctic seals, have you not? Or are you using a different definition of the
+> Arctic other than the ice bearing waters occupied by ice-associated seals?*
+
+::: author-response
+**\[author response\]** We have edited this statement to now read:
+
+The most challenging element in developing
+availability correction factors is with annual variability. It can be difficult
+to get a sufficient sample size to estimate year-specific correction factors,
+particularly because research teams would likely need to deploy bio-loggers on
+seals and conduct aerial surveys concurrently, requiring considerably more
+personnel and money. One possible suggestion is to include year as a random
+effect within models for aerial survey counts such that, without specific
+knowledge of any particular year, the among-year variance is included in the
+modeled standard errors. Regardless of the specific approach, future estimates
+of Arctic seal abundance will require specific consideration of annual
+variability and changes in the timing of peak haul-out behavior when estimating
+trends, as one will not know if moderate differences in abundance estimates are
+attributable to changes in abundance or changes in haul-out behavior.
+
+The interpreted emphasis on 'Arctic seals' was not intended. Instead, the key
+point is that there is likely unmodeled heterogeneity related to the seasonal
+distribution and annual timing of peak haul-out behavior. This is an
+acknowledgement that our model is not a complete picture or perfect
+representation of seal haul-out behavior. In an environment like the Arctic that
+is undergoing dramatic change and inter-annual variation in climate and sea-ice
+distribution, this is important to keep in mind.
+:::
+
+> ***Line 459-464:** Here you provide the discussion required to make the
+> statement that inclusion of weather and other covariates results in a
+> different estimate of haulout proportion. Can you extend this to indicate an
+> improvement in the confidence interval on this estimated proportion?*
 
@@
 
-> ***Line 465-466:** Please be specific. Here you linked the telemetered haulout behavior to available covariate data then modeled haulout behavior.*
-
-::: author-response
-**\[author response\]** Noted and additional specifics have been added to this statement
-:::
-
-> ***Line 467-473:** This can be stated in the introduction so that you can clearly indicate the framing of your study.
-"We modeled availability from telemetered haulout behavior records of seals that haulout on ice where they may be viewed by aerial surveys (list the species considered).*
+> ***Line 465-466:** Please be specific. Here you linked the telemetered haulout
+> behavior to available covariate data then modeled haulout behavior.*
+
+::: author-response
+**\[author response\]** Noted and additional specifics have been added to this
+statement
+:::
+
+> ***Line 467-473:** This can be stated in the introduction so that you can
+> clearly indicate the framing of your study. "We modeled availability from
+> telemetered haulout behavior records of seals that haulout on ice where they
+> may be viewed by aerial surveys (list the species considered).*
 
@@
 
-> *One suggestion I would share with the authors is the addition of a limitations section in the discussion.
-The limitations were outlined throughout the methods, but a succinct accounting of these limitations in the discussion would allow readers who do not want to carefully dissect all of the modeling in the methods to be able to better interpret the robustness of the results and conclusions.*
-
-::: author-response
-**\[author response\]** We thank the reviewer for this question and have added a section about limitations to the Discussion section
-:::
-
-> *I still had one remaining question on why the authors couldn't run the models with and without the sea ice concentration as a predictor.
-The comparative results might be interesting (I understand why the correction factor might be biased\...but given that this is a baseline study, it would have been an interesting comparison to see whether it was or wasn't a decent predictor for the different seal species).*
-
-::: author-response
-**\[author response\]** We think we've addressed this in edits and comments to reviewer #1.
-We think the inclusion of sea ice as a predictor is better suited for a habitat use/habitat preference study and we hope to pursue that in future analyses.
-:::
-
-> *The authors need to explain more how they are attributing statistical significance. For example, in lines 344-345, they ascribe strong influence of temperature with a p-value of 0.064.\
+> *One suggestion I would share with the authors is the addition of a
+> limitations section in the discussion. The limitations were outlined
+> throughout the methods, but a succinct accounting of these limitations in the
+> discussion would allow readers who do not want to carefully dissect all of the
+> modeling in the methods to be able to better interpret the robustness of the
+> results and conclusions.*
+
+::: author-response
+**\[author response\]** We thank the reviewer for this question and, in response
+to this and other reviewer feedback, have moved much of the limitations outlined
+in the methods to the Discussion section and hope this provides a clearer
+picture to the reader regarding limits of our analysis.
+:::
+
+> *I still had one remaining question on why the authors couldn't run the models
+> with and without the sea ice concentration as a predictor. The comparative
+> results might be interesting (I understand why the correction factor might be
+> biased...but given that this is a baseline study, it would have been an
+> interesting comparison to see whether it was or wasn't a decent predictor for
+> the different seal species).*
+
+::: author-response
+**\[author response\]** We think we've addressed this in edits and comments to
+reviewer #1. We think the inclusion of sea ice as a predictor is better suited
+for a habitat use/habitat preference study and we hope to pursue that in future
+analyses. We have also emphasized that our reliance on flipper-mounted
+bio-loggers for a portion of our data limited our understanding of movement
+and use away from sea-ice.
+:::
+
+> *The authors need to explain more how they are attributing statistical
+> significance. For example, in lines 344-345, they ascribe strong influence of
+> temperature with a p-value of 0.064.\
 > \
-> Not to be one who is fixated on a p-value of 0.05 denoting significance, I was a bit surprised to see the same authors, in lines 364-366, to state that there is 'no indication that the observed trend is meaningful'for spotted seal adult females having an R2 of 0.767, p=0.022.
-If the extent of sea ice explains nearly 80% of the variance in peak haul-out probability, how is it that the author's call this a 'trend'?
-This requires more explanation.
-As does the statement in line 427-428.*
-
-::: author-response
-**\[author response\]** We appreciate the reviewer highlighting this confusion and apologize for it.
-These were typos or mis-coded values that snuck into the section.
-We have reviewed and revised all of them to ensure they are correct and that we are consistent in our attribution of significance.
-:::
-
-> *I found 463-464 confusing.
-This is an important part of the discussion (using your framework to assess previous results from another study), but it could be more clearly written.
-The sentence, 'Applying models that ignore age, sex, and year effects\...under the current analysis framework' is very confusing and doesn't seem to follow the argument of the paper easily.
-Could this be reworded so that the full intent of the paragraph is communicated?*
-
-::: author-response
-**\[author response\]** Reviewer #1 also commented on this section and offered suggestions for editing.
-With both comments in mind, we have made edits to the text and re-arranged the location so that this is more closely linked with the discussion regarding the importance of weather covariates in the model.
-:::
-
-> *I appreciate the acknowledgement that it is difficult to know whether differences in abundance estimates are attributable to changes in abundance or changes in haul-out behavior and the potential proposed solution (lines 450-458) Has anyone done a smaller-scale study that would address this inherent problem?*
+> Not to be one who is fixated on a p-value of 0.05 denoting significance, I was
+> a bit surprised to see the same authors, in lines 364-366, to state that there
+> is 'no indication that the observed trend is meaningful'for spotted seal adult
+> females having an R2 of 0.767, p=0.022. If the extent of sea ice explains
+> nearly 80% of the variance in peak haul-out probability, how is it that the
+> author's call this a 'trend'? This requires more explanation. As does the
+> statement in line 427-428.*
+
+::: author-response
+**\[author response\]** We appreciate the reviewer highlighting this confusion
+and apologize for it. These were typos or mis-coded values that snuck into the
+section. We have reviewed and revised all of them to ensure they are correct and
+that we are consistent in our attribution of significance. We have tried to be
+judicious in our use of the work 'significant'.
+:::
+
+> *I found 463-464 confusing. This is an important part of the discussion (using
+> your framework to assess previous results from another study), but it could be
+> more clearly written. The sentence, 'Applying models that ignore age, sex, and
+> year effects...under the current analysis framework' is very confusing and
+> doesn't seem to follow the argument of the paper easily. Could this be
+> reworded so that the full intent of the paragraph is communicated?*
+
+::: author-response
+**\[author response\]** Reviewer #1 also commented on this section and offered
+suggestions for editing. With both comments in mind, we have made edits to the
+text and hope this, now, more clearly communicates how our results compare with
+previous approaches.
+:::
+
+> *I appreciate the acknowledgement that it is difficult to know whether
+> differences in abundance estimates are attributable to changes in abundance or
+> changes in haul-out behavior and the potential proposed solution (lines
+> 450-458) Has anyone done a smaller-scale study that would address this
+> inherent problem?*
+
+::: author-response
+**\[author response\]** To our knowledge a smaller-scale study addressing this
+issue has not been done
+:::
 
@@
 
-> *I have only a few minor comments, mostly where the text would benefit from extra clarity:\
+> *I have only a few minor comments, mostly where the text would benefit from
+> extra clarity:\
 > *
@@
 
-> **L63:** It may be useful to highlight somewhere that matching seal haulout behaviour to environmental conditions with any degree of certainty requires location data as well as wet-dry data, and that modern bio-logging devices can collect both data streams simultaneously.\
-
-::: author-response
-**\[author response\]** thank you for this suggestion and we have edited this paragraph to include mention regarding the ability of modern bio-logging devices to collect both behavioral and location data streams which allows inclusion of environmental covariates.
-:::
-
-> **L82-83:** All citations given here relate to ice-associated seals, but these techniques are also relevant to seals that haul-out on land.
-Lonergan et al.
-(2013) Aquatic Conservation: Marine and Freshwater Ecosystems 23: 135-144 is a key example from non-polar regions.\
-
-::: author-response
-**\[author response\]** we have expanded the list of references to include Lonergan et al as well as a few other studies from non ice-associated seals
-:::
-
-> **L110:** Clarify why meteorological factors affecting walruses may be relevant to the study species.\
-
-::: author-response
-**\[author response\]** thank you for the suggestion. We've edited the sentence to, now, read:
-
-"Previously estimated availability correction factors accounted for variables such as time-of-day and day-of-year but did not investigate the impact of meteorological variables.
-Such variables have been shown to influence walrus haul-out behavior and we expect weather conditions to also influence seal haul-out behavior and including them within the model framework will benefit our estimates of seal availability during aerial surveys."
-:::
-
-> **L120-124:** A map of the study area with deployment areas indicated would be useful for those readers not familiar with the local geography.\
-
-::: author-response
-**\[author response\]** Reviewer 1 had a similar suggestion and we now include such a figure.
-:::
-
-> **L126-128:** More information is required on the location data type (e.g. GPS or ARGOS?) and data retrieval (e.g. archival loggers or ARGOS-relay?).\
+> **L63:** It may be useful to highlight somewhere that matching seal haulout
+> behaviour to environmental conditions with any degree of certainty requires
+> location data as well as wet-dry data, and that modern bio-logging devices can
+> collect both data streams simultaneously.\
+
+::: author-response
+**\[author response\]** thank you for this suggestion and we have edited this
+paragraph to include mention regarding the ability of modern bio-logging devices
+to collect both behavioral and location data streams which allows inclusion of
+environmental covariates.
+:::
+
+> **L82-83:** All citations given here relate to ice-associated seals, but these
+> techniques are also relevant to seals that haul-out on land. Lonergan et al.
+> (2013) Aquatic Conservation: Marine and Freshwater Ecosystems 23: 135-144 is a
+> key example from non-polar regions.\
+
+::: author-response
+**\[author response\]** we have expanded the list of references to include
+Lonergan et al
+:::
+
+> **L110:** Clarify why meteorological factors affecting walruses may be
+> relevant to the study species.\
+
+::: author-response
+**\[author response\]** thank you for the suggestion. We've edited the sentence
+to, now, read:
+
+"Previously estimated availability correction factors accounted for variables
+such as time-of-day and day-of-year but did not investigate the impact of
+meteorological variables. Such variables have been shown to influence walrus
+haul-out behavior and we expect weather conditions to also influence seal
+haul-out behavior and including them within the model framework will benefit our
+estimates of seal availability during aerial surveys."
+:::
+
+> **L120-124:** A map of the study area with deployment areas indicated would be
+> useful for those readers not familiar with the local geography.\
+
+::: author-response
+**\[author response\]** Reviewer 1 had a similar suggestion and we now include
+such a figure.
+:::
+
+> **L126-128:** More information is required on the location data type (e.g. GPS
+> or ARGOS?) and data retrieval (e.g. archival loggers or ARGOS-relay?).\
 
@@
 
-"All data were transmitted by the deployed instruments via the Argos satellite network and location data were either derived from Argos transmissions or transmitted FastLoc GPS data."
-:::
-
-> **L133-135:** Please clarify, were there any cases where only head/back-mounted tags were deployed?
-If so, how did you deal with the potential for false haul-out records due to prolonged surface activity?\
-
-::: author-response
-**\[author response\]** we've clarified the language in this paragraph so it's clear that deployments consisted of head/back tag only, flipper tag only, and combined.
-Examination of the haul-out records hasn't indicated there's a propensity for these tags to record extended dry periods when surface resting at sea for the head/back tags.
-:::
-
-> **L135:** At this stage you have not yet explained what type of data each device collects, so we do not know what the hourly percent-dry observations from the flipper tag were preferred to.\
-
-::: author-response
-**\[author response\]** thank you for noting this. we have moved the later paragraph (starting line 165) up so it is before this section.
+"All data were transmitted by the deployed instruments via the Argos satellite
+network and location data were either derived from Argos transmissions or
+transmitted FastLoc GPS data."
+:::
+
+> **L133-135:** Please clarify, were there any cases where only
+> head/back-mounted tags were deployed? If so, how did you deal with the
+> potential for false haul-out records due to prolonged surface activity?\
+
+::: author-response
+**\[author response\]** we've clarified the language in this paragraph so it's
+clear that deployments consisted of head/back tag only, flipper tag only, and
+combined. Examination of the haul-out records hasn't indicated there's a
+propensity for these tags to record extended dry periods when surface resting at
+sea for the head/back tags.
+:::
+
+> **L135:** At this stage you have not yet explained what type of data each
+> device collects, so we do not know what the hourly percent-dry observations
+> from the flipper tag were preferred to.\
+
+::: author-response
+**\[author response\]** thank you for noting this. we have moved the later
+paragraph (starting line 165) up so it is before this section.
 :::
@@
 ::: author-response
-**\[author response\]** we have removed the 'for some species' from this sentence
-:::
-
-> **L136-144:** This could be clearer.
-I assume the "some species" part refers only to pelage, but it appears at first read to refer to the whole list of metrics.
-I would suggest re-wording this section to improve clarity regarding which factors were used for which species and with what degree of certainty age class could be determined.\
-
-::: author-response
-**\[author response\]** as noted in the previous response, we've removed the 'for some species' and think this improves the clarity of this section.
-Unfortunately, we don't have any way to quantify a degree of certainty for age class determination across the various field teams responsible for field identification.
+**\[author response\]** we have removed the 'for some species' from this
+sentence
+:::
+
+> **L136-144:** This could be clearer. I assume the "some species" part refers
+> only to pelage, but it appears at first read to refer to the whole list of
+> metrics. I would suggest re-wording this section to improve clarity regarding
+> which factors were used for which species and with what degree of certainty
+> age class could be determined.\
+
+::: author-response
+**\[author response\]** as noted in the previous response, we've removed the
+'for some species' and think this improves the clarity of this section.
+Unfortunately, we don't have any way to quantify a degree of certainty for age
+class determination across the various field teams responsible for field
+identification.
 :::
@@
 ::: author-response
-**\[author response\]** we've added this metric
+**\[author response\]** there were 11 YOY advanced to subadult. We considered
+adding this into the results but struggled to find a place where the information
+fit and was relevant.
 :::
@@
 ::: author-response
-**\[author response\]** we have edited this sentence to state 'also haul out on land'
-:::
-
-> **L194:** It is not clear to me if all tags for a given species were deployed in the same region, or across multiple regions (a map of the study area with deployment regions per species would help here).
-Given regional differences in phenology mentioned in the Introduction (L52-53), did you consider whether region should be included as a covariate?\
+**\[author response\]** we have edited this sentence to state 'also haul out on
+land'
+:::
+
+> **L194:** It is not clear to me if all tags for a given species were deployed
+> in the same region, or across multiple regions (a map of the study area with
+> deployment regions per species would help here). Given regional differences in
+> phenology mentioned in the Introduction (L52-53), did you consider whether
+> region should be included as a covariate?\
 
@@
 
-> **Figs 4, 6, 8:** How is that hour of the day is fixed at local solar noon if predictions are shown for all hours?
-Also, is it possible to show uncertainty (lower and upper CIs) for these figures, perhaps in Supplementary Materials?\
-
-::: author-response
-**\[author response\]** Reviewer 1 also noted the incorrect statement that hour of day was fixed at local solar noon and this has been corrected.
-We have taken the reviewer's suggestion and added supplemental materials with plots (this time fixed at solar noon) that indicated the lower/upper CIs.
+> **Figs 4, 6, 8:** How is that hour of the day is fixed at local solar noon if
+> predictions are shown for all hours? Also, is it possible to show uncertainty
+> (lower and upper CIs) for these figures, perhaps in Supplementary Materials?\
+
+::: author-response
+**\[author response\]** Reviewer 1 also noted the incorrect statement that hour
+of day was fixed at local solar noon and this has been corrected. We have taken
+the reviewer's suggestion and added supplemental materials with plots (this time
+fixed at solar noon) that indicated the lower/upper CIs.
 :::
@@
 
-> **L387-390:** It is not clear what species was studied by Thometz et al, and thus maybe a citation is needed for the information relating to ribbon and spotted seals.\
-
-::: author-response
-**\[author response\]** we've added a clarification that the Thometz et al study and data presented was for bearded seals.
+> **L387-390:** It is not clear what species was studied by Thometz et al, and
+> thus maybe a citation is needed for the information relating to ribbon and
+> spotted seals.\
+
+::: author-response
+**\[author response\]** we've added a clarification that the Thometz et al study
+and data presented was for bearded seals.
 :::
@@
 ::: author-response
-**\[author response\]** we've edited to specifically state 'ribbon and spotted seals' to avoid any confusion
+**\[author response\]** we've edited to specifically state 'ribbon and spotted
+seals' to avoid any confusion
 :::
@@
 > \
-> The paper could benefit from a brief concluding paragraph summarising the importance of the work in the context of climate change and population trends, as outlined in the Introduction.
-
-::: author-response
-**\[author response\]** thank you for this suggestion. we've added a concluding paragraph that addresses this comment.
-:::
+> The paper could benefit from a brief concluding paragraph summarising the
+> importance of the work in the context of climate change and population trends,
+> as outlined in the Introduction.
+
+::: author-response
+**\[author response\]** thank you for this suggestion. we've added a concluding
+paragraph that addresses this comment.
+:::

### 2024-02-18 analysis/paper/London_HauloutBehavior.Rmd
final commit for revision

--- 2024-02-17 analysis/paper/London_HauloutBeahvior.Rmd
@@
 
-knitr::opts_chunk$set(dev = "ragg_png")
+knitr::opts_chunk$set(dev = c("ragg_png"))
 
@@
   fig.path = "../figures/",
-  dpi = 300,
+  dpi = 600,
   fig.width = 6,
@@
 
-Again, we want to acknowledge Anthony Fisbach for the suggestion during the
+Again, we want to acknowledge Anthony Fischbach for the suggestion during the
 peer review process. We think this is an excellent example of the peer review

### 2024-02-18 analysis/paper/London_HauloutBehavior.tex
final commit for revision

%% Submissions for peer-review must enable line-numbering
%% using the lineno option in the \documentclass command.
%%
%% Preprints and camera-ready submissions do not need
%% line numbers, and should have this option removed.
%%
%% Please note that the line numbering option requires
%% version 1.1 or newer of the wlpeerj.cls file, and
%% the corresponding author info requires v1.2

\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions

% ZNK -- Adding headers for pandoc

\setlength{\emergencystretch}{3em}
\usepackage[unicode=true]{hyperref}

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{stix2}
\usepackage {booktabs}
\usepackage {setspace}
\onehalfspacing
\usepackage [T1]{fontenc}
\usepackage [semibold]{sourcesanspro}
\usepackage {hyperref}
\hypersetup {colorlinks, linkcolor={red!50!black}, citecolor={blue!50!black},urlcolor={blue!80!black}}
\usepackage {pdflscape}
\newcommand {\blandscape}{\begin{landscape}}
\newcommand {\elandscape}{\end{landscape}}
\usepackage {caption}
\captionsetup{font={normalsize,onehalfspacing}}
\newcommand{\beginsupplement}{\setcounter{table}{0}\renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0}\renewcommand{\thefigure}{S\arabic{figure}}}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}

\title{Spring haul-out behavior of seals in the Bering and Chukchi seas}

\author[1]{Josh M. London}

\corrauthor[1]{Josh M. London}{\href{mailto:josh.london@noaa.gov}{\nolinkurl{josh.london@noaa.gov}}}
\author[1]{Paul B. Conn}

\author[1]{Stacie M. Koslovsky}

\author[1]{Erin L. Richmond}

\author[1]{Jay M. Ver Hoef}

\author[1]{Michael F. Cameron}

\author[2]{Justin A. Crawford}

\author[3]{Andrew L. Von Duyke}

\author[2]{Lori T. Quakenbush}

\author[1]{Peter L. Boveng}

\affil[1]{Marine Mammal Laboratory, Alaska Fisheries Science Center, National Marine Fisheries Service, NOAA, Seattle, Washington, USA}
\affil[2]{Arctic Marine Mammals Program, Alaska Department of Fish and Game, Fairbanks, Alaska, USA}
\affil[3]{Department of Wildlife Management, North Slope Borough, Utqiaġvik, Alaska, USA}

%
% \author[1]{First Author}
% \author[2]{Second Author}
% \affil[1]{Address of first author}
% \affil[2]{Address of second author}
% \corrauthor[1]{First Author}{f.author@email.com}

%

\begin{abstract}
Ice-associated seals rely on sea ice for a variety of activities, including pupping, breeding, molting, and resting. In the Arctic, many of these activities occur in spring (April -- June) as sea ice begins to melt and retreat northward. Rapid acceleration of climate change in Arctic ecosystems is therefore of concern as the quantity and quality of suitable habitat is forecast to decrease. Improved estimates of seal population abundance are needed to properly monitor the impacts of these changes over time. In this paper, we use hourly percent-dry data from satellite-linked bio-loggers deployed between 2005 and 2021 to quantify the proportion of seals hauled out on ice. This information is needed to accurately estimate abundance from aerial survey counts of ice-associated seals (i.e., to correct for the proportion of animals that are in the water while surveys are conducted). In addition to providing essential data for survey `availability' calculations, our analysis also provides insights into the seasonal timing and environmental factors affecting haul-out behavior by ice-associated seals. We specifically focused on bearded (\emph{Erignathus barbatus}), ribbon (\emph{Histriophoca fasciata}), and spotted seals (\emph{Phoca largha}) in the Bering and Chukchi seas. Because ringed seals (\emph{Phoca (pusa) hispida}) can be out of the water but hidden from view in snow lairs analysis of their `availability' to surveys requires special consideration; therefore,they were not included in this analysis. Using generalized linear mixed pseudo-models to properly account for temporal autocorrelation, we fit models with covariates of interest (e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure, temperature, precipitation) to examine their ability to explain variation in haul-out probability. We found evidence for strong diel and within-season patterns in haul-out behavior, as well as strong weather effects (particularly wind and temperature). In general, seals were more likely to haul out on ice in the middle of the day and when wind speed was low and temperatures were higher. Haul-out probability increased through March and April, peaking in May and early June before declining again. The timing and frequency of haul-out events also varied based on species and age-sex class. For ribbon and spotted seals, models with year effects were highly supported, indicating that the timing and magnitude of haul-out behavior varied among years. However, we did not find broad evidence that haul-out timing was linked to annual sea ice extent. Our analysis emphasizes the importance of accounting for seasonal and temporal variation in haul-out behavior, as well as associated environmental covariates, when interpreting the number of seals counted in aerial surveys.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\hypertarget{introduction}{%
\section*{Introduction}\label{introduction}}
\addcontentsline{toc}{section}{Introduction}

Global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence (Meier et al., 2014; Wang et al., 2018; Kwok, 2018; Overland, 2021). These changes have tangible effects on
Arctic organisms, ecosystems, and the people who live in the region
(Huntington et al., 2020). Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering and Chukchi seas as a platform for important life history
functions, such as pupping, nursing, breeding behavior, and molting
(Boveng et al., 2009, 2013; Cameron et al., 2010; Kelly et al., 2010). Limited data and large
knowledge gaps complicate predictions about the ultimate effects of changes in
sea ice on the behavior, health, abundance, and distribution of these seals. To
date, indices of seal health sampled during periods of declining sea ice differ
regionally (Crawford, Quakenbush \& Citta, 2015; Harwood et al., 2020). Knowledge about evolutionary
constraints on the timing of reproductive and molting behavior is generally
lacking, so it is difficult to predict how or if ice-associated seal species
will adjust to future changes (e.g., by adjusting pupping or molting schedules to
earlier dates or different locales). This is further complicated by the
spatio-temporal variation in the phenology of these life history events within
regions and throughout their full ranges. Additionally, trends in abundance of
these species are unknown, so it is difficult to assess the effect, if any,
declines in sea ice habitat have had, or will have, on seal demography.

Statutory requirements (e.g., United States Marine Mammal Protection Act (MMPA),
United States Endangered Species Act (ESA)) for timely estimates of population
abundance and trend mean improved aerial survey effort is needed for these
species. Those survey efforts must also be paired with improved knowledge of
haul-out behavior to ensure appropriate survey design, robust methods, and
accurate estimates. Several studies have contributed estimates of the
distribution and abundance of ice-associated seal species in the Arctic using
aerial surveys (e.g.,
Bengtson et al. (2005), Conn et al. (2014), and Ver Hoef et al. (2014)) and more recent efforts have
significantly expanded on previous survey effort. Such abundance studies are
conducted over very large areas and estimation of absolute abundance requires
making inference about numerous issues affecting the observation of seals on
ice. These include availability (only seals on ice are available to be counted),
detection probability (observers or automated detection systems may miss some
seals on ice), species misclassification, and possible disturbance of seals by
aircraft (Ver Hoef et al., 2014; Conn et al., 2014). Refining these inferences will improve the
accuracy of abundance estimates and, hopefully, allow credible predictions about
the effects of climate disruptions on the abundance and distribution of Arctic
seal populations.

How ice-associated seals use sea ice as a haul-out platform varies among
species. Ribbon seals (\emph{Histriophoca fasciata}) haul out of the water almost
exclusively on sea ice and are mostly pelagic outside the spring pupping,
breeding, and molting season (Boveng \& Lowry, 2018). Although spotted (\emph{Phoca largha})
and bearded (\emph{Erignathus barbatus}) seals occasionally rest on coastal land,
they primarily use sea ice as a haul-out platform during the spring and early
summer (Frost \& Burns, 2018). Ringed seals (\emph{Phoca (pusa) hispida}) --- not included in this
study --- haul out on sea-ice but primarily use snow lairs on sea-ice during
winter and spring.

The remoteness of the Bering and Chukchi seas means direct scientific
observation of seal behavior is impractical. Thus, bio-logging devices are
especially useful tools for collecting key information on movement and haul-out
behavior for these species. Bio-logging records of time spent out of the water
provide valuable data for identifying covariates that explain variation in
haul-out behavior. For instance,Von Duyke et al. (2020) used
satellite-linked bio-loggers to corroborate seasonal changes between diurnal and
nocturnal haul-out behavior of ringed seals previously described by Kelly and
Quakenbush (1990) using VHF radio tags and direct observation. Bengtson
et al. (2005) documented a higher propensity for ringed seal basking
near solar noon, as did Ver Hoef et al. (2014) in an analysis of
bearded, ribbon, and spotted seals using much larger sample sizes. Olnes et al.
(2020) showed that the proportion of time bearded seals spent hauled out
progressively increased through spring and summer, and Ver Hoef et al.
(2014) found haul-out probabilities increased gradually starting in
March and peaked in May and June for bearded, ribbon, and spotted seals. Such
analyses have not been limited to the Arctic. In the Antarctic, Bengtson and
Cameron (2004) relied on bio-logging data to demonstrate greater
haul-out propensity in juvenile crabeater seals (\emph{Lobodon carcinophaga}) than
adults, with highest probabilities in February and at times close to solar noon.

Knowledge of haul-out patterns is not only important for understanding natural
history and ecology, but also for developing ``availability'' correction factors
for aerial surveys. Specifically, researchers need to know the fraction of seals
hauled out (versus in the water) when aerial surveys are conducted. Studies
estimating availability correction factors for seals typically use logistic
regression-style analyses to estimate the time-specific probability of being
hauled out based on `wet/dry' data relayed by bio-loggers. In these models,
haul-out probabilities were expressed as a function of predictive covariates,
such as time-of-day, day-of-year, sex, age class, and environmental conditions
(e.g., Reder et al. (2003), Bengtson \& Cameron (2004), Bengtson et al. (2005), Udevitz et al. (2009), Lonergan et al. (2013),
Ver Hoef et al. (2014), Southwell et al. (2008), and Niemi et al. (2023)). However, sample sizes have often been
insufficient to permit strong inference about demographic and/or seasonal
variation in haul-out probabilities. For instance, Bengtson and Cameron's
(2004) study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s (2005) study was based on 6 ringed seals in the
Chukchi and Beaufort seas. These studies were often further limited by
logistical constraints on fieldwork and the attachment duration or operational
life of bio-loggers. In this study, we addressed some of these limitations by
deploying small bio-loggers designed for longer-term attachment on rear flippers
of a subset of the study individuals. These devices are designed to collect data
through the molt period (when those adhered to the hair would fall off) and, in
some situations, provide multiple years of data.

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we sought to refine
estimates of haul-out availability corrections for aerial surveys in order to
improve estimates of seal abundance. Previously estimated availability
correction factors (e.g., Bengtson et al. (2005), Conn et al. (2014), and Ver Hoef et al. (2014))
accounted for variables such as the time-of-day and day-of-year, but did not
investigate the impact of weather variables. Such variables have been
shown to influence walrus haul-out behavior (Udevitz et al., 2009) and we expect weather
conditions to also influence seal haul-out behavior and including them within
the model framework will benefit our estimates of seal availability during
aerial surveys. Third, we aimed to assess the annual variability in haul-out
timing and possible linkage to changes in the extent of seasonal sea ice between
2005 and 2020. Our work extends the scope of previous haul-out analyses,
includes the influence of weather variability, and investigates the
potential impact of changing sea-ice extent on the behavior of these species.

\hypertarget{methods}{%
\section*{Methods}\label{methods}}
\addcontentsline{toc}{section}{Methods}

\hypertarget{data-collection}{%
\subsection*{Data collection}\label{data-collection}}
\addcontentsline{toc}{subsection}{Data collection}

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2021. Seals were captured using
nets and bio-loggers were attached during studies based in coastal communities
or on research ships. Ship-based capture events occurred during spring near the
southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
events occurred between 2005 and 2020 from May to October, generally between the
Alaska coastal communities of Scammon Bay in the Bering Sea, Utqiaġvik
in the Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental
Material, \protect\hyperlink{s1}{S1}). Data from additional
deployments along the Kamchatka peninsula in the western Bering Sea are also
included. We refer readers to Figure \ref{fig:deployplot} and the primary literature
for detailed capture and bio-logger attachment methods (see also Supplemental
Material, \protect\hyperlink{s1}{S1}).

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/deployplot-1} \caption{\textbf{Initial bio-logger deployment areas for bearded, ribbon, and spotted seals in the study between 2005 and 2020 across multiple regions within the Bering, Chukchi, and western Beaufort seas.} \linebreak Solid regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility. The larger, shaded region indicates the spatial extent of aerial survey effort to date in the Bering, Chukchi, and Beaufort seas. Deployments were initiated across a range of months but only haul-out data from 1 March to 15 July were included in the analysis. See Figure \ref{fig:dataMap} for the spatial distribution of observed data used in the study and Supplemental Materials \ref{tab:deploymentTable} for additional deployment details.}\label{fig:deployplot}
\end{figure}

We created a subset of haul-out behavior data from
250 bio-loggers deployed on
35 bearded,
110 ribbon, and
105 spotted seals
to include only records from 1 March to 15 July between
2005 and
2021. Bio-loggers were of the
`SPLASH' or `SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA). Deployments consisted of either a single `SPLASH' device, a
single `SPOT' device, or both types. Devices were either adhered to the hair on
the seal or attached through the rear flipper inter-digital webbing. The use of
bio-loggers adhered to the back or head provides some benefits over flipper
mounted devices (e.g.~increased satellite transmittal rates, locations at sea)
but these fall off during the following annual molt, which, depending on
deployment date, limits the duration of haul-out data they provide especially
during the focus months of our study. Additionally, bio-loggers attached to the
head or dorsal region are often dry while the seal is floating at the surface,
inducing a slight positive bias in the hourly percent-dry values reported by the
bio-logger. For this study, in cases where both bio-logger types were deployed,
we preferred hourly percent-dry observations from the flipper tag. All data were
transmitted by the deployed instruments via the Argos satellite network and
location data were either derived from Argos transmissions or transmitted
FastLoc GPS data.

Sex and age class (non-dependent \emph{young-of-the-year}, sexually immature
\emph{subadults}, and mature \emph{adults}) were estimated at the time of deployment by
various combinations of length, claw growth ridges (McLaren, 1958), and pelage
characteristics. Seals determined to be less than one year were classified as
young-of-the-year. For those bio-loggers deployed on young-of-the-year and
transmitting into the next year, the age class was advanced to subadult on 1
March of the following year -- the assumed anniversary of their birth. Subadults
are those seals likely greater than one year of age but less than four years.
Adults are individuals that are likely older than four years. Table
\ref{tab:tblDeploy} provides a summary of these deployments and data received
from them.

\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}

\global\setlength{\Oldtabcolsep}{\tabcolsep}

\setlength{\tabcolsep}{0pt}

\renewcommand*{\arraystretch}{1.5}

\providecommand{\ascline}[3]{\noalign{\global\arrayrulewidth #1}\arrayrulecolor[HTML]{#2}\cline{#3}}

\begin{longtable}[c]{|p{0.90in}|p{0.40in}|p{1.35in}|p{1.35in}|p{1.35in}}

\caption{\textbf{Summary of bio-logger data by seal species and age classification from 1 March to 15 July 2005-2021.} \linebreak Seal hours represents the sum of hourly observations across all seals used in the analysis. Because young-of-the-year are advanced to subadult on 1 March of the following year, some individual seals are represented in both columns in this table}\label{tab:tblDeploy}\\

\ascline{1.5pt}{666666}{1-5}

\multicolumn{2}{>{\centering}m{\dimexpr 1.3in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{3}{>{\centering}m{\dimexpr 4.05in+4\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Age\ Class}}} \\

\ascline{1.5pt}{666666}{1-5}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.9in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Species}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Sex}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Adult}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Subadult}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Young-of-the-Year}}} \\

\ascline{1.5pt}{666666}{1-5}\endfirsthead \caption[]{\textbf{Summary of bio-logger data by seal species and age classification from 1 March to 15 July 2005-2021.} \linebreak Seal hours represents the sum of hourly observations across all seals used in the analysis. Because young-of-the-year are advanced to subadult on 1 March of the following year, some individual seals are represented in both columns in this table}\label{tab:tblDeploy}\\

\ascline{1.5pt}{666666}{1-5}

\multicolumn{2}{>{\centering}m{\dimexpr 1.3in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{3}{>{\centering}m{\dimexpr 4.05in+4\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Age\ Class}}} \\

\ascline{1.5pt}{666666}{1-5}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.9in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Species}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Sex}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Adult}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Subadult}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Young-of-the-Year}}} \\

\ascline{1.5pt}{666666}{1-5}\endhead

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.9in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bearded\ seal}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1\ (\ 1,776\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{16\ (21,648\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.9in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bearded\ seal}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2\ (\ 2,108\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{16\ (17,232\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.9in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Ribbon\ seal}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{33\ (35,128\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{18\ (15,984\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{13\ (\ 3,734\ seal\ hours)}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.9in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Ribbon\ seal}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{24\ (27,465\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{19\ (13,046\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{9\ (\ 4,228\ seal\ hours)}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.9in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Spotted\ seal}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{23\ (21,588\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{21\ (19,559\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{11\ (13,417\ seal\ hours)}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.9in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Spotted\ seal}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{20\ (31,793\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{21\ (17,210\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{12\ (11,285\ seal\ hours)}}} \\

\ascline{1.5pt}{666666}{1-5}

\end{longtable}

\arrayrulecolor[HTML]{000000}

\global\setlength{\arrayrulewidth}{\Oldarrayrulewidth}

\global\setlength{\tabcolsep}{\Oldtabcolsep}

\renewcommand*{\arraystretch}{1}

Haul-out behavior data were recorded in a manner standard across Wildlife
Computers bio-loggers and transmitted via the Argos satellite network as hourly
percent-dry timelines. For each hour of a day, the wet/dry sensor was polled by
the tag firmware every few seconds and the percent of the hour in the dry state
was calculated (Figure \ref{fig:examplePlot}). On board the bio-logger, hourly
percent-dry calculations were rounded to the nearest 10\% inclusive of 0\% and
100\% along with additional values at 3\% and 98\%. This compression resulted in
additional data transmission as each message consisted of two complete 24-hour
records. Memory capacity allowed caching of percent-dry records for several
weeks or months and each message was transmitted several times to ensure
reception at the satellite. Bio-loggers were deployed and programmed in a manner
to maximize data transmission during the spring pupping and molting period,
though hourly percent-dry data were not always successfully transmitted. This is
due to a variety of factors including satellite coverage, tag availability
(e.g., tags mounted to the rear flipper often do not transmit while at sea), tag
performance, duty cycling, and atmospheric interference. Fortunately, missing
records do not substantially bias inference about haul-out probabilities
(Conn et al., 2012).

Tags that fell off due to molt, attachment failure, or seal mortality and remained
on ice or land may have continued to send data to satellites; i.e., a detached
bio-logger that is dry (either on ice or land) will record and transmit data
suggesting the seal is hauled out. Therefore, end times of each deployment were
identified by examining bio-logger locations, percent-dry records, and dive
behavior (if available) to determine when bio-loggers ceased providing data
consistent with seal behavior. For example, a data record that ended with several
consecutive days (\textasciitilde10+ days) of 100\% dry observations and with locations
indicating the tag was on land were truncated to the final
stretch of 100\% dry observations. The vast majority of deployments ended with the
device detaching in the water and the deployment end date was obvious. There is
no perfect algorithm for identifying deployment end dates and each deployment in
question was considered separately. While not perfect, we are confident our
reliance on expert opinion and examination of multiple data streams provided the
best option. Data outside of the deployment start and end times were discarded
prior to analysis.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/examplePlot-1} \caption{\textbf{Example percent-dry actogram from bio-logger data}
\linebreak Haul-out behavior observations recorded by a bio-logger deployed on a
ribbon seal over two years during the months of March, April, and May. This
actogram plot represents the transmitted hourly percent-dry values. Not all data
during a deployment are reliably transmitted from the bio-logger and because data
are transmitted in 24-hour chunks occasional missing days are present in the
record. Here, missing data that were not successfully received from the device
are represented as empty rectangles.}\label{fig:examplePlot}
\end{figure}

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and seal location. The use of modern
bio-loggers that record and transmit behavioral data while simultaneously
providing location estimates was key to this objective. We explored the use
of a continuous-time correlated random walk (Johnson et al., 2008) movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
\emph{3} = 250 m, \emph{2} = 500 m, \emph{1} = 1500 m, \emph{0} = 2500 m (Lopez et al., 2013); we chose 2500 m for
location classes \emph{A} and \emph{B}) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50 m. On days when haul-out observations were
present but location data were missing we used the seal's last calculated weighted
average daily location; days when the location intersected with land
were removed from the seal's record. We recognize that bearded and spotted
seals haul out on land. However, assessing the relationship between haul-out
behavior and weather covariates and seals' availability for aerial surveys on
land was outside the scope of this study. Additionally, any daily locations on
land were likely more reflective of coordinate averaging and measurement error,
rather than actual use of coastal features.

\hypertarget{explanatory-variables}{%
\subsection*{Explanatory variables}\label{explanatory-variables}}
\addcontentsline{toc}{subsection}{Explanatory variables}

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). We calculated local solar
hour using the \{solaR\} package (Perpiñán, 2012) within the R statistical
environment (R Core Team, 2021) based on the weighted daily average locations. We
also linked the weighted average daily locations to weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction (Mesinger et al., 2006). The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made available
across the region 8 times daily. For this study, NARR weather values were limited
to the extent of our study area over the Bering, Chukchi, and Beaufort seas at 3-hr
intervals based on the original grid resolution of 32 km (1024 km\textsuperscript{2}). The
following weather variables are known to affect haul-out behavior in
other Arctic pinnipeds (Reder et al., 2003; Udevitz et al., 2009; Perry, Stenson \& Buren, 2017) and were
interpolated and assigned to daily seal locations using a bilinear method: 1)
air temperature at 2 m above the Earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the
Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation
(Table \ref{tab:tblCovariates}).

For all seal species, we considered the following variables when modeling the
hourly haul-out behavior: day-of-year, solar hour, temperature, wind speed,
barometric pressure, precipitation, and wind chill (represented by a
\emph{wind:temperature} interaction (Udevitz et al., 2009)). Ribbon and spotted seal models
included age-sex class and interactions between day-of-year and age-sex class,
but we omitted these from bearded seal models due to poor representation of
age-sex classes (Table \ref{tab:tblDeploy}). Bearded seal models included a
latitudinal effect (and an interaction with day-of-year), because bearded seals
occupy a substantial range during the spring and we were interested in possible
differences in the timing of haul-out behavior along a latitudinal gradient. We
omitted the latitudinal effect from ribbon and spotted seal models because,
during the spring, these species are most prevalent near the southern ice edge
in the Bering Sea (Conn et al., 2014).

\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}

\global\setlength{\Oldtabcolsep}{\tabcolsep}

\setlength{\tabcolsep}{0pt}

\renewcommand*{\arraystretch}{1.5}

\providecommand{\ascline}[3]{\noalign{\global\arrayrulewidth #1}\arrayrulecolor[HTML]{#2}\cline{#3}}

\begin{longtable}[c]{|p{1.00in}|p{1.00in}|p{2.50in}}

\caption{\textbf{Explanatory covariates used in analyses of binary haul-out records for bearded, spotted, and ribbon seals.} \linebreak Note that we also considered select interactions (see article text) between these primary covariates. For instance, wind chill was represented by the interaction temperature:wind.}\label{tab:tblCovariates}\\

\ascline{1.5pt}{666666}{1-3}

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Covariate}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Type}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Description}}} \\

\ascline{1.5pt}{666666}{1-3}\endfirsthead \caption[]{\textbf{Explanatory covariates used in analyses of binary haul-out records for bearded, spotted, and ribbon seals.} \linebreak Note that we also considered select interactions (see article text) between these primary covariates. For instance, wind chill was represented by the interaction temperature:wind.}\label{tab:tblCovariates}\\

\ascline{1.5pt}{666666}{1-3}

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Covariate}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Type}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Description}}} \\

\ascline{1.5pt}{666666}{1-3}\endhead

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Age-sex\ class}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Categorical}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{young-of-the-year,\ subadult,\ adult\ male\ and\ adult\ female}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Hour}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Continuous;\ Fourier\ basis}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{local\ solar\ hour\ using\ 6\ variables\ of\ a\ Fourier-series\ basis}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Day}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Continuous}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{linear,\ quadratic,\ and\ cubic\ effects\ of\ day-of-year}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Precip}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Continuous}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{convective\ precipitation\ kg/m\textsuperscript{2}\ (NARR)}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Pressure}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Continuous}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{atmospheric\ pressure\ at\ sea\ level\ (kPa)\ (NARR)}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Temp}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Continuous}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{air\ temperature\ (C)\ at\ 2m\ above\ the\ earth’s\ surface\ (NARR)}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Wind}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Continuous}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{northerly\ and\ easterly\ vector\ components\ for\ wind\ (NARR)\ converted\ into\ a\ single\ wind\ speed\ via\ the\ Euclidean\ norm}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Northing}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Continuous}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{latitude\ divided\ by\ the\ mean\ latitude\ across\ all\ locations\ (for\ bearded\ seals\ only)}}} \\

\ascline{1pt}{666666}{1-3}

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Year}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Continuous}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{For\ the\ set\ of\ models\ examining\ inter-annual\ variation\ in\ sea\ ice\ use,\ we\ fitted\ models\ with\ the\ addition\ of\ year\ by\ day-of-year\ interactions.}}} \\

\ascline{1.5pt}{666666}{1-3}

\end{longtable}

\arrayrulecolor[HTML]{000000}

\global\setlength{\arrayrulewidth}{\Oldarrayrulewidth}

\global\setlength{\tabcolsep}{\Oldtabcolsep}

\renewcommand*{\arraystretch}{1}

\hypertarget{haul-out-modeling}{%
\subsection*{Haul-out modeling}\label{haul-out-modeling}}
\addcontentsline{toc}{subsection}{Haul-out modeling}

Haul-out records for seals are often characterized by sequential hours spent
hauled out on ice alternating with long periods in the water (Figure
\ref{fig:examplePlot}). Commonly used statistical models for binary data (e.g.
logistic regression) assume independence among responses, an assumption that is
clearly violated if hourly responses are modeled. Any analysis that ignores
temporal autocorrelation in responses will thus have overstated precision
(Betts et al., 2006).

To properly account for temporal dependence and to take advantage of
computational efficiency, we used generalized linear mixed pseudo-models
(GLMPMs; (Ver Hoef, London \& Boveng, 2010)) to model variation in haul-out behavior as a function
of (1) covariate predictors, (2) temporally autocorrelated random effects, and
(3) individual random effects representing heterogeneity in individual behavior.
We used the glmmLDTS package (Ver Hoef, London \& Boveng, 2010) to implement GLMPMs. We explored
two different model formulations for our data and we fit separate models to
bearded, ribbon, and spotted seal data sets as we expected differing behavior by
species. Separate models for each species were also needed because a single,
very large data set proved computationally intractable. In our first model
formulation and for each species, we fitted a year-independent model that
predicted average haul-out behavior as a function of demographic, weather,
seasonal, and diurnal effects. Second, for ribbon and spotted seals (which had
considerably more data than bearded seals), we fitted models that included all
the effects from the first model, but also permitted annual variation in
haul-out timing. This second set of models was used to examine whether
haul-out patterns varied by year and to determine the annual timing of apparent
peaks in haul-out behavior. For both models, we assumed an hourly Bernoulli
response (a binary 0-1 response dependent upon whether the tag was greater than
50\% dry for a given hour) where the linear predictor was
modeled on the logit scale. This is
consistent with previous approaches (London et al., 2012; Ver Hoef et al., 2014) and only
6.995\% of our
observations fell between 10\% and 90\% hourly percent-dry.

We followed Ver Hoef et al. (2014) in using linear, quadratic, and cubic
effects of day-of-year to represent temporal changes
in behavior during the season. However, unlike previous models for harbor seals
(London et al., 2012) and ice-associated seals (Ver Hoef et al., 2014), which treated
hour-of-day as a 24-level categorical variable to capture diurnal cycles, we
adopted a continuous formulation based on Fourier series that provides a
flexible model while preserving the inherent circularity needed for time-of-day
effects (i.e., hour 0 should be equal to hour 24). It also represents
hour-of-day with 6 parameters, which is a considerable reduction when compared
to a 24-parameter variable. According to this approach, we used the following
specification for hour-of-day effects:

\[
H_t=\alpha_1 \cos(\frac{\pi t}{4}) + \alpha_2 \sin(\frac{\pi t}{4}) + \alpha_3 \cos(\frac{\pi t}{6}) + \alpha_4 \sin(\frac{\pi t}{6}) + \alpha_5 \cos(\frac{\pi t}{12}) + \alpha_6 \sin(\frac{\pi t}{12})
\]
where \(H_t\) gives the effect for solar hour \(t\) and \(\alpha_{i}\) are
estimated parameters (regression coefficients).

For the second set of models examining inter-annual variation in sea ice use, we
fitted models with year by day-of-year interactions. However, in this case we
only included \emph{year:day} and \emph{year:day\textsuperscript{2}}, omitting the main effects of year as
well as \emph{year:day\textsuperscript{3}} interactions because models with the latter effects were
numerically unstable. However, the modeled interactions were sufficient to allow
shifts in haul-out distribution, as one can show mathematically that a simple
horizontal shift in timing of haul-out distributions does not affect the main
effects or cubic terms in a polynomial regression model. Bearded seals were
not included in this examination of inter-annual variation because of limited
data across many years in the study.

A typical model fitting exercise would also include a model selection process.
However, AIC (and similar criteria) is not suitable when using
pseudo-likelihoods, because pseudo-data generated in the model fitting process
(Ver Hoef, London \& Boveng, 2010) differ between models (Ten Eyck \& Cavanaugh, 2018). After fitting GLMPM
models, we instead used ``type III'' \(F\)-tests to calculate \(p\)-values
(Ver Hoef, London \& Boveng, 2010) to evaluate model performance and important terms. We also
produced predictions of haul-out behavior as a function of three influential
predictors (e.g.~solar hour, day-of-year, age-sex). Weather covariates for these
predictions were based on daily or hourly smoothed weather covariate values
across the study region. Such predictions were then used to develop haul-out
probability surfaces, explore conditional effects of weather covariates, and
determine annual peaks in haul-out activity. The timing of peak haul-out
behavior was further used to regress against the annual maximum sea ice extent
in the study region. Predictions before 15 March and after 30 June were not
included in visualizations or other evaluations to avoid spurious model
predictions at the edge of the data range.

Visualizing the marginal or conditional effect of an individual weather
covariate (where all other weather covariates are being held at mean values) on
haul-out probability was difficult in this analysis because of the collinearity
between covariates as well as the spatial and temporal variation across such a
large region. The relationship of each weather covariate with haul-out
probability, averaged over the other weather conditions, was more variable than
model coefficients would imply. That said, important insights can be gained from
plots of marginal effects. To create these plots, we predicted haul-out
probability across the full range of each weather covariate while fixing hour
of the day at local solar noon and day-of-year at 15 May. For the other weather
covariate values, we chose not to use a fixed mean value because we expect weather
to vary within day over the season (e.g.~the temperature at solar noon will
gradually increase from March through June). To account for this, we fit a simple generalized
additive model for each weather covariate with smooth terms for day-of-year and
solar hour. We used predicted values from the generalized additive model in lieu
of holding other weather covariates at a fixed mean value which would not capture
seasonal change. The visualizations
also include vertical lines representing 95\% confidence intervals around the
predicted haul-out probability to better communicate the variation in model
uncertainty.

We also assessed whether the annual variation in maximum spring sea ice extent in the
Bering Sea influenced the seasonal peak of seal haul-out probability. In
particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
SSM/I-SSMIS Passive Microwave Dataset, Version 1 (Cavalieri et al., 1996) to calculate
maximum sea ice extent. All sea ice concentration grid cells (25 km\textsuperscript{2}) in the
study area with greater than 15\% concentration were counted daily to get the total sea
ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily count of grid cells
with greater than 15\% concentration for each year. A separate regression model,
built on the results of the haul-out model, was used to evaluate the relationship
between the annual computed peak haul-out day (as the response) with the
maximum sea ice extent (as the predictor).

\hypertarget{results}{%
\section*{Results}\label{results}}
\addcontentsline{toc}{section}{Results}

Figure \ref{fig:dataMap} shows the spatial distribution of weighted locations
with available haul-out behavior data used for analysis of each species across
the study area. Figure \ref{fig:dataCal} shows the temporal distribution of all
haul-out data across the study season for each species. Observations of ribbon
and spotted seals were concentrated in the months of May and June due to the
timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the
water and more data are transmitted. Molt timing also impacts when many
deployments end as any bio-loggers adhered to the hair will fall off. Relative
to the other species in the study, there were fewer deployments of bio-loggers
on bearded seals. This resulted in fewer data observations overall and
noticeably lower in numbers May and June. The majority of deployments on bearded
seals occurred in August and September and, by May, bio-loggers had either
fallen off or their batteries were depleted. Therefore, observations for bearded
seals were concentrated in March (Figure \ref{fig:dataCal}).

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/dataMap-1} \caption{\textbf{Spatial distribution of haul-out records from 1 March through July 15 for bearded, ribbon, and spotted seal.} \linebreak Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model. For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km\textsuperscript{2}}\label{fig:dataMap}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/dataCal-1} \caption{\textbf{Seasonal distribution of haul-out behavior observations by species} \linebreak Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species. Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020. Seal-hours of data are represented as a color gradient and bar height. The higher amount of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, many bio-logger deployments started in April and May. The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study and the fact that the majority of deployments on bearded seals occurred in August and September; therefore most bio-loggers deployed had exhausted their batteries.}\label{fig:dataCal}
\end{figure}

Across all three seal species, generally, models omitting year effects suggested
that day-of-year, solar hour, age-sex class, temperature, and wind substantially
influenced haul-out behavior of all three species, with \(F\) tests producing
\(p\)-values less than 0.05 for variables embodying these effects and/or their
interactions. Haul-out probabilities typically increased throughout March and
April, reaching a peak in May and early June before declining again. Diurnal
patterns were present, with maximum haul-out behavior centered around local
solar noon.

\hypertarget{bearded-seals}{%
\subsection*{Bearded Seals}\label{bearded-seals}}
\addcontentsline{toc}{subsection}{Bearded Seals}

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages (Figure \ref{fig:beardedHOCal}; see also
\ref{fig:beardedPredSE}). Additionally, after approximately 9 months, 7 devices
deployed on the rear flipper of bearded seals reported implausible hourly
percent-dry data (100\% dry for several weeks but indicative of movement and
increasing transmission rates (see Boveng \& Cameron (2013))). All data after the first
instance of unrealistic values were censored from this analysis. In addition to
a peak around local solar noon, the bearded seal model predicted additional
haul-out activity around local midnight. In concert with the lower magnitude of
haul-out probability, bearded seal haul-out behavior was also more protracted
throughout the spring season compared to ribbon (Figure \ref{fig:ribbonHOCal})
and spotted seals (Figure \ref{fig:spottedHOCal}). Overall, bearded seals were
less likely to haul out and had a bi-modal distribution of haul-out probability
across the day.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/beardedHOCal-1} \caption{\textbf{Bearded seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 15 March to 30 June for all age and sex classes combined. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range. Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. Note, our model predicts lower haul-out probability for bearded seals overall compared to ribbon and spotted seals and the color scales are not directly comparable.}\label{fig:beardedHOCal}
\end{figure}

When exploring the influence of weather, bearded seal haul-out probability
was strongly affected by wind
(\(F_{1,42728}\)
= 130.468; \(p\) = \textless0.001)
and temperature
(\(F_{1,42728}\)
= 19.5; \(p\) =
\textless0.001) with much higher haul-out probability during
periods of higher temperatures and low wind speeds (Figure \ref{fig:beardedHOwx}).
Not surprisingly, wind chill
(\(F_{1,42728}\)
= 14.54; \(p\) =
\textless0.001) was also important. Barometric
pressure (\(F_{1,42728}\)
= 7.779; \(p\) =
0.005) was also significant factor although less
apparent (Figure \ref{fig:beardedHOwx}). Any effect of precipitation
was not a significant influence on haul-out probability
(\(F_{1,42728}\)
= 0.519; \(p\) =
0.471).

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/beardedHOwx-1} \caption{\textbf{Influence of weather covariates on bearded seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications. Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May. The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. While the marginal effect of the covariate is continuous, points and vertical lines representing the 95\% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.}\label{fig:beardedHOwx}
\end{figure}

\hypertarget{ribbon-seals}{%
\subsection*{Ribbon Seals}\label{ribbon-seals}}
\addcontentsline{toc}{subsection}{Ribbon Seals}

Ribbon seals exhibited a pattern of gradually increasing haul-out probability in
April that peaks in late May for subadults and in early June for adults (Figure
\ref{fig:ribbonHOCal}; see also \ref{fig:ribbonPredSE}). There is an apparent
seasonal progression with subadults hauling out earlier in the season followed
by adult males and, then, adult females. Haul-out behavior was clearly centered
around local solar noon and expanded to other hours later in the spring as seals
entered their molting period. Subadults showed an earlier start and more intense
haul-out activity in April and May. The young-of-the-year records began after
weaning and the model predictions seemed to indicate development of in-water
activities (e.g.~swimming, foraging) in May. Adult females had a more protracted
haul-out season compared to males, and more time was spent hauled out in June
compared to adult males and subadults.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/ribbonHOCal-1} \caption{\textbf{Ribbon seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model. Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. Predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range.}\label{fig:ribbonHOCal}
\end{figure}

The haul-out probability for ribbon seals was mostly influenced by
temperature
(\(F_{1,99540}\)
= 6.87; \(p\) = 0.009) and wind
(\(F_{1,99540}\)
= 49.314; \(p\) = \textless0.001) with barometric pressure
(\(F_{1,99540}\)
= 3.446; \(p\) =
0.063) having a milder impact.
Ribbon seals were more likely to haul out when temperatures were
relatively warm and less likely to haul out at higher winds and lower pressure
values (Figure \ref{fig:ribbonHOwx}). Neither wind chill
(\(F_{1,99540}\)
= 1.83; \(p\) =
0.176) nor precipitation
(\(F_{1,99540}\)
= 0; \(p\) =
0.989) were a significant
influence on haul-out probability. Compared with bearded seals, the effect of
weather covariates on the predicted haul-out probability for
ribbon seals was less striking. Because our ribbon seal model included age and sex class, we can
visualize the different influences of weather covariates on those classes and see
that subadults differ from adult males and females (Figure \ref{fig:ribbonHOwx}).

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/ribbonHOwx-1} \caption{\textbf{Influence of weather covariates on ribbon seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classifications. Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May. The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. While the marginal effect of the covariate is continuous, points and vertical lines representing the 95\% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.}\label{fig:ribbonHOwx}
\end{figure}

\hypertarget{spotted-seals}{%
\subsection*{Spotted Seals}\label{spotted-seals}}
\addcontentsline{toc}{subsection}{Spotted Seals}

Compared to ribbon seals, spotted seals showed a longer spring haul-out season
that was less intensely centered on solar noon (Figure \ref{fig:spottedHOCal};
see also \ref{fig:spottedPredSE}). Adults of both sexes spent considerable time
in April, May and June hauled out. Adult spotted seal males had a more
protracted haul-out season compared to females, and more time out of the water
in June (Figure \ref{fig:spottedHOCal}). This likely reflects the triad
behavior in spotted seals when suitor males haul out with nursing females. As
with ribbon seals, the young-of-the-year records began after weaning and the
model predictions reflected development of in-water activities (e.g.
swimming, foraging) in May.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/spottedHOCal-1} \caption{\textbf{Spotted seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of spotted seals from 1 March to 30 June for each age and sex class used in the model. Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. As with ribbon seals, predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range.}\label{fig:spottedHOCal}
\end{figure}

Spotted seal haul-out behavior was less affected by the weather covariates
compared to ribbon and bearded seals but their influence on the model was still
significant in some cases.
Temperature
(\(F_{1,114807}\)
= 5.462; \(p\) = 0.019), wind
(\(F_{1,114807}\)
= 46.954; \(p\) = \textless0.001),
and barometric pressure
(\(F_{1,114807}\)
= 10.214; \(p\) =
0.001) were all significant. Spotted seals were more likely to be on the ice when
temperatures were relatively warm and less likely to
haul out at higher wind speeds. Wind chill
(\(F_{1,114807}\)
= 0.559; \(p\) =
0.455) and
precipitation
(\(F_{1,114807}\)
= 0.763; \(p\) =
0.382)
were not as influential as the other covariates.
Differences in the magnitude of response between the age-sex classes were
present and consistent across each of the weather covariates (Figure
\ref{fig:spottedHOwx}). There was a consistent ranking of adult males being the most
likely to haul out, followed by adult females, and, then, subadults. This differs
from ribbon seals which showed more overlap between adult males and adult females
and that subadults were most likely to haul out across the presented range of weather
covariates.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/spottedHOwx-1} \caption{\textbf{Influence of weather covariates on spotted seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification. Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May. The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.. While the marginal effect of the covariate is continuous, points and vertical lines representing the 95\% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.}\label{fig:spottedHOwx}
\end{figure}

\hypertarget{annual-variation-in-haul-out-timing}{%
\subsection*{Annual variation in haul-out timing}\label{annual-variation-in-haul-out-timing}}
\addcontentsline{toc}{subsection}{Annual variation in haul-out timing}

The second set of models, which included annual variation in haul-out
patterns, uncovered significant contributions for linear and quadratic
interactions between day and year for only spotted seals (day:year,
\(F_{15,114762}\)
= 4.345; \(p\) =
\textless0.001; day\textsuperscript{2}:year,
\(F_{15,114762}\)
= 5.827; \(p\) =
\textless0.001). Ribbon seals showed no significant
contribution for interactions between day and year (day:year,
\(F_{10,99510}\)
= 0.516; \(p\) =
0.880; day\textsuperscript{2}:year,
\(F_{10,99510}\)
= 0.549; \(p\) =
0.856). Predicted distributions of
haul-out activity were largely unimodal, but varied some among and
within years with respect to the timing and magnitude of haul-out peaks
(Figure \ref{fig:annualHO}). It is important to note that predicted
variation in annual haul-out patterns likely reflected both process error
and sampling variability. While we did remove any years where only one
deployment in a species + age:sex group was present, there were still
some years where the pattern shown was informed by a small number of
individuals that may not represent population-level patterns.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/annualHO-1} \caption{\textbf{Annual variability in the timing of peak haul-out probability for ribbon and spotted seals.} \linebreak Spring haul-out probability (colored markers) for ribbon and spotted seals across 14 years are shown. The full seasonal range of annual predictions along with uncertainty are presented as grey lines. Covariates were fixed at local solar noon and under smoothed weather conditions. Only those groups (age:sex + year) that included observations from more than one seal are shown. Additionally, any groups where data were only available after 1 June or before 1 May are not included.}\label{fig:annualHO}
\end{figure}

The annual timing of peak haul-out probability for ribbon seals and adult
male spotted seals appeared to have
no relationship with the amount of yearly maximum sea ice extent in the
study area. For ribbon seals and adult male spotted seals,
\(p\)-values were substantially larger than 0.05 (ribbon seal adult
females; \(R^{2}\) = 0.004, \(p\) = 0.896;
ribbon seal adult males: \(R^{2}\) = 0.059, \(p\) =
0.693; ribbon seal subadults: \(R^{2}\) =
0.007, \(p\) = 0.828; spotted seals
adult males: \(R^{2}\) = 0.003, \(p\) =
0.903). Adult female and subadult spotted seals both
showed a negative trend (peak haul-out occurs later in years with less
sea ice) with a significant relationship for adult female spotted seals
(spotted seal adult female: \(R^{2}\) =
0.663, \(p\) = 0.049; spotted seal
subadults: \(R^{2}\) = 0.384, \(p\) =
0.056).

\hypertarget{discussion}{%
\section*{Discussion}\label{discussion}}
\addcontentsline{toc}{section}{Discussion}

We modeled data from bio-loggers deployed on bearded, ribbon, and spotted seals
to examine factors affecting their haul-out behavior on sea ice in the Bering
and Chukchi seas. Our analysis shows all three species of seal haul out
progressively more through the spring and peak near mid-May to early June before
declining again. This pattern aligns well with what has been previously
documented qualitatively (Boveng et al., 2009; Cameron et al., 2010; Boveng \& Cameron, 2013) and confirms
our haul-out data are likely quantifying population-level behavioral patterns.
Beginning in spring, seals preferentially haul out on ice shortly after solar
noon, which allows seals to maximize absorption of solar radiation thought to
facilitate the molting process (Feltz \& Fay, 1966). Interestingly, bearded seals appear
to have two peaks in haul-out activity within a day, one shortly after solar
noon, and one centered near solar midnight. This, of course, could be an
artifact of our limited sample size for bearded seal deployments across all age
classes. However, a similar bi-modal pattern has been seen in ringed seals
(Von Duyke et al., 2020) and suggests that bearded and ringed seals may be operating
under different constraints than ribbon and spotted seals. Bearded and ringed
seals are distributed across higher latitudes higher latitudes that experience
extended daylight hours during spring which may allow more flexibility in
alternating resting and foraging events. Other factors such as predation by
polar bears (which is rare for ribbon and spotted seals in the Bering Sea) may
also explain differing haul-out patterns. The change in haul-out behavior during
the season was less pronounced in bearded seals compared to ribbon and spotted
seals. This aligns with findings from Thometz et al. (Thometz et al., 2021) who observed
a mean molting period of 119±2 days and a relatively stable resting metabolic
rate for bearded seals during that time. While ribbon seals were not considered
in that study, spotted and ringed seals underwent molt periods of just 33±4
and 28±6 days and had increased resting metabolic rates.

Unlike previous analyses of seal haul-out behavior in spotted and ribbon seals
(e.g. (Ver Hoef, London \& Boveng, 2010), (Conn et al., 2014)), we also investigated the influence of
sex-age class on haul-out probabilities of both species (but not for bearded seals
because of low sample size). Field identification of age class can be inexact,
particularly when differentiating subadults from adults. In the case of ribbon
seals, subadults often have less distinct ribbons than adults. Spotted seal
pelage cannot be used to reliably discern adults from subadults. Despite these
challenges, we feel the age classifications used in this analysis are useful in
testing for age-related effects on haul-out behavior.

Although ribbon and spotted seals exhibited a
unimodal diel haul-out pattern generally centered around local solar noon, there
were key differences across species, age, and sex that match our understanding
from natural history descriptions of their ecological behavior. Spotted seals
are known to form triads during the breeding season where a female and dependent
pup are accompanied on the ice by a suitor male (Frost \& Burns, 2018). The male waits for
the female to wean the pup and enter estrus, and fends off any other potential
suitor males. Triad formation results in both males and females spending a large
portion of the day hauled out on ice and a protracted spring haul-out season for
both sexes. While females are still nursing the pup and not yet in estrus they
may be less inclined to interrupt their haul out and enter the water where the
suitor male would attempt mating. We see this reflected in the predicted
haul-out patterns, with both males and females exhibiting a broad distribution
of time out of the water throughout the solar day and the season. Ribbon seals
are not known to form triads and our model predicts a progression of increased
haul-out behavior with females starting earlier in the season than males.
Notably, female ribbon seals spend a large portion of the day in the water
during the pupping period, aligning with the hypothesis that ribbon seal females
continue foraging while nursing. Subadult ribbon and spotted seals begin
elevated haul-out behavior earlier in the spring and follow a pattern seen in
other phocids where yearlings and subadults molt first followed by adult females
and males (Thompson \& Rothery, 1987; Kirkman et al., 2003; Reder et al., 2003).

We also investigated the influence of weather on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind
chill. These have been investigated for walruses (e.g.~Udevitz et al.
(2009)) and a few select studies of ice-associated seals
(Perry, Stenson \& Buren, 2017; Hamilton \& Evans, 2018). In our study, ribbon seal haul-out behavior seemed to be the most
influenced by weather, with wind, temperature, and barometric pressure all being
important components of the model. Spotted seals were most affected by wind and
barometric pressure. For bearded seals, the model indicated wind and temperature
had the greatest impact. In general, and as might be expected, seals were more
likely to haul-out when daily temperatures were warmer, winds speeds were lower,
barometric pressure was higher, and precipitation was lower. Those weather
conditions are general indicators of increased solar radiation and lower
convective heat loss, both of which provides energetic benefits (see additional
discussion in Supplemental Material \protect\hyperlink{exploring-insolation-solar-radiation-as-a-model-covariate}{Exploring Insolation (Solar Radiation) as a
Model Covariate} regarding the potential use of solar radiation directly). Low
winds and precipitation could also enhance predator detection. Our results
highlight the importance of incorporating weather covariates when analyzing
haul-out behavior and calculating availability corrections for aerial surveys.

Notably missing from our list of haul-out model explanatory variables is any
spatial-temporal representation of sea ice concentration, area, or extent. This
may seem counterintuitive when modeling the haul-out behavior of seal species
with such a close association to sea ice; seals haul out in the presence of sea
ice, and we could assess the local concentration of sea ice during these events
(see (Olnes et al., 2020)). This, however, expands the scope of our analysis into the
realm of habitat selection and many of our deployments consisted of a single
device attached to the rear flipper of the seal which meant we only received
locations when seals were hauled out on sea-ice, limiting our ability to fully
evaluate fine-scale habitat preferences related to sea-ice. Insight into how
seals use and interact with sea-ice during an extended period when the
availability and characteristics of sea-ice is rapidly changing is important
(Breed et al., 2018; Cameron et al., 2018) but ancillary to the focus of this analysis -- and,
in the end, not possible given key limitations of our data. Additionally, our
study was limited to the spring season when seal haul-out behaviors are strongly
influenced by pupping, nursing, breeding behavior, and molt. These drivers are
likely more influential on haul-out behavior than sea-ice concentration.
Crawford et al. (2019) compared haul-out probability models for ringed
seals and found those that only included season (and not sea-ice concentration)
were the most parsimonious. For all of these reasons, we have elected not to use
sea ice concentration \emph{as a predictor for haul-out probability} in the present
study.

Our models detected small deviations in the timing and magnitude of annual peaks
in haul-out behavior for ribbon and spotted seals. The timing of peak haul-out
activity appears to fall within a relatively narrow time window of 3-4 weeks in
late May and early June. This consistency across 15 years is likely a reflection
of the relationship between a critical photoperiod and the timing of important
life history stages (Temte, 1994; Bronson, 2009). However, along with a critical
photoperiod, ribbon and spotted seals are dependent upon the presence of sea ice
for pupping and molt. We did not find large support in our models for a
relationship between the timing of peaks in haul-out behavior and the amount of
yearly maximum sea ice. This could indicate that, while the extent of spring sea
ice in the Bering Sea varied widely during our study period, seals were still
able to locate sea ice and haul out. Only a small portion of our data was from
2018 -- 2019, years of extreme low spring ice extent in the Bering Sea that
appeared to impact body condition of ribbon and spotted seals (Boveng et al., 2020), so we may
currently lack sufficient contrast in ice extent to characterize an effect on
the timing of peak haul-out probability. We should expect, however, that some minimal threshold in
the spatial extent or density of sea ice will have a meaningful impact on the
timing of peak haul-out behavior --- if there is no sea-ice, seals will not haul
out or they will be forced to use terrestrial haul-out sites which were not part of the
evolution of their normal behaviors. Additionally, while from an
ecological perspective the haul-out behavior appears consistent, the interannual
differences in timing and magnitude are large enough to have important
ramifications on calculations of abundance and trend. Those ramifications will
only be exacerbated if climate variability amplifies interannual differences.

Previous attempts to estimate the abundance of phocid seals from aerial survey
data in the Bering and Chukchi seas (e.g. Bengtson et al. (2005), Conn et al. (2014),
Ver Hoef et al. (2014)) have used estimated haul-out probabilities to correct for the
proportion of animals that are in the water and thus unavailable to be counted.
Although several of these studies allowed haul-out probabilities to vary by
day-of-year and time-of-day, they have not accounted for variability among
years, weather conditions, or in the age-sex class of the sample. In this paper,
we have shown that there can be considerable differences in the haul-out
probability of seals on ice based on these factors and subsequent analyses have
shown the potential for considerable bias in abundance estimates if such
covariates are unaccounted for (Conn \& Trukhanova, 2023). We recommend that future abundance
analyses employ availability models that account for them. For instance, it is
relatively straightforward to obtain weather reanalysis products for times and
locations that are surveyed and to construct a relevant correction factor based
on predictions of GLMPMs. The most challenging element in developing
availability correction factors is with annual variability. It can be difficult
to get a sufficient sample size to estimate year-specific correction factors,
particularly because research teams would likely need to deploy bio-loggers on
seals and conduct aerial surveys concurrently, requiring considerably more
personnel and money. One possible suggestion is to include year as a random
effect within models for aerial survey counts such that, without specific
knowledge of any particular year, the among-year variance is included in the
modeled standard errors. Regardless of the specific approach, future estimates
of Arctic seal abundance will require specific consideration of annual
variability and changes in the timing of peak haul-out behavior when estimating
trends, as one will not know if moderate differences in abundance estimates are
attributable to changes in abundance or changes in haul-out behavior.

Predictions of absolute haul-out probability in this paper were somewhat
different than those previously reported for these species, especially for
bearded seals. For instance, Ver Hoef et al. (2014) and Conn et al.
(2014) used haul-out correction factors with maximums of 0.66 for bearded
seals, 0.62 for ribbon seals, and 0.54 for spotted seals, where these maximums
corresponded to times near local solar noon in mid-late May. Results here
suggest a haul-out probability on May 20 at local solar noon of 0.304
(95\% CI: 0.258 -- 0.354)
for bearded seals across all age and sex classes,
0.715
(95\% CI: 0.62 -- 0.794) for
adult male ribbon seals, 0.661
(95\% CI: 0.576 -- 0.738) for
adult female ribbon seals, 0.742
(95\% CI: 0.655 -- 0.813) for
adult male spotted seals, and 0.663
(95\% CI: 0.574 -- 0.742) for
adult female spotted seals. Our estimates
of haul-out probability reflect increased sample sizes in terms of
number of individuals, inclusion of weather covariates, and improvements to the
way data were prepared prior to analysis and should be the basis for any
future estimates of seal abundance from aerial surveys in the Bering and
Chukchi seas.

We focused this paper on haul-out behavior of bearded, ribbon, and spotted
seals. Ringed seals are also present in the Bering and Chukchi seas but exhibit
a unique complicating factor. Adult ringed seals build subnivean lairs under the
snow on top of the sea ice, where they haul out and where females rear pups
until conditions are good for basking (Frost et al., 2004). Thus, the wet-dry sensor on
a bio-logger could indicate that an animal is hauled out, but if it is within a
lair, it is not available to be detected during an aerial survey. We hope to
address availability of ringed seals using data from satellite tags, replicate
survey tracks, and auxiliary information about snow depth and timing of melt in
a future study (see, e.g., Lindsay et al. (2021)).

Our analysis emphasizes the importance of accounting for seasonal and temporal
variation in haul-out behavior, as well as associated weather covariates,
when interpreting the number of seals counted in aerial surveys. The rapid
acceleration of climate change in Arctic ecosystems is already occurring and is
forecast to decrease the quantity and quality of suitable habitat for
ice-associated seals that rely on sea-ice for a variety of activities. Improved
estimates of seal population abundance from aerial surveys are needed to
properly monitor the impacts of these changes on seal populations over time.
Those monitoring surveys will need to be paired with continued investigation and
assessment of seal haul-out behavior to allow credible predictions about the
effects of climate disruptions on the abundance and distribution of Arctic seal
populations.

\hypertarget{author-contributions}{%
\section*{Author Contributions}\label{author-contributions}}
\addcontentsline{toc}{section}{Author Contributions}

\begin{itemize}
\tightlist
\item
  \textbf{Josh M. London}: investigation, conceptualization, methodology,
  formal analysis, validation, software, writing: original draft, writing:
  review and editing, visualization, and data curation
\item
  \textbf{Paul B. Conn}: conceptualization, methodology, formal analysis,
  software, validation, writing: original draft, writing: review and
  editing
\item
  \textbf{Stacie K. Hardy}: investigation, data curation, methodology,
  validation, writing: review and editing
\item
  \textbf{Erin L. Richmond}: data curation, investigation, methodology,
  validation, writing: review and editing
\item
  \textbf{Jay M. Ver Hoef}: conceptualization, methodology, software, writing:
  review and editing
\item
  \textbf{Michael F. Cameron}: investigation, project administration, writing:
  review and editing
\item
  \textbf{Justin A. Crawford}: investigation, methodology, validation, data curation,
  writing: review and editing
\item
  \textbf{Lori T. Quakenbush}: investigation, methodology, supervision, project
  administration, writing: review and editing
\item
  \textbf{Andrew L. Von Duyke}: investigation, methodology, validation, data curation,
  writing: review and editing
\item
  \textbf{Peter L. Boveng}: investigation, conceptualization, supervision,
  project administration, writing: review and editing
\end{itemize}

\hypertarget{data-availability}{%
\section*{Data Availability}\label{data-availability}}
\addcontentsline{toc}{section}{Data Availability}

This manuscript was developed as a reproducible research compendium. All data and
code are available on GitHub (\url{https://github.com/noaa-afsc/berchukseals-haulout}) and
major versions archived at Zenodo (\url{https://doi.org/10.5281/zenodo.4638221}).
Original data sources for telemetry are archived
at the United States Animal Telemetry Network, archived at Movebank, or
associated with other published manuscripts (see supplemental
material \ref{tab:deploymentTable}). Collated and cleaned data products
needed to replicate the analysis along with the results of all model fits are
also available and versioned as an R package on
GitHub (\url{https://github.com/noaa-afsc/berchuksealsHauloutFits}) and archived at
Zenodo.

\hypertarget{acknowledgments}{%
\section*{Acknowledgments}\label{acknowledgments}}
\addcontentsline{toc}{section}{Acknowledgments}

We recognize that the species and ecosystems we studied are within the ancestral
and present-day environs of the Inpuiat and Yup'ik people who, through many
uncredited contributions of traditional knowledge, provided early western
naturalists and scientists with much of what gets described as the `basic
biology' of Arctic seals. The deployment of bio-logging devices used in this
study were often done in collaboration with Alaska Native seal hunters and the
approval of
their communities. We would like to especially acknowledge the communities of
Kotzebue, Koyuk, Nome, Nuiqsut, Scammon Bay, St.~Michael, Utqiaġvik, and Ulguniq
(Wainwright) and the following individuals: James Adams, Jeff Barger, David
Barr, Wendell Booth, Cyrus Harris, Nereus `Doc' Harris, Grover Harris, Lee
Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin,
Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr.,
Virgil Naylor Sr., Dan Savetilik, Chuck Schaeffer, Ross Schaeffer, Allen Stone,
and Randy Toshavik from Kotzebue; Merlin Henry from Koyuk; Tom
Gray from Nome; Vernon Long and Richard Tukle from Nuiqsuit; Morgan
Simon, River Simon, and Al Smith from Scammon Bay; Alex Niksik Jr.~from
St.~Michael; Billy Adams, James Aiken, Tim Aiken, Howard Kittick,
Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, and Joe Skin from Utqiaġvik,
Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, and
Fred Rexford from Ulguniq.

We are grateful for the assistance in catching and sampling seals by Ryan Adam,
James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov,
Cynthia Christman, Sarah Coburn, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey Goldstein, Jeff
Harris, Jason Herreman, Markus Horning, John Jansen, Shawn Johnson, Charles
Littnan, Lloyd Lowry, Brett McClintock, Erin Moreland, Mark Nelson, Justin
Olnes, Lorrie Rea, Bob Shears, Gay Sheffield, Brent Stewart, Dave Withrow, and
Heather Ziel. We also appreciate the commitment to science and safety by all
officers and crew of the NOAA ship \emph{Oscar Dyson}, the NOAA ship \emph{MacArthur II}, and
the RV \emph{Thomas G. Thompson}.

Telemetry data from the Alaska Department of Fish and Game (ADF\&G) and the North
Slope Borough Department of Wildlife Management (NSB) were important
contributions to the findings presented here. Deployments in the western Bering
Sea were done in collaboration with Russian colleagues and North Pacific
Wildlife.

The findings and conclusions in the paper are those of the author(s) and do not
necessarily represent the views of the National Marine Fisheries Service, NOAA.
Any use of trade, product, or firm names does not imply an endorsement by the
U.S. Government. Funding for this study was provided by the U.S. National
Oceanic and Atmospheric Administration. The field work was conducted under the
authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
Fisheries Science Center Institutional Animal Care and Use Committee (IACUC). ).
Funding to ADF\&G for tagging seals was provided by the Bureau of Ocean Energy
Management (No.~M13PC0015 for work in Kotzebue, Alaska in 2009) and the Office
of Naval Research (No.~N00014-16-1-3019). ADF\&G and NSB field work was covered by
Research Permits Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF\&G IACUC
permits Nos. 06-16, 09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27,
0027-2018-29, 0027-2019-041.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-bengtson2004}{}}%
Bengtson JL, Cameron MF. 2004. Seasonal haulout patterns of crabeater seals (\emph{Lobodon carcinophaga}). \emph{Polar Biology} 27:344--349. DOI: \href{https://doi.org/DOI\%2010.1007/s00300-004-0597-1}{DOI 10.1007/s00300-004-0597-1}.

\leavevmode\vadjust pre{\hypertarget{ref-bengtson2005a}{}}%
Bengtson JL, Hiruki-Raring LM, Simpkins MA, Boveng PL. 2005. Ringed and bearded seal densities in the eastern Chukchi Sea, 1999-2000. \emph{Polar Biology} 28:833--845.

\leavevmode\vadjust pre{\hypertarget{ref-betts2006}{}}%
Betts MG, Diamond A, Forbes G, Villard M-A, Gunn J. 2006. The importance of spatial autocorrelation, extent and resolution in predicting forest bird occurrence. \emph{Ecological Modelling} 191:197--224.

\leavevmode\vadjust pre{\hypertarget{ref-boveng2009}{}}%
Boveng PL, Bengtson JL, Buckley TW, Cameron MF, Dahle SP, Kelly BP, Megrey BA, Overland JE, Williamson NJ. 2009. \href{https://repository.library.noaa.gov/view/noaa/3671}{Status review of the spotted seal (\emph{Phoca largha})}. \emph{U.S. Dep. Commer., NOAA Tech. Memo.} NMFS-AFSC-200:153.

\leavevmode\vadjust pre{\hypertarget{ref-boveng2013a}{}}%
Boveng PL, Bengtson JL, Cameron MF, Dahle SP, Logerwell EA(ElizabethA), London JM, Overland JE, Sterling JT, Stevenson DE 1970-, Taylor BL 1954-, Ziel HL 1974-. 2013. \href{https://repository.library.noaa.gov/view/noaa/4451}{Status review of the ribbon seal (\emph{Histriophoca fasciata})}. \emph{U.S. Dep. Commer., NOAA Tech. Memo.} NMFS-AFSC-255:174.

\leavevmode\vadjust pre{\hypertarget{ref-boveng2013}{}}%
Boveng PL, Cameron MF. 2013. \emph{Pinniped movements and foraging: seasonal movements, habitat selection, foraging and haul-out behavior of adult bearded seals in the Chukchi Sea. Final Report, BOEM Report 2013-01150}. Anchorage, Alaska: Bureau of Ocean Energy Management, Alaska Outer Continental Shelf Region.

\leavevmode\vadjust pre{\hypertarget{ref-boveng2018a}{}}%
Boveng P, Lowry L. 2018. Ribbon seal (\emph{Histriophoca fasciata}). In: Bernd Würsig, J. G. M. Thewissen, Kit Kovacs eds. \emph{Encyclopedia of Marine Mammals: Third Edition}. London: Academic Press, 811--813.

\leavevmode\vadjust pre{\hypertarget{ref-boveng2020}{}}%
Boveng PL, Ziel HL, McClintock BT, Cameron MF. 2020. Body condition of phocid seals during a period of rapid environmental change in the Bering Sea and Aleutian Islands, Alaska. \emph{Deep Sea Research Part II: Topical Studies in Oceanography} 181-182:104904. DOI: \href{https://doi.org/10.1016/j.dsr2.2020.104904}{10.1016/j.dsr2.2020.104904}.

\leavevmode\vadjust pre{\hypertarget{ref-breed2018}{}}%
Breed GA, Cameron MF, Ver Hoef JM, Boveng PL, Whiting A, Frost KJ. 2018. Seasonal sea ice dynamics drive movement and migration of juvenile bearded seals \emph{Erignathus barbatus}. \emph{Marine Ecology Progress Series} 600:223--237.

\leavevmode\vadjust pre{\hypertarget{ref-bronson2009}{}}%
Bronson FH. 2009. Climate change and seasonal reproduction in mammals. \emph{Philosophical Transactions of the Royal Society B} 364:3331--3340. DOI: \href{https://doi.org/10.1098/rstb.2009.0140}{10.1098/rstb.2009.0140}.

\leavevmode\vadjust pre{\hypertarget{ref-cameron2010}{}}%
Cameron MF, Bengtson JL, Boveng PL, Jansen JK(JohnK, Kelly BP, Dahle SP, Logerwell EA(ElizabethA), Overland JE, Sabine CL, Waring GT, Wilder JM. 2010. \href{https://repository.library.noaa.gov/view/noaa/3761}{Status review of the bearded seal (\emph{Erignathus barbatus})}. \emph{U.S. Dep. Commer., NOAA Tech. Memo.} NMFS-AFSC-211:246.

\leavevmode\vadjust pre{\hypertarget{ref-cameron2018}{}}%
Cameron MF, Frost KJ, Ver Hoef JM, Breed GA, Whiting AV, Goodwin J, Boveng PL. 2018. Habitat selection and seasonal movements of young bearded seals (\emph{Erignathus barbatus}) in the Bering Sea. \emph{PLoS ONE} 13:e0192743. DOI: \href{https://doi.org/10.1371/journal.pone.0192743}{10.1371/journal.pone.0192743}.

\leavevmode\vadjust pre{\hypertarget{ref-cavalieri1996}{}}%
Cavalieri DJ, Parkinson CL, Gloersen P, Zwally H. 1996.Sea ice concentrations from Nimbus-7 SMMR and DMSP SSM/I-SSMIS passive microwave data. Boulder, Colorado USA: NASA National Snow and Ice Data Center Distributed Active Archive Center

\leavevmode\vadjust pre{\hypertarget{ref-conn2014a}{}}%
Conn PB, Hoef JMV, McClintock BT, Moreland EE, London JM, Cameron MF, Dahle SP, Boveng PL. 2014. \href{https://onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.12127}{Estimating multispecies abundance using automated detection systems: Ice-associated seals in the Bering Sea}. \emph{Methods in Ecology and Evolution} 5:1280--1293.

\leavevmode\vadjust pre{\hypertarget{ref-conn2012c}{}}%
Conn PB, Johnson DS, London JM, Boveng PL. 2012. Accounting for missing data when assessing availability in animal population surveys: an application to ice-associated seals in the Bering Sea. \emph{Methods in Ecology and Evolution}:1039--1046.

\leavevmode\vadjust pre{\hypertarget{ref-conn2023}{}}%
Conn PB, Trukhanova IS. 2023. Modeling vital rates and age-sex structure of Pacific Arctic phocids: influence on aerial survey correction factors. \emph{Marine Mammal Science} 39:648--661. DOI: \href{https://doi.org/10.1111/mms.12992}{10.1111/mms.12992}.

\leavevmode\vadjust pre{\hypertarget{ref-crawford2019}{}}%
Crawford JA, Frost KJ, Quakenbush LT, Whiting A. 2019. Seasonal and diel differences in dive and haul-out behavior of adult and subadult ringed seals (\emph{Pusa hispida}) in the Bering and Chukchi seas. \emph{Polar Biology} 42:65--80. DOI: \href{https://doi.org/10.1007/s00300-018-2399-x}{10.1007/s00300-018-2399-x}.

\leavevmode\vadjust pre{\hypertarget{ref-crawford2015}{}}%
Crawford JA, Quakenbush LT, Citta JJ. 2015. A comparison of ringed and bearded seal diet, condition and productivity between historical (1975-1984) and recent (2003-2012) periods in the Alaskan Bering and Chukchi seas. \emph{Progress in Oceanography} 136:133--150. DOI: \href{https://doi.org/10.1016/j.pocean.2015.05.011}{10.1016/j.pocean.2015.05.011}.

\leavevmode\vadjust pre{\hypertarget{ref-feltz1966}{}}%
Feltz ET, Fay FH. 1966. Thermal requirements in vitro of epidermal cells from seals. \emph{Cryobiology} 3:261--264.

\leavevmode\vadjust pre{\hypertarget{ref-frost2018}{}}%
Frost KJ, Burns JJ. 2018. Spotted Seal: \emph{Phoca largha}. In: Würsig B, Thewissen JGM, Kovacs KM eds. \emph{Encyclopedia of Marine Mammals (Third Edition)}. Academic Press, 928--931. DOI: \href{https://doi.org/10.1016/B978-0-12-804327-1.00244-2}{10.1016/B978-0-12-804327-1.00244-2}.

\leavevmode\vadjust pre{\hypertarget{ref-frost2004}{}}%
Frost KJ, Lowry LF, Pendleton G, Nute HR. 2004. Factors affecting the observed densities of ringed seals, \emph{Phoca hispida}, in the Alaskan Beaufort Sea, 1996-99. \emph{Arctic} 57:115--128.

\leavevmode\vadjust pre{\hypertarget{ref-hamilton2018}{}}%
Hamilton V, Evans K. 2018. Establishing growth chronologies from marine mammal teeth: a method applicable across species. \emph{Journal of Experimental Marine Biology and Ecology} 505:24--34. DOI: \url{https://doi.org/10.1016/j.jembe.2018.04.006}.

\leavevmode\vadjust pre{\hypertarget{ref-harwood2020}{}}%
Harwood LA, Smith TG, Alikamik J, Alikamik E, Lea EV, Stirling I, Wright H, Melling H, Zhu X. 2020. Long-term, harvest-based monitoring of ringed seal body condition and reproduction in Canada's western arctic: an update through 2019. \emph{Arctic} 73:206--220. DOI: \href{https://doi.org/10.14430/arctic70428}{10.14430/arctic70428}.

\leavevmode\vadjust pre{\hypertarget{ref-huntington2020}{}}%
Huntington HP, Danielson SL, Wiese FK, Baker M, Boveng P, Citta JJ, De Robertis A, Dickson DMS, Farley E, George JC, Iken K, Kimmel DG, Kuletz K, Ladd C, Levine R, Quakenbush L, Stabeno P, Stafford KM, Stockwell D, Wilson C. 2020. Evidence suggests potential transformation of the Pacific Arctic ecosystem is underway. \emph{Nature Climate Change}:342--348. DOI: \href{https://doi.org/10.1038/s41558-020-0695-2}{10.1038/s41558-020-0695-2}.

\leavevmode\vadjust pre{\hypertarget{ref-johnson2008a}{}}%
Johnson DS, London JM, Lea M-A, Durban JW. 2008. Continuous-time correlated random walk model for animal telemetry data. \emph{Ecology} 89:1208--1215.

\leavevmode\vadjust pre{\hypertarget{ref-kelly2010}{}}%
Kelly BP, Bengtson JL, Boveng PL, Cameron MF, Dahle SP, Jansen JK, Logerwell EA, Overland JE, Sabine CL, Waring GT, Wilder JM. 2010. \emph{\href{https://repository.library.noaa.gov/view/noaa/3762}{Status review of the ringed seal (\emph{Phoca hispida})}}. U.S. Department of Commerce.

\leavevmode\vadjust pre{\hypertarget{ref-kelly1990c}{}}%
Kelly BP, Quakenbush LT. 1990. Spatiotemporal use of lairs by ringed seals (\emph{Phoca hispida}). \emph{Canadian Journal of Zoology} 68:2503--2512.

\leavevmode\vadjust pre{\hypertarget{ref-kirkman2003}{}}%
Kirkman SP, Bester MN, Pistorius PA, Hofmeyr GJG, Jonker FC, Owen R, Strydom N. 2003. Variation in the timing of moult in southern elephant seals at Marion Island : research article. \emph{South African Journal of Wildlife Research - 24-month delayed open access} 33:79--84. DOI: \href{https://doi.org/10.10520/EJC117173}{10.10520/EJC117173}.

\leavevmode\vadjust pre{\hypertarget{ref-kwok2018}{}}%
Kwok R. 2018. Arctic sea ice thickness, volume, and multiyear ice coverage: losses and coupled variability (1958--2018). \emph{Environmental Research Letters} 13:105005. DOI: \href{https://doi.org/10.1088/1748-9326/aae3ec}{10.1088/1748-9326/aae3ec}.

\leavevmode\vadjust pre{\hypertarget{ref-lindsay2021}{}}%
Lindsay JM, Laidre KL, Conn PB, Moreland EE, Boveng PL. 2021. \href{https://www.int-res.com/abstracts/esr/v46/p1-17/}{Modeling ringed seal \emph{Pusa hispida} habitat and lair emergence timing in the eastern Bering and Chukchi Seas}. \emph{Endangered Species Research} 46:1--17.

\leavevmode\vadjust pre{\hypertarget{ref-london2012}{}}%
London JM, Ver Hoef JM, Jeffries SJ, Lance MM, Boveng PL. 2012. Haul-out behavior of harbor seals (\emph{Phoca vitulina}) in Hood Canal, Washington. \emph{PLoS ONE} 7:e38180. DOI: \href{https://doi.org/10.1371/journal.pone.0038180}{10.1371/journal.pone.0038180}.

\leavevmode\vadjust pre{\hypertarget{ref-lonergan2013}{}}%
Lonergan M, Duck C, Moss S, Morris C, Thompson D. 2013. Rescaling of aerial survey data with information from small numbers of telemetry tags to estimate the size of a declining harbour seal population. \emph{Aquatic Conservation-Marine and Freshwater Ecosystems} 23:135--144. DOI: \href{https://doi.org/Doi\%2010.1002/Aqc.2277}{Doi 10.1002/Aqc.2277}.

\leavevmode\vadjust pre{\hypertarget{ref-lopez2013}{}}%
Lopez R, Malardé JP, Royer F, Gaspar P. 2013. Improving Argos Doppler location using multiple-model Kalman filtering. \emph{Geoscience and Remote Sensing, IEEE Transactions on} PP:1--12. DOI: \href{https://doi.org/10.1109/tgrs.2013.2284293}{10.1109/tgrs.2013.2284293}.

\leavevmode\vadjust pre{\hypertarget{ref-mclaren1958a}{}}%
McLaren IA. 1958. Some aspects of growth and reproduction of the bearded seal, \emph{Erignathus barbartus} (Erxleben). \emph{Journal of the Fisheries Research Board of Canada} 15:219--227.

\leavevmode\vadjust pre{\hypertarget{ref-meier2014}{}}%
Meier WN, Hovelsrud GK, Oort BEH van, Key JR, Kovacs KM, Michel C, Haas C, Granskog MA, Gerland S, Perovich DK, Makshtas A, Reist JD. 2014. \href{https://agupubs.onlinelibrary.wiley.com/doi/abs/10.1002/2013RG000431}{Arctic sea ice in transformation: A review of recent observed changes and impacts on biology and human activity}. \emph{Reviews of Geophysics} 52:185--217.

\leavevmode\vadjust pre{\hypertarget{ref-mesinger2006a}{}}%
Mesinger F, DiMego G, Kalnay E, Mitchell K, Shafran PC, Ebisuzaki W, Jović D, Woollen J, Rogers E, Berbery EH, Ek MB, Fan Y, Grumbine R, Higgins W, Li H, Lin Y, Manikin G, Parrish D, Shi W. 2006. North American Regional Reanalysis. \emph{Bulletin of the American Meteorological Society} 87:343--360. DOI: \href{https://doi.org/10.1175/BAMS-87-3-343}{10.1175/BAMS-87-3-343}.

\leavevmode\vadjust pre{\hypertarget{ref-niemi2023}{}}%
Niemi M, Nykänen M, Biard V, Kunnasranta M. 2023. Seasonal changes in diel haul-out patterns of a lacustrine ringed seal (\emph{Pusa hispida saimensis}). \emph{Ecology and Evolution} 13:e10264. DOI: \href{https://doi.org/10.1002/ece3.10264}{10.1002/ece3.10264}.

\leavevmode\vadjust pre{\hypertarget{ref-olnes2020}{}}%
Olnes J, Crawford J, Citta JJ, Druckenmiller ML, Von Duyke AL, Quakenbush L. 2020. Movement, diving, and haul-out behaviors of juvenile bearded seals in the Bering, Chukchi and Beaufort seas, 2014--2018. \emph{Polar Biology}. DOI: \href{https://doi.org/10.1007/s00300-020-02710-6}{10.1007/s00300-020-02710-6}.

\leavevmode\vadjust pre{\hypertarget{ref-overland2021}{}}%
Overland JE. 2021. Rare events in the Arctic. \emph{Climatic Change} 168:1--13.

\leavevmode\vadjust pre{\hypertarget{ref-perpinan2012}{}}%
Perpiñán O. 2012. \href{http://www.jstatsoft.org/v50/i09/}{solaR: Solar Radiation and Photovoltaic Systems with R}. \emph{Journal of Statistical Software} 50:1--32.

\leavevmode\vadjust pre{\hypertarget{ref-perry2017}{}}%
Perry EA, Stenson GB, Buren AD. 2017. Attendance and nursing patterns of harp seals in the harsh environment of the northwest Atlantic. \emph{Polar Biology} 40:151--160. DOI: \href{https://doi.org/10.1007/s00300-016-1938-6}{10.1007/s00300-016-1938-6}.

\leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2021}{}}%
R Core Team. 2021. \emph{\href{https://www.R-project.org/}{R: A language and environment for statistical computing}}. Vienna, Austria.

\leavevmode\vadjust pre{\hypertarget{ref-reder2003}{}}%
Reder S, Lydersen C, Arnold W, Kovacs KM. 2003. Haulout behaviour of High Arctic harbour seals (\emph{Phoca vitulina vitulina}) in Svalbard, Norway. \emph{Polar Biology} 27:6--16.

\leavevmode\vadjust pre{\hypertarget{ref-southwell2008}{}}%
Southwell CJ, Paxton CGM, Borchers DL, Boveng PL, Nordøy ES, Blix AS, De La Mare WK. 2008. Estimating population status under conditions of uncertainty: the Ross seal in East Antarctica. \emph{Antarctic Science} 20:123--133.

\leavevmode\vadjust pre{\hypertarget{ref-temte1994}{}}%
Temte JL. 1994. Photoperiod control of birth timing in the harbour seal (\emph{Phoca vitulina}). \emph{J. Zool.} 233:369--384.

\leavevmode\vadjust pre{\hypertarget{ref-teneyck2018}{}}%
Ten Eyck P, Cavanaugh JE. 2018. An Alternate Approach to Pseudo-Likelihood Model Selection in the Generalized Linear Mixed Modeling Framework. \emph{Sankhya B} 80:98--122. DOI: \href{https://doi.org/10.1007/s13571-017-0130-5}{10.1007/s13571-017-0130-5}.

\leavevmode\vadjust pre{\hypertarget{ref-thometz2021}{}}%
Thometz NM, Hermann-Sorensen H, Russell B, Rosen DAS, Reichmuth C. 2021. Molting strategies of Arctic seals drive annual patterns in metabolism. \emph{Conservation Physiology} 9:coaa112. DOI: \href{https://doi.org/10.1093/conphys/coaa112}{10.1093/conphys/coaa112}.

\leavevmode\vadjust pre{\hypertarget{ref-thompson1987}{}}%
Thompson P, Rothery P. 1987. Age and sex differences in the timing of moult in the common seal, \emph{Phoca vitulina}. \emph{Journal of Zoology} 212:597--603.

\leavevmode\vadjust pre{\hypertarget{ref-udevitz2009}{}}%
Udevitz MS, Jay CV, Fischback AS, Garlich-Miller JL. 2009. Modeling haul-out behavior of walruses in Bering Sea ice. \emph{Can. J. Zool} 87:1111--1128. DOI: \href{https://doi.org/10.1139/Z09-098}{10.1139/Z09-098}.

\leavevmode\vadjust pre{\hypertarget{ref-verhoef2014a}{}}%
Ver Hoef JM, Cameron MF, Boveng PL, London JM, Moreland EE. 2014. A spatial hierarchical model for abundance of three ice-associated seal species in the eastern Bering Sea. \emph{Statistical Methodology} 17:46--66. DOI: \url{http://dx.doi.org/10.1016/j.stamet.2013.03.001}.

\leavevmode\vadjust pre{\hypertarget{ref-verhoef2010a}{}}%
Ver Hoef JM, London JM, Boveng PL. 2010. Fast computing of some generalized linear mixed pseudo-models with temporal autocorrelation. \emph{Computational Statistics} 25:39--55. DOI: \href{https://doi.org/10.1007/s00180-009-0160-1}{10.1007/s00180-009-0160-1}.

\leavevmode\vadjust pre{\hypertarget{ref-vonduyke2020}{}}%
Von Duyke AL, Douglas DC, Herreman JK, Crawford JA. 2020. Ringed seal (\emph{Pusa hispida}) seasonal movements, diving, and haul-out behavior in the Beaufort, Chukchi, and Bering Seas (2011--2017). \emph{Ecology and Evolution} 10:5595--5616. DOI: \href{https://doi.org/10.1002/ece3.6302}{10.1002/ece3.6302}.

\leavevmode\vadjust pre{\hypertarget{ref-wang2018}{}}%
Wang M, Yang Q, Overland JE, Stabeno P. 2018. Sea-ice cover timing in the Pacific Arctic: The present and projections to mid-century by selected CMIP5 models. \emph{Deep Sea Research Part II: Topical Studies in Oceanography} 152:22--34. DOI: \url{https://doi.org/10.1016/j.dsr2.2017.11.017}.

\end{CSLReferences}

\newpage
\setcounter{table}{0}\renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0}\renewcommand{\thefigure}{S\arabic{figure}}
\begin{landscape}

\hypertarget{supplemental-material}{%
\section*{Supplemental Material}\label{supplemental-material}}
\addcontentsline{toc}{section}{Supplemental Material}

\hypertarget{additional-bio-logger-deployment-details}{%
\subsection{Additional Bio-logger Deployment Details}\label{additional-bio-logger-deployment-details}}

\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}

\global\setlength{\Oldtabcolsep}{\tabcolsep}

\setlength{\tabcolsep}{0pt}

\renewcommand*{\arraystretch}{1.5}

\providecommand{\ascline}[3]{\noalign{\global\arrayrulewidth #1}\arrayrulecolor[HTML]{#2}\cline{#3}}

\begin{longtable}[c]{|p{1.00in}|p{0.80in}|p{1.50in}|p{1.50in}|p{0.80in}|p{0.80in}|p{0.80in}|p{1.50in}}

\caption{The\ \ timing,\ location,\ and\ institutions\ responsible\ for\ the\ bio-logger\ deployments\ used\ in\ this\ study\ along\ with\ research\ permits\ and\ any\ associated\ publications.}\label{tab:deploymentTable}\\

\ascline{1.5pt}{666666}{1-8}

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Institution}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Year\ Deployed}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Location}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Publication(s)}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Permits}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{No.\ Seals}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Sex}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Age\ Class}}}} \\

\ascline{1.5pt}{666666}{1-8}\endfirsthead

\ascline{1.5pt}{666666}{1-8}

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Institution}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Year\ Deployed}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Location}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Publication(s)}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Permits}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{No.\ Seals}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Sex}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Age\ Class}}}} \\

\ascline{1.5pt}{666666}{1-8}\endhead

\multicolumn{8}{>{\raggedright}m{\dimexpr 8.7in+14\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{\textbf{ADFG}}}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{=Alaska\ Department\ of\ Fish\ and\ Game;\ }}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{\textbf{NSB}}}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{=North\ Slope\ Borough;\ }}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{\textbf{NMFS}}}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{=NOAA\ National\ Marine\ Fisheries\ Service;\ }}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{\textbf{NPWC}}}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{=North\ Pacific\ Wildlife\ Consortium}}} \\

\endfoot

\multicolumn{8}{>{\raggedright}m{\dimexpr 8.7in+14\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{\textbf{Bearded\ seal}}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2005}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Cameron\ et\ al\ 2018}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1585}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{6}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2006}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Cameron\ et\ al\ 2018}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1585}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2009}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Breed\ et\ al\ 2018}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1787}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{4}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2014}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ Koyuk\ River}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2014}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ Nome}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2015}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ St.\ Michael}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Elson\ Lagoon,\ Utqiagvik}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ Koyuk\ River}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ Nome}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ St.\ Michael}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Colville\ River,\ Nuiqsut}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ Koyuk\ River}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ Nome}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2019}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Dease\ Inlet,\ Utqiagvik}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2021}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{20466}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2005}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1585}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2009}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{McClintock\ et\ al\ 2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1765}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2011}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{McClintock\ et\ al\ 2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15126}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{3}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2012}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{McClintock\ et\ al\ 2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15126}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2012}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Elson\ Lagoon,\ Utqiagvik}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2019}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Pittalugruaq\ Lake}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{20466}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{8}{>{\raggedright}m{\dimexpr 8.7in+14\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{\textbf{Ribbon\ seal}}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2005}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Ozemoy\ Gulf,\ Russia}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1765}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{9}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{young\ of\ year,\ adult,\ subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2006}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1765}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{7}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult,\ young\ of\ year}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2007}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1765}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{28}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult,\ young\ of\ year}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2008}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1765}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2009}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1765}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{28}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult,\ subadult,\ young\ of\ year}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2010}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1787,\ 15126}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{17}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{young\ of\ year,\ adult,\ subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2014}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15126}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{13}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult,\ young\ of\ year}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{19309}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{7}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult}}} \\

\multicolumn{8}{>{\raggedright}m{\dimexpr 8.7in+14\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{\textbf{Spotted\ seal}}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2005}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1585}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{3}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Dease\ Inlet,\ Utqiagvik}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{4}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Colville\ River,\ Nuiqsut}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Scammon\ Bay}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{3}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2018}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Dease\ Inlet,\ Utqiagvik}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{20466}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2018}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Scammon\ Bay}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{20466}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2019}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Dease\ Inlet,\ Utqiagvik}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{20466}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{6}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult,\ subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2006}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1676}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{5}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{young\ of\ year,\ subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2007}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1676}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{12}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult,\ young\ of\ year,\ subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2009}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1787}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{23}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult,\ subadult,\ young\ of\ year}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2010}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1787,\ 15126}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{8}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{young\ of\ year,\ adult,\ subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2014}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15126}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{5}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{young\ of\ year,\ adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{19309}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{6}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2018}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{19309}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{5}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NPWC}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2009}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kamchatka\ Peninsula}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NA}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{3}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2012}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Tiny\ Island}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2014}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Oarlock\ Island}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{6}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2014}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Seal\ Island}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2015}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Oarlock\ Island}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{6}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Pittalugruaq\ Lake}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{3}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Pittalugruaq\ Lake}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\

\ascline{1.5pt}{666666}{1-8}

\end{longtable}

\arrayrulecolor[HTML]{000000}

\global\setlength{\arrayrulewidth}{\Oldarrayrulewidth}

\global\setlength{\tabcolsep}{\Oldtabcolsep}

\renewcommand*{\arraystretch}{1}

\end{landscape}

\hypertarget{supplemental-figures-showing-confidence-intervals-associated-with-predictions}{%
\subsection{Supplemental Figures Showing Confidence Intervals Associated with Predictions}\label{supplemental-figures-showing-confidence-intervals-associated-with-predictions}}

The following series of figures (\ref{fig:beardedPredSE}, \ref{fig:ribbonPredSE},
and \ref{fig:spottedPredSE}) show the seasonal variability in predicted haul-out
probability and the associated 95\% confidence intervals for bearded, ribbon, and
spotted seals. The predictions shown are based on the same data used in
\ref{fig:beardedHOCal}, \ref{fig:ribbonHOCal}, and \ref{fig:spottedHOCal} but
selected for three local solar hours (07:00, 12:00, and 17:00) so the
confidence intervals can also be shown and comparisons can be made.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/beardedPredSE-1} \caption{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for bearded seals. Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00) with smoothed weather covariates. Age and sex classes are combined into a single 'all ages' category.}\label{fig:beardedPredSE}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/ribbonPredSE-1} \caption{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals. Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00) with smoothed weather covariates. Age and sex classes are separated to allow comparisons.}\label{fig:ribbonPredSE}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/spottedPredSE-1} \caption{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for spotted seals. Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00) with smoothed weather covariates. Age and sex classes are separated to allow comparisons.}\label{fig:spottedPredSE}
\end{figure}

\hypertarget{exploring-insolation-solar-radiation-as-a-model-covariate}{%
\subsection{Exploring Insolation (Solar Radiation) as a Model Covariate}\label{exploring-insolation-solar-radiation-as-a-model-covariate}}

\hypertarget{introduction-1}{%
\subsubsection{Introduction}\label{introduction-1}}

During the peer review process for this manuscript, Anthony Fischbach suggested
the possibility of using predicted insolation (or solar radiation) values from
the reanalysis model as a more direct and, potentially, more informative
predictor of the daily haul-out cycle in seals compared to time of day. The
notion being that seals are, likely, directly responding to changes in solar radiation
throughout the day and not what time of day it is (i.e.~seals don't have human
watches). Additionally, given the energetic benefits of increased solar
radiation it could be more informative as we would expect seals might have a
higher haul-out probability on sunnier days and for there to be geographic
variability in haul-out behavior associated with geographical differences in
insolation. This approach has an additional benefit of being more parsimonious
compared to our use of the Fourier series or other approaches to represent
hour-of-day in the model (e.g.~24 factors for each hour).

Because of these reasons, we considered and explored this
possibility for our model and the analysis presented in this manuscript. A key
drawback to reliance on solar radiation, in our minds, is that we would lose
insight regarding potential diel patterns -- solar radiation does not
differentiate between dusk or dawn. Bi-modal patterns have been
previously observed in ringed seals and our results in this analysis show some indication of
increased haul-out probability during dawn compared to dusk periods for bearded seals
and some age and sex classes for ribbon and spotted seals.
For other phocid species, increased haul-out probability before solar noon or after
solar noon has been observed. Importantly, understanding these relationships
between haul-out probability and hour-of-day can have important ramifications
on aerial survey study design -- a key focus of this paper.\\
Another hesitation we had was that solar radiation estimates from reanalysis models have
not been previously used as a model covariate within a published study of
pinniped haul-out behavior. Thus, for this analysis, we chose to keep our
original approach and rely on the Fourier series to capture any hour-of-day
effects.

That said, we think the idea of solar radiation as a model covariate in pinniped
haul-out models is intriguing and worth further exploration. The current
availability and increased accessibility to detailed climate reanalysis
products that include solar radiation is exciting and we encourage future, more
detailed exploration of this as a component in pinniped haul-out analysis. To
provide some inspiration, we present some initial efforts and examples for
comparison.

\hypertarget{methods-1}{%
\subsubsection{Methods}\label{methods-1}}

In this manuscript, we rely on the NARR reanalysis model as the source for our
weather covariates. However, since our initiation of this analysis, the ERA5
reanalysis model (\url{https://doi.org/10.24381/cds.adbb2d47}) has become one of the
go-to standards for global climate reanalysis and provides an increased temporal
resolution to hourly (compared to the 3-hour resolution of NARR). The global
coverage of ERA5 provides additional flexibility in that the area of interest is
not limited to North America. The ERA5 model provides a number of solar
radiation parameters and it was important to evaluate and understand each of
these estimates in order to select the one that was likely most relevant to
seals. Here, we used the `surface short-wave (solar) radiation downwards'
parameter. This
parameter is described as \emph{``the amount of solar radiation (also known as shortwave radiation)
that reaches a horizontal plane at the surface of the Earth and comprises both
direct and diffuse solar radiation. To a reasonably good approximation, this
parameter is the model equivalent of what would be measured by a pyranometer (an
instrument used for measuring solar radiation) at the surface''}
(\url{https://codes.ecmwf.int/grib/param-db/?id=169}). Thus, this is the
value which most closely represents the amount of solar radiation likely
felt by a seal hauled out of the water.

ERA5 data is available via the Copernicus climate data store API which can be
queried with the CDS-API Python package
(\url{https://cds.climate.copernicus.eu/api-how-to}). The R code provided here
documents the download of the \emph{surface\_solar\_radiation\_downwards} parameter for
our study area of interest and years of interest. The \texttt{reticulate} R package
(\url{https://CRAN.R-project.org/package=reticulate}) allowed interaction with Python.
Additionally, note, extra steps are required to download data on either side
of the 180 anti-meridian.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(tidyverse)}
\FunctionTok{library}\NormalTok{(reticulate)}
\FunctionTok{library}\NormalTok{(sf)}
\FunctionTok{library}\NormalTok{(terra)}

\CommentTok{\#import python CDS{-}API}
\NormalTok{cdsapi }\OtherTok{\textless{}{-}} \FunctionTok{import}\NormalTok{(}\StringTok{\textquotesingle{}cdsapi\textquotesingle{}}\NormalTok{)}
\CommentTok{\#for this step there must exist the file .cdsapirc}
\NormalTok{server }\OtherTok{=}\NormalTok{ cdsapi}\SpecialCharTok{$}\FunctionTok{Client}\NormalTok{() }\CommentTok{\#start the connection}

\NormalTok{get\_era5 }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(y) \{}
  \CommentTok{\#we create the query}
\NormalTok{  query }\OtherTok{\textless{}{-}} \FunctionTok{r\_to\_py}\NormalTok{(}
    \FunctionTok{list}\NormalTok{(}
      \AttributeTok{variable =} \StringTok{"surface\_solar\_radiation\_downwards"}\NormalTok{,}
      \AttributeTok{product\_type =} \StringTok{"reanalysis"}\NormalTok{,}
      \AttributeTok{area =} \StringTok{"75/152/47/180"}\NormalTok{, }\CommentTok{\# North, West, South, East}
      \AttributeTok{year =}\NormalTok{ y,}
      \AttributeTok{month =} \FunctionTok{str\_pad}\NormalTok{(}\DecValTok{2}\SpecialCharTok{:}\DecValTok{7}\NormalTok{, }\DecValTok{2}\NormalTok{, }\StringTok{"left"}\NormalTok{, }\StringTok{"0"}\NormalTok{),}
      \AttributeTok{day =} \FunctionTok{str\_pad}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\DecValTok{31}\NormalTok{, }\DecValTok{2}\NormalTok{, }\StringTok{"left"}\NormalTok{, }\StringTok{"0"}\NormalTok{),}
      \AttributeTok{time =} \FunctionTok{str\_c}\NormalTok{(}\DecValTok{0}\SpecialCharTok{:}\DecValTok{23}\NormalTok{, }\StringTok{"00"}\NormalTok{, }\AttributeTok{sep =} \StringTok{":"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{str\_pad}\NormalTok{(}\DecValTok{5}\NormalTok{, }\StringTok{"left"}\NormalTok{, }\StringTok{"0"}\NormalTok{),}
      \AttributeTok{format =} \StringTok{"netcdf"}
\NormalTok{    )}
\NormalTok{  )}
  \CommentTok{\#query to get the ncdf}
\NormalTok{  server}\SpecialCharTok{$}\FunctionTok{retrieve}\NormalTok{(}\StringTok{"reanalysis{-}era5{-}single{-}levels"}\NormalTok{,}
\NormalTok{                  query,}
                  \FunctionTok{paste0}\NormalTok{(}\StringTok{"era5\_ssrd\_"}\NormalTok{, y, }\StringTok{"\_left.nc"}\NormalTok{))}

\NormalTok{  query }\OtherTok{\textless{}{-}} \FunctionTok{r\_to\_py}\NormalTok{(}
    \FunctionTok{list}\NormalTok{(}
      \AttributeTok{variable =} \StringTok{"surface\_solar\_radiation\_downwards"}\NormalTok{,}
      \AttributeTok{product\_type =} \StringTok{"reanalysis"}\NormalTok{,}
      \AttributeTok{area =} \StringTok{"75/{-}180/47/{-}142"}\NormalTok{, }\CommentTok{\# North, West, South, East}
      \AttributeTok{year =}\NormalTok{ y,}
      \AttributeTok{month =} \FunctionTok{str\_pad}\NormalTok{(}\DecValTok{2}\SpecialCharTok{:}\DecValTok{7}\NormalTok{, }\DecValTok{2}\NormalTok{, }\StringTok{"left"}\NormalTok{, }\StringTok{"0"}\NormalTok{),}
      \AttributeTok{day =} \FunctionTok{str\_pad}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\DecValTok{31}\NormalTok{, }\DecValTok{2}\NormalTok{, }\StringTok{"left"}\NormalTok{, }\StringTok{"0"}\NormalTok{),}
      \AttributeTok{time =} \FunctionTok{str\_c}\NormalTok{(}\DecValTok{0}\SpecialCharTok{:}\DecValTok{23}\NormalTok{, }\StringTok{"00"}\NormalTok{, }\AttributeTok{sep =} \StringTok{":"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{str\_pad}\NormalTok{(}\DecValTok{5}\NormalTok{, }\StringTok{"left"}\NormalTok{, }\StringTok{"0"}\NormalTok{),}
      \AttributeTok{format =} \StringTok{"netcdf"}
\NormalTok{    )}
\NormalTok{  )}
  \CommentTok{\#query to get the ncdf}
\NormalTok{  server}\SpecialCharTok{$}\FunctionTok{retrieve}\NormalTok{(}\StringTok{"reanalysis{-}era5{-}single{-}levels"}\NormalTok{,}
\NormalTok{                  query,}
                  \FunctionTok{paste0}\NormalTok{(}\StringTok{"era5\_ssrd\_"}\NormalTok{, y, }\StringTok{"\_right.nc"}\NormalTok{))}

\NormalTok{\}}

\NormalTok{years }\OtherTok{\textless{}{-}} \FunctionTok{as.character}\NormalTok{(}\DecValTok{2005}\SpecialCharTok{:}\DecValTok{2021}\NormalTok{)}
\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(years)) \{}
  \FunctionTok{get\_era5}\NormalTok{(years[i])}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

To explore performance of our solar radiation parameter within a haul-out model
we replaced the various Fourier series parameters in our model from the manuscript
with the ERA5 \emph{surface solar radiation downwards} (\texttt{era\_ssrd\_watts}) parameter. As
with other reanalysis values (from NARR) in the manuscript, the \texttt{era-ssrd-watts}
values are matched in time and space to the seal haul-out observation data; we
use the full hourly temporal resolution from ERA5. The glmmLDTS framework used
in the paper does not allow for model comparisons with AIC because of
the reliance on pseudo-likelihood. The \texttt{bam()} function within the \texttt{mgcv} package
provides a quick model fitting option that also allowed us to do some model
comparison with AIC. This approach was sufficient for the general demonstration and
exploration purposes here
but future research should consider a range of model fitting frameworks and
approaches that might be more appropriate.

The model specification below was used to specify an \texttt{mgcv::bam()} model that
matched the formula used in the manuscript for ribbon seals. The \texttt{s(speno,\ bs\ =\ "re")}
term is the smooth term for the random effect. All other predictors were
the same.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{m1\_ribbon }\OtherTok{\textless{}{-}}\NormalTok{ mgcv}\SpecialCharTok{::}\FunctionTok{bam}\NormalTok{(}
\NormalTok{  dry }\SpecialCharTok{\textasciitilde{}}\NormalTok{ age\_sex }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(speno, }\AttributeTok{bs =} \StringTok{"re"}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{    sin1 }\SpecialCharTok{+}\NormalTok{ cos1 }\SpecialCharTok{+}\NormalTok{ sin2 }\SpecialCharTok{+}\NormalTok{ cos2 }\SpecialCharTok{+}\NormalTok{ sin3 }\SpecialCharTok{+}\NormalTok{ cos3 }\SpecialCharTok{+} 
    \FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{    sin1}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    cos1}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    sin2}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    cos2}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    sin3}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    cos3}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    wind}\SpecialCharTok{*}\NormalTok{temp2m }\SpecialCharTok{+}\NormalTok{ pressure }\SpecialCharTok{+}\NormalTok{ precip }\SpecialCharTok{+} 
\NormalTok{    age\_sex}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{4}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{),}
  \AttributeTok{data =}\NormalTok{ ribbon\_model\_data,}
  \AttributeTok{family =}\NormalTok{ binomial,}
  \AttributeTok{discrete =} \ConstantTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Note, the specification for \emph{m1\_ribbon} here does not include
any AR1 structure for temporal autocorrelation. To include this, we needed to
provide a value for \(\rho\) (or \emph{rho}). We examined the autocorrelation within
the model and used the lag-1 value for \(\rho\) .The value for lag-1
autocorrelation was 0.8082 which is
rather high but not surprising. We then updated our model specification with
a value for \(\rho\) as well as the \texttt{A1.start} argument which specifies (as either
\textbf{TRUE} or \textbf{FALSE}) the start point of each block.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{m2\_ribbon }\OtherTok{\textless{}{-}}\NormalTok{ mgcv}\SpecialCharTok{::}\FunctionTok{bam}\NormalTok{(}
\NormalTok{  dry }\SpecialCharTok{\textasciitilde{}}\NormalTok{ age\_sex }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(speno, }\AttributeTok{bs =} \StringTok{"re"}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{    sin1 }\SpecialCharTok{+}\NormalTok{ cos1 }\SpecialCharTok{+}\NormalTok{ sin2 }\SpecialCharTok{+}\NormalTok{ cos2 }\SpecialCharTok{+}\NormalTok{ sin3 }\SpecialCharTok{+}\NormalTok{ cos3 }\SpecialCharTok{+} 
    \FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{    sin1}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    cos1}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    sin2}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    cos2}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    sin3}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    cos3}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    wind}\SpecialCharTok{*}\NormalTok{temp2m }\SpecialCharTok{+}\NormalTok{ pressure }\SpecialCharTok{+}\NormalTok{ precip }\SpecialCharTok{+} 
\NormalTok{    age\_sex}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{),}
  \AttributeTok{data =}\NormalTok{ ribbon\_model\_data,}
  \AttributeTok{family =}\NormalTok{ binomial,}
  \AttributeTok{AR.start =}\NormalTok{ ar1\_start,}
  \AttributeTok{rho =}\NormalTok{ lag1\_ribbon,}
  \AttributeTok{discrete =} \ConstantTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

The model specification for exploring the use of solar radiation was
specified similarly but without all of the Fourier series parameters and
interactions.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{m2\_ssrd\_ribbon }\OtherTok{\textless{}{-}}\NormalTok{ mgcv}\SpecialCharTok{::}\FunctionTok{bam}\NormalTok{(}
\NormalTok{  dry }\SpecialCharTok{\textasciitilde{}}\NormalTok{ age\_sex }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(speno, }\AttributeTok{bs =} \StringTok{"re"}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{    era5\_ssrd\_watts }\SpecialCharTok{+}
    \FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{    era5\_ssrd\_watts}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{    wind}\SpecialCharTok{*}\NormalTok{temp2m }\SpecialCharTok{+}\NormalTok{ pressure }\SpecialCharTok{+}\NormalTok{ precip }\SpecialCharTok{+} 
\NormalTok{    age\_sex}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{),}
  \AttributeTok{data =}\NormalTok{ ribbon\_model\_data,}
  \AttributeTok{family =}\NormalTok{ binomial,}
  \AttributeTok{AR.start =}\NormalTok{ ar1\_start,}
  \AttributeTok{rho =}\NormalTok{ lag1\_ribbon,}
  \AttributeTok{discrete =} \ConstantTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

The two models were compared with AIC to evaluate whether the
reduction in degrees of freedom with fewer terms in the solar radiation
model was matched with improved explanatory power in the model fit. While the
model and code specified above is for ribbon seals, the same approach was
repeated for bearded and spotted seals.

A similar approach to that presented in this manuscript for prediction was
employed with solar radiation values in lieu of hour of day. For prediction
values, quantiles (5\% increments) of the observed range of ERA5 solar radiation values were
used with 100\% representing the maximum observed solar radiation value. This
allowed similar data visualizations and easier comparisons to those
predictions in the manuscript that include hour of day.

\hypertarget{results-1}{%
\subsubsection{Results}\label{results-1}}

To evaluate whether the solar radiation parameter matched our expectations and
compared well with hour of the day, we visualized the variability of the
\texttt{era5\_ssrd} values within
our study area as they relate to hour of the day (\ref{fig:era5-ssrd-plot}).
The unimodal distribution
is centered around the middle of the solar day with peak solar radiation
coinciding with 13:00 local solar. This suggests solar radiation could be an
informative covariate for capturing unimodal diel patterns in haul-out
behavior.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/era5-ssrd-plot-1} \caption{Downward surface solar radiation estimates from the ERA5 climate reanalysis for 5000 random points within the study area between 2005 and 2021. Solar radiation values are presented in Watts per square-meter and the smoothed line highlights the strong diel pattern.}\label{fig:era5-ssrd-plot}
\end{figure}

The bearded seal model matching the specification from the manuscript resulted in
126.13 degrees of freedom and an AIC value of
-7428.929. The model with solar radiation resulted in
39.829 degrees of freedom and an AIC value of
-6874.298. The ribbon seal model matching the specification from the manuscript resulted in
131.478 degrees of freedom and an AIC value of
-16372.29. The model with solar radiation resulted in
114.72 degrees of freedom and an AIC value of
-15223.346. The spotted seal model matching the specification from the manuscript resulted in
125.54 degrees of freedom and an AIC value of
-23482.051. The model with solar radiation resulted in
108.985 degrees of freedom and an AIC value of
-22917.167. Despite the additional terms, the
models with the Fourier series representation of hour of day resulted in a
lower AIC value and were still preferred models for each of the species.

Predictions from the model fits and visualization of those predictions were
produced for each species but, here, we only present visualizations from
ribbon seals as an example (Figure \ref{fig:ssrd-prediction-heatmap} and Figure
\ref{fig:ssrd-ribbonPredSE}). Similar seasonal patterns previously observed were
still apparent with subadults hauling out earlier in the season followed by
adult males and, then, adult females. The observed relationship with hour of
day and the centering of peak haul-out probability around solar noon was
reflected in these predictions as a one-sided distribution with maximum solar
radiation having the highest haul-out probability and minimal solar radiation
the least. The seasonal distribution of haul-out probability along with 95\%
confidence intervals also provided comparable insights (see figures
\ref{fig:ribbonPredSE} and \ref{fig:ssrd-ribbonPredSE}). That said, subtle
differences in the shape and extent of confidence limits were present.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/ssrd-prediction-heatmap-1} \caption{Predicted haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model. In this model, solar radiation was used in lieu of hour of day. The apparent seasonal progression with subadults hauling out earlier in the season followed by adult males and, then, adult females is still notable although maybe not as clear. Predictions for young of the year still show their transition from newly weaned pups resting on the ice to more in-water activities. The overall pattern is in agreement with a one-sided view of Figure 7 where maximum solar radiation is equivalent to local solar noon.}\label{fig:ssrd-prediction-heatmap}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/ssrd-ribbonPredSE-1} \caption{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals. In this model predictions are shown for low, medium, and high values of solar radiation (as percentages of the maximum value observed) in lieu of local solar hour. There's general agreement in the overall seasonal patterns between the two approaches but sublte differences in shape and extent of the confidence limits were present (see Figure S2 for comparisons).}\label{fig:ssrd-ribbonPredSE}
\end{figure}

\hypertarget{discussion-1}{%
\subsubsection{Discussion}\label{discussion-1}}

Solar radiation has potential as an informative
covariate in pinniped haul-out models that can be directly linked to seal
physiology and expected behavioral changes. The ERA5's \emph{surface solar radiation downwards}
values aligned with hour of day and maximum values
occurred at or just after local solar noon. This highlighted the informative
potential for this approach. However, despite an overall reduction in the
total number of parameters and degrees of freedom, AIC comparison still favored
the models for each species that included hour of day as a Fourier series. This
analysis was not intended to be a full comparison -- we simply want to
demonstrate the potential and inspire further investigation -- but, there are
three possibilities that might explain the preference for hour of day.

First,
there are a broad range of solar radiation values represented for each hour of the
day. Cloud cover, fog, and precipitation all reduce downward solar radiation at
the surface and we might expect this to impact haul-out probability. However, the photoperiod and the timing of sunrise and sunset are
not impacted by weather and seals may be responding to these signals more than
the amount of solar radiation. Additionally, this study spans a range of
physiological cycles and energetic needs and higher solar radiation may not be
a consistently driving influence on seals. Increased energy from the sun may
be important during molt but less so during pupping and breeding periods. Second,
the timing and duration of haul-out behavior may also be influenced by diel
patterns in weather (e.g.~lower winds in the morning) or ecosystem dynamics (e.g.
prey availability) that lead to a skewness in the distribution of haul-out
behavior that wouldn't be reliably captured by solar radiation values. Third,
this effort is only an initial effort to explore the use of solar radiation in
pinniped haul-out models. A more in depth and rigorous exploration of this topic
might discover an approach that results in a more parsimonious and preferred
model formulation.

Again, we want to acknowledge Anthony Fischbach for the suggestion during the
peer review process. We think this is an excellent example of the peer review
process working to improve the quality of our manuscript and advance the
scientific process. We hope others will take our example and expand on it
within future analyses.

\end{document}

### 2024-03-28 R/wx_plots.R
update weather plots
add lines, fix ylim, decrease alpha for point-range lines

@@ -33,7 +33,9 @@ plot_ribbon_wx <- function(ribbon_fit_obj,age_sex_colors,age_sex_labels) {
                     fill = age_sex), linetype = 0, alpha = 0.2,
                 show.legend = FALSE) +
     geom_pointrange(aes(ymin = lower95, ymax = upper95), 
-                    size = 0.25, alpha = 1, stroke = 0) +
+                    size = 0, alpha = 0.25, stroke = 0,
+                    show.legend = FALSE) +
+    geom_line() +
     scale_color_manual(values= age_sex_colors,
                        labels = age_sex_labels,
                        name = "Ribbon seal",
@@ -62,7 +64,8 @@ plot_ribbon_wx <- function(ribbon_fit_obj,age_sex_colors,age_sex_labels) {
                     fill = age_sex), linetype = 0, alpha = 0.2,
                 show.legend = FALSE) +
     geom_pointrange(aes(ymin = lower95, ymax = upper95), 
-                    size = 0.25, alpha = 1, stroke = 0) +
+                    size = 0, alpha = 0.25, stroke = 0) +
+    geom_line() +
     scale_color_manual(values= age_sex_colors,
                        labels = age_sex_labels,
                        guide = guide_legend(override.aes = list(alpha = 1),
@@ -87,7 +90,8 @@ plot_ribbon_wx <- function(ribbon_fit_obj,age_sex_colors,age_sex_labels) {
                     fill = age_sex), linetype = 0, alpha = 0.2,
                 show.legend = FALSE) +
     geom_pointrange(aes(ymin = lower95, ymax = upper95), 
-                    size = 0.25, alpha = 1, stroke = 0) +
+                    size = 0, alpha = 0.25, stroke = 0) +
+    geom_line() +
     scale_color_manual(values= age_sex_colors,
                        labels = age_sex_labels,
                        guide = guide_legend(override.aes = list(alpha = 1),
@@ -113,7 +117,8 @@ plot_ribbon_wx <- function(ribbon_fit_obj,age_sex_colors,age_sex_labels) {
                     fill = age_sex), linetype = 0, alpha = 0.2,
                 show.legend = FALSE) +
     geom_pointrange(aes(ymin = lower95, ymax = upper95), 
-                    size = 0.25, alpha = 1, stroke = 0) +
+                    size = 0, alpha = 0.25, stroke = 0) +
+    geom_line() +
[…]

### 2024-03-28 analysis/paper/London_HauloutBehavior.Rmd
move fig titles to caption
remove figure titles and include within the figure caption; also adjustments to the legend(s) to provide more detail to reader

@@
 Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
-The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season..
+The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
 While the marginal effect of the covariate is continuous, points and vertical lines representing the 95% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.
@@
 The full seasonal range of annual predictions along with uncertainty are presented as grey lines.
-Covariates were fixed at local solar noon and under smoothed weather conditions.
+Covariates were fixed at local solar noon and weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
 Only those groups (age:sex + year) that included observations from more than one seal are shown.
@@
   ylab("haul-out probability") +
-  labs(caption = "predictions are shown for local solar noon and under mean weather conditions") +
   theme(legend.position = "top",
@@
 
+\clearpage
 ## Supplemental Figures Showing Confidence Intervals Associated with Predictions
@@
 
-```{r beardedPredSE,fig.asp = 0.5, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for bearded seals.
-Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00) with smoothed weather covariates.
-Age and sex classes are combined into a single 'all ages' category."}
+(ref:beardedPredSEcaption) **Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for bearded seals.** \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
+Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
+Age and sex classes are combined into a single 'all ages' category.
+
+```{r beardedPredSE,fig.asp = 0.5, include = TRUE, fig.cap="(ref:beardedPredSEcaption)"}
 plot_df <- bearded_newdata %>%
@@
          lubridate::yday(date) >73,
-         solar_hour %in% c(7,12,17))
+         solar_hour %in% c(7,12,17)) |> 
+  mutate(species = 'Bearded seal')
 
 bearded_predse <- ggplot(plot_df, aes(date,ho_prob)) +
+  geom_ribbon(aes(ymin=lower95,ymax=upper95, fill = age_sex), 
+              fill = species_colors['Bearded seal'], alpha = 0.3) +
+  geom_line(aes(date,ho_prob, color = species)) +
+  scale_color_manual(name = 'Predicted haul-out probability',
+                     values = "#E58606",
+                     guide = guide_legend(title.position = "top",
+                                          title.hjust = 0,
+                                          direction = "horizontal")) +
+  facet_grid(age_sex ~ solar_hour,
+             scales = "free_x",
+                    space = "free_x",
+                    labeller = labeller(
+                      age_sex = age_sex_labels,
+                      solar_hour = local_solar_labels
+                    )) +
+  scale_y_continuous(breaks = c(0.2,0.4,0.6,0.8), 
+                     labels = c("0.2","0.4", "0.6", "0.8"), 
+                     limits = c(0,1)) +
+  xlab("day of year") +
+  ylab("haul-out probability") +
+  theme(legend.position = "top",
+        legend.justification = "left")
+
+bearded_predse
+```
+
+(ref:ribbonPredSEcaption) **Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals.** \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
+Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
+Age and sex classes are separated to allow comparisons.
+
+```{r ribbonPredSE,fig.asp = 0.7, include = TRUE, fig.cap="(ref:ribbonPredSEcaption)"}
+plot_df <- ribbon_newdata %>%
+  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
+         month = lubridate::month(date,label = TRUE),
+         day = lubridate::day(date)) %>%
+  filter(!month %in% c("Jul","Aug"),
+         lubridate::yday(date) >73,
+         solar_hour %in% c(7,12,17)) |> 
+  mutate(species = 'Ribbon seal')
+
+ribbon_predse <- ggplot(plot_df, aes(date,ho_prob)) +
   geom_ribbon(aes(ymin=lower95,ymax=upper95), 
-              fill = species_colors['Bearded seal'], alpha = 0.3) +
-  geom_line(aes(date,ho_prob),
-            color = species_colors['Bearded seal']) +
+              fill = species_colors['Ribbon seal'], alpha = 0.3) +
+  geom_line(aes(date,ho_prob, color = species)) +
+  scale_color_manual(name = 'Predicted haul-out probability',
+                     values = rep("#5D69B1",4),
+                     guide = guide_legend(title.position = "top",
+                                          title.hjust = 0,
+                                          direction = "horizontal")) +
   facet_grid(age_sex ~ solar_hour,
@@
   ylab("haul-out probability") +
-  labs(title = "Bearded seal predicted haul-out probability with uncertainty",
-       subtitle = "Predictions and associated 95% confidence intervals are shown for three **local solar** times and with smoothed weather covariates") +
-  theme(legend.position = "none")
-
-bearded_predse
-```
-
-```{r ribbonPredSE,fig.asp = 0.7, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for ribbon seals.
-Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00) with smoothed weather covariates.
-Age and sex classes are separated to allow comparisons."}
-plot_df <- ribbon_newdata %>%
+  theme(legend.position = "top",
+        legend.justification = "left")
+
+ribbon_predse
+```
+
+(ref:spottedPredSEcaption) **Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for spotted seals.** \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
+Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
+Age and sex classes are separated to allow comparisons.
+
+```{r spottedPredSE,fig.asp = 0.7, include = TRUE, fig.cap="(ref:spottedPredSEcaption)"}
+plot_df <- spotted_newdata %>%
   mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
@@
          lubridate::yday(date) >73,
-         solar_hour %in% c(7,12,17))
-
-ribbon_predse <- ggplot(plot_df, aes(date,ho_prob)) +
+         solar_hour %in% c(7,12,17)) |> 
+  mutate(species = 'Spotted seal')
+
+spotted_predse <- ggplot(plot_df, aes(date,ho_prob)) +
   geom_ribbon(aes(ymin=lower95,ymax=upper95), 
-              fill = species_colors['Ribbon seal'], alpha = 0.3) +
-  geom_line(aes(date,ho_prob),
-            color = species_colors['Ribbon seal']) +
+              fill = species_colors['Spotted seal'], alpha = 0.3) +
+  geom_line(aes(date,ho_prob, color = species)) +
+  scale_color_manual(name = 'Predicted haul-out probability',
+                     values = rep("#52BCA3",4),
+                     guide = guide_legend(title.position = "top",
+                                          title.hjust = 0,
+                                          direction = "horizontal")) +
   facet_grid(age_sex ~ solar_hour,
@@
   ylab("haul-out probability") +
-  labs(title = "Ribbon seal predicted haul-out probability with uncertainty",
-       subtitle = "Predictions and associated 95% confidence intervals are shown for three **local solar** times and with smoothed weather covariates") +
-  theme(legend.position = "none")
-
-ribbon_predse
-```
-
-```{r spottedPredSE,fig.asp = 0.7, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for spotted seals.
-Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00) with smoothed weather covariates.
-Age and sex classes are separated to allow comparisons."}
-plot_df <- spotted_newdata %>%
-  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
-         month = lubridate::month(date,label = TRUE),
-         day = lubridate::day(date)) %>%
-  filter(!month %in% c("Jul","Aug"),
-         lubridate::yday(date) >73,
-         solar_hour %in% c(7,12,17))
-
-spotted_predse <- ggplot(plot_df, aes(date,ho_prob)) +
-  geom_ribbon(aes(ymin=lower95,ymax=upper95), 
-              fill = species_colors['Spotted seal'], alpha = 0.3) +
-  geom_line(aes(date,ho_prob),
-            color = species_colors['Spotted seal']) +
-  facet_grid(age_sex ~ solar_hour,
-             scales = "free_x",
-                    space = "free_x",
-                    labeller = labeller(
-                      age_sex = age_sex_labels,
-                      solar_hour = local_solar_labels
-                    )) +
-  scale_y_continuous(breaks = c(0.2,0.4,0.6,0.8), labels = c("0.2","0.4", "0.6", "0.8"), limits = c(0,1)) +
-  xlab("day of year") +
-  ylab("haul-out probability") +
-  labs(title = "Spotted seal predicted haul-out probability with uncertainty",
-       subtitle = "Predictions and associated 95% confidence intervals are shown for three **local solar** times and with smoothed weather covariates") +
-  theme(legend.position = "none")
+  theme(legend.position = "top",
+        legend.justification = "left")
 
@@
 
+\clearpage
 ## Exploring Insolation (Solar Radiation) as a Model Covariate
@@
 
-```{r era5-ssrd-plot, fig.asp = 0.75, include=TRUE, fig.cap="Downward surface solar radiation estimates from the ERA5 climate reanalysis for 5000 random points within the study area between 2005 and 2021.
-Solar radiation values are presented in Watts per square-meter and the smoothed line highlights the strong diel pattern."}
-
-pts_ssrd_sf <- readRDS(here::here('data/random_ssrd_pts.rds'))
+(ref:era5ssrdcaption) **Diel Pattern of Solar Radiation Values from ERA5 Reanalysis.** \linebreak Downward surface solar radiation estimates from the ERA5 climate reanalysis for 5000 random points within the study area between 2005 and 2021.
+Solar radiation values are presented in Watts per square-meter and the smoothed line highlights the strong diel pattern.
+
+```{r era5-ssrd-plot, fig.asp = 0.75, include=TRUE, fig.cap="(ref:era5ssrdcaption)"}
+
+pts_ssrd_sf <- readRDS(here::here('data/random_ssrd_pts.rds')) |> 
+  mutate(era5_value = "downward surface solar radiation")
 
 ssrd_solarhour_plot <- ggplot(data = pts_ssrd_sf, aes(x=solar_hour, y=value)) + 
-  geom_point(size = 0.5, alpha = 0.3, color = "#edd9a3") + 
+  geom_point(aes(color = era5_value), size = 0.8, alpha = 0.3) +
+  scale_color_manual(name = 'ERA5 Reanalysis',
+                     values = "#edd9a3",
+                     guide = guide_legend(title.position = "top",
+                                          title.hjust = 0,
+                                          direction = "horizontal",
+                                          override.aes = list(alpha = 1,
+                                                              size = 2))) +
   geom_smooth(se=TRUE, linewidth = 0.2, color = "#4b2991") +
@@
   ylab("Solar Radiation") +
-  labs(title = "Diel Pattern of Solar Radiation Values from ERA5 Reanalysis",
-       subtitle = "A random sample of hourly ERA5 downward surface solar radiation values for points within the study area confirms a unimodal relationship with local solar hour.",
-       caption = "solar radiation values are presented in Watts per meter^2")
+  labs(caption = "solar radiation values are presented in Watts per meter^2") +
+  theme(legend.position = "top",
+        legend.justification = "left")
 
@@
 
-```{r ssrd-prediction-heatmap, include = TRUE, fig.asp=0.9, fig.cap="Predicted haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
+(ref:ssrdHeatMapcaption) **Solar radiation as a predictor of ribbon seal haul-out probability.** \linebreak Predicted haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
 In this model, solar radiation was used in lieu of hour of day.
@@
 Predictions for young of the year still show their transition from newly weaned pups resting on the ice to more in-water activities.
-The overall pattern is in agreement with a one-sided view of Figure 7 where maximum solar radiation is equivalent to local solar noon."}
+The overall pattern is in agreement with a one-sided view of Figure 7 where maximum solar radiation is equivalent to local solar noon.
+
+```{r ssrd-prediction-heatmap, include = TRUE, fig.asp=0.9, fig.cap="(ref:ssrdHeatMapcaption)"}
 r <- ribbon_newdata |> 
@@
                           begin = 0.1, end = 1,
-                       name = "haul-out probability", 
+                       name = "Ribbon seal haul-out probability", 
                        limits = c(0,1), breaks = c(0.25, 0.50, 0.75),
-                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
-                                              barheight = 0.5, title.hjust = 0.5)
+                       guide = guide_colorbar(title.position = "top", barwidth = 15,
+                                              barheight = 0.5, title.hjust = 0)
   ) +
@@
   coord_cartesian(expand=FALSE)
-p <- p + theme(panel.spacing.x = unit(0.1, "lines"),
-               panel.spacing.y = unit(0.2, "lines"),
-               legend.position = "bottom",
+
+p <- p + theme(panel.spacing = unit(0.1, "lines"),
+               legend.position = "top", legend.justification='left',
+               legend.text.align = 0,
                strip.background = element_rect(colour="white"),
                axis.ticks=element_blank()) +
-  xlab("day of month") + ylab("% of max solar radiation") +
-  ggtitle("Solar radiation as a predictor of ribbon seal haul-out probability",
-          subtitle = "Increased haul-out probability at higher levels of solar
-          radiation is consistent with maximum predicted haul-out at solar noon")
+  xlab("day of month") + ylab("% of max solar radiation")
+
 p
@@
 
-```{r ssrd-ribbonPredSE,fig.asp = 0.7, include = TRUE, fig.cap="Seasonal variability in haul-out probability and the associated 95\\% confidence intervals (shaded area) for ribbon seals.
+(ref:ssrdRibbonPredSEcaption) **Solar radiation as a predictor of ribbon seal haul-out probability (with uncertainty).** \linebreak Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals.
 In this model predictions are shown for low, medium, and high values of solar radiation (as percentages of the maximum value observed) in lieu of local solar hour.
-There's general agreement in the overall seasonal patterns between the two approaches but sublte differences in shape and extent of the confidence limits were present (see Figure S2 for comparisons)."}
+There's general agreement in the overall seasonal patterns between the two approaches but sublte differences in shape and extent of the confidence limits were present (see Figure S2 for comparisons).
+
+```{r ssrd-ribbonPredSE,fig.asp = 0.7, include = TRUE, fig.cap="(ref:ssrdRibbonPredSEcaption)"}
 
@@
   inner_join(ssrd_facet_labels, by = join_by(ssrd_quants == ssrd_facet)) |> 
-  rename(ssrd_facet_label = value)
+  rename(ssrd_facet_label = value) |> 
+  mutate(species = 'Ribbon seal')
 
@@
               fill = species_colors['Ribbon seal'], alpha = 0.3) +
-  geom_line(aes(date,fit),
-            color = species_colors['Ribbon seal']) +
+  geom_line(aes(date,fit, color = species)) +
+  scale_color_manual(name = 'Predicted haul-out probability',
+                     values = "#5D69B1",
+                     guide = guide_legend(title.position = "top",
+                                          title.hjust = 0,
+                                          direction = "horizontal")) +
   facet_grid(age_sex ~ ssrd_facet_label,
@@
   ylab("haul-out probability") +
-  labs(title = "Solar radiation as a predictor of haul-out probability (with uncertainty)",
-       subtitle = "Predictions and associated 95% confidence intervals for three **solar radiation** quantile categories show similar seasonal patterns to those based on local solar hour") +
-  theme(legend.position = "none")
+  theme(legend.position = "top",
+        legend.justification = "left")
 
@@
 
-<!-- - ear5 solar radiation has potential as an informative covariate that is a more -->
-<!-- directly linked with seal behavior and physiology -->
-<!-- - the unimodal distribution and correlation with hour of day also demonstrates -->
-<!-- it could be considered a replacement for hour of day; though, with increased -->
-<!-- variability given the impact of cloud cover and other weather parameters -->
-<!-- - despite the reduced number of parameters, AIC comparison for all species -->
-<!-- still preferred the model with local solar hour.  -->
-<!-- - there is some a slight skewness present in the predicted distributions with -->
-<!-- higher predictions in the morning compared to the late afternoon; this isn't -->
-<!-- captured with solar radiation alone -->
-
-Solar radiation has potential as an informative
-covariate in pinniped haul-out models that can be directly linked to seal
-physiology and expected behavioral changes. The ERA5's _surface solar radiation downwards_
-values aligned with hour of day and maximum values
-occurred at or just after local solar noon. This highlighted the informative
-potential for this approach. However, despite an overall reduction in the
-total number of parameters and degrees of freedom, AIC comparison still favored
-the models for each species that included hour of day as a Fourier series. This
-analysis was not intended to be a full comparison -- we simply want to 
-demonstrate the potential and inspire further investigation -- but, there are
-three possibilities that might explain the preference for hour of day.
-
-First,
-there are a broad range of solar radiation values represented for each hour of the
-day. Cloud cover, fog, and precipitation all reduce downward solar radiation at
-the surface and we might expect this to impact haul-out probability. However, the photoperiod and the timing of sunrise and sunset are
+Solar radiation has potential as an informative covariate in pinniped haul-out
+models that can be directly linked to seal physiology and expected behavioral
+changes. The ERA5's _surface solar radiation downwards_ values aligned with hour
+of day and maximum values occurred at or just after local solar noon. This
+highlighted the informative potential for this approach. However, despite an
+overall reduction in the total number of parameters and degrees of freedom, AIC
+comparison still favored the models for each species that included hour of day
+as a Fourier series.
+
+This analysis was not intended to be a full comparison --
+we simply want to demonstrate the potential and inspire further investigation --
+but, there are three possibilities that might explain the preference for hour of
+day. First, there are a broad range of solar radiation values represented for each
+hour of the day. Cloud cover, fog, and precipitation all reduce downward solar
+radiation at the surface and we might expect this to impact haul-out
+probability. However, the photoperiod and the timing of sunrise and sunset are
 not impacted by weather and seals may be responding to these signals more than
 the amount of solar radiation. Additionally, this study spans a range of
-physiological cycles and energetic needs and higher solar radiation may not be
-a consistently driving influence on seals. Increased energy from the sun may
-be important during molt but less so during pupping and breeding periods. Second,
+physiological cycles and energetic needs and higher solar radiation may not be a
+consistent driving influence on seals. Increased energy from the sun may be
+important during molt but less so during pupping and breeding periods. Second,
 the timing and duration of haul-out behavior may also be influenced by diel
-patterns in weather (e.g. lower winds in the morning) or ecosystem dynamics (e.g.
-prey availability) that lead to a skewness in the distribution of haul-out 
+patterns in weather (e.g. lower winds in the morning) or ecosystem dynamics
+(e.g. prey availability) that lead to a skewness in the distribution of haul-out
 behavior that wouldn't be reliably captured by solar radiation values. Third,

### 2024-03-28 analysis/paper/London_HauloutBehavior.tex
move fig titles to caption
remove figure titles and include within the figure caption; also adjustments to the legend(s) to provide more detail to reader

@@
 Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
-The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season..
+The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
 While the marginal effect of the covariate is continuous, points and vertical lines representing the 95\% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.}\label{fig:spottedHOwx}
@@
 
+\clearpage
+
 \hypertarget{supplemental-figures-showing-confidence-intervals-associated-with-predictions}{%
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/beardedPredSE-1} \caption{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for bearded seals.
-Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00) with smoothed weather covariates.
-Age and sex classes are combined into a single 'all ages' category.}\label{fig:beardedPredSE}
+\includegraphics[width=1\linewidth]{../figures/beardedPredSE-1} \caption{\textbf{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for bearded seals.} \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
+Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
+Age and sex classes are combined into a single `all ages' category.}\label{fig:beardedPredSE}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/ribbonPredSE-1} \caption{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals.
-Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00) with smoothed weather covariates.
+\includegraphics[width=1\linewidth]{../figures/ribbonPredSE-1} \caption{\textbf{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals.} \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
+Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
 Age and sex classes are separated to allow comparisons.}\label{fig:ribbonPredSE}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/spottedPredSE-1} \caption{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for spotted seals.
-Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00) with smoothed weather covariates.
+\includegraphics[width=1\linewidth]{../figures/spottedPredSE-1} \caption{\textbf{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for spotted seals.} \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
+Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
 Age and sex classes are separated to allow comparisons.}\label{fig:spottedPredSE}
 \end{figure}
+
+\clearpage
 
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/era5-ssrd-plot-1} \caption{Downward surface solar radiation estimates from the ERA5 climate reanalysis for 5000 random points within the study area between 2005 and 2021.
+\includegraphics[width=1\linewidth]{../figures/era5-ssrd-plot-1} \caption{\textbf{Diel Pattern of Solar Radiation Values from ERA5 Reanalysis.} \linebreak Downward surface solar radiation estimates from the ERA5 climate reanalysis for 5000 random points within the study area between 2005 and 2021.
 Solar radiation values are presented in Watts per square-meter and the smoothed line highlights the strong diel pattern.}\label{fig:era5-ssrd-plot}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/ssrd-prediction-heatmap-1} \caption{Predicted haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
+\includegraphics[width=1\linewidth]{../figures/ssrd-prediction-heatmap-1} \caption{\textbf{Solar radiation as a predictor of ribbon seal haul-out probability.} \linebreak Predicted haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
 In this model, solar radiation was used in lieu of hour of day.
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/ssrd-ribbonPredSE-1} \caption{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals.
+\includegraphics[width=1\linewidth]{../figures/ssrd-ribbonPredSE-1} \caption{\textbf{Solar radiation as a predictor of ribbon seal haul-out probability (with uncertainty).} \linebreak Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals.
 In this model predictions are shown for low, medium, and high values of solar radiation (as percentages of the maximum value observed) in lieu of local solar hour.
@@
 
-Solar radiation has potential as an informative
-covariate in pinniped haul-out models that can be directly linked to seal
-physiology and expected behavioral changes. The ERA5's \emph{surface solar radiation downwards}
-values aligned with hour of day and maximum values
-occurred at or just after local solar noon. This highlighted the informative
-potential for this approach. However, despite an overall reduction in the
-total number of parameters and degrees of freedom, AIC comparison still favored
-the models for each species that included hour of day as a Fourier series. This
-analysis was not intended to be a full comparison -- we simply want to
-demonstrate the potential and inspire further investigation -- but, there are
-three possibilities that might explain the preference for hour of day.
-
-First,
-there are a broad range of solar radiation values represented for each hour of the
-day. Cloud cover, fog, and precipitation all reduce downward solar radiation at
-the surface and we might expect this to impact haul-out probability. However, the photoperiod and the timing of sunrise and sunset are
+Solar radiation has potential as an informative covariate in pinniped haul-out
+models that can be directly linked to seal physiology and expected behavioral
+changes. The ERA5's \emph{surface solar radiation downwards} values aligned with hour
+of day and maximum values occurred at or just after local solar noon. This
+highlighted the informative potential for this approach. However, despite an
+overall reduction in the total number of parameters and degrees of freedom, AIC
+comparison still favored the models for each species that included hour of day
+as a Fourier series.
+
+This analysis was not intended to be a full comparison --
+we simply want to demonstrate the potential and inspire further investigation --
+but, there are three possibilities that might explain the preference for hour of
+day. First, there are a broad range of solar radiation values represented for each
+hour of the day. Cloud cover, fog, and precipitation all reduce downward solar
+radiation at the surface and we might expect this to impact haul-out
+probability. However, the photoperiod and the timing of sunrise and sunset are
 not impacted by weather and seals may be responding to these signals more than
 the amount of solar radiation. Additionally, this study spans a range of
-physiological cycles and energetic needs and higher solar radiation may not be
-a consistently driving influence on seals. Increased energy from the sun may
-be important during molt but less so during pupping and breeding periods. Second,
+physiological cycles and energetic needs and higher solar radiation may not be a
+consistent driving influence on seals. Increased energy from the sun may be
+important during molt but less so during pupping and breeding periods. Second,
 the timing and duration of haul-out behavior may also be influenced by diel
-patterns in weather (e.g.~lower winds in the morning) or ecosystem dynamics (e.g.
-prey availability) that lead to a skewness in the distribution of haul-out
+patterns in weather (e.g.~lower winds in the morning) or ecosystem dynamics
+(e.g.~prey availability) that lead to a skewness in the distribution of haul-out
 behavior that wouldn't be reliably captured by solar radiation values. Third,

### 2024-03-28 analysis/paper/Letter_Response_to_Reviewers.qmd
final edits, reviewer response, compare doc

@@
 
-```{=tex}
 \newenvironment{author-response}{\color{Bittersweet}}
-
-```
 
@@
 While many of the comments and suggestions from reviewers were minor, we feel
-they deserved appropriate time, consideration, and exploration. We hope our
-responses here reflect that consideration and we are grateful to the reviewers
-for their thoughtful comments. Lastly, we also used this opportunity to improve
-the quality of the underlying telemetry data set and expand the reproducibility
-of the analysis to meet or exceed expectations for open science, data
-availability, and code availability.
+they deserved appropriate time, consideration, and exploration. We have made
+several changes to the manuscript. We hope our responses here reflect that
+consideration and communicate our edits. We are grateful to the reviewers for
+their thoughtful comments. Lastly, we also used this opportunity to improve the
+quality of the underlying telemetry data set and expand the reproducibility of
+the analysis to meet or exceed expectations for open science, data availability,
+and code availability.
 
 Below, please find detailed responses to each of the reviewer comments. In
-addition to this letter, several of the suggestions are documented as specific
-issues within the GitHub repository and the full suite of updates are all
-contained and documented within a single pull request. We have also provided a
-pdf document that indicates all changes between the revised manuscript and the
-submitted manuscript.
+addition to this letter, the full suite of updates are all contained and
+documented within a single pull request. We have also provided a pdf document
+that indicates all changes between the revised manuscript and the submitted
+manuscript.
 
@@
 inclusion of additional covariates (such as weather products) in haul-out models
-only serves to increase standard errors in predictions.  The benefit of
-including weather covariates is to more accurately explain variation in survey
-counts; it is thus more relevant to reducing bias in aerial survey estimates
-than it is to increasing precision.
+only serves to increase standard errors in predictions. The benefit of including
+weather covariates is to more accurately explain variation in survey counts; it
+is thus more relevant to reducing bias in aerial survey estimates than it is to
+increasing precision.
 :::
@@
 ::: author-response
-**\[author response\]** Clearly there are a large number impacts of loss of
-sea ice to human communities (and ecosystems) beyond the Arctic region to the
-rest of the world. However, for this sentence, we wanted to keep the reader
-focused on the Arctic region.
+**\[author response\]** Clearly there are a large number impacts of loss of sea
+ice to human communities (and ecosystems) beyond the Arctic region to the rest
+of the world. However, for this sentence, we wanted to keep the reader focused
+on the Arctic region.
 :::
@@
 Chukchi, and western Beaufort seas by multiple organizations as part of
-collaborative investigations from 2005 through 2020."
+collaborative investigations from 2005 through 2021."
 :::
@@
 of 100% dry observations and with locations indicating the tag was on land would
-be truncated to the final stretch of 100% dry observations. The vast
-majority of deployments end with the device detaching in the water and the
-deployment end date is obvious. There is no perfect algorithm for identifying
-deployment end dates and each deployment in question must be considered
-separately. While not perfect, we are confident our reliance on expert judgement
-and examination of multiple data streams provides the best option."
+be truncated to the final stretch of 100% dry observations. The vast majority of
+deployments end with the device detaching in the water and the deployment end
+date is obvious. There is no perfect algorithm for identifying deployment end
+dates and each deployment in question must be considered separately. While not
+perfect, we are confident our reliance on expert judgement and examination of
+multiple data streams provides the best option."
 :::
@@
 communicate our procedure for processing the large NARR data set and that we
-used the original grid size of 32km. Regarding the calculation of wind vectors, we
-include this detail because many readers, like us initially, may not be well
+used the original grid size of 32km. Regarding the calculation of wind vectors,
+we include this detail because many readers, like us initially, may not be well
 versed in how wind vectors components are converted to wind speed.
@@
 
-::: author-response
 **\[author response\]** We have revised the text in this section to try to
@@
 ice, and we could assess the local concentration of sea ice during these events
-(see [@olnes2020]). This, however, expands the scope of our analysis into the
-realm of habitat selection and many of our deployments consisted of a single
+(see Olnes et al., 2020). This, however, expands the scope of our analysis into
+the realm of habitat selection and many of our deployments consisted of a single
 device attached to the rear flipper of the seal which meant we only received
@@
 availability and characteristics of sea-ice is rapidly changing is important
-[@breed2018; @cameron2018] but ancillary to the focus of this analysis -- and,
-in the end, not possible given key limitations of our data. Additionally, our
-study was limited to the spring season when seal haul-out behaviors are strongly
-influenced by pupping, nursing, breeding behavior, and molt. These drivers are
-likely more influential on haul-out behavior than sea-ice concentration.
-Crawford et al. [-@crawford2019] compared haul-out probability models for ringed
-seals and found those that only included season (and not sea-ice concentration)
-were the most parsimonious. For all of these reasons, we have elected not to use
-sea ice concentration _as a predictor for haul-out probability_ in the present
-study.
-:::
+(Breed et al., 2018; Cameron et al., 2018) but ancillary to the focus of this
+analysis -- and, in the end, not possible given key limitations of our data.
+Additionally, our study was limited to the spring season when seal haul-out
+behaviors are strongly influenced by pupping, nursing, breeding behavior, and
+molt. These drivers are likely more influential on haul-out behavior than
+sea-ice concentration. Crawford et al. (2019) compared haul-out probability
+models for ringed seals and found those that only included season (and not
+sea-ice concentration) were the most parsimonious. For all of these reasons, we
+have elected not to use sea ice concentration *as a predictor for haul-out
+probability* in the present study.
 
@@
 captions to include the title information to be consistent with the journal
-styling and requirements. We've also revised many of the
-figure legends to better communicate the information to the reader and ensure
-if any figures are presented out of context (e.g. without the figure caption) in
-other venues the information is clear.
+styling and requirements. We've also revised many of the figure legends to
+better communicate the information to the reader and ensure if any figures are
+presented out of context (e.g. without the figure caption) in other venues the
+information is clear.
 :::
@@
 > ***Figure 6:** Please use standard conventions for plot titles. Your title
-> deserves a statement in your Results or Discussion section. It does not
-> follow the expectations for a publication figure title.*
+> deserves a statement in your Results or Discussion section. It does not follow
+> the expectations for a publication figure title.*
 
@@
 > co-variates. Once you perform this, you may state that inclusion of weather
-> co-variates improved haulout state estimates. However, you have not done
-> this. If you wish to state this, please do this.*
+> co-variates improved haulout state estimates. However, you have not done this.
+> If you wish to state this, please do this.*
 
@@
 
-The most challenging element in developing
-availability correction factors is with annual variability. It can be difficult
-to get a sufficient sample size to estimate year-specific correction factors,
-particularly because research teams would likely need to deploy bio-loggers on
-seals and conduct aerial surveys concurrently, requiring considerably more
-personnel and money. One possible suggestion is to include year as a random
-effect within models for aerial survey counts such that, without specific
-knowledge of any particular year, the among-year variance is included in the
-modeled standard errors. Regardless of the specific approach, future estimates
-of Arctic seal abundance will require specific consideration of annual
-variability and changes in the timing of peak haul-out behavior when estimating
-trends, as one will not know if moderate differences in abundance estimates are
-attributable to changes in abundance or changes in haul-out behavior.
+The most challenging element in developing availability correction factors is
+with annual variability. It can be difficult to get a sufficient sample size to
+estimate year-specific correction factors, particularly because research teams
+would likely need to deploy bio-loggers on seals and conduct aerial surveys
+concurrently, requiring considerably more personnel and money. One possible
+suggestion is to include year as a random effect within models for aerial survey
+counts such that, without specific knowledge of any particular year, the
+among-year variance is included in the modeled standard errors. Regardless of
+the specific approach, future estimates of Arctic seal abundance will require
+specific consideration of annual variability and changes in the timing of peak
+haul-out behavior when estimating trends, as one will not know if moderate
+differences in abundance estimates are attributable to changes in abundance or
+changes in haul-out behavior.
 
@@
 analyses. We have also emphasized that our reliance on flipper-mounted
-bio-loggers for a portion of our data limited our understanding of movement
-and use away from sea-ice.
+bio-loggers for a portion of our data limited our understanding of movement and
+use away from sea-ice.
 :::

### 2024-03-28 analysis/paper/London_HauloutBehavior.tex
final edits, reviewer response, compare doc

@@
 
-\author[2]{Lori T. Quakenbush}
+\author[2]{Lori Quakenbush}
 
@@
 For ribbon and spotted seals, models with year effects were highly supported, indicating that the timing and magnitude of haul-out behavior varied among years.
-However, we did not find broad evidence that haul-out timing was linked to annual sea ice extent.
+However, we did not find broad evidence that haul-out timing was linked to annual sea-ice extent.
 Our analysis emphasizes the importance of accounting for seasonal and temporal variation in haul-out behavior, as well as associated environmental covariates, when interpreting the number of seals counted in aerial surveys.
@@
 these species are unknown, so it is difficult to assess the effect, if any,
-declines in sea ice habitat have had, or will have, on seal demography.
+declines in sea-ice habitat have had, or will have, on seal demography.
 
@@
 summer (Frost \& Burns, 2018). Ringed seals (\emph{Phoca (pusa) hispida}) --- not included in this
-study --- haul out on sea-ice but primarily use snow lairs on sea-ice during
+study --- haul out on sea ice but primarily use snow lairs on sea ice during
 winter and spring.
@@
 timing and possible linkage to changes in the extent of seasonal sea ice between
-2005 and 2020. Our work extends the scope of previous haul-out analyses,
+2005 and 2021. Our work extends the scope of previous haul-out analyses,
 includes the influence of weather variability, and investigates the
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Year}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Continuous}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{For\ the\ set\ of\ models\ examining\ inter-annual\ variation\ in\ sea\ ice\ use,\ we\ fitted\ models\ with\ the\ addition\ of\ year\ by\ day-of-year\ interactions.}}} \\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Year}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Continuous}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 2.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{For\ the\ set\ of\ models\ examining\ inter-annual\ variation\ in\ sea-ice\ use,\ we\ fitted\ models\ with\ the\ addition\ of\ year\ by\ day-of-year\ interactions.}}} \\
 
@@
 computational efficiency, we used generalized linear mixed pseudo-models
-(GLMPMs; (Ver Hoef, London \& Boveng, 2010)) to model variation in haul-out behavior as a function
-of (1) covariate predictors, (2) temporally autocorrelated random effects, and
-(3) individual random effects representing heterogeneity in individual behavior.
-We used the glmmLDTS package (Ver Hoef, London \& Boveng, 2010) to implement GLMPMs. We explored
-two different model formulations for our data and we fit separate models to
-bearded, ribbon, and spotted seal data sets as we expected differing behavior by
-species. Separate models for each species were also needed because a single,
-very large data set proved computationally intractable. In our first model
-formulation and for each species, we fitted a year-independent model that
-predicted average haul-out behavior as a function of demographic, weather,
-seasonal, and diurnal effects. Second, for ribbon and spotted seals (which had
-considerably more data than bearded seals), we fitted models that included all
-the effects from the first model, but also permitted annual variation in
-haul-out timing. This second set of models was used to examine whether
-haul-out patterns varied by year and to determine the annual timing of apparent
-peaks in haul-out behavior. For both models, we assumed an hourly Bernoulli
-response (a binary 0-1 response dependent upon whether the tag was greater than
-50\% dry for a given hour) where the linear predictor was
-modeled on the logit scale. This is
+(GLMPMs; Ver Hoef, London \& Boveng (2010)) to model variation in
+haul-out behavior as a function of (1) covariate predictors, (2) temporally
+autocorrelated random effects, and (3) individual random effects representing
+heterogeneity in individual behavior. We used the glmmLDTS package
+(Ver Hoef, London \& Boveng, 2010) to implement GLMPMs. We explored two different model
+formulations for our data and we fit separate models to bearded, ribbon, and
+spotted seal data sets as we expected differing behavior by species. Separate
+models for each species were also needed because a single, very large data set
+proved computationally intractable. In our first model formulation and for each
+species, we fitted a year-independent model that predicted average haul-out
+behavior as a function of demographic, weather, seasonal, and diurnal effects.
+Second, for ribbon and spotted seals (which had considerably more data than
+bearded seals), we fitted models that included all the effects from the first
+model, but also permitted annual variation in haul-out timing. This second set
+of models was used to examine whether haul-out patterns varied by year and to
+determine the annual timing of apparent peaks in haul-out behavior. For both
+models, we assumed an hourly Bernoulli response (a binary 0-1 response dependent
+upon whether the tag was greater than 50\% dry for a given hour) where the linear
+predictor was modeled on the logit scale. This is
 consistent with previous approaches (London et al., 2012; Ver Hoef et al., 2014) and only
@@
 
-For the second set of models examining inter-annual variation in sea ice use, we
+For the second set of models examining inter-annual variation in sea-ice use, we
 fitted models with year by day-of-year interactions. However, in this case we
@@
 determine annual peaks in haul-out activity. The timing of peak haul-out
-behavior was further used to regress against the annual maximum sea ice extent
+behavior was further used to regress against the annual maximum sea-ice extent
 in the study region. Predictions before 15 March and after 30 June were not
@@
 
-We also assessed whether the annual variation in maximum spring sea ice extent in the
+We also assessed whether the annual variation in maximum spring sea-ice extent in the
 Bering Sea influenced the seasonal peak of seal haul-out probability. In
-particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
+particular, we used sea-ice concentration data from the Nimbus-7 SMMR and DMSP
 SSM/I-SSMIS Passive Microwave Dataset, Version 1 (Cavalieri et al., 1996) to calculate
-maximum sea ice extent. All sea ice concentration grid cells (25 km\textsuperscript{2}) in the
+maximum sea-ice extent. All sea-ice concentration grid cells (25 km\textsuperscript{2}) in the
 study area with greater than 15\% concentration were counted daily to get the total sea
 ice extent for each day between 15 February and 15 July across all years.
-Maximum spring sea ice extent was simply the largest daily count of grid cells
+Maximum spring sea-ice extent was simply the largest daily count of grid cells
 with greater than 15\% concentration for each year. A separate regression model,
@@
 between the annual computed peak haul-out day (as the response) with the
-maximum sea ice extent (as the predictor).
+maximum sea-ice extent (as the predictor).
 
@@
 male spotted seals appeared to have
-no relationship with the amount of yearly maximum sea ice extent in the
+no relationship with the amount of yearly maximum sea-ice extent in the
 study area. For ribbon seals and adult male spotted seals,
@@
 Notably missing from our list of haul-out model explanatory variables is any
-spatial-temporal representation of sea ice concentration, area, or extent. This
+spatial-temporal representation of sea-ice concentration, area, or extent. This
 may seem counterintuitive when modeling the haul-out behavior of seal species
@@
 device attached to the rear flipper of the seal which meant we only received
-locations when seals were hauled out on sea-ice, limiting our ability to fully
-evaluate fine-scale habitat preferences related to sea-ice. Insight into how
-seals use and interact with sea-ice during an extended period when the
-availability and characteristics of sea-ice is rapidly changing is important
+locations when seals were hauled out on sea ice, limiting our ability to fully
+evaluate fine-scale habitat preferences related to sea ice. Insight into how
+seals use and interact with sea ice during an extended period when the
+availability and characteristics of sea ice is rapidly changing is important
 (Breed et al., 2018; Cameron et al., 2018) but ancillary to the focus of this analysis -- and,
@@
 were the most parsimonious. For all of these reasons, we have elected not to use
-sea ice concentration \emph{as a predictor for haul-out probability} in the present
+sea-ice concentration \emph{as a predictor for haul-out probability} in the present
 study.
@@
 2018 -- 2019, years of extreme low spring ice extent in the Bering Sea that
-appeared to impact body condition of ribbon and spotted seals (Boveng et al., 2020), so we may
-currently lack sufficient contrast in ice extent to characterize an effect on
-the timing of peak haul-out probability. We should expect, however, that some minimal threshold in
-the spatial extent or density of sea ice will have a meaningful impact on the
-timing of peak haul-out behavior --- if there is no sea-ice, seals will not haul
-out or they will be forced to use terrestrial haul-out sites which were not part of the
-evolution of their normal behaviors. Additionally, while from an
-ecological perspective the haul-out behavior appears consistent, the interannual
-differences in timing and magnitude are large enough to have important
-ramifications on calculations of abundance and trend. Those ramifications will
-only be exacerbated if climate variability amplifies interannual differences.
+appeared to impact body condition of ribbon and spotted seals (Boveng et al., 2020), so
+we may currently lack sufficient contrast in ice extent to characterize an
+effect on the timing of peak haul-out probability. We should expect, however,
+that some minimal threshold in the spatial extent or density of sea ice will
+have a meaningful impact on the timing of peak haul-out behavior --- if there is
+no sea-ice, seals will not haul out or they will be forced to use terrestrial
+haul-out sites which were not part of the evolution of their normal behaviors.
+Additionally, while from an ecological perspective the haul-out behavior appears
+consistent, the interannual differences in timing and magnitude are large enough
+to have important ramifications on calculations of abundance and trend. Those
+ramifications will only be exacerbated if climate variability amplifies
+interannual differences.
 

### 2024-03-29 analysis/paper/Letter_Response_to_Reviewers.qmd
update readme (Rmd and md) for GitHub

@@
 addition to this letter, the full suite of updates are all contained and
-documented within a single pull request. We have also provided a pdf document
-that indicates all changes between the revised manuscript and the submitted
-manuscript.
+documented within a single pull request (see
+<https://github.com/noaa-afsc/berchukseals-haulout/pull/51>). We have also
+provided a pdf document (*\[Compare Report\] London_HauloutBehavior.pdf\]*) that
+indicates all changes between the compedium supporting this revised manuscript
+and the compendium supporting the originally submitted manuscript.
 

### 2024-03-29 README.Rmd

@@
 
-## Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas
+## Spring haul-out behavior of seals in the Bering and Chukchi seas
 Josh M.
 London^1^,✉, Paul B.
-Conn^1^, Stacie K.
-Hardy^1^, Erin L.
+Conn^1^, Stacie M.
+Koslovsky^1^, Erin L.
 Richmond^1^, Jay M.
@@
 Cameron^1^, Justin Crawford^2^, Andrew L.
-Von Duyke^3^, Lori T.
-Quakenbush^2^, and Peter L.
+Von Duyke^3^, Lori Quakenbush^2^, and Peter L.
 Boveng^1^
@@
 As the manuscript works its way through peer review and the publications process,
-additional reproducibility features (e.g. Dockerfile, MyBinder.org) and
-an R/Shiny application to explore results will be made developed and made available.
+additional reproducibility features (e.g. devcontainer) and
+documentation of key revisions and releases
 
@@
 
-Please note this manuscript is still in draft form and under active development.
-Changes to results, code, and the manuscript are likely and this **should not
-be cited without author permission**. We plan to provide a preprint to 
-[bioRxiv](https://www.biorxiv.org/) once internal NOAA reviews and final 
-revisions are completed.
+Please note this manuscript is still in peer review and not final.
+Changes to results, code, and the manuscript are still possible and, until,
+final publication the latest pre-print at bioRxiv should be cited.
 
@@
     generated by the analysis. It also has a rendered PDF,
-    [London_HauloutBehavior.pdf](https://github.com/jmlondon/berchukHaulout/blob/main/analysis/paper/London_HauloutBeahvior.pdf), suitable for reading (the code is replaced by figures
+    [London_HauloutBehavior.pdf](https://github.com/jmlondon/berchukHaulout/blob/main/analysis/paper/London_HauloutBehavior.pdf), suitable for reading (the code is replaced by figures
     and tables in this file)
@@
 We recognize that the species and ecosystems we studied are within the ancestral
-and present-day environs of the Inupiat and Yup'ik people who, through many
+and present-day environs of the Inpuiat and Yup'ik people who, through many
 uncredited contributions of traditional knowledge, provided early western
@@
 Heather Ziel. We also appreciate the commitment to science and safety by all
-officers and crew of the NOAA Ship Oscar Dyson, the NOAA Ship MacArthur II, and
-the RV Thomas G. Thompson.
+officers and crew of the NOAA ship _Oscar Dyson_, the NOAA ship _MacArthur II_, and
+the RV _Thomas G. Thompson_.
 
-Telemetry data from the Alaska Department of Fish and Game (ADFG) and the North
+Telemetry data from the Alaska Department of Fish and Game (ADF&G) and the North
 Slope Borough Department of Wildlife Management (NSB) were important
-contributions to the findings presented here. Both agencies collaborate closely
-with Alaska Native communities and the contributions of local hunters to the
-successful capture, deployment, and understanding of seal ecology are
-invaluable. Deployments in the western Bering Sea were done in collaboration
-with Russian colleagues and North Pacific Wildlife.
+contributions to the findings presented here. Deployments in the western Bering
+Sea were done in collaboration with Russian colleagues and North Pacific
+Wildlife.
+
+The findings and conclusions in the paper are those of the author(s) and do not
+necessarily represent the views of the National Marine Fisheries Service, NOAA.
+Any use of trade, product, or firm names does not imply an endorsement by the
+U.S. Government. Funding for this study was provided by the U.S. National
+Oceanic and Atmospheric Administration. The field work was conducted under the
+authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
+782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
+Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
+A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
+Fisheries Science Center Institutional Animal Care and Use Committee (IACUC). ).
+Funding to ADF&G for tagging seals was provided by the Bureau of Ocean Energy
+Management (No. M13PC0015 for work in Kotzebue, Alaska in 2009) and the Office
+of Naval Research (No. N00014-16-1-3019). ADF&G and NSB field work was covered by
+Research Permits Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF&G IACUC
+permits Nos. 06-16, 09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27,
+0027-2018-29, 0027-2019-041.
 

### 2024-03-29 analysis/paper/London_HauloutBehavior.Rmd

@@
     affiliation: 3
-  - name: Lori T. Quakenbush
+  - name: Lori Quakenbush
     affiliation: 2
@@
   magnitude of haul-out behavior varied among years. However, we did not find
-  broad evidence that haul-out timing was linked to annual sea ice extent. Our
+  broad evidence that haul-out timing was linked to annual sea-ice extent. Our
   analysis emphasizes the importance of accounting for seasonal and temporal
@@
 summer [@frost2018]. Ringed seals (_Phoca (pusa) hispida_) --- not included in this
-study --- haul out on sea-ice but primarily use snow lairs on sea-ice during
+study --- haul out on sea ice but primarily use snow lairs on sea ice during
 winter and spring.
@@
   "Northing", "Continuous", "latitude divided by the mean latitude across all locations (for bearded seals only)",
-  "Year", "Continuous", "For the set of models examining inter-annual variation in sea ice use, we fitted models with the addition of year by day-of-year interactions." 
+  "Year", "Continuous", "For the set of models examining inter-annual variation in sea-ice use, we fitted models with the addition of year by day-of-year interactions." 
 )
@@
 computational efficiency, we used generalized linear mixed pseudo-models
-(GLMPMs; [@verhoef2010a]) to model variation in haul-out behavior as a function
-of (1) covariate predictors, (2) temporally autocorrelated random effects, and
-(3) individual random effects representing heterogeneity in individual behavior.
-We used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored
-two different model formulations for our data and we fit separate models to
-bearded, ribbon, and spotted seal data sets as we expected differing behavior by
-species. Separate models for each species were also needed because a single,
-very large data set proved computationally intractable. In our first model
-formulation and for each species, we fitted a year-independent model that
-predicted average haul-out behavior as a function of demographic, weather,
-seasonal, and diurnal effects. Second, for ribbon and spotted seals (which had
-considerably more data than bearded seals), we fitted models that included all
-the effects from the first model, but also permitted annual variation in
-haul-out timing. This second set of models was used to examine whether
-haul-out patterns varied by year and to determine the annual timing of apparent
-peaks in haul-out behavior. For both models, we assumed an hourly Bernoulli
-response (a binary 0-1 response dependent upon whether the tag was greater than 
-50% dry for a given hour) where the linear predictor was 
-modeled on the logit scale. This is
+(GLMPMs; Ver Hoef, London & Boveng [-@verhoef2010a]) to model variation in
+haul-out behavior as a function of (1) covariate predictors, (2) temporally
+autocorrelated random effects, and (3) individual random effects representing
+heterogeneity in individual behavior. We used the glmmLDTS package
+[@verhoef2010a] to implement GLMPMs. We explored two different model
+formulations for our data and we fit separate models to bearded, ribbon, and
+spotted seal data sets as we expected differing behavior by species. Separate
+models for each species were also needed because a single, very large data set
+proved computationally intractable. In our first model formulation and for each
+species, we fitted a year-independent model that predicted average haul-out
+behavior as a function of demographic, weather, seasonal, and diurnal effects.
+Second, for ribbon and spotted seals (which had considerably more data than
+bearded seals), we fitted models that included all the effects from the first
+model, but also permitted annual variation in haul-out timing. This second set
+of models was used to examine whether haul-out patterns varied by year and to
+determine the annual timing of apparent peaks in haul-out behavior. For both
+models, we assumed an hourly Bernoulli response (a binary 0-1 response dependent
+upon whether the tag was greater than 50% dry for a given hour) where the linear
+predictor was modeled on the logit scale. This is
 consistent with previous approaches [@verhoef2014a; @london2012] and only
@@
 
-We also assessed whether the annual variation in maximum spring sea ice extent in the
+We also assessed whether the annual variation in maximum spring sea-ice extent in the
 Bering Sea influenced the seasonal peak of seal haul-out probability. In
-particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
+particular, we used sea-ice concentration data from the Nimbus-7 SMMR and DMSP
 SSM/I-SSMIS Passive Microwave Dataset, Version 1 [@cavalieri1996] to calculate
-maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
+maximum sea-ice extent. All sea-ice concentration grid cells (25 km^2^) in the
 study area with greater than 15% concentration were counted daily to get the total sea
 ice extent for each day between 15 February and 15 July across all years.
-Maximum spring sea ice extent was simply the largest daily count of grid cells
+Maximum spring sea-ice extent was simply the largest daily count of grid cells
 with greater than 15% concentration for each year. A separate regression model,
@@
 percent-dry data (100% dry for several weeks but indicative of movement and
-increasing transmission rates (see @boveng2013)). All data after the first
+increasing transmission rates (see Boveng & Cameron [-@boveng2013])). All data after the first
 instance of unrealistic values were censored from this analysis. In addition to
@@
 
-```{r maxextentHO, fig.height = 5.5, include=FALSE, fig.cap="Annual timing of peak haul-out behavior analyzed against the yearly maximum area of sea ice extent in the study area.
+```{r maxextentHO, fig.height = 5.5, include=FALSE, fig.cap="Annual timing of peak haul-out behavior analyzed against the yearly maximum area of sea-ice extent in the study area.
 While spotted seal adult females and subadults show a negative trend line, none of the relationships were determined to be significant."}
@@
         legend.justification = "right") +
-  labs(x = "date of peak haul-out", y = "maximum sea ice extent (km<sup>2</sup> / 1000)",
-       title = "The relationship between yearly maximum sea ice extent and the \nannual timing of peak haul-out probability") +
+  labs(x = "date of peak haul-out", y = "maximum sea-ice extent (km<sup>2</sup> / 1000)",
+       title = "The relationship between yearly maximum sea-ice extent and the \nannual timing of peak haul-out probability") +
   theme(axis.title.y = element_markdown())
@@
 device attached to the rear flipper of the seal which meant we only received
-locations when seals were hauled out on sea-ice, limiting our ability to fully
-evaluate fine-scale habitat preferences related to sea-ice. Insight into how
-seals use and interact with sea-ice during an extended period when the
-availability and characteristics of sea-ice is rapidly changing is important
+locations when seals were hauled out on sea ice, limiting our ability to fully
+evaluate fine-scale habitat preferences related to sea ice. Insight into how
+seals use and interact with sea ice during an extended period when the
+availability and characteristics of sea ice is rapidly changing is important
 [@breed2018; @cameron2018] but ancillary to the focus of this analysis -- and,
@@
 were the most parsimonious. For all of these reasons, we have elected not to use
-sea ice concentration _as a predictor for haul-out probability_ in the present
+sea-ice concentration _as a predictor for haul-out probability_ in the present
 study.
@@
 2018 – 2019, years of extreme low spring ice extent in the Bering Sea that
-appeared to impact body condition of ribbon and spotted seals [@boveng2020], so we may
-currently lack sufficient contrast in ice extent to characterize an effect on
-the timing of peak haul-out probability. We should expect, however, that some minimal threshold in
-the spatial extent or density of sea ice will have a meaningful impact on the
-timing of peak haul-out behavior --- if there is no sea-ice, seals will not haul
-out or they will be forced to use terrestrial haul-out sites which were not part of the
-evolution of their normal behaviors. Additionally, while from an
-ecological perspective the haul-out behavior appears consistent, the interannual
-differences in timing and magnitude are large enough to have important
-ramifications on calculations of abundance and trend. Those ramifications will
-only be exacerbated if climate variability amplifies interannual differences.
+appeared to impact body condition of ribbon and spotted seals [@boveng2020], so
+we may currently lack sufficient contrast in ice extent to characterize an
+effect on the timing of peak haul-out probability. We should expect, however,
+that some minimal threshold in the spatial extent or density of sea ice will
+have a meaningful impact on the timing of peak haul-out behavior --- if there is
+no sea-ice, seals will not haul out or they will be forced to use terrestrial
+haul-out sites which were not part of the evolution of their normal behaviors.
+Additionally, while from an ecological perspective the haul-out behavior appears
+consistent, the interannual differences in timing and magnitude are large enough
+to have important ramifications on calculations of abundance and trend. Those
+ramifications will only be exacerbated if climate variability amplifies
+interannual differences.
 

### 2024-04-05 analysis/paper/London_HauloutBehavior.Rmd
additional color blind friendly improvements

@@
 
-pal <- c(rcartocolor::carto_pal(12,'Bold')[1:11],
-         rcartocolor::carto_pal(12,'Antique')[1:3])
+# pal <- c(rcartocolor::carto_pal(12,'Bold')[1:11],
+#          rcartocolor::carto_pal(12,'Antique')[1:3])
+pal <- c("#A6CEE3","#3385BB","#84BF96","#6DBD57","#F3E587","#F57C7C","#E42622",
+         "#FBB268","#FE8D19","#DE9E83","#9D7BBA", "#977899","#7F9D55","#B15928")
 names(pal) <- as.character(sort(unique(peak_ho_df$year)))

### 2024-04-05 R/get_age_sex_colors.R

@@ -1,9 +1,9 @@
 get_age_sex_colors <- function() {
   age_sex_colors <- c(
-    "ADULT.F" = "#CC6677",
-    "SUBADULT" = "#117733",
-    "ADULT.M" = "#88CCEE",
-    "ALL AGES" = "#DDCC77"
+    "ADULT.F" = "#0072B2",
+    "SUBADULT" = "#009E73",
+    "ADULT.M" = "#D55E00",
+    "ALL AGES" = "#E69F00"
   )
   return(age_sex_colors)
 }

### 2024-05-29 analysis/paper/London_HauloutBehavior.Rmd
move permit details to methods section

@@
 events occurred between 2005 and 2020 from May to October, generally between the
-Alaska coastal communities of Scammon Bay in the Bering Sea, Utqiaġvik
-in the Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental
-Material, [S1](#s1)). Data from additional
-deployments along the Kamchatka peninsula in the western Bering Sea are also
-included. We refer readers to Figure \@ref(fig:deployplot) and the primary literature
-for detailed capture and bio-logger attachment methods (see also Supplemental
-Material, [S1](#s1)).
+Alaska coastal communities of Scammon Bay in the Bering Sea, Utqiaġvik in the
+Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental Material, [S1](#s1)).
+Data from additional deployments along the Kamchatka peninsula in the western
+Bering Sea are also included. We refer readers to Figure \@ref(fig:deployplot)
+and the primary literature for detailed capture and bio-logger attachment
+methods (see also Supplemental Material, [S1](#s1)). NOAA-led research was
+conducted under the authority of Marine Mammal Protection Act Research Permits
+Nos. 782-1676, 782-1765, 15126, and 19309 issued by the National Marine
+Fisheries Service, and Letters of Assurance of Compliance with Animal Welfare
+Act regulations, Nos. A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries
+Science Center/Northwest Fisheries Science Center Institutional Animal Care and
+Use Committee (IACUC). ADF&G and NSB field work was covered by Research Permits
+Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF&G IACUC permits Nos. 06-16,
+09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27, 0027-2018-29, 0027-2019-041.
 
@@
 
-```{r deployplot, include=TRUE, fig.asp = 0.4, fig.cap = "(ref:deployMapCaption)"}
+```{r deployplot, include=TRUE, fig.asp = 0.42, fig.cap = "(ref:deployMapCaption)"}
 targets::tar_load(deploy_details, store = here::here("_targets"))
@@
                     name = "species") + 
-  labs(caption = "<span style = 'color: #A26E7C;'><span style = 'font-family: \"Font Awesome 6 Free Regular\";'>&#xf0c8;</span> spatial extent of aerial survey effort in US and Russian waters</span>") +
+  labs(caption = "<span style = 'color: #A26E7C;'><span style = 'font-family: \"Font Awesome 6 Free Regular\";'>&#xf0c8;</span> spatial extent of aerial survey effort in US and Russian waters</span><br />world land vector data from naturalearthdata.com") +
   theme(legend.position = "none",
@@
   labs(
-    caption = "One hexagonal cell represents 50 km^2^"
+    caption = "One hexagonal cell represents 50 km^2^<br />world land vector data from naturalearthdata.com"
   ) +
@@
 U.S. Government. Funding for this study was provided by the U.S. National
-Oceanic and Atmospheric Administration. The field work was conducted under the
-authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
-782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
-Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
-A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
-Fisheries Science Center Institutional Animal Care and Use Committee (IACUC). ).
-Funding to ADF&G for tagging seals was provided by the Bureau of Ocean Energy
+Oceanic and Atmospheric Administration. Funding to ADF&G for tagging seals was provided by the Bureau of Ocean Energy
 Management (No. M13PC0015 for work in Kotzebue, Alaska in 2009) and the Office
-of Naval Research (No. N00014-16-1-3019). ADF&G and NSB field work was covered by
-Research Permits Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF&G IACUC
-permits Nos. 06-16, 09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27,
-0027-2018-29, 0027-2019-041.
+of Naval Research (No. N00014-16-1-3019).
 

### 2024-05-29 analysis/paper/London_HauloutBehavior.Rmd
reference pre-print with link
remove funding details form acknowledgements

@@
 
-This manuscript was developed as a reproducible research compendium. All data and
-code are available on GitHub (https://github.com/noaa-afsc/berchukseals-haulout) and
-major versions archived at Zenodo (https://doi.org/10.5281/zenodo.4638221).
-Original data sources for telemetry are archived
-at the United States Animal Telemetry Network, archived at Movebank, or
-associated with other published manuscripts (see supplemental 
-material \@ref(tab:deploymentTable)). Collated and cleaned data products
-needed to replicate the analysis along with the results of all model fits are 
-also available and versioned as an R package on 
-GitHub (https://github.com/noaa-afsc/berchuksealsHauloutFits) and archived at 
-Zenodo.
+This manuscript was developed as a reproducible research compendium and was
+originally published as a pre-print (https://doi.org/10.1101/2022.04.07.487572).
+All data and code are available on GitHub
+(https://github.com/noaa-afsc/berchukseals-haulout) and major versions archived
+at Zenodo (https://doi.org/10.5281/zenodo.4638221). Original data sources for
+telemetry are archived at the United States Animal Telemetry Network, archived
+at Movebank, or associated with other published manuscripts (see supplemental
+material \@ref(tab:deploymentTable)). Collated and cleaned data products needed
+to replicate the analysis along with the results of all model fits are also
+available and versioned as an R package on GitHub
+(https://github.com/noaa-afsc/berchuksealsHauloutFits) and archived at Zenodo.
 
@@
 Any use of trade, product, or firm names does not imply an endorsement by the
-U.S. Government. Funding for this study was provided by the U.S. National
-Oceanic and Atmospheric Administration. Funding to ADF&G for tagging seals was provided by the Bureau of Ocean Energy
-Management (No. M13PC0015 for work in Kotzebue, Alaska in 2009) and the Office
-of Naval Research (No. N00014-16-1-3019).
+U.S. Government.
 

### 2024-05-29 analysis/paper/London_HauloutBehavior.Rmd
remove early references to Fig 7 and Fig 9

@@
 haul-out probability, bearded seal haul-out behavior was also more protracted
-throughout the spring season compared to ribbon (Figure \@ref(fig:ribbonHOCal))
-and spotted seals (Figure \@ref(fig:spottedHOCal)). Overall, bearded seals were
+throughout the spring season compared to ribbon
+and spotted seals (see below). Overall, bearded seals were
 less likely to haul out and had a bi-modal distribution of haul-out probability

### 2024-05-29 analysis/paper/London_HauloutBehavior.Rmd
rename figures to Figure-##

@@
 
+make_plot_namer <- function(prefix="Figure-", fig_n=1L, width=2){
+  function(x){
+    x2 <- file.path(dirname(x), 
+                    paste0(prefix, 
+                           formatC(fig_n, width=width, flag="0"),
+                           ".", 
+                           tools::file_ext(x)))
+    fig_n <<- fig_n + 1L
+    if (file.rename(x, x2)) x2 else x
+  }
+}
+
+knitr::opts_chunk$set(fig.process = make_plot_namer(width=3))
+
 options(scipen = 3, digits = 3)

### 2024-05-31 analysis/paper/London_HauloutBehavior.Rmd
rebuild/re-run analysis; change figure names

@@
 
+options(tinytex.verbose = TRUE)
+
 options(scipen = 3, digits = 3)
@@
 young-of-the-year. For those bio-loggers deployed on young-of-the-year and
-transmitting into the next year, the age class was advanced to subadult on 1
-March of the following year -- the assumed anniversary of their birth. Subadults
-are those seals likely greater than one year of age but less than four years.
-Adults are individuals that are likely older than four years. Table
-\@ref(tab:tblDeploy) provides a summary of these deployments and data received
-from them.
+transmitting into the next year (6 ribbon seals; 3 spotted seals), the age class
+was advanced to subadult on 1 March of the following year -- the assumed
+anniversary of their birth. Subadults are those seals likely greater than one
+year of age but less than four years. Adults are individuals that are likely
+older than four years. Table \@ref(tab:tblDeploy) provides a summary of these
+deployments and data received from them.
 
@@
 This manuscript was developed as a reproducible research compendium and was
-originally published as a pre-print (https://doi.org/10.1101/2022.04.07.487572).
-All data and code are available on GitHub
-(https://github.com/noaa-afsc/berchukseals-haulout) and major versions archived
-at Zenodo (https://doi.org/10.5281/zenodo.4638221). Original data sources for
-telemetry are archived at the United States Animal Telemetry Network, archived
-at Movebank, or associated with other published manuscripts (see supplemental
-material \@ref(tab:deploymentTable)). Collated and cleaned data products needed
-to replicate the analysis along with the results of all model fits are also
+originally published as a pre-print at bioRxiv (@london2022;
+https://doi.org/10.1101/2022.04.07.487572). All data used in the study and code
+are available on GitHub (https://github.com/noaa-afsc/berchukseals-haulout) and
+major versions archived at Zenodo (https://doi.org/10.5281/zenodo.4638221).
+Original data sources for telemetry are archived as part of datasets at
+the United States Animal Telemetry Network (https://portal.atn.ioos.us/), archived at Movebank (see
+Movebank ID	732321226), or associated
+with other published manuscripts (see supplemental material
+\@ref(tab:deploymentTable)). Collated and cleaned data products needed to
+replicate the analysis along with the results of all model fits are also
 available and versioned as an R package on GitHub
-(https://github.com/noaa-afsc/berchuksealsHauloutFits) and archived at Zenodo.
+(https://github.com/noaa-afsc/berchuksealsHauloutFits) and archived at Zenodo
+(https://doi.org/10.5281/zenodo.10056308).
 

### 2024-05-31 analysis/paper/London_HauloutBehavior.Rmd

@@
   fig.path = "../figures/",
-  dpi = 600,
+  dpi = 500,
   fig.width = 6,

### 2024-05-31 analysis/paper/London_HauloutBehavior.tex

@@
 % Pandoc citation processing
+%From Pandoc 3.1.8
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
@@
 
-\hypertarget{introduction}{%
-\section*{Introduction}\label{introduction}}
+\section*{Introduction}\label{introduction}
 \addcontentsline{toc}{section}{Introduction}
@@
 
-\hypertarget{methods}{%
-\section*{Methods}\label{methods}}
+\section*{Methods}\label{methods}
 \addcontentsline{toc}{section}{Methods}
 
-\hypertarget{data-collection}{%
-\subsection*{Data collection}\label{data-collection}}
+\subsection*{Data collection}\label{data-collection}
 \addcontentsline{toc}{subsection}{Data collection}
@@
 events occurred between 2005 and 2020 from May to October, generally between the
-Alaska coastal communities of Scammon Bay in the Bering Sea, Utqiaġvik
-in the Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental
-Material, \protect\hyperlink{s1}{S1}). Data from additional
-deployments along the Kamchatka peninsula in the western Bering Sea are also
-included. We refer readers to Figure \ref{fig:deployplot} and the primary literature
-for detailed capture and bio-logger attachment methods (see also Supplemental
-Material, \protect\hyperlink{s1}{S1}).
+Alaska coastal communities of Scammon Bay in the Bering Sea, Utqiaġvik in the
+Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental Material, \hyperref[s1]{S1}).
+Data from additional deployments along the Kamchatka peninsula in the western
+Bering Sea are also included. We refer readers to Figure \ref{fig:deployplot}
+and the primary literature for detailed capture and bio-logger attachment
+methods (see also Supplemental Material, \hyperref[s1]{S1}). NOAA-led research was
+conducted under the authority of Marine Mammal Protection Act Research Permits
+Nos. 782-1676, 782-1765, 15126, and 19309 issued by the National Marine
+Fisheries Service, and Letters of Assurance of Compliance with Animal Welfare
+Act regulations, Nos. A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries
+Science Center/Northwest Fisheries Science Center Institutional Animal Care and
+Use Committee (IACUC). ADF\&G and NSB field work was covered by Research Permits
+Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF\&G IACUC permits Nos. 06-16,
+09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27, 0027-2018-29, 0027-2019-041.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/deployplot-1} \caption{\textbf{Initial bio-logger deployment areas for bearded, ribbon, and spotted seals in the study between 2005 and 2020 across multiple regions within the Bering, Chukchi, and western Beaufort seas.} \linebreak Solid regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility.
+\includegraphics[width=1\linewidth]{../figures/Figure-001} \caption{\textbf{Initial bio-logger deployment areas for bearded, ribbon, and spotted seals in the study between 2005 and 2020 across multiple regions within the Bering, Chukchi, and western Beaufort seas.} \linebreak Solid regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility.
 The larger, shaded region indicates the spatial extent of aerial survey effort to date in the Bering, Chukchi, and Beaufort seas.
@@
 2005 and
-2021. Bio-loggers were of the
+2022. Bio-loggers were of the
 `SPLASH' or `SPOT' family of tags developed by Wildlife Computers (Redmond,
@@
 young-of-the-year. For those bio-loggers deployed on young-of-the-year and
-transmitting into the next year, the age class was advanced to subadult on 1
-March of the following year -- the assumed anniversary of their birth. Subadults
-are those seals likely greater than one year of age but less than four years.
-Adults are individuals that are likely older than four years. Table
-\ref{tab:tblDeploy} provides a summary of these deployments and data received
-from them.
+transmitting into the next year (6 ribbon seals; 3 spotted seals), the age class
+was advanced to subadult on 1 March of the following year -- the assumed
+anniversary of their birth. Subadults are those seals likely greater than one
+year of age but less than four years. Adults are individuals that are likely
+older than four years. Table \ref{tab:tblDeploy} provides a summary of these
+deployments and data received from them.
 
@@
 
-\setlength{\tabcolsep}{0pt}
+\setlength{\tabcolsep}{2pt}
 
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.9in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Spotted\ seal}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{23\ (21,588\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{21\ (19,559\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{11\ (13,417\ seal\ hours)}}} \\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.9in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Spotted\ seal}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{23\ (21,970\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{21\ (19,559\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{11\ (13,417\ seal\ hours)}}} \\
 
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/examplePlot-1} \caption{\textbf{Example percent-dry actogram from bio-logger data}
+\includegraphics[width=1\linewidth]{../figures/Figure-002} \caption{\textbf{Example percent-dry actogram from bio-logger data}
 \linebreak Haul-out behavior observations recorded by a bio-logger deployed on a
@@
 
-\hypertarget{explanatory-variables}{%
-\subsection*{Explanatory variables}\label{explanatory-variables}}
+\subsection*{Explanatory variables}\label{explanatory-variables}
 \addcontentsline{toc}{subsection}{Explanatory variables}
@@
 
-\hypertarget{haul-out-modeling}{%
-\subsection*{Haul-out modeling}\label{haul-out-modeling}}
+\subsection*{Haul-out modeling}\label{haul-out-modeling}
 \addcontentsline{toc}{subsection}{Haul-out modeling}
@@
 consistent with previous approaches (London et al., 2012; Ver Hoef et al., 2014) and only
-6.995\% of our
+7.005\% of our
 observations fell between 10\% and 90\% hourly percent-dry.
@@
 
-\hypertarget{results}{%
-\section*{Results}\label{results}}
+\section*{Results}\label{results}
 \addcontentsline{toc}{section}{Results}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/dataMap-1} \caption{\textbf{Spatial distribution of haul-out records from 1 March through July 15 for bearded, ribbon, and spotted seal.} \linebreak Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model.
+\includegraphics[width=1\linewidth]{../figures/Figure-003} \caption{\textbf{Spatial distribution of haul-out records from 1 March through July 15 for bearded, ribbon, and spotted seal.} \linebreak Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model.
 For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km\textsuperscript{2}}\label{fig:dataMap}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/dataCal-1} \caption{\textbf{Seasonal distribution of haul-out behavior observations by species} \linebreak Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
+\includegraphics[width=1\linewidth]{../figures/Figure-004} \caption{\textbf{Seasonal distribution of haul-out behavior observations by species} \linebreak Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species.
 Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020.
@@
 
-\hypertarget{bearded-seals}{%
-\subsection*{Bearded Seals}\label{bearded-seals}}
+\subsection*{Bearded Seals}\label{bearded-seals}
 \addcontentsline{toc}{subsection}{Bearded Seals}
@@
 haul-out probability, bearded seal haul-out behavior was also more protracted
-throughout the spring season compared to ribbon (Figure \ref{fig:ribbonHOCal})
-and spotted seals (Figure \ref{fig:spottedHOCal}). Overall, bearded seals were
+throughout the spring season compared to ribbon
+and spotted seals (see below). Overall, bearded seals were
 less likely to haul out and had a bi-modal distribution of haul-out probability
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/beardedHOCal-1} \caption{\textbf{Bearded seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 15 March to 30 June for all age and sex classes combined.
+\includegraphics[width=1\linewidth]{../figures/Figure-005} \caption{\textbf{Bearded seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 15 March to 30 June for all age and sex classes combined.
 Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range.
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/beardedHOwx-1} \caption{\textbf{Influence of weather covariates on bearded seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
+\includegraphics[width=1\linewidth]{../figures/Figure-006} \caption{\textbf{Influence of weather covariates on bearded seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications.
 Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
@@
 
-\hypertarget{ribbon-seals}{%
-\subsection*{Ribbon Seals}\label{ribbon-seals}}
+\subsection*{Ribbon Seals}\label{ribbon-seals}
 \addcontentsline{toc}{subsection}{Ribbon Seals}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/ribbonHOCal-1} \caption{\textbf{Ribbon seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
+\includegraphics[width=1\linewidth]{../figures/Figure-007} \caption{\textbf{Ribbon seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
 Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/ribbonHOwx-1} \caption{\textbf{Influence of weather covariates on ribbon seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classifications.
+\includegraphics[width=1\linewidth]{../figures/Figure-008} \caption{\textbf{Influence of weather covariates on ribbon seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classifications.
 Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
@@
 
-\hypertarget{spotted-seals}{%
-\subsection*{Spotted Seals}\label{spotted-seals}}
+\subsection*{Spotted Seals}\label{spotted-seals}
 \addcontentsline{toc}{subsection}{Spotted Seals}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/spottedHOCal-1} \caption{\textbf{Spotted seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of spotted seals from 1 March to 30 June for each age and sex class used in the model.
+\includegraphics[width=1\linewidth]{../figures/Figure-009} \caption{\textbf{Spotted seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of spotted seals from 1 March to 30 June for each age and sex class used in the model.
 Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
@@
 Temperature
-(\(F_{1,114807}\)
-= 5.462; \(p\) = 0.019), wind
-(\(F_{1,114807}\)
-= 46.954; \(p\) = \textless0.001),
+(\(F_{1,115189}\)
+= 5.384; \(p\) = 0.020), wind
+(\(F_{1,115189}\)
+= 45.718; \(p\) = \textless0.001),
 and barometric pressure
-(\(F_{1,114807}\)
-= 10.214; \(p\) =
-0.001) were all significant. Spotted seals were more likely to be on the ice when
+(\(F_{1,115189}\)
+= 9.445; \(p\) =
+0.002) were all significant. Spotted seals were more likely to be on the ice when
 temperatures were relatively warm and less likely to
 haul out at higher wind speeds. Wind chill
-(\(F_{1,114807}\)
-= 0.559; \(p\) =
-0.455) and
+(\(F_{1,115189}\)
+= 0.72; \(p\) =
+0.396) and
 precipitation
-(\(F_{1,114807}\)
-= 0.763; \(p\) =
-0.382)
+(\(F_{1,115189}\)
+= 0.773; \(p\) =
+0.379)
 were not as influential as the other covariates.
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/spottedHOwx-1} \caption{\textbf{Influence of weather covariates on spotted seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
+\includegraphics[width=1\linewidth]{../figures/Figure-010} \caption{\textbf{Influence of weather covariates on spotted seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification.
 Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May.
@@
 
-\hypertarget{annual-variation-in-haul-out-timing}{%
-\subsection*{Annual variation in haul-out timing}\label{annual-variation-in-haul-out-timing}}
+\subsection*{Annual variation in haul-out timing}\label{annual-variation-in-haul-out-timing}
 \addcontentsline{toc}{subsection}{Annual variation in haul-out timing}
@@
 interactions between day and year for only spotted seals (day:year,
-\(F_{15,114762}\)
-= 4.345; \(p\) =
+\(F_{15,115144}\)
+= 4.445; \(p\) =
 \textless0.001; day\textsuperscript{2}:year,
-\(F_{15,114762}\)
-= 5.827; \(p\) =
+\(F_{15,115144}\)
+= 5.854; \(p\) =
 \textless0.001). Ribbon seals showed no significant
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/annualHO-1} \caption{\textbf{Annual variability in the timing of peak haul-out probability for ribbon and spotted seals.} \linebreak Spring haul-out probability (colored markers) for ribbon and spotted seals across 14 years are shown.
+\includegraphics[width=1\linewidth]{../figures/Figure-011} \caption{\textbf{Annual variability in the timing of peak haul-out probability for ribbon and spotted seals.} \linebreak Spring haul-out probability (colored markers) for ribbon and spotted seals across 14 years are shown.
 The full seasonal range of annual predictions along with uncertainty are presented as grey lines.
@@
 0.007, \(p\) = 0.828; spotted seals
-adult males: \(R^{2}\) = 0.003, \(p\) =
-0.903). Adult female and subadult spotted seals both
+adult males: \(R^{2}\) = 0.004, \(p\) =
+0.889). Adult female and subadult spotted seals both
 showed a negative trend (peak haul-out occurs later in years with less
@@
 (spotted seal adult female: \(R^{2}\) =
-0.663, \(p\) = 0.049; spotted seal
-subadults: \(R^{2}\) = 0.384, \(p\) =
-0.056).
-
-\hypertarget{discussion}{%
-\section*{Discussion}\label{discussion}}
+0.456, \(p\) = 0.141; spotted seal
+subadults: \(R^{2}\) = 0.369, \(p\) =
+0.062).
+
+\section*{Discussion}\label{discussion}
 \addcontentsline{toc}{section}{Discussion}
@@
 convective heat loss, both of which provides energetic benefits (see additional
-discussion in Supplemental Material \protect\hyperlink{exploring-insolation-solar-radiation-as-a-model-covariate}{Exploring Insolation (Solar Radiation) as a
+discussion in Supplemental Material \hyperref[exploring-insolation-solar-radiation-as-a-model-covariate]{Exploring Insolation (Solar Radiation) as a
 Model Covariate} regarding the potential use of solar radiation directly). Low
@@
 (95\% CI: 0.576 -- 0.738) for
-adult female ribbon seals, 0.742
-(95\% CI: 0.655 -- 0.813) for
-adult male spotted seals, and 0.663
-(95\% CI: 0.574 -- 0.742) for
+adult female ribbon seals, 0.74
+(95\% CI: 0.654 -- 0.811) for
+adult male spotted seals, and 0.66
+(95\% CI: 0.571 -- 0.739) for
 adult female spotted seals. Our estimates
@@
 
-\hypertarget{author-contributions}{%
-\section*{Author Contributions}\label{author-contributions}}
+\section*{Author Contributions}\label{author-contributions}
 \addcontentsline{toc}{section}{Author Contributions}
@@
 
-\hypertarget{data-availability}{%
-\section*{Data Availability}\label{data-availability}}
+\section*{Data Availability}\label{data-availability}
 \addcontentsline{toc}{section}{Data Availability}
 
-This manuscript was developed as a reproducible research compendium. All data and
-code are available on GitHub (\url{https://github.com/noaa-afsc/berchukseals-haulout}) and
+This manuscript was developed as a reproducible research compendium and was
+originally published as a pre-print at bioRxiv (London et al. (2022);
+\url{https://doi.org/10.1101/2022.04.07.487572}). All data used in the study and code
+are available on GitHub (\url{https://github.com/noaa-afsc/berchukseals-haulout}) and
 major versions archived at Zenodo (\url{https://doi.org/10.5281/zenodo.4638221}).
-Original data sources for telemetry are archived
-at the United States Animal Telemetry Network, archived at Movebank, or
-associated with other published manuscripts (see supplemental
-material \ref{tab:deploymentTable}). Collated and cleaned data products
-needed to replicate the analysis along with the results of all model fits are
-also available and versioned as an R package on
-GitHub (\url{https://github.com/noaa-afsc/berchuksealsHauloutFits}) and archived at
-Zenodo.
-
-\hypertarget{acknowledgments}{%
-\section*{Acknowledgments}\label{acknowledgments}}
+Original data sources for telemetry are archived as part of datasets at
+the United States Animal Telemetry Network (\url{https://portal.atn.ioos.us/}), archived at Movebank (see
+Movebank ID 732321226), or associated
+with other published manuscripts (see supplemental material
+\ref{tab:deploymentTable}). Collated and cleaned data products needed to
+replicate the analysis along with the results of all model fits are also
+available and versioned as an R package on GitHub
+(\url{https://github.com/noaa-afsc/berchuksealsHauloutFits}) and archived at Zenodo
+(\url{https://doi.org/10.5281/zenodo.10056308}).
+
+\section*{Acknowledgments}\label{acknowledgments}
 \addcontentsline{toc}{section}{Acknowledgments}
@@
 Any use of trade, product, or firm names does not imply an endorsement by the
-U.S. Government. Funding for this study was provided by the U.S. National
-Oceanic and Atmospheric Administration. The field work was conducted under the
-authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
-782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
-Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
-A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
-Fisheries Science Center Institutional Animal Care and Use Committee (IACUC). ).
-Funding to ADF\&G for tagging seals was provided by the Bureau of Ocean Energy
-Management (No.~M13PC0015 for work in Kotzebue, Alaska in 2009) and the Office
-of Naval Research (No.~N00014-16-1-3019). ADF\&G and NSB field work was covered by
-Research Permits Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF\&G IACUC
-permits Nos. 06-16, 09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27,
-0027-2018-29, 0027-2019-041.
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
+U.S. Government.
+
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-bengtson2004}{}}%
+\bibitem[\citeproctext]{ref-bengtson2004}
 Bengtson JL, Cameron MF.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bengtson2005a}{}}%
+\bibitem[\citeproctext]{ref-bengtson2005a}
 Bengtson JL, Hiruki-Raring LM, Simpkins MA, Boveng PL. 2005. Ringed and bearded seal densities in the eastern Chukchi Sea, 1999-2000. \emph{Polar Biology} 28:833--845.
 
-\leavevmode\vadjust pre{\hypertarget{ref-betts2006}{}}%
+\bibitem[\citeproctext]{ref-betts2006}
 Betts MG, Diamond A, Forbes G, Villard M-A, Gunn J.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-boveng2009}{}}%
+\bibitem[\citeproctext]{ref-boveng2009}
 Boveng PL, Bengtson JL, Buckley TW, Cameron MF, Dahle SP, Kelly BP, Megrey BA, Overland JE, Williamson NJ.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-boveng2013a}{}}%
+\bibitem[\citeproctext]{ref-boveng2013a}
 Boveng PL, Bengtson JL, Cameron MF, Dahle SP, Logerwell EA(ElizabethA), London JM, Overland JE, Sterling JT, Stevenson DE 1970-, Taylor BL 1954-, Ziel HL 1974-.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-boveng2013}{}}%
+\bibitem[\citeproctext]{ref-boveng2013}
 Boveng PL, Cameron MF.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-boveng2018a}{}}%
+\bibitem[\citeproctext]{ref-boveng2018a}
 Boveng P, Lowry L.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-boveng2020}{}}%
+\bibitem[\citeproctext]{ref-boveng2020}
 Boveng PL, Ziel HL, McClintock BT, Cameron MF.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-breed2018}{}}%
+\bibitem[\citeproctext]{ref-breed2018}
 Breed GA, Cameron MF, Ver Hoef JM, Boveng PL, Whiting A, Frost KJ.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bronson2009}{}}%
+\bibitem[\citeproctext]{ref-bronson2009}
 Bronson FH.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-cameron2010}{}}%
+\bibitem[\citeproctext]{ref-cameron2010}
 Cameron MF, Bengtson JL, Boveng PL, Jansen JK(JohnK, Kelly BP, Dahle SP, Logerwell EA(ElizabethA), Overland JE, Sabine CL, Waring GT, Wilder JM.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-cameron2018}{}}%
+\bibitem[\citeproctext]{ref-cameron2018}
 Cameron MF, Frost KJ, Ver Hoef JM, Breed GA, Whiting AV, Goodwin J, Boveng PL.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-cavalieri1996}{}}%
+\bibitem[\citeproctext]{ref-cavalieri1996}
 Cavalieri DJ, Parkinson CL, Gloersen P, Zwally H.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-conn2014a}{}}%
+\bibitem[\citeproctext]{ref-conn2014a}
 Conn PB, Hoef JMV, McClintock BT, Moreland EE, London JM, Cameron MF, Dahle SP, Boveng PL.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-conn2012c}{}}%
+\bibitem[\citeproctext]{ref-conn2012c}
 Conn PB, Johnson DS, London JM, Boveng PL.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-conn2023}{}}%
+\bibitem[\citeproctext]{ref-conn2023}
 Conn PB, Trukhanova IS.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-crawford2019}{}}%
+\bibitem[\citeproctext]{ref-crawford2019}
 Crawford JA, Frost KJ, Quakenbush LT, Whiting A.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-crawford2015}{}}%
+\bibitem[\citeproctext]{ref-crawford2015}
 Crawford JA, Quakenbush LT, Citta JJ.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-feltz1966}{}}%
+\bibitem[\citeproctext]{ref-feltz1966}
 Feltz ET, Fay FH. 1966. Thermal requirements in vitro of epidermal cells from seals. \emph{Cryobiology} 3:261--264.
 
-\leavevmode\vadjust pre{\hypertarget{ref-frost2018}{}}%
+\bibitem[\citeproctext]{ref-frost2018}
 Frost KJ, Burns JJ.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-frost2004}{}}%
+\bibitem[\citeproctext]{ref-frost2004}
 Frost KJ, Lowry LF, Pendleton G, Nute HR. 2004. Factors affecting the observed densities of ringed seals, \emph{Phoca hispida}, in the Alaskan Beaufort Sea, 1996-99. \emph{Arctic} 57:115--128.
 
-\leavevmode\vadjust pre{\hypertarget{ref-hamilton2018}{}}%
+\bibitem[\citeproctext]{ref-hamilton2018}
 Hamilton V, Evans K.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-harwood2020}{}}%
+\bibitem[\citeproctext]{ref-harwood2020}
 Harwood LA, Smith TG, Alikamik J, Alikamik E, Lea EV, Stirling I, Wright H, Melling H, Zhu X.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-huntington2020}{}}%
+\bibitem[\citeproctext]{ref-huntington2020}
 Huntington HP, Danielson SL, Wiese FK, Baker M, Boveng P, Citta JJ, De Robertis A, Dickson DMS, Farley E, George JC, Iken K, Kimmel DG, Kuletz K, Ladd C, Levine R, Quakenbush L, Stabeno P, Stafford KM, Stockwell D, Wilson C.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-johnson2008a}{}}%
+\bibitem[\citeproctext]{ref-johnson2008a}
 Johnson DS, London JM, Lea M-A, Durban JW. 2008. Continuous-time correlated random walk model for animal telemetry data. \emph{Ecology} 89:1208--1215.
 
-\leavevmode\vadjust pre{\hypertarget{ref-kelly2010}{}}%
+\bibitem[\citeproctext]{ref-kelly2010}
 Kelly BP, Bengtson JL, Boveng PL, Cameron MF, Dahle SP, Jansen JK, Logerwell EA, Overland JE, Sabine CL, Waring GT, Wilder JM.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kelly1990c}{}}%
+\bibitem[\citeproctext]{ref-kelly1990c}
 Kelly BP, Quakenbush LT. 1990. Spatiotemporal use of lairs by ringed seals (\emph{Phoca hispida}). \emph{Canadian Journal of Zoology} 68:2503--2512.
 
-\leavevmode\vadjust pre{\hypertarget{ref-kirkman2003}{}}%
+\bibitem[\citeproctext]{ref-kirkman2003}
 Kirkman SP, Bester MN, Pistorius PA, Hofmeyr GJG, Jonker FC, Owen R, Strydom N.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kwok2018}{}}%
+\bibitem[\citeproctext]{ref-kwok2018}
 Kwok R.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lindsay2021}{}}%
+\bibitem[\citeproctext]{ref-lindsay2021}
 Lindsay JM, Laidre KL, Conn PB, Moreland EE, Boveng PL.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-london2012}{}}%
+\bibitem[\citeproctext]{ref-london2022}
+London JM, Conn PB, Hardy SK, Richmond EL, Ver Hoef JM, Cameron MF, Crawford JA, Von Duyke AL, Quakenbush LT, Boveng PL.
+2022.
+Spring haul-out behavior of seals in the Bering and Chukchi seas.
+\emph{bioRxiv}:2022--04.
+DOI: \url{https://doi.org/10.1101/2022.04.07.487572}.
+
+\bibitem[\citeproctext]{ref-london2012}
 London JM, Ver Hoef JM, Jeffries SJ, Lance MM, Boveng PL.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lonergan2013}{}}%
+\bibitem[\citeproctext]{ref-lonergan2013}
 Lonergan M, Duck C, Moss S, Morris C, Thompson D.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lopez2013}{}}%
+\bibitem[\citeproctext]{ref-lopez2013}
 Lopez R, Malardé JP, Royer F, Gaspar P.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-mclaren1958a}{}}%
+\bibitem[\citeproctext]{ref-mclaren1958a}
 McLaren IA. 1958. Some aspects of growth and reproduction of the bearded seal, \emph{Erignathus barbartus} (Erxleben). \emph{Journal of the Fisheries Research Board of Canada} 15:219--227.
 
-\leavevmode\vadjust pre{\hypertarget{ref-meier2014}{}}%
+\bibitem[\citeproctext]{ref-meier2014}
 Meier WN, Hovelsrud GK, Oort BEH van, Key JR, Kovacs KM, Michel C, Haas C, Granskog MA, Gerland S, Perovich DK, Makshtas A, Reist JD.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-mesinger2006a}{}}%
+\bibitem[\citeproctext]{ref-mesinger2006a}
 Mesinger F, DiMego G, Kalnay E, Mitchell K, Shafran PC, Ebisuzaki W, Jović D, Woollen J, Rogers E, Berbery EH, Ek MB, Fan Y, Grumbine R, Higgins W, Li H, Lin Y, Manikin G, Parrish D, Shi W.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-niemi2023}{}}%
+\bibitem[\citeproctext]{ref-niemi2023}
 Niemi M, Nykänen M, Biard V, Kunnasranta M.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-olnes2020}{}}%
+\bibitem[\citeproctext]{ref-olnes2020}
 Olnes J, Crawford J, Citta JJ, Druckenmiller ML, Von Duyke AL, Quakenbush L.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-overland2021}{}}%
+\bibitem[\citeproctext]{ref-overland2021}
 Overland JE. 2021. Rare events in the Arctic. \emph{Climatic Change} 168:1--13.
 
-\leavevmode\vadjust pre{\hypertarget{ref-perpinan2012}{}}%
+\bibitem[\citeproctext]{ref-perpinan2012}
 Perpiñán O. 2012. \href{http://www.jstatsoft.org/v50/i09/}{solaR: Solar Radiation and Photovoltaic Systems with R}. \emph{Journal of Statistical Software} 50:1--32.
 
-\leavevmode\vadjust pre{\hypertarget{ref-perry2017}{}}%
+\bibitem[\citeproctext]{ref-perry2017}
 Perry EA, Stenson GB, Buren AD.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2021}{}}%
+\bibitem[\citeproctext]{ref-rcoreteam2021}
 R Core Team. 2021. \emph{\href{https://www.R-project.org/}{R: A language and environment for statistical computing}}. Vienna, Austria.
 
-\leavevmode\vadjust pre{\hypertarget{ref-reder2003}{}}%
+\bibitem[\citeproctext]{ref-reder2003}
 Reder S, Lydersen C, Arnold W, Kovacs KM. 2003. Haulout behaviour of High Arctic harbour seals (\emph{Phoca vitulina vitulina}) in Svalbard, Norway. \emph{Polar Biology} 27:6--16.
 
-\leavevmode\vadjust pre{\hypertarget{ref-southwell2008}{}}%
+\bibitem[\citeproctext]{ref-southwell2008}
 Southwell CJ, Paxton CGM, Borchers DL, Boveng PL, Nordøy ES, Blix AS, De La Mare WK.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-temte1994}{}}%
+\bibitem[\citeproctext]{ref-temte1994}
 Temte JL. 1994. Photoperiod control of birth timing in the harbour seal (\emph{Phoca vitulina}). \emph{J. Zool.} 233:369--384.
 
-\leavevmode\vadjust pre{\hypertarget{ref-teneyck2018}{}}%
+\bibitem[\citeproctext]{ref-teneyck2018}
 Ten Eyck P, Cavanaugh JE.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-thometz2021}{}}%
+\bibitem[\citeproctext]{ref-thometz2021}
 Thometz NM, Hermann-Sorensen H, Russell B, Rosen DAS, Reichmuth C.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-thompson1987}{}}%
+\bibitem[\citeproctext]{ref-thompson1987}
 Thompson P, Rothery P. 1987. Age and sex differences in the timing of moult in the common seal, \emph{Phoca vitulina}. \emph{Journal of Zoology} 212:597--603.
 
-\leavevmode\vadjust pre{\hypertarget{ref-udevitz2009}{}}%
+\bibitem[\citeproctext]{ref-udevitz2009}
 Udevitz MS, Jay CV, Fischback AS, Garlich-Miller JL.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-verhoef2014a}{}}%
+\bibitem[\citeproctext]{ref-verhoef2014a}
 Ver Hoef JM, Cameron MF, Boveng PL, London JM, Moreland EE.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-verhoef2010a}{}}%
+\bibitem[\citeproctext]{ref-verhoef2010a}
 Ver Hoef JM, London JM, Boveng PL.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-vonduyke2020}{}}%
+\bibitem[\citeproctext]{ref-vonduyke2020}
 Von Duyke AL, Douglas DC, Herreman JK, Crawford JA.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wang2018}{}}%
+\bibitem[\citeproctext]{ref-wang2018}
 Wang M, Yang Q, Overland JE, Stabeno P.
@@
 
-\hypertarget{supplemental-material}{%
-\section*{Supplemental Material}\label{supplemental-material}}
+\section*{Supplemental Material}\label{supplemental-material}
 \addcontentsline{toc}{section}{Supplemental Material}
 
-\hypertarget{additional-bio-logger-deployment-details}{%
-\subsection{Additional Bio-logger Deployment Details}\label{additional-bio-logger-deployment-details}}
+\subsection{Additional Bio-logger Deployment Details}\label{additional-bio-logger-deployment-details}
 
@@
 
-\endfoot
+\endlastfoot
 
@@
 
-\hypertarget{supplemental-figures-showing-confidence-intervals-associated-with-predictions}{%
-\subsection{Supplemental Figures Showing Confidence Intervals Associated with Predictions}\label{supplemental-figures-showing-confidence-intervals-associated-with-predictions}}
+\subsection{Supplemental Figures Showing Confidence Intervals Associated with Predictions}\label{supplemental-figures-showing-confidence-intervals-associated-with-predictions}
 
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/beardedPredSE-1} \caption{\textbf{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for bearded seals.} \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
+\includegraphics[width=1\linewidth]{../figures/Figure-013} \caption{\textbf{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for bearded seals.} \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
 Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/ribbonPredSE-1} \caption{\textbf{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals.} \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
+\includegraphics[width=1\linewidth]{../figures/Figure-014} \caption{\textbf{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals.} \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
 Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/spottedPredSE-1} \caption{\textbf{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for spotted seals.} \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
+\includegraphics[width=1\linewidth]{../figures/Figure-015} \caption{\textbf{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for spotted seals.} \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
 Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
@@
 
-\hypertarget{exploring-insolation-solar-radiation-as-a-model-covariate}{%
-\subsection{Exploring Insolation (Solar Radiation) as a Model Covariate}\label{exploring-insolation-solar-radiation-as-a-model-covariate}}
-
-\hypertarget{introduction-1}{%
-\subsubsection{Introduction}\label{introduction-1}}
+\subsection{Exploring Insolation (Solar Radiation) as a Model Covariate}\label{exploring-insolation-solar-radiation-as-a-model-covariate}
+
+\subsubsection{Introduction}\label{introduction-1}
 
@@
 
-\hypertarget{methods-1}{%
-\subsubsection{Methods}\label{methods-1}}
+\subsubsection{Methods}\label{methods-1}
 
@@
 
-\hypertarget{results-1}{%
-\subsubsection{Results}\label{results-1}}
+\subsubsection{Results}\label{results-1}
 
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/era5-ssrd-plot-1} \caption{\textbf{Diel Pattern of Solar Radiation Values from ERA5 Reanalysis.} \linebreak Downward surface solar radiation estimates from the ERA5 climate reanalysis for 5000 random points within the study area between 2005 and 2021.
+\includegraphics[width=1\linewidth]{../figures/Figure-016} \caption{\textbf{Diel Pattern of Solar Radiation Values from ERA5 Reanalysis.} \linebreak Downward surface solar radiation estimates from the ERA5 climate reanalysis for 5000 random points within the study area between 2005 and 2021.
 Solar radiation values are presented in Watts per square-meter and the smoothed line highlights the strong diel pattern.}\label{fig:era5-ssrd-plot}
@@
 -7428.929. The model with solar radiation resulted in
-39.829 degrees of freedom and an AIC value of
--6874.298. The ribbon seal model matching the specification from the manuscript resulted in
+39.619 degrees of freedom and an AIC value of
+-6797.378. The ribbon seal model matching the specification from the manuscript resulted in
 131.478 degrees of freedom and an AIC value of
 -16372.29. The model with solar radiation resulted in
-114.72 degrees of freedom and an AIC value of
--15223.346. The spotted seal model matching the specification from the manuscript resulted in
-125.54 degrees of freedom and an AIC value of
--23482.051. The model with solar radiation resulted in
-108.985 degrees of freedom and an AIC value of
--22917.167. Despite the additional terms, the
+115.126 degrees of freedom and an AIC value of
+-16038.175. The spotted seal model matching the specification from the manuscript resulted in
+125.506 degrees of freedom and an AIC value of
+-23584.373. The model with solar radiation resulted in
+109.163 degrees of freedom and an AIC value of
+-23302.772. Despite the additional terms, the
 models with the Fourier series representation of hour of day resulted in a
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/ssrd-prediction-heatmap-1} \caption{\textbf{Solar radiation as a predictor of ribbon seal haul-out probability.} \linebreak Predicted haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
+\includegraphics[width=1\linewidth]{../figures/Figure-017} \caption{\textbf{Solar radiation as a predictor of ribbon seal haul-out probability.} \linebreak Predicted haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
 In this model, solar radiation was used in lieu of hour of day.
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{../figures/ssrd-ribbonPredSE-1} \caption{\textbf{Solar radiation as a predictor of ribbon seal haul-out probability (with uncertainty).} \linebreak Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals.
+\includegraphics[width=1\linewidth]{../figures/Figure-018} \caption{\textbf{Solar radiation as a predictor of ribbon seal haul-out probability (with uncertainty).} \linebreak Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals.
 In this model predictions are shown for low, medium, and high values of solar radiation (as percentages of the maximum value observed) in lieu of local solar hour.
@@
 
-\hypertarget{discussion-1}{%
-\subsubsection{Discussion}\label{discussion-1}}
+\subsubsection{Discussion}\label{discussion-1}
 

### 2024-08-17 analysis/paper/Letter_Response_to_Reviewers.qmd
Squashed commit of the following:
commit cb17749e30b8733a2c326510d023904d5c2c2f35
Author: Josh M. London <josh.london@noaa.gov>
Date:   Sat Aug 17 10:37:44 2024 -0700

    Squashed commit of the following:

    commit 23f199555d6fdbd4b7d2b932ea2f7724a8ccdab6
    Author: Josh M. London <josh.london@noaa.gov>
    Date:   Sat Aug 17 10:35:18 2024 -0700

        ignore _targets.zip

commit 33f7f1920c55847a5739b37968e3ed3b9ae3f153
Author: Josh M. London <josh.london@noaa.gov>
Date:   Fri Jun 7 10:57:22 2024 -0700

    additional changes requested from staff

    - added additional link/url to the f

--- 2024-03-28 analysis/paper/Letter_Response_to_Reviewers.qmd
@@
 
+```{=tex}
 \newenvironment{author-response}{\color{Bittersweet}}
 
-Josh M. London, PhD\newline
+J
+```
+osh M. London, PhD\newline
 7600 Sand Point Way NE\newline Seattle, Washington,
@@
 addition to this letter, the full suite of updates are all contained and
-documented within a single pull request. We have also provided a pdf document
-that indicates all changes between the revised manuscript and the submitted
-manuscript.
+documented within two pull requests (see
+<https://github.com/noaa-afsc/berchukseals-haulout/pull/51> (initial
+resubmission) and <https://github.com/noaa-afsc/berchukseals-haulout/pull/52>
+(additional suggestions from PeerJ staff). We have also provided a pdf document
+(*\[Compare Report\] London_HauloutBehavior.pdf\]*) that indicates all changes
+between the compendium supporting this revised manuscript and the compendium
+supporting the originally submitted manuscript.
 

### 2024-08-17 analysis/paper/London_HauloutBehavior.Rmd
Squashed commit of the following:
commit cb17749e30b8733a2c326510d023904d5c2c2f35
Author: Josh M. London <josh.london@noaa.gov>
Date:   Sat Aug 17 10:37:44 2024 -0700

    Squashed commit of the following:

    commit 23f199555d6fdbd4b7d2b932ea2f7724a8ccdab6
    Author: Josh M. London <josh.london@noaa.gov>
    Date:   Sat Aug 17 10:35:18 2024 -0700

        ignore _targets.zip

commit 33f7f1920c55847a5739b37968e3ed3b9ae3f153
Author: Josh M. London <josh.london@noaa.gov>
Date:   Fri Jun 7 10:57:22 2024 -0700

    additional changes requested from staff

    - added additional link/url to the f

@@
 See Figure \@ref(fig:dataMap) for the spatial distribution of observed data used in the study and Supplemental Materials \@ref(tab:deploymentTable) for additional deployment details.
+World land vector data from https://naturalearthdata.com.
 
@@
 (ref:mapCaption) **Spatial distribution of haul-out records from 1 March through July 15 for bearded, ribbon, and spotted seal.** \linebreak Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model.
-For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km^2^
+For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km^2^.
+World land vector data from https://naturalearthdata.com.
 
@@
 
-# pal <- c(rcartocolor::carto_pal(12,'Bold')[1:11],
-#          rcartocolor::carto_pal(12,'Antique')[1:3])
-pal <- c("#A6CEE3","#3385BB","#84BF96","#6DBD57","#F3E587","#F57C7C","#E42622",
+# see https://davidmathlogic.com/colorblind/#%23A6CEE3-%233385BB-%2384BF96-%23004D40-%23F3E587-%23F57C7C-%23D81B60-%23FBB268-%23FE8D19-%23DE9E83-%239D7BBA-%23977899-%237F9D55-%23B15928
+# for simulation of colorblind scenarios
+
+pal <- c("#A6CEE3","#3385BB","#84BF96","#004D40","#F3E587","#F57C7C","#D81B60",
          "#FBB268","#FE8D19","#DE9E83","#9D7BBA", "#977899","#7F9D55","#B15928")
@@
 Original data sources for telemetry are archived as part of datasets at
-the United States Animal Telemetry Network (https://portal.atn.ioos.us/), archived at Movebank (see
+the United States Animal Telemetry Network (https://portal.atn.ioos.us/; 
+https://doi.org/10.24431/rw1k8er), archived at Movebank (see
 Movebank ID	732321226), or associated

### 2024-08-17 analysis/paper/London_HauloutBehavior.tex
Squashed commit of the following:
commit cb17749e30b8733a2c326510d023904d5c2c2f35
Author: Josh M. London <josh.london@noaa.gov>
Date:   Sat Aug 17 10:37:44 2024 -0700

    Squashed commit of the following:

    commit 23f199555d6fdbd4b7d2b932ea2f7724a8ccdab6
    Author: Josh M. London <josh.london@noaa.gov>
    Date:   Sat Aug 17 10:35:18 2024 -0700

        ignore _targets.zip

commit 33f7f1920c55847a5739b37968e3ed3b9ae3f153
Author: Josh M. London <josh.london@noaa.gov>
Date:   Fri Jun 7 10:57:22 2024 -0700

    additional changes requested from staff

    - added additional link/url to the f

@@
 Deployments were initiated across a range of months but only haul-out data from 1 March to 15 July were included in the analysis.
-See Figure \ref{fig:dataMap} for the spatial distribution of observed data used in the study and Supplemental Materials \ref{tab:deploymentTable} for additional deployment details.}\label{fig:deployplot}
+See Figure \ref{fig:dataMap} for the spatial distribution of observed data used in the study and Supplemental Materials \ref{tab:deploymentTable} for additional deployment details.
+World land vector data from \url{https://naturalearthdata.com}.}\label{fig:deployplot}
 \end{figure}
@@
 \includegraphics[width=1\linewidth]{../figures/Figure-003} \caption{\textbf{Spatial distribution of haul-out records from 1 March through July 15 for bearded, ribbon, and spotted seal.} \linebreak Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model.
-For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km\textsuperscript{2}}\label{fig:dataMap}
+For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km\textsuperscript{2}.
+World land vector data from \url{https://naturalearthdata.com}.}\label{fig:dataMap}
 \end{figure}
@@
 Original data sources for telemetry are archived as part of datasets at
-the United States Animal Telemetry Network (\url{https://portal.atn.ioos.us/}), archived at Movebank (see
+the United States Animal Telemetry Network (\url{https://portal.atn.ioos.us/};
+\url{https://doi.org/10.24431/rw1k8er}), archived at Movebank (see
 Movebank ID 732321226), or associated

### 2024-08-17 analysis/paper/London_HauloutBehavior.Rmd
typo fix Ribon to Ribbon

@@
                           begin = 0.1, end = 1,
-                       name = "Ribon seal haul-out probability", 
+                       name = "Ribbon seal haul-out probability", 
                        limits = c(0,1), breaks = c(0.25, 0.50, 0.75),

### 2024-08-17 analysis/paper/London_HauloutBehavior.Rmd
change title per reviewer comment
M Carter suggested this title to better capture the paper's focus on the implications of abundance estimation
reword abstract per M Carter suggestions

@@
 ---
-title: Spring haul-out behavior of seals in the Bering and Chukchi seas
+title: Spring haul-out behaviour of seals in the Bering and Chukchi Seas: implications for abundance estimation
 preprint: false
@@
   is therefore of concern as the quantity and quality of suitable habitat is
-  forecast to decrease. Improved estimates of seal population abundance are
-  needed to properly monitor the impacts of these changes over time. In this
-  paper, we use hourly percent-dry data from satellite-linked bio-loggers
-  deployed between 2005 and 2021 to quantify the proportion of seals hauled out
-  on ice. This information is needed to accurately estimate abundance from
-  aerial survey counts of ice-associated seals (i.e., to correct for the
-  proportion of animals that are in the water while surveys are conducted). In
-  addition to providing essential data for survey 'availability' calculations,
-  our analysis also provides insights into the seasonal timing and environmental
-  factors affecting haul-out behavior by ice-associated seals. We specifically
-  focused on bearded (_Erignathus barbatus_), ribbon (_Histriophoca fasciata_),
-  and spotted seals (_Phoca largha_) in the Bering and Chukchi seas. Because
-  ringed seals (_Phoca (pusa) hispida_) can be out of the water but hidden from view in
-  snow lairs analysis of their 'availability' to surveys requires special
-  consideration; therefore,they were not included in this analysis. Using
-  generalized linear mixed pseudo-models to properly account for temporal
-  autocorrelation, we fit models with covariates of interest (e.g., day-of-year,
-  solar hour, age-sex class, wind speed, barometric pressure, temperature,
-  precipitation) to examine their ability to explain variation in haul-out
-  probability. We found evidence for strong diel and within-season patterns in
-  haul-out behavior, as well as strong weather effects (particularly wind and
-  temperature). In general, seals were more likely to haul out on ice in the
-  middle of the day and when wind speed was low and temperatures were higher.
-  Haul-out probability increased through March and April, peaking in May and
-  early June before declining again. The timing and frequency of haul-out events
-  also varied based on species and age-sex class. For ribbon and spotted seals,
-  models with year effects were highly supported, indicating that the timing and
-  magnitude of haul-out behavior varied among years. However, we did not find
-  broad evidence that haul-out timing was linked to annual sea-ice extent. Our
-  analysis emphasizes the importance of accounting for seasonal and temporal
-  variation in haul-out behavior, as well as associated environmental
-  covariates, when interpreting the number of seals counted in aerial surveys.
+  forecast to decrease. Robust estimates of seal population abundance are needed
+  to properly monitor the impacts of these changes over time. Aerial surveys of
+  seals on ice are an efficient method for counting seals but must be paired
+  with estimates of the proportion of seals out of the water to derive
+  population abundance. In this paper, we use hourly percent-dry data from
+  satellite-linked bio-loggers deployed between 2005 and 2021 to quantify the
+  proportion of seals hauled out on ice. This information is needed to
+  accurately estimate abundance from aerial survey counts of ice-associated
+  seals (i.e., to correct for the proportion of animals that are in the water
+  while surveys are conducted). In addition to providing essential data for
+  survey 'availability' calculations, our analysis also provides insights into
+  the seasonal timing and environmental factors affecting haul-out behavior by
+  ice-associated seals. We specifically focused on bearded (_Erignathus
+  barbatus_), ribbon (_Histriophoca fasciata_), and spotted seals (_Phoca
+  largha_) in the Bering and Chukchi seas. Because ringed seals (_Phoca (pusa)
+  hispida_) can be out of the water but hidden from view in snow lairs analysis
+  of their 'availability' to surveys requires special consideration;
+  therefore,they were not included in this analysis. Using generalized linear
+  mixed pseudo-models to properly account for temporal autocorrelation, we fit
+  models with covariates of interest (e.g., day-of-year, solar hour, age-sex
+  class, wind speed, barometric pressure, temperature, precipitation) to examine
+  their ability to explain variation in haul-out probability. We found evidence
+  for strong diel and within-season patterns in haul-out behavior, as well as
+  strong weather effects (particularly wind and temperature). In general, seals
+  were more likely to haul out on ice in the middle of the day and when wind
+  speed was low and temperatures were higher. Haul-out probability increased
+  through March and April, peaking in May and early June before declining again.
+  The timing and frequency of haul-out events also varied based on species and
+  age-sex class. For ribbon and spotted seals, models with year effects were
+  highly supported, indicating that the timing and magnitude of haul-out
+  behavior varied among years. However, we did not find broad evidence that
+  haul-out timing was linked to annual sea-ice extent. Our analysis emphasizes
+  the importance of accounting for seasonal and temporal variation in haul-out
+  behavior, as well as associated environmental covariates, when interpreting
+  the number of seals counted in aerial surveys.
 output:

### 2024-08-17 analysis/paper/London_HauloutBehavior.Rmd
minor suggestions from M Carter

@@
 United States Endangered Species Act (ESA)) for timely estimates of population
-abundance and trend mean improved aerial survey effort is needed for these
+abundance and trends mean improved aerial survey effort is needed for these
 species. Those survey efforts must also be paired with improved knowledge of
@@
 of a subset of the study individuals. These devices are designed to collect data
-through the molt period (when those adhered to the hair would fall off) and, in
-some situations, provide multiple years of data.
-
-In this study, we used 16 years of bio-logging data to investigate the haul-out
+through the molt period (when those adhered to the hair -- a more conventional
+method -- would fall off) and, in some situations, provide multiple years of
+data.
+
+In this study, we used data collected from multiple bio-logging deployments
+spanning a 16-year period to investigate the haul-out
 behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
@@
 nets and bio-loggers were attached during studies based in coastal communities
-or on research ships. Ship-based capture events occurred during spring near the
-southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
-events occurred between 2005 and 2020 from May to October, generally between the
-Alaska coastal communities of Scammon Bay in the Bering Sea, Utqiaġvik in the
-Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental Material, [S1](#s1)).
-Data from additional deployments along the Kamchatka peninsula in the western
-Bering Sea are also included. We refer readers to Figure \@ref(fig:deployplot)
-and the primary literature for detailed capture and bio-logger attachment
-methods (see also Supplemental Material, [S1](#s1)). NOAA-led research was
-conducted under the authority of Marine Mammal Protection Act Research Permits
-Nos. 782-1676, 782-1765, 15126, and 19309 issued by the National Marine
-Fisheries Service, and Letters of Assurance of Compliance with Animal Welfare
-Act regulations, Nos. A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries
-Science Center/Northwest Fisheries Science Center Institutional Animal Care and
-Use Committee (IACUC). ADF&G and NSB field work was covered by Research Permits
-Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF&G IACUC permits Nos. 06-16,
-09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27, 0027-2018-29, 0027-2019-041.
+or on research ships (Figure \@ref(fig:deployplot)). Ship-based capture events
+occurred during spring near the southern ice edge in the Bering Sea between 2005
+and 2018. Land-based capture events occurred between 2005 and 2020 from May to
+October, generally between the Alaska coastal communities of Scammon Bay in the
+Bering Sea, Utqiaġvik in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. Data
+from additional deployments along the Kamchatka peninsula in the western Bering
+Sea are also included. We refer readers to the primary literature for detailed
+capture and bio-logger attachment methods (see publications listed in
+Supplemental Material, [S1](#s1)). NOAA-led research was conducted under the
+authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
+782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
+Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
+A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
+Fisheries Science Center Institutional Animal Care and Use Committee (IACUC).
+ADF&G and NSB field work was covered by Research Permits Nos. 358-1585,
+358-1787, 15324, and 20466 and by ADF&G IACUC permits Nos. 06-16, 09-21,
+2014-03, 2015-25, 2016-23, 0027-2017-27, 0027-2018-29, 0027-2019-041.
 

### 2024-08-19 analysis/paper/London_HauloutBehavior.Rmd
final edits in response to M Carter review

@@
 weaning and the model predictions seemed to indicate development of in-water
-activities (e.g. swimming, foraging) in May. Adult females had a more protracted
+activities (e.g. swimming, foraging) in May and, like adults, haul-out behavior
+was centered around solar noon. Adult females had a more protracted
 haul-out season compared to males, and more time was spent hauled out in June
@@
 model predictions reflected development of in-water activities (e.g.
-swimming, foraging) in May.
+swimming, foraging) in May and haul-out behavior centered around solar noon.
 
@@
 under different constraints than ribbon and spotted seals. Bearded and ringed
-seals are distributed across higher latitudes higher latitudes that experience
+seals are distributed across higher latitudes that experience
 extended daylight hours during spring which may allow more flexibility in
@@
 other phocids where yearlings and subadults molt first followed by adult females
-and males [@reder2003; @thompson1987; @kirkman2003].
+and males [@reder2003; @thompson1987; @kirkman2003]. Also of note is the early
+development in newly weaned pups of haul-out behavior centered around solar noon
+observed in this study.
 
@@
 conditions are general indicators of increased solar radiation and lower
-convective heat loss, both of which provides energetic benefits (see additional
+convective heat loss, both of which provide energetic benefits (see additional
 discussion in Supplemental Material [Exploring Insolation (Solar Radiation) as a
@@
 shown the potential for considerable bias in abundance estimates if such
-covariates are unaccounted for [@conn2023]. We recommend that future abundance
-analyses employ availability models that account for them. For instance, it is
-relatively straightforward to obtain weather reanalysis products for times and
-locations that are surveyed and to construct a relevant correction factor based
-on predictions of GLMPMs. The most challenging element in developing
-availability correction factors is with annual variability. It can be difficult
-to get a sufficient sample size to estimate year-specific correction factors,
-particularly because research teams would likely need to deploy bio-loggers on
-seals and conduct aerial surveys concurrently, requiring considerably more
-personnel and money. One possible suggestion is to include year as a random
-effect within models for aerial survey counts such that, without specific
-knowledge of any particular year, the among-year variance is included in the
-modeled standard errors. Regardless of the specific approach, future estimates
-of Arctic seal abundance will require specific consideration of annual
-variability and changes in the timing of peak haul-out behavior when estimating
-trends, as one will not know if moderate differences in abundance estimates are
-attributable to changes in abundance or changes in haul-out behavior.
+covariates are unaccounted for (see @conn2023 for discussion about the
+importance of including stable age- and stage-proportions). We recommend that
+future abundance analyses employ availability models that account for them. For
+instance, it is relatively straightforward to obtain weather reanalysis products
+(e.g. NARR, ERA5) for times and locations that are surveyed and to construct a
+relevant correction factor based on predictions of GLMPMs. The most challenging
+element in developing availability correction factors is with annual
+variability. It can be difficult to get a sufficient sample size to estimate
+year-specific correction factors, particularly because research teams would
+likely need to deploy bio-loggers on seals and conduct aerial surveys
+concurrently, requiring considerably more personnel and money. One possible
+suggestion is to include year as a random effect within models for aerial survey
+counts such that, without specific knowledge of any particular year, the
+among-year variance is included in the modeled standard errors. Regardless of
+the specific approach, future estimates of Arctic seal abundance will require
+specific consideration of annual variability and changes in the timing of peak
+haul-out behavior when estimating trends, as one will not know if moderate
+differences in abundance estimates are attributable to changes in abundance or
+changes in haul-out behavior.
 

### 2024-08-19 analysis/paper/London_HauloutBehavior.tex
final render with updated revisions, round 2

@@
 
-\title{Spring haul-out behavior of seals in the Bering and Chukchi seas}
+\title{Spring haul-out behavior of seals in the Bering and Chukchi Seas: implications for abundance estimation}
 
@@
 Rapid acceleration of climate change in Arctic ecosystems is therefore of concern as the quantity and quality of suitable habitat is forecast to decrease.
-Improved estimates of seal population abundance are needed to properly monitor the impacts of these changes over time.
+Robust estimates of seal population abundance are needed to properly monitor the impacts of these changes over time.
+Aerial surveys of seals on ice are an efficient method for counting seals but must be paired with estimates of the proportion of seals out of the water to derive population abundance.
 In this paper, we use hourly percent-dry data from satellite-linked bio-loggers deployed between 2005 and 2021 to quantify the proportion of seals hauled out on ice.
@@
 nets and bio-loggers were attached during studies based in coastal communities
-or on research ships. Ship-based capture events occurred during spring near the
-southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
-events occurred between 2005 and 2020 from May to October, generally between the
-Alaska coastal communities of Scammon Bay in the Bering Sea, Utqiaġvik in the
-Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental Material, \hyperref[s1]{S1}).
-Data from additional deployments along the Kamchatka peninsula in the western
-Bering Sea are also included. We refer readers to Figure \ref{fig:deployplot}
-and the primary literature for detailed capture and bio-logger attachment
-methods (see also Supplemental Material, \hyperref[s1]{S1}). NOAA-led research was
-conducted under the authority of Marine Mammal Protection Act Research Permits
-Nos. 782-1676, 782-1765, 15126, and 19309 issued by the National Marine
-Fisheries Service, and Letters of Assurance of Compliance with Animal Welfare
-Act regulations, Nos. A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries
-Science Center/Northwest Fisheries Science Center Institutional Animal Care and
-Use Committee (IACUC). ADF\&G and NSB field work was covered by Research Permits
-Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF\&G IACUC permits Nos. 06-16,
-09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27, 0027-2018-29, 0027-2019-041.
+or on research ships (Figure \ref{fig:deployplot}). Ship-based capture events
+occurred during spring near the southern ice edge in the Bering Sea between 2005
+and 2018. Land-based capture events occurred between 2005 and 2020 from May to
+October, generally between the Alaska coastal communities of Scammon Bay in the
+Bering Sea, Utqiaġvik in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. Data
+from additional deployments along the Kamchatka peninsula in the western Bering
+Sea are also included. We refer readers to the primary literature for detailed
+capture and bio-logger attachment methods (see publications listed in
+Supplemental Material, \hyperref[s1]{S1}). NOAA-led research was conducted under the
+authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
+782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
+Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
+A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
+Fisheries Science Center Institutional Animal Care and Use Committee (IACUC).
+ADF\&G and NSB field work was covered by Research Permits Nos. 358-1585,
+358-1787, 15324, and 20466 and by ADF\&G IACUC permits Nos. 06-16, 09-21,
+2014-03, 2015-25, 2016-23, 0027-2017-27, 0027-2018-29, 0027-2019-041.
 
@@
 weaning and the model predictions seemed to indicate development of in-water
-activities (e.g.~swimming, foraging) in May. Adult females had a more protracted
+activities (e.g.~swimming, foraging) in May and, like adults, haul-out behavior
+was centered around solar noon. Adult females had a more protracted
 haul-out season compared to males, and more time was spent hauled out in June
@@
 other phocids where yearlings and subadults molt first followed by adult females
-and males (Thompson \& Rothery, 1987; Kirkman et al., 2003; Reder et al., 2003).
+and males (Thompson \& Rothery, 1987; Kirkman et al., 2003; Reder et al., 2003). Also of note is the early
+development in newly weaned pups of haul-out behavior centered around solar noon
+observed in this study.
 
@@
 conditions are general indicators of increased solar radiation and lower
-convective heat loss, both of which provides energetic benefits (see additional
+convective heat loss, both of which provide energetic benefits (see additional
 discussion in Supplemental Material \hyperref[exploring-insolation-solar-radiation-as-a-model-covariate]{Exploring Insolation (Solar Radiation) as a
@@
 shown the potential for considerable bias in abundance estimates if such
-covariates are unaccounted for (Conn \& Trukhanova, 2023). We recommend that future abundance
-analyses employ availability models that account for them. For instance, it is
-relatively straightforward to obtain weather reanalysis products for times and
-locations that are surveyed and to construct a relevant correction factor based
-on predictions of GLMPMs. The most challenging element in developing
-availability correction factors is with annual variability. It can be difficult
-to get a sufficient sample size to estimate year-specific correction factors,
-particularly because research teams would likely need to deploy bio-loggers on
-seals and conduct aerial surveys concurrently, requiring considerably more
-personnel and money. One possible suggestion is to include year as a random
-effect within models for aerial survey counts such that, without specific
-knowledge of any particular year, the among-year variance is included in the
-modeled standard errors. Regardless of the specific approach, future estimates
-of Arctic seal abundance will require specific consideration of annual
-variability and changes in the timing of peak haul-out behavior when estimating
-trends, as one will not know if moderate differences in abundance estimates are
-attributable to changes in abundance or changes in haul-out behavior.
+covariates are unaccounted for (see Conn \& Trukhanova (2023) for discussion about the
+importance of including stable age- and stage-proportions). We recommend that
+future abundance analyses employ availability models that account for them. For
+instance, it is relatively straightforward to obtain weather reanalysis products
+(e.g.~NARR, ERA5) for times and locations that are surveyed and to construct a
+relevant correction factor based on predictions of GLMPMs. The most challenging
+element in developing availability correction factors is with annual
+variability. It can be difficult to get a sufficient sample size to estimate
+year-specific correction factors, particularly because research teams would
+likely need to deploy bio-loggers on seals and conduct aerial surveys
+concurrently, requiring considerably more personnel and money. One possible
+suggestion is to include year as a random effect within models for aerial survey
+counts such that, without specific knowledge of any particular year, the
+among-year variance is included in the modeled standard errors. Regardless of
+the specific approach, future estimates of Arctic seal abundance will require
+specific consideration of annual variability and changes in the timing of peak
+haul-out behavior when estimating trends, as one will not know if moderate
+differences in abundance estimates are attributable to changes in abundance or
+changes in haul-out behavior.
 

### 2024-08-19 analysis/paper/Letter_Response_to_Reviewers_2.qmd
response letter, round 2

---
subject: "Additional Revisions to Maunscript 2022:04:72978"
sender: |
  Josh M. London, PhD \
  7600 Sand Point Way NE . Seattle, Washington, 98115 . josh.london@noaa.gov"
recipient: |
  To: Suzanne Prange \ 
  Academic Editor \ 
  PeerJ Life & Environment
sent: "19 August 2024"
format:
  letter-typst: default
---

Dr. Prange

I sincerely appreciate the additional feedback from the reviewers and and
pleased to hear you support publication of this article in *PeerJ*. I was,
unfortunately, on leave for a portion of the past few weeks and thus was unable
to make the initial 15 day deadline. I have revised the manuscript in response
to feedback from reviewer M. Carter and have provided additional details below.
I especially appreciated Matt's suggestion for a new title and some of this
thoughts regarding the ordering of topics.

I also want to take a moment and thank the *PeerJ* staff for their previous
efforts and dedication ensuring the formatting details, accessibility, and data
access were all finalized during the previous submission.

Sincerely

Josh M. London, PhD\
Wildlife Biologist\
Alaska Fisheries Science Center, NOAA Fisheries

--------------------------------------------------------------------------------

#### Response to Comments from M Carter

The reviewer provided a useful suggestion regarding the manuscript title and the
order in which particular topics are presented. We have re-titled the manuscript
per their suggestion to *Spring haul-out behavior of seals in the Bering and
Chukchi Seas: implications for abundance estimation*. Additionally we have made
minor edits to the Abstract and Introduction in response to the reviewer's
suggestions regarding topic order.

*L59: Typo? “Trends” rather than “trend”?*

Fixed

*L111: Maybe outline for non-specialist readers that adhering bio-loggers to
hair is the conventional approach for most devices on pinnipeds.*

The text now reads:

> These devices are designed to collect data through the molt period (when those
> adhered to the hair -- a more conventional method -- would fall off) and, in
> some situations, provide multiple years of data.

*L113: Given the final sentence of the previous paragraph, I think it’s
important to detail that you’re not using a continuous 16-year dataset from one
device here, rather you have compiled data collected from multiple devices
spanning a 16-year period.*

The text now reads:

> In this study, we used data collected from multiple bio-logging deployments
> spanning a 16-year period to investigate the haul-out behavior of bearded,
> ribbon, and spotted seals in the Bering and Chukchi seas.

*L138: What is meant by “primary literature” here? I can’t see any references
associated with the deployments.*

The supplemental table lists primary literature associated with each study where
additional details regarding the capture techniques, etc can be found. I've
update the text for additional clarification:

> We refer readers to the primary literature for detailed capture and bio-logger
> attachment methods (see publications listed in Supplemental Material S1)

*L156-157: Can you clarify this? Is it that you get erroneous haulout records
where the animal is actually at sea and thus there is a bias introduced into the
percentage time hauled out estimate that is used to scale the abundance
estimates? In cases where flipper-mounted tags were not deployed, did you use
any spatial cleaning method to identify and remove these (i.e. if they are
clearly offshore according to the associated location data)? Or is this
complicated by lack of knowledge about sea ice presence?*

The text currently states *bio-loggers attached to the head or dorsal region are
often dry while the seal is floating at the surface, inducing a slight positive
bias in the hourly percent-dry values reported by the bio-logger*. We're not
sure how we could efficiently improve the specifics of this statement. Because
seals must put their heads out of the water to breath, tags attached to the head
or dorsal region will naturally spend more time out of the water compared to the
flipper. This results in a slight bias.

*L365 and 383: Perhaps interesting to note from an ontogeny perspective that the
diurnal pattern of haulout seems to develop in YOY from a very young age for
both ribbon and spotted seals.*

An additional sentence was added to highlight this

*L430: Repetition of “higher latitudes”.*

Fixed

*L463: Can you comment on the ontogeny of haulout behaviour seen in pups for
these species? It seems interesting that they adopt the diel pattern in haulout
behaviour so soon after weaning as presumably they are not moulting? Can you
unpack this in the context of existing knowledge, if any exists? If none exists,
perhaps draw attention to its novelty?*

Additional text was added to note this ontogeny.

*L474: “Both of which provide” (not “provides”).*

Fixed

*L523-524: Can you give an idea of the magnitude of this bias?*

The Conn & Trukhanova 2023 paper describes the bias in detail related to the
inclusion of age and stage so the text was edited to more directly encourage the
readers to see that paper for more details. We don't currently have the
framework to measure the magnitude of the bias but it might be possible after
we've completed updated abundance estimates from recent surveys.

*L567-569: I think this statement would be much more impactful if we had some
idea of the magnitude of bias in abundance estimates from not accounting for
these factors.*

Agreed. Unfortunately, including this and conducting the additional analysis
would exceed the expectations of 'minor revisions' and delay publication. We
will keep this in mind for future analyses and manuscripts.

### 2024-08-20 analysis/paper/London_HauloutBehavior.tex
edit abstract and remove supplemental from main

@@
 Ice-associated seals rely on sea ice for a variety of activities, including pupping, breeding, molting, and resting.
-In the Arctic, many of these activities occur in spring (April -- June) as sea ice begins to melt and retreat northward.
+In the Arctic, many of these activities occur in spring (April through June) as sea ice begins to melt and retreat northward.
 Rapid acceleration of climate change in Arctic ecosystems is therefore of concern as the quantity and quality of suitable habitat is forecast to decrease.
@@
 We specifically focused on bearded (\emph{Erignathus barbatus}), ribbon (\emph{Histriophoca fasciata}), and spotted seals (\emph{Phoca largha}) in the Bering and Chukchi seas.
-Because ringed seals (\emph{Phoca (pusa) hispida}) can be out of the water but hidden from view in snow lairs analysis of their `availability' to surveys requires special consideration; therefore,they were not included in this analysis.
+Because ringed seals (\emph{Phoca (pusa) hispida}) can be out of the water but hidden from view in snow lairs analysis of their `availability' to surveys requires special consideration; therefore, they were not included in this analysis.
 Using generalized linear mixed pseudo-models to properly account for temporal autocorrelation, we fit models with covariates of interest (e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure, temperature, precipitation) to examine their ability to explain variation in haul-out probability.
@@
 
-\newpage
-\setcounter{table}{0}\renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0}\renewcommand{\thefigure}{S\arabic{figure}}
-\begin{landscape}
-
-\section*{Supplemental Material}\label{supplemental-material}
-\addcontentsline{toc}{section}{Supplemental Material}
-
-\subsection{Additional Bio-logger Deployment Details}\label{additional-bio-logger-deployment-details}
-
-\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}
-
-\global\setlength{\Oldtabcolsep}{\tabcolsep}
-
-\setlength{\tabcolsep}{2pt}
-
-\renewcommand*{\arraystretch}{1.5}
-
-\providecommand{\ascline}[3]{\noalign{\global\arrayrulewidth #1}\arrayrulecolor[HTML]{#2}\cline{#3}}
-
-\begin{longtable}[c]{|p{1.00in}|p{0.80in}|p{1.50in}|p{1.50in}|p{0.80in}|p{0.80in}|p{0.80in}|p{1.50in}}
-
-\caption{The\ \ timing,\ location,\ and\ institutions\ responsible\ for\ the\ bio-logger\ deployments\ used\ in\ this\ study\ along\ with\ research\ permits\ and\ any\ associated\ publications.}\label{tab:deploymentTable}\\
-
-\ascline{1.5pt}{666666}{1-8}
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Institution}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Year\ Deployed}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Location}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Publication(s)}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Permits}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{No.\ Seals}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Sex}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Age\ Class}}}} \\
-
-\ascline{1.5pt}{666666}{1-8}\endfirsthead
-
-\ascline{1.5pt}{666666}{1-8}
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Institution}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Year\ Deployed}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Location}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Publication(s)}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Permits}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{No.\ Seals}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Sex}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\textbf{Age\ Class}}}} \\
-
-\ascline{1.5pt}{666666}{1-8}\endhead
-
-\multicolumn{8}{>{\raggedright}m{\dimexpr 8.7in+14\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{\textbf{ADFG}}}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{=Alaska\ Department\ of\ Fish\ and\ Game;\ }}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{\textbf{NSB}}}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{=North\ Slope\ Borough;\ }}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{\textbf{NMFS}}}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{=NOAA\ National\ Marine\ Fisheries\ Service;\ }}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{\textbf{NPWC}}}\textcolor[HTML]{000000}{\fontsize{8}{8}\selectfont{=North\ Pacific\ Wildlife\ Consortium}}} \\
-
-\endlastfoot
-
-\multicolumn{8}{>{\raggedright}m{\dimexpr 8.7in+14\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{\textbf{Bearded\ seal}}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2005}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Cameron\ et\ al\ 2018}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1585}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{6}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2006}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Cameron\ et\ al\ 2018}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1585}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2009}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Breed\ et\ al\ 2018}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1787}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{4}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2014}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ Koyuk\ River}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2014}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ Nome}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2015}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ St.\ Michael}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Elson\ Lagoon,\ Utqiagvik}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ Koyuk\ River}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ Nome}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ St.\ Michael}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Colville\ River,\ Nuiqsut}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ Koyuk\ River}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Norton\ Sound,\ Nome}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2020}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2019}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Dease\ Inlet,\ Utqiagvik}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Olnes\ et\ al\ 2021}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{20466}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2005}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1585}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2009}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{McClintock\ et\ al\ 2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1765}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2011}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{McClintock\ et\ al\ 2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15126}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{3}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2012}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{McClintock\ et\ al\ 2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15126}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2012}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Elson\ Lagoon,\ Utqiagvik}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2019}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Pittalugruaq\ Lake}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{20466}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{8}{>{\raggedright}m{\dimexpr 8.7in+14\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{\textbf{Ribbon\ seal}}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2005}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Ozemoy\ Gulf,\ Russia}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1765}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{9}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{young\ of\ year,\ adult,\ subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2006}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1765}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{7}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult,\ young\ of\ year}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2007}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1765}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{28}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult,\ young\ of\ year}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2008}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1765}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2009}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1765}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{28}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult,\ subadult,\ young\ of\ year}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2010}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1787,\ 15126}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{17}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{young\ of\ year,\ adult,\ subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2014}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15126}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{13}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult,\ young\ of\ year}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{19309}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{7}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult}}} \\
-
-\multicolumn{8}{>{\raggedright}m{\dimexpr 8.7in+14\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{\textbf{Spotted\ seal}}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2005}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kotzebue\ Sound}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1585}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{3}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Dease\ Inlet,\ Utqiagvik}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{4}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Colville\ River,\ Nuiqsut}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Scammon\ Bay}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{3}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2018}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Dease\ Inlet,\ Utqiagvik}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{20466}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2018}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Scammon\ Bay}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{20466}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{ADFG}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2019}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Dease\ Inlet,\ Utqiagvik}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{20466}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{6}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult,\ subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2006}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1676}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{5}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{young\ of\ year,\ subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2007}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{782-1676}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{12}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult,\ young\ of\ year,\ subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2009}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1787}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{23}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult,\ subadult,\ young\ of\ year}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2010}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{358-1787,\ 15126}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{8}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F,\ M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{young\ of\ year,\ adult,\ subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2014}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15126}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{5}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{young\ of\ year,\ adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{19309}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{6}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NMFS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2018}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Bering\ Sea}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{19309}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{5}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NPWC}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2009}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Kamchatka\ Peninsula}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NA}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{3}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2012}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Tiny\ Island}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2014}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Oarlock\ Island}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{6}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2014}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Seal\ Island}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2015}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Oarlock\ Island}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{6}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M,\ F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult,\ adult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2016}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Pittalugruaq\ Lake}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{3}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 1in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{NSB}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{2017}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Pittalugruaq\ Lake}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Von\ Duyke\ et\ al\ in\ prep}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{15324}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{1}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.8in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{M}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.5in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{subadult}}} \\
-
-\ascline{1.5pt}{666666}{1-8}
-
-\end{longtable}
-
-\arrayrulecolor[HTML]{000000}
-
-\global\setlength{\arrayrulewidth}{\Oldarrayrulewidth}
-
-\global\setlength{\tabcolsep}{\Oldtabcolsep}
-
-\renewcommand*{\arraystretch}{1}
-
-\end{landscape}
-
-\clearpage
-
-\subsection{Supplemental Figures Showing Confidence Intervals Associated with Predictions}\label{supplemental-figures-showing-confidence-intervals-associated-with-predictions}
-
-The following series of figures (\ref{fig:beardedPredSE}, \ref{fig:ribbonPredSE},
-and \ref{fig:spottedPredSE}) show the seasonal variability in predicted haul-out
-probability and the associated 95\% confidence intervals for bearded, ribbon, and
-spotted seals. The predictions shown are based on the same data used in
-\ref{fig:beardedHOCal}, \ref{fig:ribbonHOCal}, and \ref{fig:spottedHOCal} but
-selected for three local solar hours (07:00, 12:00, and 17:00) so the
-confidence intervals can also be shown and comparisons can be made.
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{../figures/Figure-013} \caption{\textbf{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for bearded seals.} \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
-Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
-Age and sex classes are combined into a single `all ages' category.}\label{fig:beardedPredSE}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{../figures/Figure-014} \caption{\textbf{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals.} \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
-Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
-Age and sex classes are separated to allow comparisons.}\label{fig:ribbonPredSE}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{../figures/Figure-015} \caption{\textbf{Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for spotted seals.} \linebreak Model predictions are shown for three local solar hours (07:00, 12:00, and 17:00).
-Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season.
-Age and sex classes are separated to allow comparisons.}\label{fig:spottedPredSE}
-\end{figure}
-
-\clearpage
-
-\subsection{Exploring Insolation (Solar Radiation) as a Model Covariate}\label{exploring-insolation-solar-radiation-as-a-model-covariate}
-
-\subsubsection{Introduction}\label{introduction-1}
-
-During the peer review process for this manuscript, Anthony Fischbach suggested
-the possibility of using predicted insolation (or solar radiation) values from
-the reanalysis model as a more direct and, potentially, more informative
-predictor of the daily haul-out cycle in seals compared to time of day. The
-notion being that seals are, likely, directly responding to changes in solar radiation
-throughout the day and not what time of day it is (i.e.~seals don't have human
-watches). Additionally, given the energetic benefits of increased solar
-radiation it could be more informative as we would expect seals might have a
-higher haul-out probability on sunnier days and for there to be geographic
-variability in haul-out behavior associated with geographical differences in
-insolation. This approach has an additional benefit of being more parsimonious
-compared to our use of the Fourier series or other approaches to represent
-hour-of-day in the model (e.g.~24 factors for each hour).
-
-Because of these reasons, we considered and explored this
-possibility for our model and the analysis presented in this manuscript. A key
-drawback to reliance on solar radiation, in our minds, is that we would lose
-insight regarding potential diel patterns -- solar radiation does not
-differentiate between dusk or dawn. Bi-modal patterns have been
-previously observed in ringed seals and our results in this analysis show some indication of
-increased haul-out probability during dawn compared to dusk periods for bearded seals
-and some age and sex classes for ribbon and spotted seals.
-For other phocid species, increased haul-out probability before solar noon or after
-solar noon has been observed. Importantly, understanding these relationships
-between haul-out probability and hour-of-day can have important ramifications
-on aerial survey study design -- a key focus of this paper.\\
-Another hesitation we had was that solar radiation estimates from reanalysis models have
-not been previously used as a model covariate within a published study of
-pinniped haul-out behavior. Thus, for this analysis, we chose to keep our
-original approach and rely on the Fourier series to capture any hour-of-day
-effects.
-
-That said, we think the idea of solar radiation as a model covariate in pinniped
-haul-out models is intriguing and worth further exploration. The current
-availability and increased accessibility to detailed climate reanalysis
-products that include solar radiation is exciting and we encourage future, more
-detailed exploration of this as a component in pinniped haul-out analysis. To
-provide some inspiration, we present some initial efforts and examples for
-comparison.
-
-\subsubsection{Methods}\label{methods-1}
-
-In this manuscript, we rely on the NARR reanalysis model as the source for our
-weather covariates. However, since our initiation of this analysis, the ERA5
-reanalysis model (\url{https://doi.org/10.24381/cds.adbb2d47}) has become one of the
-go-to standards for global climate reanalysis and provides an increased temporal
-resolution to hourly (compared to the 3-hour resolution of NARR). The global
-coverage of ERA5 provides additional flexibility in that the area of interest is
-not limited to North America. The ERA5 model provides a number of solar
-radiation parameters and it was important to evaluate and understand each of
-these estimates in order to select the one that was likely most relevant to
-seals. Here, we used the `surface short-wave (solar) radiation downwards'
-parameter. This
-parameter is described as \emph{``the amount of solar radiation (also known as shortwave radiation)
-that reaches a horizontal plane at the surface of the Earth and comprises both
-direct and diffuse solar radiation. To a reasonably good approximation, this
-parameter is the model equivalent of what would be measured by a pyranometer (an
-instrument used for measuring solar radiation) at the surface''}
-(\url{https://codes.ecmwf.int/grib/param-db/?id=169}). Thus, this is the
-value which most closely represents the amount of solar radiation likely
-felt by a seal hauled out of the water.
-
-ERA5 data is available via the Copernicus climate data store API which can be
-queried with the CDS-API Python package
-(\url{https://cds.climate.copernicus.eu/api-how-to}). The R code provided here
-documents the download of the \emph{surface\_solar\_radiation\_downwards} parameter for
-our study area of interest and years of interest. The \texttt{reticulate} R package
-(\url{https://CRAN.R-project.org/package=reticulate}) allowed interaction with Python.
-Additionally, note, extra steps are required to download data on either side
-of the 180 anti-meridian.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(tidyverse)}
-\FunctionTok{library}\NormalTok{(reticulate)}
-\FunctionTok{library}\NormalTok{(sf)}
-\FunctionTok{library}\NormalTok{(terra)}
-
-\CommentTok{\#import python CDS{-}API}
-\NormalTok{cdsapi }\OtherTok{\textless{}{-}} \FunctionTok{import}\NormalTok{(}\StringTok{\textquotesingle{}cdsapi\textquotesingle{}}\NormalTok{)}
-\CommentTok{\#for this step there must exist the file .cdsapirc}
-\NormalTok{server }\OtherTok{=}\NormalTok{ cdsapi}\SpecialCharTok{$}\FunctionTok{Client}\NormalTok{() }\CommentTok{\#start the connection}
-
-\NormalTok{get\_era5 }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(y) \{}
-  \CommentTok{\#we create the query}
-\NormalTok{  query }\OtherTok{\textless{}{-}} \FunctionTok{r\_to\_py}\NormalTok{(}
-    \FunctionTok{list}\NormalTok{(}
-      \AttributeTok{variable =} \StringTok{"surface\_solar\_radiation\_downwards"}\NormalTok{,}
-      \AttributeTok{product\_type =} \StringTok{"reanalysis"}\NormalTok{,}
-      \AttributeTok{area =} \StringTok{"75/152/47/180"}\NormalTok{, }\CommentTok{\# North, West, South, East}
-      \AttributeTok{year =}\NormalTok{ y,}
-      \AttributeTok{month =} \FunctionTok{str\_pad}\NormalTok{(}\DecValTok{2}\SpecialCharTok{:}\DecValTok{7}\NormalTok{, }\DecValTok{2}\NormalTok{, }\StringTok{"left"}\NormalTok{, }\StringTok{"0"}\NormalTok{),}
-      \AttributeTok{day =} \FunctionTok{str\_pad}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\DecValTok{31}\NormalTok{, }\DecValTok{2}\NormalTok{, }\StringTok{"left"}\NormalTok{, }\StringTok{"0"}\NormalTok{),}
-      \AttributeTok{time =} \FunctionTok{str\_c}\NormalTok{(}\DecValTok{0}\SpecialCharTok{:}\DecValTok{23}\NormalTok{, }\StringTok{"00"}\NormalTok{, }\AttributeTok{sep =} \StringTok{":"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{str\_pad}\NormalTok{(}\DecValTok{5}\NormalTok{, }\StringTok{"left"}\NormalTok{, }\StringTok{"0"}\NormalTok{),}
-      \AttributeTok{format =} \StringTok{"netcdf"}
-\NormalTok{    )}
-\NormalTok{  )}
-  \CommentTok{\#query to get the ncdf}
-\NormalTok{  server}\SpecialCharTok{$}\FunctionTok{retrieve}\NormalTok{(}\StringTok{"reanalysis{-}era5{-}single{-}levels"}\NormalTok{,}
-\NormalTok{                  query,}
-                  \FunctionTok{paste0}\NormalTok{(}\StringTok{"era5\_ssrd\_"}\NormalTok{, y, }\StringTok{"\_left.nc"}\NormalTok{))}
-
-\NormalTok{  query }\OtherTok{\textless{}{-}} \FunctionTok{r\_to\_py}\NormalTok{(}
-    \FunctionTok{list}\NormalTok{(}
-      \AttributeTok{variable =} \StringTok{"surface\_solar\_radiation\_downwards"}\NormalTok{,}
-      \AttributeTok{product\_type =} \StringTok{"reanalysis"}\NormalTok{,}
-      \AttributeTok{area =} \StringTok{"75/{-}180/47/{-}142"}\NormalTok{, }\CommentTok{\# North, West, South, East}
-      \AttributeTok{year =}\NormalTok{ y,}
-      \AttributeTok{month =} \FunctionTok{str\_pad}\NormalTok{(}\DecValTok{2}\SpecialCharTok{:}\DecValTok{7}\NormalTok{, }\DecValTok{2}\NormalTok{, }\StringTok{"left"}\NormalTok{, }\StringTok{"0"}\NormalTok{),}
-      \AttributeTok{day =} \FunctionTok{str\_pad}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\DecValTok{31}\NormalTok{, }\DecValTok{2}\NormalTok{, }\StringTok{"left"}\NormalTok{, }\StringTok{"0"}\NormalTok{),}
-      \AttributeTok{time =} \FunctionTok{str\_c}\NormalTok{(}\DecValTok{0}\SpecialCharTok{:}\DecValTok{23}\NormalTok{, }\StringTok{"00"}\NormalTok{, }\AttributeTok{sep =} \StringTok{":"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{str\_pad}\NormalTok{(}\DecValTok{5}\NormalTok{, }\StringTok{"left"}\NormalTok{, }\StringTok{"0"}\NormalTok{),}
-      \AttributeTok{format =} \StringTok{"netcdf"}
-\NormalTok{    )}
-\NormalTok{  )}
-  \CommentTok{\#query to get the ncdf}
-\NormalTok{  server}\SpecialCharTok{$}\FunctionTok{retrieve}\NormalTok{(}\StringTok{"reanalysis{-}era5{-}single{-}levels"}\NormalTok{,}
-\NormalTok{                  query,}
-                  \FunctionTok{paste0}\NormalTok{(}\StringTok{"era5\_ssrd\_"}\NormalTok{, y, }\StringTok{"\_right.nc"}\NormalTok{))}
-
-\NormalTok{\}}
-
-\NormalTok{years }\OtherTok{\textless{}{-}} \FunctionTok{as.character}\NormalTok{(}\DecValTok{2005}\SpecialCharTok{:}\DecValTok{2021}\NormalTok{)}
-\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(years)) \{}
-  \FunctionTok{get\_era5}\NormalTok{(years[i])}
-\NormalTok{\}}
-\end{Highlighting}
-\end{Shaded}
-
-To explore performance of our solar radiation parameter within a haul-out model
-we replaced the various Fourier series parameters in our model from the manuscript
-with the ERA5 \emph{surface solar radiation downwards} (\texttt{era\_ssrd\_watts}) parameter. As
-with other reanalysis values (from NARR) in the manuscript, the \texttt{era-ssrd-watts}
-values are matched in time and space to the seal haul-out observation data; we
-use the full hourly temporal resolution from ERA5. The glmmLDTS framework used
-in the paper does not allow for model comparisons with AIC because of
-the reliance on pseudo-likelihood. The \texttt{bam()} function within the \texttt{mgcv} package
-provides a quick model fitting option that also allowed us to do some model
-comparison with AIC. This approach was sufficient for the general demonstration and
-exploration purposes here
-but future research should consider a range of model fitting frameworks and
-approaches that might be more appropriate.
-
-The model specification below was used to specify an \texttt{mgcv::bam()} model that
-matched the formula used in the manuscript for ribbon seals. The \texttt{s(speno,\ bs\ =\ "re")}
-term is the smooth term for the random effect. All other predictors were
-the same.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{m1\_ribbon }\OtherTok{\textless{}{-}}\NormalTok{ mgcv}\SpecialCharTok{::}\FunctionTok{bam}\NormalTok{(}
-\NormalTok{  dry }\SpecialCharTok{\textasciitilde{}}\NormalTok{ age\_sex }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(speno, }\AttributeTok{bs =} \StringTok{"re"}\NormalTok{) }\SpecialCharTok{+} 
-\NormalTok{    sin1 }\SpecialCharTok{+}\NormalTok{ cos1 }\SpecialCharTok{+}\NormalTok{ sin2 }\SpecialCharTok{+}\NormalTok{ cos2 }\SpecialCharTok{+}\NormalTok{ sin3 }\SpecialCharTok{+}\NormalTok{ cos3 }\SpecialCharTok{+}
-    \FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+} 
-\NormalTok{    sin1}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    cos1}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    sin2}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    cos2}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    sin3}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    cos3}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    wind}\SpecialCharTok{*}\NormalTok{temp2m }\SpecialCharTok{+}\NormalTok{ pressure }\SpecialCharTok{+}\NormalTok{ precip }\SpecialCharTok{+} 
-\NormalTok{    age\_sex}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{4}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{),}
-  \AttributeTok{data =}\NormalTok{ ribbon\_model\_data,}
-  \AttributeTok{family =}\NormalTok{ binomial,}
-  \AttributeTok{discrete =} \ConstantTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Note, the specification for \emph{m1\_ribbon} here does not include
-any AR1 structure for temporal autocorrelation. To include this, we needed to
-provide a value for \(\rho\) (or \emph{rho}). We examined the autocorrelation within
-the model and used the lag-1 value for \(\rho\) .The value for lag-1
-autocorrelation was 0.8082 which is
-rather high but not surprising. We then updated our model specification with
-a value for \(\rho\) as well as the \texttt{A1.start} argument which specifies (as either
-\textbf{TRUE} or \textbf{FALSE}) the start point of each block.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{m2\_ribbon }\OtherTok{\textless{}{-}}\NormalTok{ mgcv}\SpecialCharTok{::}\FunctionTok{bam}\NormalTok{(}
-\NormalTok{  dry }\SpecialCharTok{\textasciitilde{}}\NormalTok{ age\_sex }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(speno, }\AttributeTok{bs =} \StringTok{"re"}\NormalTok{) }\SpecialCharTok{+} 
-\NormalTok{    sin1 }\SpecialCharTok{+}\NormalTok{ cos1 }\SpecialCharTok{+}\NormalTok{ sin2 }\SpecialCharTok{+}\NormalTok{ cos2 }\SpecialCharTok{+}\NormalTok{ sin3 }\SpecialCharTok{+}\NormalTok{ cos3 }\SpecialCharTok{+}
-    \FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+} 
-\NormalTok{    sin1}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    cos1}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    sin2}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    cos2}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    sin3}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    cos3}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    wind}\SpecialCharTok{*}\NormalTok{temp2m }\SpecialCharTok{+}\NormalTok{ pressure }\SpecialCharTok{+}\NormalTok{ precip }\SpecialCharTok{+} 
-\NormalTok{    age\_sex}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{),}
-  \AttributeTok{data =}\NormalTok{ ribbon\_model\_data,}
-  \AttributeTok{family =}\NormalTok{ binomial,}
-  \AttributeTok{AR.start =}\NormalTok{ ar1\_start,}
-  \AttributeTok{rho =}\NormalTok{ lag1\_ribbon,}
-  \AttributeTok{discrete =} \ConstantTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-The model specification for exploring the use of solar radiation was
-specified similarly but without all of the Fourier series parameters and
-interactions.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{m2\_ssrd\_ribbon }\OtherTok{\textless{}{-}}\NormalTok{ mgcv}\SpecialCharTok{::}\FunctionTok{bam}\NormalTok{(}
-\NormalTok{  dry }\SpecialCharTok{\textasciitilde{}}\NormalTok{ age\_sex }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(speno, }\AttributeTok{bs =} \StringTok{"re"}\NormalTok{) }\SpecialCharTok{+} 
-\NormalTok{    era5\_ssrd\_watts }\SpecialCharTok{+}
-    \FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+} 
-\NormalTok{    era5\_ssrd\_watts}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{    wind}\SpecialCharTok{*}\NormalTok{temp2m }\SpecialCharTok{+}\NormalTok{ pressure }\SpecialCharTok{+}\NormalTok{ precip }\SpecialCharTok{+} 
-\NormalTok{    age\_sex}\SpecialCharTok{:}\FunctionTok{poly}\NormalTok{(day, }\DecValTok{3}\NormalTok{, }\AttributeTok{raw=}\ConstantTok{TRUE}\NormalTok{),}
-  \AttributeTok{data =}\NormalTok{ ribbon\_model\_data,}
-  \AttributeTok{family =}\NormalTok{ binomial,}
-  \AttributeTok{AR.start =}\NormalTok{ ar1\_start,}
-  \AttributeTok{rho =}\NormalTok{ lag1\_ribbon,}
-  \AttributeTok{discrete =} \ConstantTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-The two models were compared with AIC to evaluate whether the
-reduction in degrees of freedom with fewer terms in the solar radiation
-model was matched with improved explanatory power in the model fit. While the
-model and code specified above is for ribbon seals, the same approach was
-repeated for bearded and spotted seals.
-
-A similar approach to that presented in this manuscript for prediction was
-employed with solar radiation values in lieu of hour of day. For prediction
-values, quantiles (5\% increments) of the observed range of ERA5 solar radiation values were
-used with 100\% representing the maximum observed solar radiation value. This
-allowed similar data visualizations and easier comparisons to those
-predictions in the manuscript that include hour of day.
-
-\subsubsection{Results}\label{results-1}
-
-To evaluate whether the solar radiation parameter matched our expectations and
-compared well with hour of the day, we visualized the variability of the
-\texttt{era5\_ssrd} values within
-our study area as they relate to hour of the day (\ref{fig:era5-ssrd-plot}).
-The unimodal distribution
-is centered around the middle of the solar day with peak solar radiation
-coinciding with 13:00 local solar. This suggests solar radiation could be an
-informative covariate for capturing unimodal diel patterns in haul-out
-behavior.
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{../figures/Figure-016} \caption{\textbf{Diel Pattern of Solar Radiation Values from ERA5 Reanalysis.} \linebreak Downward surface solar radiation estimates from the ERA5 climate reanalysis for 5000 random points within the study area between 2005 and 2021.
-Solar radiation values are presented in Watts per square-meter and the smoothed line highlights the strong diel pattern.}\label{fig:era5-ssrd-plot}
-\end{figure}
-
-The bearded seal model matching the specification from the manuscript resulted in
-126.13 degrees of freedom and an AIC value of
--7428.929. The model with solar radiation resulted in
-39.619 degrees of freedom and an AIC value of
--6797.378. The ribbon seal model matching the specification from the manuscript resulted in
-131.478 degrees of freedom and an AIC value of
--16372.29. The model with solar radiation resulted in
-115.126 degrees of freedom and an AIC value of
--16038.175. The spotted seal model matching the specification from the manuscript resulted in
-125.506 degrees of freedom and an AIC value of
--23584.373. The model with solar radiation resulted in
-109.163 degrees of freedom and an AIC value of
--23302.772. Despite the additional terms, the
-models with the Fourier series representation of hour of day resulted in a
-lower AIC value and were still preferred models for each of the species.
-
-Predictions from the model fits and visualization of those predictions were
-produced for each species but, here, we only present visualizations from
-ribbon seals as an example (Figure \ref{fig:ssrd-prediction-heatmap} and Figure
-\ref{fig:ssrd-ribbonPredSE}). Similar seasonal patterns previously observed were
-still apparent with subadults hauling out earlier in the season followed by
-adult males and, then, adult females. The observed relationship with hour of
-day and the centering of peak haul-out probability around solar noon was
-reflected in these predictions as a one-sided distribution with maximum solar
-radiation having the highest haul-out probability and minimal solar radiation
-the least. The seasonal distribution of haul-out probability along with 95\%
-confidence intervals also provided comparable insights (see figures
-\ref{fig:ribbonPredSE} and \ref{fig:ssrd-ribbonPredSE}). That said, subtle
-differences in the shape and extent of confidence limits were present.
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{../figures/Figure-017} \caption{\textbf{Solar radiation as a predictor of ribbon seal haul-out probability.} \linebreak Predicted haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model.
-In this model, solar radiation was used in lieu of hour of day.
-The apparent seasonal progression with subadults hauling out earlier in the season followed by adult males and, then, adult females is still notable although maybe not as clear.
-Predictions for young of the year still show their transition from newly weaned pups resting on the ice to more in-water activities.
-The overall pattern is in agreement with a one-sided view of Figure 7 where maximum solar radiation is equivalent to local solar noon.}\label{fig:ssrd-prediction-heatmap}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{../figures/Figure-018} \caption{\textbf{Solar radiation as a predictor of ribbon seal haul-out probability (with uncertainty).} \linebreak Seasonal variability in haul-out probability and the associated 95\% confidence intervals (shaded area) for ribbon seals.
-In this model predictions are shown for low, medium, and high values of solar radiation (as percentages of the maximum value observed) in lieu of local solar hour.
-There's general agreement in the overall seasonal patterns between the two approaches but sublte differences in shape and extent of the confidence limits were present (see Figure S2 for comparisons).}\label{fig:ssrd-ribbonPredSE}
-\end{figure}
-
-\subsubsection{Discussion}\label{discussion-1}
-
-Solar radiation has potential as an informative covariate in pinniped haul-out
-models that can be directly linked to seal physiology and expected behavioral
-changes. The ERA5's \emph{surface solar radiation downwards} values aligned with hour
-of day and maximum values occurred at or just after local solar noon. This
-highlighted the informative potential for this approach. However, despite an
-overall reduction in the total number of parameters and degrees of freedom, AIC
-comparison still favored the models for each species that included hour of day
-as a Fourier series.
-
-This analysis was not intended to be a full comparison --
-we simply want to demonstrate the potential and inspire further investigation --
-but, there are three possibilities that might explain the preference for hour of
-day. First, there are a broad range of solar radiation values represented for each
-hour of the day. Cloud cover, fog, and precipitation all reduce downward solar
-radiation at the surface and we might expect this to impact haul-out
-probability. However, the photoperiod and the timing of sunrise and sunset are
-not impacted by weather and seals may be responding to these signals more than
-the amount of solar radiation. Additionally, this study spans a range of
-physiological cycles and energetic needs and higher solar radiation may not be a
-consistent driving influence on seals. Increased energy from the sun may be
-important during molt but less so during pupping and breeding periods. Second,
-the timing and duration of haul-out behavior may also be influenced by diel
-patterns in weather (e.g.~lower winds in the morning) or ecosystem dynamics
-(e.g.~prey availability) that lead to a skewness in the distribution of haul-out
-behavior that wouldn't be reliably captured by solar radiation values. Third,
-this effort is only an initial effort to explore the use of solar radiation in
-pinniped haul-out models. A more in depth and rigorous exploration of this topic
-might discover an approach that results in a more parsimonious and preferred
-model formulation.
-
-Again, we want to acknowledge Anthony Fischbach for the suggestion during the
-peer review process. We think this is an excellent example of the peer review
-process working to improve the quality of our manuscript and advance the
-scientific process. We hope others will take our example and expand on it
-within future analyses.
-
 \end{document}

### 2024-09-03 analysis/paper/London_HauloutBehavior.Rmd
remove british spelling of behavior from title

@@
 ---
-title: Spring haul-out behaviour of seals in the Bering and Chukchi Seas: implications for abundance estimation
+title: "Spring haul-out behavior of seals in the Bering and Chukchi Seas: implications for abundance estimation"
 preprint: false
@@
   pupping, breeding, molting, and resting. In the Arctic, many of these
-  activities occur in spring (April -- June) as sea ice begins to melt and
+  activities occur in spring (April through June) as sea ice begins to melt and
   retreat northward. Rapid acceleration of climate change in Arctic ecosystems
@@
   ice-associated seals. We specifically focused on bearded (_Erignathus
-  barbatus_), ribbon (_Histriophoca fasciata_), and spotted seals (_Phoca
-  largha_) in the Bering and Chukchi seas. Because ringed seals (_Phoca (pusa)
-  hispida_) can be out of the water but hidden from view in snow lairs analysis
-  of their 'availability' to surveys requires special consideration;
-  therefore,they were not included in this analysis. Using generalized linear
-  mixed pseudo-models to properly account for temporal autocorrelation, we fit
-  models with covariates of interest (e.g., day-of-year, solar hour, age-sex
-  class, wind speed, barometric pressure, temperature, precipitation) to examine
-  their ability to explain variation in haul-out probability. We found evidence
-  for strong diel and within-season patterns in haul-out behavior, as well as
-  strong weather effects (particularly wind and temperature). In general, seals
-  were more likely to haul out on ice in the middle of the day and when wind
-  speed was low and temperatures were higher. Haul-out probability increased
-  through March and April, peaking in May and early June before declining again.
-  The timing and frequency of haul-out events also varied based on species and
-  age-sex class. For ribbon and spotted seals, models with year effects were
-  highly supported, indicating that the timing and magnitude of haul-out
-  behavior varied among years. However, we did not find broad evidence that
-  haul-out timing was linked to annual sea-ice extent. Our analysis emphasizes
-  the importance of accounting for seasonal and temporal variation in haul-out
-  behavior, as well as associated environmental covariates, when interpreting
-  the number of seals counted in aerial surveys.
+  barbatus_), ribbon (_Histriophoca fasciata_), and spotted seals (_Phoca largha_) in
+  the Bering and Chukchi seas. Because ringed seals (_Phoca (pusa) hispida_) can
+  be out of the water but hidden from view in snow lairs analysis of their
+  'availability' to surveys requires special consideration; therefore, they were
+  not included in this analysis. Using generalized linear mixed pseudo-models to
+  properly account for temporal autocorrelation, we fit models with covariates
+  of interest (e.g., day-of-year, solar hour, age-sex class, wind speed,
+  barometric pressure, temperature, precipitation) to examine their ability to
+  explain variation in haul-out probability. We found evidence for strong diel
+  and within-season patterns in haul-out behavior, as well as strong weather
+  effects (particularly wind and temperature). In general, seals were more
+  likely to haul out on ice in the middle of the day and when wind speed was low
+  and temperatures were higher. Haul-out probability increased through March and
+  April, peaking in May and early June before declining again. The timing and
+  frequency of haul-out events also varied based on species and age-sex class.
+  For ribbon and spotted seals, models with year effects were highly supported,
+  indicating that the timing and magnitude of haul-out behavior varied among
+  years. However, we did not find broad evidence that haul-out timing was linked
+  to annual sea-ice extent. Our analysis emphasizes the importance of accounting
+  for seasonal and temporal variation in haul-out behavior, as well as
+  associated environmental covariates, when interpreting the number of seals
+  counted in aerial surveys.
 output:

### 2024-09-22 analysis/paper/London_HauloutBehavior.tex
final edits for the proofs

--- 2024-08-19 analysis/paper/London_HauloutBehavior.tex
@@
 In this paper, we use hourly percent-dry data from satellite-linked bio-loggers deployed between 2005 and 2021 to quantify the proportion of seals hauled out on ice.
-This information is needed to accurately estimate abundance from aerial survey counts of ice-associated seals (i.e., to correct for the proportion of animals that are in the water while surveys are conducted).
+This information is needed to accurately estimate abundance from aerial survey counts of ice-associated seals (i.e., to correct for the proportion of animals that are in the water, and so are not counted, while surveys are conducted).
 In addition to providing essential data for survey `availability' calculations, our analysis also provides insights into the seasonal timing and environmental factors affecting haul-out behavior by ice-associated seals.
 We specifically focused on bearded (\emph{Erignathus barbatus}), ribbon (\emph{Histriophoca fasciata}), and spotted seals (\emph{Phoca largha}) in the Bering and Chukchi seas.
-Because ringed seals (\emph{Phoca (pusa) hispida}) can be out of the water but hidden from view in snow lairs analysis of their `availability' to surveys requires special consideration; therefore,they were not included in this analysis.
+Because ringed seals (\emph{Phoca (pusa) hispida}) can be out of the water but hidden from view in snow lairs analysis of their `availability' to surveys requires special consideration; therefore, they were not included in this analysis.
 Using generalized linear mixed pseudo-models to properly account for temporal autocorrelation, we fit models with covariates of interest (e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure, temperature, precipitation) to examine their ability to explain variation in haul-out probability.
@@
 Haul-out probability increased through March and April, peaking in May and early June before declining again.
-The timing and frequency of haul-out events also varied based on species and age-sex class.
+The timing and frequency of haul-out events also varied based on species and age and sex class.
 For ribbon and spotted seals, models with year effects were highly supported, indicating that the timing and magnitude of haul-out behavior varied among years.
@@
 Knowledge of haul-out patterns is not only important for understanding natural
-history and ecology, but also for developing ``availability'' correction factors
+history and ecology, but also for developing `availability' correction factors
 for aerial surveys. Specifically, researchers need to know the fraction of seals
@@
 investigate the impact of weather variables. Such variables have been
-shown to influence walrus haul-out behavior (Udevitz et al., 2009) and we expect weather
+shown to influence walrus haul-out behavior (Udevitz et al., 2009) and we expected weather
 conditions to also influence seal haul-out behavior and including them within
@@
 capture and bio-logger attachment methods (see publications listed in
-Supplemental Material, \hyperref[s1]{S1}). NOAA-led research was conducted under the
+Supplemental Information Table \hyperref[s1]{S1}). NOAA-led research was conducted under the
 authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
@@
 (Ver Hoef, London \& Boveng, 2010) differ between models (Ten Eyck \& Cavanaugh, 2018). After fitting GLMPM
-models, we instead used ``type III'' \(F\)-tests to calculate \(p\)-values
+models, we instead used `type III' \(F\)-tests to calculate \(p\)-values
 (Ver Hoef, London \& Boveng, 2010) to evaluate model performance and important terms. We also
@@
 lower sample size for adult and young-of-year age classes. As such, results are
-shown for all ages (Figure \ref{fig:beardedHOCal}; see also
+shown for all ages (Figure \ref{fig:beardedHOCal}; see also Supplemental Information Fig.
 \ref{fig:beardedPredSE}). Additionally, after approximately 9 months, 7 devices
@@
 April that peaks in late May for subadults and in early June for adults (Figure
-\ref{fig:ribbonHOCal}; see also \ref{fig:ribbonPredSE}). There is an apparent
-seasonal progression with subadults hauling out earlier in the season followed
-by adult males and, then, adult females. Haul-out behavior was clearly centered
-around local solar noon and expanded to other hours later in the spring as seals
-entered their molting period. Subadults showed an earlier start and more intense
-haul-out activity in April and May. The young-of-the-year records began after
-weaning and the model predictions seemed to indicate development of in-water
-activities (e.g.~swimming, foraging) in May and, like adults, haul-out behavior
-was centered around solar noon. Adult females had a more protracted
-haul-out season compared to males, and more time was spent hauled out in June
-compared to adult males and subadults.
+\ref{fig:ribbonHOCal}; see also Supplemental Information Fig.
+\ref{fig:ribbonPredSE}). There is an apparent seasonal progression with
+subadults hauling out earlier in the season followed by adult males and, then,
+adult females. Haul-out behavior was clearly centered around local solar noon
+and expanded to other hours later in the spring as seals entered their molting
+period. Subadults showed an earlier start and more intense haul-out activity in
+April and May. The young-of-the-year records began after weaning and the model
+predictions seemed to indicate development of in-water activities (e.g.
+swimming, foraging) in May and, like adults, haul-out behavior was centered
+around solar noon. Adult females had a more protracted haul-out season compared
+to males, and more time was spent hauled out in June compared to adult males and
+subadults.
 
@@
 that was less intensely centered on solar noon (Figure \ref{fig:spottedHOCal};
-see also \ref{fig:spottedPredSE}). Adults of both sexes spent considerable time
-in April, May and June hauled out. Adult spotted seal males had a more
-protracted haul-out season compared to females, and more time out of the water
-in June (Figure \ref{fig:spottedHOCal}). This likely reflects the triad
-behavior in spotted seals when suitor males haul out with nursing females. As
-with ribbon seals, the young-of-the-year records began after weaning and the
-model predictions reflected development of in-water activities (e.g.
-swimming, foraging) in May and haul-out behavior centered around solar noon.
+see also Supplemental Information Fig.\ref{fig:spottedPredSE}). Adults of both
+sexes spent considerable time in April, May and June hauled out. Adult spotted
+seal males had a more protracted haul-out season compared to females, and more
+time out of the water in June (Figure \ref{fig:spottedHOCal}). This likely
+reflects the triad behavior in spotted seals when suitor males haul out with
+nursing females. As with ribbon seals, the young-of-the-year records began after
+weaning and the model predictions reflected development of in-water activities
+(e.g.~swimming, foraging) in May and haul-out behavior centered around solar
+noon.
 
@@
 patterns, uncovered significant contributions for linear and quadratic
-interactions between day and year for only spotted seals (day:year,
+interactions between day and year for only spotted seals (\emph{day:year},
 \(F_{15,115144}\)
 = 4.445; \(p\) =
-\textless0.001; day\textsuperscript{2}:year,
+\textless0.001; \emph{day\textsuperscript{2}:year},
 \(F_{15,115144}\)
@@
 \textless0.001). Ribbon seals showed no significant
-contribution for interactions between day and year (day:year,
+contribution for interactions between day and year (\emph{day:year},
 \(F_{10,99510}\)
 = 0.516; \(p\) =
-0.880; day\textsuperscript{2}:year,
+0.880; \emph{day\textsuperscript{2}:year},
 \(F_{10,99510}\)
@@
 showed a negative trend (peak haul-out occurs later in years with less
-sea ice) with a significant relationship for adult female spotted seals
+sea ice) but neither with a significant relationship\\
 (spotted seal adult female: \(R^{2}\) =
@@
 (2009)) and a few select studies of ice-associated seals
-(Perry, Stenson \& Buren, 2017; Hamilton \& Evans, 2018). In our study, ribbon seal haul-out behavior seemed to be the most
+(Perry, Stenson \& Buren, 2017; Hamilton \& Evans, 2018). In our study, ribbon seal haul-out behavior was notably
 influenced by weather, with wind, temperature, and barometric pressure all being
@@
 We recognize that the species and ecosystems we studied are within the ancestral
-and present-day environs of the Inpuiat and Yup'ik people who, through many
+and present-day environs of the Inupiat and Yup'ik people who, through many
 uncredited contributions of traditional knowledge, provided early western
@@
 biology' of Arctic seals. The deployment of bio-logging devices used in this
-study were often done in collaboration with Alaska Native seal hunters and the
+study were often done in collaboration with Alaska Native seal hunters and wiht the
 approval of
-their communities. We would like to especially acknowledge the communities of
+their communities. We would like to especially acknowledge the Alaska communities of
 Kotzebue, Koyuk, Nome, Nuiqsut, Scammon Bay, St.~Michael, Utqiaġvik, and Ulguniq
@@
 St.~Michael; Billy Adams, James Aiken, Tim Aiken, Howard Kittick,
-Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, and Joe Skin from Utqiaġvik,
-Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, and
-Fred Rexford from Ulguniq.
+Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Bobby Sarren, and Joe Skin from Utqiaġvik;
+Mary Ellen Ahmaogak, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, and
+Fred Rexford from Ulguniq. We would also like to acknowledge the support of the
+Ice Seal Committee.
 
 We are grateful for the assistance in catching and sampling seals by Ryan Adam,
-James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov,
-Cynthia Christman, Sarah Coburn, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
-Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey Goldstein, Jeff
-Harris, Jason Herreman, Markus Horning, John Jansen, Shawn Johnson, Charles
-Littnan, Lloyd Lowry, Brett McClintock, Erin Moreland, Mark Nelson, Justin
-Olnes, Lorrie Rea, Bob Shears, Gay Sheffield, Brent Stewart, Dave Withrow, and
+James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Anna Bryan,
+Vladamir Burkanov, Cynthia Christman, Sarah Coburn, Shawn Dahle, Rob Delong,
+Stacy DiRocco, Deb Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Craig
+George, Tracey Goldstein, Jeff Harris, Jason Herreman, Markus Horning, John
+Jansen, Shawn Johnson, Charles Littnan, Lloyd Lowry, Brett McClintock, Erin
+Moreland, Aaron Morris, Mark Nelson, Justin Olnes, Lorrie Rea, Bob Shears, Gay
+Sheffield, Kayla Scheimreif, Brent Stewart, Alexy Trukhin, Dave Withrow, and
 Heather Ziel. We also appreciate the commitment to science and safety by all
-officers and crew of the NOAA ship \emph{Oscar Dyson}, the NOAA ship \emph{MacArthur II}, and
-the RV \emph{Thomas G. Thompson}.
+officers and crew of the NOAA ship \emph{Oscar Dyson}, the NOAA ship \emph{MacArthur II},
+the \emph{MV Tayfun}, and the RV \emph{Thomas G. Thompson}.
 
@@
 Sea were done in collaboration with Russian colleagues and North Pacific
-Wildlife.
+Wildlife Consulting, LLC.
 

### 2024-09-22 analysis/paper/London_HauloutBehavior.Rmd

--- 2024-08-19 analysis/paper/London_HauloutBehavior.Rmd
@@
   accurately estimate abundance from aerial survey counts of ice-associated
-  seals (i.e., to correct for the proportion of animals that are in the water
-  while surveys are conducted). In addition to providing essential data for
-  survey 'availability' calculations, our analysis also provides insights into
-  the seasonal timing and environmental factors affecting haul-out behavior by
-  ice-associated seals. We specifically focused on bearded (_Erignathus
-  barbatus_), ribbon (_Histriophoca fasciata_), and spotted seals (_Phoca
-  largha_) in the Bering and Chukchi seas. Because ringed seals (_Phoca (pusa)
-  hispida_) can be out of the water but hidden from view in snow lairs analysis
-  of their 'availability' to surveys requires special consideration;
-  therefore,they were not included in this analysis. Using generalized linear
+  seals (i.e., to correct for the proportion of animals that are in the water,
+  and so are not counted, while surveys are conducted). In addition to providing
+  essential data for survey 'availability' calculations, our analysis also
+  provides insights into the seasonal timing and environmental factors affecting
+  haul-out behavior by ice-associated seals. We specifically focused on bearded
+  (_Erignathus barbatus_), ribbon (_Histriophoca fasciata_), and spotted seals
+  (_Phoca largha_) in the Bering and Chukchi seas. Because ringed seals (_Phoca
+  (pusa) hispida_) can be out of the water but hidden from view in snow lairs
+  analysis of their 'availability' to surveys requires special consideration;
+  therefore, they were not included in this analysis. Using generalized linear
   mixed pseudo-models to properly account for temporal autocorrelation, we fit
@@
   The timing and frequency of haul-out events also varied based on species and
-  age-sex class. For ribbon and spotted seals, models with year effects were
+  age and sex class. For ribbon and spotted seals, models with year effects were
   highly supported, indicating that the timing and magnitude of haul-out
@@
 Knowledge of haul-out patterns is not only important for understanding natural
-history and ecology, but also for developing "availability" correction factors
+history and ecology, but also for developing 'availability' correction factors
 for aerial surveys. Specifically, researchers need to know the fraction of seals
@@
 investigate the impact of weather variables. Such variables have been
-shown to influence walrus haul-out behavior [@udevitz2009] and we expect weather
+shown to influence walrus haul-out behavior [@udevitz2009] and we expected weather
 conditions to also influence seal haul-out behavior and including them within
@@
 capture and bio-logger attachment methods (see publications listed in
-Supplemental Material, [S1](#s1)). NOAA-led research was conducted under the
+Supplemental Information Table [S1](#s1)). NOAA-led research was conducted under the
 authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
@@
 [@verhoef2010a] differ between models [@teneyck2018]. After fitting GLMPM
-models, we instead used "type III" $F$-tests to calculate $p$-values
+models, we instead used 'type III' $F$-tests to calculate $p$-values
 [@verhoef2010a] to evaluate model performance and important terms. We also
@@
 lower sample size for adult and young-of-year age classes. As such, results are
-shown for all ages (Figure \@ref(fig:beardedHOCal); see also
+shown for all ages (Figure \@ref(fig:beardedHOCal); see also Supplemental Information Fig. 
 \@ref(fig:beardedPredSE)). Additionally, after approximately 9 months, 7 devices
@@
 April that peaks in late May for subadults and in early June for adults (Figure
-\@ref(fig:ribbonHOCal); see also \@ref(fig:ribbonPredSE)). There is an apparent
-seasonal progression with subadults hauling out earlier in the season followed
-by adult males and, then, adult females. Haul-out behavior was clearly centered
-around local solar noon and expanded to other hours later in the spring as seals
-entered their molting period. Subadults showed an earlier start and more intense
-haul-out activity in April and May. The young-of-the-year records began after
-weaning and the model predictions seemed to indicate development of in-water
-activities (e.g. swimming, foraging) in May and, like adults, haul-out behavior
-was centered around solar noon. Adult females had a more protracted
-haul-out season compared to males, and more time was spent hauled out in June
-compared to adult males and subadults.
+\@ref(fig:ribbonHOCal); see also Supplemental Information Fig.
+\@ref(fig:ribbonPredSE)). There is an apparent seasonal progression with
+subadults hauling out earlier in the season followed by adult males and, then,
+adult females. Haul-out behavior was clearly centered around local solar noon
+and expanded to other hours later in the spring as seals entered their molting
+period. Subadults showed an earlier start and more intense haul-out activity in
+April and May. The young-of-the-year records began after weaning and the model
+predictions seemed to indicate development of in-water activities (e.g.
+swimming, foraging) in May and, like adults, haul-out behavior was centered
+around solar noon. Adult females had a more protracted haul-out season compared
+to males, and more time was spent hauled out in June compared to adult males and
+subadults.
 
@@
 that was less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal);
-see also \@ref(fig:spottedPredSE)). Adults of both sexes spent considerable time
-in April, May and June hauled out. Adult spotted seal males had a more
-protracted haul-out season compared to females, and more time out of the water
-in June (Figure \@ref(fig:spottedHOCal)). This likely reflects the triad
-behavior in spotted seals when suitor males haul out with nursing females. As
-with ribbon seals, the young-of-the-year records began after weaning and the
-model predictions reflected development of in-water activities (e.g.
-swimming, foraging) in May and haul-out behavior centered around solar noon.
+see also Supplemental Information Fig.\@ref(fig:spottedPredSE)). Adults of both
+sexes spent considerable time in April, May and June hauled out. Adult spotted
+seal males had a more protracted haul-out season compared to females, and more
+time out of the water in June (Figure \@ref(fig:spottedHOCal)). This likely
+reflects the triad behavior in spotted seals when suitor males haul out with
+nursing females. As with ribbon seals, the young-of-the-year records began after
+weaning and the model predictions reflected development of in-water activities
+(e.g. swimming, foraging) in May and haul-out behavior centered around solar
+noon.
 
@@
 patterns, uncovered significant contributions for linear and quadratic
-interactions between day and year for only spotted seals (day:year,
+interactions between day and year for only spotted seals (_day:year_,
 $F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$
 = `r spotted_year_type3$day_year$F.value`; $p$ =
-`r spotted_year_type3$day_year$Prob.F`; day^2^:year,
+`r spotted_year_type3$day_year$Prob.F`; _day^2^:year_,
 $F_{`r paste(spotted_year_type3$day_2_year$Num.df, spotted_year_type3$day_2_year$Den.df, sep=",")`}$
@@
 `r spotted_year_type3$day_2_year$Prob.F`). Ribbon seals showed no significant
-contribution for interactions between day and year (day:year,
+contribution for interactions between day and year (_day:year_,
 $F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$
 = `r ribbon_year_type3$day_year$F.value`; $p$ =
-`r ribbon_year_type3$day_year$Prob.F`; day^2^:year,
+`r ribbon_year_type3$day_year$Prob.F`; _day^2^:year_,
 $F_{`r paste(ribbon_year_type3$day_2_year$Num.df, ribbon_year_type3$day_2_year$Den.df, sep=",")`}$
@@
 showed a negative trend (peak haul-out occurs later in years with less
-sea ice) with a significant relationship for adult female spotted seals 
+sea ice) but neither with a significant relationship  
 (spotted seal adult female: $R^{2}$ =
@@
 [-@udevitz2009]) and a few select studies of ice-associated seals
-[@hamilton2018; @perry2017]. In our study, ribbon seal haul-out behavior seemed to be the most
+[@hamilton2018; @perry2017]. In our study, ribbon seal haul-out behavior was notably
 influenced by weather, with wind, temperature, and barometric pressure all being
@@
 biology’ of Arctic seals. The deployment of bio-logging devices used in this
-study were often done in collaboration with Alaska Native seal hunters and the
+study were often done in collaboration with Alaska Native seal hunters and wiht the
 approval of
-their communities. We would like to especially acknowledge the communities of
+their communities. We would like to especially acknowledge the Alaska communities of
 Kotzebue, Koyuk, Nome, Nuiqsut, Scammon Bay, St. Michael, Utqiaġvik, and Ulguniq
@@
 St. Michael; Billy Adams, James Aiken, Tim Aiken, Howard Kittick,
-Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, and Joe Skin from Utqiaġvik,
-Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, and
-Fred Rexford from Ulguniq.
+Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Bobby Sarren, and Joe Skin from Utqiaġvik; 
+Mary Ellen Ahmaogak, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, and
+Fred Rexford from Ulguniq. We would also like to acknowledge the support of the
+Ice Seal Committee.
 
 We are grateful for the assistance in catching and sampling seals by Ryan Adam,
-James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov,
-Cynthia Christman, Sarah Coburn, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
-Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey Goldstein, Jeff
-Harris, Jason Herreman, Markus Horning, John Jansen, Shawn Johnson, Charles
-Littnan, Lloyd Lowry, Brett McClintock, Erin Moreland, Mark Nelson, Justin
-Olnes, Lorrie Rea, Bob Shears, Gay Sheffield, Brent Stewart, Dave Withrow, and
+James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Anna Bryan,
+Vladamir Burkanov, Cynthia Christman, Sarah Coburn, Shawn Dahle, Rob Delong,
+Stacy DiRocco, Deb Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Craig
+George, Tracey Goldstein, Jeff Harris, Jason Herreman, Markus Horning, John
+Jansen, Shawn Johnson, Charles Littnan, Lloyd Lowry, Brett McClintock, Erin
+Moreland, Aaron Morris, Mark Nelson, Justin Olnes, Lorrie Rea, Bob Shears, Gay
+Sheffield, Kayla Scheimreif, Brent Stewart, Alexy Trukhin, Dave Withrow, and
 Heather Ziel. We also appreciate the commitment to science and safety by all
-officers and crew of the NOAA ship _Oscar Dyson_, the NOAA ship _MacArthur II_, and
-the RV _Thomas G. Thompson_.
+officers and crew of the NOAA ship _Oscar Dyson_, the NOAA ship _MacArthur II_,
+the _MV Tayfun_, and the RV _Thomas G. Thompson_.
 
